### 2017-03-02 tssim.tex
started writing

\documentclass{bioinfo}
\copyrightyear{2017} \pubyear{2017}

\newcommand{\plr}[1]{{\color{blue}\it #1}}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Application Notes}

\begin{document}
\firstpage{1}

% Bioinformatics:
% Application Notes (up to 2 pages; this is approx. 1300 words or 1000 words plus
% one figure) Applications Notes are short descriptions of novel software or new
% algorithm implementations, databases and network services (web servers, and
% interfaces). Software or data must be freely available to non-commercial users.
% Availability and Implementation must be clearly stated in the article. Authors
% must also ensure that the software is available for a full TWO YEARS following
% publication. Web services must not require mandatory registration by the user.
% Additional Supplementary data can be published online-only by the journal. This
% supplementary material should be referred to in the abstract of the Application
% Note. If describing software, the software should run under nearly all
% conditions on a wide range of machines. Web servers should not be browser
% specific. Application Notes must not describe trivial utilities, nor involve
% significant investment of time for the user to install. The name of the
% application should be included in the title.

\subtitle{Subject Section}

\title[Recording the pedigree]{ftprime: Recording the pedigree: efficient simulation of whole genomes}
\author{Corresponding Author\,$^{\text{\sfb 1,}*}$, Co-Author\,$^{\text{\sfb 2}}$ and Co-Author\,$^{\text{\sfb 2,}*}$}
\address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country and \\
$^{\text{\sf 2}}$Department, Institution, City, Post Code,
Country.}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{
    To use genomic data for inference and prediction it is often necessary to obtain whole-genome information
    from individual-based simulations,
    but the computational burden of endowing each simulated individual with an entire genome can be substantial.
    In this note we describe how to both (a) dramatically reduce this burden and 
    (b) efficiently record the entire history of the population.
    We do this by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording the entire history of genetic inheritance in an efficient data structure,
    on which neutral mutations can be quickly placed afterwards.
    \plr{make more clear data structure was already developed? refer to 'tree sequence' by name?}
    The algorithm is implemented in python,
    and is designed to be easily used by any forwards-time simulation software.
\textbf{Availability and implementation:} \\
\textbf{Contact:} \\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}

\maketitle

\section{Introduction}

% OUTLINE
% 1. motivate need for whole-genome fwd-time simulations; point out that we only recently have the computing power to do this
% 2. explain ARG: explain that for forwards-time only need selected loci as by defn all others can be put on afterwards
% 3. back-of-the-envelope calculation to compare cost of tracking whole genomes versus putting mutations on ARG
% 4. review something about your efficient methods for storing/traversing tree sequence
% 5. write down algorithm used in ftprime to produce msprime-ready coalescence records
% 6. comparison of speed in really simple case (but whole-chromosome) using simupop in both cases

In recent years the increase in computing power --
gradual only in comparison to the increase in our ability to sequence genomes --
has made it possible to simulate the evolution of whole genomes of realistically-sized populations.
This major milestone promises to remove the field's reliance on approximations of unknown applicability.
Thus far, much of our understanding of how genomes evolve
comes from models of one or a few linked loci \citep{ewens},
and/or results derived under the assumption of neutrality.
These are certainly important, but researchers today have widely differing views
on the practical importance of widespread selection \citep{encode,graur,backgroundsel}.
Thus far what little analytical progress has been made on models of ubiquitous selection
are hard to check due to a lack of realistic simulation.

Simulations of organisms endowed with functional, varying genomes
interacting with each other and spatially and temporally varying environments
will not only allow us to develop and test our understanding of evolution,
but also to produce quantitative predictions of population dynamics --
for instance, the spread of insecticide resistance and behavior modification of mosquitos.

The most commonly used simulation methods to date rely on neutrality,
as this assumption coupled with random mating allows the use of coalescent methods \citep{wakeley}.
These drastically reduce the amount of required computation,
because they exploit a time-reversal duality of such models
to \emph{only} simulate the portion of history that determines the genomes in the modern population.
How does this work?
Suppose that to the \emph{population pedigree} --
the entire history of parent-offspring relationships of an entire population going back to a remote time --
we add information encoding the genetic outcomes of each ancestral meiosis --
who inherited which parts of which parental chromosomes.
This embellished graph is known as the \emph{ancestral recombination graph}, or ARG \citep{griffiths}.
Combined with ancestral genotypes and the origins of new mutations,
it completely specifies the genomic sequence of any individual in the population at any time.
However, much less than the entire ARG is needed to specify relationships between any given set of samples --
only those portions of it from which those samples have actually inherited, back to their most recent common ancestors.
For instance, \plr{add diagram of simple example?}
The assumptions of coalescent theory -- random mating and neutrality --
imply that the stochastic law of this random set forms a Markov process looked at \emph{backwards} in time,
and can hence be simulated without reference to the unnecessary remainder.
Although this point was in principle known since \citet{griffithsmarjoram},
only recent algorithmic advances made it possible to actually simulate the process correctly,
without approximation, across whole genomes \citep{msprime}.

These efficient methods are not available if either key assumption of coalescent theory --
neutrality or random mating -- are broken.
(Extensions can be made for only a few loci under selection \citep{hudson,krone}.)
Forwards-time simulation, however, is burdened by recording and passing on entire genomes,
which typically have tens to hundreds of millions of varying sites.

\begin{methods}
\section{Methods and Implementation}

\end{methods}
\section{Results and Discussion}

\section{Conclusion}

\section*{Acknowledgements}

\section*{Funding}

\bibliographystyle{natbib}
\bibliography{refs.bib}

\end{document}

### 2017-03-03 tssim.tex
algorithm, begun

@@
 (Extensions can be made for only a few loci under selection \citep{hudson,krone}.)
-Forwards-time simulation, however, is burdened by recording and passing on entire genomes,
+Forwards-time, individual-based simulation, however, is burdened by recording and passing on entire genomes,
 which typically have tens to hundreds of millions of varying sites.
+Such simulations must explicitly record any non-neutral genetic variants,
+because these affect the demographic process.
+However, neutral variants, by definition, do not, 
+and so if the demographic history is recorded
+they can be simulated afterwards.
+
+Wrap-up paragraph.  
+This is what we do, which lets us both record history
+and be more efficient.
 
@@
 \section{Methods and Implementation}
+
+Since msprime provides a number of tools for manipulating tree sequences,
+adding mutations,
+and producing output (e.g., VCF files)
+we need only determine a method for properly recording history and passing it to msprime.
+
+\subsection{msprime: Tree sequences and coalescence records}
+
+First we recall the data storage methods used by msprime \citep[see][for more details]{msprime}.
+
+To record the history and identity of a set of sampled genomes,
+it is only necessary to record commmon ancestry (or, \emph{coalescence}) events --
+most recent common ancestors of distinct segments of sampled genomes --
+and the identity and location of any mutations that these most recent common ancestors carry
+that are not shared by any other recorded ancestor.
+\plr{(not quite right)}
+
+The topology of the ARG
+is recorded by msprime as a list of Coalescence Records, which are tuples of the form
+\texttt{(left,right,parent,children)},
+which indicates that the individuals listed in \texttt{children}
+have inherited the segment of genome between \texttt{left} and \texttt{right}
+from the ancestor \texttt{parent}.
+Individuals are identified by integers,
+and there is also a table recording the \emph{birth time} of each individual.
+
+For a set of coalescence records to be consistent,
+it is required that:
+\begin{enumerate}
+    \item The birth time of any \texttt{children} must be strictly after the birth time of the \texttt{parent}.
+    \item The set of intervals on which individual $a$ is a child must be disjoint, for every $a$.
+    \item The set of intervals on which individual $a$ is a parent must be disjoint, for every $a$.
+\end{enumerate}
+These ensure that the coalescence records unambiguously specify a subset of the genealogy at any location on the genome,
+but do not ensure that the entire history of the samples is specified.
+Furthermore, to allow for efficient algorithms, msprime also requires that
+\begin{enumerate} \setcounter{enumi}{3}
+    \item The samples must be numbered \texttt{0,...,n-1}, and the smallest non-sampled label must be an internal node.
+        \plr{perhaps remove this one}
+    \item The list of \texttt{children} in a coalescence record must be sorted.
+    \item Records must be sorted in nondecreasing order by birth time of the \texttt{parent}.
+    \item Birth times must be strictly greater than zero.
+\end{enumerate}
+
+\subsection{Recording history}
+
+Since we do not know from which ancestors
+the eventual genetic samples will inherit,
+to record the history of the samples
+we must record the results of all reproduction events -- 
+the entire ARG.
+To do this, and conform with the format above, we maintain
+an ordered dictionary \texttt{arg},
+with one entry for each parent,
+whose entries are of the form
+\texttt{id:(time,records)},
+where \texttt{id} is the integer ID of the parent,
+\texttt{time} is the birth time of the parent,
+and \texttt{records} is a list of nonoverlapping coalescence records, sorted by left endpoint.
+We update this structure each time a new individual is born.
+Suppose that individual $I$ is born at time $t$ to parents $p_0$ and $p_1$,
+with recombination breakpoints $0=x_0 < x_1 < x_2 < \cdots < x_n \le 1$,
+so that $I$ inherits the genomic segment $[0,x_1)$ from $p_0$,
+the segment $[x_1,x_2)$ from $p_1$, and so forth (here $n \ge 1$).
+When this occurs, we:
+\begin{enumerate}
+    \item append a new entry \texttt{I:(t,[])} to the list \texttt{arg}; and
+    \item for each $1 \le k \le n$, merge the new record \texttt{(x[k-1],x[k],p[k mod 2],(I,))}
+        into the existing records in `arg[p[k mod 2]]`.
+\end{enumerate}
+We need to merge these records
+so as to retain sortedness described above.
 

### 2017-03-05 tssim.tex
simupop simulation script

@@
     \item The birth time of any \texttt{children} must be strictly after the birth time of the \texttt{parent}.
-    \item The set of intervals on which individual $a$ is a child must be disjoint, for every $a$.
-    \item The set of intervals on which individual $a$ is a parent must be disjoint, for every $a$.
+    \item The set of intervals on which individual \texttt{a} is a child must be disjoint, for every \texttt{a}.
+    \item The set of intervals on which individual \texttt{a} is a parent must be disjoint, for every \texttt{a}.
 \end{enumerate}
@@
 \subsection{Recording history}
+
+Simulation software often gives a unique identifier to each \emph{diploid} individual,
+but the natural exchangeable unit of consideration for the ARG (and msprime) is gamete --
+so, each diploid individual is the union of two haploid individuals: 
+one consisting of the set of chromosomes inherited maternally, the other paternally.
+To translate between the two,
+we adopt the convention that the maternally inherited set of chromosomes of diploid individual $k$
+is given haploid ID $2k$, and the paternally inherited set $2k+1$.
+ID numbers referred to below are haploid IDs.
 
@@
 Suppose that individual $I$ is born at time $t$ to parents $p_0$ and $p_1$,
-with recombination breakpoints $0=x_0 < x_1 < x_2 < \cdots < x_n \le 1$,
+with recombination breakpoints $0=x_0 < x_1 < x_2 < \cdots < x_n \le L$,
 so that $I$ inherits the genomic segment $[0,x_1)$ from $p_0$,
-the segment $[x_1,x_2)$ from $p_1$, and so forth (here $n \ge 1$).
+the segment $[x_1,x_2)$ from $p_1$, and so forth (here $n \ge 1$),
+and where $L$ is the total length of the chromosome.
 When this occurs, we:
@@
     \item append a new entry \texttt{I:(t,[])} to the list \texttt{arg}; and
-    \item for each $1 \le k \le n$, merge the new record \texttt{(x[k-1],x[k],p[k mod 2],(I,))}
+    \item for each $1 \le k \le n$, \texttt{merge} the new record \texttt{(x[k-1],x[k],p[k mod 2],(I,))}
         into the existing records in `arg[p[k mod 2]]`.
 \end{enumerate}
-We need to merge these records
-so as to retain sortedness described above.
+The second step requires we merge the new record with existing ones
+so as to retain the criteria described above.
+To \texttt{merge} a record saying that individual \texttt{I} also inherits from this \texttt{parent} on interval \texttt{[a,b)}
+into an existing set of records of the form \texttt{(left,right,parent,children)},
+we need to 
+\begin{enumerate}
+    \item split any record for which \texttt{left<a<right} or \texttt{left<b<right} into two at the corresponding endpoint, 
+    \item and then add \texttt{I} to \texttt{children} for any intervals overlapping \texttt{[a,b)}.
+\end{enumerate}
+This is done easily, maintaining the fact that records are not overlapping and are maintained in sorted order.
+\plr{could alternatively write out algorithm but it is a bit nitpicky}
+
+% To do this, to \texttt{merge} a record \texttt{(new.left,new.right,parent,new.children)}
+% into an existing sorted list of \texttt{records},
+% we do the following.
+% \begin{enumerate} \setcounter{enumi}{-1}
+%     \item Initialize \texttt{here=new.left} and \texttt{k=0}.
+%     \item While \texttt{here<right}, let the next entry of \texttt{records} 
+%         (ordered by left endpoint) be \texttt{(left,right,parent,children)}.
+%         \begin{enumerate}
+%             \item If \texttt{right<here}, go to the next record.
+%             \item If \texttt{here<left}: 
+%                 \begin{enumerate}
+%                     \item insert a new record \texttt{(here,min(new.right,left),parent,new.children)} before the current record
+%                     \item and set \texttt{here=min(new.right,left)}.
+%                 \end{enumerate}
+%             \item otherwise, if  TOO MANY STEPS
+%         \end{enumerate}
+% \end{enumerate}
 

### 2017-03-05 sims/run-sim.py

@@ -1,0 +1,208 @@
+#!/usr/bin/env python3.5
+description = '''
+Simulate a chromosome with scattered selected loci using simuPOP,
+postprocessed with msprime to add neutral loci.
+'''
+
+import gzip
+import sys
+from optparse import OptionParser
+import math
+import time
+import random
+from ftprime import RecombCollector
+import msprime
+
+parser = OptionParser(description=description)
+parser.add_option("-T","--generations",dest="generations",help="number of generations to run for")
+parser.add_option("-N","--popsize",dest="popsize",help="size of each subpopulation",default=100)
+parser.add_option("-L","--length",dest="length",help="number of bp in the chromosome",default=1e4)
+parser.add_option("-l","--nloci",dest="nloci",help="number of selected loci",default=20)
+parser.add_option("-m","--migr",dest="migr",help="migration proportion between adjacent populations",default=.01)
+parser.add_option("-u","--sel_mut_rate",dest="sel_mut_rate",help="mutation rate of selected alleles",default=1e-7)
+parser.add_option("-r","--recomb_rate",dest="recomb_rate",help="recombination rate",default=2.5e-8)
+parser.add_option("-a","--gamma_alpha",dest="gamma_alpha",help="alpha parameter in gamma distributed selection coefficients",default=.23)
+parser.add_option("-b","--gamma_beta",dest="gamma_beta",help="beta parameter in gamma distributed selection coefficients",default=5.34)
+parser.add_option("-k","--nsamples",dest="nsamples",help="number of *diploid* samples, total")
+parser.add_option("-A","--ancestor_age",dest="ancestor_age",help="time to ancestor above beginning of sim")
+parser.add_option("-U","--mut_rate",dest="mut_rate",help="neutral mutation rate",default=1e-7)
+parser.add_option("-t","--treefile",dest="treefile",help="name of output file for trees (default: not output)",default=None)
+
+parser.add_option("-o","--outfile",dest="outfile",help="name of output VCF file (default: not output)",default=None)
+parser.add_option("-g","--logfile",dest="logfile",help="name of log file (or '-' for stdout)",default="-")
[…]

### 2017-03-05 sims/run-simupop.py

@@ -1,0 +1,172 @@
+#!/usr/bin/env python3.5
+description = '''
+Simulate a chromosome with scattered selected loci using simuPOP.
+'''
+
+import gzip
+import sys
+from argparse import ArgumentParser
+import math
+import time
+import random
+
+parser = ArgumentParser(description=description)
+parser.add_argument("-T","--generations", dest="generations", type=int,
+        help="number of generations to run for")
+parser.add_argument("-N","--popsize", dest="popsize", type=int,
+        help="size of the population", default=100)
+parser.add_argument("-r","--recomb_rate", dest="recomb_rate", type=float,
+        help="recombination rate", default=2.5e-8)
+parser.add_argument("-L","--length", dest="chrom_length", type=int,
+        help="number of bp in the chromosome", default=100)
+parser.add_argument("-U","--neut_mut_rate", dest="neut_mut_rate", type=float,
+        help="neutral mutation rate", default=1e-7)
+parser.add_argument("-l","--nselloci", dest="nselloci", type=int,
+        help="number of selected loci", default=20)
+parser.add_argument("-u","--sel_mut_rate", dest="sel_mut_rate", type=float,
+        help="mutation rate of selected alleles", default=1e-7)
+parser.add_argument("-a","--gamma_alpha", dest="gamma_alpha", type=float,
+        help="alpha parameter in gamma distributed selection coefficients", default=.23)
+parser.add_argument("-b","--gamma_beta", dest="gamma_beta", type=float, 
+        help="beta parameter in gamma distributed selection coefficients", default=5.34)
+parser.add_argument("-k","--nsamples", dest="nsamples", type=int,
+        help="number of *diploid* samples, total")
+parser.add_argument("-o","--outfile", dest="outfile", type=str,
+        help="name of output PED file (default: not output)", default=None)
+parser.add_argument("-g","--logfile", dest="logfile", type=str,
+        help="name of log file (or '-' for stdout)", default="-")
+parser.add_argument("-s","--selloci_file", dest="selloci_file", type=str,
+        help="name of file to output selected locus information", default="sel_loci.txt")
+args = parser.parse_args()
+
+if args.generations is None:
+    parser.print_help()
+    sys.exit()
+
[…]

### 2017-03-05 sims/run-sim.py
and version using msprime

@@ -6,32 +6,53 @@ postprocessed with msprime to add neutral loci.
 
 import gzip
 import sys
-from optparse import OptionParser
+from argparse import ArgumentParser
 import math
 import time
 import random
 from ftprime import RecombCollector
 import msprime
 
-parser = OptionParser(description=description)
-parser.add_option("-T","--generations",dest="generations",help="number of generations to run for")
-parser.add_option("-N","--popsize",dest="popsize",help="size of each subpopulation",default=100)
-parser.add_option("-L","--length",dest="length",help="number of bp in the chromosome",default=1e4)
-parser.add_option("-l","--nloci",dest="nloci",help="number of selected loci",default=20)
-parser.add_option("-m","--migr",dest="migr",help="migration proportion between adjacent populations",default=.01)
-parser.add_option("-u","--sel_mut_rate",dest="sel_mut_rate",help="mutation rate of selected alleles",default=1e-7)
-parser.add_option("-r","--recomb_rate",dest="recomb_rate",help="recombination rate",default=2.5e-8)
-parser.add_option("-a","--gamma_alpha",dest="gamma_alpha",help="alpha parameter in gamma distributed selection coefficients",default=.23)
-parser.add_option("-b","--gamma_beta",dest="gamma_beta",help="beta parameter in gamma distributed selection coefficients",default=5.34)
-parser.add_option("-k","--nsamples",dest="nsamples",help="number of *diploid* samples, total")
-parser.add_option("-A","--ancestor_age",dest="ancestor_age",help="time to ancestor above beginning of sim")
-parser.add_option("-U","--mut_rate",dest="mut_rate",help="neutral mutation rate",default=1e-7)
-parser.add_option("-t","--treefile",dest="treefile",help="name of output file for trees (default: not output)",default=None)
+parser = ArgumentParser(description=description)
+parser.add_argument("-T","--generations", dest="generations", type=int,
+        help="number of generations to run for")
+parser.add_argument("-N","--popsize", dest="popsize", type=int,
+        help="size of the population", default=100)
+parser.add_argument("-r","--recomb_rate", dest="recomb_rate", type=float,
+        help="recombination rate", default=2.5e-8)
[…]

### 2017-03-05 sims/run-simupop.py

@@ -37,6 +37,7 @@ parser.add_argument("-g","--logfile", dest="logfile", type=str,
         help="name of log file (or '-' for stdout)", default="-")
 parser.add_argument("-s","--selloci_file", dest="selloci_file", type=str,
         help="name of file to output selected locus information", default="sel_loci.txt")
+
 args = parser.parse_args()
 
 if args.generations is None:

### 2017-03-05 sims/run-sim.py
notes for simulation parameters

@@ -19,7 +19,7 @@ parser.add_argument("-T","--generations", dest="generations", type=int,
 parser.add_argument("-N","--popsize", dest="popsize", type=int,
         help="size of the population", default=100)
 parser.add_argument("-r","--recomb_rate", dest="recomb_rate", type=float,
-        help="recombination rate", default=2.5e-8)
+        help="recombination rate", default=1e-7)
 parser.add_argument("-L","--length", dest="chrom_length", type=int,
         help="number of bp in the chromosome", default=100)
 parser.add_argument("-U","--neut_mut_rate", dest="neut_mut_rate", type=float,
@@ -33,7 +33,7 @@ parser.add_argument("-a","--gamma_alpha", dest="gamma_alpha", type=float,
 parser.add_argument("-b","--gamma_beta", dest="gamma_beta", type=float, 
         help="beta parameter in gamma distributed selection coefficients", default=5.34)
 parser.add_argument("-k","--nsamples", dest="nsamples", type=int,
-        help="number of *diploid* samples, total")
+        help="number of *diploid* samples, total", default=100)
 parser.add_argument("-o","--outfile", dest="outfile", type=str,
         help="name of output PED file (default: not output)", default=None)
 parser.add_argument("-g","--logfile", dest="logfile", type=str,
@@ -50,9 +50,6 @@ args = parser.parse_args()
 if args.generations is None:
     parser.print_help()
     sys.exit()
-
-if args.nsamples is None:
-    args.nsamples = args.popsize
 
 import simuOpt
 simuOpt.setOptions(alleleType='mutant')

### 2017-03-05 sims/run-simupop.py

@@ -16,7 +16,7 @@ parser.add_argument("-T","--generations", dest="generations", type=int,
 parser.add_argument("-N","--popsize", dest="popsize", type=int,
         help="size of the population", default=100)
 parser.add_argument("-r","--recomb_rate", dest="recomb_rate", type=float,
-        help="recombination rate", default=2.5e-8)
+        help="recombination rate", default=1e-7)
 parser.add_argument("-L","--length", dest="chrom_length", type=int,
         help="number of bp in the chromosome", default=100)
 parser.add_argument("-U","--neut_mut_rate", dest="neut_mut_rate", type=float,
@@ -30,7 +30,7 @@ parser.add_argument("-a","--gamma_alpha", dest="gamma_alpha", type=float,
 parser.add_argument("-b","--gamma_beta", dest="gamma_beta", type=float, 
         help="beta parameter in gamma distributed selection coefficients", default=5.34)
 parser.add_argument("-k","--nsamples", dest="nsamples", type=int,
-        help="number of *diploid* samples, total")
+        help="number of *diploid* samples, total", default=100)
 parser.add_argument("-o","--outfile", dest="outfile", type=str,
         help="name of output PED file (default: not output)", default=None)
 parser.add_argument("-g","--logfile", dest="logfile", type=str,
@@ -43,9 +43,6 @@ args = parser.parse_args()
 if args.generations is None:
     parser.print_help()
     sys.exit()
-
-if args.nsamples is None:
-    args.nsamples = args.popsize
 
 import simuOpt
 simuOpt.setOptions(alleleType='mutant')

### 2017-03-06 tssim.tex
a few tweaks

@@
     from individual-based simulations,
-    but the computational burden of endowing each simulated individual with an entire genome can be substantial.
+    but the computational burden of tracking the genome of each simulated individual can be substantial.
     In this note we describe how to both (a) dramatically reduce this burden and 
@@
     We do this by simulating only those loci that may affect reproduction (those having non-neutral variants),
-    and recording the entire history of genetic inheritance in an efficient data structure,
+    and recording the entire history of genetic inheritance in an efficient representation of the ancestral recombination graph,
     on which neutral mutations can be quickly placed afterwards.
@@
 \maketitle
+
+Other title ideas: (I think it has to start with the `software name', for Bioinformatics applications notes anyhow)
+
+Coalescence theory in forwards time
+
+The ancestral recombination graph from individual-based simulations
+
+History, and whole genomes too
 
@@
 has made it possible to simulate the evolution of whole genomes of realistically-sized populations.
-This major milestone promises to remove the field's reliance on approximations of unknown applicability.
-Thus far, much of our understanding of how genomes evolve
+This major milestone allows researchers to investigate models of much greater complexity
+than was previously acheivable.
+Thus far, much of our understanding of how populations evolve
 comes from models of one or a few linked loci \citep{ewens},
@@
 on the practical importance of widespread selection \citep{encode,graur,backgroundsel}.
-Thus far what little analytical progress has been made on models of ubiquitous selection
-are hard to check due to a lack of realistic simulation.
-
-Simulations of organisms endowed with functional, varying genomes
+% Thus far what little analytical progress has been made understanding
+% how potentially many functional genomic variants might evolve in ecologically realistic environments
+% are hard to check due to a lack of realistic simulation methods.
+Simulations of organisms carrying with functional, varying genomes
 interacting with each other and spatially and temporally varying environments
 will not only allow us to develop and test our understanding of evolution,
-but also to produce quantitative predictions of population dynamics --
+but also to produce quantitative predictions of eco-evolutionary dynamics --
 for instance, the spread of insecticide resistance and behavior modification of mosquitos.

### 2017-08-26 forwards_paper.tex
moved over writeup from ftprime; merged here

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{enumitem}
\bibliographystyle{plainnat}

\include{macros}

\newcommand{\msprime}{\texttt{msprime}}

\usepackage{color}
\newcommand{\plr}[1]{{\em \color{blue} #1}}

\begin{document}

\title{Efficient forwards-time simulation with ancestral recombination graphs}
\author{Jerome Kelleher, 
        Kevin R. Thornton,
        Jaime Ashander, and
        Peter L. Ralph}
\maketitle

\begin{abstract}
    To use genomic data for inference and prediction it is often necessary to obtain whole-genome information
    from individual-based simulations,
    but the computational burden of tracking the genome of each simulated individual can be substantial.
    In this note we describe how to both (a) dramatically reduce this burden and 
    (b) efficiently record the entire history of the population.
    We do this by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording the entire history of genetic inheritance in an efficient representation of the ancestral recombination graph,
    on which neutral mutations can be quickly placed afterwards.
    \plr{make more clear data structure was already developed? refer to 'tree sequence' by name?}
    The algorithm is implemented in python,
    and is designed to be easily used by any forwards-time simulation software.

    Coalescent simulations are very helpful but require random mating and neutrality.
    For continuous space, polygenic selection, or detailed dissection of life history, 
    we must use forwards-time, individual-based simulation.
    These are much slower due in part to carrying around neutral genotypes irrelevant to the process.
    Here we show how to efficiently produce and store the entire history of ancestry and recombination
    (the ARG) from an individual-based simulation,
    on which neutral mutations can be placed afterwards.
    This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
\end{abstract}

OUTLINE
\begin{itemize}
    \item motivate need for whole-genome fwd-time simulations; point out that we only recently have the computing power to do this
    \item explain ARG: explain that for forwards-time only need selected loci as by defn all others can be put on afterwards
    \item back-of-the-envelope calculation to compare cost of tracking whole genomes versus putting mutations on ARG
    \item review something about your efficient methods for storing/traversing tree sequence
    \item write down algorithm used in ftprime to produce msprime-ready coalescence records
    \item describe simplify algorithm
    \item comparison of speed with simupop, fwdpp
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}

\plr{two main points below: can put down neutral mutations afterwards, and could be useful to compute stats even before that}

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g. $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of (a) random mating, and (b) neutrality.
\plr{Brief statement why this is.  Also include stationarity?}
Both assumptions can be side-stepped to a limited extent, and so coalescent methods are now commonly used to
simulate the results of population dynamics of a collection of randomly mating populations exchanging migrants,
having a small number of loci under selection.
Mapping results of such models onto real species can be challenging,
as these are often distributed across geographical space and may have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, allee effects, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
These considerations, and increasing computational power, have led to a resurgence of interest in forwards-time, individual-based simulations.

With modern computing power, pure demographic calculations are not a barrier,
even though biological population sizes are often above $10^6$,
and coalescent theory tells us that a population of size $N$ 
must be run for a multiple of $N$ generations to produce stable genetic patterns.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must somehow produce at the end data for each individual on a genomic scale.
As samples of most species genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simluation becomes quickly prohibitive.

However, it is thought that much of that variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
one can lay down neutral mutations on top of that history afterwards,
without loss of generality.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.

There is another compelling reason for recording the entire genealogical history.
In many applications, simulations are run for the purpose of comparing summary statistics of the resulting genomes
to those computed from real data,
with the purpose of inferring demographic history or identifying regions under selection.
However, for the same reason that neutral mutations can be laid down afterwards on the sequence of genealogical trees resulting from a simulation,
any summary statistic of genetic variation can be expressed as a statistic of the distribution of genealogical trees along the genome,
with additional noise introduced by mutations.
For instance, mean pairwise sequence diversity, $\theta$,
is most easily thought of as an estimator of $2 \mu t_{TMRCA}$, the mean time to most recent common ancestor
multiplied by twice the mutation rate,
with a relative error that is independent of the actual $t_{TMRCA}$.
As additional measurement noise makes simulation-based inference methods such as ABC or trained classifiers more diffcult,
it could be beneficial to compare the statistics observed in the data
to their expected values conditioned on the genealogical history,
as remaining mutational noise is not informative.
\plr{Say better.  A quick way to demonstrate this point?}

There is an obvious cost to recording all of history from a simulation,
rather than just the current state.
In this paper, we show how to use algorithmic tools and data structures developed for the coalescent simulator \msprime
to efficiently record, and later process, this history.

\plr{Quick review of the ARG and discussion of whether the ARG means the data structure or the reverse-time markov process (the former).}
Suppose that to the \emph{population pedigree} --
the entire history of parent-offspring relationships of an entire population going back to a remote time --
we add information encoding the genetic outcomes of each ancestral meiosis --
who inherited which parts of which parental chromosomes.
This embellished graph is known as the \emph{ancestral recombination graph}, or \emph{ARG} \citep{griffiths}.
Combined with ancestral genotypes and the origins of new mutations,
it completely specifies the genomic sequence of any individual in the population at any time.
However, much less than the entire ARG is needed to specify relationships between any given set of samples --
only those portions of it from which those samples have actually inherited, back to their most recent common ancestors.
The assumptions of coalescent theory -- random mating and neutrality --
imply that the stochastic law of this random set forms a Markov process looked at \emph{backwards} in time,
and can hence be simulated without reference to the unnecessary remainder.
Although this point was in principle known since \citet{griffithsmarjoram},
only recent algorithmic advances made it possible to actually simulate the process correctly,
without approximation, across whole chromosomes \citep{msprime}.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Methods}

\plr{Description of what we actually need to know in the end (the trees).}

\plr{Quick review of \msprime methods: sparse trees, tree differences.}

First we recall the data storage methods used by msprime \citep[see][for more details]{msprime}.

To record the history and identity of a set of sampled genomes,
it is only necessary to record commmon ancestry (or, \emph{coalescence}) events --
most recent common ancestors of distinct segments of sampled genomes --
and the identity and location of any mutations that these most recent common ancestors carry
that are not shared by any other recorded ancestor.
\plr{(not quite right)}

The topology of the ARG
is recorded by msprime as a list of Coalescence Records, which are tuples of the form
\texttt{(left,right,parent,children)},
which indicates that the individuals listed in \texttt{children}
have inherited the segment of genome between \texttt{left} and \texttt{right}
from the ancestor \texttt{parent}.
Individuals are identified by integers,
and there is also a table recording the \emph{birth time} of each individual.

For a set of coalescence records to be consistent,
it is required that:
\begin{enumerate}
    \item The birth time of any \texttt{children} must be strictly after the birth time of the \texttt{parent}.
    \item The set of intervals on which individual \texttt{a} is a child must be disjoint, for every \texttt{a}.
    \item The set of intervals on which individual \texttt{a} is a parent must be disjoint, for every \texttt{a}.
\end{enumerate}
These ensure that the coalescence records unambiguously specify a subset of the genealogy at any location on the genome,
but do not ensure that the entire history of the samples is specified.
Furthermore, to allow for efficient algorithms, msprime also requires that
\begin{enumerate} \setcounter{enumi}{3}
    \item The samples must be numbered \texttt{0,...,n-1}, and the smallest non-sampled label must be an internal node.
        \plr{perhaps remove this one}
    \item The list of \texttt{children} in a coalescence record must be sorted.
    \item Records must be sorted in nondecreasing order by birth time of the \texttt{parent}.
    \item Birth times must be strictly greater than zero.
\end{enumerate}

\plr{Specification for tables of nodes, edgesets, sites, and mutations (migrations?).}
\plr{Calling a collection of unordered tables a ``relaxed'' tree sequence
    but will make it clear that to do anything with these must sort first.}

\begin{itemize}
    \item[Nodes:]
        (flags, population, time)
    \item[Edgesets:]
        (left,  right,  parent,  children)
    \item[Sites:]
        (position,  ancestral\_state)
    \item[Mutations:]
        (site,  node,  derived\_state)
    \item \plr{Migrations?}
\end{itemize}

\plr{Steps to record information in forwards time.}
In the forwards simulation,
we record the results of each meiosis --
for each distinctly inherited segment of offspring chromosome,
we record the parental chromosome,
the left and right endpoints,
and the identities of any new mutations that have occurred on it.
This is recorded easily, without further processing,
in the relaxed tree sequence format.

Concretely, this is done as follows.
Suppose that the forwards simulation algorithm labels (haploid) individuals by integers,
which we call ``input labels'', 
to distinguish them from the ``node IDs'' given to these same individuals in the (output) tree sequence.
The algorithm maintains at all times a set of tables $(\nodes, \edgesets, \sites, \mutations)$
that record a relaxed tree sequence,
and an associative array $L$ that maps input labels to output node IDs, so that if $x$ is an input label,
then $L[x]$ is the corresponding output node ID.
We also always maintain $n$ to be the number of rows currently in the node table,
(so that with zero-indexed IDs, the next to be added will have node ID $n$),
$T_0$ to be the time of last simplification,
and $n_0$ the number of rows in the node table at that time.

Initially, we begin with $n$ and $n_0$ equal to the number of rows in the initial node table,
and $L[j] = i_j$ for each $0 \le j < N$ if the initial input generation is labeled
$i_0$, \ldots, $i_{N-1}$.

At a reproduction event where haploid parents $x$ and $y$ produce offspring $u$ 
at time $t$ in population $p$,
\begin{enumerate}
    \item add a 
        $(
        \text{flags}=0,
        \text{population}=p, 
        \text{time}=t)$ row to the node table,
    \item set $L[u] = n$,
    \item and increment $n\mathrel{+}=1$.
\end{enumerate}
Then, for each interval $[\ell,r)$ that $u$ inherits from parent $z$
(where $z$ is either $x$ or $y$),
\begin{enumerate}[resume]
    \item add a 
        $( \text{left}=\ell,
        \text{right}=r, 
        \text{parent}=z, 
        \text{children}=(u,))$ row to the edgeset table.
\end{enumerate}
If furthermore there have been mutations at genomic locations $s_1$, \ldots, $s_k$
on this interval,
with derived states $q_1$, \ldots, $q_k$,
then for each $1 \le j \le k$,
\begin{enumerate}[resume]
    \item if $s_j$ is not in the site table, add a row
        $( \text{position}=s_j,
        \text{ancestral\_state}=0)$,
    \item find the site index $i$ whose position is $s_j$,
    \item and add a row
        $( \text{site}=i$,
           \text{node}=u$,
        \text{derived\_state}=q_j)$ to the mutation table.
\end{enumerate}

To \textbf{simplify} the tree sequence at time $T$,
\begin{enumerate}
    \item add $T-T_0$ to each time in the first $n_0$ rows of the node table,
        and replace each remaining time $t$ with $T-t$.
\end{enumerate}
Then, pass the set of currently alive input IDs,
$i_0, \ldots, i_{N-1}$ to the simplificaiton algorithm,
which produces a tree sequence that has node ID $j$ corresponding to input ID $i_j$
for $0 \le j < N$, and
\begin{enumerate}[resume]
    \item empty $L$,
    \item let $L[j] = i_j$ for $0 \le j < N$,
    \item set $n$ to be the number of nodes in the new tree sequence and set $n_0 = n$, and finally
    \item set $T_0 = T$.
\end{enumerate}

Simplification keeps the tables to a managable size.
Since the map $L$ is updated to maintain the association between individuals in the simulation
and nodes in the tree sequence, simplification can be run regularly, 
as the simulation progresses.

\begin{figure}
    \begin{center}
        \includegraphics{method_diagram}
    \end{center}
    \caption{
        A simple example of the method.
        \textbf{Top:} the ARG shown on the left relates ten haploid individuals to each other.
        It is recorded, in forwards time, 
        in 10 node records (one for each individual)
        and 14 edgeset records (one for each distinctly inherited segment).
        Blue numbers denote crossing over locations in each meiosis.
        The individuals $B$, $C$, and $D$ inherit clonally from $A$
        to ensure rootnedness of resulting tree.
        \textbf{Center:} the five distinct trees relating all individuals to each other
        found across the chromosome (blue numbers denote locations on the chromosome).
        Labels after simplification are shown in red.
        \textbf{Bottom:} tables recording the tree sequence after simplification 
        with nodes $I$ and $J$ as samples.
        The mapping from labels in the forwards time simulation to nodes in the tree sequence
        is shown in red,
        which allows additional records to be added as the simulation progresses.
        \label{fig:method_diagram}
    }
\end{figure}

\plr{Something about simplification (but not the actual algorithm)
and estimates of how much memory usage is reduced by each time it is carried out.}
The resulting tables encode everything we need to know --
in fact, they record all of history for everyone alive at any time through the simulation.
This is much more than we need to reconstruct the genealogies and sequences
of a smaller sample of indidivuals.
Reducing this larger tree sequence to a smaller one relevant to a given set of ``sample'' nodes
we call \emph{simplification}.
Roughly, this works by tracing ancestry from the samples backwards through the recorded history,
adding node and edgeset records to the output only when coalescent events are reached.
This works exactly as in \msprime, allowing substantial re-use of algorithms;
the main difference being that parental choice, mutations, and recombination locations 
are determined by the input tree sequence
rather than randomly generated.
An example is shown in Figure \ref{fig:method_diagram}.
\plr{(trying to make the point strongly in the figure, that everything can be stored quite concisely)}

\plr{Note that you can begin the forwards-time simulation with the results of a previous (coalescent) one,
which is helpful so you don't need to run for so long (cite Wilkins).}
In this scheme,
at any point in the simulation genealogical history is recorded in a tree sequence.
This has two additional advantages.
First, simplification can be run periodically through the simulation,
taking the set of samples to be the entire currently alive population.
This is important as it keeps memory usage from growing linearly (and quickly) with time.
Second, the simulation can be \emph{begun} with a tree sequence produced by some other method --
for instance, by a coalescent simulation with \msprime.
This allows for incorporation of deep-time history beyond the reach of individual-based simulations.
Since geographic structure from times longer ago than the mixing time of migration across the range
has limited effect on modern genealogies \citep{wilkins_separation}
(other than possibly changing effective population size \citet{durretspatial}),
this may not negatively affect realism.

\plr{Something about putting mutations down on the tree sequence.}

\plr{Quick overview of how to efficiently hook this up with other code.}

%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

\plr{Estimates of run-time complexity}
Suppose that we wish to run a forwards-time simulation of $N$ individuals for $T$ generations,
in which there are $S$ selected loci and $L$ neutral loci.
We will estimate run-time complexity and memory usage for both a ``naive'' strategy that carries along neutral loci
and an ``ancestry-tracking'' strategy like that we consider here.
To do this, we assume that each individual must carry along its entire genotype.
More advanced schemes are used in some simulators,
but these increase efficiency by utilizing redundancy introduced by shared ancestry, 
which is effectively an intermediate scheme.
We omit the cost of computing a fitness function.

Both schemes must choose mates and recombination breakpoints,
and pass on selected genotypes.
The difference between the two comes from the tradeoff between
(a) passing on neutral genotypes, and
(b) recording and simplifying the tree sequence, and adding neutral genotypes afterwards.
(We assume here that selected genotypes are stored in the same way for both.)
Passing on neutral genotypes naively records $L$ items per individual each generation, discarding the previous generation.
\plr{do this better with numbers from nodes, edgesets}
Recording a tree sequence stores 2 parents and $\rho$ breakpoints on average each generation;
after $T$ generations this grows to $N \times T \times (2 + \rho)$ stored items.
However, after simplification a tree sequence for all $N$ individuals
only takes of order $N + \rho \log N$ records.
Simplification requires processing each of the initial records -- so, order of $N \times T \times \rho$ operations;

\begin{table}
\begin{center}
    \begin{tabular}{l||r|r||r|r|}
                    & \multicolumn{2}{|c|}{naive} & \multicolumn{2}{|c|}{ancestry-tracking} \\
                    & operations  & memory & operations  & memory \\
        mate choice & $N$       &   --  &   $N$     & -- \\
        recombination & $N\rho$ &   --  &   $N\rho$ & -- \\
    \end{tabular}
\end{center}
\end{table}

Comparison of simulation with/without msprime, using simuPOP 
or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).

Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
in a day.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusion}

This is a general-purpose strategy that can be applied to other methods.

All sorts of good reasons to want to have whole-genome simulations.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}

Kevin

\bibliography{references}

\end{document}

### 2017-08-26 forwards_paper.tex
added writeup on data structures and wrote out simple WF sim

@@
 \usepackage{enumitem}
+\usepackage{listings}
 \bibliographystyle{plainnat}
+\usepackage[T1]{fontenc}
 
@@
 \newcommand{\msprime}{\texttt{msprime}}
+\newcommand{\nodes}{\texttt{nodes}}
+\newcommand{\edgesets}{\texttt{edgesets}}
+\newcommand{\sites}{\texttt{sites}}
+\newcommand{\mutations}{\texttt{mutations}}
 
@@
 
+\emph{Note:} author order not determined
+
 \begin{abstract}
-    To use genomic data for inference and prediction it is often necessary to obtain whole-genome information
+    To use genomic data for inference and prediction 
+    it is often necessary to obtain whole-genome information
     from individual-based simulations,
@@
 
-OUTLINE
-\begin{itemize}
+\textbf{OUTLINE}
+\begin{enumerate}
     \item motivate need for whole-genome fwd-time simulations; point out that we only recently have the computing power to do this
     \item explain ARG: explain that for forwards-time only need selected loci as by defn all others can be put on afterwards
+    \item review something about msprime methods for storing/traversing tree sequence
+    \item describe tables and write out conditions to have valid tables
+    \item write down algorithm used to do simple WF simulation
+    \item describe simplify algorithm
     \item back-of-the-envelope calculation to compare cost of tracking whole genomes versus putting mutations on ARG
-    \item review something about your efficient methods for storing/traversing tree sequence
-    \item write down algorithm used in ftprime to produce msprime-ready coalescence records
-    \item describe simplify algorithm
     \item comparison of speed with simupop, fwdpp
-\end{itemize}
+\end{enumerate}
 
@@
 
-\plr{two main points below: can put down neutral mutations afterwards, and could be useful to compute stats even before that}
+\plr{perhaps some of this veers into methods rather than introduction}
 
@@
 
-\plr{Quick review of the ARG and discussion of whether the ARG means the data structure or the reverse-time markov process (the former).}
-Suppose that to the \emph{population pedigree} --
+In so doing we record the \emph{population pedigree} --
 the entire history of parent-offspring relationships of an entire population going back to a remote time --
-we add information encoding the genetic outcomes of each ancestral meiosis --
+as well as information encoding the genetic outcomes of each ancestral meiosis --
 who inherited which parts of which parental chromosomes.
-This embellished graph is known as the \emph{ancestral recombination graph}, or \emph{ARG} \citep{griffiths}.
+This embellished graph is known as the \emph{ancestral recombination graph}, or {ARG} \citep{griffiths}.
 Combined with ancestral genotypes and the origins of new mutations,
 it completely specifies the genomic sequence of any individual in the population at any time.
-However, much less than the entire ARG is needed to specify relationships between any given set of samples --
-only those portions of it from which those samples have actually inherited, back to their most recent common ancestors.
-The assumptions of coalescent theory -- random mating and neutrality --
-imply that the stochastic law of this random set forms a Markov process looked at \emph{backwards} in time,
-and can hence be simulated without reference to the unnecessary remainder.
-Although this point was in principle known since \citet{griffithsmarjoram},
-only recent algorithmic advances made it possible to actually simulate the process correctly,
-without approximation, across whole chromosomes \citep{msprime}.
+However, much less than the entire ARG is needed to specify relationships 
+between any given set of samples --
+only those portions of it from which those samples have actually inherited, 
+back to their most recent common ancestors.
+The assumptions of coalescent theory 
+% -- random mating and neutrality --
+imply that this random set describes a Markov process looked at \emph{backwards} in time,
+and can hence be simulated without reference to the unnecessary remainder 
+\citet{griffithsmarjoram,msprime}.
+This stochastic process is also called the ``ARG'',
+but we reserve this name for the embellished graph itself.
+% Although this point was in principle known since \citet{griffithsmarjoram},
+% only recent algorithmic advances made it possible to actually simulate the process correctly,
+% without approximation, across whole chromosomes \citep{msprime}.
 
@@
 
-\plr{Description of what we actually need to know in the end (the trees).}
+\plr{Reminder of what we need to know in the end (the trees), and
+    quick review of \msprime{} methods: sparse trees, tree differences.}
 
-\plr{Quick review of \msprime methods: sparse trees, tree differences.}
+%%%%%%
+\subsection*{Data structures:}
 
-First we recall the data storage methods used by msprime \citep[see][for more details]{msprime}.
+First we describe the data structures we use for recording genealogical history,
+as implemented in \msprime{}.
+These derive from those described in \citet{msprime},
+but have been modified to remove redundancy.
 
-To record the history and identity of a set of sampled genomes,
-it is only necessary to record commmon ancestry (or, \emph{coalescence}) events --
-most recent common ancestors of distinct segments of sampled genomes --
-and the identity and location of any mutations that these most recent common ancestors carry
-that are not shared by any other recorded ancestor.
-\plr{(not quite right)}
+To clarify terminology,
+below a \emph{tree} refers to
+a genealogical tree describing how a collection of individuals are related to each other.  
+A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
+tree relating all samples to each other at any point along the genome.
+In the context of a tree sequence,
+\emph{nodes} refer to distinct ancestors,
+and may be identified with points in the trees of a tree sequence.
+Every tip or branching point in a tree is associated with a node.
+A tree sequence does not necessarily carry complete genealogical information about every node,
+but calling a node a \emph{sample} implies that complete information is present for that ancestor,
+i.e., the node is associated with a point in every tree in the tree sequence.
+Since each node represents a certain ancestor, it has a
+unique ``time``, thought of as her birth time, which determines the height
+of any branching points she is associated with.  A given node will be
+associated with branching points of all trees across a region if that node
+is the most recent common ancestor to the subtending tips across that region.  
+This information is stored in the columns of a \textbf{Node Table}:
+\begin{align*}
+    \text{flags,} \qquad \text{population,} \qquad \text{time} \\ 
+\end{align*}
+where ``flags'' records other information 
+(e.g., a binary mask of `1' indicates the node is a sample).
+Importantly, the \textbf{node ID} of a node is given implicitly
+by the (zero-based) index of its corresponding row in the Node Table.
 
-The topology of the ARG
-is recorded by msprime as a list of Coalescence Records, which are tuples of the form
-\texttt{(left,right,parent,children)},
-which indicates that the individuals listed in \texttt{children}
-have inherited the segment of genome between \texttt{left} and \texttt{right}
-from the ancestor \texttt{parent}.
-Individuals are identified by integers,
-and there is also a table recording the \emph{birth time} of each individual.
+Tree sequences are constructed by specifying over which segments of genome
+which nodes inherit from which other nodes.  This information is stored by
+recording the endpoints of each distinctly inherited ancestral segment,
+the parental node, and a list of children nodes who have inherited that segment.
+As each such record describes a collection of edges across a swatch of trees in the tree sequence,
+we call these \emph{edgesets} 
+and store them in the columns of an \textbf{Edgeset Table:}
+\begin{align*}
+    \text{left,} \qquad \text{right,} \qquad \text{parent,} \qquad \text{children}
+\end{align*}
 
-For a set of coalescence records to be consistent,
-it is required that:
+To record information about genetic variants we need to also record each mutation
+and which nodes have inherited that mutation.
+The tree structure takes care of inheritance -- all we need to do
+is to record the highest node in the tree at the mutated site
+that inherited that mutation.
+As more than one mtuation may occur at a given site,
+we separate this information into two tables,
+first, the \textbf{Site Table} records for each variant site
+\begin{align*}
+    \text{position,} \qquad \text{ancestral state}
+\end{align*}
+Here ``position'' is a (floating point) position along the chromosome,
+and ``ancestral state'' is the genotype of the root of the tree at that site.
+As for nodes, \textbf{site IDs} are given implicitly
+by the (zero-based) index of the rows.
+Then, we record in a \textbf{Mutation Table}
+\begin{align*}
+    \text{site,} \qquad \text{node,} \qquad \text{derived state}
+\end{align*}
+in which ``site'' is the ID of the site at wihch this mutation occurred,
+``node'' is the ID of the highest node that has inherited this mutation,
+and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
+unless another mutation occurs.
+\plr{Omit migrations?}
+
+\paragraph{Definition of valid tables}
+Here are the formal requirements for a set of nodes and edgesets to make sense,
+and to allow ``msprime``'s algorithms to work properly.
+
+To disallow time travel and multiple inheritance:
 \begin{enumerate}
-    \item The birth time of any \texttt{children} must be strictly after the birth time of the \texttt{parent}.
-    \item The set of intervals on which individual \texttt{a} is a child must be disjoint, for every \texttt{a}.
-    \item The set of intervals on which individual \texttt{a} is a parent must be disjoint, for every \texttt{a}.
+    \item Offspring must be born after their parents (and hence, no loops).
+    \item The set of intervals on which each individual is a child must be disjoint.
 \end{enumerate}
-These ensure that the coalescence records unambiguously specify a subset of the genealogy at any location on the genome,
-but do not ensure that the entire history of the samples is specified.
-Furthermore, to allow for efficient algorithms, msprime also requires that
-\begin{enumerate} \setcounter{enumi}{3}
-    \item The samples must be numbered \texttt{0,...,n-1}, and the smallest non-sampled label must be an internal node.
-        \plr{perhaps remove this one}
-    \item The list of \texttt{children} in a coalescence record must be sorted.
-    \item Records must be sorted in nondecreasing order by birth time of the \texttt{parent}.
-    \item Birth times must be strictly greater than zero.
+For algorithmic reasons, we also require:
+\begin{enumerate} \setcounter{enumi}{2}
+    \item The leftmost endpoint of each chromosome is 0.0.
+    \item Node times must be strictly greater than zero.
+    \item The list of offspring in an edgeset must be sorted.
+    \item Edgesets must be sorted in nondecreasing time order.
+    \item The set of intervals on which each individual is a parent must be disjoint.
+    \item Each edgeset must contain at least two children.
 \end{enumerate}
+Note that since each node time is equal to the amount of time since the \emph{birth} of the
+corresponding parent, time is measured in clock time, not in meioses.
 
-\plr{Specification for tables of nodes, edgesets, sites, and mutations (migrations?).}
-\plr{Calling a collection of unordered tables a ``relaxed'' tree sequence
-    but will make it clear that to do anything with these must sort first.}
+A forwards-time simulation does \textbf{not} naturally emit genealogical information
+satisfying requirements 5--8.
+However, \msprime{} implements two algorithms that will take a set of tables
+satisfying only 1--4 and produce tables satisfying all requirements.
+Trivially, \texttt{sort\_tables} enforces requirements 5 and 6 and does not renumber nodes;
+then, \texttt{simplify} enforces requirements 7 and 8 (and does much more; see below).
 
-\begin{itemize}
-    \item[Nodes:]
-        (flags, population, time)
-    \item[Edgesets:]
-        (left,  right,  parent,  children)
-    \item[Sites:]
-        (position,  ancestral\_state)
-    \item[Mutations:]
-        (site,  node,  derived\_state)
-    \item \plr{Migrations?}
-\end{itemize}
+%%%%%%
+\subsection*{Recording the ARG in forwards time}
 
-\plr{Steps to record information in forwards time.}
-In the forwards simulation,
-we record the results of each meiosis --
-for each distinctly inherited segment of offspring chromosome,
-we record the parental chromosome,
-the left and right endpoints,
-and the identities of any new mutations that have occurred on it.
+To record the genealogical history of a simulation
+we need to record two things for each new chromosome:
+the birth time,
+and the endpoints and parental IDs of each distinctly inherited segment.
 This is recorded easily, without further processing,
-in the relaxed tree sequence format.
+in the tables described above.
+
+For concreteness, here we write out in pseudocode
+how to run a neutral Wright--Fisher simulation
+with overlapping generations
+that records genealogical history in this way.
+The simulation will run for $T$ generations,
+and has $N$ haploid individuals, each carrying a single chromosome of length L,
+on which for simplicity we assume there is exactly one crossover per generation.
+The probability of death per individual each generation is \texttt{death\_prob},
+and the whole-chromosome mutation rate per individual is \texttt{mut\_rate}.
+
+\paragraph{Initialize:}
+We will build the tables \nodes, \edgesets, \sites, and \mutations,
+and keep track of the IDs of the current generation in the list \texttt{pop}.
+\begin{lstlisting}[language=Python]
+for i in 0:N-1:
+    nodes.add_row(time=T)
+    pop[i] = i
+\end{lstlisting}
+(Alternatively, the tables could be initialized by the results of a coalescent simulation.)
+
+\paragraph{Iterate:}
+We then step through the generations:
+\plr{Add mutations to this:}
+\begin{lstlisting}[language=Python]
+for t in 1:T:
+    for i in 0:N-1:
+        if random.uniform() > death_prob:
+            new_pop[i] = pop[i]
+        else:
+            u = nodes.num_rows  # the ID of the new individual
+            new_pop[i] = u
+            nodes.add_row(time = T-t)
+            a = random.sample(pop)
+            b = random.sample(pop)
+            bp = random.sample(0:L)
+            if bp > 0:
+                edgesets.add_row(left=0, right=bp,
+                                  parent=a, children=(u,))
+            if bp < L:
+                edgesets.add_row(left=bp, right=L,
+                                  parent=b, children=(u,))
+    pop = new_pop
+\end{lstlisting}
+
+\paragraph{Finalize:}
+to obtain a tree sequence, we need only transform the tables to the required format:
+\begin{lstlisting}
+sort_tables(nodes, edgesets, sites, mutations)
+simplify_tables(nodes, edgesets, sites, mutations)
+\end{lstlisting}
+which can then be loaded into a tree sequence with \msprime.
+(These two functions operate on the tables in place.)
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{method_diagram}
+    \end{center}
+    \caption{
+        A simple example of the method.
+        \textbf{Top:} the ARG shown on the left relates ten haploid individuals to each other.
+        It is recorded, in forwards time, 
+        in 10 node records (one for each individual)
+        and 14 edgeset records (one for each distinctly inherited segment).
+        Blue numbers denote crossing over locations in each meiosis.
+        The individuals $B$, $C$, and $D$ inherit clonally from $A$
+        to ensure rootnedness of resulting tree.
+        \textbf{Center:} the five distinct trees relating all individuals to each other
+        found across the chromosome (blue numbers denote locations on the chromosome).
+        Labels after simplification are shown in red.
+        \textbf{Bottom:} tables recording the tree sequence after simplification 
+        with nodes $I$ and $J$ as samples.
+        The mapping from labels in the forwards time simulation to nodes in the tree sequence
+        is shown in red,
+        which allows additional records to be added as the simulation progresses.
+        \label{fig:method_diagram}
+    }
+\end{figure}
+
+%%%%%%
+\subsection*{Tree sequence simplification}
+
+The resulting tables encode everything we need to know --
+in fact, they record all of history for everyone alive at any time through the simulation.
+This is much more than we need to reconstruct the genealogies and sequences
+of a smaller sample of indidivuals.
+Reducing this larger tree sequence to a smaller one relevant to a given set of ``sample'' nodes
+we call \emph{simplification}.
+Roughly, this works by tracing ancestry from the samples backwards through the recorded history,
+adding node and edgeset records to the output only when coalescent events are reached.
+This works exactly as in \msprime, allowing substantial re-use of algorithms;
+the main difference being that parental choice, mutations, and recombination locations 
+are determined by the input tree sequence
+rather than randomly generated.
+An example is shown in Figure \ref{fig:method_diagram}.
+\plr{(trying to make the point strongly in the figure, that everything can be stored quite concisely)}
+
+\plr{Note that you can begin the forwards-time simulation with the results of a previous (coalescent) one,
+which is helpful so you don't need to run for so long (cite Wilkins).}
+In this scheme,
+at any point in the simulation genealogical history is recorded in a tree sequence.
+This has two additional advantages.
+First, simplification can be run periodically through the simulation,
+taking the set of samples to be the entire currently alive population.
+This is important as it keeps memory usage from growing linearly (and quickly) with time.
+Second, the simulation can be \emph{begun} with a tree sequence produced by some other method --
+for instance, by a coalescent simulation with \msprime.
+This allows for incorporation of deep-time history beyond the reach of individual-based simulations.
+Since geographic structure from times longer ago than the mixing time of migration across the range
+has limited effect on modern genealogies \citep{wilkins_separation}
+(other than possibly changing effective population size \citet{durretspatial}),
+this may not negatively affect realism.
+
+\plr{Something about putting mutations down on the tree sequence.}
+
+%%%%%%
+\subsection*{Overview of the API}
+
+\plr{Quick overview of how to efficiently hook this up with other code.}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Results}
+
+\plr{Estimates of run-time complexity}
+Suppose that we wish to run a forwards-time simulation of $N$ individuals for $T$ generations,
+in which there are $S$ selected loci and $L$ neutral loci.
+We will estimate run-time complexity and memory usage for both a ``naive'' strategy that carries along neutral loci
+and an ``ancestry-tracking'' strategy like that we consider here.
+To do this, we assume that each individual must carry along its entire genotype.
+More advanced schemes are used in some simulators,
+but these increase efficiency by utilizing redundancy introduced by shared ancestry, 
+which is effectively an intermediate scheme.
+We omit the cost of computing a fitness function.
+
+Both schemes must choose mates and recombination breakpoints,
+and pass on selected genotypes.
+The difference between the two comes from the tradeoff between
+(a) passing on neutral genotypes, and
+(b) recording and simplifying the tree sequence, and adding neutral genotypes afterwards.
+(We assume here that selected genotypes are stored in the same way for both.)
+Passing on neutral genotypes naively records $L$ items per individual each generation, discarding the previous generation.
+\plr{do this better with numbers from nodes, edgesets}
+Recording a tree sequence stores 2 parents and $\rho$ breakpoints on average each generation;
+after $T$ generations this grows to $N \times T \times (2 + \rho)$ stored items.
+However, after simplification a tree sequence for all $N$ individuals
+only takes of order $N + \rho \log N$ records.
+Simplification requires processing each of the initial records -- so, order of $N \times T \times \rho$ operations;
+
+% \begin{table}
+% \begin{center}
+%     \begin{tabular}{l||r|r||r|r|}
+%                     & \multicolumn{2}{|c|}{naive} & \multicolumn{2}{|c|}{ancestry-tracking} \\
+%                     & operations  & memory & operations  & memory \\
+%         mate choice & $N$       &   --  &   $N$     & -- \\
+%         recombination & $N\rho$ &   --  &   $N\rho$ & -- \\
+%     \end{tabular}
+% \end{center}
+% \end{table}
+
+Comparison of simulation with/without msprime, using simuPOP 
+or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
+
+Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
+so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
+in a day.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Conclusion}
+
+This is a general-purpose strategy that can be applied to other methods.
+
+All sorts of good reasons to want to have whole-genome simulations.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements}
+
+\bibliography{references}
+
+\appendix
+
+\section{More general method for recording the ARG}
+
+\plr{moved from above}
 
@@
 
-\begin{figure}
-    \begin{center}
-        \includegraphics{method_diagram}
-    \end{center}
-    \caption{
-        A simple example of the method.
-        \textbf{Top:} the ARG shown on the left relates ten haploid individuals to each other.
-        It is recorded, in forwards time, 
-        in 10 node records (one for each individual)
-        and 14 edgeset records (one for each distinctly inherited segment).
-        Blue numbers denote crossing over locations in each meiosis.
-        The individuals $B$, $C$, and $D$ inherit clonally from $A$
-        to ensure rootnedness of resulting tree.
-        \textbf{Center:} the five distinct trees relating all individuals to each other
-        found across the chromosome (blue numbers denote locations on the chromosome).
-        Labels after simplification are shown in red.
-        \textbf{Bottom:} tables recording the tree sequence after simplification 
-        with nodes $I$ and $J$ as samples.
-        The mapping from labels in the forwards time simulation to nodes in the tree sequence
-        is shown in red,
-        which allows additional records to be added as the simulation progresses.
-        \label{fig:method_diagram}
-    }
-\end{figure}
-
-\plr{Something about simplification (but not the actual algorithm)
-and estimates of how much memory usage is reduced by each time it is carried out.}
-The resulting tables encode everything we need to know --
-in fact, they record all of history for everyone alive at any time through the simulation.
-This is much more than we need to reconstruct the genealogies and sequences
-of a smaller sample of indidivuals.
-Reducing this larger tree sequence to a smaller one relevant to a given set of ``sample'' nodes
-we call \emph{simplification}.
-Roughly, this works by tracing ancestry from the samples backwards through the recorded history,
-adding node and edgeset records to the output only when coalescent events are reached.
-This works exactly as in \msprime, allowing substantial re-use of algorithms;
-the main difference being that parental choice, mutations, and recombination locations 
-are determined by the input tree sequence
-rather than randomly generated.
-An example is shown in Figure \ref{fig:method_diagram}.
-\plr{(trying to make the point strongly in the figure, that everything can be stored quite concisely)}
-
-\plr{Note that you can begin the forwards-time simulation with the results of a previous (coalescent) one,
-which is helpful so you don't need to run for so long (cite Wilkins).}
-In this scheme,
-at any point in the simulation genealogical history is recorded in a tree sequence.
-This has two additional advantages.
-First, simplification can be run periodically through the simulation,
-taking the set of samples to be the entire currently alive population.
-This is important as it keeps memory usage from growing linearly (and quickly) with time.
-Second, the simulation can be \emph{begun} with a tree sequence produced by some other method --
-for instance, by a coalescent simulation with \msprime.
-This allows for incorporation of deep-time history beyond the reach of individual-based simulations.
-Since geographic structure from times longer ago than the mixing time of migration across the range
-has limited effect on modern genealogies \citep{wilkins_separation}
-(other than possibly changing effective population size \citet{durretspatial}),
-this may not negatively affect realism.
-
-\plr{Something about putting mutations down on the tree sequence.}
-
-\plr{Quick overview of how to efficiently hook this up with other code.}
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Results}
-
-\plr{Estimates of run-time complexity}
-Suppose that we wish to run a forwards-time simulation of $N$ individuals for $T$ generations,
-in which there are $S$ selected loci and $L$ neutral loci.
-We will estimate run-time complexity and memory usage for both a ``naive'' strategy that carries along neutral loci
-and an ``ancestry-tracking'' strategy like that we consider here.
-To do this, we assume that each individual must carry along its entire genotype.
-More advanced schemes are used in some simulators,
-but these increase efficiency by utilizing redundancy introduced by shared ancestry, 
-which is effectively an intermediate scheme.
-We omit the cost of computing a fitness function.
-
-Both schemes must choose mates and recombination breakpoints,
-and pass on selected genotypes.
-The difference between the two comes from the tradeoff between
-(a) passing on neutral genotypes, and
-(b) recording and simplifying the tree sequence, and adding neutral genotypes afterwards.
-(We assume here that selected genotypes are stored in the same way for both.)
-Passing on neutral genotypes naively records $L$ items per individual each generation, discarding the previous generation.
-\plr{do this better with numbers from nodes, edgesets}
-Recording a tree sequence stores 2 parents and $\rho$ breakpoints on average each generation;
-after $T$ generations this grows to $N \times T \times (2 + \rho)$ stored items.
-However, after simplification a tree sequence for all $N$ individuals
-only takes of order $N + \rho \log N$ records.
-Simplification requires processing each of the initial records -- so, order of $N \times T \times \rho$ operations;
-
-\begin{table}
-\begin{center}
-    \begin{tabular}{l||r|r||r|r|}
-                    & \multicolumn{2}{|c|}{naive} & \multicolumn{2}{|c|}{ancestry-tracking} \\
-                    & operations  & memory & operations  & memory \\
-        mate choice & $N$       &   --  &   $N$     & -- \\
-        recombination & $N\rho$ &   --  &   $N\rho$ & -- \\
-    \end{tabular}
-\end{center}
-\end{table}
-
-Comparison of simulation with/without msprime, using simuPOP 
-or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
-
-Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
-so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
-in a day.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Conclusion}
-
-This is a general-purpose strategy that can be applied to other methods.
-
-All sorts of good reasons to want to have whole-genome simulations.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements}
-
-Kevin
-
-\bibliography{references}
-
 \end{document}

### 2017-08-26 forwards_paper.tex
added mutations to WF writeup

@@
 \paragraph{Iterate:}
-We then step through the generations:
-\plr{Add mutations to this:}
+We then step through the generations,
+using the functions \texttt{random\_mutation} and \texttt{random\_allele}
+to choose positions and alleles for mutations, respectively:
 \begin{lstlisting}[language=Python]
@@
                                   parent=b, children=(u,))
+            num_muts = random.poisson(mut_rate)
+            for j in 1:num_muts:
+                pos = random_mutation()
+                if pos not in sites.position:
+                    s = sites.num_rows
+                    sites.add_row(position=pos, 
+                           ancestral_state=random_allele())
+                else:
+                    s = sites.position.index(pos)
+                mutations.add_row(site=s, node=u, 
+                           derived_state=random_allele())
     pop = new_pop

### 2017-09-12 sims/benchmark-simuPOP.py
Benchmark script and docs
- Adds reproducible conda setup
- For how to run see README.md or scripts sims/test.sh and sims/big_test.sh

@@ -1,0 +1,336 @@
+#!/usr/bin/env python3
+import gzip
+import sys
+import time
+import csv
+import random
+from ftprime import RecombCollector
+import msprime
+import argparse
+
+description = '''
+Benchmark
+
+Desired interface:
+
+parallel --timeout 2 python benchmark-simuPOP.py -N 1000 --theta 100 --rho 100 \
+    --pdel 0.1 --nsam 10 --gc 100 --seed ::: 42
+'''
+# Notes from K Thornton / P Ralph discussion (summary by KT):
+# 1. Fitness at a single site for the genotypes AA, AA, and aa
+# 1, 1+sh, 1+s.
+# **I vote for h = 0.5.
+
+# X2. How to calculate fitness across sites.
+# ** I vote for multiplicative.
+
+# X 3. The DFE itself.
+# I think we decided that the DFE will be on 2Ns rather than on s.
+# ** I vote for (-1.0)*Gamma(shape=1.0,scale=5.0).  This gives us a mean |2Ns|
+# of 5 and 30% of new mutations have 2 <= |2Ns| <= 5, which is an interesting
+# domain where deleterious variants can persists for some time.
+
+# X 4. Mutation and recombination rates.
+# These are scaled as 4Nu and 4Nr, respectively.  We also decided 4Nu = 4Nr.
+# ** I propose 4Nu \in {1e3, 1e4, 1e5}, which is roughly 1, 10, and 100Mb of a
+# "human" chromosome.
+# ** I propose that rates are uniform along the regions.
+
+# X 5. Population size.
+# ** I propose N \in {1e3, 1e4, 1e5}
+
+# X 6. Simulation length
+# ** 20N generations.  We kinda need this until Jerome addresses the issues
+# w/incompletely-coalesced marginal trees.
+
+# X (solved in submission script) 7. The max run time.
+# ** I propose 72 hours.  In the past, I've done one week, but my preliminary
+# results suggest I can do N=1e4, 4Nu=4Nr=1e5 in just a few hours (!!!!!).  I
+# fully expect sims including all neutral variants to hit this max time for when
+# N and/or 4Nu/r are large.
+
+# X 8. The rate at which deleterious mutations arise.
+# ** I propose we keep this as a constant function of 4Nu.  For no particular
+# reason, let's say 0.1.  Thus, for a given \theta = 4Nu, the rate at which
+# deleterious variants arise will be 0.1u.  As we increase 4Nu/r, we are
+# simulating a bigger and bigger region, holding the fraction of deleterious new
[…]

### 2017-09-12 sims/big_test.sh

@@ -1,0 +1,2 @@
+#!/bin/bash
+python3 benchmark-simuPOP.py --seed 42 --nsam 10 --pdel 0.1 --theta 100 --gc 100 --rho 100 -N 1000 --logfile tmp.log --csvfile out.csv

### 2017-09-12 sims/setup-simuPOP.sh

@@ -1,0 +1,10 @@
+conda config --add channels conda-forge
+conda update -q conda
+conda info -a
+if [[ ! -d "/opt/conda/envs/ftprime" ]] ; then
+	conda create -q -n ftprime python=3.5
+fi
+wget https://raw.githubusercontent.com/ashander/ftprime/5ad8bcb3a49a19c9cfd52bdf615140af6f83d218/environment.yml
+conda env update -q -n ftprime -f environment.yml
+source activate ftprime
+pip install --editable=git+https://github.com/ashander/ftprime.git@master#egg=ftprime

### 2017-09-12 sims/test.sh

@@ -1,0 +1,2 @@
+#!/bin/bash
+python3 benchmark-simuPOP.py --seed 43 --nsam 10 --pdel 0.1 --theta 1 --rho 1 -N 100 --csvfile out.csv

### 2017-09-13 forwards_paper.tex
added example tree seq

@@
 \bibliographystyle{plainnat}
-\usepackage[T1]{fontenc}
+% \usepackage[T1]{fontenc}
 
@@
 at each position along the genome.
-
-There is another compelling reason for recording the entire genealogical history.
-In many applications, simulations are run for the purpose of comparing summary statistics of the resulting genomes
-to those computed from real data,
-with the purpose of inferring demographic history or identifying regions under selection.
-However, for the same reason that neutral mutations can be laid down afterwards on the sequence of genealogical trees resulting from a simulation,
-any summary statistic of genetic variation can be expressed as a statistic of the distribution of genealogical trees along the genome,
-with additional noise introduced by mutations.
-For instance, mean pairwise sequence diversity, $\theta$,
-is most easily thought of as an estimator of $2 \mu t_{TMRCA}$, the mean time to most recent common ancestor
-multiplied by twice the mutation rate,
-with a relative error that is independent of the actual $t_{TMRCA}$.
-As additional measurement noise makes simulation-based inference methods such as ABC or trained classifiers more diffcult,
-it could be beneficial to compare the statistics observed in the data
-to their expected values conditioned on the genealogical history,
-as remaining mutational noise is not informative.
-\plr{Say better.  A quick way to demonstrate this point?}
-
-There is an obvious cost to recording all of history from a simulation,
-rather than just the current state.
 In this paper, we show how to use algorithmic tools and data structures developed for the coalescent simulator \msprime
@@
 but have been modified to remove redundancy.
+The tables below give the example tree sequence of Figure \ref{fig:ex_tree_seq}.
+
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=0.6\textwidth]{example_tree_seq}
+    \end{center}
+    \caption{
+        A pictorial representation of the ARG
+        relating three samples to each other
+        over a chromosome of length 1.0.
+        Mutations shown in the example tables of the text
+        are marked with red stars.
+        \label{fig:ex_tree_seq}
+    }
+\end{figure}
 
@@
 Since each node represents a certain ancestor, it has a
-unique ``time``, thought of as her birth time, which determines the height
+unique ``time'', thought of as her birth time, which determines the height
 of any branching points she is associated with.  A given node will be
@@
 This information is stored in the columns of a \textbf{Node Table}:
-\begin{align*}
-    \text{flags,} \qquad \text{population,} \qquad \text{time} \\ 
-\end{align*}
+\begin{center}
+\begin{tabular}{cccc}
+    id   &  is\_sample &  population &  time \\
+    0    &  1          & 0           & 0 \\
+    1    &  1          & 0           & 0 \\
+    2    &  1          & 0           & 0 \\
+    3    &  0          & 0           & 0.3 \\
+    4    &  0          & 0           & 0.5 \\
+    5    &  0          & 0           & 0.6 \\
+    6    &  0          & 0           & 0.7 \\
+\end{tabular}
+\end{center}
 where ``flags'' records other information 
@@
 and store them in the columns of an \textbf{Edgeset Table:}
-\begin{align*}
-    \text{left,} \qquad \text{right,} \qquad \text{parent,} \qquad \text{children}
-\end{align*}
+\begin{center}
+\begin{tabular}{cccc}
+    left & right & parent & children \\
+    0.2  &  0.8  &  3     & 0,2 \\
+    0.0  &  0.2  &  4     & 1,2 \\
+    0.2  &  0.8  &  4     & 1,3 \\
+    0.8  &  1.0  &  4     & 1,2 \\
+    0.8  &  1.0  &  5     & 0,4 \\
+    0.0  &  0.2  &  6     & 0,4  \\
+\end{tabular}
+\end{center}
 
@@
 first, the \textbf{Site Table} records for each variant site
-\begin{align*}
-    \text{position,} \qquad \text{ancestral state}
-\end{align*}
+\begin{center}
+\begin{tabular}{ccc}
+    id  &   position  & ancestral\_state \\
+    0   &  0.1        & 0 \\
+    1   &  0.5        & 0 \\
+\end{tabular}
+\end{center}
 Here ``position'' is a (floating point) position along the chromosome,
@@
 Then, we record in a \textbf{Mutation Table}
-\begin{align*}
-    \text{site,} \qquad \text{node,} \qquad \text{derived state}
-\end{align*}
+\begin{center}
+\begin{tabular}{ccc}
+    site  & node  & derived\_state \\
+    0	  & 4	  & 1 \\
+    1	  & 3	  & 1 \\
+    1	  & 2	  & 0 \\
+\end{tabular}
+\end{center}
 in which ``site'' is the ID of the site at wihch this mutation occurred,
@@
 unless another mutation occurs.
-\plr{Omit migrations?}
+% \plr{Omit migrations?}
 
@@
 An example is shown in Figure \ref{fig:method_diagram}.
-\plr{(trying to make the point strongly in the figure, that everything can be stored quite concisely)}
 
-\plr{Note that you can begin the forwards-time simulation with the results of a previous (coalescent) one,
-which is helpful so you don't need to run for so long (cite Wilkins).}
 In this scheme,

### 2017-10-20 sims/benchmark-simuPOP.py
Upgrade to edges; track more data on timing

@@ -192,9 +192,9 @@ class GammaDistributedFitness:
         else:
             return max(0.0, 1. - s)
 
-time_dict = {'prepping': None,
-             'simulating': None,
-             'finalizing': None}
+time_dict = {'time_prepping': None,
+             'time_simulating': None,
+             'time_finalizing': None}
 
 start = time.time()
 
@@ -218,9 +218,10 @@ haploid_labels = [(k,p) for k in first_gen
                         for p in (0,1)]
 node_ids = {x:j for x, j in zip(haploid_labels, init_ts.samples())}
 rc = RecombCollector(ts=init_ts, node_ids=node_ids,
-                     locus_position=locus_position)
+                     locus_position=locus_position,
+                     benchmark=True)
 end_prep = time.time()
-time_dict['prepping'] = end_prep - start
+time_dict['time_prepping'] = end_prep - start
 
 pop.evolve(
     initOps=[
@@ -273,6 +274,8 @@ logfile.write("----------\n")
 logfile.flush()
 
 ts = rc.args.tree_sequence()
+times = rc.args.timings.times
+
 del rc
 
 logfile.write("Loaded into tree sequence!\n")
@@ -295,18 +298,18 @@ logfile.flush()
 
 rng = msprime.RandomGenerator(mut_seed)
 nodes = msprime.NodeTable()
-edgesets = msprime.EdgesetTable()
+edges = msprime.EdgeTable()
 sites = msprime.SiteTable()
-ts.dump_tables(nodes=nodes, edgesets=edgesets)
+ts.dump_tables(nodes=nodes, edges=edges)
 if args.record_neutral:
-    mutated_ts = msprime.load_tables(nodes=nodes, edgesets=edgesets,
+    mutated_ts = msprime.load_tables(nodes=nodes, edges=edges,
                                      sites=sites)
 elif not args.record_neutral:
     mutations = msprime.MutationTable()
     mutgen = msprime.MutationGenerator(rng, args.mut_rate)
-    mutgen.generate(nodes, edgesets, sites, mutations)
+    mutgen.generate(nodes, edges, sites, mutations)
     mutated_ts = msprime.load_tables(
-        nodes=nodes, edgesets=edgesets, sites=sites, mutations=mutations)
+        nodes=nodes, edges=edges, sites=sites, mutations=mutations)
 
 del ts
 
@@ -321,12 +324,25 @@ logfile.write("All done!\n")
 
 end_fin = time.time()
[…]

### 2017-10-20 sims/setup-simuPOP.sh

@@ -1,10 +1,10 @@
 conda config --add channels conda-forge
 conda update -q conda
 conda info -a
-if [[ ! -d "/opt/conda/envs/ftprime" ]] ; then
-	conda create -q -n ftprime python=3.5
+if [[ ! -d "/opt/conda/envs/ftprime-benchmark" ]] ; then
+	conda create -q -n ftprime-benchmark python=3.5
 fi
 wget https://raw.githubusercontent.com/ashander/ftprime/5ad8bcb3a49a19c9cfd52bdf615140af6f83d218/environment.yml
-conda env update -q -n ftprime -f environment.yml
-source activate ftprime
+conda env update -q -n ftprime-benchmark -f environment.yml
+source activate ftprime-benchmark
 pip install --editable=git+https://github.com/ashander/ftprime.git@master#egg=ftprime

### 2017-10-20 forwards_paper.tex
Added n<<N assumption paragraph & refs.

@@
 \newcommand{\plr}[1]{{\em \color{blue} #1}}
+\newcommand{\jk}[1]{{\em \color{red} #1}}
 
@@
 first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
-second, producing analytical methods to estimate parameters of interest from genetic data (e.g. $\theta = 4N_e \mu$);
+second, producing analytical methods to estimate parameters of interest from genetic data (e.g.\ $\theta = 4N_e \mu$);
 and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
@@
 
+Another critical assumption of the coalescent is that the sample size $n$ is much smaller
+than the effective population size $N_e$. With the very large sample sizes becoming common
+in human genetics, this is no longer a safe assumption. For example, a recent study~\citep{martin2017human}
+simulated 600,000 samples of human chromosome 20 to assess the impact of European biased
+reference panels in genome wide association studies. While many results of the coalescent are surpisingly
+robust to violations of the assumption that $n \ll N$, there are instances where genealogical properties
+are distorted and coalescent simulations may be
+misleading~\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
+A hybrid approach, in which the recent history of a large sample of
+individuals is simulated under a detailed forwards-in-time model and the deep history
+under the coalescent, is an attractive prospect.
+\jk{Not sure where this should go; perhaps in the discussion? We can make the point that
+by keeping track of the genealogies forwards in time we can easily combine these with the
+deep history simulated by msprime}
+
 With modern computing power, pure demographic calculations are not a barrier,
@@
 As samples of most species genomes harbor tens or hundreds of millions of variant sites,
-naively carrying full genotypes for even modest numbers of individuals through a simluation becomes quickly prohibitive.
+naively carrying full genotypes for even modest numbers of individuals through a simulation becomes quickly prohibitive.
 
@@
 at each position along the genome.
-In this paper, we show how to use algorithmic tools and data structures developed for the coalescent simulator \msprime
+In this paper, we show how to use algorithmic tools and data structures developed for the coalescent simulator \msprime\
 to efficiently record, and later process, this history.

### 2017-10-20 sims/benchmark-simuPOP.py
Use binary allele

@@ -129,6 +129,9 @@ if args.record_neutral:
 if args.generations is None:
     args.generations = args.popsize * 20
 
+from simuOpt import setOptions
+setOptions(alleleType = 'binary')
+
 import simuPOP as sim
 
 sim.setRNG(seed=args.seed)

### 2017-10-20 sims/benchmark-simuPOP.py
Update csv output; don't log by default

@@ -2,11 +2,11 @@
 import gzip
 import sys
 import time
-import csv
 import random
 from ftprime import RecombCollector
 import msprime
 import argparse
+import pandas as pd
 
 description = '''
 Benchmark
@@ -107,9 +107,9 @@ parser.add_argument("--gamma_scale","-b", type=float, dest="gamma_scale",
 parser.add_argument("--gc", "-G", dest="simplify_interval", type=int,
         help="Interval between simplify steps.", default=500)
 parser.add_argument("--logfile","-g", type=str, dest="logfile",
-        help="name of log file (or '-' for stdout)",default="-")
-parser.add_argument("--csvfile","-c", type=str, dest="csvfile",
-        help="name of csv file")
+        help="name of log file (or '-' for stdout)",default="/dev/null")
+parser.add_argument("--outfile1","-o", type=str, dest="outfile1",
+        help="name of csv file to write")
 parser.add_argument("--seed", "-d", dest="seed", type=int, help="random seed")
 
 # optional args
@@ -138,7 +138,6 @@ sim.setRNG(seed=args.seed)
 random.seed(args.seed)
 
 logfile = fileopt(args.logfile, "w")
-csvfile = fileopt(args.csvfile, "w+")
 
 logfile.write("Options:\n")
 logfile.write(str(args)+"\n")
@@ -195,11 +194,7 @@ class GammaDistributedFitness:
         else:
             return max(0.0, 1. - s)
 
-time_dict = {'time_prepping': None,
-             'time_simulating': None,
-             'time_finalizing': None}
-
-start = time.time()
+before_time = time.time()
 
 init_geno=[sim.InitGenotype(freq=1.0)]
 
@@ -223,8 +218,6 @@ node_ids = {x:j for x, j in zip(haploid_labels, init_ts.samples())}
 rc = RecombCollector(ts=init_ts, node_ids=node_ids,
                      locus_position=locus_position,
                      benchmark=True)
-end_prep = time.time()
-time_dict['time_prepping'] = end_prep - start
 
 pop.evolve(
     initOps=[
@@ -254,8 +247,8 @@ pop.evolve(
     ],
     gen = args.generations
 )
-end_sim = time.time()
-time_dict['total_runtime'] = end_sim - end_prep
+
+tsim = time.time()- before_time
 
 logfile.write("Done simulating!\n")
 logfile.write(time.strftime('%X %x %Z')+"\n")
[…]

### 2017-10-20 sims/test.sh

@@ -1,2 +1,2 @@
 #!/bin/bash
-python3 benchmark-simuPOP.py --seed 43 --nsam 10 --pdel 0.1 --theta 1 --rho 1 -N 100 --csvfile out.csv
+python3 benchmark-simuPOP.py --seed 43 --nsam 10 --pdel 0.1 --theta 1 --rho 1 -N 100 --outfile1 out.csv.gz

### 2017-10-20 sims/benchmark-simuPOP.py
Cleanup output; make clear we record total_time
total time not time **in** sim, which differs from KT

@@ -109,7 +109,7 @@ parser.add_argument("--gc", "-G", dest="simplify_interval", type=int,
 parser.add_argument("--logfile","-g", type=str, dest="logfile",
         help="name of log file (or '-' for stdout)",default="/dev/null")
 parser.add_argument("--outfile1","-o", type=str, dest="outfile1",
-        help="name of csv file to write")
+        help="name of csv file to write timing statisitcs.")
 parser.add_argument("--seed", "-d", dest="seed", type=int, help="random seed")
 
 # optional args
@@ -120,14 +120,18 @@ parser.add_argument("--mut_rate","-U", type=float, dest="mut_rate",
 parser.add_argument("--treefile","-t", type=str, dest="treefile",
         help="name of output file for trees (default: not output)",default=None)
 parser.add_argument("--generations","-T", type=int, dest="generations",
-        help="number of generations to run for (default 20N)", default=None)
+        help="number of generations to run for (default SIMLEN * N)",
+                    default=None)
+parser.add_argument("--simlen","-s", type=int, dest="multiplier",
+        help="multiplier of popsize to run for (unless `generations` is specified)",
+                    default=10)
 
 args = parser.parse_args()
 if args.record_neutral:
     sys.exit(1)
 
 if args.generations is None:
-    args.generations = args.popsize * 20
+    args.generations = args.popsize * args.multiplier
 
 from simuOpt import setOptions
 setOptions(alleleType = 'binary')
@@ -194,7 +198,7 @@ class GammaDistributedFitness:
         else:
             return max(0.0, 1. - s)
 
-before_time = time.time()
+before_time = time.process_time()
 
 init_geno=[sim.InitGenotype(freq=1.0)]
 
@@ -248,7 +252,7 @@ pop.evolve(
     gen = args.generations
 )
 
-tsim = time.time()- before_time
+tsim = time.process_time()- before_time
 
 logfile.write("Done simulating!\n")
 logfile.write(time.strftime('%X %x %Z')+"\n")
@@ -276,14 +280,12 @@ args.gc = args.simplify_interval  # so use same name below
 
 ## begin block similar to K Thornton's
 # Take times from args before they change.
-times['fwd_sim_runtime'] = tsim
-times['N'] = args.popsize
-times['theta'] = args.theta
-times['rho'] = args.rho
[…]

### 2017-10-20 forwards_paper.tex

@@
 % without approximation, across whole chromosomes \citep{msprime}.
+\jk{I find this confusing I'm afraid. I agree we should mention the ARG but I think only in passing.
+I have never found a situation in which an ARG made things easier to understand!}
 
@@
 %%%%%%
+\subsection*{Tree sequences}
+A tree sequence is an efficient encoding for a sequence of correlated trees such as
+is produced by recombination. The encoding is efficient because branches that are
+shared by adjacent trees are stored once, rather than repeatedely for each tree.
+The topology of a tree sequence is defined via a set of \emph{nodes} and
+\emph{edges}. A node in a tree sequence refers to a distinct ancestor and
+corresponds to the vertices in the individual genealogies along the sequence.
+Since each node represents a specific ancestor, it has a unique ``time'',
+thought of as her birth time, which determines the height of any vertices
+she is associated with. In the example of Figure~\ref{fig:example_tree_sequence}
+we have a total of five nodes. Nodes 0, 1 and 2 occur at time 0 and are our
+samples. Nodes 3 and 4 represent the ancestors of these samples, and were born at
+time 1.5 and 2.5 respectively.
+
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=0.6\textwidth]{example_tree_sequence}
+    \end{center}
+    \caption{
+        An example tree sequence with three samples over a chromosome of length 10. In the
+        left-most panel we show the tree sequence pictorially in two different ways:
+        in the top are the tree topologies and the bottom shows the spatial extent of the
+        edges that define these topologies. The right-hand panels show the specific encoding
+        of this tree sequence in the four tables (nodes, edges, sites and mutations) defined
+        by \msprime.
+        \label{fig:example_tree_sequence}
+    }
+\end{figure}
+
+The topology of a tree sequence is defined by the \emph{edges} which define how
+nodes relate to each other over specific genomic intervals. Each edge is a tuple
+$(\ell, r, p, c)$, where $[\ell, r)$ is a half-open genomic interval defining the
+spatial extend of the edge; $p$ and $c$ are the parent and child nodes specifying
+a single branch in the tree over this interval. The spatial extent of the
+edges defining the topology of Figure~\ref{fig:example_tree_sequence} in the
+bottom left panel. For example, the branch joining nodes 1 to 3 is shared in both trees,
+and is recorded in a single edge extending over the whole chromosome. It is this
+method of capturing the shared structure between adjacent trees that makes the
+tree sequence encoding very compact and algorithmically efficient.
+
+\jk{Continue along these lines with sites and mutations after updating the figure}
+
+\jk{Other stuff we can review here:}
+\begin{enumerate}
+\item Encoding a tree sequence topology like this is very efficient, much, much more compact than newick.
+Can refer to example in msprime paper, or could do a large running example here. Say 500K samples of
+human CHR1, something like that. Newick will be hundreds of TB.
+
+\item Can also make the point that it's much more efficient than, say, VCF. We could drop some
+mutations on the example above, and then report how big it would be if encoded with VCF and BCF.
+
+\item The \msprime\ API is very efficient for working with these large datasets. Examples on
+how long it takes to iterate over all the trees, compute pi, etc.
+
+\item The point of all this is to emphasise that not only is it efficient to generate the
+data as a tree sequence, it's also very efficient to store and process it this way too. By the
+time we get to the end of this section we should have convinced the reader of this.
+
+\end{enumerate}
+
 \subsection*{Data structures:}
@@
 First we describe the data structures we use for recording genealogical history,
-as implemented in \msprime{}.
-These derive from those described in \citet{msprime},
-but have been modified to remove redundancy.
+as implemented in \msprime.
+These derive from those described by \citet{kelleher2016efficient},
+but have been generalised and modified to remove redundancy.
 The tables below give the example tree sequence of Figure \ref{fig:ex_tree_seq}.
@@
 tree relating all samples to each other at any point along the genome.
-In the context of a tree sequence,
-\emph{nodes} refer to distinct ancestors,
-and may be identified with points in the trees of a tree sequence.
-Every tip or branching point in a tree is associated with a node.
-A tree sequence does not necessarily carry complete genealogical information about every node,
-but calling a node a \emph{sample} implies that complete information is present for that ancestor,
-i.e., the node is associated with a point in every tree in the tree sequence.
-Since each node represents a certain ancestor, it has a
-unique ``time'', thought of as her birth time, which determines the height
-of any branching points she is associated with.  A given node will be
+In the context of a tree sequence, \emph{nodes} refer to distinct ancestors,
+and correspond to the vertices in the trees of a tree sequence.
+Since each node represents a certain ancestor, it has a unique ``time'', thought of as her birth time,
+which determines the height of any branching points she is associated with.  A given node will be
 associated with branching points of all trees across a region if that node

### 2017-10-23 forwards_paper.tex
Added asy source for rough version of diagram.
Also short discussion of sites and mutations.

@@
     Here we show how to efficiently produce and store the entire history of ancestry and recombination
-    (the ARG) from an individual-based simulation,
+    from an individual-based simulation,
     on which neutral mutations can be placed afterwards.
@@
 who inherited which parts of which parental chromosomes.
-This embellished graph is known as the \emph{ancestral recombination graph}, or {ARG} \citep{griffiths}.
+This embellished pedigree contains all the information necessary
+to construct the genealogical tree that relates each individual to each other
+at each position on the genome, i.e., the \emph{tree sequence}.
 Combined with ancestral genotypes and the origins of new mutations,
-it completely specifies the genomic sequence of any individual in the population at any time.
-However, much less than the entire ARG is needed to specify relationships
-between any given set of samples --
-only those portions of it from which those samples have actually inherited,
-back to their most recent common ancestors.
-The assumptions of coalescent theory
-% -- random mating and neutrality --
-imply that this random set describes a Markov process looked at \emph{backwards} in time,
-and can hence be simulated without reference to the unnecessary remainder
-\citet{griffithsmarjoram,msprime}.
-This stochastic process is also called the ``ARG'',
-but we reserve this name for the embellished graph itself.
-% Although this point was in principle known since \citet{griffithsmarjoram},
-% only recent algorithmic advances made it possible to actually simulate the process correctly,
-% without approximation, across whole chromosomes \citep{msprime}.
-\jk{I find this confusing I'm afraid. I agree we should mention the ARG but I think only in passing.
-I have never found a situation in which an ARG made things easier to understand!}
+it also completely specifies the genomic sequence of any individual in the population at any time.
+This is much more than we need to know, however,
+so we discard all information irrelevant to the genetic history
+of the \emph{sampled} individuals,
+which results in considerable savings.
+
+Another way of representing this same information
+is known as the \emph{ancestral recombination graph}, or {ARG} \citep{griffiths1997ancestral},
+which has been the subject of substantial study
+under the assumptions of coalescent theory \citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
 
@@
     \begin{center}
-        \includegraphics[width=0.6\textwidth]{example_tree_sequence}
+        \includegraphics[width=\textwidth]{example_tree_sequence}
     \end{center}
@@
         by \msprime.
+        \jk{The figure needs to be rescaled a bit to give more space to the trees. We could also use a
+        time scale next to the trees.}
         \label{fig:example_tree_sequence}
@@
 
-\jk{Continue along these lines with sites and mutations after updating the figure}
+Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
+encode the sequence information for each sample in an efficient way. Each site
+is associated with a position on the genome and an ancestral state. For example,
+in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
+2 and ancestral state `A' and the other at position 7 with ancestral state `G'. If
+no mutations occur, all samples inherit the ancestral state at a given site.
+A mutation occurs over a specific node at a given site, and results in a specific
+derived state. Thus, all samples below the mutation node in the tree will inherit
+this state (unless further mutations are encountered). Three mutations are shown
+in Figure~\ref{fig:example_tree_sequence}, illustrated by red hexagons. The first
+mutation occurs at site zero (which is in the left-hand tree), which is a simple
+mutation resulting in node $2$ inheriting the state 'T'. The second side (in the
+right hand tree) has two mutations: one occuring over node $3$ changing the state to
+`C' and a back mutation over node $1$ changing the state to `G'.
+\jk{Not sure if I'm labouring the point here: this should be reasonably obvious.
+Perhaps if we added the state changes to the trees, and the states of the leaves
+below the samples, we could get rid of all the example discussion here?}
 
@@
 
+%%%%%%
+\subsection*{Recording the pedigree in forwards time}
+
+To record the genealogical history of a simulation
+we need to record two things for each new chromosome:
+the birth time,
+and the endpoints and parental IDs of each distinctly inherited segment.
+This is recorded easily, without further processing,
+in the tables described above.
+
+For concreteness, here we write out in pseudocode
+how to run a neutral Wright--Fisher simulation
+with overlapping generations
+that records genealogical history in this way.
+The simulation will run for $T$ generations,
+and has $N$ haploid individuals, each carrying a single chromosome of length L,
+on which for simplicity we assume there is exactly one crossover per generation.
+The probability of death per individual each generation is \texttt{death\_prob},
+and the whole-chromosome mutation rate per individual is \texttt{mut\_rate}.
+
+\paragraph{Initialize:}
+We will build the tables \nodes, \edgesets, \sites, and \mutations,
+and keep track of the IDs of the current generation in the list \texttt{pop}.
+\begin{lstlisting}[language=Python]
+for i in 0:N-1:
+    nodes.add_row(time=T)
+    pop[i] = i
+\end{lstlisting}
+(Alternatively, the tables could be initialized by the results of a coalescent simulation.)
+
+\paragraph{Iterate:}
+We then step through the generations,
+using the functions \texttt{random\_mutation} and \texttt{random\_allele}
+to choose positions and alleles for mutations, respectively:
+\begin{lstlisting}[language=Python]
+for t in 1:T:
+    for i in 0:N-1:
+        if random.uniform() > death_prob:
+            new_pop[i] = pop[i]
+        else:
+            u = nodes.num_rows  # the ID of the new individual
+            new_pop[i] = u
+            nodes.add_row(time = T-t)
+            a = random.sample(pop)
+            b = random.sample(pop)
+            bp = random.sample(0:L)
+            if bp > 0:
+                edgesets.add_row(left=0, right=bp,
+                                  parent=a, children=(u,))
+            if bp < L:
+                edgesets.add_row(left=bp, right=L,
+                                  parent=b, children=(u,))
+            num_muts = random.poisson(mut_rate)
+            for j in 1:num_muts:
+                pos = random_mutation()
+                if pos not in sites.position:
+                    s = sites.num_rows
+                    sites.add_row(position=pos,
+                           ancestral_state=random_allele())
+                else:
+                    s = sites.position.index(pos)
+                mutations.add_row(site=s, node=u,
+                           derived_state=random_allele())
+    pop = new_pop
+\end{lstlisting}
+
+\paragraph{Finalize:}
+to obtain a tree sequence, we need only transform the tables to the required format:
+\begin{lstlisting}
+sort_tables(nodes, edgesets, sites, mutations)
+simplify_tables(nodes, edgesets, sites, mutations)
+\end{lstlisting}
+which can then be loaded into a tree sequence with \msprime.
+(These two functions operate on the tables in place.)
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{method_diagram}
+    \end{center}
+    \caption{
+        A simple example of the method.
+        \textbf{Top:} the diagram on the left relates ten haploid individuals to each other.
+        It is recorded, in forwards time,
+        in 10 node records (one for each individual)
+        and 14 edgeset records (one for each distinctly inherited segment).
+        Blue numbers denote crossing over locations in each meiosis.
+        The individuals $B$, $C$, and $D$ inherit clonally from $A$
+        to ensure rootnedness of resulting tree.
+        \textbf{Center:} the five distinct trees relating all individuals to each other
+        found across the chromosome (blue numbers denote locations on the chromosome).
+        Labels after simplification are shown in red.
+        \textbf{Bottom:} tables recording the tree sequence after simplification
+        with nodes $I$ and $J$ as samples.
+        The mapping from labels in the forwards time simulation to nodes in the tree sequence
+        is shown in red,
+        which allows additional records to be added as the simulation progresses.
+        \label{fig:method_diagram}
+    }
+\end{figure}
+
+%%%%%%
+\subsection*{Tree sequence simplification}
+
+The resulting tables encode everything we need to know --
+in fact, they record all of history for everyone alive at any time through the simulation.
+This is much more than we need to reconstruct the genealogies and sequences
+of a smaller sample of indidivuals.
+Reducing this larger tree sequence to a smaller one relevant to a given set of ``sample'' nodes
+we call \emph{simplification}.
+Roughly, this works by tracing ancestry from the samples backwards through the recorded history,
+adding node and edgeset records to the output only when coalescent events are reached.
+This works exactly as in \msprime, allowing substantial re-use of algorithms;
+the main difference being that parental choice, mutations, and recombination locations
+are determined by the input tree sequence
+rather than randomly generated.
+An example is shown in Figure \ref{fig:method_diagram}.
+
+In this scheme,
+at any point in the simulation genealogical history is recorded in a tree sequence.
+This has two additional advantages.
+First, simplification can be run periodically through the simulation,
+taking the set of samples to be the entire currently alive population.
+This is important as it keeps memory usage from growing linearly (and quickly) with time.
+Second, the simulation can be \emph{begun} with a tree sequence produced by some other method --
+for instance, by a coalescent simulation with \msprime.
+This allows for incorporation of deep-time history beyond the reach of individual-based simulations.
+Since geographic structure from times longer ago than the mixing time of migration across the range
+has limited effect on modern genealogies \citep{wilkins_separation}
+(other than possibly changing effective population size \citet{durretspatial}),
+this may not negatively affect realism.
+
+\plr{Something about putting mutations down on the tree sequence.}
+
+%%%%%%
+\subsection*{Overview of the API}
+
+\plr{Quick overview of how to efficiently hook this up with other code.}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Results}
+
+\plr{Estimates of run-time complexity}
+Suppose that we wish to run a forwards-time simulation of $N$ individuals for $T$ generations,
+in which there are $S$ selected loci and $L$ neutral loci.
+We will estimate run-time complexity and memory usage for both a ``naive'' strategy that carries along neutral loci
+and an ``ancestry-tracking'' strategy like that we consider here.
+To do this, we assume that each individual must carry along its entire genotype.
+More advanced schemes are used in some simulators,
+but these increase efficiency by utilizing redundancy introduced by shared ancestry,
+which is effectively an intermediate scheme.
+We omit the cost of computing a fitness function.
+
+Both schemes must choose mates and recombination breakpoints,
+and pass on selected genotypes.
+The difference between the two comes from the tradeoff between
+(a) passing on neutral genotypes, and
+(b) recording and simplifying the tree sequence, and adding neutral genotypes afterwards.
+(We assume here that selected genotypes are stored in the same way for both.)
+Passing on neutral genotypes naively records $L$ items per individual each generation, discarding the previous generation.
+\plr{do this better with numbers from nodes, edgesets}
+Recording a tree sequence stores 2 parents and $\rho$ breakpoints on average each generation;
+after $T$ generations this grows to $N \times T \times (2 + \rho)$ stored items.
+However, after simplification a tree sequence for all $N$ individuals
+only takes of order $N + \rho \log N$ records.
+Simplification requires processing each of the initial records -- so, order of $N \times T \times \rho$ operations;
+
+% \begin{table}
+% \begin{center}
+%     \begin{tabular}{l||r|r||r|r|}
+%                     & \multicolumn{2}{|c|}{naive} & \multicolumn{2}{|c|}{ancestry-tracking} \\
+%                     & operations  & memory & operations  & memory \\
+%         mate choice & $N$       &   --  &   $N$     & -- \\
+%         recombination & $N\rho$ &   --  &   $N\rho$ & -- \\
+%     \end{tabular}
+% \end{center}
+% \end{table}
+
+Comparison of simulation with/without msprime, using simuPOP
+or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
+
+Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
+so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
+in a day.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Conclusion}
+
+This is a general-purpose strategy that can be applied to other methods.
+
+All sorts of good reasons to want to have whole-genome simulations.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements}
+
+\bibliography{references}
+
+\appendix
+
 \subsection*{Data structures:}
+\jk{Moved from above.}
 
@@
     \caption{
-        A pictorial representation of the ARG
+        A pictorial representation of a tree sequence
         relating three samples to each other
@@
 
-%%%%%%
-\subsection*{Recording the ARG in forwards time}
-
-To record the genealogical history of a simulation
-we need to record two things for each new chromosome:
-the birth time,
-and the endpoints and parental IDs of each distinctly inherited segment.
-This is recorded easily, without further processing,
-in the tables described above.
-
-For concreteness, here we write out in pseudocode
-how to run a neutral Wright--Fisher simulation
-with overlapping generations
-that records genealogical history in this way.
-The simulation will run for $T$ generations,
-and has $N$ haploid individuals, each carrying a single chromosome of length L,
-on which for simplicity we assume there is exactly one crossover per generation.
-The probability of death per individual each generation is \texttt{death\_prob},
-and the whole-chromosome mutation rate per individual is \texttt{mut\_rate}.
-
-\paragraph{Initialize:}
-We will build the tables \nodes, \edgesets, \sites, and \mutations,
-and keep track of the IDs of the current generation in the list \texttt{pop}.
-\begin{lstlisting}[language=Python]
-for i in 0:N-1:
-    nodes.add_row(time=T)
-    pop[i] = i
-\end{lstlisting}
-(Alternatively, the tables could be initialized by the results of a coalescent simulation.)
-
-\paragraph{Iterate:}
-We then step through the generations,
-using the functions \texttt{random\_mutation} and \texttt{random\_allele}
-to choose positions and alleles for mutations, respectively:
-\begin{lstlisting}[language=Python]
-for t in 1:T:
-    for i in 0:N-1:
-        if random.uniform() > death_prob:
-            new_pop[i] = pop[i]
-        else:
-            u = nodes.num_rows  # the ID of the new individual
-            new_pop[i] = u
-            nodes.add_row(time = T-t)
-            a = random.sample(pop)
-            b = random.sample(pop)
-            bp = random.sample(0:L)
-            if bp > 0:
-                edgesets.add_row(left=0, right=bp,
-                                  parent=a, children=(u,))
-            if bp < L:
-                edgesets.add_row(left=bp, right=L,
-                                  parent=b, children=(u,))
-            num_muts = random.poisson(mut_rate)
-            for j in 1:num_muts:
-                pos = random_mutation()
-                if pos not in sites.position:
-                    s = sites.num_rows
-                    sites.add_row(position=pos,
-                           ancestral_state=random_allele())
-                else:
-                    s = sites.position.index(pos)
-                mutations.add_row(site=s, node=u,
-                           derived_state=random_allele())
-    pop = new_pop
-\end{lstlisting}
-
-\paragraph{Finalize:}
-to obtain a tree sequence, we need only transform the tables to the required format:
-\begin{lstlisting}
-sort_tables(nodes, edgesets, sites, mutations)
-simplify_tables(nodes, edgesets, sites, mutations)
-\end{lstlisting}
-which can then be loaded into a tree sequence with \msprime.
-(These two functions operate on the tables in place.)
-
-\begin{figure}
-    \begin{center}
-        \includegraphics{method_diagram}
-    \end{center}
-    \caption{
-        A simple example of the method.
-        \textbf{Top:} the ARG shown on the left relates ten haploid individuals to each other.
-        It is recorded, in forwards time,
-        in 10 node records (one for each individual)
-        and 14 edgeset records (one for each distinctly inherited segment).
-        Blue numbers denote crossing over locations in each meiosis.
-        The individuals $B$, $C$, and $D$ inherit clonally from $A$
-        to ensure rootnedness of resulting tree.
-        \textbf{Center:} the five distinct trees relating all individuals to each other
-        found across the chromosome (blue numbers denote locations on the chromosome).
-        Labels after simplification are shown in red.
-        \textbf{Bottom:} tables recording the tree sequence after simplification
-        with nodes $I$ and $J$ as samples.
-        The mapping from labels in the forwards time simulation to nodes in the tree sequence
-        is shown in red,
-        which allows additional records to be added as the simulation progresses.
-        \label{fig:method_diagram}
-    }
-\end{figure}
-
-%%%%%%
-\subsection*{Tree sequence simplification}
-
-The resulting tables encode everything we need to know --
-in fact, they record all of history for everyone alive at any time through the simulation.
-This is much more than we need to reconstruct the genealogies and sequences
-of a smaller sample of indidivuals.
-Reducing this larger tree sequence to a smaller one relevant to a given set of ``sample'' nodes
-we call \emph{simplification}.
-Roughly, this works by tracing ancestry from the samples backwards through the recorded history,
-adding node and edgeset records to the output only when coalescent events are reached.
-This works exactly as in \msprime, allowing substantial re-use of algorithms;
-the main difference being that parental choice, mutations, and recombination locations
-are determined by the input tree sequence
-rather than randomly generated.
-An example is shown in Figure \ref{fig:method_diagram}.
-
-In this scheme,
-at any point in the simulation genealogical history is recorded in a tree sequence.
-This has two additional advantages.
-First, simplification can be run periodically through the simulation,
-taking the set of samples to be the entire currently alive population.
-This is important as it keeps memory usage from growing linearly (and quickly) with time.
-Second, the simulation can be \emph{begun} with a tree sequence produced by some other method --
-for instance, by a coalescent simulation with \msprime.
-This allows for incorporation of deep-time history beyond the reach of individual-based simulations.
-Since geographic structure from times longer ago than the mixing time of migration across the range
-has limited effect on modern genealogies \citep{wilkins_separation}
-(other than possibly changing effective population size \citet{durretspatial}),
-this may not negatively affect realism.
-
-\plr{Something about putting mutations down on the tree sequence.}
-
-%%%%%%
-\subsection*{Overview of the API}
-
-\plr{Quick overview of how to efficiently hook this up with other code.}
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Results}
-
-\plr{Estimates of run-time complexity}
-Suppose that we wish to run a forwards-time simulation of $N$ individuals for $T$ generations,
-in which there are $S$ selected loci and $L$ neutral loci.
-We will estimate run-time complexity and memory usage for both a ``naive'' strategy that carries along neutral loci
-and an ``ancestry-tracking'' strategy like that we consider here.
-To do this, we assume that each individual must carry along its entire genotype.
-More advanced schemes are used in some simulators,
-but these increase efficiency by utilizing redundancy introduced by shared ancestry,
-which is effectively an intermediate scheme.
-We omit the cost of computing a fitness function.
-
-Both schemes must choose mates and recombination breakpoints,
-and pass on selected genotypes.
-The difference between the two comes from the tradeoff between
-(a) passing on neutral genotypes, and
-(b) recording and simplifying the tree sequence, and adding neutral genotypes afterwards.
-(We assume here that selected genotypes are stored in the same way for both.)
-Passing on neutral genotypes naively records $L$ items per individual each generation, discarding the previous generation.
-\plr{do this better with numbers from nodes, edgesets}
-Recording a tree sequence stores 2 parents and $\rho$ breakpoints on average each generation;
-after $T$ generations this grows to $N \times T \times (2 + \rho)$ stored items.
-However, after simplification a tree sequence for all $N$ individuals
-only takes of order $N + \rho \log N$ records.
-Simplification requires processing each of the initial records -- so, order of $N \times T \times \rho$ operations;
-
-% \begin{table}
-% \begin{center}
-%     \begin{tabular}{l||r|r||r|r|}
-%                     & \multicolumn{2}{|c|}{naive} & \multicolumn{2}{|c|}{ancestry-tracking} \\
-%                     & operations  & memory & operations  & memory \\
-%         mate choice & $N$       &   --  &   $N$     & -- \\
-%         recombination & $N\rho$ &   --  &   $N\rho$ & -- \\
-%     \end{tabular}
-% \end{center}
-% \end{table}
-
-Comparison of simulation with/without msprime, using simuPOP
-or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
-
-Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
-so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
-in a day.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Conclusion}
-
-This is a general-purpose strategy that can be applied to other methods.
-
-All sorts of good reasons to want to have whole-genome simulations.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements}
-
-\bibliography{references}
-
-\appendix
-
-\section{More general method for recording the ARG}
+\section{More general method for recording the pedigree}
 

### 2017-10-23 forwards_paper.tex
changed ARG references

@@
     \begin{center}
-        \includegraphics[width=\textwidth]{example_tree_sequence}
+        \includegraphics[width=0.6\textwidth]{example_tree_sequence}
     \end{center}
@@
         by \msprime.
-        \jk{The figure needs to be rescaled a bit to give more space to the trees. We could also use a
-        time scale next to the trees.}
         \label{fig:example_tree_sequence}
@@
 
-Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
-encode the sequence information for each sample in an efficient way. Each site
-is associated with a position on the genome and an ancestral state. For example,
-in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
-2 and ancestral state `A' and the other at position 7 with ancestral state `G'. If
-no mutations occur, all samples inherit the ancestral state at a given site.
-A mutation occurs over a specific node at a given site, and results in a specific
-derived state. Thus, all samples below the mutation node in the tree will inherit
-this state (unless further mutations are encountered). Three mutations are shown
-in Figure~\ref{fig:example_tree_sequence}, illustrated by red hexagons. The first
-mutation occurs at site zero (which is in the left-hand tree), which is a simple
-mutation resulting in node $2$ inheriting the state 'T'. The second side (in the
-right hand tree) has two mutations: one occuring over node $3$ changing the state to
-`C' and a back mutation over node $1$ changing the state to `G'.
-\jk{Not sure if I'm labouring the point here: this should be reasonably obvious.
-Perhaps if we added the state changes to the trees, and the states of the leaves
-below the samples, we could get rid of all the example discussion here?}
+\jk{Continue along these lines with sites and mutations after updating the figure}
 
@@
 \end{enumerate}
+
+\subsection*{Data structures:}
+
+First we describe the data structures we use for recording genealogical history,
+as implemented in \msprime.
+These derive from those described by \citet{kelleher2016efficient},
+but have been generalised and modified to remove redundancy.
+The tables below give the example tree sequence of Figure \ref{fig:ex_tree_seq}.
+
+\begin{figure}
+    \begin{center}
+        \includegraphics[width=0.6\textwidth]{example_tree_seq}
+    \end{center}
+    \caption{
+        A pictorial representation of a tree sequence
+        relating three samples to each other
+        over a chromosome of length 1.0.
+        Mutations shown in the example tables of the text
+        are marked with red stars.
+        \label{fig:ex_tree_seq}
+    }
+\end{figure}
+
+To clarify terminology,
+below a \emph{tree} refers to
+a genealogical tree describing how a collection of individuals are related to each other.
+A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
+tree relating all samples to each other at any point along the genome.
+In the context of a tree sequence, \emph{nodes} refer to distinct ancestors,
+and correspond to the vertices in the trees of a tree sequence.
+Since each node represents a certain ancestor, it has a unique ``time'', thought of as her birth time,
+which determines the height of any branching points she is associated with.  A given node will be
+associated with branching points of all trees across a region if that node
+is the most recent common ancestor to the subtending tips across that region.
+This information is stored in the columns of a \textbf{Node Table}:
+\begin{center}
+\begin{tabular}{cccc}
+    id   &  is\_sample &  population &  time \\
+    0    &  1          & 0           & 0 \\
+    1    &  1          & 0           & 0 \\
+    2    &  1          & 0           & 0 \\
+    3    &  0          & 0           & 0.3 \\
+    4    &  0          & 0           & 0.5 \\
+    5    &  0          & 0           & 0.6 \\
+    6    &  0          & 0           & 0.7 \\
+\end{tabular}
+\end{center}
+where ``flags'' records other information
+(e.g., a binary mask of `1' indicates the node is a sample).
+Importantly, the \textbf{node ID} of a node is given implicitly
+by the (zero-based) index of its corresponding row in the Node Table.
+
+Tree sequences are constructed by specifying over which segments of genome
+which nodes inherit from which other nodes.  This information is stored by
+recording the endpoints of each distinctly inherited ancestral segment,
+the parental node, and a list of children nodes who have inherited that segment.
+As each such record describes a collection of edges across a swatch of trees in the tree sequence,
+we call these \emph{edgesets}
+and store them in the columns of an \textbf{Edgeset Table:}
+\begin{center}
+\begin{tabular}{cccc}
+    left & right & parent & children \\
+    0.2  &  0.8  &  3     & 0,2 \\
+    0.0  &  0.2  &  4     & 1,2 \\
+    0.2  &  0.8  &  4     & 1,3 \\
+    0.8  &  1.0  &  4     & 1,2 \\
+    0.8  &  1.0  &  5     & 0,4 \\
+    0.0  &  0.2  &  6     & 0,4  \\
+\end{tabular}
+\end{center}
+
+To record information about genetic variants we need to also record each mutation
+and which nodes have inherited that mutation.
+The tree structure takes care of inheritance -- all we need to do
+is to record the highest node in the tree at the mutated site
+that inherited that mutation.
+As more than one mtuation may occur at a given site,
+we separate this information into two tables,
+first, the \textbf{Site Table} records for each variant site
+\begin{center}
+\begin{tabular}{ccc}
+    id  &   position  & ancestral\_state \\
+    0   &  0.1        & 0 \\
+    1   &  0.5        & 0 \\
+\end{tabular}
+\end{center}
+Here ``position'' is a (floating point) position along the chromosome,
+and ``ancestral state'' is the genotype of the root of the tree at that site.
+As for nodes, \textbf{site IDs} are given implicitly
+by the (zero-based) index of the rows.
+Then, we record in a \textbf{Mutation Table}
+\begin{center}
+\begin{tabular}{ccc}
+    site  & node  & derived\_state \\
+    0	  & 4	  & 1 \\
+    1	  & 3	  & 1 \\
+    1	  & 2	  & 0 \\
+\end{tabular}
+\end{center}
+in which ``site'' is the ID of the site at wihch this mutation occurred,
+``node'' is the ID of the highest node that has inherited this mutation,
+and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
+unless another mutation occurs.
+% \plr{Omit migrations?}
+
+\paragraph{Definition of valid tables}
+Here are the formal requirements for a set of nodes and edgesets to make sense,
+and to allow ``msprime``'s algorithms to work properly.
+
+To disallow time travel and multiple inheritance:
+\begin{enumerate}
+    \item Offspring must be born after their parents (and hence, no loops).
+    \item The set of intervals on which each individual is a child must be disjoint.
+\end{enumerate}
+For algorithmic reasons, we also require:
+\begin{enumerate} \setcounter{enumi}{2}
+    \item The leftmost endpoint of each chromosome is 0.0.
+    \item Node times must be strictly greater than zero.
+    \item The list of offspring in an edgeset must be sorted.
+    \item Edgesets must be sorted in nondecreasing time order.
+    \item The set of intervals on which each individual is a parent must be disjoint.
+    \item Each edgeset must contain at least two children.
+\end{enumerate}
+Note that since each node time is equal to the amount of time since the \emph{birth} of the
+corresponding parent, time is measured in clock time, not in meioses.
+
+A forwards-time simulation does \textbf{not} naturally emit genealogical information
+satisfying requirements 5--8.
+However, \msprime{} implements two algorithms that will take a set of tables
+satisfying only 1--4 and produce tables satisfying all requirements.
+Trivially, \texttt{sort\_tables} enforces requirements 5 and 6 and does not renumber nodes;
+then, \texttt{simplify} enforces requirements 7 and 8 (and does much more; see below).
 
@@
 
-\subsection*{Data structures:}
-\jk{Moved from above.}
-
-First we describe the data structures we use for recording genealogical history,
-as implemented in \msprime.
-These derive from those described by \citet{kelleher2016efficient},
-but have been generalised and modified to remove redundancy.
-The tables below give the example tree sequence of Figure \ref{fig:ex_tree_seq}.
-
-\begin{figure}
-    \begin{center}
-        \includegraphics[width=0.6\textwidth]{example_tree_seq}
-    \end{center}
-    \caption{
-        A pictorial representation of a tree sequence
-        relating three samples to each other
-        over a chromosome of length 1.0.
-        Mutations shown in the example tables of the text
-        are marked with red stars.
-        \label{fig:ex_tree_seq}
-    }
-\end{figure}
-
-To clarify terminology,
-below a \emph{tree} refers to
-a genealogical tree describing how a collection of individuals are related to each other.
-A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
-tree relating all samples to each other at any point along the genome.
-In the context of a tree sequence, \emph{nodes} refer to distinct ancestors,
-and correspond to the vertices in the trees of a tree sequence.
-Since each node represents a certain ancestor, it has a unique ``time'', thought of as her birth time,
-which determines the height of any branching points she is associated with.  A given node will be
-associated with branching points of all trees across a region if that node
-is the most recent common ancestor to the subtending tips across that region.
-This information is stored in the columns of a \textbf{Node Table}:
-\begin{center}
-\begin{tabular}{cccc}
-    id   &  is\_sample &  population &  time \\
-    0    &  1          & 0           & 0 \\
-    1    &  1          & 0           & 0 \\
-    2    &  1          & 0           & 0 \\
-    3    &  0          & 0           & 0.3 \\
-    4    &  0          & 0           & 0.5 \\
-    5    &  0          & 0           & 0.6 \\
-    6    &  0          & 0           & 0.7 \\
-\end{tabular}
-\end{center}
-where ``flags'' records other information
-(e.g., a binary mask of `1' indicates the node is a sample).
-Importantly, the \textbf{node ID} of a node is given implicitly
-by the (zero-based) index of its corresponding row in the Node Table.
-
-Tree sequences are constructed by specifying over which segments of genome
-which nodes inherit from which other nodes.  This information is stored by
-recording the endpoints of each distinctly inherited ancestral segment,
-the parental node, and a list of children nodes who have inherited that segment.
-As each such record describes a collection of edges across a swatch of trees in the tree sequence,
-we call these \emph{edgesets}
-and store them in the columns of an \textbf{Edgeset Table:}
-\begin{center}
-\begin{tabular}{cccc}
-    left & right & parent & children \\
-    0.2  &  0.8  &  3     & 0,2 \\
-    0.0  &  0.2  &  4     & 1,2 \\
-    0.2  &  0.8  &  4     & 1,3 \\
-    0.8  &  1.0  &  4     & 1,2 \\
-    0.8  &  1.0  &  5     & 0,4 \\
-    0.0  &  0.2  &  6     & 0,4  \\
-\end{tabular}
-\end{center}
-
-To record information about genetic variants we need to also record each mutation
-and which nodes have inherited that mutation.
-The tree structure takes care of inheritance -- all we need to do
-is to record the highest node in the tree at the mutated site
-that inherited that mutation.
-As more than one mtuation may occur at a given site,
-we separate this information into two tables,
-first, the \textbf{Site Table} records for each variant site
-\begin{center}
-\begin{tabular}{ccc}
-    id  &   position  & ancestral\_state \\
-    0   &  0.1        & 0 \\
-    1   &  0.5        & 0 \\
-\end{tabular}
-\end{center}
-Here ``position'' is a (floating point) position along the chromosome,
-and ``ancestral state'' is the genotype of the root of the tree at that site.
-As for nodes, \textbf{site IDs} are given implicitly
-by the (zero-based) index of the rows.
-Then, we record in a \textbf{Mutation Table}
-\begin{center}
-\begin{tabular}{ccc}
-    site  & node  & derived\_state \\
-    0	  & 4	  & 1 \\
-    1	  & 3	  & 1 \\
-    1	  & 2	  & 0 \\
-\end{tabular}
-\end{center}
-in which ``site'' is the ID of the site at wihch this mutation occurred,
-``node'' is the ID of the highest node that has inherited this mutation,
-and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
-unless another mutation occurs.
-% \plr{Omit migrations?}
-
-\paragraph{Definition of valid tables}
-Here are the formal requirements for a set of nodes and edgesets to make sense,
-and to allow ``msprime``'s algorithms to work properly.
-
-To disallow time travel and multiple inheritance:
-\begin{enumerate}
-    \item Offspring must be born after their parents (and hence, no loops).
-    \item The set of intervals on which each individual is a child must be disjoint.
-\end{enumerate}
-For algorithmic reasons, we also require:
-\begin{enumerate} \setcounter{enumi}{2}
-    \item The leftmost endpoint of each chromosome is 0.0.
-    \item Node times must be strictly greater than zero.
-    \item The list of offspring in an edgeset must be sorted.
-    \item Edgesets must be sorted in nondecreasing time order.
-    \item The set of intervals on which each individual is a parent must be disjoint.
-    \item Each edgeset must contain at least two children.
-\end{enumerate}
-Note that since each node time is equal to the amount of time since the \emph{birth} of the
-corresponding parent, time is measured in clock time, not in meioses.
-
-A forwards-time simulation does \textbf{not} naturally emit genealogical information
-satisfying requirements 5--8.
-However, \msprime{} implements two algorithms that will take a set of tables
-satisfying only 1--4 and produce tables satisfying all requirements.
-Trivially, \texttt{sort\_tables} enforces requirements 5 and 6 and does not renumber nodes;
-then, \texttt{simplify} enforces requirements 7 and 8 (and does much more; see below).
-
 \section{More general method for recording the pedigree}

### 2017-10-25 forwards_paper.tex
Added benchmarks and discussion.

--- 2017-10-23 forwards_paper.tex
@@
 mutation resulting in node $2$ inheriting the state 'T'. The second side (in the
-right hand tree) has two mutations: one occuring over node $3$ changing the state to
+right hand tree) has two mutations: one occurring over node $3$ changing the state to
 `C' and a back mutation over node $1$ changing the state to `G'.
@@
 
-\jk{Other stuff we can review here:}
-\begin{enumerate}
-\item Encoding a tree sequence topology like this is very efficient, much, much more compact than newick.
-Can refer to example in msprime paper, or could do a large running example here. Say 500K samples of
-human CHR1, something like that. Newick will be hundreds of TB.
+This encoding of a sequence of trees and accompanying mutational information is
+very concise. To illustrate this, we ran a simulation of $500,000$ samples of a
+$200$ megabase human-like chromosome ($N_e=10^4$ and per-base mutation and
+recombination rates of $10^{-8}$ per generation) using \msprime. This resulted
+in about 1 million distinct marginal trees and $1.1$ million infinite-sites
+mutations. The HDF5 file encoding the nodes, edges, sites and mutations (as
+described above) for this simulation consumed 157MiB of storage space. Using
+the \msprime\ Python API, the time required to load this file into memory was
+around 1.5 seconds, and the time required to iterate over all 1 million tree
+was 2.7 seconds. In contrast, recording the topological information in Newick
+format would require around 20 TiB and storing the genotype information
+in VCF would require about 1 TiB. Working with either the Newick or VCF encoding
+of this dataset would be exceedingly cumbersome, likely requiring several
+days of CPU time simply to read the information info memory. \jk{Could
+add concrete timing estimates based on subsets here, but maybe overdoing it?}
 
-\item Can also make the point that it's much more efficient than, say, VCF. We could drop some
-mutations on the example above, and then report how big it would be if encoded with VCF and BCF.
-
-\item The \msprime\ API is very efficient for working with these large datasets. Examples on
-how long it takes to iterate over all the trees, compute pi, etc.
-
-\item The point of all this is to emphasise that not only is it efficient to generate the
-data as a tree sequence, it's also very efficient to store and process it this way too. By the
-time we get to the end of this section we should have convinced the reader of this.
-
-\end{enumerate}
+The tree sequence encoding is very compact but it is also very efficient to process.
+Many algorithms to compute statistics of interest for population genetics
+are naturally expressed in terms of tree topologies. For example, the
+pairwise nucleotide diversity $\pi$, is defined as the average number of
+differences between sequences in the sample. Computing $\pi$ directly
+from observed sequence data requires $O(n^2 m)$ time, since we must
+compare all pairs of samples at every site. However, given the topology of
+the trees at every site and the locations of mutations much better algorithms
+are possible. Using the fact that we can count the number of samples below
+a given node efficiently using previously described tree sequence
+algorithms~\citep{kelleher2016efficient}, the time required to compute $\pi$
+becomes roughly $O(n \log n + m)$.
+The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
+samples in a tree sequence, and calculating $\pi$ over all 500K samples
+in the example above required about 1.2 seconds.
+\jk{I haven't checked the analysis of computing $\pi$ here and not thought about
+it too deeply. I guess we should also compare this with something else in order to
+show that this time is pretty good. We could get a subset of the sites and push
+the data into pylibseq maybe, or scikit-allel??}
 

### 2017-10-25 forwards_paper.tex

@@
 
-To record the genealogical history of a simulation
+To record the genealogical history of a forwards in time simulation
 we need to record two things for each new chromosome:
-the birth time,
-and the endpoints and parental IDs of each distinctly inherited segment.
-This is recorded easily, without further processing,
-in the tables described above.
+the birth time, and the endpoints and parental IDs of each distinctly inherited segment.
+For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
+with overlapping generations that records genealogical history in this way.
+The simulation will run for $T$ generations,
+and has $N$ haploid individuals, each carrying a single chromosome of length L.
+For simplicity we assume there is exactly one crossover per generation.
+The probability of death per individual each generation is $\delta$.
 
-For concreteness, here we write out in pseudocode
-how to run a neutral Wright--Fisher simulation
-with overlapping generations
-that records genealogical history in this way.
-The simulation will run for $T$ generations,
-and has $N$ haploid individuals, each carrying a single chromosome of length L,
-on which for simplicity we assume there is exactly one crossover per generation.
-The probability of death per individual each generation is \texttt{death\_prob},
-and the whole-chromosome mutation rate per individual is \texttt{mut\_rate}.
+Let $\randomuniform(A)$ to be an element of the set $A$ chosen
+uniformly at random (note that each instance of $\randomuniform(A)$ within
+an algorithm listing represents an independent draw from the set $A$).
 
-\paragraph{Initialize:}
-We will build the tables \nodes, \edgesets, \sites, and \mutations,
-and keep track of the IDs of the current generation in the list \texttt{pop}.
-\begin{lstlisting}[language=Python]
-for i in 0:N-1:
-    nodes.add_row(time=T)
-    pop[i] = i
-\end{lstlisting}
-(Alternatively, the tables could be initialized by the results of a coalescent simulation.)
+\begin{taocpalg}{W}{Forwards-time tree sequence}
+{Runs a forwards time Wright--Fisher simulation of a populations with $N$
+individuals with a probability of death per generation $\delta$ and a chromosome of length
+$L$ for $T$ generations. On termination, there will be $n$ nodes with times recorded
+in the vector $\tau$, and $E$ will contain a set of $(\ell, r, p, c)$ tuples
+describing the recorded edges.
+}
 
-\paragraph{Iterate:}
-We then step through the generations,
-using the functions \texttt{random\_mutation} and \texttt{random\_allele}
-to choose positions and alleles for mutations, respectively:
-\begin{lstlisting}[language=Python]
-for t in 1:T:
-    for i in 0:N-1:
-        if random.uniform() > death_prob:
-            new_pop[i] = pop[i]
-        else:
-            u = nodes.num_rows  # the ID of the new individual
-            new_pop[i] = u
-            nodes.add_row(time = T-t)
-            a = random.sample(pop)
-            b = random.sample(pop)
-            bp = random.sample(0:L)
-            if bp > 0:
-                edgesets.add_row(left=0, right=bp,
-                                  parent=a, children=(u,))
-            if bp < L:
-                edgesets.add_row(left=bp, right=L,
-                                  parent=b, children=(u,))
-            num_muts = random.poisson(mut_rate)
-            for j in 1:num_muts:
-                pos = random_mutation()
-                if pos not in sites.position:
-                    s = sites.num_rows
-                    sites.add_row(position=pos,
-                           ancestral_state=random_allele())
-                else:
-                    s = sites.position.index(pos)
-                mutations.add_row(site=s, node=u,
-                           derived_state=random_allele())
-    pop = new_pop
-\end{lstlisting}
+\algstep{W1.}{Initialisation.}{
+ For $0 \leq j < N$, set $P_j \leftarrow j$ and $\tau_j \leftarrow T$.
+Set $E\leftarrow \emptyset$ and $n \leftarrow N$.
+}
 
-\paragraph{Finalize:}
-to obtain a tree sequence, we need only transform the tables to the required format:
-\begin{lstlisting}
-sort_tables(nodes, edgesets, sites, mutations)
-simplify_tables(nodes, edgesets, sites, mutations)
-\end{lstlisting}
-which can then be loaded into a tree sequence with \msprime.
-(These two functions operate on the tables in place.)
+\algstep{W2.}{Generation loop.}{ If $t = 0$ terminate.
+Set $t \leftarrow t - 1$ and $j \leftarrow 0$.
+}
+
+\algstep{W3.}{Individual loop.}{ If $j = N$ set $P \leftarrow P'$ and go to \algref{W2}.
+}
+
+\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) > \delta$, set $P'_j \leftarrow P_j$
+    and go to \algref{W7}.
+}
+
+\algstep{W5.}{New node.}{Set $P'_j \leftarrow n$ and $\tau_n \leftarrow t$.
+}
+
+\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
+    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
+}
+
+\algstep{W7.}{Record edges}{Set $E \leftarrow E \cup \left\{ (0, x, P_a, n), (x, L, P_b, n) \right\}$
+    and $n \leftarrow n + 1$.
+}
+
+\algstep{W8.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W4}.
+}
+
+\jk{I haven't checked this carefully. If we used 1-based indexes it might make things a bit
+simpler}.
+\end{taocpalg}
+
+Algorithm~\algref{W} is a very simple implementation of a Wright--Fisher model
+intended to illustrate how we can record the nodes and edges of a tree sequence
+forwards in time. We begin in~\algref{W1} by allocating our initial population $P$
+and creating $N$ nodes with birth time $T$ generations ago, recorded in the
+vector $\tau$. The set $E$ is used to store the edges that we output during the
+simulation, and $n$ is the number of nodes created so far (and so, the ID of
+the next node we create). Steps~\algref{W2} and~\algref{W3} simply loop over
+the generation clock $t$ and individual index $j$. In~\algref{W4} we check if
+an individual $P_j$ has died in this generation. If it has, we replace it in
+steps \algref{W5}--\algref{W7}; if not, we proceed immediately to the next
+individual. When an individual in the population dies, we first allocate
+a new node with ID $n$ in~\algref{W5} and record its birth time. Then,
+in step~\algref{W6} we choose two indexes $a$ and $b$ uniformly (giving us
+parents $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
+event by storing two new edges: one recording that the parent of node $n$
+from $0$ to $x$ is $P_a$, and another recording that the parent of $n$
+from $x$ to $L$ is $P_b$. We then complete the replacement event by incrementing
+$n$, ready to represent the next new node.
+
+\begin{enumerate}
+\item Explain how to add mutational information. We can also do this afterwards on the
+output trees if they are nuetral.
+\item Explain that we can input these trees into an mpsrime tree sequence and they will
+work fine. Perhaps a diagram to show the kind of trees we get? (Multiroots, lots of unary nodes.
+Can use the msprime SVG API to draw these?) We could refer ahead to figure 2
+instead though.
+\item However, there is huge amount of redundancy in these. The amount of information stored
+grows linearly with time.
+
+\end{enumerate}
 

### 2017-11-03 forwards_paper.tex
Added some text on mutations.

@@
         by \msprime.
-        \jk{The figure needs to be rescaled a bit to give more space to the trees. We could also use a
-        time scale next to the trees.}
+        \jk{Some minor alignment issues present here.}
         \label{fig:example_tree_sequence}
@@
 `C' and a back mutation over node $1$ changing the state to `G'.
-\jk{Not sure if I'm labouring the point here: this should be reasonably obvious.
-Perhaps if we added the state changes to the trees, and the states of the leaves
-below the samples, we could get rid of all the example discussion here?}
 
@@
 of this dataset would be exceedingly cumbersome, likely requiring several
-days of CPU time simply to read the information info memory. \jk{Could
-add concrete timing estimates based on subsets here, but maybe overdoing it?}
+days of CPU time simply to read the information info memory.
 
-The tree sequence encoding is very compact but it is also very efficient to process.
+The tree sequence encoding is compact but it is also efficient to process.
 Many algorithms to compute statistics of interest for population genetics
@@
 
+This algorithm records only the topological information resulting from the
+forwards in time Wright-Fisher process, but it is straightforward to add
+mutational information. This can be done in two different ways.
+We can record mutations that occur during the simulation quite simply.
+For example, in Algorithm~\algref{W} we would generate mutations after
+we have recorded the edges joining the new node $n$ to its parents
+$P_a$ and $P_b$ in step \algref{W7}.
+For example, if we assume that infinite sites mutations
+occur at rate $\mu$ per generation then there are $\mbox{Poisson}\left((r - \ell)(\tau_p
+- \tau_c)\right)$ mutations on each edge $(\ell, r, p, c)$. Each of these mutations
+will occur on a discinct site $x$ drawn uniformly from $(\ell, r]$ and be over node $c$.
+It is straightforward to record this information during the simulation, but it
+is significantly simpler and more efficient to generate these mutations
+after the simulation has completed. In the case of infinite sites mutations
+there is no difference between generating these mutations during the forwards
+in time simulation and after it has completed. We are using precisely the same
+information.
+
+It is significantly more efficient to generate mutations as a separate process
+after the topological simulation has completed because we will generate many
+mutations that will be lost in the population.
+
 \begin{enumerate}
-\item Explain how to add mutational information. We can also do this afterwards on the
-output trees if they are nuetral.
 \item Explain that we can input these trees into an mpsrime tree sequence and they will

### 2017-11-03 forwards_paper.tex
Initial draft of WF trees + updated algorithm.

@@
  For $0 \leq j < N$, set $P_j \leftarrow j$ and $\tau_j \leftarrow T$.
-Set $E\leftarrow \emptyset$ and $n \leftarrow N$.
+Set $E\leftarrow \emptyset$, $t \leftarrow T$ and $n \leftarrow N$.
 }
@@
 \algstep{W2.}{Generation loop.}{ If $t = 0$ terminate.
-Set $t \leftarrow t - 1$ and $j \leftarrow 0$.
+Set $P'_j \leftarrow P_j$ for $0\leq j < N$ and then set $t \leftarrow t - 1$ and $j \leftarrow 0$.
 }
@@
 
-\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) > \delta$, set $P'_j \leftarrow P_j$
-    and go to \algref{W7}.
+\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) \geq \delta$ go to \algref{W8}.
 }
@@
 
-\algstep{W8.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W4}.
+\algstep{W8.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W3}.
 }
@@
 information.
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{wf-before}
+        \includegraphics{wf-after}
+    \end{center}
+    \caption{\label{fig:wf-trees}
+    }
+\end{figure}
 

### 2017-11-03 sims/wf-figures.py

@@ -1,0 +1,54 @@
+"""
+Simulation of WF model and output resulting tree sequence.
+"""
+import random
+import msprime
+
+def wright_fisher(N, delta, L, T):
+    edges = msprime.EdgeTable()
+    tau = []
+    P = [j for j in range(N)]
+    for j in range(N):
+        tau.append(T)
+    t = T
+    n = N
+    while t > 0:
+        t -= 1
+        j = 0
+        Pp = [P[j] for j in range(N)]
+        while j < N:
+            if random.random() < delta:
+                Pp[j] = n
+                tau.append(t)
+                a = random.randint(0, N - 1)
+                b = random.randint(0, N - 1)
+                x = random.uniform(0, L)
+                edges.add_row(0, x, P[a], n)
+                edges.add_row(x, L, P[b], n)
+                n += 1
+            j += 1
+        P = Pp
+    nodes = msprime.NodeTable()
+    P = set(P)
+    for j in range(n):
+        nodes.add_row(time=tau[j], flags=int(j in P))
+    msprime.sort_tables(nodes=nodes, edges=edges)
+    return msprime.load_tables(nodes=nodes, edges=edges)
+
+
+
+if __name__ == "__main__":
+
+    random.seed(1)
+    ts = wright_fisher(5, 0.95, 1, 10)
+    t = next(ts.trees())
+    print(t.draw(format="unicode"))
+    t.draw(path="wf-before.svg")
+    tss, node_map = ts.simplify(map_nodes=True)
+    node_labels = {
+        node_map[j]: str(j) for j in range(node_map.shape[0])}
+    t = next(tss.trees())
+    print(t.draw(format="unicode", node_label_text=node_labels))
+    t.draw(path="wf-after.svg", node_label_text=node_labels)
+
+

### 2017-11-06 sims/run_check.sh
Formatting issue in run script

@@ -1,9 +1,6 @@
 echo "Must have fwdpy11_arg_example installed!!"
 wget https://github.com/ashander/fwdpy11_arg_example/blob/check-trees/benchmarking.py
-python ./benchmark-simuPOP.py --theta 10 --rho 10 --nsam 10 --pdel 0 --gc 50  \
-	--outfile1 out.log.gz --seed 42 --popsize 100 --treefile
-simupoptree.hdf5
-python ./benchmarking.py --theta 10 --rho 10 --nsam 10 --pdel 0 --gc 50 \
-	--outfile1 out.log.gz --seed 42 --popsize 100 --treefile fwdpptree.hdf5
+python ./benchmark-simuPOP.py --theta 10 --rho 10 --nsam 10 --pdel 0 --gc 50  --outfile1 out.log.gz --seed 42 --popsize 100 --treefile simupoptree.hdf5
+python ./benchmarking.py --theta 10 --rho 10 --nsam 10 --pdel 0 --gc 50 --outfile1 out.log.gz --seed 42 --popsize 100 --treefile fwdpptree.hdf5
 python ./check.py
 

### 2017-11-06 sims/benchmark-simuPOP.py
oops we weren't 'optimized'

@@ -135,6 +135,8 @@ if args.generations is None:
 
 from simuOpt import setOptions
 setOptions(alleleType = 'binary')
+setOptions(optimized=True)
+
 
 import simuPOP as sim
 

### 2017-11-06 sims/benchmark-simuPOP.py
The 'mutant' allele state = big speedup

@@ -134,7 +134,7 @@ if args.generations is None:
     args.generations = args.popsize * args.multiplier
 
 from simuOpt import setOptions
-setOptions(alleleType = 'binary')
+setOptions(alleleType = 'mutant')
 setOptions(optimized=True)
 
 

### 2017-11-06 forwards_paper.tex

@@
 simulated 600,000 samples of human chromosome 20 to assess the impact of European biased
-reference panels in genome wide association studies. While many results of the coalescent are surpisingly
+reference panels in genome wide association studies. While many results of the
+coalescent are surprisingly
 robust to violations of the assumption that $n \ll N$, there are instances where genealogical properties
@@
 is produced by recombination. The encoding is efficient because branches that are
-shared by adjacent trees are stored once, rather than repeatedely for each tree.
+shared by adjacent trees are stored once, rather than repeatedly for each tree.
 The topology of a tree sequence is defined via a set of \emph{nodes} and
@@
 - \tau_c)\right)$ mutations on each edge $(\ell, r, p, c)$. Each of these mutations
-will occur on a discinct site $x$ drawn uniformly from $(\ell, r]$ and be over node $c$.
+will occur on a distinct site $x$ drawn uniformly from $(\ell, r]$ and be over node $c$.
 It is straightforward to record this information during the simulation, but it
@@
     \end{center}
-    \caption{\label{fig:wf-trees}
+    \caption{An example of a marginal genealogy from a Wright-Fisher simulation
+    with $N=5$. On the left we have the original tree including all
+    intermediate nodes and dead-ends. On the right we have the minimal tree
+    relating all of the currently-alive individuals (27--31).
+    \jk{This figure needs a little tidying up. We should combine the two SVGs
+        into a single image with some (a) (b) annotations. Also the node labels
+        are poorly aligned. The SVGs are generated by the sims/wf-figures.py.}
+    \label{fig:wf-trees}
     }
@@
 after the topological simulation has completed because we will generate many
-mutations that will be lost in the population.
+mutations that will be lost in the population. Figure~\ref{fig:wf-trees} shows
+an example of a marginal genealogy produced by a forwards-time Wright-Fisher
+process like Algorithm~\algref{W}. There are two different trees in this
+figure: one shows all the edges output by the simulation and the other
+is the minimal tree representing the ancestry of the currently alive
+population. Clearly there is a great deal of redundancy in the topological
+information output by the simulation.
 
+There are two sources of redundancy here. The first type of redundancy arises
+from nodes in tree that have only one child. In Algorithm~\algref{W} we do
+not attempt to track coalescence events but simply record all parent-child
+relationships in the history of the population. As such, my of these edges
+will record the simple passing of genealogical information from parent to child
+and only some small subset will correspond to coalesences within the marginal
+trees. The second source of redundancy in the output of Algorithm~\algref{W}
+is due to the fact that lineages die out: there will be a large number of
+individuals in the simulation that leave no ancestors in the present day
+population. Node 26 in Figure~\ref{fig:wf-trees}a, for example, leaves no
+ancestors in the current population and so the entire path tracing back to
+the root is redundant.
+
+One of the reasons that it is more efficient to generate mutations on the
+genealogies after we have completed the forwards in time simulation is
+because we avoid the cost of generating mutations on these dead-end branches.
+It is only with the benefit of hindsight that we can know which lineages end
+up being ancestral to the sample, and we can avoid a substantial cost of
+generating and maintaining mutational information on these evolutionary
+dead-ends. Assuming mutations are rare, another  reason that it is more efficient
+to place mutations on the trees afterwards is that we have far fewer edges
+in the trees: because we have removed all the intermediate `unary' nodes, we
+have fewer edges to generate mutations on.
+
+There are many advantages to having minimal genealogies such as show in
+Figure~\ref{fig:wf-trees}. Computing this minimal representation of the
+edges output by a forward-time simulation is an instance of a more
+general problem that we refer to as `tree sequence simplification'. We discuss
+this problem and an efficient solution in the next subsection.
+
+%%%%%%
+\subsection*{Tree sequence simplification}
+
+Suppose we have a tree sequence consisting of a set of nodes and edges as
+described earlier, and we have some subset of the nodes in this tree sequence
+that we are interested in (our `samples').  We wish to reduce the input tree sequence
+into the minimum representation of the topologies that include the specified
+samples. The output tree sequence must have the following properties:
 \begin{enumerate}
-\item Explain that we can input these trees into an mpsrime tree sequence and they will
-work fine. Perhaps a diagram to show the kind of trees we get? (Multiroots, lots of unary nodes.
-Can use the msprime SVG API to draw these?) We could refer ahead to figure 2
-instead though.
-\item However, there is huge amount of redundancy in these. The amount of information stored
-grows linearly with time.
-
+\item we must have the same marginal trees with respect to the samples as
+the input;
+\item within the marginal trees, all non-sample vertices must has at least
+two children (i.e., unary tree vertices are removed);
+\item any nodes and edges not reachable from the sampled nodes are removed;
+\item there are no adjacent redundant edges, i.e, pairs of edges $(\ell, x, p,
+c)$ and $(x, r, p, c)$ which can be represented with a single edge
+$(\ell, r, p, c)$.
 \end{enumerate}
@@
         in 10 node records (one for each individual)
-        and 14 edgeset records (one for each distinctly inherited segment).
+        and 14 edge records (one for each distinctly inherited segment).
         Blue numbers denote crossing over locations in each meiosis.
-        The individuals $B$, $C$, and $D$ inherit clonally from $A$
-        to ensure rootnedness of resulting tree.
+        The individuals $B$, $C$, and $D$ inherit clonally from $A$.
         \textbf{Center:} the five distinct trees relating all individuals to each other
@@
 
-%%%%%%
-\subsection*{Tree sequence simplification}
+In the current context of running forwards in time simulations, the tables of
+nodes and edges that we record all of history for everyone alive at any time through the simulation.
+This is much more than we need to reconstruct the genealogies and sequences
+of the current day population. We can use the process of simplification here to
+reduce this down to the minimal set of nodes and edges required to represent
+the information that we are interested in. We can also use simplification if  we have some very
+large tree sequence representing a large dataset and we wish to extract the
+information relevant to a subset of the samples.
 
-The resulting tables encode everything we need to know --
-in fact, they record all of history for everyone alive at any time through the simulation.
-This is much more than we need to reconstruct the genealogies and sequences
-of a smaller sample of indidivuals.
-Reducing this larger tree sequence to a smaller one relevant to a given set of ``sample'' nodes
-we call \emph{simplification}.
-Roughly, this works by tracing ancestry from the samples backwards through the recorded history,
-adding node and edgeset records to the output only when coalescent events are reached.
-This works exactly as in \msprime, allowing substantial re-use of algorithms;
-the main difference being that parental choice, mutations, and recombination locations
-are determined by the input tree sequence
-rather than randomly generated.
-An example is shown in Figure \ref{fig:method_diagram}.
+Roughly, simplification works by tracing ancestry from the samples backwards
+through the recorded history, adding node and edge records to the output
+only when coalescent events are reached. This works exactly as in \msprime,
+allowing substantial re-use of algorithms; the main difference being that
+parental choice, mutations, and recombination locations are determined by the
+input tree sequence rather than randomly generated. An example is shown in
+Figure \ref{fig:method_diagram}.
 
-In this scheme,
-at any point in the simulation genealogical history is recorded in a tree sequence.
-This has two additional advantages.
-First, simplification can be run periodically through the simulation,
-taking the set of samples to be the entire currently alive population.
-This is important as it keeps memory usage from growing linearly (and quickly) with time.
-Second, the simulation can be \emph{begun} with a tree sequence produced by some other method --
-for instance, by a coalescent simulation with \msprime.
-This allows for incorporation of deep-time history beyond the reach of individual-based simulations.
-Since geographic structure from times longer ago than the mixing time of migration across the range
-has limited effect on modern genealogies \citep{wilkins_separation}
-(other than possibly changing effective population size \citet{durretspatial}),
-this may not negatively affect realism.
+The approach that we take is based on Hudson's algorithm for simulating
+the coalescent with
+recombination~\citep{hudson1983properties,kelleher2016efficient}. The main
+state of the algorithm is a set of ancestral segments; each segment $(\ell, r,
+u)$ represents a tract of ancestral material in which the node $u$ is
+associated with the genomic interval $[\ell, r)$. In the same manner
+as Hudson's algorithm, we represent the set of extant lineages as a
+set of individuals, where each ancestral individual is composed of a
+linked list of these segments. The node $u$ in each ancestral sequence
+is the node ID associated with this segment in the output tree sequence.
+We maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
+the input tree sequence.
 
-\plr{Something about putting mutations down on the tree sequence.}
+\jk{I'm not sure this is the right level of detail to tackle this description
+at. What's the goal of this algorithm description? Are we trying to
+establish some properties of the algorithm or just give the reader a high-level
+flavour of how it works??}
+
+In this scheme, at any point in the simulation genealogical history is recorded
+in a tree sequence. This has two additional advantages. First, simplification
+can be run periodically through the simulation, taking the set of samples to be
+the entire currently alive population. This is important as it keeps memory
+usage from growing linearly (and quickly) with time. Second, the simulation can
+be \emph{begun} with a tree sequence produced by some other method -- for
+instance, by a coalescent simulation with \msprime. This allows for
+incorporation of deep-time history beyond the reach of individual-based
+simulations. Since geographic structure from times longer ago than the mixing
+time of migration across the range has limited effect on modern genealogies
+\citep{wilkins_separation} (other than possibly changing effective population
+size \citet{durretspatial}), this may not negatively affect realism.
 
@@
 
-\plr{Quick overview of how to efficiently hook this up with other code.}
+The \msprime\ Python API provides a powerful platform for working with
+tree topology and mutation data. We refer to the part of \msprime\
+that is dedicated to tree sequence input and output as the `Tables API',
+as the API is organised around simple tables of data. There are four key
+tables: nodes, edges, sites and mutations. Briefly, nodes and edges define
+the topology of a tree sequence (as defined above) and the sites and mutations
+define mutational processes on this topology.
+Figure~\ref{fig:example_tree_sequence} gives an informal depiction of this
+encoding in terms of these tables.
+
+The tables API is primarily designed to facilitate efficient interchange of
+data between programs or between different modules of the same program.
+Following the current best-practises [citations: apache arrow, etc] data is stored
+in a columnar format. There are many advantages to this approach, but the
+principle advantage for our purposes is that it allows for very efficient
+interchange of large amounts of numerical data. In principle, this enables
+zero-copy semantics, where a data consumer can read the information directly
+from the memory of a producer without incurring the overhead of a copy
+[citation?] Our implementation uses the numpy C API [citation] to efficiently copy
+data from Python into the low-level C library used to manipulate
+tree sequences.
+
+The tables API provides basic input and output operations via the numpy
+array interface, which provides a great deal of flexibility as well
+as efficiently. This makes transferring data from various sources
+such as HDF5 [citation], Dask, Zarr etc, straightforward. (For small
+scale data and debugging purposes, a simple text format is also supported.)
+Along with
+these operations we also provide a function to sort a set of tables
+to ensure that the records are in the form required to input
+into an \msprime\ tree sequence object. The \texttt{simplify\_tables}
+function implements tree sequence simplification, as described in the
+previous subsection.
+
+This interchange API is very efficient. [Describe a quick example where we generate
+a many-gigabyte tree sequence using fwdpp, and the time required
+to copy the node and edge data into the tables]. By using a simple numerical
+encoding of tree topologies and contiguous arrays of data to store this
+encoding, we can achieve data transfer rates that would be impossible under
+any text-based approach while retaining excellent portability.
 
@@
 Suppose that we wish to run a forwards-time simulation of $N$ individuals for $T$ generations,
-in which there are $S$ selected loci and $L$ neutral loci.
+in which there are $S$ selected loci and $L$ neutral loci using a similar
+method to Algorithm~\algref{W}.
 We will estimate run-time complexity and memory usage for both a ``naive'' strategy that carries along neutral loci
@@
 (We assume here that selected genotypes are stored in the same way for both.)
-Passing on neutral genotypes naively records $L$ items per individual each generation, discarding the previous generation.
-\plr{do this better with numbers from nodes, edgesets}
-Recording a tree sequence stores 2 parents and $\rho$ breakpoints on average each generation;
-after $T$ generations this grows to $N \times T \times (2 + \rho)$ stored items.
-However, after simplification a tree sequence for all $N$ individuals
-only takes of order $N + \rho \log N$ records.
-Simplification requires processing each of the initial records -- so, order of $N \times T \times \rho$ operations;
+Passing on neutral genotypes naively records $L$ items per individual each
+generation, discarding the previous generation, giving an overall complexity of
+$O((L + S) NT)$. Tracking the $S$ selected loci and recording a tree sequence
+using Algorithm~\algref{W} requires $O(SNT)$ time and space, since we
+record at most two edges for each individual in each generation.
+However, after simplification the tree sequence contains
+$O(??)$ edges since []. The cost of generating $L$ neutral mutations
+is then $O(L + ??)$. Assuming that $N$, $T$ and $L$ are large, this is
+very much smaller than the straightforward $O((L + S) NT)$ time required
+to simulate all mutations directly.
+
+\jk{Not sure where $\rho$ comes in here now. Thinking about this, shouldn't we
+be able to come up with some expression for the number of edges? Assuming
+the very simple single breakpoint per generation of Alg W, what is the
+expected number of breakpoints? This is a function of $N$ and $T$, right?
+It's not clear to me how many breakpoints this ends up giving you in
+the output tree sequence and how many edges this corresponds to.
+It's probably worthwhile thinking about this a bit.}
+
+% However, after simplification a tree sequence for all $N$ individuals
+% only takes of order $N + \rho \log N$ records.
+% Simplification requires processing each of the initial records -- so, order of $N \times T \times \rho$ operations;
 

### 2017-11-06 sims/benchmark-simuPOP.py
Make initial ts via same parameters

@@ -217,6 +217,8 @@ id_tagger.apply(pop)
 # NB: we have to simulate an initial tree sequence
 first_gen = pop.indInfo("ind_id")
 init_ts = msprime.simulate(2*len(first_gen),
+                           Ne=args.popsize,
+                           recombination_rate=args.recomb_rate,
                            length=max(locus_position))
 haploid_labels = [(k,p) for k in first_gen
                         for p in (0,1)]

### 2017-11-08 sims/check.py
Correct scaling, run checks

@@ -1,54 +1,60 @@
 # coding: utf-8
 import msprime
 import random
+import argparse
 from matplotlib import pyplot as plt
 
 
-random.seed(111)
+parser = argparse.ArgumentParser(description='small test script')
+parser.add_argument("--popsize","-N", type=int, dest="popsize",
+        help="size of each subpopulation",default=100)
+parser.add_argument("--theta","-ϴ", type=int, dest="theta",
+                    help="total mutation rate for whole chromosome: theta = 4 N u/site (sites)", default=100)
+parser.add_argument("--rho","-ρ", type=int, dest="rho",
+                    help="total recombination rate for whole chromosome: rho = 4 N r/site (sites)", default=100)
+parser.add_argument("--nsam","-k", type=int, dest="nsam",
+                    help="total number of samples with msprime", default=10)
+parser.add_argument("--sims","-n", type=int, dest="NUM",
+                    help="total number of sims with msprime", default=100)
+parser.add_argument("--seed","-s", type=int, dest="seed",
+                    help="random seed")
+parser.add_argument("--out-fig","-o", type=str, dest="figname",
+                    help="desination, e.g. check-results.png")
+args = parser.parse_args()
+
+random.seed(args.seed)
 ft_ts = msprime.load('simupoptree.hdf5')
 L = ft_ts.sequence_length
-fwd_ts = msprime.load('fwdpptree.hdf5')
+
 L = ft_ts.sequence_length
-#assert L == fwd_ts.sequence_length, "fwdpp {} and simupop {} don't match".format(fwd_ts.sequence_length, L)
-NUM = 100
-
-tss = {'simupop': ft_ts, 'fwdpp':  fwd_ts}
-
-L = fwd_ts.sequence_length
-theta = 10.
-rho = 10.
-mut_rate = theta/4./ L
-rec_rate = rho/4. / L
+theta = args.theta
+rho = args.rho
+Ne = args.popsize
+mut_rate = theta/4./ L /Ne
+rec_rate = rho/4. / L /Ne
 S_msprime = []
 
-for _ in range(NUM):
-    g = msprime.simulate(10, Ne=100, length=L, mutation_rate=mut_rate,
+for _ in range(args.NUM):
+    g = msprime.simulate(args.nsam, Ne=Ne, length=L, mutation_rate=mut_rate,
[…]

### 2017-11-08 sims/run_check.sh

@@ -1,6 +1,8 @@
-echo "Must have fwdpy11_arg_example installed!!"
-wget https://github.com/ashander/fwdpy11_arg_example/blob/check-trees/benchmarking.py
-python ./benchmark-simuPOP.py --theta 10 --rho 10 --nsam 10 --pdel 0 --gc 50  --outfile1 out.log.gz --seed 42 --popsize 100 --treefile simupoptree.hdf5
-python ./benchmarking.py --theta 10 --rho 10 --nsam 10 --pdel 0 --gc 50 --outfile1 out.log.gz --seed 42 --popsize 100 --treefile fwdpptree.hdf5
-python ./check.py
+python ./benchmark-simuPOP.py --theta 10 --rho 10 --nsam 10 --pdel 0 --gc 50 --outfile1 /dev/null --seed 42 --popsize 100 --treefile simupoptree.hdf5
+python ./check.py --rho 10 --theta 10 --popsize 100 --seed 111 -o check-10-10-100.png
 
+python ./benchmark-simuPOP.py --theta 5 --rho 5 --nsam 10 --pdel 0 --gc 50 --outfile1 /dev/null --seed 42 --popsize 200 --treefile simupoptree.hdf5
+python ./check.py --rho 5 --theta 5 --popsize 200 --seed 111 -o check-5-5-200.png
+
+python ./benchmark-simuPOP.py --theta 20 --rho 20 --nsam 10 --pdel 0 --gc 50 --outfile1 /dev/null --seed 42 --popsize 100 --treefile simupoptree.hdf5
+python ./check.py --rho 20 --theta 20 --popsize 100 --seed 111 -o check-20-20-100.png -N 500

### 2017-11-08 sims/run_check.sh
Bump theta a bit

@@ -6,3 +6,6 @@ python ./check.py --rho 5 --theta 5 --popsize 200 --seed 111 -o check-5-5-200.pn
 
 python ./benchmark-simuPOP.py --theta 20 --rho 20 --nsam 10 --pdel 0 --gc 50 --outfile1 /dev/null --seed 42 --popsize 100 --treefile simupoptree.hdf5
 python ./check.py --rho 20 --theta 20 --popsize 100 --seed 111 -o check-20-20-100.png -N 500
+
+python ./benchmark-simuPOP.py --theta 30 --rho 30 --nsam 10 --pdel 0 --gc 50 --outfile1 /dev/null --seed 42 --popsize 200 --treefile simupoptree.hdf5
+python ./check.py --rho 30 --theta 30 --popsize 200 --seed 111 -o check-30-30-200.png -N 500

### 2017-11-08 sims/run_check.sh
We've a problem

@@ -9,3 +9,6 @@ python ./check.py --rho 20 --theta 20 --popsize 100 --seed 111 -o check-20-20-10
 
 python ./benchmark-simuPOP.py --theta 30 --rho 30 --nsam 10 --pdel 0 --gc 50 --outfile1 /dev/null --seed 42 --popsize 200 --treefile simupoptree.hdf5
 python ./check.py --rho 30 --theta 30 --popsize 200 --seed 111 -o check-30-30-200.png -N 500
+
+python ./benchmark-simuPOP.py --theta 40 --rho 40 --nsam 10 --pdel 0 --gc 50 --outfile1 /dev/null --seed 42 --popsize 200 --treefile simupoptree.hdf5
+python ./check.py --rho 40 --theta 40 --popsize 200 --seed 111 -o check-40-40-200.png -N 500

### 2017-11-08 sims/benchmark-simuPOP.py
Add var-printing options

@@ -125,6 +125,10 @@ parser.add_argument("--generations","-T", type=int, dest="generations",
 parser.add_argument("--simlen","-s", type=int, dest="multiplier",
         help="multiplier of popsize to run for (unless `generations` is specified)",
                     default=10)
+parser.add_argument("--verbose", "-v", dest="verbose", action='store_true')
+parser.add_argument("--num-vars", "-n", dest="num_vars", action='store_true',
+                    help="print the number of variants post mut")
+
 
 args = parser.parse_args()
 if args.record_neutral:
@@ -136,6 +140,8 @@ if args.generations is None:
 from simuOpt import setOptions
 setOptions(alleleType = 'mutant')
 setOptions(optimized=True)
+if not args.verbose:
+    setOptions(quiet=True)
 
 
 import simuPOP as sim
@@ -152,6 +158,9 @@ logfile.write("----------\n")
 logfile.flush()
 
 # define defaults for benchmark
+# scalings:
+# theta = 4 * Ne * mutation rate per bp * total number of bp
+# rho = 4 * Ne * mutation rate per bp * total number of bp
 npops = 1
 rloci = args.rho / (args.popsize * 4 * args.recomb_rate)
 uloci = args.theta / (args.popsize * 4 * args.mut_rate)
@@ -247,9 +256,6 @@ pop.evolve(
                              infoFields="ind_id"),
         ] ),
     postOps=[
-        sim.Stat(numOfSegSites=sim.ALL_AVAIL, step=REPORTING_STEP,
-                 vars=['numOfSegSites', 'numOfFixedSites']),
-        sim.PyEval(r"'Gen: %2d #seg/#fixed sites: %d / %d\n' % (gen, numOfSegSites, numOfFixedSites)", step=REPORTING_STEP),
         sim.PyOperator(lambda pop: rc.simplify(pop.indInfo("ind_id")) or True,
                        step=args.simplify_interval),
     ],
@@ -326,7 +332,8 @@ elif not args.record_neutral:
     mutgen.generate(nodes, edges, sites, mutations)
     mutated_ts = msprime.load_tables(
         nodes=nodes, edges=edges, sites=sites, mutations=mutations)
-
+if args.num_vars:
+    print(len([i for i in mutated_ts.variants()]))
 del ts
 
 logfile.write("Generated mutations!\n")

### 2017-11-08 forwards_paper.tex
added wf figs
various fig updates

@@
 \subsection*{Tree sequences}
+
 A tree sequence is an efficient encoding for a sequence of correlated trees such as
@@
 
+Recovering the sequence of trees from a collection of edges is straightforward:
+each point along the genome that the tree topology changes
+is accompanied by the end of some \emph{edges} and the beginning of others.
+Since each \emph{edge} records the genomic interval 
+over which a given node inherits from a particular ancestor,
+to construct the tree at a certain point in the genome
+we need only retrieve all edges overlapping that point
+and construct the corresponding edges.
+To construct the tree at a nearby location,
+we need only remove those edges whose intervals do not overlap that location,
+and add those new edges whose intervals do.
+Incidentally, this property that edges naturally encode \emph{differences}
+between nearby trees (e.g., as ``subtree prune and regraft'' moves)
+allows for efficient algorithms that take advantage
+of the highly correlated nature of nearby trees.
+
 Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
@@
 days of CPU time simply to read the information info memory.
-
-The tree sequence encoding is compact but it is also efficient to process.
-Many algorithms to compute statistics of interest for population genetics
-are naturally expressed in terms of tree topologies. For example, the
-pairwise nucleotide diversity $\pi$, is defined as the average number of
-differences between sequences in the sample. Computing $\pi$ directly
-from observed sequence data requires $O(n^2 m)$ time, since we must
-compare all pairs of samples at every site. However, given the topology of
-the trees at every site and the locations of mutations much better algorithms
-are possible. Using the fact that we can count the number of samples below
-a given node efficiently using previously described tree sequence
-algorithms~\citep{kelleher2016efficient}, the time required to compute $\pi$
-becomes roughly $O(n \log n + m)$.
-The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
-samples in a tree sequence, and calculating $\pi$ over all 500K samples
-in the example above required about 1.2 seconds.
-\jk{I haven't checked the analysis of computing $\pi$ here and not thought about
-it too deeply. I guess we should also compare this with something else in order to
-show that this time is pretty good. We could get a subset of the sites and push
-the data into pylibseq maybe, or scikit-allel??}
 
@@
     \begin{center}
-        \includegraphics{wf-before}
-        \includegraphics{wf-after}
+        \includegraphics{wf-before-after}
     \end{center}
     \caption{An example of a marginal genealogy from a Wright-Fisher simulation
-    with $N=5$. On the left we have the original tree including all
-    intermediate nodes and dead-ends. On the right we have the minimal tree
+    with $N=5$. \textbf{(A)} the original tree including all
+    intermediate nodes and dead-ends, and \textbf{(B)} the minimal tree
     relating all of the currently-alive individuals (27--31).
-    \jk{This figure needs a little tidying up. We should combine the two SVGs
-        into a single image with some (a) (b) annotations. Also the node labels
-        are poorly aligned. The SVGs are generated by the sims/wf-figures.py.}
     \label{fig:wf-trees}
@@
 
+This works great and very fast.
+
+Having results in a tree sequence is really good downstream too:
+The tree sequence encoding is compact but it is also efficient to process.
+Many algorithms to compute statistics of interest for population genetics
+are naturally expressed in terms of tree topologies. For example, the
+pairwise nucleotide diversity $\pi$, is defined as the average number of
+differences between sequences in the sample. Computing $\pi$ directly
+from observed sequence data requires $O(n^2 m)$ time, since we must
+compare all pairs of samples at every site. However, given the topology of
+the trees at every site and the locations of mutations much better algorithms
+are possible. Using the fact that we can count the number of samples below
+a given node efficiently using previously described tree sequence
+algorithms~\citep{kelleher2016efficient}, the time required to compute $\pi$
+becomes roughly $O(n \log n + m)$.
+The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
+samples in a tree sequence, and calculating $\pi$ over all 500K samples
+in the example above required about 1.2 seconds.
+\jk{I haven't checked the analysis of computing $\pi$ here and not thought about
+it too deeply. I guess we should also compare this with something else in order to
+show that this time is pretty good. We could get a subset of the sites and push
+the data into pylibseq maybe, or scikit-allel??}
+
 This is a general-purpose strategy that can be applied to other methods.

### 2017-11-08 sims/benchmark-simuPOP.py
Switching to unit length doesn't help

@@ -172,7 +172,7 @@ args.sel_mut_rate = args.mut_rate * args.pdel
 # >The default positions are 1, 2, 3, 4, ... on each
 # >chromosome.
 
-locus_position = list(range(0, args.nloci))
+locus_position = list(i / args.nloci for i in range(0, args.nloci))
 
 # logfile.write("Locus positions:\n")
 # logfile.write(str(locus_position)+"\n")

### 2017-11-08 sims/benchmark-simuPOP.py
Revert to long sequence

@@ -172,7 +172,7 @@ args.sel_mut_rate = args.mut_rate * args.pdel
 # >The default positions are 1, 2, 3, 4, ... on each
 # >chromosome.
 
-locus_position = list(i / args.nloci for i in range(0, args.nloci))
+locus_position = list(range(0, args.nloci))
 
 # logfile.write("Locus positions:\n")
 # logfile.write(str(locus_position)+"\n")

### 2017-11-12 forwards_paper.tex
another fig
editing...

@@
 
-\title{Efficient forwards-time simulation with ancestral recombination graphs}
+\title{Efficient forwards-time simulation with tree sequences}
 \author{Jerome Kelleher,
@@
 \begin{abstract}
-    To use genomic data for inference and prediction
-    it is often necessary to obtain whole-genome information
-    from individual-based simulations,
-    but the computational burden of tracking the genome of each simulated individual can be substantial.
+    The computational burden of tracking individual genomes 
+    in a population genetics simulation can be substantial.
     In this note we describe how to both (a) dramatically reduce this burden and
@@
     and recording the entire history of genetic inheritance in an efficient representation of the ancestral recombination graph,
-    on which neutral mutations can be quickly placed afterwards.
-    \plr{make more clear data structure was already developed? refer to 'tree sequence' by name?}
-    The algorithm is implemented in python,
-    and is designed to be easily used by any forwards-time simulation software.
-
-    Coalescent simulations are very helpful but require random mating and neutrality.
-    For continuous space, polygenic selection, or detailed dissection of life history,
-    we must use forwards-time, individual-based simulation.
-    These are much slower due in part to carrying around neutral genotypes irrelevant to the process.
-    Here we show how to efficiently produce and store the entire history of ancestry and recombination
-    from an individual-based simulation,
-    on which neutral mutations can be placed afterwards.
-    This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
+    on which neutral mutations can be quickly placed afterwards --
+    the \emph{tree sequence} introduced in the software package \msprime.
+    We implement the method in two popular fowards-time simulation frameworks,
+    and show it leads to order-of-magnitude speedups.
+    % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
 \end{abstract}
 
-\textbf{OUTLINE}
+% NOTE: I tend to use semantic linebreaks.
+% Please excuse the ragged right margins in the source.
+
+previous \textbf{OUTLINE}, possibly out-of-date but left for comparison.
 \begin{enumerate}
@@
 
-\plr{perhaps some of this veers into methods rather than introduction}
-
 Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
@@
 are only \emph{Markovian}, and thus feasible to work with,
-thanks to the important assumptions of (a) random mating, and (b) neutrality.
-\plr{Brief statement why this is.  Also include stationarity?}
-Both assumptions can be side-stepped to a limited extent, and so coalescent methods are now commonly used to
-simulate the results of population dynamics of a collection of randomly mating populations exchanging migrants,
-having a small number of loci under selection.
-Mapping results of such models onto real species can be challenging,
-as these are often distributed across geographical space and may have large numbers of loci under various sorts of selection.
+thanks to the important assumptions of 
+(a) random mating, 
+(b) neutrality,
+and (c) small sample size relative to the population size.
+The first two assumptions can be side-stepped to a limited extent, 
+but it remains a challenge to map results of coalescent models
+onto species that are distributed across continuous geographical space 
+and/or have large numbers of loci under various sorts of selection.
 Furthermore, the relationship between the life history of a species --
@@
 are all absorbed into a single compound parameter, the coalescence rate.
+The third assumption is no longer safe, either --
+for example, a recent study~\citep{martin2017human}
+simulated 600,000 samples of human chromosome 20
+% to assess the impact of European biased reference panels in genome wide association studies. 
+Several studies have now shown that in samples of size approaching that of the population,
+genealogical properties may be distorted relative to the coalescent expectation
+\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
 These considerations, and increasing computational power, have led to a resurgence of interest in forwards-time, individual-based simulations.
 
-Another critical assumption of the coalescent is that the sample size $n$ is much smaller
-than the effective population size $N_e$. With the very large sample sizes becoming common
-in human genetics, this is no longer a safe assumption. For example, a recent study~\citep{martin2017human}
-simulated 600,000 samples of human chromosome 20 to assess the impact of European biased
-reference panels in genome wide association studies. While many results of the
-coalescent are surprisingly
-robust to violations of the assumption that $n \ll N$, there are instances where genealogical properties
-are distorted and coalescent simulations may be
-misleading~\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
-A hybrid approach, in which the recent history of a large sample of
-individuals is simulated under a detailed forwards-in-time model and the deep history
-under the coalescent, is an attractive prospect.
-\jk{Not sure where this should go; perhaps in the discussion? We can make the point that
-by keeping track of the genealogies forwards in time we can easily combine these with the
-deep history simulated by msprime}
+%% Merged this above; moved the "hybrid approach" to the conclusion.
+% Another critical assumption of the coalescent is that the sample size $n$ is much smaller
+% than the effective population size $N_e$. 
+% With the very large sample sizes becoming common
+% in human genetics, this is no longer a safe assumption. 
+% For example, a recent study~\citep{martin2017human}
+% simulated 600,000 samples of human chromosome 20 to assess the impact of European biased
+% reference panels in genome wide association studies. While many results of the
+% coalescent are surprisingly
+% robust to violations of the assumption that $n \ll N$, there are instances where genealogical properties
+% are distorted and coalescent simulations may be
+% misleading~\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
+% A hybrid approach, in which the recent history of a large sample of
+% individuals is simulated under a detailed forwards-in-time model and the deep history
+% under the coalescent, is an attractive prospect.
+% \jk{Not sure where this should go; perhaps in the discussion? We can make the point that
+% by keeping track of the genealogies forwards in time we can easily combine these with the
+% deep history simulated by msprime}
 
-With modern computing power, pure demographic calculations are not a barrier,
-even though biological population sizes are often above $10^6$,
-and coalescent theory tells us that a population of size $N$
-must be run for a multiple of $N$ generations to produce stable genetic patterns.
+Modern computing power easily allows purely \emph{demographic} simulations
+of a population having even hundreds of millions of individuals,
+even though theory tells us that a population of size $N$
+must be run for many multiples of $N$ generations to produce stable genetic patterns.
 However, if our interest lies in the resulting genetic patterns of variation
@@
 As samples of most species genomes harbor tens or hundreds of millions of variant sites,
-naively carrying full genotypes for even modest numbers of individuals through a simulation becomes quickly prohibitive.
+naively carrying full genotypes for even modest numbers of individuals through a simulation 
+becomes quickly prohibitive.
 
-However, it is thought that much of that variation is selectively neutral (or nearly so).
+However, it is thought that most genetic variation is selectively neutral (or nearly so).
 By definition, the alleles carried by individuals in a population at neutral sites
@@
 at each position along the genome.
-In this paper, we show how to use algorithmic tools and data structures developed for the coalescent simulator \msprime\
+In this paper, we show how to use algorithmic tools and data structures developed for the 
+coalescent simulator \msprime{}
 to efficiently record, and later process, this history.
@@
 which results in considerable savings.
-
 Another way of representing this same information
@@
 
+\plr{Quick review of simulation software, esp that other one that does the same thing.}
+
+\plr{Make clear that the main product here is the \msprime{} interface,}
+part of which is the tables schema.
+
 %%%%%%%%%%%%%%%%%%%%%%
-\section*{Methods}
+\section*{Results}
 
-\plr{Reminder of what we need to know in the end (the trees), and
-    quick review of \msprime{} methods: sparse trees, tree differences.}
+\plr{Order is intro-results-discussion-methods.}
+
+Below, we first describe the conceptual and algorithmic foundations for the method:
+(a) the new \emph{tables} format for storing tree sequences;
+(b) an algorithm to output tree sequence information on the fly
+    from a forwards-time simulation;
+and (c) a method to \emph{simplify} a tree sequence, i.e., remove redundant information.
+This is followed by benchmarking comparisons
+with real, forwards-time simulations.
 
 %%%%%%
-\subsection*{Tree sequences}
+\subsection*{Recording a tree sequence}
 
-A tree sequence is an efficient encoding for a sequence of correlated trees such as
-is produced by recombination. The encoding is efficient because branches that are
-shared by adjacent trees are stored once, rather than repeatedly for each tree.
-The topology of a tree sequence is defined via a set of \emph{nodes} and
-\emph{edges}. A node in a tree sequence refers to a distinct ancestor and
-corresponds to the vertices in the individual genealogies along the sequence.
+A \emph{tree sequence} is an encoding for the sequence of correlated trees 
+that describes the history of a sexual population.
+It is efficient because branches that are shared by adjacent trees are stored once, 
+rather than repeatedly for each tree.
+The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
+while information about variants are recorded as \emph{sites} and \emph{mutations}.
+The format for these four tables is given in more detail in the Methods;
+here we explain conceptually,
+using the example of Figure~\ref{fig:example_tree_sequence}.
+
+The \emph{nodes} of a tree sequence 
+correspond to the vertices in the individual genealogies along the sequence,
+and each node refers to a distinct ancestor.
 Since each node represents a specific ancestor, it has a unique ``time'',
 thought of as her birth time, which determines the height of any vertices
-she is associated with. In the example of Figure~\ref{fig:example_tree_sequence}
-we have a total of five nodes. Nodes 0, 1 and 2 occur at time 0 and are our
-samples. Nodes 3 and 4 represent the ancestors of these samples, and were born at
-time 1.5 and 2.5 respectively.
+she is associated with. 
+The example of Figure~\ref{fig:example_tree_sequence} has five nodes:
+nodes 0, 1 and 2 occur at time 0 and are the \emph{samples},
+while nodes 3 and 4 represent those ancestors necessary to record their genealogy,
+who were born at time 1.5 and 2.5 respectively.
 
@@
     \caption{
-        An example tree sequence with three samples over a chromosome of length 10. In the
-        left-most panel we show the tree sequence pictorially in two different ways:
-        in the top are the tree topologies and the bottom shows the spatial extent of the
-        edges that define these topologies. The right-hand panels show the specific encoding
-        of this tree sequence in the four tables (nodes, edges, sites and mutations) defined
-        by \msprime.
+        An example tree sequence with three samples over a chromosome of length 10.
+        The left-hand panel shows the tree sequence pictorially in two different ways:
+        (top) as a sequence of tree topologies 
+        and (bottom) the spatial extent of the edges that define these topologies. 
+        The right-hand panels show the specific encoding
+        of this tree sequence in the four tables (nodes, edges, sites and mutations) 
+        defined by \msprime.
+        \plr{Whoops!  genotype of 1 in the second tree should be `G'!}
         \jk{Some minor alignment issues present here.}
+        \plr{turns out this is no fun to edit in inkscape.  
+            I can re-make to match the other figs
+            but will leave alone for now: it's good!}
         \label{fig:example_tree_sequence}
@@
 
-The topology of a tree sequence is defined by the \emph{edges} which define how
-nodes relate to each other over specific genomic intervals. Each edge is a tuple
-$(\ell, r, p, c)$, where $[\ell, r)$ is a half-open genomic interval defining the
-spatial extend of the edge; $p$ and $c$ are the parent and child nodes specifying
-a single branch in the tree over this interval. The spatial extent of the
-edges defining the topology of Figure~\ref{fig:example_tree_sequence} in the
-bottom left panel. For example, the branch joining nodes 1 to 3 is shared in both trees,
-and is recorded in a single edge extending over the whole chromosome. It is this
-method of capturing the shared structure between adjacent trees that makes the
+The \emph{edges} define how nodes relate to each other over specific genomic intervals. 
+Each edge records: 
+% is a tuple $(\ell, r, p, c)$, where
+the endpoints $[\ell, r)$ of the half-open genomic interval defining the
+spatial extent of the edge; 
+and the identities $p$ and $c$ of the parent and child nodes
+of a single branch that occurs in all trees covering this interval.
+The spatial extent of the edges defining the topology of Figure~\ref{fig:example_tree_sequence}
+are shown in the bottom left panel. 
+For example, the branch joining nodes 1 to 3 appears both trees,
+and so is recorded as a single edge extending over the whole chromosome. 
+It is this method of capturing the shared structure between adjacent trees that makes the
 tree sequence encoding very compact and algorithmically efficient.
 
-Recovering the sequence of trees from a collection of edges is straightforward:
+Recovering the sequence of trees from this information is straightforward:
 each point along the genome that the tree topology changes
@@
 we need only retrieve all edges overlapping that point
-and construct the corresponding edges.
-To construct the tree at a nearby location,
+and construct the corresponding tree.
+To modify the tree to reflect the genealogy at a nearby location,
 we need only remove those edges whose intervals do not overlap that location,
@@
 in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
-2 and ancestral state `A' and the other at position 7 with ancestral state `G'. If
+2 wieth ancestral state `A' and the other at position 7 with ancestral state `G'. If
 no mutations occur, all samples inherit the ancestral state at a given site.
-A mutation occurs over a specific node at a given site, and results in a specific
-derived state. Thus, all samples below the mutation node in the tree will inherit
-this state (unless further mutations are encountered). Three mutations are shown
-in Figure~\ref{fig:example_tree_sequence}, illustrated by red hexagons. The first
-mutation occurs at site zero (which is in the left-hand tree), which is a simple
-mutation resulting in node $2$ inheriting the state 'T'. The second side (in the
-right hand tree) has two mutations: one occurring over node $3$ changing the state to
-`C' and a back mutation over node $1$ changing the state to `G'.
+Each mutation occurs above a specific node at a given site, 
+and results in a specific derived state. 
+Thus, all samples below the mutation node in the tree will inherit this state 
+(unless further mutations are encountered). 
+Three mutations are shown in Figure~\ref{fig:example_tree_sequence}, 
+illustrated by red hexagons. 
+The first mutation occurs at site zero (in the left-hand tree), and is a simple
+mutation resulting in node $2$ inheriting the state 'T'. 
+The second site (in the right hand tree) has two mutations: 
+one occurring over node $3$ changing the state to `C', 
+and a back mutation over node $1$ changing the state to `G'.
 
@@
 the \msprime\ Python API, the time required to load this file into memory was
-around 1.5 seconds, and the time required to iterate over all 1 million tree
+around 1.5 seconds, and the time required to iterate over all 1 million trees
 was 2.7 seconds. In contrast, recording the topological information in Newick
 format would require around 20 TiB and storing the genotype information
-in VCF would require about 1 TiB. Working with either the Newick or VCF encoding
-of this dataset would be exceedingly cumbersome, likely requiring several
+in VCF would require about 1 TiB --
+a compression factor of 200,000.
+Working with either the Newick or VCF encoding
+of this dataset would likely require several
 days of CPU time simply to read the information info memory.
@@
 
+\begin{figure}
+    \begin{center}
+        \includegraphics{simplify-state-diagram}
+    \end{center}
+    \caption{
+        A depiction of the state of the simplification algorithm.
+        \label{fig:simplify_state}
+    }
+\end{figure}
+
 The approach that we take is based on Hudson's algorithm for simulating
@@
 size \citet{durretspatial}), this may not negatively affect realism.
+
+\paragraph{Sequential simplification}
+Even without storing genomes,
+a forwards-time simulation algorithm can rapidly fill a large amount of memory.
+Since much of this is redundant, the \emph{simplify} algorithm can be periodically applied
+to the tree sequence accumulated so far,
+pruning out unnecessary history and greatly reducing the required amount of memory.
+To do this, 
+\emph{simplify} treats the entire current generation as the ``samples'';
+and since node IDs change,
+the simulation procedure must maintain a mapping between the individual IDs 
+used by the simulation and the node IDs in the tree sequence.
+Examples of this procedure are implmemented in our simulation methods (below).
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Results}
+% Outline:
+% 1. estimates and analytical results
+% 2. in practice: measurements from simulations
+\subsection*{Estimates of run-time complexity}
 
-\plr{Estimates of run-time complexity}
+For simple WF alg with exactly one recomb per gen, should be 2Nt edges; post-simplification should be 2N log N * rho edges
+because each tree change is the start of a new edge;
+since 2N log N is also the number of segregating sites, this same thing gives us the computation for mutations
+
 Suppose that we wish to run a forwards-time simulation of $N$ individuals for $T$ generations,
@@
 
+\subsection*{Simulation benchmarks}
+
 Comparison of simulation with/without msprime, using simuPOP
@@
 
-All sorts of good reasons to want to have whole-genome simulations.
+A hybrid approach, in which the recent history of a large sample of
+individuals is simulated under a detailed forwards-in-time model and the deep history
+under the coalescent, is an attractive prospect.
+\jk{Not sure where this should go; perhaps in the discussion? We can make the point that
+by keeping track of the genealogies forwards in time we can easily combine these with the
+deep history simulated by msprime}
+
+Parallelization will be important;
+this requires only one-way communication.
+
+A final note:
+in preparing this manuscript, 
+we debated a number of possible terms for the embellished pedigree,
+i.e., the ``pedigree with ancestral recombination information''.
+Current etymologica consensus \citep{liberman2014little} has
+``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
+(whose branching pattern resembes the bifurcation of a single parent-offspring relationship).
+An analogous term for the embellished pedigree might be \emph{nedigree},
+from ``nid de grue'',
+as the nest of a crange is a large jumble of (forking) branches.
+We thought it unwise to use this term throughout the manuscript,
+but perhaps it will prove useful elsewhere.
 
 %%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements}
+\section*{Methods}
 
-\bibliography{references}
+%%%%%%%%%%%%%%
+\subsection*{Simuation methods}
 
-\appendix
+Jaime and Kevin to describe simulation methods here (separately).
+
+\plr{some of this can move back to an appendix, but since they want methods at the end,
+it's the sort of thing they want here.}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements}
+
+\bibliography{references}
+
+\appendix
+
 \end{document}

### 2017-11-12 forwards_paper.tex
.
more edits

@@
 These considerations, and increasing computational power, have led to a resurgence of interest in forwards-time, individual-based simulations.
-
-%% Merged this above; moved the "hybrid approach" to the conclusion.
-% Another critical assumption of the coalescent is that the sample size $n$ is much smaller
-% than the effective population size $N_e$. 
-% With the very large sample sizes becoming common
-% in human genetics, this is no longer a safe assumption. 
-% For example, a recent study~\citep{martin2017human}
-% simulated 600,000 samples of human chromosome 20 to assess the impact of European biased
-% reference panels in genome wide association studies. While many results of the
-% coalescent are surprisingly
-% robust to violations of the assumption that $n \ll N$, there are instances where genealogical properties
-% are distorted and coalescent simulations may be
-% misleading~\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
-% A hybrid approach, in which the recent history of a large sample of
-% individuals is simulated under a detailed forwards-in-time model and the deep history
-% under the coalescent, is an attractive prospect.
-% \jk{Not sure where this should go; perhaps in the discussion? We can make the point that
-% by keeping track of the genealogies forwards in time we can easily combine these with the
-% deep history simulated by msprime}
 
@@
 in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
-2 wieth ancestral state `A' and the other at position 7 with ancestral state `G'. If
+2 with ancestral state `A' and the other at position 7 with ancestral state `G'. If
 no mutations occur, all samples inherit the ancestral state at a given site.
@@
 in VCF would require about 1 TiB --
-a compression factor of 200,000.
+a compression factor of 144,000.
 Working with either the Newick or VCF encoding
@@
 we need to record two things for each new chromosome:
-the birth time, and the endpoints and parental IDs of each distinctly inherited segment.
+the birth time, and the endpoints and parental IDs of each distinctly inherited segment,
+which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
 For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
@@
 and has $N$ haploid individuals, each carrying a single chromosome of length L.
-For simplicity we assume there is exactly one crossover per generation.
+For simplicity we sample exactly one crossover per generation.
 The probability of death per individual each generation is $\delta$.
-
-Let $\randomuniform(A)$ to be an element of the set $A$ chosen
-uniformly at random (note that each instance of $\randomuniform(A)$ within
-an algorithm listing represents an independent draw from the set $A$).
+We will use
+$\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random 
+(and each such instance represents and independent draw).
 
 \begin{taocpalg}{W}{Forwards-time tree sequence}
-{Runs a forwards time Wright--Fisher simulation of a populations with $N$
-individuals with a probability of death per generation $\delta$ and a chromosome of length
+{Simulates a randomly mating population of $N$ haploid individuals,
+with a probability of death per generation $\delta$ and a chromosome of length
 $L$ for $T$ generations. On termination, there will be $n$ nodes with times recorded
@@
 \algstep{W2.}{Generation loop.}{ If $t = 0$ terminate.
-Set $P'_j \leftarrow P_j$ for $0\leq j < N$ and then set $t \leftarrow t - 1$ and $j \leftarrow 0$.
+Set $P' \leftarrow P$ and then set $t \leftarrow t - 1$ and $j \leftarrow 0$.
 }
@@
 
-\jk{I haven't checked this carefully. If we used 1-based indexes it might make things a bit
-simpler}.
 \end{taocpalg}
@@
 $P_a$ and $P_b$ in step \algref{W7}.
-For example, if we assume that infinite sites mutations
-occur at rate $\mu$ per generation then there are $\mbox{Poisson}\left((r - \ell)(\tau_p
+For example, if we assume that mutations occur according to the infinite sites model
+at rate $\mu$ per generation per unit length then there are 
+$\mbox{Poisson}\left(\mu (r - \ell)(\tau_p
 - \tau_c)\right)$ mutations on each edge $(\ell, r, p, c)$. Each of these mutations
-will occur on a distinct site $x$ drawn uniformly from $(\ell, r]$ and be over node $c$.
+will occur on a distinct site $x$ drawn uniformly from $(\ell, r]$ and be associated with node $c$.
 It is straightforward to record this information during the simulation, but it
 is significantly simpler and more efficient to generate these mutations
-after the simulation has completed. In the case of infinite sites mutations
-there is no difference between generating these mutations during the forwards
-in time simulation and after it has completed. We are using precisely the same
-information.
+after the simulation has completed. 
+% In the case of infinite sites mutations
+% there is no difference between generating these mutations during the forwards
+% in time simulation and after it has completed. We are using precisely the same
+% information.
 
@@
 
-It is significantly more efficient to generate mutations as a separate process
-after the topological simulation has completed because we will generate many
-mutations that will be lost in the population. Figure~\ref{fig:wf-trees} shows
-an example of a marginal genealogy produced by a forwards-time Wright-Fisher
-process like Algorithm~\algref{W}. There are two different trees in this
-figure: one shows all the edges output by the simulation and the other
+Besides the cost of transmitting genotypes,
+generating mutations as a separate process after the topological simulation has completed 
+avoids generating the many mutations that are lost in the population. 
+Figure~\ref{fig:wf-trees} shows
+an example of a marginal genealogy produced by a forwards-time Wright--Fisher
+process like Algorithm~\algref{W}. 
+On the left is the tree showing all the edges output by the simulation, 
+while on the right
 is the minimal tree representing the ancestry of the currently alive
@@
 not attempt to track coalescence events but simply record all parent-child
-relationships in the history of the population. As such, my of these edges
+relationships in the history of the population. As such, many of these edges
 will record the simple passing of genealogical information from parent to child
@@
 trees. The second source of redundancy in the output of Algorithm~\algref{W}
-is due to the fact that lineages die out: there will be a large number of
-individuals in the simulation that leave no ancestors in the present day
-population. Node 26 in Figure~\ref{fig:wf-trees}a, for example, leaves no
-ancestors in the current population and so the entire path tracing back to
+is due to the fact that lineages die out: a large number of
+individuals in the simulation leave no descendants in the present day population. 
+Node 26 in Figure~\ref{fig:wf-trees}a, for example, leaves no
+ancestors in the current population, and so the entire path tracing back to
 the root is redundant.
 
-One of the reasons that it is more efficient to generate mutations on the
-genealogies after we have completed the forwards in time simulation is
-because we avoid the cost of generating mutations on these dead-end branches.
-It is only with the benefit of hindsight that we can know which lineages end
-up being ancestral to the sample, and we can avoid a substantial cost of
-generating and maintaining mutational information on these evolutionary
-dead-ends. Assuming mutations are rare, another  reason that it is more efficient
-to place mutations on the trees afterwards is that we have far fewer edges
-in the trees: because we have removed all the intermediate `unary' nodes, we
-have fewer edges to generate mutations on.
-
-There are many advantages to having minimal genealogies such as show in
-Figure~\ref{fig:wf-trees}. Computing this minimal representation of the
+A minimal genealogy such as in Figure~\ref{fig:wf-trees}
+has fewer edges on which to place mutations.
+There are many more advantages to simplifying the genealogies to this point.
+Computing this minimal representation of the
 edges output by a forward-time simulation is an instance of a more
-general problem that we refer to as `tree sequence simplification'. We discuss
-this problem and an efficient solution in the next subsection.
+general problem that cover in the next section.
 

### 2017-11-12 forwards_paper.tex
moar edits

@@
 
-Suppose we have a tree sequence consisting of a set of nodes and edges as
-described earlier, and we have some subset of the nodes in this tree sequence
-that we are interested in (our `samples').  We wish to reduce the input tree sequence
+Suppose that we are only interested in a subset of the nodes (our `samples')
+of a tree sequence,
+and so wish to reduce this input tree sequence
 into the minimum representation of the topologies that include the specified
@@
 \begin{enumerate}
-\item we must have the same marginal trees with respect to the samples as
-the input;
-\item within the marginal trees, all non-sample vertices must has at least
-two children (i.e., unary tree vertices are removed);
-\item any nodes and edges not reachable from the sampled nodes are removed;
-\item there are no adjacent redundant edges, i.e, pairs of edges $(\ell, x, p,
+\item All marginal trees must match the subtree induced by the samples of the corresponding tree in the input tree sequence.
+\item Within the marginal trees, all non-sample vertices must has at least
+two children (i.e., unary tree vertices are removed), unless the vertex is a sample.
+\item Any nodes and edges not ancestral to the sampled nodes are removed.
+\item There are no adjacent redundant edges, i.e, pairs of edges $(\ell, x, p,
 c)$ and $(x, r, p, c)$ which can be represented with a single edge
@@
         \textbf{Top:} the diagram on the left relates ten haploid individuals to each other.
-        It is recorded, in forwards time,
-        in 10 node records (one for each individual)
+        It has 10 node records (one for each individual)
         and 14 edge records (one for each distinctly inherited segment).
-        Blue numbers denote crossing over locations in each meiosis.
+        Blue numbers denote crossing over locations in each meiosis --
+        for instance, $C$ and $D$ were parents to $F$,
+        who inherited the left 70\% of the chromosome from $D$ and the remainder from $C$.
         The individuals $B$, $C$, and $D$ inherit clonally from $A$.
@@
         The mapping from labels in the forwards time simulation to nodes in the tree sequence
-        is shown in red,
-        which allows additional records to be added as the simulation progresses.
+        is shown in red.
+        % which allows additional records to be added as the simulation progresses.
         \label{fig:method_diagram}
@@
 
-In the current context of running forwards in time simulations, the tables of
-nodes and edges that we record all of history for everyone alive at any time through the simulation.
+The tree sequences produced by forwards simulations, 
+record all of history for everyone alive at any time through the simulation.
 This is much more than we need to reconstruct the genealogies and sequences
-of the current day population. We can use the process of simplification here to
-reduce this down to the minimal set of nodes and edges required to represent
-the information that we are interested in. We can also use simplification if  we have some very
+of the current population.
+Simplification is essential to reduce this 
+to a manageable quantity that still contains all
+the information that we are interested in. 
+Simplification is also useful if we have a
 large tree sequence representing a large dataset and we wish to extract the
@@
 
-Roughly, simplification works by tracing ancestry from the samples backwards
-through the recorded history, adding node and edge records to the output
-only when coalescent events are reached. This works exactly as in \msprime,
-allowing substantial re-use of algorithms; the main difference being that
-parental choice, mutations, and recombination locations are determined by the
-input tree sequence rather than randomly generated. An example is shown in
-Figure \ref{fig:method_diagram}.
+The approach that we take is based on Hudson's algorithm for simulating
+the coalescent with recombination~\citep{hudson1983properties,kelleher2016efficient};
+the implementaiton of simplify parallels the implementation of Hudson's algorithm in \msprime.
+Conceptually, this works by 
+(a) beginning by painting the chromosome each sample a distinct color;
+(b) moving back through history, 
+copying the colors of each chromosome to the portions of its' parental chromosomes
+from which it was inherited;
+(c) each time we would paint two colors in the same spot (a coalescence),
+record that information as an edge and instead paint a brand-new color; 
+and
+(d) once all colors have coalesced on a given segment,
+stop propagating it.
+This ``paint pot'' description misses some details --
+for instance, we must ensure that all coalescing segments in a given individual
+are assigned the \emph{same} new color --
+but is reasonably close.
+The process is depicted in Figure~\ref{fig:simplify_state}.
 
@@
     \caption{
-        A depiction of the state of the simplification algorithm.
+        A depiction of the state of the simplification algorithm
+        at each point in time,
+        in the example of Figure~\ref{fig:method_diagram}.
         \label{fig:simplify_state}
@@
 
-The approach that we take is based on Hudson's algorithm for simulating
-the coalescent with
-recombination~\citep{hudson1983properties,kelleher2016efficient}. The main
-state of the algorithm is a set of ancestral segments; each segment $(\ell, r,
-u)$ represents a tract of ancestral material in which the node $u$ is
-associated with the genomic interval $[\ell, r)$. In the same manner
-as Hudson's algorithm, we represent the set of extant lineages as a
-set of individuals, where each ancestral individual is composed of a
-linked list of these segments. The node $u$ in each ancestral sequence
+More concretely,
+the algorithm works by moving back through time,
+processing each edge in the input tree sequence in chronological order.
+The main state of the algorithm at each point in time is a set of ancestral segments:
+each segment $(\ell, r, u)$ represents a tract of ancestral material in which the node $u$ is
+associated with the genomic interval $[\ell, r)$. 
+These segments are stored in a collection of linked lists,
+one list of segments for each ancestral lineage present at that time.
+(These lineages are the ``colors'' above.)
+The node $u$ in each ancestral sequence
 is the node ID associated with this segment in the output tree sequence.
@@
 
+\plr{some of this can move back to an appendix, but since they want methods at the end,
+it's the sort of thing they want here.}
+
 %%%%%%%%%%%%%%
@@
 
-\plr{some of this can move back to an appendix, but since they want methods at the end,
-it's the sort of thing they want here.}
+%%%%%%%%%%%%%%
+\subsection*{Simplification}
 
+Here we describe the simplification algorithm.
+\plr{I am trying it out to see what it looks like.}
+
+\begin{taocpalg}{S}{Simplify a tree sequence}
+{Simplifies the input tree sequence.
+}
+
+\algstep{S0.}{Initialize.}{
+    Set the samples.
+}
+
+\algstep{S1.}{Input parent loop.}{Loop over parents, chronologically.
+}
+
+\algstep{S2.}{Process parent edges.}{Loop over edges associated with this parent, removing ancestry from each child.
+}
+
+\algstep{S3.}{Merge ancestors.}{Merge the resulting segments into a new ancestor for this parent.
+}
+
+\end{taocpalg}
+
+This uses
+\begin{taocpalg}{R}{Remove ancestry}
+    {Remove ancestry on a segment from an ancestor.}
+
+\end{taocpalg}
+and
+\begin{taocpalg}{M}{Merge ancestors}
+    {Merge a collection of segments and create a new ancestor.}
+
+\end{taocpalg}
+
+%%%%%%%%%%%%%%
 \subsection*{Data structures:}

### 2017-11-15 forwards_paper.tex
changed title; begun simplify alg writeup

@@
 
-\title{Efficient forwards-time simulation with tree sequences}
+\title{Efficient pedigree recording for fast population genetics simulation}
 \author{Jerome Kelleher,
@@
 
-Modern computing power easily allows purely \emph{demographic} simulations
-of a population having even hundreds of millions of individuals,
+Modern computing power easily allows simulations of birth, death and reproduction
+in a population having even hundreds of millions of individuals,
 even though theory tells us that a population of size $N$
@@
 
-A \emph{tree sequence} is an encoding for the sequence of correlated trees 
-that describes the history of a sexual population.
+A \emph{tree sequence} is an encoding for the sequence of correlated trees,
+such as those that describe the history of a sexual population.
 It is efficient because branches that are shared by adjacent trees are stored once, 
@@
 edges output by a forward-time simulation is an instance of a more
-general problem that cover in the next section.
+general problem that we cover in the next section.
 
@@
 Here we describe the simplification algorithm.
-\plr{I am trying it out to see what it looks like.}
+The flow is close to our implementation,
+but the description below uses set operations,
+while in the implementation,
+all ancestors are maintained as linked lists of segments
+ordered by left endpoint.
 
 \begin{taocpalg}{S}{Simplify a tree sequence}
-{Simplifies the input tree sequence.
+{Simplifies the input tree sequence:
+    input consists of 
+    a list of sample IDs, $S$, of length $n$,
+    a list of edges, $\mathcal{E_I}$,
+    and a list of nodes $\mathcal{N_I}$.
+    The genome length is $L$.
+    The output will be stored in a list of edges $\mathcal{E_O}$ 
+    and nodes $\mathcal{N}_O$.
+    This also maintains the list $N$ that gives the mapping 
+    from $\mathcal{N_I}$ to $\mathcal{N_O}$,
+    so that if $M[p] = u$ then the $p^\text{th}$ ancestor in $N_I$
+    is the same as the $u^\text{th}$ ancestor in $N_0$.
+    The internal state is $\mathcal{A}$, the list of ancestors
+    (which are collections of ancestral segments).
 }
@@
 \algstep{S0.}{Initialize.}{
-    Set the samples.
+    Set $\mathcal{A}[j] \leftarrow (0, L, j)$ 
+    and $\mathcal{N}_O[j] \leftarrow \mathcal{N}_I[S[j]]$ for $0 \le j < N$.
+    Set $Q \leftarrow \emptyset$,
+    and $M[p] \leftarrow -1$ for $0 \le p < \text{length}(N_I)$.
 }
@@
 
-\algstep{S2.}{Process parent edges.}{Loop over edges associated with this parent, removing ancestry from each child.
+\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} 
+    on each edge corresponding to the current parent.
 }
@@
 
-This uses
+The first step is, for each edge $(\ell, r, p, c)$,
+simply to replace any ancestry segment $[a, b)$ held by ancestor $c$
+with $[a,b) \setdiff [\ell, r)$ 
+(deleting the segment or splitting in two if necessary),
+and adding the removed segment $[a,b) \cap [\ell, r)$ to the merge queue $Q$:
 \begin{taocpalg}{R}{Remove ancestry}
-    {Remove ancestry on a segment from an ancestor.}
+    {Given an edge $(\ell, r, p, c)$,
+    and a list $\mathcal{A}[c]$ of the $m$ ancestral segments corresponding to $c$,
+    remove segments overlapping $[\ell, r)$ from $\mathcal{A}[c]$,
+    and add those overlapping segments to $Q$.
+    }
+
+    \algstep{R1.}{Segment loop.}{Loop over ancestral segments in $\mathcal{A}[c]$:
+    if $j=m$, terminate; else let $(a, b, u) = \mathcal{A}[c][j]$.
+    }
+
+    \algstep{R2.}{Add to merge queue.}{If $[a,b) \cap [\ell,r) \neq \emptyset$,
+    append $(\max(a,\ell), \min(b,r), u)$ to $Q$.
+    }
+
+    \algstep{R3.}{Remove ancestry.}{If $[a,b) \cap [\ell,r) != \emptyset$,
+    delete $\mathcal{A}[c][j]$ and insert in its place
+    $(\min(a,\ell), \max(a,\ell), u)$ (if $a<\ell$) 
+    and/or $(\min(b,r), \max(b,r), u)$ (if $r<b$).
+    }
 
 \end{taocpalg}
-and
+
+Next we must merge these ancestry segments,
+creating the new ancestor $p$,
+outputting edges and a node if coalescence has occurred.
 \begin{taocpalg}{M}{Merge ancestors}
-    {Merge a collection of segments and create a new ancestor.}
+    {Merge a collection of segments and create a new ancestor.
+    This operates on the merge queue $Q$ above,
+    which is a list of ancestral segments,
+    and creates the new ancestor $\mathcal{A}[p]$.}
+
+    \algstep{M1.}{Segment loop.}{While $\mathrm{length}(Q)>0$:}
+
+    \algstep{M2.}{Find next segment overlap boundaries}{
+    Let $\ell \leftarrow \min\{x.\ell : x \in Q\}$
+    and $H \leftarrow \{x \in Q : x.\ell = \ell\}$.
+    Also let $\ell_+ \leftarrow \min\{x.\ell : x \in Q \setdiff H\}$
+    and $r = \min(\ell_+, \min\{x.r : x \in H\})$.
+    }
+
+    \algstep{M3.}{Copy single segment.}{If $\mathrm{length}(H) == 1$,
+    set $\alpha \leftarrow (\ell, r, H[0].n)$.
+    If also $p \in S$, then append $(\alpha.\ell, \alpha.r, N[p], \alpha.c)$ to $\mathcal{E}_O$
+    and then set $\alpha.c \leftarrow M[p]$.
+    Go to \algref{M7}.
+    }
+
+    \algstep{M4.}{Coalescence head.}{If $(h \leftarrow \mathrm{length}(H)) > 1$,
+    if $M[p] == -1$ set $M[p] \leftarrow \text{length}(N_O)$
+    and append $N_I[p]$ to $N_O$.
+    Set $\alpha \leftarrow (\ell, r, M[p])$.
+    }
+
+    \algstep{M6.}{Record edges.}{
+    For each $x \in H$, append $(\ell, r, M[p], x.n)$ to $\mathcal{E}_O$.
+    }
+
+    \algstep{M7.}{Update $Q$.}{
+    Set $Q \leftarrow Q \setdiff \{x \in Q : x.r == r\}$
+    and then set $x.\ell = \ell$ for all remaining $x \in Q$.
+    }
+
+    \algstep{M8.}{Add segment to ancestor.}{
+    Append $\alpha$ to $\mathcal{A}[p]$ and return to \algref{M1}.
+    }
 

### 2017-11-15 sims/space-complexity.py
First pass at script to compute space complexity.

@@ -1,0 +1,90 @@
+"""
+Code for investigating the space complexity of algorithm W along
+with serial simplification.
+"""
+import random
+import numpy as np
+import msprime
+
+import matplotlib
+# Force matplotlib to not use any Xwindows backend.
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+
+def wright_fisher(N, T, simplify_interval=1):
+    """
+    An implementation of algorithm T where we simplify after every generation.
+    The goal here is to measure the number of edges in the tree sequence
+    representing the history as a function of time.
+
+    For simplicity we assume that the genome length L = 1 and the probability
+    of death delta = 1.
+    """
+    L = 1
+    edges = msprime.EdgeTable()
+    nodes = msprime.NodeTable()
+    P = [j for j in range(N)]
+    for j in range(N):
+        nodes.add_row(time=T, flags=1)
+    t = T
+    S = np.zeros(T, dtype=int)
+    while t > 0:
+        t -= 1
+        Pp = [P[j] for j in range(N)]
+        for j in range(N):
+            n = len(nodes)
+            nodes.add_row(time=t, flags=1)
+            Pp[j] = n
+            a = random.randint(0, N - 1)
+            b = random.randint(0, N - 1)
+            x = random.uniform(0, L)
+            edges.add_row(0, x, P[a], n)
+            edges.add_row(x, L, P[b], n)
+        P = Pp
+        if t % simplify_interval == 0:
+            msprime.sort_tables(nodes=nodes, edges=edges)
+            msprime.simplify_tables(Pp, nodes, edges)
+            P = list(range(N))
+        S[T - t - 1] = len(edges)
+    # We will always simplify at t = 0, so no need for special case at the end
+    return msprime.load_tables(nodes=nodes, edges=edges), S
+
+
+def verify():
+    """
+    Crude check to see if we get the same result when we serial simplify and we
+    don't.
+    """
+    for seed in range(1, 10):
+        for n in [5, 10, 20]:
+            for T in [1, 10, 100]:
+                print("Checking ", n, T, seed)
+                random.seed(seed)
+                ts1, _ = wright_fisher(n, T, T + 1)
+                random.seed(seed)
+                ts2, _ = wright_fisher(n, T, 1)
+                assert ts1.tables.nodes == ts2.tables.nodes
[…]

### 2017-11-15 forwards_paper.tex
pdf compiles now
wrote scaling section

--- 2017-11-15 forwards_paper.tex
@@
 \newcommand{\mutations}{\texttt{mutations}}
+\newcommand{\setdiff}{\smallsetminus}
 
@@
 the algorithm works by moving back through time,
-processing each edge in the input tree sequence in chronological order.
-The main state of the algorithm at each point in time is a set of ancestral segments:
-each segment $(\ell, r, u)$ represents a tract of ancestral material in which the node $u$ is
-associated with the genomic interval $[\ell, r)$. 
-These segments are stored in a collection of linked lists,
-one list of segments for each ancestral lineage present at that time.
+processing each parent in the input tree sequence in chronological order.
+The main state of the algorithm at each point in time is a set of ancestral lineages,
+and each lineage is a collection of ancestral segments.
+An ancestral segment $(\ell, r, u)$ is found in a lineage
+if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
+% These segments are stored in a collection of linked lists,
+% one list of segments for each ancestral lineage present at that time.
 (These lineages are the ``colors'' above.)
-The node $u$ in each ancestral sequence
-is the node ID associated with this segment in the output tree sequence.
-We maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
+We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
 the input tree sequence.
-
-\jk{I'm not sure this is the right level of detail to tackle this description
-at. What's the goal of this algorithm description? Are we trying to
-establish some properties of the algorithm or just give the reader a high-level
-flavour of how it works??}
-
-In this scheme, at any point in the simulation genealogical history is recorded
+Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence..
+
+\paragraph{Sequential simplification}
+At any point the simulation scheme above, genealogical history is recorded
 in a tree sequence. This has two additional advantages. First, simplification
@@
 
-\paragraph{Sequential simplification}
-Even without storing genomes,
-a forwards-time simulation algorithm can rapidly fill a large amount of memory.
-Since much of this is redundant, the \emph{simplify} algorithm can be periodically applied
-to the tree sequence accumulated so far,
-pruning out unnecessary history and greatly reducing the required amount of memory.
-To do this, 
-\emph{simplify} treats the entire current generation as the ``samples'';
-and since node IDs change,
-the simulation procedure must maintain a mapping between the individual IDs 
-used by the simulation and the node IDs in the tree sequence.
-Examples of this procedure are implmemented in our simulation methods (below).
-
 %%%%%%
 \subsection*{Overview of the API}
+
+\plr{leaving this alone for the moment: to stay here? move?}
 
@@
 
-% Outline:
-% 1. estimates and analytical results
-% 2. in practice: measurements from simulations
 \subsection*{Estimates of run-time complexity}
 
-For simple WF alg with exactly one recomb per gen, should be 2Nt edges; post-simplification should be 2N log N * rho edges
-because each tree change is the start of a new edge;
-since 2N log N is also the number of segregating sites, this same thing gives us the computation for mutations
-
-Suppose that we wish to run a forwards-time simulation of $N$ individuals for $T$ generations,
-in which there are $S$ selected loci and $L$ neutral loci using a similar
-method to Algorithm~\algref{W}.
-We will estimate run-time complexity and memory usage for both a ``naive'' strategy that carries along neutral loci
-and an ``ancestry-tracking'' strategy like that we consider here.
-To do this, we assume that each individual must carry along its entire genotype.
-More advanced schemes are used in some simulators,
-but these increase efficiency by utilizing redundancy introduced by shared ancestry,
-which is effectively an intermediate scheme.
-We omit the cost of computing a fitness function.
-
-Both schemes must choose mates and recombination breakpoints,
-and pass on selected genotypes.
-The difference between the two comes from the tradeoff between
-(a) passing on neutral genotypes, and
-(b) recording and simplifying the tree sequence, and adding neutral genotypes afterwards.
-(We assume here that selected genotypes are stored in the same way for both.)
-Passing on neutral genotypes naively records $L$ items per individual each
-generation, discarding the previous generation, giving an overall complexity of
-$O((L + S) NT)$. Tracking the $S$ selected loci and recording a tree sequence
-using Algorithm~\algref{W} requires $O(SNT)$ time and space, since we
-record at most two edges for each individual in each generation.
-However, after simplification the tree sequence contains
-$O(??)$ edges since []. The cost of generating $L$ neutral mutations
-is then $O(L + ??)$. Assuming that $N$, $T$ and $L$ are large, this is
-very much smaller than the straightforward $O((L + S) NT)$ time required
-to simulate all mutations directly.
-
-\jk{Not sure where $\rho$ comes in here now. Thinking about this, shouldn't we
-be able to come up with some expression for the number of edges? Assuming
-the very simple single breakpoint per generation of Alg W, what is the
-expected number of breakpoints? This is a function of $N$ and $T$, right?
-It's not clear to me how many breakpoints this ends up giving you in
-the output tree sequence and how many edges this corresponds to.
-It's probably worthwhile thinking about this a bit.}
-
-% However, after simplification a tree sequence for all $N$ individuals
-% only takes of order $N + \rho \log N$ records.
-% Simplification requires processing each of the initial records -- so, order of $N \times T \times \rho$ operations;
-
-% \begin{table}
-% \begin{center}
-%     \begin{tabular}{l||r|r||r|r|}
-%                     & \multicolumn{2}{|c|}{naive} & \multicolumn{2}{|c|}{ancestry-tracking} \\
-%                     & operations  & memory & operations  & memory \\
-%         mate choice & $N$       &   --  &   $N$     & -- \\
-%         recombination & $N\rho$ &   --  &   $N\rho$ & -- \\
-%     \end{tabular}
-% \end{center}
-% \end{table}
+Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals 
+using Algorithm~\algref{W}
+for $T$ generations,
+with the probability of death, $\delta$, set to 1 for simplicity.
+Since there is exactly one crossing over every generation,
+this produces tables of 
+$2NT$ nodes and
+$4NT$ edges.
+Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
+so that $T \sim 20N$, say.
+After simplification, we are left with the tree sequence describing the history
+of only the current generation of $2N$ individuals.
+This tree sequence has $4N-2$ edges to describe the leftmost tree,
+and one additional edge each time the marginal tree changes along the sequence.
+Coalescent theory tells us that 
+the expected total length of edges in a marginal tree is approximately $4N\log(2N)$,
+which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
+since we have taken one crossover per generation.
+Not all such recombinations actually change the tree topology,
+and so this gives us an upper bound on the number of edges.
+Similarly, not every new edge derives from a never-before-seen node,
+but the number of nodes is at most equal to the number of edges plus the sample size.
+With $T=20N$, this reduces the initial storage of $160 N^2$ items to $2N(4\log(2N) + 3)$;
+at $N=10^4$, a factor of 20,000.
+
+\plr{We could compute the recursions for these quantities explicitly for this simple WF case...}
+
+If we were to add mutations in forwards time 
+under the infinite-sites model with total mutation rate per generation $\mu$,
+there would also be around $\mu 2NT$ mutations (and the same number of sites).
+Since mutations fall as recombinations do on the marginal trees,
+adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations.
+This furthermore avoids the computational burden of propagating mutations forwards through the generations.
+
+Our method stores genealogies, and so records substantially more information 
+than would a method only recording genotypes.
+However, since the simplification algorithm requires some computational effort,
+it is still informative to compare computational complexity of the algorithm
+to one that propagates neutral genotypes.
+A typical individual will differ at $2 N \mu$ sites from the population's consensus sequence,
+so propagating these to offspring by simple copying will take $4 N^2 \mu$ operations per generation.
+On the other hand, we must store $13N$ quantities per generation (two edges and one node per individual).
+The simplification algorithm is linear in the number of edges of the input tree sequence,
+and so multiplies this by a constant factor.
+Concretely, propagating neutral genotypes for $T$ generations has time complexity $O(\mu T N^2)$,
+while our implementation is only $O(T N \log(2N))$.
 

### 2017-11-15 sims/space-complexity.py
Updated plots and corrected error.

@@ -15,7 +15,7 @@ import seaborn as sns
 
 def wright_fisher(N, T, simplify_interval=1):
     """
-    An implementation of algorithm T where we simplify after every generation.
+    An implementation of algorithm W where we simplify after every generation.
     The goal here is to measure the number of edges in the tree sequence
     representing the history as a function of time.
 
@@ -71,16 +71,20 @@ def verify():
 def plot():
     num_reps = 10
     for n in [10, 100, 1000]:
+        T = 10 * n
+        A = np.zeros((num_reps, T))
         for j in range(num_reps):
-            _, S = wright_fisher(n, 100)
-            plt.plot(S)
+            _, S = wright_fisher(n, T)
+            A[j] = S
+            plt.plot(S, alpha=0.5)
             print(n, j, "done")
+        mean = np.mean(A, axis=0)
+        plt.plot(mean, color="black", lw=3)
         plt.title("N = {}".format(n))
         plt.xlabel("Generations")
         plt.ylabel("Number of edges")
         plt.savefig("{}.png".format(n))
         plt.clf()
-
 
 
 

### 2017-11-16 sims/wf_expectations.R
script to calculate total expected height of WF tree

@@ -1,0 +1,35 @@
+library(matrixcalc)
+
+N <- 100
+
+# get the WF transition matrix
+# stirling number of the second kind:
+# S[n,k] = number of ways to partition n elements into k nonempty sets
+S <- stirling.matrix(N)[-1,-1]
+lS <- log(S)
+# P[n,k] = prob put n balls in N boxes and have exactly k nonempty boxes somewhere
+#        = S[n,k] * choose(N,k) * factorial(k) / N^n
+P <- exp( sweep(lS, 2, lchoose(N,1:N) + lfactorial(1:N), "+") - (1:N) * log(N) )
+stopifnot(all(abs(P[upper.tri(P,diag=FALSE)]) < 1e-8))
+P[upper.tri(P,diag=FALSE)] <- 0
+P[!is.finite(P)] <- 0
+stopifnot(all(abs(rowSums(P) - 1) < 1e-8))
+
+
+# Solve for the mean occupation times:
+#  v[n,k] = mean time spent at k starting at n
+#         = delta(n,k) + sum_j( P[n,j] * v[j,k] )
+# for k > 1
+#  v = I + P %*% v
+V <- forwardsolve(diag(N-1) - P[-1,-1], diag(N-1))
+# T[n] = expected total length of tree starting at n
+T0 <- rowSums(sweep(V, 2, 2:N, "*"))
+
+# Or, do this directly:
+#  T[n] = n + sum_j( P[n,j] * T[j] )
+#  T = (2:N) + P %*% T
+T1 <- forwardsolve(diag(N-1) - P[-1,-1], 2:N)
+
+plot(T1, type='l', ylim=range(0,T1,finite=TRUE))
+lines(T0, col='green')
+lines(2:N, 2 * N * cumsum(1/(2:N-1)), col='red')

### 2017-11-16 sims/space-complexity.py
added tree length

@@ -30,6 +30,7 @@ def wright_fisher(N, T, simplify_interval=1):
         nodes.add_row(time=T, flags=1)
     t = T
     S = np.zeros(T, dtype=int)
+    L = np.zeros(T, dtype=float)
     while t > 0:
         t -= 1
         Pp = [P[j] for j in range(N)]
@@ -48,9 +49,16 @@ def wright_fisher(N, T, simplify_interval=1):
             msprime.simplify_tables(Pp, nodes, edges)
             P = list(range(N))
         S[T - t - 1] = len(edges)
+        L[T - t - 1] = total_length(nodes, edges)
     # We will always simplify at t = 0, so no need for special case at the end
-    return msprime.load_tables(nodes=nodes, edges=edges), S
+    return msprime.load_tables(nodes=nodes, edges=edges), S, L
 
+def total_length(nodes, edges):
+    ts = msprime.load_tables(nodes=nodes, edges=edges)
+    L = 0.0
+    for t in ts.trees():
+        L += t.length * t.total_branch_length
+    return L
 
 def verify():
     """
@@ -62,9 +70,9 @@ def verify():
             for T in [1, 10, 100]:
                 print("Checking ", n, T, seed)
                 random.seed(seed)
-                ts1, _ = wright_fisher(n, T, T + 1)
+                ts1, _, _ = wright_fisher(n, T, T + 1)
                 random.seed(seed)
-                ts2, _ = wright_fisher(n, T, 1)
+                ts2, _, _ = wright_fisher(n, T, 1)
                 assert ts1.tables.nodes == ts2.tables.nodes
                 assert ts1.tables.edges == ts2.tables.edges
 
@@ -73,16 +81,28 @@ def plot():
     for n in [10, 100, 1000]:
         T = 10 * n
         A = np.zeros((num_reps, T))
+        B = np.zeros((num_reps, T))
         for j in range(num_reps):
-            _, S = wright_fisher(n, T)
+            _, S, L = wright_fisher(n, T)
             A[j] = S
+            B[j] = L
+            plt.subplot(211)
             plt.plot(S, alpha=0.5)
+            plt.subplot(212)
+            plt.plot(L, alpha=0.5)
             print(n, j, "done")
-        mean = np.mean(A, axis=0)
-        plt.plot(mean, color="black", lw=3)
+        mean_S = np.mean(A, axis=0)
+        mean_L = np.mean(B, axis=0)
+        plt.subplot(211)
[…]

### 2017-11-16 forwards_paper.tex
did through discussion

@@
 be \emph{begun} with a tree sequence produced by some other method -- for
-instance, by a coalescent simulation with \msprime. This allows for
-incorporation of deep-time history beyond the reach of individual-based
-simulations. Since geographic structure from times longer ago than the mixing
-time of migration across the range has limited effect on modern genealogies
-\citep{wilkins_separation} (other than possibly changing effective population
-size \citet{durretspatial}), this may not negatively affect realism.
+instance, by a coalescent simulation with \msprime.
 
@@
 
-\plr{We could compute the recursions for these quantities explicitly for this simple WF case...}
-
 If we were to add mutations in forwards time 
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\section*{Conclusion}
-
-This works great and very fast.
-
-Having results in a tree sequence is really good downstream too:
-The tree sequence encoding is compact but it is also efficient to process.
-Many algorithms to compute statistics of interest for population genetics
-are naturally expressed in terms of tree topologies. For example, the
-pairwise nucleotide diversity $\pi$, is defined as the average number of
-differences between sequences in the sample. Computing $\pi$ directly
-from observed sequence data requires $O(n^2 m)$ time, since we must
-compare all pairs of samples at every site. However, given the topology of
-the trees at every site and the locations of mutations much better algorithms
-are possible. Using the fact that we can count the number of samples below
-a given node efficiently using previously described tree sequence
-algorithms~\citep{kelleher2016efficient}, the time required to compute $\pi$
-becomes roughly $O(n \log n + m)$.
+\section*{Discussion}
+
+In this paper, we have shown that storing pedigrees
+and associated recombination events
+in a forwards-time simulation
+not only results in having available a great deal more information about the simulation,
+but also can speed up the simulation by orders of magnitude.
+To make this feasible, 
+we have described how to efficiently store this information in tables,
+and have described a fundamental algorithm for simplificication of tree sequences.
+Conceptually, recording of genealogical and recombination events
+can happen independently of the details of simulation;
+for this reason, we provide a well-defined and well-tested API in C and in python
+for use in other code bases.
+
+The tree sequences produced by default by this method
+are very compact, storing genotype \emph{and} genealogical information
+in a small fraction of the space taken by a compressed VCF file.
+The format is also highly efficient to process,
+leading to advantages for downstream analysis as well.
+This is because many algorithms to compute statistics of interest for population genetics
+are naturally expressed in terms of tree topologies,
+and so can be quickly computed from the trees underlying the tree sequence format. 
+For example, pairwise nucleotide diversity $\pi$, is the average density of
+differences between sequences in the sample. 
+To compute this directly from sequence data at $m$ sites in $n$ samples
+requires computing allele frequencies, taking $O(nm)$ operations.
+By using the locations of the mutations on the marginal trees,
+and the fact that these are correlated, 
+dynamic programming algorithms similar to those in~\citep{kelleher2016efficient}
+can do this in roughly $O(n + m \log n)$ operations.
 The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
-samples in a tree sequence, and calculating $\pi$ over all 500K samples
-in the example above required about 1.2 seconds.
-\jk{I haven't checked the analysis of computing $\pi$ here and not thought about
-it too deeply. I guess we should also compare this with something else in order to
-show that this time is pretty good. We could get a subset of the sites and push
-the data into pylibseq maybe, or scikit-allel??}
-
-This is a general-purpose strategy that can be applied to other methods.
-
-A hybrid approach, in which the recent history of a large sample of
-individuals is simulated under a detailed forwards-in-time model and the deep history
-under the coalescent, is an attractive prospect.
-\jk{Not sure where this should go; perhaps in the discussion? We can make the point that
-by keeping track of the genealogies forwards in time we can easily combine these with the
-deep history simulated by msprime}
-
-Parallelization will be important;
-this requires only one-way communication.
+samples in a tree sequence, which took about 1.2 seconds
+applied to the example tree sequence above with 1.1 million mutations 
+in 200 megabases for 500,000 samples.
+
+Another attractive feature of this set of tools
+is that it makes it easy to incorporate \emph{prior history},
+simply by seeding the simulation with a (relatively inexpensive) coalescent simulation.
+This allows for incorporation of deep-time history beyond the reach of individual-based simulations. 
+Since geographic structure from times longer ago than the mixing
+time of migration across the range has limited effect on modern genealogies
+\citep{wilkins2004separation} (other than possibly changing effective population
+size \citet{barton2002neutral,cox2002stepping}), this may not negatively affect realism.
 

### 2017-11-16 forwards_paper.tex
intro paragraphs

@@
 
-\plr{Quick review of simulation software, esp that other one that does the same thing.}
-
-\plr{Make clear that the main product here is the \msprime{} interface,}
-part of which is the tables schema.
+\plr{this paragraph is supposed to cite the fwds time simulation methods we need to}
+Forwards-time population genetics simulations are growing in importance,
+as the genome-wide impacts of hitchhiking and background selection are more widely appreciated.
+For instance, \citet{harris2016genetic} used SLiM \citep{haller2017flexible} to simulate tens of thousands of human exomes
+to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
+This is roughly state-of-the-art 
+-- the fastest simulation frameworks such as SLiM and fwdpp \citep{fwdpp}
+can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
+for tens of thousands of generations --
+but still substantially short of whole genomes for most biological population sizes.
+There has been some progress at parallelization \citep{lawrie2017accelerating},
+which has substantial promise.
+The general idea of storing genealogical information to speed up simulations is not new --
+it was entirely implemented in AnA-FiTS \citep{aberer2013rapid}, for instance --
+but this program did not implement the critical step of discarding irrelevant genealogical information.
+A similar but more limited method for discarding this information does appear in \citet{padhukasahasram2008exploring}.
+
+In this paper we discuss storage methods for genealogies (and hence, genome sequence),
+an algorithm for simplifying these,
+and their use in forwards-time simulation.
+Although we were motivated by a need for more efficient simulations,
+these tools may prove more widely useful.
 

### 2017-11-16 forwards_paper.tex
moved data format

@@
 \section*{Results}
-
-\plr{Order is intro-results-discussion-methods.}
 
@@
 %%%%%%%%%%%%%%
+\subsection*{Data structures:}
+\jk{Moved from above.}
+
+Here we describe in more detail the data structures underlying a tree sequence in \msprime.
+These derive from those described by \citet{kelleher2016efficient},
+but have been generalised and modified to remove redundancy.
+
+To clarify terminology,
+below a \emph{tree} refers to
+a genealogical tree describing how a collection of individuals are related to each other.
+A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
+tree relating all samples to each other at any point along the genome.
+In the context of a tree sequence, \emph{nodes} refer to distinct ancestors,
+and correspond to the vertices in the trees of a tree sequence.
+Since each node represents a certain ancestor, it has a unique ``time'', thought of as her birth time,
+which determines the height of any branching points she is associated with.  A given node will be
+associated with branching points of all trees across a region if that node
+is the most recent common ancestor to the subtending tips across that region.
+This information is stored in the columns of a \textbf{Node Table},
+having columns \emph{time}, \emph{population}, and \emph{flags}.
+The column ``flags'' records other information
+(e.g., a binary mask of `1' indicates the node is a sample).
+Importantly, the \textbf{node ID} of a node is given implicitly
+by the (zero-based) index of its corresponding row in the Node Table.
+
+Tree sequences are constructed by specifying over which segments of genome
+which nodes inherit from which other nodes.  This information is stored by
+recording the endpoints of each distinctly inherited ancestral segment,
+the parental node, and the child node who inherited that segment.
+As each such record describes particular edge across a swatch of trees in the tree sequence,
+we call these \emph{edges}
+and store them in the columns of an \textbf{Edge Table},
+having columns \emph{left}, \emph{right}, \emph{parent}, and \emph{child}.
+
+To record information about genetic variants we need to also record each mutation
+and which nodes have inherited that mutation.
+The tree structure takes care of inheritance -- all we need to do
+is to record the highest node in the tree at the mutated site
+that inherited that mutation.
+As more than one mtuation may occur at a given site,
+we separate this information into two tables,
+first, the \textbf{Site Table} records for each variant site its \emph{position} and \emph{ancestral state}.
+Here \emph{position} is a (floating point) position along the chromosome,
+and \emph{ancestral state} is the genotype of the root of the tree at that site.
+As for nodes, \textbf{site IDs} are given implicitly
+by the (zero-based) index of the rows.
+Then, we record in a \textbf{Mutation Table} the \emph{site}, \emph{node}, and \emph{derived state}
+for each mutation.
+Here ``site'' is the ID of the site at wihch this mutation occurred,
+``node'' is the ID of the highest node that has inherited this mutation,
+and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
+unless another mutation occurs.
+% \plr{Omit migrations?}
+
+\paragraph{Definition of valid tables}
+Here are the formal requirements for a set of nodes and edges to make sense,
+and to allow ``msprime``'s algorithms to work properly.
+
+To disallow time travel and multiple inheritance:
+\begin{enumerate}
+    \item Offspring must be born after their parents (and hence, no loops).
+    \item The set of intervals on which each individual is a child must be disjoint.
+\end{enumerate}
+For algorithmic reasons, we also require:
+\begin{enumerate} \setcounter{enumi}{2}
+    \item The leftmost endpoint of each chromosome is 0.0.
+    \item Node times must be strictly greater than zero.
+    \item Edges must be sorted in nondecreasing time order.
+\end{enumerate}
+Note that since each node time is equal to the amount of time since the \emph{birth} of the
+corresponding parent, time is measured in clock time, not in meioses.
+
+A forwards-time simulation does \textbf{not} naturally emit genealogical information
+satisfying all these requirements.
+However, \msprime{} implements two algorithms that will take a set of tables
+satisfying only 1--4 and produce tables satisfying all requirements.
+These are \texttt{sort\_tables}, to enforce sortedness requirements (without renumbering nodes),
+and \texttt{simplify}, which works on sorted tables
+and is guarenteed to produce a valid, \msprime{}-ready tree sequence.
+
+%%%%%%%%%%%%%%
 \subsection*{Simplification}
@@
 
-%%%%%%%%%%%%%%
-\subsection*{Data structures:}
-\jk{Moved from above.}
-
-First we describe the data structures we use for recording genealogical history,
-as implemented in \msprime.
-These derive from those described by \citet{kelleher2016efficient},
-but have been generalised and modified to remove redundancy.
-The tables below give the example tree sequence of Figure \ref{fig:ex_tree_seq}.
-
-\begin{figure}
-    \begin{center}
-        \includegraphics[width=0.6\textwidth]{example_tree_seq}
-    \end{center}
-    \caption{
-        A pictorial representation of a tree sequence
-        relating three samples to each other
-        over a chromosome of length 1.0.
-        Mutations shown in the example tables of the text
-        are marked with red stars.
-        \label{fig:ex_tree_seq}
-    }
-\end{figure}
-
-To clarify terminology,
-below a \emph{tree} refers to
-a genealogical tree describing how a collection of individuals are related to each other.
-A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
-tree relating all samples to each other at any point along the genome.
-In the context of a tree sequence, \emph{nodes} refer to distinct ancestors,
-and correspond to the vertices in the trees of a tree sequence.
-Since each node represents a certain ancestor, it has a unique ``time'', thought of as her birth time,
-which determines the height of any branching points she is associated with.  A given node will be
-associated with branching points of all trees across a region if that node
-is the most recent common ancestor to the subtending tips across that region.
-This information is stored in the columns of a \textbf{Node Table}:
-\begin{center}
-\begin{tabular}{cccc}
-    id   &  is\_sample &  population &  time \\
-    0    &  1          & 0           & 0 \\
-    1    &  1          & 0           & 0 \\
-    2    &  1          & 0           & 0 \\
-    3    &  0          & 0           & 0.3 \\
-    4    &  0          & 0           & 0.5 \\
-    5    &  0          & 0           & 0.6 \\
-    6    &  0          & 0           & 0.7 \\
-\end{tabular}
-\end{center}
-where ``flags'' records other information
-(e.g., a binary mask of `1' indicates the node is a sample).
-Importantly, the \textbf{node ID} of a node is given implicitly
-by the (zero-based) index of its corresponding row in the Node Table.
-
-Tree sequences are constructed by specifying over which segments of genome
-which nodes inherit from which other nodes.  This information is stored by
-recording the endpoints of each distinctly inherited ancestral segment,
-the parental node, and a list of children nodes who have inherited that segment.
-As each such record describes a collection of edges across a swatch of trees in the tree sequence,
-we call these \emph{edgesets}
-and store them in the columns of an \textbf{Edgeset Table:}
-\begin{center}
-\begin{tabular}{cccc}
-    left & right & parent & children \\
-    0.2  &  0.8  &  3     & 0,2 \\
-    0.0  &  0.2  &  4     & 1,2 \\
-    0.2  &  0.8  &  4     & 1,3 \\
-    0.8  &  1.0  &  4     & 1,2 \\
-    0.8  &  1.0  &  5     & 0,4 \\
-    0.0  &  0.2  &  6     & 0,4  \\
-\end{tabular}
-\end{center}
-
-To record information about genetic variants we need to also record each mutation
-and which nodes have inherited that mutation.
-The tree structure takes care of inheritance -- all we need to do
-is to record the highest node in the tree at the mutated site
-that inherited that mutation.
-As more than one mtuation may occur at a given site,
-we separate this information into two tables,
-first, the \textbf{Site Table} records for each variant site
-\begin{center}
-\begin{tabular}{ccc}
-    id  &   position  & ancestral\_state \\
-    0   &  0.1        & 0 \\
-    1   &  0.5        & 0 \\
-\end{tabular}
-\end{center}
-Here ``position'' is a (floating point) position along the chromosome,
-and ``ancestral state'' is the genotype of the root of the tree at that site.
-As for nodes, \textbf{site IDs} are given implicitly
-by the (zero-based) index of the rows.
-Then, we record in a \textbf{Mutation Table}
-\begin{center}
-\begin{tabular}{ccc}
-    site  & node  & derived\_state \\
-    0	  & 4	  & 1 \\
-    1	  & 3	  & 1 \\
-    1	  & 2	  & 0 \\
-\end{tabular}
-\end{center}
-in which ``site'' is the ID of the site at wihch this mutation occurred,
-``node'' is the ID of the highest node that has inherited this mutation,
-and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
-unless another mutation occurs.
-% \plr{Omit migrations?}
-
-\paragraph{Definition of valid tables}
-Here are the formal requirements for a set of nodes and edgesets to make sense,
-and to allow ``msprime``'s algorithms to work properly.
-
-To disallow time travel and multiple inheritance:
-\begin{enumerate}
-    \item Offspring must be born after their parents (and hence, no loops).
-    \item The set of intervals on which each individual is a child must be disjoint.
-\end{enumerate}
-For algorithmic reasons, we also require:
-\begin{enumerate} \setcounter{enumi}{2}
-    \item The leftmost endpoint of each chromosome is 0.0.
-    \item Node times must be strictly greater than zero.
-    \item The list of offspring in an edgeset must be sorted.
-    \item Edgesets must be sorted in nondecreasing time order.
-    \item The set of intervals on which each individual is a parent must be disjoint.
-    \item Each edgeset must contain at least two children.
-\end{enumerate}
-Note that since each node time is equal to the amount of time since the \emph{birth} of the
-corresponding parent, time is measured in clock time, not in meioses.
-
-A forwards-time simulation does \textbf{not} naturally emit genealogical information
-satisfying requirements 5--8.
-However, \msprime{} implements two algorithms that will take a set of tables
-satisfying only 1--4 and produce tables satisfying all requirements.
-Trivially, \texttt{sort\_tables} enforces requirements 5 and 6 and does not renumber nodes;
-then, \texttt{simplify} enforces requirements 7 and 8 (and does much more; see below).
-
-\section{More general method for recording the pedigree}
-
-\plr{moved from above}
-
-Concretely, this is done as follows.
-Suppose that the forwards simulation algorithm labels (haploid) individuals by integers,
-which we call ``input labels'',
-to distinguish them from the ``node IDs'' given to these same individuals in the (output) tree sequence.
-The algorithm maintains at all times a set of tables $(\nodes, \edgesets, \sites, \mutations)$
-that record a relaxed tree sequence,
-and an associative array $L$ that maps input labels to output node IDs, so that if $x$ is an input label,
-then $L[x]$ is the corresponding output node ID.
-We also always maintain $n$ to be the number of rows currently in the node table,
-(so that with zero-indexed IDs, the next to be added will have node ID $n$),
-$T_0$ to be the time of last simplification,
-and $n_0$ the number of rows in the node table at that time.
-
-Initially, we begin with $n$ and $n_0$ equal to the number of rows in the initial node table,
-and $L[j] = i_j$ for each $0 \le j < N$ if the initial input generation is labeled
-$i_0$, \ldots, $i_{N-1}$.
-
-At a reproduction event where haploid parents $x$ and $y$ produce offspring $u$
-at time $t$ in population $p$,
-\begin{enumerate}
-    \item add a
-        $(
-        \text{flags}=0,
-        \text{population}=p,
-        \text{time}=t)$ row to the node table,
-    \item set $L[u] = n$,
-    \item and increment $n\mathrel{+}=1$.
-\end{enumerate}
-Then, for each interval $[\ell,r)$ that $u$ inherits from parent $z$
-(where $z$ is either $x$ or $y$),
-\begin{enumerate}[resume]
-    \item add a
-        $( \text{left}=\ell,
-        \text{right}=r,
-        \text{parent}=z,
-        \text{children}=(u,))$ row to the edgeset table.
-\end{enumerate}
-If furthermore there have been mutations at genomic locations $s_1$, \ldots, $s_k$
-on this interval,
-with derived states $q_1$, \ldots, $q_k$,
-then for each $1 \le j \le k$,
-\begin{enumerate}[resume]
-    \item if $s_j$ is not in the site table, add a row
-        $( \text{position}=s_j,
-        \text{ancestral\_state}=0)$,
-    \item find the site index $i$ whose position is $s_j$,
-    \item and add a row
-        $( \text{site}=i$,
-           \text{node}=u$,
-        \text{derived\_state}=q_j)$ to the mutation table.
-\end{enumerate}
-
-To \textbf{simplify} the tree sequence at time $T$,
-\begin{enumerate}
-    \item add $T-T_0$ to each time in the first $n_0$ rows of the node table,
-        and replace each remaining time $t$ with $T-t$.
-\end{enumerate}
-Then, pass the set of currently alive input IDs,
-$i_0, \ldots, i_{N-1}$ to the simplificaiton algorithm,
-which produces a tree sequence that has node ID $j$ corresponding to input ID $i_j$
-for $0 \le j < N$, and
-\begin{enumerate}[resume]
-    \item empty $L$,
-    \item let $L[j] = i_j$ for $0 \le j < N$,
-    \item set $n$ to be the number of nodes in the new tree sequence and set $n_0 = n$, and finally
-    \item set $T_0 = T$.
-\end{enumerate}
-
-Simplification keeps the tables to a managable size.
-Since the map $L$ is updated to maintain the association between individuals in the simulation
-and nodes in the tree sequence, simplification can be run regularly,
-as the simulation progresses.
-
 %%%%%%%%%%%%%%%%%%%%%%
 \section*{Acknowledgements}
+Thanks to Brad Shaffer and Evan McCartney-Melstad for useful discussions.
+Funding for this project cam from XXX
 

### 2017-11-16 sims/space-complexity.py
added plot for total branch length

@@ -30,7 +30,7 @@ def wright_fisher(N, T, simplify_interval=1):
         nodes.add_row(time=T, flags=1)
     t = T
     S = np.zeros(T, dtype=int)
-    L = np.zeros(T, dtype=float)
+    Q = np.zeros(T, dtype=float)
     while t > 0:
         t -= 1
         Pp = [P[j] for j in range(N)]
@@ -49,16 +49,16 @@ def wright_fisher(N, T, simplify_interval=1):
             msprime.simplify_tables(Pp, nodes, edges)
             P = list(range(N))
         S[T - t - 1] = len(edges)
-        L[T - t - 1] = total_length(nodes, edges)
+        Q[T - t - 1] = total_length(nodes, edges)
     # We will always simplify at t = 0, so no need for special case at the end
-    return msprime.load_tables(nodes=nodes, edges=edges), S, L
+    return msprime.load_tables(nodes=nodes, edges=edges), S, Q
 
 def total_length(nodes, edges):
     ts = msprime.load_tables(nodes=nodes, edges=edges)
-    L = 0.0
+    Q = 0.0
     for t in ts.trees():
-        L += t.length * t.total_branch_length
-    return L
+        Q += t.length * t.total_branch_length
+    return Q
 
 def verify():
     """
@@ -77,29 +77,29 @@ def verify():
                 assert ts1.tables.edges == ts2.tables.edges
 
 def plot():
-    num_reps = 10
-    for n in [10, 100, 1000]:
+    num_reps = 5
+    for n in [10, 40, 100]:
         T = 10 * n
         A = np.zeros((num_reps, T))
         B = np.zeros((num_reps, T))
         for j in range(num_reps):
-            _, S, L = wright_fisher(n, T)
+            _, S, Q = wright_fisher(n, T)
             A[j] = S
-            B[j] = L
+            B[j] = Q
             plt.subplot(211)
             plt.plot(S, alpha=0.5)
             plt.subplot(212)
-            plt.plot(L, alpha=0.5)
+            plt.plot(Q, alpha=0.5)
             print(n, j, "done")
         mean_S = np.mean(A, axis=0)
-        mean_L = np.mean(B, axis=0)
+        mean_Q = np.mean(B, axis=0)
         plt.subplot(211)
         plt.plot(mean_S, color="black", lw=3)
         plt.title("N = {}".format(n))
         plt.xlabel("Generations")
         plt.ylabel("Number of edges")
[…]

### 2017-11-16 sims/space-complexity.py
added theoretical expectation to plots

@@ -30,7 +30,6 @@ def wright_fisher(N, T, simplify_interval=1):
         nodes.add_row(time=T, flags=1)
     t = T
     S = np.zeros(T, dtype=int)
-    Q = np.zeros(T, dtype=float)
     while t > 0:
         t -= 1
         Pp = [P[j] for j in range(N)]
@@ -49,9 +48,8 @@ def wright_fisher(N, T, simplify_interval=1):
             msprime.simplify_tables(Pp, nodes, edges)
             P = list(range(N))
         S[T - t - 1] = len(edges)
-        Q[T - t - 1] = total_length(nodes, edges)
     # We will always simplify at t = 0, so no need for special case at the end
-    return msprime.load_tables(nodes=nodes, edges=edges), S, Q
+    return msprime.load_tables(nodes=nodes, edges=edges), S
 
 def total_length(nodes, edges):
     ts = msprime.load_tables(nodes=nodes, edges=edges)
@@ -59,6 +57,10 @@ def total_length(nodes, edges):
     for t in ts.trees():
         Q += t.length * t.total_branch_length
     return Q
+
+def ub(T, N):
+    return 2 * (N - 1) + 6 * N * np.log(N/(1+2*N/(T+2)))
+
 
 def verify():
     """
@@ -78,31 +80,22 @@ def verify():
 
 def plot():
     num_reps = 5
-    for n in [10, 40, 100]:
+    for n in [10, 20, 30, 40, 50]:
         T = 10 * n
         A = np.zeros((num_reps, T))
         B = np.zeros((num_reps, T))
+        x = [ub(t,n) for t in range(1,T)]
+        plt.plot(x, ls="dotted", lw=3)
         for j in range(num_reps):
-            _, S, Q = wright_fisher(n, T)
+            _, S = wright_fisher(n, T)
             A[j] = S
-            B[j] = Q
-            plt.subplot(211)
             plt.plot(S, alpha=0.5)
-            plt.subplot(212)
-            plt.plot(Q, alpha=0.5)
             print(n, j, "done")
         mean_S = np.mean(A, axis=0)
-        mean_Q = np.mean(B, axis=0)
-        plt.subplot(211)
         plt.plot(mean_S, color="black", lw=3)
         plt.title("N = {}".format(n))
         plt.xlabel("Generations")
         plt.ylabel("Number of edges")
-        plt.subplot(212)
-        plt.plot(mean_Q, color="black", lw=3)
[…]

### 2017-11-16 sims/wf_expectations.R
update to wf tree length plot

@@ -1,6 +1,6 @@
 library(matrixcalc)
 
-N <- 100
+N <- 60
 
 # get the WF transition matrix
 # stirling number of the second kind:
@@ -30,6 +30,22 @@ T0 <- rowSums(sweep(V, 2, 2:N, "*"))
 #  T = (2:N) + P %*% T
 T1 <- forwardsolve(diag(N-1) - P[-1,-1], 2:N)
 
-plot(T1, type='l', ylim=range(0,T1,finite=TRUE))
-lines(T0, col='green')
-lines(2:N, 2 * N * cumsum(1/(2:N-1)), col='red')
+# coal expectation
+Tcoal <- 2 * N * cumsum(1/(2:N-1))
+
+layout(1:2)
+
+plot(2:N, T1, type='l', ylim=range(0,T1,finite=TRUE),
+     xlab='number of initial leaves',
+     ylab='expected total tree length',
+     main=paste("N =", N))
+lines(2:N, T0, col='green')
+lines(2:N, Tcoal, col='red')
+legend("bottomright", lty=c(1, 1, 1), col=c('black', 'green', 'red'),
+       legend=c('WF', 'WF', 'coalescent'))
+
+plot(2:N, T1/Tcoal, type='l',
+     xlab='number of initial leaves',
+     ylab='ratio of total tree length, WF to coal',
+     main=paste("N =", N))
+

### 2017-11-16 sims/space-complexity.py
Added quick hack to compute E[number of edges] per tree transition.

@@ -58,8 +58,15 @@ def total_length(nodes, edges):
         Q += t.length * t.total_branch_length
     return Q
 
-def ub(T, N):
-    return 2 * (N - 1) + 6 * N * np.log(N/(1+2*N/(T+2)))
+def ub(T, N, num_edges=3):
+    return 2 * (N - 1) + 2 * num_edges * N * np.log(N/(1+2*N/(T+2)))
+
+def get_mean_edges_per_transition(ts):
+    d = np.zeros(ts.num_trees - 1)
+    for j, (_, edges_out, edges_in) in enumerate(ts.edge_diffs()):
+        if len(edges_out) > 0:
+            d[j - 1] = len(edges_in)
+    return np.mean(d)
 
 
 def verify():
@@ -79,18 +86,25 @@ def verify():
                 assert ts1.tables.edges == ts2.tables.edges
 
 def plot():
-    num_reps = 5
-    for n in [10, 20, 30, 40, 50]:
+    num_reps = 20
+    # for n in [10, 20, 30, 40, 50]:
+    for n in [100, 200]:
         T = 10 * n
         A = np.zeros((num_reps, T))
-        B = np.zeros((num_reps, T))
-        x = [ub(t,n) for t in range(1,T)]
-        plt.plot(x, ls="dotted", lw=3)
+        E = np.zeros((num_reps, T))
         for j in range(num_reps):
-            _, S = wright_fisher(n, T)
+            ts, S = wright_fisher(n, T)
             A[j] = S
+            E[j] = get_mean_edges_per_transition(ts)
             plt.plot(S, alpha=0.5)
             print(n, j, "done")
+
+        x = [ub(t,n) for t in range(1,T)]
+        plt.plot(x, ls="dotted", color="blue", lw=3)
+
+        x = [ub(t,n, np.mean(E)) for t in range(1,T)]
+        plt.plot(x, ls="dashed", color="red", lw=3)
+
         mean_S = np.mean(A, axis=0)
         plt.plot(mean_S, color="black", lw=3)
         plt.title("N = {}".format(n))
@@ -104,4 +118,5 @@ def plot():
 if __name__ == "__main__":
     # verify()
     plot()
+
 

### 2017-11-16 forwards_paper.tex
added factor of three

@@
 of only the current generation of $2N$ individuals.
-This tree sequence has $4N-2$ edges to describe the leftmost tree,
-and one additional edge each time the marginal tree changes along the sequence.
+This tree sequence has $4N-2$ edges to describe the leftmost tree;
+and each time the marginal tree changes along the sequence,
+three edges end and three new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
 Coalescent theory tells us that 
@@
 Not all such recombinations actually change the tree topology,
-and so this gives us an upper bound on the number of edges.
+and three times this gives us an upper bound on the expected number of edges.
 Similarly, not every new edge derives from a never-before-seen node,
 but the number of nodes is at most equal to the number of edges plus the sample size.
-With $T=20N$, this reduces the initial storage of $160 N^2$ items to $2N(4\log(2N) + 3)$;
-at $N=10^4$, a factor of 20,000.
+With $T=20N$, this reduces the initial storage of $160 N^2$ items to $6N(4\log(2N) + 3)$;
+at $N=10^4$, a factor of 3,000.
 

### 2017-11-16 forwards_paper.tex
moved API to methods

@@
 
-%%%%%%
-\subsection*{Overview of the API}
-
-\plr{leaving this alone for the moment: to stay here? move?}
-
-The \msprime\ Python API provides a powerful platform for working with
-tree topology and mutation data. We refer to the part of \msprime\
-that is dedicated to tree sequence input and output as the `Tables API',
-as the API is organised around simple tables of data. There are four key
-tables: nodes, edges, sites and mutations. Briefly, nodes and edges define
-the topology of a tree sequence (as defined above) and the sites and mutations
-define mutational processes on this topology.
-Figure~\ref{fig:example_tree_sequence} gives an informal depiction of this
-encoding in terms of these tables.
-
-The tables API is primarily designed to facilitate efficient interchange of
-data between programs or between different modules of the same program.
-Following the current best-practises [citations: apache arrow, etc] data is stored
-in a columnar format. There are many advantages to this approach, but the
-principle advantage for our purposes is that it allows for very efficient
-interchange of large amounts of numerical data. In principle, this enables
-zero-copy semantics, where a data consumer can read the information directly
-from the memory of a producer without incurring the overhead of a copy
-[citation?] Our implementation uses the numpy C API [citation] to efficiently copy
-data from Python into the low-level C library used to manipulate
-tree sequences.
-
-The tables API provides basic input and output operations via the numpy
-array interface, which provides a great deal of flexibility as well
-as efficiently. This makes transferring data from various sources
-such as HDF5 [citation], Dask, Zarr etc, straightforward. (For small
-scale data and debugging purposes, a simple text format is also supported.)
-Along with
-these operations we also provide a function to sort a set of tables
-to ensure that the records are in the form required to input
-into an \msprime\ tree sequence object. The \texttt{simplify\_tables}
-function implements tree sequence simplification, as described in the
-previous subsection.
-
-This interchange API is very efficient. [Describe a quick example where we generate
-a many-gigabyte tree sequence using fwdpp, and the time required
-to copy the node and edge data into the tables]. By using a simple numerical
-encoding of tree topologies and contiguous arrays of data to store this
-encoding, we can achieve data transfer rates that would be impossible under
-any text-based approach while retaining excellent portability.
-
+%%%%%%%%%%%%
 \subsection*{Estimates of run-time complexity}
@@
 
+%%%%%%
+\subsection*{Overview of the API}
+
+The \msprime{} Python API provides a powerful platform for working with tree topology and mutation data. 
+The new portions of \msprime{} that we cover here
+are dedicated to tree sequence input and output using simple tables of data,
+so we refer to this as the `Tables API'.
+The four key tables: nodes, edges, sites and mutations, are described above.
+
+The tables API is primarily designed to facilitate efficient interchange of
+data between programs or between different modules of the same program.
+Following the current best-practises [citations: apache arrow, etc] data is stored
+in a columnar format.
+The principal advantage of this for our purposes is that it allows for very efficient
+interchange of large amounts of numerical data. In principle, this enables
+zero-copy semantics, where a data consumer can read the information directly
+from the memory of a producer without incurring the overhead of a copy
+[citation?] Our implementation uses the numpy C API [citation] to efficiently copy
+data from Python into the low-level C library used to manipulate
+tree sequences.
+\plr{@molpopgen: note on how this works with fwdpp?}
+
+The tables API provides basic input and output operations via the numpy
+array interface, which provides a great deal of flexibility as well
+as efficiency, and makes it straightforward to transfer data from sources
+such as HDF5 [citation], Dask, or Zarr. 
+(For small scale data and debugging purposes, a simple text format is also supported.)
+Along with these operations we also provide a function to sort a set of tables
+to ensure that the records are in the form required to input
+into an \msprime{} tree sequence object. (Simplification may also be required.)
+
+\plr{Maybe we don't need this paragraph? Already said it, more or less, above, esp if we insert something about how fwdpp works.}
+This interchange API is very efficient. [Describe a quick example where we generate
+a many-gigabyte tree sequence using fwdpp, and the time required
+to copy the node and edge data into the tables]. By using a simple numerical
+encoding of tree topologies and contiguous arrays of data to store this
+encoding, we can achieve data transfer rates that would be impossible under
+any text-based approach while retaining excellent portability.
+
 %%%%%%%%%%%%%%
 \subsection*{Data structures:}
-\jk{Moved from above.}
 

### 2017-11-16 sims/msprime-examples.py
added numpy method for calculating pi to TODO

@@ -9,6 +9,21 @@ import numpy as np
 
 import msprime
 
+# TODO:
+if False:
+    # compare speed of pi calc to this:
+
+    ts = msprime.simulate(100)
+    G = ts.genotype_matrix()
+
+    def np_pi(ts, G):
+      n = ts.num_samples
+      m = np.sum(G, 1)
+      return sum(2 * m * (n - m) / (n * (n - 1))) / ts.sequence_length
+
+    np.pi(ts, G)
+
+
 
 def run_simulation(args):
     before = time.process_time()

### 2017-11-16 forwards_paper.tex
removed some of the tables requirements

@@
 %%%%%%%%%%%%%%
-\subsection*{Data structures:}
+\subsection*{The table format}
 
-Here we describe in more detail the data structures underlying a tree sequence in \msprime.
-These derive from those described by \citet{kelleher2016efficient},
-but have been generalised and modified to remove redundancy.
-
-To clarify terminology,
-below a \emph{tree} refers to
-a genealogical tree describing how a collection of individuals are related to each other.
-A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
-tree relating all samples to each other at any point along the genome.
-In the context of a tree sequence, \emph{nodes} refer to distinct ancestors,
-and correspond to the vertices in the trees of a tree sequence.
-Since each node represents a certain ancestor, it has a unique ``time'', thought of as her birth time,
-which determines the height of any branching points she is associated with.  A given node will be
-associated with branching points of all trees across a region if that node
-is the most recent common ancestor to the subtending tips across that region.
-This information is stored in the columns of a \textbf{Node Table},
-having columns \emph{time}, \emph{population}, and \emph{flags}.
-The column ``flags'' records other information
-(e.g., a binary mask of `1' indicates the node is a sample).
-Importantly, the \textbf{node ID} of a node is given implicitly
-by the (zero-based) index of its corresponding row in the Node Table.
-
-Tree sequences are constructed by specifying over which segments of genome
-which nodes inherit from which other nodes.  This information is stored by
-recording the endpoints of each distinctly inherited ancestral segment,
-the parental node, and the child node who inherited that segment.
-As each such record describes particular edge across a swatch of trees in the tree sequence,
-we call these \emph{edges}
-and store them in the columns of an \textbf{Edge Table},
-having columns \emph{left}, \emph{right}, \emph{parent}, and \emph{child}.
-
-To record information about genetic variants we need to also record each mutation
-and which nodes have inherited that mutation.
-The tree structure takes care of inheritance -- all we need to do
-is to record the highest node in the tree at the mutated site
-that inherited that mutation.
-As more than one mtuation may occur at a given site,
-we separate this information into two tables,
-first, the \textbf{Site Table} records for each variant site its \emph{position} and \emph{ancestral state}.
-Here \emph{position} is a (floating point) position along the chromosome,
-and \emph{ancestral state} is the genotype of the root of the tree at that site.
-As for nodes, \textbf{site IDs} are given implicitly
-by the (zero-based) index of the rows.
-Then, we record in a \textbf{Mutation Table} the \emph{site}, \emph{node}, and \emph{derived state}
-for each mutation.
-Here ``site'' is the ID of the site at wihch this mutation occurred,
-``node'' is the ID of the highest node that has inherited this mutation,
-and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
-unless another mutation occurs.
-% \plr{Omit migrations?}
-
-\paragraph{Definition of valid tables}
-Here are the formal requirements for a set of nodes and edges to make sense,
-and to allow ``msprime``'s algorithms to work properly.
-
-To disallow time travel and multiple inheritance:
+Given a set of node and edge tables as described above,
+there are a small set of requirements that ensure the tables
+describe a valid tree sequence.
+(There are essentially no such requirements on the site and mutation tables.)
+These are:
 \begin{enumerate}
@@
 \end{enumerate}
-For algorithmic reasons, we also require:
-\begin{enumerate} \setcounter{enumi}{2}
-    \item The leftmost endpoint of each chromosome is 0.0.
-    \item Node times must be strictly greater than zero.
-    \item Edges must be sorted in nondecreasing time order.
-\end{enumerate}
+A pair of node and edge tables that satisfy these two requirements
+is guarenteed to uniquely describe at each point on the genome
+a collection of directed, acyclic graphs -- in other words, a forest of trees.
+For some applications it is necessary to check that at every point
+there is only a \emph{single} tree.  
+Checking this is more difficult, but is implemented in \msprime{}'s API.
+
+For efficiency, \msprime{} makes several other sortedness requirements on the tables,
+that are not necessarily satisfied by tabled emitted by a forwards-time simulation.
+These can be rectified by first sorting (using \texttt{sort\_tables})
+and then using \texttt{simplify}, which works on sorted tables
+and is guarenteed to produce a valid, \msprime{}-ready tree sequence.
+
 Note that since each node time is equal to the amount of time since the \emph{birth} of the
 corresponding parent, time is measured in clock time, not in meioses.
-
-A forwards-time simulation does \textbf{not} naturally emit genealogical information
-satisfying all these requirements.
-However, \msprime{} implements two algorithms that will take a set of tables
-satisfying only 1--4 and produce tables satisfying all requirements.
-These are \texttt{sort\_tables}, to enforce sortedness requirements (without renumbering nodes),
-and \texttt{simplify}, which works on sorted tables
-and is guarenteed to produce a valid, \msprime{}-ready tree sequence.
 

### 2017-11-16 forwards_paper.tex
.

@@
 
-\plr{this paragraph is supposed to cite the fwds time simulation methods we need to}
+\plr{this paragraph is supposed to cite the fwds time simulation methods we need to - @molpopgen fix this up?}
 Forwards-time population genetics simulations are growing in importance,

### 2017-11-16 forwards_paper.tex
overview of how things work using fwdpp

--- 2017-11-16 forwards_paper.tex
@@
 
+\newcommand{\fwdpp}{\texttt{fwdpp}}
 \newcommand{\msprime}{\texttt{msprime}}
@@
 \usepackage{color}
+\newcommand{\krt}[1]{{\em \color{green} #1}}
 \newcommand{\plr}[1]{{\em \color{blue} #1}}
@@
 tree sequences.
+
+The \msprime{} tables API is independent of the details of a particular simulation engine.  In other words, one simply
+needs to fill the relevant columns and send them to \msprime{} at regular intervals for simplification.  For example,
+using the \fwdpp{} C++ API \cite{Thornton2014-hx}, we fill arrays of simple structures representing the node and edge data.
+Doing so requires no modification of the \fwdpp{} code base.  Rather, we simply need to book-keep the parent/offspring
+labels and perform simple processing of the recombination breakpoints from each mating event.  Using \texttt{pybind11}
+(\url{https://github.com/pybind/pybind11/}),
+the node/edge arrays on
+the C++ side are made visible, without copy, to Python as a multi-column numpy array. This array is in turn processed by
+the \msprime{} tables API.  In order to create a standalone user interface, our implementation uses \texttt{pybind11} to
+create a Python package extending \texttt{fwdpy11} (\url{https://github.com/molpopgen/fwdpy11}), which is a Python
+package based on \fwdpp{}.  (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
+API from a command-line C/C++ application via an embedded Python interpreter.)
 \plr{@molpopgen: note on how this works with fwdpp?}
+\krt{@plr: done, but perhaps too verbose?}
 

### 2017-11-16 sims/space-complexity.py
Changed to num_edges = 4 and tweaked plot.

@@ -58,8 +58,8 @@ def total_length(nodes, edges):
         Q += t.length * t.total_branch_length
     return Q
 
-def ub(T, N, num_edges=3):
-    return 2 * (N - 1) + 2 * num_edges * N * np.log(N/(1+2*N/(T+2)))
+def ub(T, N):
+    return 2 * (N - 1) + 8 * N * np.log(N/(1+2*N/(T+2)))
 
 def get_mean_edges_per_transition(ts):
     d = np.zeros(ts.num_trees - 1)
@@ -86,27 +86,22 @@ def verify():
                 assert ts1.tables.edges == ts2.tables.edges
 
 def plot():
-    num_reps = 20
+    num_reps = 50
     # for n in [10, 20, 30, 40, 50]:
-    for n in [100, 200]:
+    # for n in [100, 200]:
+    for n in [50]:
         T = 10 * n
         A = np.zeros((num_reps, T))
-        E = np.zeros((num_reps, T))
         for j in range(num_reps):
             ts, S = wright_fisher(n, T)
             A[j] = S
-            E[j] = get_mean_edges_per_transition(ts)
             plt.plot(S, alpha=0.5)
             print(n, j, "done")
 
         x = [ub(t,n) for t in range(1,T)]
-        plt.plot(x, ls="dotted", color="blue", lw=3)
-
-        x = [ub(t,n, np.mean(E)) for t in range(1,T)]
-        plt.plot(x, ls="dashed", color="red", lw=3)
-
+        plt.plot(x, ls="dashed", color="black", lw=3)
         mean_S = np.mean(A, axis=0)
-        plt.plot(mean_S, color="black", lw=3)
+        plt.plot(mean_S, lw=3)
         plt.title("N = {}".format(n))
         plt.xlabel("Generations")
         plt.ylabel("Number of edges")
@@ -118,5 +113,4 @@ def plot():
 if __name__ == "__main__":
     # verify()
     plot()
-
 

### 2017-11-16 forwards_paper.tex
added space complexity figure

@@
 but the number of nodes is at most equal to the number of edges plus the sample size.
-With $T=20N$, this reduces the initial storage of $160 N^2$ items to $6N(4\log(2N) + 3)$;
-at $N=10^4$, a factor of 3,000.
+With $T=20N$, this reduces the initial storage of $160 N^2$ items to $8N(4\log(2N) + 3)$;
+at $N=10^4$, a factor of 4,700.
+
+To get an idea of how required space depends on the length of time,
+suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
+begun with no prior history.
+This produces $4NT$ edges; how many are required after simplification?
+As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
+Roughly, the length of time for which a marginal tree has k tips is 
+$2N/k(k-1) = 2N(1/(k-1) - 1/k)$,
+and so the time to go from N to n tips is $2N(1/(n-1) - 1/(N-1))$.
+This implies that after running for time $T$ we expect to 
+have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
+The total tree length over this time is 
+$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which 
+% is approximately
+% \begin{align*}
+%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
+% \end{align*}
+leads to an upper bound on the number of edges of
+\begin{align*}
+    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
+\end{align*}
+This holds up quite well in practice --
+the number of edges actually required for 50 independent Wright--Fisher simulations,
+as a function of time (obtained by running simplify every generation)
+is shown in \ref{fig:num_edges}, compared to this prediction.
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{sims/num_edges_50}
+    \end{center}
+    \caption{
+        Number of edges in the simplified tree sequence
+        for 50 Wright-Fisher simulations begun with no prior history
+        as a function of number of generations.
+        Each line is one simulation, the black line gives the average,
+        and the dotted line is the upper bound of the text.
+        \label{fig:num_edges}
+    }
+\end{figure}
 

### 2017-11-17 sims/msprime-examples.py
Added benchmark for simplify wrt input sample size.

@@ -7,6 +7,13 @@ import os.path
 import io
 import numpy as np
 
+import matplotlib
+# Force matplotlib to not use any Xwindows backend.
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+import seaborn as sns
+import pandas as pd
+
 import msprime
 
 # TODO:
@@ -14,15 +21,6 @@ if False:
     # compare speed of pi calc to this:
 
     ts = msprime.simulate(100)
-    G = ts.genotype_matrix()
-
-    def np_pi(ts, G):
-      n = ts.num_samples
-      m = np.sum(G, 1)
-      return sum(2 * m * (n - m) / (n * (n - 1))) / ts.sequence_length
-
-    np.pi(ts, G)
-
 
 
 def run_simulation(args):
@@ -105,6 +103,7 @@ def run_stats(args):
     print("length     \t {} MB".format(ts.sequence_length / 10**6))
     print("num_sites\t", ts.num_sites)
     print("num_trees\t", ts.num_trees)
+    print("variant matrix size\t", ts.num_sites * ts.num_samples / (1024**3), "GiB")
     before = time.process_time()
     num_trees = 0
     for tree in ts.trees():
@@ -115,37 +114,93 @@ def run_stats(args):
     assert num_trees == ts.num_trees
 
 
+def run_simplify_benchmark(args):
+    ts = msprime.load(args.file)
+    np.random.seed(1)
+    print("Running simplify benchmarks")
+    N = 20
+    num_replicates = 5
+    subsample_size = np.logspace(1, 5, N).astype(int)
+    print(subsample_size)
+    T = np.zeros(N)
+    for j in range(N):
+        X = np.zeros(num_replicates)
+        for k in range(num_replicates):
+            sample = np.random.choice(
+                ts.num_samples, subsample_size[j], replace=False).astype(np.int32)
+            before = time.process_time()
+            sub_ts = ts.simplify(sample)
+            X[k] = time.process_time() - before
+        T[j] = np.mean(X)
+        print(subsample_size[j], T[j])
+
+    df = pd.DataFrame({"subsample_size": subsample_size, "time": T})
+    df.to_csv("data/simplify_subsample.dat")
+
+    plt.semilogx(subsample_size, T, marker="o")
+    plt.xlabel("Subsample size")
+    plt.ylabel("Time to simplify (s)")
+    plt.savefig("simplify_subsample_perf.pdf", format='pdf')
+
+
+def run_benchmark_pi(args):
[…]

### 2017-11-17 sims/msprime-examples.py
Added benchmark for pi computation time.

@@ -144,23 +144,33 @@ def run_simplify_benchmark(args):
 
 
 def run_benchmark_pi(args):
+    before = time.process_time()
     ts = msprime.simulate(
         sample_size=args.sample_size, length=args.length * 10**6, Ne=10**4,
         recombination_rate=1e-8, mutation_rate=1e-8, random_seed=10)
     duration = time.process_time() - before
     print("Ran simulation in {} seconds".format(duration))
+    print("sample size = ", ts.sample_size)
+    print("num_sites = ", ts.num_sites)
 
     G = ts.genotype_matrix()
+    print("Genotype matrix = ", G.nbytes / 1024**3, "GB")
 
     def np_pi(G):
         n = ts.num_samples
         m = np.sum(G, 1)
-        return sum(2 * m * (n - m) / (n * (n - 1))) / ts.sequence_length
+        return sum(2 * m * (n - m) / (n * (n - 1)))
 
+    before = time.process_time()
     pi1 = np_pi(G)
-    pi2 = ts.get_pairwise_diversity()
+    np_time = time.process_time() - before
+    print("np time = ", np_time)
 
-    print(pi1, pi2)
+    before = time.process_time()
+    pi2 = ts.get_pairwise_diversity()
+    msp_time = time.process_time() - before
+    print("msp time = ", msp_time)
+    # print("pi values = ", pi1, pi2)
 
 
 if __name__ == "__main__":
@@ -198,7 +208,7 @@ if __name__ == "__main__":
     subparser.set_defaults(func=run_simplify_benchmark)
 
     subparser = subparsers.add_parser('benchmark-pi')
-    subparser.add_argument("--sample-size", "-s", type=int, default=10**4)
+    subparser.add_argument("--sample-size", "-s", type=int, default=100**4)
     subparser.add_argument("--length", "-l", type=int, default=100)
     subparser.set_defaults(func=run_benchmark_pi)
 

### 2017-11-17 forwards_paper.tex
Add most overview comments

@@
 \include{macros}
-
+\newcommand{\simupop}{\texttt{simuPOP}}
 \newcommand{\fwdpp}{\texttt{fwdpp}}
@@
 \newcommand{\jk}[1]{{\em \color{red} #1}}
+\newcommand{\jda}[1]{{\em \color{cyan} #1}}
 
@@
     and recording the entire history of genetic inheritance in an efficient representation of the ancestral recombination graph,
-    on which neutral mutations can be quickly placed afterwards --
+    on which neutral mutations can be quickly placed afterwards ---
     the \emph{tree sequence} introduced in the software package \msprime.
@@
 For this reason, if one records the entire genealogical history of a population over the course of a simulation,
-one can lay down neutral mutations on top of that history afterwards,
-without loss of generality.
+one can lay down neutral mutations on top of that history afterwards.
+\jda{WLOG seems unclear/ not needed}
 Precisely, we would need to know the genealogical tree relating all sampled individuals
 at each position along the genome.
-In this paper, we show how to use algorithmic tools and data structures developed for the 
+In this paper, describe algorithmic tools and data structures developed for the 
 coalescent simulator \msprime{}
@@
 Another way of representing this same information
-is known as the \emph{ancestral recombination graph}, or {ARG} \citep{griffiths1997ancestral},
+is known as the \emph{ancestral recombination graph} ---
+the {ARG} \citep{griffiths1997ancestral},
+where nodes are common ancestor or recombination events ---
 which has been the subject of substantial study
 under the assumptions of coalescent theory \citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
+\jda{would be nice to say, concisely the difference}
 
+\jda{Is this the right place for this paragraph? should it maybe be integrated
+	with graf beginning 'Modern computing' above?}
 \plr{this paragraph is supposed to cite the fwds time simulation methods we need to - @molpopgen fix this up?}
@@
 which has substantial promise.
+\jda{this is vague, promise for what?}
 The general idea of storing genealogical information to speed up simulations is not new --
@@
 Below, we first describe the conceptual and algorithmic foundations for the method:
-(a) the new \emph{tables} format for storing tree sequences;
-(b) an algorithm to output tree sequence information on the fly
+(a) a format, implemented in the \msprime{} Python API,
+for storing tree sequences in several \emph{tables};
+(b) an algorithm, \algref{W}, to record tree sequence information into these tables on the fly
     from a forwards-time simulation;
 and (c) a method to \emph{simplify} a tree sequence, i.e., remove redundant information.
-This is followed by benchmarking comparisons
-with real, forwards-time simulations.
+We then describe a method that combines (b) and (d) to allows efficiently record simplified a tree sequence,
+and present analyses of its run time complexity.
+This is followed by benchmarking comparisons between two implementations of our method,
+each using previously published forwards-time simulation libraries.
 
 %%%%%%
-\subsection*{Recording a tree sequence}
+\subsection*{Storing a tree sequence}
 
@@
 For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
-with overlapping generations that records genealogical history in this way.
+(but with overlapping generations) that records genealogical history in this way.
 The simulation will run for $T$ generations,
@@
 The probability of death per individual each generation is $\delta$.
+\jda{In the below we are using the integer vector encoding of a tree as in
+	\msprime{} should say so and reference it as the role of node ids is
+	otherwise more confusing than it need be}
 We will use
@@
 \begin{taocpalg}{W}{Forwards-time tree sequence}
-{Simulates a randomly mating population of $N$ haploid individuals,
+{Simulates a randomly mating population $P$ of $N$ haploid individuals,
 with a probability of death per generation $\delta$ and a chromosome of length
@@
 
-\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
-    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
+\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, n - 1\})$,
+    $b \leftarrow \randomuniform(\{0, \dots, n - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
 }
 
-\algstep{W7.}{Record edges}{Set $E \leftarrow E \cup \left\{ (0, x, P_a, n), (x, L, P_b, n) \right\}$
-    and $n \leftarrow n + 1$.
+\algstep{W7.}{Record edges}{Set $E \leftarrow E \cup \left\{ (0, x, P_a, n), (x,
+		L, P_b, n) \right\}$.
+}
+
+\algstep{W8.}{Increment the current node.}{
+$n \leftarrow n + 1$.
 }
@@
     \caption{
-        A simple example of the method.
+        A simple example of the simplify method.
         \textbf{Top:} the diagram on the left relates ten haploid individuals to each other.
@@
 
-The tree sequences produced by forwards simulations, 
+The tree sequences produced by forwards simulations 
 record all of history for everyone alive at any time through the simulation.
@@
 The process is depicted in Figure~\ref{fig:simplify_state}.
+\jda{Need refs to Figure~\ref{fig:method_diagram} in the text}
 
@@
 the input tree sequence.
-Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence..
+Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.
 
-\paragraph{Sequential simplification}
-At any point the simulation scheme above, genealogical history is recorded
-in a tree sequence. This has two additional advantages. First, simplification
+\subsection*{Forward simulation and sequential simplification to record simplifed tree sequences}
+
+At any point in Algorithm~\algref{W}, genealogical history is recorded
+in an unsimplified tree sequence. This has two additional advantages. First, simplification
 can be run periodically through the simulation, taking the set of samples to be
@@
 instance, by a coalescent simulation with \msprime. 
+This process of forward simulation coupled with sequential simplification is our
+method
+
+\jda{I didn't write down the full thing, but mightn't things be clearer if we
+	have a statement of the combined method?}
+\begin{taocpalg}{F}{Forwards-time simplifed tree sequence}{
+Simulates a population starting with $N$ diploid individuals on a chromosome of length $L$ for $T$
+generations.
+}
+
+\algstep{F0.}{Initialize nodes.}{
+Generate initial haploid nodes $\mathcal N$:
+1.) using coalsecent simulation with \msprime{} for $2N$ initial haploid samples $S$ on
+a chromosome of length $L$ and taking $\mathcal N = \{\texttt{id}(u) : u \in S \}$; or
+2.) simply defining $\mathcal N = \{1\dots 2N\}$ as the node ids.
+Define a mapping from  $\mathcal M: \{1\dots 2N\} \to \mathcal N$. In the second
+case $\mathcal M$ is the identity.
+}
+
+\algstep{F1.}{Initialize edges.}{
+Define the initial edges $E_0$: 1) either from the coalescent simulation's final
+state, or 2) as \dots
+}
+
+\algstep{F2.}{Record trees till time $T - \delta$.}{ 
+Simulate, using Algorithm~\algref{W} or another algorithm, $2N$ haploid
+individuals using $\mathcal M$ to renumber the nodes, producing nodes $n$ and
+edgesets $E$.
+}
+
+\algstep{F3.}{Simplification over $\delta$.}{
+Select final samples $S'$ and using the simplification Algorithm~\algref{S},
+simplify $n$ and $E$ to produce a simplified tree sequence.
+}
+
+\algstep{F4.}{Continue?}{
+	Stop if $T - \delta  = 0$, else continue.
+}
+\end{taocpalg}
 
 %%%%%%%%%%%%
-\subsection*{Estimates of run-time complexity}
+\subsubsection*{Estimates of run-time complexity}
 
+\jda{I'd like to outline what we're doing here with reference to alg F}
 Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals 
@@
 
-Comparison of simulation with/without msprime, using simuPOP
+Comparison of simulation with/without msprime, using \simupop{}
 or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
@@
 from ``nid de grue'',
-as the nest of a crange is a large jumble of (forking) branches.
+as the nest of a crane is a large jumble of (forking) branches.
 We thought it unwise to use this term throughout the manuscript,

### 2017-11-17 forwards_paper.tex
reorg idea for API overview

@@
 tree sequences.
+Thus, by using a simple numerical
+encoding of tree topologies and contiguous arrays of data to store this
+encoding, we can achieve data transfer rates that would be impossible under
+any text-based approach while retaining excellent portability.
 
@@
 \plr{Maybe we don't need this paragraph? Already said it, more or less, above, esp if we insert something about how fwdpp works.}
+\jda{agree re the next two lines. propose moving the other above}
 This interchange API is very efficient. [Describe a quick example where we generate
 a many-gigabyte tree sequence using fwdpp, and the time required
-to copy the node and edge data into the tables]. By using a simple numerical
-encoding of tree topologies and contiguous arrays of data to store this
-encoding, we can achieve data transfer rates that would be impossible under
-any text-based approach while retaining excellent portability.
+to copy the node and edge data into the tables].
 

### 2017-11-17 forwards_paper.tex
Comment on 'biological population sizes'

@@
 but still substantially short of whole genomes for most biological population sizes.
+\jda{I'd find this more compelling with something added like: "e.g., Xs of
+	gigabases for Ys of individuals, implying 10Ys of generations needed to
+	equilibrate}
 There has been some progress at parallelization \citep{lawrie2017accelerating},
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%
 \section*{Results}

### 2017-11-18 forwards_paper.tex
Add formatting comment

@@
 allows for efficient algorithms that take advantage
-of the highly correlated nature of nearby trees.
+of the highly correlated nature of nearby trees~\citep{kelleher2016efficient}.
 
@@
 
-The tables API is primarily designed to facilitate efficient interchange of
+The Tables API is primarily designed to facilitate efficient interchange of
 data between programs or between different modules of the same program.
+\jda{should capitalize Tables, or not, as above}
 Following the current best-practises [citations: apache arrow, etc] data is stored

### 2017-11-18 forwards_paper.tex
Comments on tables section

@@
 Checking this is more difficult, but is implemented in \msprime{}'s API.
+\jda{TODO add function name}
 
 For efficiency, \msprime{} makes several other sortedness requirements on the tables,
-that are not necessarily satisfied by tabled emitted by a forwards-time simulation.
-These can be rectified by first sorting (using \texttt{sort\_tables})
+that are not necessarily satisfied by tables emitted by a forwards-time simulation.
+Fortunately, \msprime{}'s API includes tools to rectify this by first sorting (using \texttt{sort\_tables})
 and then using \texttt{simplify}, which works on sorted tables
@@
 corresponding parent, time is measured in clock time, not in meioses.
+\jda{is this the right place? I'd put it in discussion of Figure~\ref{fig:example_tree_sequence}}
 

### 2017-11-18 forwards_paper.tex
edited edits

@@
 For this reason, if one records the entire genealogical history of a population over the course of a simulation,
-one can lay down neutral mutations on top of that history afterwards.
-\jda{WLOG seems unclear/ not needed}
+simply laying down neutral mutations on top of that history afterwards
+is equivalent to having generated them during the simulation.
 Precisely, we would need to know the genealogical tree relating all sampled individuals
 at each position along the genome.
-In this paper, describe algorithmic tools and data structures developed for the 
+In this paper, we describe how algorithmic tools and data structures developed for the 
 coalescent simulator \msprime{}
-to efficiently record, and later process, this history.
+can be used to efficiently record, and later process, this history.
 
@@
 can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
-for tens of thousands of generations --
-but still substantially short of whole genomes for most biological population sizes.
-\jda{I'd find this more compelling with something added like: "e.g., Xs of
-	gigabases for Ys of individuals, implying 10Ys of generations needed to
-	equilibrate}
-There has been some progress at parallelization \citep{lawrie2017accelerating},
-which has substantial promise.
-\jda{this is vague, promise for what?}
+for tens of thousands of generations.
+This is still substantially short of whole genomes (hundreds to thousands of megabases)
+for many biological population sizes of interest 
+(e.g., millions for a restricted population of mosquitos).
+Achieving these sizes will likely require parallelization \citep{lawrie2017accelerating}.
 The general idea of storing genealogical information to speed up simulations is not new --
@@
 (a) a format, implemented in the \msprime{} Python API,
-for storing tree sequences in several \emph{tables};
+for recording tree sequences in several \emph{tables};
 (b) an algorithm, \algref{W}, to record tree sequence information into these tables on the fly
     from a forwards-time simulation;
-and (c) a method to \emph{simplify} a tree sequence, i.e., remove redundant information.
-We then describe a method that combines (b) and (d) to allows efficiently record simplified a tree sequence,
-and present analyses of its run time complexity.
-This is followed by benchmarking comparisons between two implementations of our method,
-each using previously published forwards-time simulation libraries.
+and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
+We then describe a general-purpose method that uses these tools
+(as implemented in \msprime{}) 
+to efficiently record simplified a tree sequence,
+and analyze of its run time complexity.
+This is followed by benchmarking the performance improvement achieved by connecting
+two previously published forwards-time simulation libraries,
+\simupop{} \citep{simupop} and \fwdpp{} \citep{fwdpp}
+to the tools in \msprime{}.
 
 %%%%%%
-\subsection*{Storing a tree sequence}
+\subsection*{Recording a tree sequence}
 
@@
 The probability of death per individual each generation is $\delta$.
-\jda{In the below we are using the integer vector encoding of a tree as in
-	\msprime{} should say so and reference it as the role of node ids is
-	otherwise more confusing than it need be}
 We will use
@@
 (and each such instance represents and independent draw).
+
+We begin in~\algref{W1} by allocating our initial population $P$
+(a vector of $N$ node IDs),
+and creating $N$ nodes with birth time $T$ generations ago, recorded in the
+vector $\tau$. The set $E$ is used to store the edges that we output during the
+simulation, and $n$ is the number of nodes created so far (and so, the ID of
+the next node we create). Steps~\algref{W2} and~\algref{W3} simply loop over
+the generation clock $t$ and individual index $j$. In~\algref{W4} we check if
+an individual $P_j$ has died in this generation. If it has, we replace it in
+steps \algref{W5}--\algref{W7}; if not, we proceed immediately to the next
+individual. When an individual in the population dies, we first allocate
+a new node with ID $n$ in~\algref{W5} and record its birth time. Then,
+in step~\algref{W6} we choose two indexes $a$ and $b$ uniformly (giving us
+parental IDs $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
+event by storing two new edges: one recording that the parent of node $n$
+from $0$ to $x$ is $P_a$, and another recording that the parent of $n$
+from $x$ to $L$ is $P_b$. We then complete the replacement event by incrementing
+$n$, ready to represent the next new node.
 
@@
 
-\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) \geq \delta$ go to \algref{W8}.
+\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) \geq \delta$ go to \algref{W9}.
 }
@@
 
-\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, n - 1\})$,
-    $b \leftarrow \randomuniform(\{0, \dots, n - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
+\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
+    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
 }
 
-\algstep{W7.}{Record edges}{Set $E \leftarrow E \cup \left\{ (0, x, P_a, n), (x,
-		L, P_b, n) \right\}$.
+\algstep{W7.}{Record edges}{Set 
+    $E \leftarrow E \cup \left\{ (0, x, P_a, n), (x, L, P_b, n) \right\}$.
 }
@@
 
-\algstep{W8.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W3}.
+\algstep{W9.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W3}.
 }
@@
 \end{taocpalg}
-
-Algorithm~\algref{W} is a very simple implementation of a Wright--Fisher model
-intended to illustrate how we can record the nodes and edges of a tree sequence
-forwards in time. We begin in~\algref{W1} by allocating our initial population $P$
-and creating $N$ nodes with birth time $T$ generations ago, recorded in the
-vector $\tau$. The set $E$ is used to store the edges that we output during the
-simulation, and $n$ is the number of nodes created so far (and so, the ID of
-the next node we create). Steps~\algref{W2} and~\algref{W3} simply loop over
-the generation clock $t$ and individual index $j$. In~\algref{W4} we check if
-an individual $P_j$ has died in this generation. If it has, we replace it in
-steps \algref{W5}--\algref{W7}; if not, we proceed immediately to the next
-individual. When an individual in the population dies, we first allocate
-a new node with ID $n$ in~\algref{W5} and record its birth time. Then,
-in step~\algref{W6} we choose two indexes $a$ and $b$ uniformly (giving us
-parents $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
-event by storing two new edges: one recording that the parent of node $n$
-from $0$ to $x$ is $P_a$, and another recording that the parent of $n$
-from $x$ to $L$ is $P_b$. We then complete the replacement event by incrementing
-$n$, ready to represent the next new node.
 
@@
 but is reasonably close.
-The process is depicted in Figure~\ref{fig:simplify_state}.
-\jda{Need refs to Figure~\ref{fig:method_diagram} in the text}
+The process is depicted in Figures~\ref{fig:method_diagram} and \ref{fig:simplify_state}.
 
@@
 
-\subsection*{Forward simulation and sequential simplification to record simplifed tree sequences}
+\subsection*{Forward simulation and sequential simplification}
 
-At any point in Algorithm~\algref{W}, genealogical history is recorded
-in an unsimplified tree sequence. This has two additional advantages. First, simplification
+At any point in a simulation scheme like Algorithm~\algref{W}, genealogical history is recorded
+in a tree sequence. This has two additional advantages. First, simplification
 can be run periodically through the simulation, taking the set of samples to be
@@
 that are not necessarily satisfied by tables emitted by a forwards-time simulation.
-Fortunately, \msprime{}'s API includes tools to rectify this by first sorting (using \texttt{sort\_tables})
+\msprime{}'s API includes tools to rectify this by first sorting (using \texttt{sort\_tables})
 and then using \texttt{simplify}, which works on sorted tables

### 2017-11-18 forwards_paper.tex
Remove old comment; s/delta/Delta in F

@@
 under the assumptions of coalescent theory \citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
-\jda{would be nice to say, concisely the difference}
 
@@
 
-\algstep{F2.}{Record trees till time $T - \delta$.}{ 
+\algstep{F2.}{Record trees till time $T - \Delta$.}{ 
 Simulate, using Algorithm~\algref{W} or another algorithm, $2N$ haploid
@@
 
-\algstep{F3.}{Simplification over $\delta$.}{
+\algstep{F3.}{Simplification over $\Delta$.}{
 Select final samples $S'$ and using the simplification Algorithm~\algref{S},

### 2017-11-18 forwards_paper.tex
Very rough outline of Alg W using table notation.
closes #24

@@
 with a probability of death per generation $\delta$ and a chromosome of length
-$L$ for $T$ generations. On termination, there will be $n$ nodes with times recorded
-in the vector $\tau$, and $E$ will contain a set of $(\ell, r, p, c)$ tuples
-describing the recorded edges.
+$L$ for $T$ generations. Events are recorded in a node table $\mathcal{N}$
+and edge table $\mathcal{E}$. Simplify is run every $s$ generations.
 }
@@
 \algstep{W1.}{Initialisation.}{
- For $0 \leq j < N$, set $P_j \leftarrow j$ and $\tau_j \leftarrow T$.
-Set $E\leftarrow \emptyset$, $t \leftarrow T$ and $n \leftarrow N$.
+Set $\mathcal{N} \leftarrow \mbox{NodeTable}()$ and $\mathcal{E}
+\leftarrow \mbox{EdgeTable}()$ and $t \leftarrow T$.
+ For $0 \leq j < N$, set $P_j \leftarrow \mbox{add}(\mathcal{N}, T)$.
 }
@@
 
-\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) \geq \delta$ go to \algref{W9}.
+\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) \geq \delta$ go to \algref{W8}.
 }
 
-\algstep{W5.}{New node.}{Set $P'_j \leftarrow n$ and $\tau_n \leftarrow t$.
+\algstep{W5.}{New node.}{Set $u \leftarrow \mbox{add}(\mathcal{N}, t)$ and $P'_j \leftarrow u$.
 }
@@
 
-\algstep{W7.}{Record edges}{Set 
-    $E \leftarrow E \cup \left\{ (0, x, P_a, n), (x, L, P_b, n) \right\}$.
+\algstep{W7.}{Record edges}{
+Call $\mbox{add}(\mathcal{E}, (0, x, P_a, u))$ and $\mbox{add}(\mathcal{E}, (x,
+L, P_b, u))$.
 }
 
-\algstep{W8.}{Increment the current node.}{
-$n \leftarrow n + 1$.
+\algstep{W8.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W3}.
 }
 
-\algstep{W9.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W3}.
-}
+\algstep{W9.}{Simplify}{TODO}
 

### 2017-11-19 forwards_paper.tex
Revise tables subhed per plr suggestion
Remove F, clean up a bit

@@
 %%%%%%
-\subsection*{Recording a tree sequence}
+\subsection*{Working with a tree sequence through tables}
 
@@
 
+\jda{TODO: rewrite with tables}
 We begin in~\algref{W1} by allocating our initial population $P$
@@
 
-At any point in a simulation scheme like Algorithm~\algref{W}, genealogical history is recorded
-in a tree sequence. This has two additional advantages. First, simplification
+At any point in a simulation scheme that, like Algorithm~\algref{W}, records
+data into tables the genealogical history is recorded
+in a tree sequence.
+This has two additional advantages, as seen in Algorithm~\algref{W}.
+First, simplification
 can be run periodically through the simulation, taking the set of samples to be
@@
 instance, by a coalescent simulation with \msprime. 
-This process of forward simulation coupled with sequential simplification is our
-method
-
-\jda{I didn't write down the full thing, but mightn't things be clearer if we
-	have a statement of the combined method?}
-\begin{taocpalg}{F}{Forwards-time simplifed tree sequence}{
-Simulates a population starting with $N$ diploid individuals on a chromosome of length $L$ for $T$
-generations.
-}
-
-\algstep{F0.}{Initialize nodes.}{
-Generate initial haploid nodes $\mathcal N$:
-1.) using coalsecent simulation with \msprime{} for $2N$ initial haploid samples $S$ on
-a chromosome of length $L$ and taking $\mathcal N = \{\texttt{id}(u) : u \in S \}$; or
-2.) simply defining $\mathcal N = \{1\dots 2N\}$ as the node ids.
-Define a mapping from  $\mathcal M: \{1\dots 2N\} \to \mathcal N$. In the second
-case $\mathcal M$ is the identity.
-}
-
-\algstep{F1.}{Initialize edges.}{
-Define the initial edges $E_0$: 1) either from the coalescent simulation's final
-state, or 2) as \dots
-}
-
-\algstep{F2.}{Record trees till time $T - \Delta$.}{ 
-Simulate, using Algorithm~\algref{W} or another algorithm, $2N$ haploid
-individuals using $\mathcal M$ to renumber the nodes, producing nodes $n$ and
-edgesets $E$.
-}
-
-\algstep{F3.}{Simplification over $\Delta$.}{
-Select final samples $S'$ and using the simplification Algorithm~\algref{S},
-simplify $n$ and $E$ to produce a simplified tree sequence.
-}
-
-\algstep{F4.}{Continue?}{
-	Stop if $T - \delta  = 0$, else continue.
-}
-\end{taocpalg}
+We provide results for simulation timing from two specific implementations of this
+method (forward simulation coupled with sequential simplification)
+in the next section, but first we analyze the expected resource use of the
+general Algorithm~\algref{W}.
 
@@
 
-\jda{I'd like to outline what we're doing here with reference to alg F}
+\jda{It'd be good to outline what readers should expect here.}
+\jda{and note we're doing the space analysis for with reference to alg W with $s=1$}
 Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals 

### 2017-11-19 forwards_paper.tex
finished simplify description

@@
 \newcommand{\setdiff}{\smallsetminus}
+\newcommand{\Nt}{\mathcal{N}}  % node table
+\newcommand{\Et}{\mathcal{E}}  % edge table
+\newcommand{\St}{\mathcal{S}}  % site table
+\newcommand{\Mt}{\mathcal{M}}  % mutation table
+\newcommand{\Al}{\mathcal{A}}  % list of ancestors
+\DeclareMathOperator{\len}{length}
+\DeclareMathOperator{\add}{add} % for adding to tables
+\newcommand{\lleft}{\mathrm{left}}  % left label
+\newcommand{\lright}{\mathrm{right}}
+\newcommand{\lid}{\mathrm{id}}
 
@@
 ordered by left endpoint.
+As it goes, the algorithm 
+records a list $M$ that gives the mapping
+from input nodes to output nodes,
+so that if $M[p] = u$ then the $p^\text{th}$ node in the input
+is the same as the $u^\text{th}$ node in the output.
+The current state of the algorithm is recorded in $\Al$, 
+a list of ancestors
+each of which is a collections of ancestral segments.
+An ancestral segment is a triple $(\ell, r, x)$,
+indicating that the output node $x$ inherits from this ancestor
+on the genomic segment $[\ell, r)$.
 
 \begin{taocpalg}{S}{Simplify a tree sequence}
-{Simplifies the input tree sequence:
-    input consists of 
-    a list of sample IDs, $S$, of length $n$,
-    a list of edges, $\mathcal{E_I}$,
-    and a list of nodes $\mathcal{N_I}$.
-    The genome length is $L$.
-    The output will be stored in a list of edges $\mathcal{E_O}$ 
-    and nodes $\mathcal{N}_O$.
-    This also maintains the list $N$ that gives the mapping 
-    from $\mathcal{N_I}$ to $\mathcal{N_O}$,
-    so that if $M[p] = u$ then the $p^\text{th}$ ancestor in $N_I$
-    is the same as the $u^\text{th}$ ancestor in $N_0$.
-    The internal state is $\mathcal{A}$, the list of ancestors
-    (which are collections of ancestral segments).
+{Input consists of
+    a list $S$ of sample IDs,
+    a list $\Nt_I$ of nodes (ordered by birth time),
+    a list $\Et_I$ of edges,
+    and the genome length, $L$.
+    The output is a 
+    nodes $\Nt_O$,
+    and list of edges $\Et_O$.
 }
@@
 \algstep{S0.}{Initialize.}{
-    Set $\mathcal{A}[j] \leftarrow (0, L, j)$ 
-    and $\mathcal{N}_O[j] \leftarrow \mathcal{N}_I[S[j]]$ for $0 \le j < N$.
-    Set $Q \leftarrow \emptyset$,
-    and $M[p] \leftarrow -1$ for $0 \le p < \text{length}(N_I)$.
+    Set $\Al[j] \leftarrow (0, L, j)$
+    and $\Nt_O[j] \leftarrow \Nt_I[S[j]]$ for $0 \le j < \len(S)$.
+    Set the current input parent index to $p=0$
+    and $M[u] \leftarrow -1$ for $0 \le u < \len(\Nt_I)$.
 }
 
-\algstep{S1.}{Input parent loop.}{Loop over parents, chronologically.
+\algstep{S1.}{Input parent loop.}{Set $Q \leftarrow \emptyset$.
 }
 
-\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} 
-    on each edge corresponding to the current parent.
+\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} on each edge $(\ell, r, p, c)$ in $\Et_I$
+    corresponding to parent $p$ (which stores the resulting segments in the merge queue $Q$).
 }
 
-\algstep{S3.}{Merge ancestors.}{Merge the resulting segments into a new ancestor for this parent.
+\algstep{S3.}{Merge ancestors.}{Call Algorithm \algref{M} on $Q$
+    (which adds to $\Nt_O$ and $\Et_O$).
+}
+
+\algstep{S4.}{Next parent.}{Set $p \leftarrow p+1$;
+    if $p = \len(\Nt_I)$, terminate and return $\Nt_O$ and $\Et_O$;
+    otherwise, return to \algref{S1}.
 }
@@
 
-The first step is, for each edge $(\ell, r, p, c)$,
-simply to replace any ancestry segment $[a, b)$ held by ancestor $c$
-with $[a,b) \setdiff [\ell, r)$ 
+The algorithm steps back up through the pedigree,
+since nodes are ordered by time-since-birth.
+First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
+so, for every edge $(\ell, r, p, c)$,
+Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
+with $[u,v) \setdiff [\ell, r)$
 (deleting the segment or splitting in two if necessary),
-and adding the removed segment $[a,b) \cap [\ell, r)$ to the merge queue $Q$:
+and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:
+
 \begin{taocpalg}{R}{Remove ancestry}
     {Given an edge $(\ell, r, p, c)$,
-    and a list $\mathcal{A}[c]$ of the $m$ ancestral segments corresponding to $c$,
-    remove segments overlapping $[\ell, r)$ from $\mathcal{A}[c]$,
+    and a list $\Al[c]$ of the $m$ ancestral segments corresponding to $c$,
+    remove segments overlapping $[\ell, r)$ from $\Al[c]$,
     and add those overlapping segments to $Q$.
@@
 
-    \algstep{R1.}{Segment loop.}{Loop over ancestral segments in $\mathcal{A}[c]$:
-    if $j=m$, terminate; else let $(a, b, u) = \mathcal{A}[c][j]$.
+    \algstep{R0.}{Initialize.}{
+        Set $j \leftarrow 0$.
     }
 
-    \algstep{R2.}{Add to merge queue.}{If $[a,b) \cap [\ell,r) \neq \emptyset$,
-    append $(\max(a,\ell), \min(b,r), u)$ to $Q$.
+    \algstep{R1.}{Segment loop.}{
+    If $j=m$, terminate; else let $(u, v, x) = \Al[c][j]$.
+    (Here $x$ is the output node that inherits from this segment.)
     }
 
-    \algstep{R3.}{Remove ancestry.}{If $[a,b) \cap [\ell,r) != \emptyset$,
-    delete $\mathcal{A}[c][j]$ and insert in its place
-    $(\min(a,\ell), \max(a,\ell), u)$ (if $a<\ell$) 
-    and/or $(\min(b,r), \max(b,r), u)$ (if $r<b$).
+    \algstep{R2.}{Add to merge queue.}{If $[u,v) \cap [\ell,r) = \emptyset$,
+    skip to \algref{R4}.
+    Otherwise, append $(\max(u,\ell), \min(v,r), x)$ to $Q$. 
+    }
+
+    \algstep{R3.}{Remove ancestry.}{
+    Delete $\Al[c][j]$ and insert in its place
+    $(\min(u,\ell), \max(v,\ell), x)$ (if $u<\ell$)
+    and(/or) $(\min(v,r), \max(v,r), x)$ (if $r<v$).
+    }
+
+    \algstep{R4.}{Iterate.}{Set $j \leftarrow j+1$ and return to \algref{R1}.
     }
@@
 
-Next we must merge these ancestry segments,
-creating the new ancestor $p$,
-outputting edges and a node if coalescence has occurred.
+Next we must merge the ancestry segments in $Q$,
+creating the new ancestor corresponding to input node $p$,
+adding a node and edges to the output if any segments overlap,
+which implies that an output coalescence has occurred.
+Below we index elements of an ancestral segment $z = (\ell, r, u)$
+by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
+and $z.\lid = u$ for the output node ID.
+
 \begin{taocpalg}{M}{Merge ancestors}
-    {Merge a collection of segments and create a new ancestor.
-    This operates on the merge queue $Q$ above,
-    which is a list of ancestral segments,
-    and creates the new ancestor $\mathcal{A}[p]$.}
+    {Merge the ancestral segments in $Q$,
+    and create the new ancestor $\Al[p]$.
+    This algorithm also appends to $\Nt_O$ and $\Et_O$,
+    and needs to know about $S$, 
+    to check if input nodes are samples.
+    }
 
-    \algstep{M1.}{Segment loop.}{While $\mathrm{length}(Q)>0$:}
+    \algstep{M0.}{Initialize.}{
+        Set $z \leftarrow (0, 0, -1)$ and $d = \text{False}$.
+    }
+
+    \algstep{M1.}{Segment loop.}{If $\len(Q) = 0$, go to \algref{M9}.}
 
     \algstep{M2.}{Find next segment overlap boundaries}{
-    Let $\ell \leftarrow \min\{x.\ell : x \in Q\}$
-    and $H \leftarrow \{x \in Q : x.\ell = \ell\}$.
-    Also let $\ell_+ \leftarrow \min\{x.\ell : x \in Q \setdiff H\}$
-    and $r = \min(\ell_+, \min\{x.r : x \in H\})$.
+        Let $\ell_0 \leftarrow \min\{x.\lleft : x \in Q\}$
+        and $H \leftarrow \{x \in Q : x.\lleft = \ell_0\}$.
+        Also let $\ell_+ \leftarrow \min\{x.\lleft : x \in Q \setdiff H\}$
+        and $r = \min(\ell_+, \min\{x.\lright : x \in H\})$.
     }
 
-    \algstep{M3.}{Copy single segment.}{If $\mathrm{length}(H) == 1$,
-    set $\alpha \leftarrow (\ell, r, H[0].n)$.
-    If also $p \in S$, then append $(\alpha.\ell, \alpha.r, N[p], \alpha.c)$ to $\mathcal{E}_O$
-    and then set $\alpha.c \leftarrow M[p]$.
+    \algstep{M3.}{Copy single segment.}{
+        If $\len(H) = 1$,
+        set $\alpha \leftarrow (\ell, r, H[0].\lid)$.
+        If also $p \in S$, then
+        set $\alpha.\lid \leftarrow M[p]$ and
+        $\add(\Et_O, (\alpha.\lleft, \alpha.\lright, \Nt[p], \alpha.\lid))$.
         Go to \algref{M7}.
@@
 
-    \algstep{M4.}{Coalescence head.}{If $(h \leftarrow \mathrm{length}(H)) > 1$,
-    if $M[p] == -1$ set $M[p] \leftarrow \text{length}(N_O)$
-    and append $N_I[p]$ to $N_O$.
+    \algstep{M4.}{New output node.}{
+        If $M[p] = -1$ set $M[p] \leftarrow \add(\Nt_O, \Nt_I[p])$
         Set $\alpha \leftarrow (\ell, r, M[p])$.
@@
 
-    \algstep{M6.}{Record edges.}{
-    For each $x \in H$, append $(\ell, r, M[p], x.n)$ to $\mathcal{E}_O$.
+    \algstep{M6.}{Record edges and update $Q$.}{
+        For each $x \in H$, do $\add(\Et_O, (\ell, r, M[p], x.\lid))$.
     }
@@
     \algstep{M7.}{Update $Q$.}{
-    Set $Q \leftarrow Q \setdiff \{x \in Q : x.r == r\}$
-    and then set $x.\ell = \ell$ for all remaining $x \in Q$.
+        Remove from $Q$ any segment $x \in H$ with $x.\lright = r$
+        and then set $x.\lleft = \ell$ for all remaining $x \in H$.
     }
@@
     \algstep{M8.}{Add segment to ancestor.}{
-    Append $\alpha$ to $\mathcal{A}[p]$ and return to \algref{M1}.
+        If $z.\lright = \alpha.\lleft$ and $z.\lid = \alpha.\lid$ set $d = \text{True}$.
+        Append $\alpha$ to $\Al[p]$, set $z \leftarrow \alpha$, and return to \algref{M1}.
+    }
+
+    \algstep{M9.}{Defragment.}{
+        If $d$ is True, find and merge any adjacent segments in $\Al[p]$
+        with the same $\lid$.  Terminate.
     }

### 2017-11-20 sims/msprime-examples.py
Added script to benchmark simplify by number of edges.

@@ -114,7 +114,7 @@ def run_stats(args):
     assert num_trees == ts.num_trees
 
 
-def run_simplify_benchmark(args):
+def run_simplify_subsample_size_benchmark(args):
     ts = msprime.load(args.file)
     np.random.seed(1)
     print("Running simplify benchmarks")
@@ -141,6 +141,56 @@ def run_simplify_benchmark(args):
     plt.xlabel("Subsample size")
     plt.ylabel("Time to simplify (s)")
     plt.savefig("simplify_subsample_perf.pdf", format='pdf')
+
+
+def run_simplify_num_edges_benchmark(args):
+    ts = msprime.load(args.file)
+    np.random.seed(1)
+    print("num_nodes = ", ts.num_nodes)
+    print("num_edges = ", ts.num_edges)
+    num_slices = 20
+    N = 1000
+
+    tables = ts.dump_tables()
+    nodes = tables.nodes
+    edges = tables.edges
+
+    node_time = nodes.time
+    left = edges.left
+    right = edges.right
+    parent = edges.parent
+    child = edges.child
+
+    size = left.nbytes + right.nbytes + parent.nbytes + child.nbytes
+    print("Total edge size = ", size / 1024**3, "GiB")
+
+    num_edges = np.zeros(num_slices)
+    simplify_time = np.zeros(num_slices)
+    slice_size = ts.num_edges // num_slices
+    for j, start in enumerate(range(ts.num_edges - slice_size, 0, -slice_size)):
+        max_node = np.max(child[start:])
+        samples = np.arange(max_node - N, max_node, dtype=np.int32)
+        subset_nodes = msprime.NodeTable()
+        subset_nodes.set_columns(
+            time=node_time[:max_node + 1], flags=np.ones(max_node + 1, dtype=np.uint32))
+        subset_edges = msprime.EdgeTable()
+        subset_edges.set_columns(
+            left=left[start:], right=right[start:], parent=parent[start:],
+            child=child[start:])
+        before = time.process_time()
+        msprime.simplify_tables(samples=samples, nodes=subset_nodes, edges=subset_edges)
+        duration = time.process_time() - before
+        num_edges[j] = ts.num_edges - start
+        simplify_time[j] = duration
+        print(num_edges[j], duration, num_edges[j] / duration, "per second")
+
+    df = pd.DataFrame({"num_edges": num_edges, "time": simplify_time})
[…]

### 2017-11-20 sims/bulk_evolve.py
Added script used to generate data.

@@ -1,0 +1,34 @@
+
+# Script used to generate the data for the plot benchmarking simplify performance
+# against the number of edges. This was copied from the fwdpy11_arg_example repository
+# where it was originally in the update_to_edge branch.
+import fwdpy11_arg_example.evolve_without_simplify
+import numpy as np
+import msprime
+import time
+
+before = time.process_time()
+x = fwdpy11_arg_example.evolve_without_simplify.evolve(1000, 10000)
+# x = fwdpy11_arg_example.evolve_without_simplify.evolve(100, 100)
+duration = time.process_time() - before
+
+print("Ran simulation in {} seconds".format(duration))
+
+before = time.process_time()
+sim_nodes = np.array(x.nodes, copy=False)
+sim_edges = np.array(x.edges, copy=False)
+nodes = msprime.NodeTable()
+edges = msprime.EdgeTable()
+flags = np.ones(len(sim_nodes), dtype=np.uint32)
+nodes.set_columns(
+    flags=flags, population=sim_nodes['population'], time=sim_nodes['generation'])
+edges.set_columns(left=sim_edges['left'], right=sim_edges['right'],
+                  parent=sim_edges['parent'], child=sim_edges['child'])
+duration = time.process_time() - before
+
+print("copies data into tables in {} seconds".format(duration))
+
+msprime.sort_tables(nodes=nodes,edges=edges)
+ts = msprime.load_tables(nodes=nodes, edges=edges)
+ts.dump("tmp__NOBACKUP__/raw-simulation.hdf5")
+

### 2017-11-20 forwards_paper.tex
removed trailing spaces
fix citations

@@
 as the genome-wide impacts of hitchhiking and background selection are more widely appreciated.
-For instance, \citet{harris2016genetic} used SLiM \citep{haller2017flexible} to simulate tens of thousands of human exomes
+For instance, \citet{harris2016genetic} used SLiM \citep{slim`} to simulate tens of thousands of human exomes
 to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
@@
 needs to fill the relevant columns and send them to \msprime{} at regular intervals for simplification.  For example,
-using the \fwdpp{} C++ API \cite{Thornton2014-hx}, we fill arrays of simple structures representing the node and edge data.
+using the \fwdpp{} C++ API \cite{fwdpp}, we fill arrays of simple structures representing the node and edge data.
 Doing so requires no modification of the \fwdpp{} code base.  Rather, we simply need to book-keep the parent/offspring

### 2017-11-20 forwards_paper.tex
remove xtra char

@@
 as the genome-wide impacts of hitchhiking and background selection are more widely appreciated.
-For instance, \citet{harris2016genetic} used SLiM \citep{slim`} to simulate tens of thousands of human exomes
+For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate tens of thousands of human exomes
 to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.

### 2017-11-20 forwards_paper.tex

@@
 as the genome-wide impacts of hitchhiking and background selection are more widely appreciated.
-For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate tens of thousands of human exomes
+For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
 to assess the impact of genetic load and Neanderthal introgression on human genetic diversity. 
-This is roughly state-of-the-art
--- the fastest simulation frameworks such as SLiM and fwdpp \citep{fwdpp}
+\cite{Sanjak2017-ko} used
+fwdpp \citep{fwdpp} to simulate a series of models of quantiative traits under mutation-selection balance with
+population sizes of $2 times 10^4$ diploids in stable populations and populations growing up to $\approx 5
+\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
+outcomes.
+
+Due to tracking neutral mutations, the fastest simulation frameworks such as SLiM, SLiM2 \cite{haller2017flexible} and fwdpp \citep{fwdpp}
 can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
-for tens of thousands of generations.
-This is still substantially short of whole genomes (hundreds to thousands of megabases)
-for many biological population sizes of interest
-(e.g., millions for a restricted population of mosquitos).
-Achieving these sizes will likely require parallelization \citep{lawrie2017accelerating}.
-The general idea of storing genealogical information to speed up simulations is not new --
-it was entirely implemented in AnA-FiTS \citep{aberer2013rapid}, for instance --
-but this program did not implement the critical step of discarding irrelevant genealogical information.
+for tens of thousands of generations, and there are still near-order-of-magnitude performance differences between these
+implementations \cite{haller2017flexible}. In practice, current state-of-the-art simulation software may take on the
+order of 
+weeks to simulate models without selection\cite{fwdpp,Hernandez2015-wf}, and different simulation engines differ in how efficiently they
+calculate fitnesses in models with selection \cite{fwdpp}.
+These population and region sizes are still substantially short of whole genomes (hundreds to thousands of megabases)
+for many biological population sizes of interest.
+Simulating very large popluations and entire genomes will likely require parallelization, which \cite{lawrie2017accelerating} used to improve
+performance for the "Poisson Random Field" family of models \cite{Sawyer1992-jw}, in which mutations are unlinked and there is no
+effect of genetic background on fitness.  However, little, and perhaps even no, attention has been paid to
+parallelization of simulations involving linkage.  Here, we show that recording genealogical information instead of
+explicitly tracking neutral mutations results in order-of-magnitude performance improvements to simulations based on
+\fwdpp{}.  The  idea of storing genealogical information to speed up simulations is not new.  AnA-FiTS implemented it
+completely \citep{aberer2013rapid}, but this program did not implement the critical step of discarding irrelevant genealogical information.
 A similar but more limited method for discarding this information does appear in \citet{padhukasahasram2008exploring}.

### 2017-11-22 forwards_paper.tex
Updated introduction.

@@
     in a population genetics simulation can be substantial.
-    In this note we describe how to both (a) dramatically reduce this burden and
-    (b) efficiently record the entire history of the population.
+    In this paper we describe how to dramatically reduce this burden by
+    efficiently recording the entire history of the population.
     We do this by simulating only those loci that may affect reproduction (those having non-neutral variants),
@@
 for example, a recent study~\citep{martin2017human}
-simulated 600,000 samples of human chromosome 20
-% to assess the impact of European biased reference panels in genome wide association studies.
+simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
 Several studies have now shown that in samples of size approaching that of the population,
@@
 \citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
-These considerations, and increasing computational power, have led to a resurgence of interest in forwards-time, individual-based simulations.
+These considerations, and increasing computational power, have led to a resurgence of
+interest in large scale forwards-time, individual-based simulations.
+For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
+to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
+\cite{Sanjak2017-ko} used
+fwdpp \citep{fwdpp} to simulate a series of models of quantiative traits under mutation-selection balance with
+population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\approx 5
+\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
+outcomes.
 
@@
 then such simulations must somehow produce at the end data for each individual on a genomic scale.
-As samples of most species genomes harbor tens or hundreds of millions of variant sites,
+As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
 naively carrying full genotypes for even modest numbers of individuals through a simulation
 becomes quickly prohibitive.
+Because of this computational burden of neutral mutations, even the fastest simulation frameworks such as SLiM,
+SLiM2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
+can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
+for tens of thousands of generations, and there are still near-order-of-magnitude performance differences between these
+implementations~\citep{haller2017flexible}.
+In practice, current state-of-the-art simulation software may take on the order of
+weeks to simulate models without selection~\citep{fwdpp,Hernandez2015-wf},
+and different simulation engines differ in how efficiently they
+calculate fitnesses in models with selection~\citep{fwdpp}.
+These population and region sizes are still substantially short of whole genomes
+(hundreds to thousands of megabases)
+for many biological population sizes of interest.
 
@@
 can be used to efficiently record, and later process, this history.
-
 In so doing we record the \emph{population pedigree} --
-the entire history of parent-offspring relationships of an entire population going back to a remote time --
-as well as information encoding the genetic outcomes of each ancestral meiosis --
-who inherited which parts of which parental chromosomes.
+the complete history of parent-offspring relationships of an entire population
+going back to a remote time including the genetic outcomes of each ancestral
+meiosis.
 This embellished pedigree contains all the information necessary
 to construct the genealogical tree that relates each individual to each other
-at each position on the genome, i.e., the \emph{tree sequence}.
+at each position on the genome, i.e., the \emph{tree
+sequence}~\citep{kelleher2016efficient}.
 Combined with ancestral genotypes and the origins of new mutations,
@@
 which has been the subject of substantial study
-under the assumptions of coalescent theory \citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
-
-\jda{Is this the right place for this paragraph? should it maybe be integrated
-	with graf beginning 'Modern computing' above?}
-\plr{this paragraph is supposed to cite the fwds time simulation methods we need to - @molpopgen fix this up?}
-Forwards-time population genetics simulations are growing in importance,
-as the genome-wide impacts of hitchhiking and background selection are more widely appreciated.
-For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
-to assess the impact of genetic load and Neanderthal introgression on human genetic diversity. 
-\cite{Sanjak2017-ko} used
-fwdpp \citep{fwdpp} to simulate a series of models of quantiative traits under mutation-selection balance with
-population sizes of $2 times 10^4$ diploids in stable populations and populations growing up to $\approx 5
-\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
-outcomes.
-
-Due to tracking neutral mutations, the fastest simulation frameworks such as SLiM, SLiM2 \cite{haller2017flexible} and fwdpp \citep{fwdpp}
-can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
-for tens of thousands of generations, and there are still near-order-of-magnitude performance differences between these
-implementations \cite{haller2017flexible}. In practice, current state-of-the-art simulation software may take on the
-order of 
-weeks to simulate models without selection\cite{fwdpp,Hernandez2015-wf}, and different simulation engines differ in how efficiently they
-calculate fitnesses in models with selection \cite{fwdpp}.
-These population and region sizes are still substantially short of whole genomes (hundreds to thousands of megabases)
-for many biological population sizes of interest.
-Simulating very large popluations and entire genomes will likely require parallelization, which \cite{lawrie2017accelerating} used to improve
-performance for the "Poisson Random Field" family of models \cite{Sawyer1992-jw}, in which mutations are unlinked and there is no
-effect of genetic background on fitness.  However, little, and perhaps even no, attention has been paid to
-parallelization of simulations involving linkage.  Here, we show that recording genealogical information instead of
-explicitly tracking neutral mutations results in order-of-magnitude performance improvements to simulations based on
-\fwdpp{}.  The  idea of storing genealogical information to speed up simulations is not new.  AnA-FiTS implemented it
-completely \citep{aberer2013rapid}, but this program did not implement the critical step of discarding irrelevant genealogical information.
-A similar but more limited method for discarding this information does appear in \citet{padhukasahasram2008exploring}.
-
+under the assumptions of coalescent
+theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
+
+\jk{And finish with a more detailed descrition of what we do in the paper}
 In this paper we discuss storage methods for genealogies (and hence, genome sequence),
@@
 
+\jk{Moved this from the introduction: here seems like the right place to
+discuss the importance of parallisation. We should rework this paragraph to
+discuss say that a nice part of our approach is that we can do the
+simplification in parallel. Also looking forward we need to parallise more
+because the simulations need so much RAM}
+Simulating very large popluations and entire genomes will likely require parallelization,
+which \cite{lawrie2017accelerating} used to improve
+performance for the ``Poisson Random Field'' family of models~\citep{Sawyer1992-jw},
+in which mutations are unlinked and there is no
+effect of genetic background on fitness.  However, little, and perhaps even no, attention has been paid to
+parallelization of simulations involving linkage.  Here, we show that recording genealogical information instead of
+explicitly tracking neutral mutations results in order-of-magnitude performance improvements to simulations based on
+\fwdpp{}.
+
+\jk{Moved this here from intro also. Basically we just need to say that others
+have done something similar. Ours is much better than Ana-Fits because ...}
+The  idea of storing genealogical information to speed up simulations is not new.  AnA-FiTS implemented it
+completely~\citep{aberer2013rapid}, but this program did not implement the critical step of
+discarding irrelevant genealogical information.  A similar but more limited method for
+discarding this information does appear in \citet{padhukasahasram2008exploring}.
+
 A final note:

### 2017-11-22 forwards_paper.tex
Updated tree sequence section.

@@
 
-A \emph{tree sequence} is an encoding for the sequence of correlated trees,
-such as those that describe the history of a sexual population.
+A \emph{tree sequence} is an encoding for a sequence of correlated trees,
+such as those describing the history of a sexual population.
 It is efficient because branches that are shared by adjacent trees are stored once,
@@
 here we explain conceptually,
-using the example of Figure~\ref{fig:example_tree_sequence}.
+using the example of Figure~\ref{fig:example_tree_sequence}. This formulation
+is derived from the ``coalescence records'' encoding of tree
+sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
+and generalised to include a much larger class of tree topology.
 
@@
 while nodes 3 and 4 represent those ancestors necessary to record their genealogy,
-who were born at time 1.5 and 2.5 respectively.
+who were born at time 1 and 2 respectively.
 
@@
 To modify the tree to reflect the genealogy at a nearby location,
-we need only remove those edges whose intervals do not overlap that location,
+we simply remove those edges whose intervals do not overlap that location,
 and add those new edges whose intervals do.
@@
 The first mutation occurs at site zero (in the left-hand tree), and is a simple
-mutation resulting in node $2$ inheriting the state 'T'.
+mutation resulting in node $2$ inheriting the state `T'.
 The second site (in the right hand tree) has two mutations:
@@
 This encoding of a sequence of trees and accompanying mutational information is
-very concise. To illustrate this, we ran a simulation of $500,000$ samples of a
+very concise. To illustrate this, we ran a simulation of 500,000 samples of a
 $200$ megabase human-like chromosome ($N_e=10^4$ and per-base mutation and
@@
 format would require around 20 TiB and storing the genotype information
-in VCF would require about 1 TiB --
-a compression factor of 144,000.
+in VCF would require about 1 TiB (giving a compression factor of 144,000 in
+this instance).
 Working with either the Newick or VCF encoding
 of this dataset would likely require several
-days of CPU time simply to read the information info memory.
+days of CPU time just to read the information into memory.
 

### 2017-11-22 forwards_paper.tex
Updated Algorithm W to include simplify.
Also general editing pass.

@@
 \newcommand{\Al}{\mathcal{A}}  % list of ancestors
+\newcommand{\taddnode}[1]{\mbox{addnode}(#1)}
+\newcommand{\taddedge}[1]{\mbox{addedge}(#1)}
+\newcommand{\tsimplify}[1]{\mbox{simplify}(#1)}
 \DeclareMathOperator{\len}{length}
@@
 $\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random
-(and each such instance represents and independent draw).
-
-\jda{TODO: rewrite with tables}
-We begin in~\algref{W1} by allocating our initial population $P$
-(a vector of $N$ node IDs),
-and creating $N$ nodes with birth time $T$ generations ago, recorded in the
-vector $\tau$. The set $E$ is used to store the edges that we output during the
-simulation, and $n$ is the number of nodes created so far (and so, the ID of
-the next node we create). Steps~\algref{W2} and~\algref{W3} simply loop over
-the generation clock $t$ and individual index $j$. In~\algref{W4} we check if
-an individual $P_j$ has died in this generation. If it has, we replace it in
-steps \algref{W5}--\algref{W7}; if not, we proceed immediately to the next
-individual. When an individual in the population dies, we first allocate
-a new node with ID $n$ in~\algref{W5} and record its birth time. Then,
-in step~\algref{W6} we choose two indexes $a$ and $b$ uniformly (giving us
+(and each such instance represents and independent draw). Given a node table
+$\Nt$, the function $\taddnode{\Nt, t}$ adds a new node to the table with time $t$
+and returns the ID of this new node. Similarly, the function $\taddedge{\Et,
+\ell, r, p, c}$ adds a new edge $(\ell, r, p, c)$ to the edge table $\Et$.
+The function $\tsimplify{P, \Nt, \Et}$ simplifies the history stored in the
+tables $\Nt$ and $\Et$ to the minimal information required to represent the
+genealogies of the list of node IDs $P$ (see section XXX).
+We first list
+the algorithm in detail, and immediately follow this with a step-by-step
+explanation.
+
+\begin{taocpalg}{W}{Forwards-time tree sequence}
+{Simulates a randomly mating population of $N$ haploid individuals with
+chromosome of length $L$ for $T$ generations, and returns the node
+and edge tables ($\Nt$ and $\Et$) recording the simulated history.
+Simplify is run every $s$ generations, removing genealogical
+information from $\Nt$ and $\Et$ unreachable from the current population $P$.
+}
+
+\algstep{W1.}{Initialisation.}{
+Set $\Nt \leftarrow \mbox{NodeTable}()$, $\mathcal{E}
+\leftarrow \mbox{EdgeTable}()$ and $t \leftarrow T$.
+ For $0 \leq j < N$, set $P_j \leftarrow \taddnode{\Nt, T}$. Set $j \leftarrow 0$.
+}
+
+\algstep{W2.}{New node.}{Set $u \leftarrow \taddnode{\Nt, t}$ and $P'_j \leftarrow u$.
+}
+
+\algstep{W3.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
+    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
+}
+
+\algstep{W4.}{Record edges}{
+Call $\taddedge{\Et, 0, x, P_a, u}$ and $\taddedge{\Et, x, r, P_b, u}$.
+}
+
+\algstep{W5.}{Individual loop}{Set $j \leftarrow j + 1$. If $j < N$ go to \algref{W2}.
+Otherwise, if $t\bmod s \neq 0$ go to \algref{W7}.
+}
+
+\algstep{W6.}{Simplify.}{Call $\tsimplify{P', \Nt, \Et}$, and set $P'_j
+\leftarrow j $ for $0 \leq j < N$. (Tables may need to be sorted.)}
+
+\algstep{W7.}{Generation loop.}{Set $t \leftarrow t - 1$. If $t = 0$ terminate.
+Set $P \leftarrow P'$, $j \leftarrow 0$, and go to \algref{W2}.
+}
+
+\end{taocpalg}
+
+We begin in~\algref{W1} by creating new node and edge tables, and setting
+our population $P$ (a vector of $N$ node IDs) to the initial population.
+This initial population is a set of $N$ nodes with birth time $T$ generations
+ago. We also initialise our generation clock $t$ and individual index $j$.
+Steps~\algref{W2}--\algref{W4} replace the individual with node ID
+$P_j$ by first creating a new node with birth time $t$ (and ID $u$).
+Then, in step~\algref{W3} we choose two indexes $a$ and $b$ uniformly (giving us
 parental IDs $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
-event by storing two new edges: one recording that the parent of node $n$
-from $0$ to $x$ is $P_a$, and another recording that the parent of $n$
-from $x$ to $L$ is $P_b$. We then complete the replacement event by incrementing
-$n$, ready to represent the next new node.
-
-\begin{taocpalg}{W}{Forwards-time tree sequence}
-{Simulates a randomly mating population $P$ of $N$ haploid individuals,
-with a probability of death per generation $\delta$ and a chromosome of length
-$L$ for $T$ generations. Events are recorded in a node table $\mathcal{N}$
-and edge table $\mathcal{E}$. Simplify is run every $s$ generations.
-}
-
-\algstep{W1.}{Initialisation.}{
-Set $\mathcal{N} \leftarrow \mbox{NodeTable}()$ and $\mathcal{E}
-\leftarrow \mbox{EdgeTable}()$ and $t \leftarrow T$.
- For $0 \leq j < N$, set $P_j \leftarrow \mbox{add}(\mathcal{N}, T)$.
-}
-
-\algstep{W2.}{Generation loop.}{ If $t = 0$ terminate.
-Set $P' \leftarrow P$ and then set $t \leftarrow t - 1$ and $j \leftarrow 0$.
-}
-
-\algstep{W3.}{Individual loop.}{ If $j = N$ set $P \leftarrow P'$ and go to \algref{W2}.
-}
-
-\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) \geq \delta$ go to \algref{W8}.
-}
-
-\algstep{W5.}{New node.}{Set $u \leftarrow \mbox{add}(\mathcal{N}, t)$ and $P'_j \leftarrow u$.
-}
-
-\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
-    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
-}
-
-\algstep{W7.}{Record edges}{
-Call $\mbox{add}(\mathcal{E}, (0, x, P_a, u))$ and $\mbox{add}(\mathcal{E}, (x,
-L, P_b, u))$.
-}
-
-\algstep{W8.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W3}.
-}
-
-\algstep{W9.}{Simplify}{TODO}
-
-\end{taocpalg}
-
-This algorithm records only the topological information resulting from the
-forwards in time Wright-Fisher process, but it is straightforward to add
-mutational information. This can be done in two different ways.
-We can record mutations that occur during the simulation quite simply.
-For example, in Algorithm~\algref{W} we would generate mutations after
-we have recorded the edges joining the new node $n$ to its parents
-$P_a$ and $P_b$ in step \algref{W7}.
-For example, if we assume that mutations occur according to the infinite sites model
-at rate $\mu$ per generation per unit length then there are
-$\mbox{Poisson}\left(\mu (r - \ell)(\tau_p
-- \tau_c)\right)$ mutations on each edge $(\ell, r, p, c)$. Each of these mutations
-will occur on a distinct site $x$ drawn uniformly from $(\ell, r]$ and be associated with node $c$.
-It is straightforward to record this information during the simulation, but it
-is significantly simpler and more efficient to generate these mutations
-after the simulation has completed.
-% In the case of infinite sites mutations
-% there is no difference between generating these mutations during the forwards
-% in time simulation and after it has completed. We are using precisely the same
-% information.
+event by storing two new edges: one recording that the parent of node $u$
+from $0$ to $x$ is $P_a$, and another recording that the parent of $u$
+from $x$ to $L$ is $P_b$. Step \algref{W5} then iterates over each of the
+$N$ individuals for each generation. At the end of a generation, we then check
+if we need to simplify (which is done every $s$ generations). If simplification
+is required, we do this in step \algref{W6} by calling the simplify function
+on the node and edge tables with the current set of population IDs $P'$ as
+the samples. This updates the tables in-place to remove all redundant
+information, and remaps the specified sample IDs to $0, \dots N - 1$
+in the updated tables. Hence, we set our current population IDs to
+$0, \dots N - 1$ after simplify has completed. (See section XXX for more
+details on simplify.) Step \algref{W7} then completes the algorithm by looping
+over generations; we decrement our clock $t$, and terminate if $t = 0$.
+Otherwise, we update our current population and individual index and loop
+back to \algref{W2}.
 
@@
 
-Besides the cost of transmitting genotypes,
-generating mutations as a separate process after the topological simulation has completed
-avoids generating the many mutations that are lost in the population.
+This algorithm records only the topological information resulting from the
+forwards in time Wright-Fisher process, but it is straightforward to add
+mutational information. This can be done in two different ways.
+We can record mutations that occur during the simulation quite simply.
+For example, in Algorithm~\algref{W} we would generate mutations after
+we have recorded the edges joining the new node $n$ to its parents
+$P_a$ and $P_b$ in step \algref{W7}.
+It is straightforward to record this information during the simulation, but it
+is significantly more efficient to generate these mutations
+after the simulation has completed becuase it
+avoids the cost of generating the many mutations that are lost in the population.
 Figure~\ref{fig:wf-trees} shows
@@
 
-A minimal genealogy such as in Figure~\ref{fig:wf-trees}
-has fewer edges on which to place mutations.
-There are many more advantages to simplifying the genealogies to this point.
-Computing this minimal representation of the
-edges output by a forward-time simulation is an instance of a more
-general problem that we cover in the next section.
-
 %%%%%%
@@
 \item Within the marginal trees, all non-sample vertices must has at least
-two children (i.e., unary tree vertices are removed), unless the vertex is a sample.
+two children (i.e., unary tree vertices are removed).
 \item Any nodes and edges not ancestral to the sampled nodes are removed.
@@
 
-More concretely,
-the algorithm works by moving back through time,
-processing each parent in the input tree sequence in chronological order.
-The main state of the algorithm at each point in time is a set of ancestral lineages,
-and each lineage is a collection of ancestral segments.
-An ancestral segment $(\ell, r, u)$ is found in a lineage
-if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
-% These segments are stored in a collection of linked lists,
-% one list of segments for each ancestral lineage present at that time.
-(These lineages are the ``colors'' above.)
-We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
-the input tree sequence.
-Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.
+\jk{Insert lost text about simplify efficiency and plot}
 
@@
 using Algorithm~\algref{W}
-for $T$ generations,
-with the probability of death, $\delta$, set to 1 for simplicity.
+for $T$ generations and simplify interval $s = 1$ so that redundant information
+is removed after every generation.
 Since there is exactly one crossing over every generation,
@@
 and each time the marginal tree changes along the sequence,
-three edges end and three new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
+four edges end and four new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
 Coalescent theory tells us that
@@
 Not all such recombinations actually change the tree topology,
-and three times this gives us an upper bound on the expected number of edges.
+and four times this gives us an upper bound on the expected number of edges.
 Similarly, not every new edge derives from a never-before-seen node,
@@
 at $N=10^4$, a factor of 4,700.
+\jk{I haven't checked these numbers, just corrected the obvious threes/fours.}
 
@@
 as a function of time (obtained by running simplify every generation)
-is shown in \ref{fig:num_edges}, compared to this prediction.
+is shown in Figure~\ref{fig:num_edges}, compared to this prediction.
 
@@
 To make this feasible,
-we have described how to efficiently store this information in tables,
+we have described how to efficiently store this information in numerical tables,
 and have described a fundamental algorithm for simplificication of tree sequences.
@@
 can happen independently of the details of simulation;
-for this reason, we provide a well-defined and well-tested API in C and in python
-for use in other code bases.
+for this reason, we provide a well-defined and well-tested API in Python
+for use in other code bases (a C API is also planned).
 
@@
 and the fact that these are correlated,
-dynamic programming algorithms similar to those in~\citep{kelleher2016efficient}
+sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
 can do this in roughly $O(n + m \log n)$ operations.
@@
 Thus, by using a simple numerical
-encoding of tree topologies and contiguous arrays of data to store this
-encoding, we can achieve data transfer rates that would be impossible under
+encoding of tree topologies and contiguous arrays of data to store table
+columns, we can achieve data transfer rates that would be impossible under
 any text-based approach while retaining excellent portability.
@@
 
+\jk{Moved from above}
+More concretely,
+the algorithm works by moving back through time,
+processing each parent in the input tree sequence in chronological order.
+The main state of the algorithm at each point in time is a set of ancestral lineages,
+and each lineage is a collection of ancestral segments.
+An ancestral segment $(\ell, r, u)$ is found in a lineage
+if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
+% These segments are stored in a collection of linked lists,
+% one list of segments for each ancestral lineage present at that time.
+(These lineages are the ``colors'' above.)
+We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
+the input tree sequence.
+Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.
+
 %%%%%%%%%%%%%%%%%%%%%%

### 2017-11-22 sims/benchmark-simuPOP.py
Use binary mode in simupop

@@ -236,7 +236,7 @@ haploid_labels = [(k,p) for k in first_gen
 node_ids = {x:j for x, j in zip(haploid_labels, init_ts.samples())}
 rc = RecombCollector(ts=init_ts, node_ids=node_ids,
                      locus_position=locus_position,
-                     benchmark=True)
+                     benchmark=True, mode='binary')
 
 pop.evolve(
     initOps=[
@@ -254,7 +254,8 @@ pop.evolve(
     matingScheme=sim.RandomMating(
         ops=[
             id_tagger,
-            sim.Recombinator(rates=args.recomb_rate, output=rc.collect_recombs,
+            sim.Recombinator(rates=args.recomb_rate,
+                             output=sim.WithMode(rc.collect_recombs, 'b'),
                              infoFields="ind_id"),
         ] ),
     postOps=[

### 2017-11-24 forwards_paper.tex
merged changes from jeromekelleher/edit-pass
merged methods and results; put benchmarks first

@@
 % Please excuse the ragged right margins in the source.
-
-previous \textbf{OUTLINE}, possibly out-of-date but left for comparison.
-\begin{enumerate}
-    \item motivate need for whole-genome fwd-time simulations; point out that we only recently have the computing power to do this
-    \item explain ARG: explain that for forwards-time only need selected loci as by defn all others can be put on afterwards
-    \item review something about msprime methods for storing/traversing tree sequence
-    \item describe tables and write out conditions to have valid tables
-    \item write down algorithm used to do simple WF simulation
-    \item describe simplify algorithm
-    \item back-of-the-envelope calculation to compare cost of tracking whole genomes versus putting mutations on ARG
-    \item comparison of speed with simupop, fwdpp
-\end{enumerate}
 
@@
 
-Below, we first describe the conceptual and algorithmic foundations for the method:
+We begin by benchmarking the performance improvement achieved by connecting
+two previously published forwards-time simulation libraries,
+\simupop{} \citep{simupop} and \fwdpp{} \citep{fwdpp}
+to the tools we describe here, implemented in \msprime{}.
+Then, we describe the conceptual and algorithmic foundations for the method:
 (a) a format, implemented in the \msprime{} Python API,
@@
 and analyze of its run time complexity.
-This is followed by benchmarking the performance improvement achieved by connecting
-two previously published forwards-time simulation libraries,
-\simupop{} \citep{simupop} and \fwdpp{} \citep{fwdpp}
-to the tools in \msprime{}.
+
+%%%%%%
+\subsection*{Simulation benchmarks}
+
+\plr{Jaime and Kevin to describe simulation methods here (separately).}
+
+Comparison of simulation with/without msprime, using \simupop{}
+or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
+
+Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
+so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
+in a day.
 
@@
 she is associated with.
+Note that since each node time is equal to the amount of time since the {birth} of the
+corresponding parent, time is measured in clock time, not in meioses.
 The example of Figure~\ref{fig:example_tree_sequence} has five nodes:
@@
 days of CPU time just to read the information into memory.
+
+\paragraph{Validity of a set of tables}
+Given a set of node and edge tables as described above,
+there are a small set of requirements that ensure the tables
+describe a valid tree sequence.
+% (There are essentially no such requirements on the site and mutation tables.)
+These are:
+\begin{enumerate}
+    \item Offspring must be born after their parents (and hence, no loops).
+    \item The set of intervals on which each individual is a child must be disjoint.
+\end{enumerate}
+A pair of node and edge tables that satisfy these two requirements
+is guarenteed to uniquely describe at each point on the genome
+a collection of directed, acyclic graphs -- in other words, a forest of trees.
+For some applications it is necessary to check that at every point
+there is only a \emph{single} tree.
+Checking this is more difficult, but is implemented in \msprime{}'s API.
+\jda{TODO add function name}
+For efficiency, \msprime{} makes several other sortedness requirements on the tables,
+that are not necessarily satisfied by tables emitted by a forwards-time simulation.
+\msprime{}'s API includes tools to rectify this by first sorting (using \texttt{sort\_tables})
+and then using \texttt{simplify}, which works on sorted tables
+and is guarenteed to produce a valid, \msprime{}-ready tree sequence.
 
@@
 
-\jk{Insert lost text about simplify efficiency and plot}
-
+More concretely,
+the algorithm works by moving back through time,
+processing each parent in the input tree sequence in chronological order.
+The main state of the algorithm at each point in time is a set of ancestral lineages,
+and each lineage is a collection of ancestral segments.
+An ancestral segment $(\ell, r, u)$ is found in a lineage
+if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
+% These segments are stored in a collection of linked lists,
+% one list of segments for each ancestral lineage present at that time.
+(These lineages are the ``colors'' above.)
+We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
+the input tree sequence.
+Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.
+
+%%%%%%%%%%%%%%
+\subsection*{Simplification algorithm}
+
+Here we describe the simplification algorithm.
+The flow is close to our implementation,
+but the description below uses set operations,
+while in the implementation,
+all ancestors are maintained as linked lists of segments
+ordered by left endpoint.
+As it goes, the algorithm
+records a list $M$ that gives the mapping
+from input nodes to output nodes,
+so that if $M[p] = u$ then the $p^\text{th}$ node in the input
+is the same as the $u^\text{th}$ node in the output.
+The current state of the algorithm is recorded in $\Al$,
+a list of ancestors
+each of which is a collections of ancestral segments.
+An ancestral segment is a triple $(\ell, r, x)$,
+indicating that the output node $x$ inherits from this ancestor
+on the genomic segment $[\ell, r)$.
+
+\begin{taocpalg}{S}{Simplify a tree sequence}
+{Input consists of
+    a list $S$ of sample IDs,
+    a list $\Nt_I$ of nodes (ordered by birth time),
+    a list $\Et_I$ of edges,
+    and the genome length, $L$.
+    The output is a
+    nodes $\Nt_O$,
+    and list of edges $\Et_O$.
+}
+
+\algstep{S0.}{Initialize.}{
+    Set $\Al[j] \leftarrow (0, L, j)$
+    and $\Nt_O[j] \leftarrow \Nt_I[S[j]]$ for $0 \le j < \len(S)$.
+    Set the current input parent index to $p=0$
+    and $M[u] \leftarrow -1$ for $0 \le u < \len(\Nt_I)$.
+}
+
+\algstep{S1.}{Input parent loop.}{Set $Q \leftarrow \emptyset$.
+}
+
+\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} on each edge $(\ell, r, p, c)$ in $\Et_I$
+    corresponding to parent $p$ (which stores the resulting segments in the merge queue $Q$).
+}
+
+\algstep{S3.}{Merge ancestors.}{Call Algorithm \algref{M} on $Q$
+    (which adds to $\Nt_O$ and $\Et_O$).
+}
+
+\algstep{S4.}{Next parent.}{Set $p \leftarrow p+1$;
+    if $p = \len(\Nt_I)$, terminate and return $\Nt_O$ and $\Et_O$;
+    otherwise, return to \algref{S1}.
+}
+
+\end{taocpalg}
+
+The algorithm steps back up through the pedigree,
+since nodes are ordered by time-since-birth.
+First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
+so, for every edge $(\ell, r, p, c)$,
+Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
+with $[u,v) \setdiff [\ell, r)$
+(deleting the segment or splitting in two if necessary),
+and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:
+
+\begin{taocpalg}{R}{Remove ancestry}
+    {Given an edge $(\ell, r, p, c)$,
+    and a list $\Al[c]$ of the $m$ ancestral segments corresponding to $c$,
+    remove segments overlapping $[\ell, r)$ from $\Al[c]$,
+    and add those overlapping segments to $Q$.
+    }
+
+    \algstep{R0.}{Initialize.}{
+        Set $j \leftarrow 0$.
+    }
+
+    \algstep{R1.}{Segment loop.}{
+    If $j=m$, terminate; else let $(u, v, x) = \Al[c][j]$.
+    (Here $x$ is the output node that inherits from this segment.)
+    }
+
+    \algstep{R2.}{Add to merge queue.}{If $[u,v) \cap [\ell,r) = \emptyset$,
+    skip to \algref{R4}.
+    Otherwise, append $(\max(u,\ell), \min(v,r), x)$ to $Q$.
+    }
+
+    \algstep{R3.}{Remove ancestry.}{
+    Delete $\Al[c][j]$ and insert in its place
+    $(\min(u,\ell), \max(v,\ell), x)$ (if $u<\ell$)
+    and(/or) $(\min(v,r), \max(v,r), x)$ (if $r<v$).
+    }
+
+    \algstep{R4.}{Iterate.}{Set $j \leftarrow j+1$ and return to \algref{R1}.
+    }
+
+\end{taocpalg}
+
+Next we must merge the ancestry segments in $Q$,
+creating the new ancestor corresponding to input node $p$,
+adding a node and edges to the output if any segments overlap,
+which implies that an output coalescence has occurred.
+Below we index elements of an ancestral segment $z = (\ell, r, u)$
+by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
+and $z.\lid = u$ for the output node ID.
+
+\begin{taocpalg}{M}{Merge ancestors}
+    {Merge the ancestral segments in $Q$,
+    and create the new ancestor $\Al[p]$.
+    This algorithm also appends to $\Nt_O$ and $\Et_O$,
+    and needs to know about $S$,
+    to check if input nodes are samples.
+    }
+
+    \algstep{M0.}{Initialize.}{
+        Set $z \leftarrow (0, 0, -1)$ and $d = \text{False}$.
+    }
+
+    \algstep{M1.}{Segment loop.}{If $\len(Q) = 0$, go to \algref{M9}.}
+
+    \algstep{M2.}{Find next segment overlap boundaries}{
+        Let $\ell_0 \leftarrow \min\{x.\lleft : x \in Q\}$
+        and $H \leftarrow \{x \in Q : x.\lleft = \ell_0\}$.
+        Also let $\ell_+ \leftarrow \min\{x.\lleft : x \in Q \setdiff H\}$
+        and $r = \min(\ell_+, \min\{x.\lright : x \in H\})$.
+    }
+
+    \algstep{M3.}{Copy single segment.}{
+        If $\len(H) = 1$,
+        set $\alpha \leftarrow (\ell, r, H[0].\lid)$.
+        If also $p \in S$, then
+        set $\alpha.\lid \leftarrow M[p]$ and
+        $\add(\Et_O, (\alpha.\lleft, \alpha.\lright, \Nt[p], \alpha.\lid))$.
+        Go to \algref{M7}.
+    }
+
+    \algstep{M4.}{New output node.}{
+        If $M[p] = -1$ set $M[p] \leftarrow \add(\Nt_O, \Nt_I[p])$
+        Set $\alpha \leftarrow (\ell, r, M[p])$.
+    }
+
+    \algstep{M6.}{Record edges and update $Q$.}{
+        For each $x \in H$, do $\add(\Et_O, (\ell, r, M[p], x.\lid))$.
+    }
+
+    \algstep{M7.}{Update $Q$.}{
+        Remove from $Q$ any segment $x \in H$ with $x.\lright = r$
+        and then set $x.\lleft = \ell$ for all remaining $x \in H$.
+    }
+
+    \algstep{M8.}{Add segment to ancestor.}{
+        If $z.\lright = \alpha.\lleft$ and $z.\lid = \alpha.\lid$ set $d = \text{True}$.
+        Append $\alpha$ to $\Al[p]$, set $z \leftarrow \alpha$, and return to \algref{M1}.
+    }
+
+    \algstep{M9.}{Defragment.}{
+        If $d$ is True, find and merge any adjacent segments in $\Al[p]$
+        with the same $\lid$.  Terminate.
+    }
+
+\end{taocpalg}
+
+\jk{Moved from above}
+More concretely,
+the algorithm works by moving back through time,
+processing each parent in the input tree sequence in chronological order.
+The main state of the algorithm at each point in time is a set of ancestral lineages,
+and each lineage is a collection of ancestral segments.
+An ancestral segment $(\ell, r, u)$ is found in a lineage
+if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
+% These segments are stored in a collection of linked lists,
+% one list of segments for each ancestral lineage present at that time.
+(These lineages are the ``colors'' above.)
+We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
+the input tree sequence.
+Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.
+
+%%%%%%%%%%%%%%
 \subsection*{Forward simulation and sequential simplification}
@@
 
-\subsection*{Simulation benchmarks}
-
-Comparison of simulation with/without msprime, using \simupop{}
-or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
-
-Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
-so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
-in a day.
+%%%%%%
+\subsection*{Overview of the API}
+
+The \msprime{} Python API provides a powerful platform for working with tree topology and mutation data.
+The new portions of \msprime{} that we cover here
+are dedicated to tree sequence input and output using simple tables of data,
+so we refer to this as the `Tables API'.
+The four key tables: nodes, edges, sites and mutations, are described above.
+
+The Tables API is primarily designed to facilitate efficient interchange of
+data between programs or between different modules of the same program.
+\jda{should capitalize Tables, or not, as above}
+Following the current best-practises [citations: apache arrow, etc] data is stored
+in a columnar format.
+The principal advantage of this for our purposes is that it allows for very efficient
+interchange of large amounts of numerical data. In principle, this enables
+zero-copy semantics, where a data consumer can read the information directly
+from the memory of a producer without incurring the overhead of a copy
+[citation?] Our implementation uses the numpy C API [citation] to efficiently copy
+data from Python into the low-level C library used to manipulate
+tree sequences.
+Thus, by using a simple numerical
+encoding of tree topologies and contiguous arrays of data to store table
+columns, we can achieve data transfer rates that would be impossible under
+any text-based approach while retaining excellent portability.
+
+The \msprime{} tables API is independent of the details of a particular simulation engine.  In other words, one simply
+needs to fill the relevant columns and send them to \msprime{} at regular intervals for simplification.  For example,
+using the \fwdpp{} C++ API \cite{fwdpp}, we fill arrays of simple structures representing the node and edge data.
+Doing so requires no modification of the \fwdpp{} code base.  Rather, we simply need to book-keep the parent/offspring
+labels and perform simple processing of the recombination breakpoints from each mating event.  Using \texttt{pybind11}
+(\url{https://github.com/pybind/pybind11/}),
+the node/edge arrays on
+the C++ side are made visible, without copy, to Python as a multi-column numpy array. This array is in turn processed by
+the \msprime{} tables API.  In order to create a standalone user interface, our implementation uses \texttt{pybind11} to
+create a Python package extending \texttt{fwdpy11} (\url{https://github.com/molpopgen/fwdpy11}), which is a Python
+package based on \fwdpp{}.  (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
+API from a command-line C/C++ application via an embedded Python interpreter.)
+\plr{@molpopgen: note on how this works with fwdpp?}
+\krt{@plr: done, but perhaps too verbose?}
+
+The tables API provides basic input and output operations via the numpy
+array interface, which provides a great deal of flexibility as well
+as efficiency, and makes it straightforward to transfer data from sources
+such as HDF5 [citation], Dask, or Zarr.
+(For small scale data and debugging purposes, a simple text format is also supported.)
+Along with these operations we also provide a function to sort a set of tables
+to ensure that the records are in the form required to input
+into an \msprime{} tree sequence object. (Simplification may also be required.)
+
+\plr{Maybe we don't need this paragraph? Already said it, more or less, above, esp if we insert something about how fwdpp works.}
+\jda{agree re the next two lines. propose moving the other above}
+This interchange API is very efficient. [Describe a quick example where we generate
+a many-gigabyte tree sequence using fwdpp, and the time required
+to copy the node and edge data into the tables].
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Methods}
-
-\plr{some of this can move back to an appendix, but since they want methods at the end,
-it's the sort of thing they want here.}
-
 %%%%%%%%%%%%%%
-\subsection*{Simuation methods}
-
-Jaime and Kevin to describe simulation methods here (separately).
-
-%%%%%%
-\subsection*{Overview of the API}
-
-The \msprime{} Python API provides a powerful platform for working with tree topology and mutation data.
-The new portions of \msprime{} that we cover here
-are dedicated to tree sequence input and output using simple tables of data,
-so we refer to this as the `Tables API'.
-The four key tables: nodes, edges, sites and mutations, are described above.
-
-The Tables API is primarily designed to facilitate efficient interchange of
-data between programs or between different modules of the same program.
-\jda{should capitalize Tables, or not, as above}
-Following the current best-practises [citations: apache arrow, etc] data is stored
-in a columnar format.
-The principal advantage of this for our purposes is that it allows for very efficient
-interchange of large amounts of numerical data. In principle, this enables
-zero-copy semantics, where a data consumer can read the information directly
-from the memory of a producer without incurring the overhead of a copy
-[citation?] Our implementation uses the numpy C API [citation] to efficiently copy
-data from Python into the low-level C library used to manipulate
-tree sequences.
-Thus, by using a simple numerical
-encoding of tree topologies and contiguous arrays of data to store table
-columns, we can achieve data transfer rates that would be impossible under
-any text-based approach while retaining excellent portability.
-
-The \msprime{} tables API is independent of the details of a particular simulation engine.  In other words, one simply
-needs to fill the relevant columns and send them to \msprime{} at regular intervals for simplification.  For example,
-using the \fwdpp{} C++ API \cite{fwdpp}, we fill arrays of simple structures representing the node and edge data.
-Doing so requires no modification of the \fwdpp{} code base.  Rather, we simply need to book-keep the parent/offspring
-labels and perform simple processing of the recombination breakpoints from each mating event.  Using \texttt{pybind11}
-(\url{https://github.com/pybind/pybind11/}),
-the node/edge arrays on
-the C++ side are made visible, without copy, to Python as a multi-column numpy array. This array is in turn processed by
-the \msprime{} tables API.  In order to create a standalone user interface, our implementation uses \texttt{pybind11} to
-create a Python package extending \texttt{fwdpy11} (\url{https://github.com/molpopgen/fwdpy11}), which is a Python
-package based on \fwdpp{}.  (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
-API from a command-line C/C++ application via an embedded Python interpreter.)
-\plr{@molpopgen: note on how this works with fwdpp?}
-\krt{@plr: done, but perhaps too verbose?}
-
-The tables API provides basic input and output operations via the numpy
-array interface, which provides a great deal of flexibility as well
-as efficiency, and makes it straightforward to transfer data from sources
-such as HDF5 [citation], Dask, or Zarr.
-(For small scale data and debugging purposes, a simple text format is also supported.)
-Along with these operations we also provide a function to sort a set of tables
-to ensure that the records are in the form required to input
-into an \msprime{} tree sequence object. (Simplification may also be required.)
-
-\plr{Maybe we don't need this paragraph? Already said it, more or less, above, esp if we insert something about how fwdpp works.}
-\jda{agree re the next two lines. propose moving the other above}
-This interchange API is very efficient. [Describe a quick example where we generate
-a many-gigabyte tree sequence using fwdpp, and the time required
-to copy the node and edge data into the tables].
-
-%%%%%%%%%%%%%%
-\subsection*{The table format}
-
-Given a set of node and edge tables as described above,
-there are a small set of requirements that ensure the tables
-describe a valid tree sequence.
-(There are essentially no such requirements on the site and mutation tables.)
-These are:
-\begin{enumerate}
-    \item Offspring must be born after their parents (and hence, no loops).
-    \item The set of intervals on which each individual is a child must be disjoint.
-\end{enumerate}
-A pair of node and edge tables that satisfy these two requirements
-is guarenteed to uniquely describe at each point on the genome
-a collection of directed, acyclic graphs -- in other words, a forest of trees.
-For some applications it is necessary to check that at every point
-there is only a \emph{single} tree.
-Checking this is more difficult, but is implemented in \msprime{}'s API.
-\jda{TODO add function name}
-
-For efficiency, \msprime{} makes several other sortedness requirements on the tables,
-that are not necessarily satisfied by tables emitted by a forwards-time simulation.
-\msprime{}'s API includes tools to rectify this by first sorting (using \texttt{sort\_tables})
-and then using \texttt{simplify}, which works on sorted tables
-and is guarenteed to produce a valid, \msprime{}-ready tree sequence.
-
-Note that since each node time is equal to the amount of time since the \emph{birth} of the
-corresponding parent, time is measured in clock time, not in meioses.
-\jda{is this the right place? I'd put it in discussion of Figure~\ref{fig:example_tree_sequence}}
-
-%%%%%%%%%%%%%%
-\subsection*{Simplification}
-
-Here we describe the simplification algorithm.
-The flow is close to our implementation,
-but the description below uses set operations,
-while in the implementation,
-all ancestors are maintained as linked lists of segments
-ordered by left endpoint.
-As it goes, the algorithm
-records a list $M$ that gives the mapping
-from input nodes to output nodes,
-so that if $M[p] = u$ then the $p^\text{th}$ node in the input
-is the same as the $u^\text{th}$ node in the output.
-The current state of the algorithm is recorded in $\Al$,
-a list of ancestors
-each of which is a collections of ancestral segments.
-An ancestral segment is a triple $(\ell, r, x)$,
-indicating that the output node $x$ inherits from this ancestor
-on the genomic segment $[\ell, r)$.
-
-\begin{taocpalg}{S}{Simplify a tree sequence}
-{Input consists of
-    a list $S$ of sample IDs,
-    a list $\Nt_I$ of nodes (ordered by birth time),
-    a list $\Et_I$ of edges,
-    and the genome length, $L$.
-    The output is a
-    nodes $\Nt_O$,
-    and list of edges $\Et_O$.
-}
-
-\algstep{S0.}{Initialize.}{
-    Set $\Al[j] \leftarrow (0, L, j)$
-    and $\Nt_O[j] \leftarrow \Nt_I[S[j]]$ for $0 \le j < \len(S)$.
-    Set the current input parent index to $p=0$
-    and $M[u] \leftarrow -1$ for $0 \le u < \len(\Nt_I)$.
-}
-
-\algstep{S1.}{Input parent loop.}{Set $Q \leftarrow \emptyset$.
-}
-
-\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} on each edge $(\ell, r, p, c)$ in $\Et_I$
-    corresponding to parent $p$ (which stores the resulting segments in the merge queue $Q$).
-}
-
-\algstep{S3.}{Merge ancestors.}{Call Algorithm \algref{M} on $Q$
-    (which adds to $\Nt_O$ and $\Et_O$).
-}
-
-\algstep{S4.}{Next parent.}{Set $p \leftarrow p+1$;
-    if $p = \len(\Nt_I)$, terminate and return $\Nt_O$ and $\Et_O$;
-    otherwise, return to \algref{S1}.
-}
-
-\end{taocpalg}
-
-The algorithm steps back up through the pedigree,
-since nodes are ordered by time-since-birth.
-First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
-so, for every edge $(\ell, r, p, c)$,
-Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
-with $[u,v) \setdiff [\ell, r)$
-(deleting the segment or splitting in two if necessary),
-and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:
-
-\begin{taocpalg}{R}{Remove ancestry}
-    {Given an edge $(\ell, r, p, c)$,
-    and a list $\Al[c]$ of the $m$ ancestral segments corresponding to $c$,
-    remove segments overlapping $[\ell, r)$ from $\Al[c]$,
-    and add those overlapping segments to $Q$.
-    }
-
-    \algstep{R0.}{Initialize.}{
-        Set $j \leftarrow 0$.
-    }
-
-    \algstep{R1.}{Segment loop.}{
-    If $j=m$, terminate; else let $(u, v, x) = \Al[c][j]$.
-    (Here $x$ is the output node that inherits from this segment.)
-    }
-
-    \algstep{R2.}{Add to merge queue.}{If $[u,v) \cap [\ell,r) = \emptyset$,
-    skip to \algref{R4}.
-    Otherwise, append $(\max(u,\ell), \min(v,r), x)$ to $Q$.
-    }
-
-    \algstep{R3.}{Remove ancestry.}{
-    Delete $\Al[c][j]$ and insert in its place
-    $(\min(u,\ell), \max(v,\ell), x)$ (if $u<\ell$)
-    and(/or) $(\min(v,r), \max(v,r), x)$ (if $r<v$).
-    }
-
-    \algstep{R4.}{Iterate.}{Set $j \leftarrow j+1$ and return to \algref{R1}.
-    }
-
-\end{taocpalg}
-
-Next we must merge the ancestry segments in $Q$,
-creating the new ancestor corresponding to input node $p$,
-adding a node and edges to the output if any segments overlap,
-which implies that an output coalescence has occurred.
-Below we index elements of an ancestral segment $z = (\ell, r, u)$
-by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
-and $z.\lid = u$ for the output node ID.
-
-\begin{taocpalg}{M}{Merge ancestors}
-    {Merge the ancestral segments in $Q$,
-    and create the new ancestor $\Al[p]$.
-    This algorithm also appends to $\Nt_O$ and $\Et_O$,
-    and needs to know about $S$,
-    to check if input nodes are samples.
-    }
-
-    \algstep{M0.}{Initialize.}{
-        Set $z \leftarrow (0, 0, -1)$ and $d = \text{False}$.
-    }
-
-    \algstep{M1.}{Segment loop.}{If $\len(Q) = 0$, go to \algref{M9}.}
-
-    \algstep{M2.}{Find next segment overlap boundaries}{
-        Let $\ell_0 \leftarrow \min\{x.\lleft : x \in Q\}$
-        and $H \leftarrow \{x \in Q : x.\lleft = \ell_0\}$.
-        Also let $\ell_+ \leftarrow \min\{x.\lleft : x \in Q \setdiff H\}$
-        and $r = \min(\ell_+, \min\{x.\lright : x \in H\})$.
-    }
-
-    \algstep{M3.}{Copy single segment.}{
-        If $\len(H) = 1$,
-        set $\alpha \leftarrow (\ell, r, H[0].\lid)$.
-        If also $p \in S$, then
-        set $\alpha.\lid \leftarrow M[p]$ and
-        $\add(\Et_O, (\alpha.\lleft, \alpha.\lright, \Nt[p], \alpha.\lid))$.
-        Go to \algref{M7}.
-    }
-
-    \algstep{M4.}{New output node.}{
-        If $M[p] = -1$ set $M[p] \leftarrow \add(\Nt_O, \Nt_I[p])$
-        Set $\alpha \leftarrow (\ell, r, M[p])$.
-    }
-
-    \algstep{M6.}{Record edges and update $Q$.}{
-        For each $x \in H$, do $\add(\Et_O, (\ell, r, M[p], x.\lid))$.
-    }
-
-    \algstep{M7.}{Update $Q$.}{
-        Remove from $Q$ any segment $x \in H$ with $x.\lright = r$
-        and then set $x.\lleft = \ell$ for all remaining $x \in H$.
-    }
-
-    \algstep{M8.}{Add segment to ancestor.}{
-        If $z.\lright = \alpha.\lleft$ and $z.\lid = \alpha.\lid$ set $d = \text{True}$.
-        Append $\alpha$ to $\Al[p]$, set $z \leftarrow \alpha$, and return to \algref{M1}.
-    }
-
-    \algstep{M9.}{Defragment.}{
-        If $d$ is True, find and merge any adjacent segments in $\Al[p]$
-        with the same $\lid$.  Terminate.
-    }
-
-\end{taocpalg}
-
-\jk{Moved from above}
-More concretely,
-the algorithm works by moving back through time,
-processing each parent in the input tree sequence in chronological order.
-The main state of the algorithm at each point in time is a set of ancestral lineages,
-and each lineage is a collection of ancestral segments.
-An ancestral segment $(\ell, r, u)$ is found in a lineage
-if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
-% These segments are stored in a collection of linked lists,
-% one list of segments for each ancestral lineage present at that time.
-(These lineages are the ``colors'' above.)
-We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
-the input tree sequence.
-Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.
+% \section*{Methods}
 

### 2017-11-27 forwards_paper.tex
add methods for benchmarking

@@
 
-\plr{Jaime and Kevin to describe simulation methods here (separately).}
-
-Comparison of simulation with/without msprime, using \simupop{}
-or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
-
-Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
-so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
-in a day.
+To measure the performance gains from ancestry recording, we ran simulations implemented in C++ (see above) with and
+without ancestry tracking.  The simulations without ancestry tracking introduced neutral mutations at a scaled mutation
+rate equal to the scaled recombination rate (\textit{e.g.}, $\theta = 4N\mu = \rho$, where $\rho = 4Nr$, $N$ is the
+diploid population size, $\mu$ is the
+mutation rate to neutral mutations (per gamete, per generation) and $r$ is the recombination rate per diploid per
+generation.  Both $\mu$ and $r$ are the means of a Poisson process and we used the infinitely-many sites mutation model
+\citep{Kimura1969-uc}.
+
+We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$ and ran a single replicate of all $(N,\rho)$
+combinations.  Deleterious mutations were introduced at rate $\rho/100$ per generation, drawing scaled selection
+coefficients ($2Ns$)
+from a Gamma distribution with a mean of -5 and a shape parameter of 1.  This distribution of fitness effects results in
+thousands of weakly-deleterious mutations segregating in the population, many of which drift to intermediate
+frequencies.
+
+We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 Gigahertz Intel E5-2650 CPU with
+hyperthreading enabled.  We ran one simulation at a time and the machine was under minimal load otherwise.  We used GNU
+parallel \citep{Tange2011a} to kill
+any simulation that did not finish within 72 hours.  We used the Linux \texttt{time} command to record run time and peak
+memory usage of each replicate and our simulation scripts separately recorded the time spent in various steps of each
+run.
+
+For efficiency reasons, we only ran \simupop{} with ancestry tracking.
+
+\krt{Jaime--any details here to add re: how simuPOP was tweaked??}
+
+% \plr{Jaime and Kevin to describe simulation methods here (separately).}
+% 
+% Comparison of simulation with/without msprime, using \simupop{}
+% or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
+% 
+% Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
+% so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
+% in a day.
 

### 2017-11-27 forwards_paper.tex
above -> below
address PLR's comment on tracking selected-vs-neutral variants

@@
 
-To measure the performance gains from ancestry recording, we ran simulations implemented in C++ (see above) with and
+To measure the performance gains from ancestry recording, we ran simulations implemented in C++ (see below) with and
 without ancestry tracking.  The simulations without ancestry tracking introduced neutral mutations at a scaled mutation
@@
 thousands of weakly-deleterious mutations segregating in the population, many of which drift to intermediate
-frequencies.
+frequencies.  The case of many mutations with selection is a non-trivial violation of exchangeability assumptions of the
+coalescent \citep{Neuhauser1997-nn}.  Therefore, these selected mutations must be explicitly tracked in our forward simulation
+and the time savings due to ancestry tracking come from not having to record \textit{neutral} mutations.
 

### 2017-11-27 sims/cpp_benchmarks/organize_total_runtimes.py
add in copy/paste of cpp results

@@ -1,0 +1,38 @@
+import sys
+import re
+import pandas as pd
+files = []
+
+for i in sys.argv[1:]:
+    OK = True
+    for line in open(i, 'r'):
+        if re.search('non-zero', line):
+            OK = False
+            break
+    if OK is True:
+        files.append(i)
+
+dflist = []
+for i in files:
+    ARG = True
+    Q = False
+    if re.search('with', i):
+        ARG = False
+    if re.search('queue',i):
+        Q = True
+    d = pd.read_csv(i, sep=" ", names=['time', 'mem'], index_col=False)
+    d['arg'] = [ARG]
+    d['queue'] =  [Q]
+    fields = i.split('.')
+    fields[1] = re.sub('N', '', fields[1])
+    fields[2] = re.sub('size', '', fields[2])
+    d['N'] = float(fields[1])
+    d['rho'] = float(fields[2])
+    dflist.append(d)
+
+df = pd.concat(dflist)
+
+# This is a bit loose:
+df['time_per_gen'] = df.time / (10. * df.N)
+
+df.to_csv("cpp_timings.txt",index=False,sep="\t")

### 2017-11-27 sims/cpp_benchmarks/run_with_neutral_mutations.sh

@@ -1,0 +1,34 @@
+#!/bin/bash
+
+# 72 hour run limit
+SECONDS_TO_KILL=259200
+
+export PYTHONPATH=../.. 
+
+for N in 1000 10000 50000
+do
+    KILL=0
+    for size in 1000 10000 100000
+    do
+        #We will arbitrarily GC every 0.1N generations
+        echo "PARAMS = " $N $size
+        GC=1000
+        TIME_MEM_FILE=time_with_neutral.N$N"."size$size".out"
+        if [ $KILL == 0 ]
+        then
+            DETAILED_TIME_FILE=detailed_time_argi_with_neutral.N$N"."size$size".out.gz"
+            /usr/bin/time -f "%e %M" -o $TIME_MEM_FILE parallel --timeout $SECONDS_TO_KILL PYTHONPATH=../.. python ../../benchmarking.py --popsize $N --neutral_mutations --theta $size --rho $size --nsam 100 --pdel 0.01 --gc $GC --outfile1 $DETAILED_TIME_FILE --seed ::: 42 
+            # If we terminated the job,
+            # then the next params will
+            # also take too long, and 
+            # so we can skip running them
+            STATUS=$?
+            if [ $STATUS -gt 0 ]
+            then
+                echo "STATUS = " $STATUS
+                KILL=1
+            fi
+        fi
+            #/usr/bin/time -f "%e %M" -o $TIME_MEM_FILE timeout $SECONDS_TO_KILL python ../../benchmarking.py --neutral_mutations --popsize $N --theta $size --rho $size --nsam 100 --pdel 0.01 --gc $GC --seed 42
+    done
+done

### 2017-11-27 sims/benchmark-simuPOP.py
Rework to use binary data xfer and sim neutral too
For neutral, determine selected / neutral loci as ratio of mutation
rates times number of loci. Adds timings results for N=100, theta=500

@@ -95,8 +95,8 @@ parser.add_argument("--nsam","-k", type=int, dest="nsamples",
         help="number of *diploid* samples, total",)
 parser.add_argument("--pdel","-p", type=float, dest="pdel",
         help="Ratio of deleterious mutations to neutral mutations",)
-parser.add_argument("--record-neutral", type=bool, dest="record_neutral",
-                    default=False)
+parser.add_argument("--record-neutral", dest="record_neutral",
+                    action='store_true')
 
 parser.add_argument("--gamma_shape","-a", type=float, dest="gamma_shape",
         help="shape parameter in gamma distributed selection coefficients",
@@ -133,8 +133,6 @@ parser.add_argument("--len-trees", "-l", dest="len_trees", action='store_true',
 
 
 args = parser.parse_args()
-if args.record_neutral:
-    sys.exit(1)
 
 if args.generations is None:
     args.generations = args.popsize * args.multiplier
@@ -173,7 +171,6 @@ args.sel_mut_rate = args.mut_rate * args.pdel
 # hard code defaults for simupop:
 # >The default positions are 1, 2, 3, 4, ... on each
 # >chromosome.
-
 locus_position = list(range(0, args.nloci))
 
 # logfile.write("Locus positions:\n")
@@ -197,7 +194,7 @@ class GammaDistributedFitness:
 
     def __call__(self, loc, alleles):
         # because s is assigned for each locus, we need to make sure the
-        # same s is used for fitness of genotypes 01 (1-s) and 11 (1-2s)
+        # same s is used for fitness of genotypes 01 (1-s/2) and 11 (1-s)
         # at each locus
         if loc in self.coefMap:
             s = self.coefMap[loc]
@@ -237,31 +234,43 @@ node_ids = {x:j for x, j in zip(haploid_labels, init_ts.samples())}
 rc = RecombCollector(ts=init_ts, node_ids=node_ids,
                      locus_position=locus_position,
                      benchmark=True, mode='binary')
+pre_ops = [sim.PyOperator(lambda pop: rc.increment_time() or True)]
+mating_ops = [id_tagger]
 
+logfile.write(time.strftime('%X %x %Z')+"\n")
+logfile.write("Started simulating! Generations:\n")
+post_ops = [sim.PyEval(r'"  %d\n" % (gen)', step=100, output=logfile)]
+
[…]

### 2017-11-27 sims/run_with_neutral_variants.sh

@@ -13,10 +13,10 @@ do
             #We will arbitrarily GC every 0.1N generations
             GC=`echo "0.1*$N"|bc -l| sed 's/\.0//'`
             GC=1000
-            TIME_MEM_FILE=time_arg.N$N"."size$size".out"
-            DETAILED_TIME_FILE=simupop_detailed_time_arg.N$N"."size$size".out.gz"
-            LOG_FILE=simupop_log_arg.N$N"."size$size".log"
-            /usr/bin/time -f "%e %M" -o $TIME_MEM_FILE parallel --timeout $SECONDS_TO_KILL python benchmark-simuPOP.py --popsize $N --theta $size --rho $size --nsam 100 --pdel 0.01 --gc $GC --outfile1 $DETAILED_TIME_FILE --logfile $LOG_FILE --seed ::: 42
+            TIME_MEM_FILE=time_neutral.N$N"."size$size".out"
+            DETAILED_TIME_FILE=simupop_detailed_time_neutral.N$N"."size$size".out.gz"
+            LOG_FILE=simupop_log_neutral.N$N"."size$size".log"
+            /usr/bin/time -f "%e %M" -o $TIME_MEM_FILE parallel --timeout $SECONDS_TO_KILL python benchmark-simuPOP.py --popsize $N --theta $size --rho $size --nsam 100 --pdel 0.01 --gc $GC --outfile1 $DETAILED_TIME_FILE --logfile $LOG_FILE --record-neutral --seed ::: 42
             # --seed 43 --nsam 10 --pdel 0.1 --theta 1 --rho 1 -N 100 --outfile1 out.csv.gz
             #/usr/bin/time -f "%e %M" -o $TIME_MEM_FILE parallel --timeout $SECONDS_TO_KILL PYTHONPATH=../.. python ../../benchmarking.py --popsize $N --theta $size --rho $size --nsam 100 --pdel 0.01 --gc $GC --outfile1 $DETAILED_TIME_FILE --seed ::: 42 
             # If we terminated the job,
@@ -26,8 +26,8 @@ do
             STATUS=$?
             if [ $STATUS -gt 0 ]
             then
-                echo "STATUS = " $STATUS
 		# don't do this for now
+                echo "STATUS = " $STATUS
                 KILL=0
             fi
         fi

### 2017-11-27 sims/setup-simuPOP.sh

@@ -4,7 +4,7 @@ conda info -a
 if [[ ! -d "/opt/conda/envs/ftprime-benchmark" ]] ; then
 	conda create -q -n ftprime-benchmark python=3.5
 fi
-wget https://raw.githubusercontent.com/ashander/ftprime/5ad8bcb3a49a19c9cfd52bdf615140af6f83d218/environment.yml
+wget https://raw.githubusercontent.com/ashander/ftprime/0.0.6-rc/environment.yml
 conda env update -q -n ftprime-benchmark -f environment.yml
 source activate ftprime-benchmark
 pip install --editable=git+https://github.com/ashander/ftprime.git@master#egg=ftprime

### 2017-11-27 sims/quick_plot.py
quick figure added

@@ -1,0 +1,51 @@
+import matplotlib.pyplot as plt
+import pandas as pd
+
+spop = pd.read_csv("simupop_timings.txt", sep="\t")
+fp11 = pd.read_csv("cpp_benchmarks/cpp_timings.txt", sep="\t")
+
+spop['engine'] = ['simuPOP'] * len(spop.index)
+fp11['engine'] = ['fwdpy11'] * len(fp11.index)
+
+data = pd.concat([spop, fp11])
+
+lcolors = ['black', 'green', 'blue']
+colors = {}
+i = 0
+for x in sorted(data.N.unique()):
+    colors[x] = lcolors[i]
+    i = i + 1
+
+groups = data.groupby(['engine', 'arg', 'queue', 'N'])
+fig, (ax_fwdpp, ax_fwdpp_arg, ax_simupop) = plt.subplots(
+    3, sharex=True, sharey=True)
+# ax.margins(0.05) # Optional, just adds 5% padding to the autoscaling
+for name, group in groups:
+    lstyle = 'solid'
+    ax = ax_fwdpp
+    if name[0] == 'fwdpy11' and name[1] is True:
+        ax = ax_fwdpp_arg
+    m = 'o'
+    if name[0] == 'simuPOP':
+        ax = ax_simupop
+    mfacecolor = colors[name[3]]
+    if name[2] is True:
+        lstyle = 'dashed'
+        mfacecolor = 'none'
+    popsize = int(name[3])
+    ax.plot(group.rho, group.time / (60.**2.), marker=m,
+            ms=6, linestyle=lstyle, label=r'$N = {}$'.format(popsize),
+            color=colors[name[3]],
+            markerfacecolor=mfacecolor)
+
+ax_fwdpp.legend(loc='best')
+ax_fwdpp.set_title("fwdpy11 with neutral mutations")
+ax_fwdpp_arg.set_title("fwdpy11 with ancestry tracking")
+ax_fwdpp_arg.set_ylabel("Run time (hours)")
+ax_simupop.set_title("simuPOP with ancestry tracking")
+ax_simupop.set_xlabel('Scaled recmbination rate (' + r'$\rho = 4Nr$)')
+ax_simupop.set_xticks([1e3,1e4,1e5])
+# ax.set_yscale("log")
+# plt.legend(bbox_to_anchor=(1.05, 1), loc=2)
+plt.tight_layout()
+plt.savefig("quickfig.png")

### 2017-11-27 sims/quick_plot.py
exploratory plot of memory use

@@ -49,3 +49,36 @@ ax_simupop.set_xticks([1e3,1e4,1e5])
 # plt.legend(bbox_to_anchor=(1.05, 1), loc=2)
 plt.tight_layout()
 plt.savefig("quickfig.png")
+
+fig, (ax_fwdpp, ax_fwdpp_arg, ax_simupop) = plt.subplots(
+    3, sharex=True, sharey=True)
+# ax.margins(0.05) # Optional, just adds 5% padding to the autoscaling
+for name, group in groups:
+    lstyle = 'solid'
+    ax = ax_fwdpp
+    if name[0] == 'fwdpy11' and name[1] is True:
+        ax = ax_fwdpp_arg
+    m = 'o'
+    if name[0] == 'simuPOP':
+        ax = ax_simupop
+    mfacecolor = colors[name[3]]
+    if name[2] is True:
+        lstyle = 'dashed'
+        mfacecolor = 'none'
+    popsize = int(name[3])
+    ax.plot(group.rho, group.mem/(1024.**2.), marker=m,
+            ms=6, linestyle=lstyle, label=r'$N = {}$'.format(popsize),
+            color=colors[name[3]],
+            markerfacecolor=mfacecolor)
+
+ax_fwdpp.legend(loc='best')
+ax_fwdpp.set_title("fwdpy11 with neutral mutations")
+ax_fwdpp_arg.set_title("fwdpy11 with ancestry tracking")
+ax_fwdpp_arg.set_ylabel("Peak RAM use (Gigabytes)")
+ax_simupop.set_title("simuPOP with ancestry tracking")
+ax_simupop.set_xlabel('Scaled recmbination rate (' + r'$\rho = 4Nr$)')
+ax_simupop.set_xticks([1e3,1e4,1e5])
+# ax.set_yscale("log")
+# plt.legend(bbox_to_anchor=(1.05, 1), loc=2)
+plt.tight_layout()
+plt.savefig("quickfig_mem.png")

### 2017-11-27 sims/big_test.sh
Cleanup

@@ -1,2 +1,0 @@
-#!/bin/bash
-python3 benchmark-simuPOP.py --seed 42 --nsam 10 --pdel 0.1 --theta 100 --gc 100 --rho 100 -N 1000 --outfile1 out.csv.gz

### 2017-11-27 sims/run-sim.py

@@ -1,208 +1,0 @@
-#!/usr/bin/env python3.5
-description = '''
-Simulate a chromosome with scattered selected loci using simuPOP,
-postprocessed with msprime to add neutral loci.
-'''
-
-import gzip
-import sys
-from argparse import ArgumentParser
-import math
-import time
-import random
-from ftprime import RecombCollector
-import msprime
-
-parser = ArgumentParser(description=description)
-parser.add_argument("-T","--generations", dest="generations", type=int,
-        help="number of generations to run for")
-parser.add_argument("-N","--popsize", dest="popsize", type=int,
-        help="size of the population", default=100)
-parser.add_argument("-r","--recomb_rate", dest="recomb_rate", type=float,
-        help="recombination rate", default=1e-7)
-parser.add_argument("-L","--length", dest="chrom_length", type=int,
-        help="number of bp in the chromosome", default=100)
-parser.add_argument("-U","--neut_mut_rate", dest="neut_mut_rate", type=float,
-        help="neutral mutation rate", default=1e-7)
-parser.add_argument("-l","--nselloci", dest="nselloci", type=int,
-        help="number of selected loci", default=20)
-parser.add_argument("-u","--sel_mut_rate", dest="sel_mut_rate", type=float,
-        help="mutation rate of selected alleles", default=1e-7)
-parser.add_argument("-a","--gamma_alpha", dest="gamma_alpha", type=float,
-        help="alpha parameter in gamma distributed selection coefficients", default=.23)
-parser.add_argument("-b","--gamma_beta", dest="gamma_beta", type=float, 
-        help="beta parameter in gamma distributed selection coefficients", default=5.34)
-parser.add_argument("-k","--nsamples", dest="nsamples", type=int,
-        help="number of *diploid* samples, total", default=100)
-parser.add_argument("-o","--outfile", dest="outfile", type=str,
-        help="name of output PED file (default: not output)", default=None)
-parser.add_argument("-g","--logfile", dest="logfile", type=str,
-        help="name of log file (or '-' for stdout)", default="-")
-parser.add_argument("-s","--selloci_file", dest="selloci_file", type=str,
[…]

### 2017-11-27 sims/run-simupop.py

@@ -1,170 +1,0 @@
-#!/usr/bin/env python3.5
-description = '''
-Simulate a chromosome with scattered selected loci using simuPOP.
-'''
-
-import gzip
-import sys
-from argparse import ArgumentParser
-import math
-import time
-import random
-
-parser = ArgumentParser(description=description)
-parser.add_argument("-T","--generations", dest="generations", type=int,
-        help="number of generations to run for")
-parser.add_argument("-N","--popsize", dest="popsize", type=int,
-        help="size of the population", default=100)
-parser.add_argument("-r","--recomb_rate", dest="recomb_rate", type=float,
-        help="recombination rate", default=1e-7)
-parser.add_argument("-L","--length", dest="chrom_length", type=int,
-        help="number of bp in the chromosome", default=100)
-parser.add_argument("-U","--neut_mut_rate", dest="neut_mut_rate", type=float,
-        help="neutral mutation rate", default=1e-7)
-parser.add_argument("-l","--nselloci", dest="nselloci", type=int,
-        help="number of selected loci", default=20)
-parser.add_argument("-u","--sel_mut_rate", dest="sel_mut_rate", type=float,
-        help="mutation rate of selected alleles", default=1e-7)
-parser.add_argument("-a","--gamma_alpha", dest="gamma_alpha", type=float,
-        help="alpha parameter in gamma distributed selection coefficients", default=.23)
-parser.add_argument("-b","--gamma_beta", dest="gamma_beta", type=float, 
-        help="beta parameter in gamma distributed selection coefficients", default=5.34)
-parser.add_argument("-k","--nsamples", dest="nsamples", type=int,
-        help="number of *diploid* samples, total", default=100)
-parser.add_argument("-o","--outfile", dest="outfile", type=str,
-        help="name of output PED file (default: not output)", default=None)
-parser.add_argument("-g","--logfile", dest="logfile", type=str,
-        help="name of log file (or '-' for stdout)", default="-")
-parser.add_argument("-s","--selloci_file", dest="selloci_file", type=str,
-        help="name of file to output selected locus information", default="sel_loci.txt")
-
-args = parser.parse_args()
-
-if args.generations is None:
-    parser.print_help()
-    sys.exit()
-
[…]

### 2017-11-27 sims/test.sh

@@ -1,2 +1,0 @@
-#!/bin/bash
-python3 benchmark-simuPOP.py --seed 43 --nsam 10 --pdel 0.1 --theta 1 --rho 1 -N 100 --outfile1 out.csv.gz

### 2017-11-28 sims/simplify-alg.py
Updated algorithm to work locus-by-locus.

@@ -5,6 +5,19 @@ import collections
 import msprime
 import numpy as np
 
+
+class Edge(object):
+    def __init__(self, left, right, parent, child):
+        self.left = left
+        self.right = right
+        self.parent= parent
+        self.child = child
+
+    def __repr__(self):
+        return repr([self.left, self.right, self.parent, self.child])
+
+
+
 def simplify(S, Ni, Ei, L):
     No = msprime.NodeTable()
     Eo = msprime.EdgeTable()
@@ -13,66 +26,105 @@ def simplify(S, Ni, Ei, L):
     child = Ei.child
     parent = Ei.parent
     time = Ni.time
-    A = collections.defaultdict(set)
-    M = {}
+    A = np.zeros((len(Ni), L), dtype=int) - 1
+    M = np.zeros(len(Ni), dtype=int) - 1
     for input_id in S:
         # TODO update node table to return ID of new row.
         output_id = len(No)
-        No.add_row(time=time[input_id])
-        M[input_id] = output_id
-        # TODO We probably need to make this a set of tuples, otherwise we lose the
-        # ability to track where individual bits have come from.
-        A[output_id] = set(range(L - 1))
-    for k, v in A.items():
-        print(k, "->", v)
+        No.add_row(time=time[input_id], flags=1)
+        A[output_id, :] = input_id
 
-    # This isn't quite right as we've lost the ability to map nodes. Seems like
-    # it's OK otherwise.
+    E = []
 
     for input_parent in range(len(Ni)):
         index = parent == input_parent
-        print("parent = ", input_parent)
+        output_id = -1
         S = []
         for l, r, c in zip(left[index], right[index], child[index]):
-            # Remove ancestry is now trivial; directly expressed as set operations.
-            x = set(range(l, r))
-            S.append(A[c] & x)
-            A[c] -= x
+            # For each locus, transfer any genetic material to the parent.
+            P = np.zeros(L, dtype=int) - 1
+            P[l: r] = A[c, l: r]
+            A[c, l: r] = -1
+            # Store the transferred genetic material for all parents in the
+            # list S so that we can find any coalescences afterwards.
+            S.append(P)
         if len(S) > 0:
-            print("S = ", S)
-            A[input_parent] = S[0].union(*S[1:])
-            for k, v in A.items():
-                if len(v) > 0:
-                    print(k, "->", v)
-            inter = S[0].intersection(*S[1:])
[…]

### 2017-12-01 forwards_paper.tex
Reorg benchmarks section; add simuPOP

--- 2017-11-27 forwards_paper.tex
@@
 \newcommand{\msprime}{\texttt{msprime}}
+\newcommand{\ftprime}{\texttt{ftprime}}
 \newcommand{\nodes}{\texttt{nodes}}
@@
 \simupop{} \citep{simupop} and \fwdpp{} \citep{fwdpp}
-to the tools we describe here, implemented in \msprime{}.
+to the tools we describe here, implemented in \msprime{}, allowing them to
+efficiently record the population pedigree in a simplified tree sequence.
 Then, we describe the conceptual and algorithmic foundations for the method:
@@
 and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
-We then describe a general-purpose method that uses these tools
-(as implemented in \msprime{})
-to efficiently record simplified a tree sequence,
-and analyze of its run time complexity.
+Finally, we analyze the run time and space complexity of our general-purpose method.
 
@@
 
-To measure the performance gains from ancestry recording, we ran simulations implemented in C++ (see above) with and
-without ancestry tracking.  The simulations without ancestry tracking introduced neutral mutations at a scaled mutation
-rate equal to the scaled recombination rate (\textit{e.g.}, $\theta = 4N\mu = \rho$, where $\rho = 4Nr$, $N$ is the
-diploid population size, $\mu$ is the
+To measure the performance gains from recording the pedigree we ran simulations with and
+without recording, for both implementations (\simupop{} \citep{simupop}
+and \fwdpp{} \citep{fwdpp}).
+All simulations implemented a discrete-time Wright-Fisher population of diploid
+population size $N$, simulated for $10N$ generations.
+The simulations without pedigree recording introduced neutral mutations at a scaled mutation
+rate equal to the scaled recombination rate (\textit{e.g.}, $\theta = 4N\mu = \rho$, where $\rho = 4Nr$, $\mu$ is the
 mutation rate to neutral mutations (per gamete, per generation) and $r$ is the recombination rate per diploid per
-generation.  Both $\mu$ and $r$ are the means of a Poisson process and we used the infinitely-many sites mutation model
-\citep{Kimura1969-uc}.
-
-We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$ and ran a single replicate of all $(N,\rho)$
-combinations.  Deleterious mutations were introduced at rate $\rho/100$ per generation, drawing scaled selection
+generation.
+Deleterious mutations were introduced at rate $\rho/100$ per generation, drawing scaled selection
 coefficients ($2Ns$)
@@
 thousands of weakly-deleterious mutations segregating in the population, many of which drift to intermediate
-frequencies.
+frequencies.  The case of many mutations with selection is a non-trivial violation of exchangeability assumptions of the
+coalescent \citep{Neuhauser1997-nn}.  Therefore, these selected mutations must be explicitly tracked in our forward simulation
+and the time savings due to pedigree recording come from not having to record \textit{neutral} mutations.
 
@@
 run.
-
-For efficiency reasons, we only ran \simupop{} with ancestry tracking.
-
-\krt{Jaime--any details here to add re: how simuPOP was tweaked??}
-
-% \plr{Jaime and Kevin to describe simulation methods here (separately).}
-% 
+We varied $N$ and $\rho$ and ran a single replicate of all $(N,\rho)$
+combinations (the actual values differed for \simupop{} and \fwdpp{}, see below).
+
+\subsubsection*{The \fwdpp{} implementation}
+
+We implemented Wright-Fisher simulations in C++ using \fwdpp{} library
+functions and interface code that records data into the \msprime{} Tables API (see below).
+In these simulations, $\mu$ and $r$ are the means of a Poisson process and we used the infinitely-many sites mutation model
+\citep{Kimura1969-uc}.
+We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.
+
+\subsubsection*{The \simupop{} implementation}
+
+We implemented Wright-Fisher simulations in Python using \simupop{} library
+functions (which wrap C++ code) and interface code (\ftprime{}, also in Python) that records data into the \msprime{} Tables API (see below).
+In these simulations, $\mu$ and $r$ define a per-base-pair mutation rate and we
+defined the length of the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.
+We varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.
+
 % Comparison of simulation with/without msprime, using \simupop{}

### 2017-12-02 forwards_paper.tex
Mention forwsim and anafits

@@
 at each position along the genome.
+This basic insight has already led development of specialized forward simulators
+(e.g., ~\citet{forwsim}, ``Ancestry aware forward in time simulator" (anafits))
+that track only recombination and neutral haplotypes.
+
 In this paper, we describe how algorithmic tools and data structures developed for the
 coalescent simulator \msprime{}
-can be used to efficiently record, and later process, this history.
+can be used to efficiently record, and later process, this history with
+\emph{any} forward-time simulator.
 In so doing we record the \emph{population pedigree} --

### 2017-12-09 forwards_paper.tex
Peter edited through

--- 2017-11-27 forwards_paper.tex
@@
 \begin{abstract}
-    The computational burden of tracking individual genomes
-    in a population genetics simulation can be substantial.
-    In this paper we describe how to dramatically reduce this burden by
-    efficiently recording the entire history of the population.
-    We do this by simulating only those loci that may affect reproduction (those having non-neutral variants),
-    and recording the entire history of genetic inheritance in an efficient representation of the ancestral recombination graph,
-    on which neutral mutations can be quickly placed afterwards ---
-    the \emph{tree sequence} introduced in the software package \msprime.
-    We implement the method in two popular fowards-time simulation frameworks,
-    and show it leads to order-of-magnitude speedups.
+    In this paper we describe how to 
+    efficiently record the entire genetic history of a population
+    during a forwards-time, individual-based population genetics simulation.
+    This allows us to dramatically reduce the computational burden of tracking individual genomes
+    by simulating only those loci that may affect reproduction (those having non-neutral variants),
+    and recording history in the \emph{tree sequence} format introduced in the software package \msprime,
+    on which neutral mutations can be quickly placed afterwards.
+    We implement the method in two popular forwards-time simulation frameworks,
+    and show that not only does this speed up large simulations by one or two orders of magnitude,
+    but saves substantially more information from the simulation
+    (all marginal genealogies).
     % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
@@
 first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
-second, producing analytical methods to estimate parameters of interest from genetic data (e.g.\ $\theta = 4N_e \mu$);
+second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
 and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
@@
 Furthermore, the relationship between the life history of a species --
-fecundity and mortality schedules, allee effects, and demographic fluctuations --
+fecundity and mortality schedules, Allee effects, and demographic fluctuations --
 are all absorbed into a single compound parameter, the coalescence rate.
@@
 These considerations, and increasing computational power, have led to a resurgence of
-interest in large scale forwards-time, individual-based simulations.
+interest in large forwards-time, individual-based simulations.
 For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
@@
 \cite{Sanjak2017-ko} used
-fwdpp \citep{fwdpp} to simulate a series of models of quantiative traits under mutation-selection balance with
+fwdpp \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
 population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\approx 5
@@
 Modern computing power easily allows simulations of birth, death and reproduction
-in a population having even hundreds of millions of individuals,
-even though theory tells us that a population of size $N$
-must be run for many multiples of $N$ generations to produce stable genetic patterns.
+in a population having even hundreds of millions of individuals.
 However, if our interest lies in the resulting genetic patterns of variation
@@
 becomes quickly prohibitive.
+To make matters worse, 
+a population of size $N$ must be simulated across many multiples of $N$ generations
+to produce stable genetic patterns.
 Because of this computational burden of neutral mutations, even the fastest simulation frameworks such as SLiM,
@@
 weeks to simulate models without selection~\citep{fwdpp,Hernandez2015-wf},
-and different simulation engines differ in how efficiently they
+and existing simulation engines differ in how efficiently they
 calculate fitnesses in models with selection~\citep{fwdpp}.
@@
 to construct the genealogical tree that relates each individual to each other
-at each position on the genome, i.e., the \emph{tree
-sequence}~\citep{kelleher2016efficient}.
+at each position on the genome, i.e., the \emph{tree sequence}~\citep{kelleher2016efficient}.
 Combined with ancestral genotypes and the origins of new mutations,
@@
 Another way of representing this same information
-is known as the \emph{ancestral recombination graph} ---
+is known as the \emph{ancestral recombination graph} --
 the {ARG} \citep{griffiths1997ancestral},
-where nodes are common ancestor or recombination events ---
+where nodes are common ancestor or recombination events --
 which has been the subject of substantial study
@@
 theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
-
-\jk{And finish with a more detailed descrition of what we do in the paper}
+We use tree sequences for the substantial algorithmic advantages exploited in~\citep{kelleher2016efficient}.
+
 In this paper we discuss storage methods for genealogies (and hence, genome sequence),
@@
 and their use in forwards-time simulation.
-Although we were motivated by a need for more efficient simulations,
+Although we were motivated by a need for more efficient genomic simulations,
 these tools may prove more widely useful.
@@
 two previously published forwards-time simulation libraries,
-\simupop{} \citep{simupop} and \fwdpp{} \citep{fwdpp}
-to the tools we describe here, implemented in \msprime{}.
+\simupop{} \citep{peng2005simupop} and \fwdpp{} \citep{fwdpp}
+to the tools we describe here, as implemented in \msprime{}.
 Then, we describe the conceptual and algorithmic foundations for the method:
@@
 for recording tree sequences in several \emph{tables};
-(b) an algorithm, \algref{W}, to record tree sequence information into these tables on the fly
+(b) an algorithm to record tree sequence information into these tables on the fly
     from a forwards-time simulation;
@@
 to efficiently record simplified a tree sequence,
-and analyze of its run time complexity.
+and analyze its run time complexity.
 
@@
 
-To measure the performance gains from ancestry recording, we ran simulations implemented in C++ (see below) with and
-without ancestry tracking.  The simulations without ancestry tracking introduced neutral mutations at a scaled mutation
-rate equal to the scaled recombination rate (\textit{e.g.}, $\theta = 4N\mu = \rho$, where $\rho = 4Nr$, $N$ is the
-diploid population size, $\mu$ is the
-mutation rate to neutral mutations (per gamete, per generation) and $r$ is the recombination rate per diploid per
-generation.  Both $\mu$ and $r$ are the means of a Poisson process and we used the infinitely-many sites mutation model
-\citep{Kimura1969-uc}.
-
-We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$ and ran a single replicate of all $(N,\rho)$
-combinations.  Deleterious mutations were introduced at rate $\rho/100$ per generation, drawing scaled selection
-coefficients ($2Ns$)
-from a Gamma distribution with a mean of -5 and a shape parameter of 1.  This distribution of fitness effects results in
-thousands of weakly-deleterious mutations segregating in the population, many of which drift to intermediate
-frequencies.  The case of many mutations with selection is a non-trivial violation of exchangeability assumptions of the
-coalescent \citep{Neuhauser1997-nn}.  Therefore, these selected mutations must be explicitly tracked in our forward simulation
+To measure the performance gains from ancestry recording, we ran simulations implemented in C++ (see below) with and without ancestry tracking.  
+The simulations without ancestry tracking introduced neutral mutations at a rate 
+equal to the recombination rate,
+so $\mu = r$, where $\mu$ and $r$ are the expected number of mutations and recombinations,
+respectively, per gamete and per generation.
+Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,
+and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
+We categorize simulations in terms of the \emph{scaled} parameters
+$\theta = 4N\mu$ and $\rho = 4Nr$, where $N$ is the diploid population size,
+since these most strongly determine run time.
+
+We ran one simulation for each combination of $N$ from $(1000, 10000, 50000)$, and $\rho$ from $(10^3, 10^4, 10^5)$.
+Deleterious mutations were introduced at rate $\rho/100$ per generation, 
+drawing scaled selection coefficients ($2Ns$)
+from a Gamma distribution with a mean of -5 and a shape parameter of 1.  
+This distribution of fitness effects results in
+thousands of weakly-deleterious mutations segregating in the population, 
+many of which drift to intermediate frequencies.  
+The case of many mutations with selection is a non-trivial violation of exchangeability assumptions of the
+coalescent \citep{Neuhauser1997-nn}.
+Therefore, these selected mutations must be explicitly tracked in our forward simulation,
 and the time savings due to ancestry tracking come from not having to record \textit{neutral} mutations.
@@
 We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 Gigahertz Intel E5-2650 CPU with
-hyperthreading enabled.  We ran one simulation at a time and the machine was under minimal load otherwise.  We used GNU
-parallel \citep{Tange2011a} to kill
-any simulation that did not finish within 72 hours.  We used the Linux \texttt{time} command to record run time and peak
-memory usage of each replicate and our simulation scripts separately recorded the time spent in various steps of each
-run.
+hyperthreading enabled.  
+We ran one simulation at a time and the machine was under minimal load otherwise.  
+We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
+and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
+Our simulation scripts separately recorded the time spent in various steps of each run.
 
@@
 \krt{Jaime--any details here to add re: how simuPOP was tweaked??}
+
+\plr{Insert figures/tables of performance here and describe them.}
 
@@
 
-A \emph{tree sequence} is an encoding for a sequence of correlated trees,
+The simulations above recorded information about each new individual 
+in collection of tables that together encode a \emph{tree sequence}.
+A {tree sequence} is an encoding for a sequence of correlated trees,
 such as those describing the history of a sexual population.
@@
 while nodes 3 and 4 represent those ancestors necessary to record their genealogy,
-who were born at time 1 and 2 respectively.
+who were born one and two units of time in the past, respectively.
 
@@
         defined by \msprime.
-        \plr{Whoops!  genotype of 1 in the second tree should be `G'!}
+        \plr{Whoops!  Genotypes in the second tree should be `GGC'}
         \jk{Some minor alignment issues present here.}
@@
 are shown in the bottom left panel.
-For example, the branch joining nodes 1 to 3 appears both trees,
+For example, the branch joining nodes 1 to 3 appears in both trees,
 and so is recorded as a single edge extending over the whole chromosome.
@@
 each point along the genome that the tree topology changes
-is accompanied by the end of some \emph{edges} and the beginning of others.
-Since each \emph{edge} records the genomic interval
+is accompanied by the end of some {edges} and the beginning of others.
+Since each {edge} records the genomic interval
 over which a given node inherits from a particular ancestor,
@@
 between nearby trees (e.g., as ``subtree prune and regraft'' moves)
-allows for efficient algorithms that take advantage
+allows for efficient algorithms to compute statistics of the genome sequence that take advantage
 of the highly correlated nature of nearby trees~\citep{kelleher2016efficient}.
@@
 and results in a specific derived state.
-Thus, all samples below the mutation node in the tree will inherit this state
-(unless further mutations are encountered).
+Thus, all samples below the mutation's node in the tree will inherit this state,
+unless further mutations are encountered.
 Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
 illustrated by red hexagons.
-The first mutation occurs at site zero (in the left-hand tree), and is a simple
-mutation resulting in node $2$ inheriting the state `T'.
-The second site (in the right hand tree) has two mutations:
+The first site, in the left-hand tree,
+has a single mutation, which results in node $2$ inheriting the state `T'.
+The second site, in the right hand tree, has two mutations:
 one occurring over node $3$ changing the state to `C',
@@
 Given a set of node and edge tables as described above,
-there are a small set of requirements that ensure the tables
+there are only two requirements that ensure the tables
 describe a valid tree sequence.
@@
 A pair of node and edge tables that satisfy these two requirements
-is guarenteed to uniquely describe at each point on the genome
+is guaranteed to uniquely describe at each point on the genome
 a collection of directed, acyclic graphs -- in other words, a forest of trees.
@@
 Checking this is more difficult, but is implemented in \msprime{}'s API.
-\jda{TODO add function name}
+% \jda{TODO add function name}  PLR: not sure this is necessary.
 For efficiency, \msprime{} makes several other sortedness requirements on the tables,
 that are not necessarily satisfied by tables emitted by a forwards-time simulation.
-\msprime{}'s API includes tools to rectify this by first sorting (using \texttt{sort\_tables})
-and then using \texttt{simplify}, which works on sorted tables
-and is guarenteed to produce a valid, \msprime{}-ready tree sequence.
+\msprime{}'s API includes tools to rectify this by first sorting % (using \texttt{sort\_tables})
+and then using the \texttt{simplify} algorithm described below, which works on sorted tables
+and is guaranteed to produce a valid, \msprime{}-ready tree sequence.
 
@@
 which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
-For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
-(but with overlapping generations) that records genealogical history in this way.
+For concreteness, we write out in pseudocode how to run a neutral Wright--Fisher simulation
+that records genealogical history in this way.
 The simulation will run for $T$ generations,
-and has $N$ haploid individuals, each carrying a single chromosome of length L.
-For simplicity we sample exactly one crossover per generation.
-The probability of death per individual each generation is $\delta$.
-We will use
-$\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random
-(and each such instance represents and independent draw). Given a node table
-$\Nt$, the function $\taddnode{\Nt, t}$ adds a new node to the table with time $t$
-and returns the ID of this new node. Similarly, the function $\taddedge{\Et,
-\ell, r, p, c}$ adds a new edge $(\ell, r, p, c)$ to the edge table $\Et$.
-The function $\tsimplify{P, \Nt, \Et}$ simplifies the history stored in the
-tables $\Nt$ and $\Et$ to the minimal information required to represent the
-genealogies of the list of node IDs $P$ (see section XXX).
-We first list
-the algorithm in detail, and immediately follow this with a step-by-step
-explanation.
+and has $N$ haploid individuals, each carrying a single chromosome of length $L$.
+For simplicity, we sample exactly one crossover per generation.
+
+We use $\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random.
+% (and all such instances is independent).  % this is understood here
+Given a node table $\Nt$, the function $\taddnode{\Nt, t}$ 
+adds a new node to the table $\Nt$ with time $t$
+and returns the ID of this new node. 
+Similarly, the function $\taddedge{\Et, \ell, r, p, c}$ 
+adds a new edge $(\ell, r, p, c)$ to the edge table $\Et$.
+The function $\tsimplify{P, \Nt, \Et}$ (described in section \ref{ss:simplify})
+simplifies the history stored in the tables $\Nt$ and $\Et$ 
+to the minimal information required to represent the genealogies of the list of node IDs $P$;
+after simplification the nodes appearing in $P$ are relabeled $(0, 1, \ldots, |P|)$.
+A step-by-step explanation follows the pseudocode.
 
@@
 and edge tables ($\Nt$ and $\Et$) recording the simulated history.
-Simplify is run every $s$ generations, removing genealogical
+The tables are simplified every $s$ generations, removing genealogical
 information from $\Nt$ and $\Et$ unreachable from the current population $P$.
@@
 Set $\Nt \leftarrow \mbox{NodeTable}()$, $\mathcal{E}
-\leftarrow \mbox{EdgeTable}()$ and $t \leftarrow T$.
- For $0 \leq j < N$, set $P_j \leftarrow \taddnode{\Nt, T}$. Set $j \leftarrow 0$.
-}
-
-\algstep{W2.}{New node.}{Set $u \leftarrow \taddnode{\Nt, t}$ and $P'_j \leftarrow u$.
+\leftarrow \mbox{EdgeTable}()$, $t \leftarrow T$, and $j \leftarrow 0$.
+ For $0 \leq k < N$, set $P_k \leftarrow \taddnode{\Nt, T}$. 
+}
+
+\algstep{W2.}{Generation loop head: new node.}{Set $u \leftarrow \taddnode{\Nt, t}$ and $P'_j \leftarrow u$.
 }
@@
 
-\algstep{W6.}{Simplify.}{Call $\tsimplify{P', \Nt, \Et}$, and set $P'_j
-\leftarrow j $ for $0 \leq j < N$. (Tables may need to be sorted.)}
+\algstep{W6.}{Simplify.}{Call $\tsimplify{P', \Nt, \Et}$, and set $P'_k
+\leftarrow k $ for $0 \leq k < N$. (Tables may need to be sorted.)}
 
@@
 $P_j$ by first creating a new node with birth time $t$ (and ID $u$).
-Then, in step~\algref{W3} we choose two indexes $a$ and $b$ uniformly (giving us
-parental IDs $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
-event by storing two new edges: one recording that the parent of node $u$
+In step~\algref{W3} we choose two indexes $a$ and $b$ uniformly, 
+giving us parental IDs $P_a$ and $P_b$, 
+and choose a chromosomal breakpoint $x$. 
+We record the effects of this event by storing two new edges: one recording that the parent of node $u$
 from $0$ to $x$ is $P_a$, and another recording that the parent of $u$
-from $x$ to $L$ is $P_b$. Step \algref{W5} then iterates over each of the
-$N$ individuals for each generation. At the end of a generation, we then check
-if we need to simplify (which is done every $s$ generations). If simplification
-is required, we do this in step \algref{W6} by calling the simplify function
-on the node and edge tables with the current set of population IDs $P'$ as
-the samples. This updates the tables in-place to remove all redundant
-information, and remaps the specified sample IDs to $0, \dots N - 1$
-in the updated tables. Hence, we set our current population IDs to
-$0, \dots N - 1$ after simplify has completed. (See section XXX for more
-details on simplify.) Step \algref{W7} then completes the algorithm by looping
-over generations; we decrement our clock $t$, and terminate if $t = 0$.
-Otherwise, we update our current population and individual index and loop
-back to \algref{W2}.
+from $x$ to $L$ is $P_b$. Step \algref{W5} then iterates these steps
+for each of the $N$ individuals for each generation. 
+At the end of a generation, we then check
+if we need to simplify (which is done every $s$ generations). 
+If simplification is required, we do this in step \algref{W6} by calling the simplify function
+on the node and edge tables with the current set of population IDs $P'$ as the samples. 
+This updates the tables in-place to remove all redundant
+information, and remaps the specified sample IDs to $0, \dots, N - 1$ in the updated tables. 
+Hence, we set our current population IDs to
+$0, \dots N - 1$ after simplify has completed.  
+Step \algref{W7} loops these steps until the required number of generations have been simulated.
+% then completes the algorithm by looping over generations; 
+% we decrement our clock $t$, and terminate if $t = 0$.
+% Otherwise, we update our current population and individual index and loop
+% back to \algref{W2}.
 
@@
 
-This algorithm records only the topological information resulting from the
-forwards in time Wright-Fisher process, but it is straightforward to add
-mutational information. This can be done in two different ways.
-We can record mutations that occur during the simulation quite simply.
-For example, in Algorithm~\algref{W} we would generate mutations after
-we have recorded the edges joining the new node $n$ to its parents
+This algorithm records only topological information about the simulated genealogies,
+but it is straightforward to add mutational information. 
+This can be done in two different ways.
+We can record mutations that occur during the simulation by simply
+randomly sampling new mutations after recording the edges joining the new node $u$ to its parents
 $P_a$ and $P_b$ in step \algref{W7}.
 It is straightforward to record this information during the simulation, but it
-is significantly more efficient to generate these mutations
-after the simulation has completed becuase it
-avoids the cost of generating the many mutations that are lost in the population.
+Alternatively, we can generate these mutations after the simulation has completed, thus
+avoiding the cost of generating the many mutations that are lost in the population.
+This is straightforward to do because we have access to the marginal genealogies.
+
 Figure~\ref{fig:wf-trees} shows
@@
 while on the right
-is the minimal tree representing the ancestry of the currently alive
-population. Clearly there is a great deal of redundancy in the topological
+is the minimal tree representing the ancestry of the currently alive population. 
+Clearly there is a great deal of redundancy in the topological
 information output by the simulation.
-
-There are two sources of redundancy here. The first type of redundancy arises
-from nodes in tree that have only one child. In Algorithm~\algref{W} we do
-not attempt to track coalescence events but simply record all parent-child
-relationships in the history of the population. As such, many of these edges
+This redundancy comes from two sources.
+First, there are a large number of nodes in tree that have only one child. 
+In Algorithm~\algref{W} we do not attempt to identify coalescence events,
+but simply record all parent-child
+relationships in the history of the population. 
+As such, many of these edges
 will record the simple passing of genealogical information from parent to child
-and only some small subset will correspond to coalesences within the marginal
+and only some small subset will correspond to coalescences within the marginal
 trees. The second source of redundancy in the output of Algorithm~\algref{W}
@@
 \subsection*{Tree sequence simplification}
-
-Suppose that we are only interested in a subset of the nodes (our `samples')
-of a tree sequence,
-and so wish to reduce this input tree sequence
-into the minimum representation of the topologies that include the specified
-samples. The output tree sequence must have the following properties:
+\label{ss:simplify}
+
+It is desirable for many reasons to remove redundant information from a tree sequence,
+such as those produced by a forwards-time simulation.
+To formalize this: 
+suppose that we are only interested in a subset of the nodes of a tree sequence
+(which we refer to as our `samples'),
+and wish to reduce this input tree sequence
+into the minimum representation of the topologies that include the specified samples. 
+The output tree sequence must have the following properties:
 \begin{enumerate}
@@
 two children (i.e., unary tree vertices are removed).
-\item Any nodes and edges not ancestral to the sampled nodes are removed.
-\item There are no adjacent redundant edges, i.e, pairs of edges $(\ell, x, p,
-c)$ and $(x, r, p, c)$ which can be represented with a single edge
+\item Any nodes and edges not ancestral to any of the sampled nodes are removed.
+\item There are no adjacent redundant edges, i.e., pairs of edges 
+    $(\ell, x, p, c)$ and $(x, r, p, c)$ which can be represented with a single edge
     $(\ell, r, p, c)$.
@@
 
-The tree sequences produced by forwards simulations
-record all of history for everyone alive at any time through the simulation.
-This is much more than we need to reconstruct the genealogies and sequences
-of the current population.
-Simplification is essential to reduce this
-to a manageable quantity that still contains all
-the information that we are interested in.
-Simplification is also useful if we have a
-large tree sequence representing a large dataset and we wish to extract the
-information relevant to a subset of the samples.
-
-The approach that we take is based on Hudson's algorithm for simulating
-the coalescent with recombination~\citep{hudson1983properties,kelleher2016efficient};
-the implementaiton of simplify parallels the implementation of Hudson's algorithm in \msprime.
+Simplification is essential 
+not only for keeping the information recorded by forwards simulation manageable,
+but also is useful for extracting subsets of a tree sequence representing a very large dataset.
+% The tree sequences produced by forwards simulations
+% record all of history for everyone alive at any time through the simulation.
+% Simplification is essential to reduce this
+% to a manageable quantity that still contains all
+% the information that we are interested in.
+% Simplification is also useful if we have a
+% tree sequence representing a large dataset and wish to extract the
+% information relevant to a subset of the samples.
+
+Our approach to simplification is based on Hudson's algorithm for simulating
+the coalescent with recombination~\citep{hudson1983properties},
+paralleling the implementation in \msprime{} \citep{kelleher2016efficient};
+an implementation in pseudocode is provided in Appendix~\ref{ss:simplify_algorithm}.
 Conceptually, this works by
@@
 but is reasonably close.
-The process is depicted in Figures~\ref{fig:method_diagram} and \ref{fig:simplify_state}.
+Figure~\ref{fig:method_diagram} shows an example tree sequence,
+before and after simplification,
+and Figure~\ref{fig:simplify_state} depicts the ``paint pot'' state of the algorithm 
+during the process of simplifying this tree sequence.
 
@@
 The main state of the algorithm at each point in time is a set of ancestral lineages,
-and each lineage is a collection of ancestral segments.
+and each lineage is a linked list of ancestral segments.
 An ancestral segment $(\ell, r, u)$ is found in a lineage
-if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
+if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage
+(and so $u$ corresponds to a ``color'' in the description above).
 % These segments are stored in a collection of linked lists,
 % one list of segments for each ancestral lineage present at that time.
-(These lineages are the ``colors'' above.)
-We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
-the input tree sequence.
-Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.
+We also maintain a map $A$ such that $A_j$ is the lineage
+corresponding to node $j$ in the input tree sequence.
+Crucially, the time required to run the algorithm is 
+linear in the number of edges of the input tree sequence.
 
 %%%%%%%%%%%%%%
-\subsection*{Simplification algorithm}
+\subsection*{Sequential simplification and prior history}
+
+Any simulation scheme that records data into tables, 
+as does Algorithm~\algref{W},
+has its genealogical history available at any time as a tree sequence.
+This has two additional advantages:
+First, simplification can be run periodically through the simulation, 
+if we take the set of samples to be the entire currently alive population. 
+This is important in practice as it keeps memory usage from growing linearly (and quickly) with time. 
+Second, the simulation can
+be \emph{begun} with a tree sequence produced by some other method -- for
+instance, by a coalescent simulation with \msprime,
+providing an easy, efficient way to specify prior history.
+A natural question is now: how often should simplification occur?
+As shown in the next section,
+there is no computational advantage to simplifying more often than is necessary
+to keep memory usage down.
+
+%%%%%%%%%%%%
+\subsubsection*{Estimates of run-time complexity}
+
+Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
+using Algorithm~\algref{W} for $T$ generations,
+with simplify interval $s = 1$ so that redundant information is removed after every generation.
+Since every chromosome inherits material from both parents,
+without simplification this would produce tables of
+$2NT$ nodes and
+$4NT$ edges.
+Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
+($T \sim 20N$, say).
+After simplification, we are left with the tree sequence describing the history
+of only the current generation of $2N$ individuals.
+This tree sequence has $4N-2$ edges to describe the leftmost tree;
+and each time the marginal tree changes along the sequence,
+four edges end and four new edges begin 
+(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
+Coalescent theory tells us that
+the expected total length of the edges of a marginal tree is approximately $4N\log(2N)$,
+which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
+since we have taken one crossover per individual per generation.
+Not all such recombinations actually change the tree topology,
+so four times this gives us an upper bound on the expected number of additional edges.
+Similarly, not every new edge derives from a never-before-seen node,
+but the number of nodes is at most equal to the number of edges plus the sample size.
+With $T=20N$, this reduces the initial storage of $120 N^2$ items to 
+% 4N - 2 + 2N + 4N log(2N) * 4 * 2 = 6N + 32 N log(2N)
+$6N + 32 N \log(2N)$;
+at $N=10^4$, a factor of 3,700.
+
+To get an idea of how required space depends on the length of time between simplification steps,
+suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
+begun with no prior history.
+How many of the resulting $4NT$ edges are required after simplification?
+As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
+Roughly, the length of time for which a marginal tree has $k$ tips is
+$2N/k(k-1) = 2N(1/(k-1) - 1/k)$,
+and so the time to go from $N$ to $n$ tips is $2N(1/(n-1) - 1/(N-1))$.
+This implies that after running for time $T$ we expect to
+have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
+The total tree length over this time is
+$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
+% is approximately
+% \begin{align*}
+%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
+% \end{align*}
+leads to an upper bound on the number of edges of
+\begin{align}
+    \label{eqn:edge_bound}
+    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
+\end{align}
+This holds up quite well in practice,
+as shown in Figure~\ref{fig:num_edges}.
+% by comparison to
+% the number of edges actually required for 50 independent Wright--Fisher simulations,
+% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
+% is shown in Figure~\ref{fig:num_edges}, compared to this prediction.
+
+\begin{figure}
+    \begin{center}
+        \includegraphics{sims/num_edges_50}
+    \end{center}
+    \caption{
+        Number of edges in the simplified tree sequence
+        for 50 Wright--Fisher simulations begun with no prior history,
+        as a function of number of generations.
+        Each line is one simulation, the black line gives the average,
+        and the dotted line is the upper bound of equation \eqref{eqn:edge_bound}.
+        \label{fig:num_edges}
+    }
+\end{figure}
+
+If we were to add mutations in forwards time
+under the infinite-sites model with total mutation rate per generation $\mu$,
+there would also be around $\mu 2NT$ mutations (and the same number of sites).
+Since mutations fall as recombinations do on the marginal trees,
+adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations,
+and the same savings as we obtained from edges above,
+even without considering the 
+computational burden of propagating mutations forwards across generations.
+
+Our method stores genealogies, and so records substantially more information
+than would a method only recording genotypes.
+However, since the simplification algorithm requires computational effort,
+it is informative to compare computational complexity of the algorithm
+to one that propagates neutral genotypes.
+A typical individual differs at around $2 N \mu$ sites from the population's consensus sequence,
+so propagating these to offspring by simple copying will take $4 N^2 \mu$ operations per generation.
+On the other hand, 
+in our scheme we must store $13N$ quantities per generation (two edges and one node per individual).
+The simplification algorithm is linear in the number of edges of the input tree sequence,
+and so multiplies this by a constant factor.
+This implies that propagating neutral genotypes for $T$ generations has time complexity $O(\mu T N^2)$,
+while our implementation is only $O(T N \log(2N))$.
+
+%%%%%%
+\subsection*{Overview of the API}
+
+These tools are implemented through the \msprime{} Python API,
+which provides a powerful platform for working with tree topology and mutation data.
+The new portions of \msprime{} that we cover here
+are dedicated to tree sequence input and output using simple tables of data,
+so we refer to this as the `Tables API'.
+The four key tables: nodes, edges, sites and mutations, are described above.
+
+The Tables API is primarily designed to facilitate efficient interchange of
+data between programs or between different modules of the same program.
+% we are capitalizing Tables API but not the more generic "tables".
+% \jda{should capitalize Tables, or not, as above}
+Following current best practises
+\plr{citations: Apache arrow, etc},
+data is stored in a columnar format,
+which allows for very efficient interchange of large amounts of numerical data. 
+In principle, this enables zero-copy semantics, 
+where a data consumer can read the information directly
+from the memory of a producer without incurring the overhead of a copy.
+Our implementation uses the numpy C API \citep{numpy_C_api} to efficiently copy
+data from Python into the low-level C library used to manipulate tree sequences.
+Since tables are encoded as contiguous arrays of numeric data,
+this achieves data transfer rates that would be impossible under
+any text-based approach while retaining excellent portability.
+
+This also provides basic input and output operations via numpy's
+array interface, which provides a great deal of flexibility and efficiency, 
+and makes it straightforward to transfer data from sources
+such as HDF5 \citep{hdf5}, Dask, or Zarr.
+(For small scale data and debugging purposes, a simple text format is also supported.)
+% PLR: this was mentioned above, and belongs in the documentation, I think.
+% The API also provides functions to sort tables
+% to ensure that the records are in the form required to input
+% into an \msprime{} tree sequence object. (Simplification may also be required.)
+
+The \msprime{} Tables API is designed to be easily ``plugged'' into any simulation engine.
+One simply needs to fill the relevant columns,
+and to send them to \msprime{} at regular intervals for simplification.  
+For example, in the benchmarking simulations above,
+we used the \fwdpp{} C++ API \citep{fwdpp}
+to fill arrays of simple structures representing the node and edge data.
+Doing so requires no modification of the \fwdpp{} code base.  
+Rather, we simply need to bookkeep parent/offspring labels, 
+and perform simple processing of the recombination breakpoints from each mating event.  
+Using \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}),
+the node/edge arrays on the C++ side are made visible, without copy, to Python as a multi-column numpy array,
+which is processed by the \msprime{} tables API.  
+Our implementation also uses \texttt{pybind11}
+to create a Python package extending \texttt{fwdpy11} 
+(a Python package based on \fwdpp{}; \url{https://github.com/molpopgen/fwdpy11}), 
+% (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
+% API from a command-line C/C++ application via an embedded Python interpreter.)
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Discussion}
+
+In this paper, we have shown that storing pedigrees
+and associated recombination events
+in a forwards-time simulation
+not only results in having available a great deal more information about the simulated population,
+but also can speed up the simulation by orders of magnitude.
+To make this feasible,
+we have described how to efficiently store this information in numerical tables,
+and have described a fundamental algorithm for simplification of tree sequences.
+Conceptually, recording of genealogical and recombination events
+can happen independently of the details of simulation;
+for this reason, we provide a well-defined and well-tested API in Python
+for use in other code bases (a C API is also planned).
+
+The idea of storing genealogical information to speed up simulations is not new.  
+It was implemented completely in AnA-FiTS~\citep{aberer2013rapid}, 
+but without the critical step of discarding irrelevant genealogical information.  
+A similar but much more limited method for
+discarding this information does appear in \citet{padhukasahasram2008exploring}.
+
+The tree sequences produced by default by this method
+are very compact, storing genotype \emph{and} genealogical information
+in a small fraction of the space taken by a compressed VCF file.
+The format is also highly efficient to process,
+leading to advantages for downstream analysis as well.
+This is because many algorithms to compute statistics of interest for population genetics
+are naturally expressed in terms of tree topologies,
+and so can be quickly computed from the trees underlying the tree sequence format.
+For example, pairwise nucleotide diversity $\pi$, is the average density of
+differences between sequences in the sample.
+To compute this directly from sequence data at $m$ sites in $n$ samples
+requires computing allele frequencies, taking $O(nm)$ operations.
+By using the locations of the mutations on the marginal trees,
+and the fact that these are correlated,
+sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
+can do this in roughly $O(n + m \log n)$ operations.
+The \msprime{} API provides a method to compute $\pi$ among arbitrary subsets of the
+samples in a tree sequence, which took about 1.2 seconds
+applied to the example tree sequence above with 1.1 million mutations
+in 200 megabases for 500,000 samples.
+
+Another attractive feature of this set of tools
+is that it makes it easy to incorporate \emph{prior history},
+simply by seeding the simulation with a (relatively inexpensive) coalescent simulation.
+This allows for incorporation of deep-time history beyond the reach of individual-based simulations.
+This may not even negatively affect realism,
+since geographic structure from times longer ago than the mixing
+time of migration across the range has limited effect on modern genealogies
+\citep{wilkins2004separation},
+other than possibly changing effective population size \citep{barton2002neutral,cox2002stepping}.
+
+Simulating very large populations and entire genomes will likely require parallelization.
+The one-way nature of information flow here makes our scheme relatively easy to incorporate
+into a parallel algorithm.
+Recently, \citet{lawrie2017accelerating} used parallelization 
+to improve performance for the ``Poisson Random Field'' family of models~\citep{Sawyer1992-jw},
+in which mutations are unlinked and there is no effect of genetic background on fitness.  
+However, thus far little attention has been paid to parallelization of simulations involving linkage.
+
+A final note:
+in preparing this manuscript,
+we debated a number of possible terms for the embellished pedigree,
+i.e., the ``pedigree with ancestral recombination information''.
+Etymological consensus \citep{liberman2014little} has
+``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
+(whose branching pattern resembles the bifurcation of a single parent-offspring relationship).
+An analogous term for the embellished pedigree might then be \emph{nedigree},
+from ``nid de grue'',
+as the nest of a crane is a large jumble of (forking) branches.
+We thought it unwise to use this term throughout the manuscript,
+but perhaps it will prove useful elsewhere.
+
+%%%%%%%%%%%%%%
+% \section*{Methods}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements}
+Thanks to Brad Shaffer and Evan McCartney--Melstad for useful discussions.
+Funding for this project came from XXX
+
+\bibliography{references}
+
+\appendix
+
+%%%%%%%%%%%%%%
+\section{Simplification algorithm}
+\label{ss:simplify_algorithm}
+
+\plr{Replace this with Jerome's implementation.}
 
@@
 
-\jk{Moved from above}
-More concretely,
-the algorithm works by moving back through time,
-processing each parent in the input tree sequence in chronological order.
-The main state of the algorithm at each point in time is a set of ancestral lineages,
-and each lineage is a collection of ancestral segments.
-An ancestral segment $(\ell, r, u)$ is found in a lineage
-if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
-% These segments are stored in a collection of linked lists,
-% one list of segments for each ancestral lineage present at that time.
-(These lineages are the ``colors'' above.)
-We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
-the input tree sequence.
-Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.
-
-%%%%%%%%%%%%%%
-\subsection*{Forward simulation and sequential simplification}
-
-At any point in a simulation scheme that, like Algorithm~\algref{W}, records
-data into tables the genealogical history is recorded
-in a tree sequence.
-This has two additional advantages, as seen in Algorithm~\algref{W}.
-First, simplification
-can be run periodically through the simulation, taking the set of samples to be
-the entire currently alive population. This is important as it keeps memory
-usage from growing linearly (and quickly) with time. Second, the simulation can
-be \emph{begun} with a tree sequence produced by some other method -- for
-instance, by a coalescent simulation with \msprime.
-We provide results for simulation timing from two specific implementations of this
-method (forward simulation coupled with sequential simplification)
-in the next section, but first we analyze the expected resource use of the
-general Algorithm~\algref{W}.
-
-%%%%%%%%%%%%
-\subsubsection*{Estimates of run-time complexity}
-
-\jda{It'd be good to outline what readers should expect here.}
-\jda{and note we're doing the space analysis for with reference to alg W with $s=1$}
-Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
-using Algorithm~\algref{W}
-for $T$ generations and simplify interval $s = 1$ so that redundant information
-is removed after every generation.
-Since there is exactly one crossing over every generation,
-this produces tables of
-$2NT$ nodes and
-$4NT$ edges.
-Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
-so that $T \sim 20N$, say.
-After simplification, we are left with the tree sequence describing the history
-of only the current generation of $2N$ individuals.
-This tree sequence has $4N-2$ edges to describe the leftmost tree;
-and each time the marginal tree changes along the sequence,
-four edges end and four new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
-Coalescent theory tells us that
-the expected total length of edges in a marginal tree is approximately $4N\log(2N)$,
-which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
-since we have taken one crossover per generation.
-Not all such recombinations actually change the tree topology,
-and four times this gives us an upper bound on the expected number of edges.
-Similarly, not every new edge derives from a never-before-seen node,
-but the number of nodes is at most equal to the number of edges plus the sample size.
-With $T=20N$, this reduces the initial storage of $160 N^2$ items to $8N(4\log(2N) + 3)$;
-at $N=10^4$, a factor of 4,700.
-\jk{I haven't checked these numbers, just corrected the obvious threes/fours.}
-
-To get an idea of how required space depends on the length of time,
-suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
-begun with no prior history.
-This produces $4NT$ edges; how many are required after simplification?
-As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
-Roughly, the length of time for which a marginal tree has k tips is
-$2N/k(k-1) = 2N(1/(k-1) - 1/k)$,
-and so the time to go from N to n tips is $2N(1/(n-1) - 1/(N-1))$.
-This implies that after running for time $T$ we expect to
-have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
-The total tree length over this time is
-$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
-% is approximately
-% \begin{align*}
-%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
-% \end{align*}
-leads to an upper bound on the number of edges of
-\begin{align*}
-    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
-\end{align*}
-This holds up quite well in practice --
-the number of edges actually required for 50 independent Wright--Fisher simulations,
-as a function of time (obtained by running simplify every generation)
-is shown in Figure~\ref{fig:num_edges}, compared to this prediction.
-
-\begin{figure}
-    \begin{center}
-        \includegraphics{sims/num_edges_50}
-    \end{center}
-    \caption{
-        Number of edges in the simplified tree sequence
-        for 50 Wright-Fisher simulations begun with no prior history
-        as a function of number of generations.
-        Each line is one simulation, the black line gives the average,
-        and the dotted line is the upper bound of the text.
-        \label{fig:num_edges}
-    }
-\end{figure}
-
-If we were to add mutations in forwards time
-under the infinite-sites model with total mutation rate per generation $\mu$,
-there would also be around $\mu 2NT$ mutations (and the same number of sites).
-Since mutations fall as recombinations do on the marginal trees,
-adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations.
-This furthermore avoids the computational burden of propagating mutations forwards through the generations.
-
-Our method stores genealogies, and so records substantially more information
-than would a method only recording genotypes.
-However, since the simplification algorithm requires some computational effort,
-it is still informative to compare computational complexity of the algorithm
-to one that propagates neutral genotypes.
-A typical individual will differ at $2 N \mu$ sites from the population's consensus sequence,
-so propagating these to offspring by simple copying will take $4 N^2 \mu$ operations per generation.
-On the other hand, we must store $13N$ quantities per generation (two edges and one node per individual).
-The simplification algorithm is linear in the number of edges of the input tree sequence,
-and so multiplies this by a constant factor.
-Concretely, propagating neutral genotypes for $T$ generations has time complexity $O(\mu T N^2)$,
-while our implementation is only $O(T N \log(2N))$.
-
-%%%%%%
-\subsection*{Overview of the API}
-
-The \msprime{} Python API provides a powerful platform for working with tree topology and mutation data.
-The new portions of \msprime{} that we cover here
-are dedicated to tree sequence input and output using simple tables of data,
-so we refer to this as the `Tables API'.
-The four key tables: nodes, edges, sites and mutations, are described above.
-
-The Tables API is primarily designed to facilitate efficient interchange of
-data between programs or between different modules of the same program.
-\jda{should capitalize Tables, or not, as above}
-Following the current best-practises [citations: apache arrow, etc] data is stored
-in a columnar format.
-The principal advantage of this for our purposes is that it allows for very efficient
-interchange of large amounts of numerical data. In principle, this enables
-zero-copy semantics, where a data consumer can read the information directly
-from the memory of a producer without incurring the overhead of a copy
-[citation?] Our implementation uses the numpy C API [citation] to efficiently copy
-data from Python into the low-level C library used to manipulate
-tree sequences.
-Thus, by using a simple numerical
-encoding of tree topologies and contiguous arrays of data to store table
-columns, we can achieve data transfer rates that would be impossible under
-any text-based approach while retaining excellent portability.
-
-The \msprime{} tables API is independent of the details of a particular simulation engine.  In other words, one simply
-needs to fill the relevant columns and send them to \msprime{} at regular intervals for simplification.  For example,
-using the \fwdpp{} C++ API \cite{fwdpp}, we fill arrays of simple structures representing the node and edge data.
-Doing so requires no modification of the \fwdpp{} code base.  Rather, we simply need to book-keep the parent/offspring
-labels and perform simple processing of the recombination breakpoints from each mating event.  Using \texttt{pybind11}
-(\url{https://github.com/pybind/pybind11/}),
-the node/edge arrays on
-the C++ side are made visible, without copy, to Python as a multi-column numpy array. This array is in turn processed by
-the \msprime{} tables API.  In order to create a standalone user interface, our implementation uses \texttt{pybind11} to
-create a Python package extending \texttt{fwdpy11} (\url{https://github.com/molpopgen/fwdpy11}), which is a Python
-package based on \fwdpp{}.  (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
-API from a command-line C/C++ application via an embedded Python interpreter.)
-\plr{@molpopgen: note on how this works with fwdpp?}
-\krt{@plr: done, but perhaps too verbose?}
-
-The tables API provides basic input and output operations via the numpy
-array interface, which provides a great deal of flexibility as well
-as efficiency, and makes it straightforward to transfer data from sources
-such as HDF5 [citation], Dask, or Zarr.
-(For small scale data and debugging purposes, a simple text format is also supported.)
-Along with these operations we also provide a function to sort a set of tables
-to ensure that the records are in the form required to input
-into an \msprime{} tree sequence object. (Simplification may also be required.)
-
-\plr{Maybe we don't need this paragraph? Already said it, more or less, above, esp if we insert something about how fwdpp works.}
-\jda{agree re the next two lines. propose moving the other above}
-This interchange API is very efficient. [Describe a quick example where we generate
-a many-gigabyte tree sequence using fwdpp, and the time required
-to copy the node and edge data into the tables].
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Discussion}
-
-In this paper, we have shown that storing pedigrees
-and associated recombination events
-in a forwards-time simulation
-not only results in having available a great deal more information about the simulation,
-but also can speed up the simulation by orders of magnitude.
-To make this feasible,
-we have described how to efficiently store this information in numerical tables,
-and have described a fundamental algorithm for simplificication of tree sequences.
-Conceptually, recording of genealogical and recombination events
-can happen independently of the details of simulation;
-for this reason, we provide a well-defined and well-tested API in Python
-for use in other code bases (a C API is also planned).
-
-The tree sequences produced by default by this method
-are very compact, storing genotype \emph{and} genealogical information
-in a small fraction of the space taken by a compressed VCF file.
-The format is also highly efficient to process,
-leading to advantages for downstream analysis as well.
-This is because many algorithms to compute statistics of interest for population genetics
-are naturally expressed in terms of tree topologies,
-and so can be quickly computed from the trees underlying the tree sequence format.
-For example, pairwise nucleotide diversity $\pi$, is the average density of
-differences between sequences in the sample.
-To compute this directly from sequence data at $m$ sites in $n$ samples
-requires computing allele frequencies, taking $O(nm)$ operations.
-By using the locations of the mutations on the marginal trees,
-and the fact that these are correlated,
-sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
-can do this in roughly $O(n + m \log n)$ operations.
-The \msprime\ API provides a method to compute $\pi$ among arbitrary subsets of the
-samples in a tree sequence, which took about 1.2 seconds
-applied to the example tree sequence above with 1.1 million mutations
-in 200 megabases for 500,000 samples.
-
-Another attractive feature of this set of tools
-is that it makes it easy to incorporate \emph{prior history},
-simply by seeding the simulation with a (relatively inexpensive) coalescent simulation.
-This allows for incorporation of deep-time history beyond the reach of individual-based simulations.
-Since geographic structure from times longer ago than the mixing
-time of migration across the range has limited effect on modern genealogies
-\citep{wilkins2004separation} (other than possibly changing effective population
-size \citet{barton2002neutral,cox2002stepping}), this may not negatively affect realism.
-
-\jk{Moved this from the introduction: here seems like the right place to
-discuss the importance of parallisation. We should rework this paragraph to
-discuss say that a nice part of our approach is that we can do the
-simplification in parallel. Also looking forward we need to parallise more
-because the simulations need so much RAM}
-Simulating very large popluations and entire genomes will likely require parallelization,
-which \cite{lawrie2017accelerating} used to improve
-performance for the ``Poisson Random Field'' family of models~\citep{Sawyer1992-jw},
-in which mutations are unlinked and there is no
-effect of genetic background on fitness.  However, little, and perhaps even no, attention has been paid to
-parallelization of simulations involving linkage.  Here, we show that recording genealogical information instead of
-explicitly tracking neutral mutations results in order-of-magnitude performance improvements to simulations based on
-\fwdpp{}.
-
-\jk{Moved this here from intro also. Basically we just need to say that others
-have done something similar. Ours is much better than Ana-Fits because ...}
-The  idea of storing genealogical information to speed up simulations is not new.  AnA-FiTS implemented it
-completely~\citep{aberer2013rapid}, but this program did not implement the critical step of
-discarding irrelevant genealogical information.  A similar but more limited method for
-discarding this information does appear in \citet{padhukasahasram2008exploring}.
-
-A final note:
-in preparing this manuscript,
-we debated a number of possible terms for the embellished pedigree,
-i.e., the ``pedigree with ancestral recombination information''.
-Current etymologica consensus \citep{liberman2014little} has
-``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
-(whose branching pattern resembes the bifurcation of a single parent-offspring relationship).
-An analogous term for the embellished pedigree might be \emph{nedigree},
-from ``nid de grue'',
-as the nest of a crane is a large jumble of (forking) branches.
-We thought it unwise to use this term throughout the manuscript,
-but perhaps it will prove useful elsewhere.
-
-%%%%%%%%%%%%%%
-% \section*{Methods}
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements}
-Thanks to Brad Shaffer and Evan McCartney-Melstad for useful discussions.
-Funding for this project cam from XXX
-
-\bibliography{references}
-
-\appendix
-
 \end{document}

### 2017-12-09 forwards_paper.tex
added other simplify complexity info

@@
 \maketitle
-
-\emph{Note:} author order not determined
 
@@
 %%%%%%%%%%%%%%
-\subsection*{Sequential simplification and prior history}
+\subsubsection*{Sequential simplification and prior history}
 
@@
 
+\plr{Refer to simulation results here.}
+
+%% Memory reduction thanks to simplification
 Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
@@
 
+%% Progressive memory reduction as a function of T
 To get an idea of how required space depends on the length of time between simplification steps,
@@
 This holds up quite well in practice,
-as shown in Figure~\ref{fig:num_edges}.
+as shown in Figure~\ref{fig:simplify_complexity}.
 % by comparison to
@@
 % as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
-% is shown in Figure~\ref{fig:num_edges}, compared to this prediction.
+% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.
 
@@
         \includegraphics{sims/num_edges_50}
+        \includegraphics{sims/simplify_timings}
     \end{center}
     \caption{
+        \textbf{(A)}
         Number of edges in the simplified tree sequence
@@
         and the dotted line is the upper bound of equation \eqref{eqn:edge_bound}.
-        \label{fig:num_edges}
+        \plr{Propose to add other figures as subfigures here:}
+        \textbf{(B)}
+        Time required to simplify the first $k$ edges
+        of \plr{what tree sequence?},
+        plotted against $k$.
+        \textbf{(C)}
+        Time required to simplify
+        \plr{what tree sequence?}
+        down to a sample of size $n$,
+        plotted against $n$
+        (note the log scale;
+        the time scales logarithmically with $n$).
+        \label{fig:simplify_complexity}
     }
@@
 
+%% Above was edges; same holds for memory usage savings due to mutations
 If we were to add mutations in forwards time
@@
 computational burden of propagating mutations forwards across generations.
+
+%% simplify run time
+Above we have described memory requirements;
+how does the computation time required for simplification scale?
+Simply because it must process each edge,
+the simplification algorithm is linear in the number of edges of the input tree sequence.
+This is seen in Figure~\ref{fig:simplify_complexity}X,
+where we show shows the time required to simplify increasingly large subsets of a large tree sequence.
+When simplifying the result of a forwards-time sequence, the number of edges is the main contributing factor.
+Suppose on the other hand we want to 
+simplify an already-minimal but large tree sequence with $N$ nodes
+to a subsample of size $n$.
+How does the required time scale with $n$?
+In this case, the computation is dominated by the size of the output tree sequence,
+which grows with $\log(n)$.
+this is also seen in Figure~\ref{fig:simplify_complexity}X,
+showing the time required to simplify progressively larger subsets of a single large tree sequence.
 
@@
 in our scheme we must store $13N$ quantities per generation (two edges and one node per individual).
-The simplification algorithm is linear in the number of edges of the input tree sequence,
+\plr{XXX}
+The simplification algorithm is linear in the number of edges of the input tree sequence
+(because it must process each edge),
 and so multiplies this by a constant factor.
@@
 in 200 megabases for 500,000 samples.
+Calculating $\pi$ from the numeric genotype matrix using numpy took \plr{HOW MANY} seconds.
 

### 2017-12-09 forwards_paper.tex
merged jaime edits

@@
 and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
-We then describe a general-purpose method that uses these tools
-(as implemented in \msprime{})
-to efficiently record simplified a tree sequence,
-and analyze its run time complexity.
+Finally, we analyze the run time and space complexity of our general-purpose method.
 
@@
 
-To measure the performance gains from ancestry recording, we ran simulations implemented in C++ (see below) with and without ancestry tracking.  
-The simulations without ancestry tracking introduced neutral mutations at a rate 
+To measure the performance gains from recording the pedigree we ran simulations with and
+without recording, for both implementations (\simupop{} \citep{simupop}
+and \fwdpp{} \citep{fwdpp}, as described below).
+All simulations implemented a discrete-time Wright-Fisher population of diploid
+population size $N$, simulated for $10N$ generations.
+The simulations without pedigree recording introduced neutral mutations at a rate 
 equal to the recombination rate,
@@
 respectively, per gamete and per generation.
-Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,
-and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
-We categorize simulations in terms of the \emph{scaled} parameters
-$\theta = 4N\mu$ and $\rho = 4Nr$, where $N$ is the diploid population size,
-since these most strongly determine run time.
-
-We ran one simulation for each combination of $N$ from $(1000, 10000, 50000)$, and $\rho$ from $(10^3, 10^4, 10^5)$.
-Deleterious mutations were introduced at rate $\rho/100$ per generation, 
-drawing scaled selection coefficients ($2Ns$)
-from a Gamma distribution with a mean of -5 and a shape parameter of 1.  
-This distribution of fitness effects results in
-thousands of weakly-deleterious mutations segregating in the population, 
-many of which drift to intermediate frequencies.  
-The case of many mutations with selection is a non-trivial violation of exchangeability assumptions of the
-coalescent \citep{Neuhauser1997-nn}.
-Therefore, these selected mutations must be explicitly tracked in our forward simulation,
-and the time savings due to ancestry tracking come from not having to record \textit{neutral} mutations.
+% Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,
+% and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
+We ran simulations with different values of $N$
+and of the \emph{scaled} parameter $\rho = 4Nr$,
+categorized by the value of $\rho$, since this
+is what most strongly determines run time
+(along with the scaled mutation rate, $\theta = 4N\mu$, which is also equal to $\rho$.)
+
+Deleterious mutations were introduced at rate $\rho/100$ per generation, drawing scaled selection
+coefficients ($2Ns$)
+from a Gamma distribution with a mean of -5 and a shape parameter of 1.  This distribution of fitness effects results in
+thousands of weakly-deleterious mutations segregating in the population, many of which drift to intermediate
+frequencies.  The case of many mutations with selection is a non-trivial violation of exchangeability assumptions of the
+coalescent \citep{Neuhauser1997-nn}.  Therefore, these selected mutations must be explicitly tracked in our forward simulation
+and the time savings due to pedigree recording come from not having to record \textit{neutral} mutations.
 
@@
 
-For efficiency reasons, we only ran \simupop{} with ancestry tracking.
-
-\krt{Jaime--any details here to add re: how simuPOP was tweaked??}
-
+\subsubsection*{The \fwdpp{} implementation}
+
+We implemented Wright-Fisher simulations in C++ using \fwdpp{} library
+functions and interface code that records data into the \msprime{} Tables API (see below).
+In these simulations, $\mu$ and $r$ are the means of a Poisson process and we used the infinitely-many sites mutation model
+\citep{Kimura1969-uc}.
+We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.
+
+\plr{Perhaps implementations should go below in Methods?}
 \plr{Insert figures/tables of performance here and describe them.}
 
-% \plr{Jaime and Kevin to describe simulation methods here (separately).}
-% 
+\subsubsection*{The \simupop{} implementation}
+
+We implemented Wright-Fisher simulations in Python using \simupop{} library
+functions (which wrap C++ code) and interface code (\ftprime{}, also in Python) that records data into the \msprime{} Tables API (see below).
+In these simulations, $\mu$ and $r$ define a per-base-pair mutation rate and we
+defined the length of the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.
+We varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.
+
 % Comparison of simulation with/without msprime, using \simupop{}

### 2017-12-09 forwards_paper.tex
a few more edits

@@
 
-To measure the performance gains from recording the pedigree we ran simulations with and
-without recording, for both implementations (\simupop{} \citep{simupop}
-and \fwdpp{} \citep{fwdpp}, as described below).
+To measure the performance gains from recording the pedigree we ran simulations both with and without recording.
 All simulations implemented a discrete-time Wright-Fisher population of diploid
@@
 
+To demonstrate the method's flexibility and to test it in different contexts,
+we implemented these simulations separately using two different simulation engines:
+\begin{enumerate}
+
+    \item 
+        In C++, using \fwdpp{} library functions and interface code (details below),
+        using a continuum-sites model for both mutation and recombination.
+        % $\mu$ and $r$ were the means of Poisson processes,
+        % and mutations used the infinite-sites model \citep{Kimura1969-uc}.
+        % We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.
+
+    \item 
+        In Python, using the \simupop{} library interfacing with a custom Python library (\ftprime{}),
+        using a discrete but large number of sites.
+        % $\mu$ and $r$ define per-base-pair mutation rate and we
+        % defined the length of the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.
+        % We varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.
+
+\end{enumerate}
+
 We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 Gigahertz Intel E5-2650 CPU with
@@
 
-\subsubsection*{The \fwdpp{} implementation}
-
-We implemented Wright-Fisher simulations in C++ using \fwdpp{} library
-functions and interface code that records data into the \msprime{} Tables API (see below).
-In these simulations, $\mu$ and $r$ are the means of a Poisson process and we used the infinitely-many sites mutation model
-\citep{Kimura1969-uc}.
-We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.
-
-\plr{Perhaps implementations should go below in Methods?}
 \plr{Insert figures/tables of performance here and describe them.}
-
-\subsubsection*{The \simupop{} implementation}
-
-We implemented Wright-Fisher simulations in Python using \simupop{} library
-functions (which wrap C++ code) and interface code (\ftprime{}, also in Python) that records data into the \msprime{} Tables API (see below).
-In these simulations, $\mu$ and $r$ define a per-base-pair mutation rate and we
-defined the length of the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.
-We varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.
 
@@
         defined by \msprime.
-        \plr{Whoops!  Genotypes in the second tree should be `GGC'}
-        \jk{Some minor alignment issues present here.}
-        \plr{turns out this is no fun to edit in inkscape.
-            I can re-make to match the other figs
-            but will leave alone for now: it's good!}
         \label{fig:example_tree_sequence}
@@
         and the dotted line is the upper bound of equation \eqref{eqn:edge_bound}.
-        \plr{Propose to add other figures as subfigures here:}
+        \plr{Propose to add other figures as subfigures here; but better than this.}
         \textbf{(B)}
@@
 
-\plr{Replace this with Jerome's implementation.}
+\plr{Replace this with Jerome's interval-tree implementation?}
 

### 2017-12-12 forwards_paper.tex
Update intro and tree sequence section.

@@
 fwdpp \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
-population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\approx 5
+population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
 \times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
@@
 so we discard all information irrelevant to the genetic history
-of the \emph{sampled} individuals,
-which results in considerable savings.
-Another way of representing this same information
-is known as the \emph{ancestral recombination graph} --
-the {ARG} \citep{griffiths1997ancestral},
-where nodes are common ancestor or recombination events --
-which has been the subject of substantial study
-under the assumptions of coalescent
-theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
-We use tree sequences for the substantial algorithmic advantages exploited in~\citep{kelleher2016efficient}.
+of the extant population. We show that we need only $O(N \log N)$
+space to store the resulting tree sequence for a single crossover per
+generation Wright-Fisher model.
 
@@
 such as those describing the history of a sexual population.
-It is efficient because branches that are shared by adjacent trees are stored once,
+A closely related structure is known as the \emph{ancestral recombination graph} --
+the {ARG} \citep{griffiths1991two,griffiths1997ancestral},
+where nodes are common ancestor or recombination events --
+which has been the subject of substantial study
+under the assumptions of coalescent
+theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
+However, the properties of the ARG as a computational structure have not
+been studied and, despite several efforts to standardise a common
+format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper.
+ARGs are rarely used
+in practise. In contrast, the algorithmic properties of tree sequence
+algorithms have been explored in detail~\citep{kelleher2016efficient},
+contributing substantially to the efficiency of the \msprime\ coalescent simulator.
+
+Tree sequences are efficient because branches that are shared by adjacent trees are stored once,
 rather than repeatedly for each tree.

### 2017-12-21 sims/cpp_benchmarks/organize_total_runtimes.py
fix organizing script for cpp output.  there are two missing data points to re-run

@@ -19,7 +19,7 @@ for i in files:
     SEL = True
     if re.search('all_neutral', i):
         SEL = False
-        if re.search('with', i) is False:
+        if re.search('neutral_all_neutral', i):
             ARG = False
     else:
         if re.search('with', i):

### 2017-12-21 sims/quick_plot.py
update data with missing data points and move placement of legend

@@ -137,7 +137,7 @@ for name, group in groups:
             color=colors[name[3]],
             markerfacecolor=mfacecolor)
 
-ax_fwdpp.legend(loc='best')
+ax_fwdpp.legend(loc='upper left')
 ax_fwdpp.set_title("fwdpy11 with neutral mutations",fontsize='medium')
 ax_fwdpp_arg.set_title("fwdpy11 with ancestry tracking",fontsize='medium')
 ax_simupop.set_ylabel("Run time (hours)")

### 2017-12-21 sims/quick_plot.py
autopep8

@@ -50,12 +50,12 @@ for name, group in groups:
             markerfacecolor=mfacecolor)
 
 ax_fwdpp.legend(loc='best')
-ax_fwdpp.set_title("fwdpy11 with neutral mutations",fontsize='medium')
-ax_fwdpp_arg.set_title("fwdpy11 with ancestry tracking",fontsize='medium')
+ax_fwdpp.set_title("fwdpy11 with neutral mutations", fontsize='medium')
+ax_fwdpp_arg.set_title("fwdpy11 with ancestry tracking", fontsize='medium')
 ax_simupop.set_ylabel("Run time (hours)")
 ax_fwdpp.set_ylabel("Run time (hours)")
-ax_simupop.set_title("simuPOP with neutral mutations",fontsize='medium')
-ax_simupop_arg.set_title("simuPOP with ancestry tracking",fontsize='medium')
+ax_simupop.set_title("simuPOP with neutral mutations", fontsize='medium')
+ax_simupop_arg.set_title("simuPOP with ancestry tracking", fontsize='medium')
 ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop_arg.set_xticks([1e3, 1e4, 1e5])
@@ -98,8 +98,8 @@ for name, group in groups:
 
 for ax in (ax_fwdpp, ax_simupop):
     ax.set_ylabel("Speedup due to\nARG tracking")
-ax_fwdpp.set_title("fwdpy11",fontsize='medium')
-ax_simupop.set_title("simuPOP",fontsize='medium')
+ax_fwdpp.set_title("fwdpy11", fontsize='medium')
+ax_simupop.set_title("simuPOP", fontsize='medium')
 ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop.set_xscale('log')
 # for tick in ax_simupop.get_xticklabels():
@@ -111,7 +111,7 @@ fig.tight_layout()
 plt.savefig("speedup.pdf")
 
 # Now, plots for sims w/o selection
-data_neut.sort_values(by='rho',inplace=True)
+data_neut.sort_values(by='rho', inplace=True)
 groups = data_neut.groupby(['engine', 'arg', 'queue', 'N'])
[…]

### 2017-12-21 sims/quick_plot.py
plot speedup for no selection sims

@@ -158,3 +158,40 @@ for ax in (ax_simupop, ax_simupop_arg):
 fig.tight_layout()
 # plt.ticklabel_format(style='sci', axis='x', scilimits=(0,2))
 plt.savefig("rawspeed_nosel.pdf")
+
+# Relative speedup for sims w/o selection
+data_arg = data_neut[(data_neut['arg'] == True) & (data_neut['queue'] == False)].copy()
+data_noarg = data_neut[(data_neut['arg'] == False) & (data_neut['queue'] == False)].copy()
+joined = data_arg.merge(data_noarg, on=[
+                        'engine', 'N', 'rho'], how='outer', suffixes=('_arg', '_noarg')).dropna()
+joined['speedup'] = joined['time_noarg'] / joined['time_arg']
+groups = joined.groupby(['engine', 'N'])
+fig, (ax_fwdpp, ax_simupop) = plt.subplots(
+    2,  sharex=True, sharey=True)
+for name, group in groups:
+    lstyle = 'solid'
+    if name[0] == 'fwdpy11':
+        ax = ax_fwdpp
+    m = 'o'
+    if name[0] == 'simuPOP':
+        ax = ax_simupop
+    mfacecolor = colors[name[1]]
+    popsize = int(name[1])
+    ax.plot(group.rho, group.speedup, marker=m,
+            ms=4, linestyle=lstyle, label=r'$N = {:.0e}$'.format(popsize),
+            color=colors[name[1]],
+            markerfacecolor=mfacecolor)
+
+for ax in (ax_fwdpp, ax_simupop):
+    ax.set_ylabel("Speedup due to\nARG tracking")
+ax_fwdpp.set_title("fwdpy11", fontsize='medium')
+ax_simupop.set_title("simuPOP", fontsize='medium')
+ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
+ax_simupop.set_xscale('log')
+# for tick in ax_simupop.get_xticklabels():
+#     tick.set_rotation(45)
+
+
+ax_fwdpp.legend(loc='best')
+fig.tight_layout()
+plt.savefig("speedup_nosel.pdf")

### 2017-12-21 sims/quick_plot.py
cleanup plotting script

@@ -61,14 +61,7 @@ ax_simupop_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop_arg.set_xticks([1e3, 1e4, 1e5])
 for ax in (ax_simupop, ax_simupop_arg):
     ax.set_xscale('log')
-#     for tick in ax.get_xticklabels():
-#         tick.set_rotation(45)
-    # ax.ticklabel_format(style='sci', axis='x', scilimits=(-3,0))
-# ax.set_yscale("log")
-# plt.legend(bbox_to_anchor=(1.05, 1), loc=2)
-# plt.xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 fig.tight_layout()
-# plt.ticklabel_format(style='sci', axis='x', scilimits=(0,2))
 plt.savefig("rawspeed.pdf")
 
 # Plot the relative speedup due to ARG tracking for the sims with selection
@@ -102,10 +95,6 @@ ax_fwdpp.set_title("fwdpy11", fontsize='medium')
 ax_simupop.set_title("simuPOP", fontsize='medium')
 ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop.set_xscale('log')
-# for tick in ax_simupop.get_xticklabels():
-#     tick.set_rotation(45)
-
-
 ax_fwdpp.legend(loc='best')
 fig.tight_layout()
 plt.savefig("speedup.pdf")
@@ -149,19 +138,14 @@ ax_simupop_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop_arg.set_xticks([1e3, 1e4, 1e5])
 for ax in (ax_simupop, ax_simupop_arg):
     ax.set_xscale('log')
-#     for tick in ax.get_xticklabels():
-#         tick.set_rotation(45)
-    # ax.ticklabel_format(style='sci', axis='x', scilimits=(-3,0))
-# ax.set_yscale("log")
-# plt.legend(bbox_to_anchor=(1.05, 1), loc=2)
-# plt.xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 fig.tight_layout()
-# plt.ticklabel_format(style='sci', axis='x', scilimits=(0,2))
 plt.savefig("rawspeed_nosel.pdf")
 
 # Relative speedup for sims w/o selection
[…]

### 2017-12-22 forwards_paper.tex
tiny tweak
add in benchmark results to manuscript

@@
 \newcommand{\fwdpp}{\texttt{fwdpp}}
+\newcommand{\fwdpy}{\texttt{fwdpy11}}
 \newcommand{\msprime}{\texttt{msprime}}
@@
 
+The total run times for simulations with selection are shown in Figure~\ref{fig:runtimes_selection}.  When tracking
+neutral mutations (instead of the ARG), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for
+these simulations).  With \fwdpy{}, simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
+With \simupop{}, we were only able to run simulations with $N=10^3$ and region sizes up to $10^4$.  ARG tracking
+dramatically improves run times for both \fwdpy{} and \simupop{} (right column of Figure~\ref{fig:runtimes_selection}).
+For parameters that ran to completion when tracking neutral mutations and when tracking
+the ARG, the relative speedup due to ARG tracking is up to $\approx 50$ fold (Figure~\ref{fig:relative_speedup_selection}).
+For \fwdpy{}, ARG tracking reduces runtimes more for larger population sizes
+(Figure~\ref{fig:relative_speedup_selection}).  We could not make such an observation with \simupop{}, which tended to
+run out of memory on our system.  Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
+partially obscures an important observation that the run times appear approximately linear in region size when ARG
+tracking.  The reason for this behavior is that the simulations are generating a constant number of new edges on average
+each generation, and the number of edges is a function of $\rho$ in expectation.
+
+We also performed a more limited set of simulations without natural selection.  The total run times are shown in
+Figure~\ref{sfig:rawspeed_nosel} and show the same qualitative behavior as simulations with selection
+(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to ARG tracking is again substantial in these
+simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to ARG tracking here with \fwdpy11{}
+than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
+close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
+more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
+and~\ref{sfig:speedup_nosel}.
+
+\begin{figure}
+    \includegraphics[]{sims/rawspeed}
+    \caption{\label{fig:runtimes_selection}Total run time for a single simulation replicate as a function of region
+        length, measured as the scaled recombination parameter $\rho = 4Nr$. The different colored lines represent
+        different diploid population sizes ($N$). The left column shows run times for
+        simulations including the tracking of neutral mutations.  Missing data points are due to either simulations
+    being killed at 72 hours or running out of memory (which only happened with \simupop{}).  The right column of
+panels shows run times tracking the ARG instead of the neutral mutations. The dashed line in the upper right panel shows
+the results for an implementation using \fwdpy{} where the ARG simplification steps were handled in a separate thread of
+execution and fitness calculations were parallelized.}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[]{sims/speedup}
+    \caption{\label{fig:relative_speedup_selection}Relative performance improvement due to ARG tracking. This figure
+    shows the ratio of the run time tracking neutral mutations over the run time tracking the ARG. Data points are taken
+from Figure~\ref{fig:runtimes_selection} for simulations that ran to completion using both methods.}
+\end{figure}
+
 % Comparison of simulation with/without msprime, using \simupop{}
@@
 
-We use $\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random.
-% (and all such instances is independent).  % this is understood here
+We use $\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random
+(and all such instances are independent).
 Given a node table $\Nt$, the function $\taddnode{\Nt, t}$
@@
 
+\newpage
+\clearpage
+\section*{Supplementary Information}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+
+\begin{figure}[!h]
+    \includegraphics[]{sims/rawspeed_nosel}
+    \caption{\label{sfig:rawspeed_nosel}Total run time for a single simulation replicate as a function of region
+        length, measured as the scaled recombination parameter $\rho = 4Nr$.  The simulations here involve no natural
+    selection.}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[]{sims/speedup_nosel}
+    \caption{\label{sfig:speedup_nosel}Relative performance improvement due to ARG tracking for simulations without
+    selection.  Data points are taken from Figure~\ref{sfig:rawspeed_nosel} and show the ratio of run times tracking
+neutral variation over the run times tracking the ARG.}
+\end{figure}
+
 \end{document}

### 2017-12-22 sims/quick_plot.py
cleanup terminology in figures

@@ -51,11 +51,11 @@ for name, group in groups:
 
 ax_fwdpp.legend(loc='best')
 ax_fwdpp.set_title("fwdpy11 with neutral mutations", fontsize='medium')
-ax_fwdpp_arg.set_title("fwdpy11 with ancestry tracking", fontsize='medium')
+ax_fwdpp_arg.set_title("fwdpy11 with pedigree tracking", fontsize='medium')
 ax_simupop.set_ylabel("Run time (hours)")
 ax_fwdpp.set_ylabel("Run time (hours)")
 ax_simupop.set_title("simuPOP with neutral mutations", fontsize='medium')
-ax_simupop_arg.set_title("simuPOP with ancestry tracking", fontsize='medium')
+ax_simupop_arg.set_title("simuPOP with pedigree tracking", fontsize='medium')
 ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop_arg.set_xticks([1e3, 1e4, 1e5])
@@ -90,7 +90,7 @@ for name, group in groups:
             markerfacecolor=mfacecolor)
 
 for ax in (ax_fwdpp, ax_simupop):
-    ax.set_ylabel("Speedup due to\nARG tracking")
+    ax.set_ylabel("Speedup due to\npedigree recording")
 ax_fwdpp.set_title("fwdpy11", fontsize='medium')
 ax_simupop.set_title("simuPOP", fontsize='medium')
 ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
@@ -128,11 +128,11 @@ for name, group in groups:
 
 ax_fwdpp.legend(loc='upper left')
 ax_fwdpp.set_title("fwdpy11 with neutral mutations", fontsize='medium')
-ax_fwdpp_arg.set_title("fwdpy11 with ancestry tracking", fontsize='medium')
+ax_fwdpp_arg.set_title("fwdpy11 with pedigree tracking", fontsize='medium')
 ax_simupop.set_ylabel("Run time (hours)")
 ax_fwdpp.set_ylabel("Run time (hours)")
 ax_simupop.set_title("simuPOP with neutral mutations", fontsize='medium')
-ax_simupop_arg.set_title("simuPOP with ancestry tracking", fontsize='medium')
[…]

### 2017-12-22 forwards_paper.tex
fwdpy11 -> fwdpy
terminology cleanup in text

@@
 The total run times for simulations with selection are shown in Figure~\ref{fig:runtimes_selection}.  When tracking
-neutral mutations (instead of the ARG), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for
+neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for
 these simulations).  With \fwdpy{}, simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
-With \simupop{}, we were only able to run simulations with $N=10^3$ and region sizes up to $10^4$.  ARG tracking
+With \simupop{}, we were only able to run simulations with $N=10^3$ and region sizes up to $10^4$.  Pedigree tracking
 dramatically improves run times for both \fwdpy{} and \simupop{} (right column of Figure~\ref{fig:runtimes_selection}).
 For parameters that ran to completion when tracking neutral mutations and when tracking
-the ARG, the relative speedup due to ARG tracking is up to $\approx 50$ fold (Figure~\ref{fig:relative_speedup_selection}).
-For \fwdpy{}, ARG tracking reduces runtimes more for larger population sizes
+the pedigree, the relative speedup due to pedigree tracking is up to $\approx 50$ fold (Figure~\ref{fig:relative_speedup_selection}).
+For \fwdpy{}, pedigree tracking reduces runtimes more for larger population sizes
 (Figure~\ref{fig:relative_speedup_selection}).  We could not make such an observation with \simupop{}, which tended to
 run out of memory on our system.  Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
-partially obscures an important observation that the run times appear approximately linear in region size when ARG
+partially obscures an important observation that the run times appear approximately linear in region size when pedigree
 tracking.  The reason for this behavior is that the simulations are generating a constant number of new edges on average
@@
 Figure~\ref{sfig:rawspeed_nosel} and show the same qualitative behavior as simulations with selection
-(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to ARG tracking is again substantial in these
-simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to ARG tracking here with \fwdpy11{}
+(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to pedigree tracking is again substantial in these
+simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to pedigree tracking here with \fwdpy{}
 than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
@@
     being killed at 72 hours or running out of memory (which only happened with \simupop{}).  The right column of
-panels shows run times tracking the ARG instead of the neutral mutations. The dashed line in the upper right panel shows
-the results for an implementation using \fwdpy{} where the ARG simplification steps were handled in a separate thread of
+panels shows run times tracking the pedigree instead of the neutral mutations. The dashed line in the upper right panel shows
+the results for an implementation using \fwdpy{} where the pedigree simplification steps were handled in a separate thread of
 execution and fitness calculations were parallelized.}
@@
     \includegraphics[]{sims/speedup}
-    \caption{\label{fig:relative_speedup_selection}Relative performance improvement due to ARG tracking. This figure
-    shows the ratio of the run time tracking neutral mutations over the run time tracking the ARG. Data points are taken
+    \caption{\label{fig:relative_speedup_selection}Relative performance improvement due to pedigree tracking. This figure
+    shows the ratio of the run time tracking neutral mutations over the run time tracking the pedigree. Data points are taken
 from Figure~\ref{fig:runtimes_selection} for simulations that ran to completion using both methods.}
@@
 %
-% Maybe an estimate of how long \emph{just} the ARG recording and simplification takes,
+% Maybe an estimate of how long \emph{just} the pedigree recording and simplification takes,
 % so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
@@
     \includegraphics[]{sims/speedup_nosel}
-    \caption{\label{sfig:speedup_nosel}Relative performance improvement due to ARG tracking for simulations without
+    \caption{\label{sfig:speedup_nosel}Relative performance improvement due to pedigree tracking for simulations without
     selection.  Data points are taken from Figure~\ref{sfig:rawspeed_nosel} and show the ratio of run times tracking
-neutral variation over the run times tracking the ARG.}
+neutral variation over the run times tracking the pedigree.}
 \end{figure}

### 2017-12-22 sims/quick_plot.py
get rid of shared y scale for relative speed up plots

@@ -74,7 +74,7 @@ joined = data_arg.merge(data_noarg, on=[
 joined['speedup'] = joined['time_noarg'] / joined['time_arg']
 groups = joined.groupby(['engine', 'N'])
 fig, (ax_fwdpp, ax_simupop) = plt.subplots(
-    2,  sharex=True, sharey=True)
+    2,  sharex=True, sharey=False)
 for name, group in groups:
     lstyle = 'solid'
     if name[0] == 'fwdpy11':
@@ -151,7 +151,7 @@ joined = data_arg.merge(data_noarg, on=[
 joined['speedup'] = joined['time_noarg'] / joined['time_arg']
 groups = joined.groupby(['engine', 'N'])
 fig, (ax_fwdpp, ax_simupop) = plt.subplots(
-    2,  sharex=True, sharey=True)
+    2,  sharex=True, sharey=False)
 for name, group in groups:
     lstyle = 'solid'
     if name[0] == 'fwdpy11':

### 2017-12-22 forwards_paper.tex
more comments on how big a speedup we get.

@@
 For parameters that ran to completion when tracking neutral mutations and when tracking
-the pedigree, the relative speedup due to pedigree tracking is up to $\approx 50$ fold (Figure~\ref{fig:relative_speedup_selection}).
-For \fwdpy{}, pedigree tracking reduces runtimes more for larger population sizes
-(Figure~\ref{fig:relative_speedup_selection}).  We could not make such an observation with \simupop{}, which tended to
-run out of memory on our system.  Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
+the pedigree, the relative speedup due to pedigree tracking is up to $\approx 50$ fold
+(Figure~\ref{fig:relative_speedup_selection}).  For \fwdpy{}, pedigree tracking reduces runtimes more for larger
+population sizes and we observe increasing relative speedups as $4Nr$ increases
+(Figure~\ref{fig:relative_speedup_selection}).  We saw a similar trend with $4Nr$ for \simupop{}, but the effect of $N$
+could not be explored because the simulations tended to run out of available memory.
+  Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
 partially obscures an important observation that the run times appear approximately linear in region size when pedigree

### 2017-12-23 sims/quick_plot.py
make choose from viridis extremes to differentiate lines a bit better

@@ -12,7 +12,7 @@ fp11['engine'] = ['fwdpy11'] * len(fp11.index)
 data = pd.concat([spop, fp11])
 
 cmap = matplotlib.cm.get_cmap('viridis')
-lcolors = [cmap(0.25), cmap(0.5), cmap(0.75)]
+lcolors = [cmap(0), cmap(125), cmap(cmap.N-25)]
 colors = {}
 i = 0
 for x in sorted(data.N.unique()):

### 2017-12-27 sims/quick_plot.py
use multiple marker points in plots

@@ -18,7 +18,12 @@ i = 0
 for x in sorted(data.N.unique()):
     colors[x] = lcolors[i]
     i = i + 1
-
+ptypes = ['o','^','x']
+points = {}
+i=0
+for x in sorted(data.N.unique()):
+    points[x] = ptypes[i]
+    i = i+1
 
 data_neut = data[data['sel'] == False].copy(deep=True)
 
@@ -29,12 +34,12 @@ groups = data.groupby(['engine', 'arg', 'queue', 'N'])
 fig, ((ax_fwdpp, ax_fwdpp_arg), (ax_simupop, ax_simupop_arg)) = plt.subplots(
     2, 2, sharex=True, sharey=True)
 for name, group in groups:
-    lstyle = 'solid'
+    lstyle = '-'
     if name[0] == 'fwdpy11' and name[1] is False:
         ax = ax_fwdpp
     if name[0] == 'fwdpy11' and name[1] is True:
         ax = ax_fwdpp_arg
-    m = 'o'
+    m = points[name[3]]
     if name[0] == 'simuPOP':
         ax = ax_simupop
         if name[1] is True:

### 2017-12-27 forwards_paper.tex
move fig code around
add todo for myself
delete comment
replace Allee effect with what it is

@@
 Furthermore, the relationship between the life history of a species --
-fecundity and mortality schedules, Allee effects, and demographic fluctuations --
+fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
 are all absorbed into a single compound parameter, the coalescence rate.
@@
 
-\plr{Insert figures/tables of performance here and describe them.}
-
 The total run times for simulations with selection are shown in Figure~\ref{fig:runtimes_selection}.  When tracking
@@
 
-We also performed a more limited set of simulations without natural selection.  The total run times are shown in
-Figure~\ref{sfig:rawspeed_nosel} and show the same qualitative behavior as simulations with selection
-(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to pedigree tracking is again substantial in these
-simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to pedigree tracking here with \fwdpy{}
-than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
-close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
-more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
-and~\ref{sfig:speedup_nosel}.
-
 \begin{figure}
@@
 
+We also performed a more limited set of simulations without natural selection.  The total run times are shown in
+Figure~\ref{sfig:rawspeed_nosel} and show the same qualitative behavior as simulations with selection
+(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to pedigree tracking is again substantial in these
+simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to pedigree tracking here with \fwdpy{}
+than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
+close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
+more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
+and~\ref{sfig:speedup_nosel}.
+
 \begin{figure}
@@
 
+\krt{I should discuss RAM use here}
 % Comparison of simulation with/without msprime, using \simupop{}

### 2017-12-27 forwards_paper.tex
clarify that it is large regions = large run times in large pops

@@
 In practice, current state-of-the-art simulation software may take on the order of
-weeks to simulate models without selection~\citep{fwdpp,Hernandez2015-wf},
+weeks to simulate models of large genomic regions without selection~\citep{fwdpp,Hernandez2015-wf},
 and existing simulation engines differ in how efficiently they

### 2017-12-27 forwards_paper.tex
there is no such thing as a neutral site
cite msprime paper at 1st occurrence

@@
 However, it is thought that most genetic variation is selectively neutral (or nearly so).
-By definition, the alleles carried by individuals in a population at neutral sites
+By definition, the neutral alleles carried by individuals in a population
 do not affect the population process.
@@
 In this paper, we describe how algorithmic tools and data structures developed for the
-coalescent simulator \msprime{}
+coalescent simulator \msprime{} \citep{kelleher2016efficient}
 can be used to efficiently record, and later process, this history.

### 2017-12-27 forwards_paper.tex
add missing words

--- 2017-12-27 forwards_paper.tex
@@
 space to store the resulting tree sequence for a single crossover per
-generation Wright-Fisher model.
+generation under the Wright-Fisher model.
 

### 2017-12-27 forwards_paper.tex
clarify units

@@
 equal to the recombination rate,
-so $\mu = r$, where $\mu$ and $r$ are the expected number of mutations and recombinations,
-respectively, per gamete and per generation.
+so $\mu = r$, where $\mu$ and $r$ are the expected number of mutations per gamete and recombination breakpoints per
+diploid,
+respectively, each generation. 
 % Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,

### 2017-12-27 forwards_paper.tex
reword sentence on what we varied

@@
 % and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
-We ran simulations with different values of $N$
-and of the \emph{scaled} parameter $\rho = 4Nr$,
-categorized by the value of $\rho$, since this
-is what most strongly determines run time
-(along with the scaled mutation rate, $\theta = 4N\mu$, which is also equal to $\rho$.)
+We ran simulations with different values of $N$ and varied the size of the region according to the scaled recombination
+parameter $\rho = 4Nr$.  When simulating neutral mutations, we kept the scaled mutation parameter, $\theta = 4N\mu$
+equal to $\rho$.
 

### 2017-12-27 sims/quick_plot.py
apply new points to rest of figs

@@ -84,7 +84,7 @@ for name, group in groups:
     lstyle = 'solid'
     if name[0] == 'fwdpy11':
         ax = ax_fwdpp
-    m = 'o'
+    m = points[name[1]]
     if name[0] == 'simuPOP':
         ax = ax_simupop
     mfacecolor = colors[name[1]]
@@ -116,7 +116,7 @@ for name, group in groups:
     if name[0] == 'fwdpy11' and name[1] is True:
         print(group)
         ax = ax_fwdpp_arg
-    m = 'o'
+    m = points[name[3]]
     if name[0] == 'simuPOP':
         ax = ax_simupop
         if name[1] is True:
@@ -161,7 +161,7 @@ for name, group in groups:
     lstyle = 'solid'
     if name[0] == 'fwdpy11':
         ax = ax_fwdpp
-    m = 'o'
+    m = points[name[1]]
     if name[0] == 'simuPOP':
         ax = ax_simupop
     mfacecolor = colors[name[1]]

### 2017-12-27 forwards_paper.tex
define fwdpy11 before we start using it below
add comment
clarification of Fig 2

@@
         In C++, using \fwdpp{} library functions and interface code (details below),
-        using a continuum-sites model for both mutation and recombination.
+        using a continuum-sites model for both mutation and recombination. Simulations were run using \fwdpy{}, a Python
+        package based on \fwdpp.
         % $\mu$ and $r$ were the means of Poisson processes,
@@
 Our simulation scripts separately recorded the time spent in various steps of each run.
+
+\krt{currently, we are not doing anything with the detailed timings.  I think that is fine, and that we can delete last
+sentence of preceding paragraph.}
 
@@
 (Figure~\ref{fig:relative_speedup_selection}).  For \fwdpy{}, pedigree tracking reduces runtimes more for larger
-population sizes and we observe increasing relative speedups as $4Nr$ increases
+population sizes and we observe increasing relative speedups as $4Nr$ increases for a given $N$
 (Figure~\ref{fig:relative_speedup_selection}).  We saw a similar trend with $4Nr$ for \simupop{}, but the effect of $N$

### 2017-12-27 forwards_paper.tex
clarify number of new edges--not a new result, but rather the standard calc.

@@
 tracking.  The reason for this behavior is that the simulations are generating a constant number of new edges on average
-each generation, and the number of edges is a function of $\rho$ in expectation.
+each generation as the number of new edges is Poisson distributed with expectation $\rho$.
 

### 2017-12-27 forwards_paper.tex
delete SI section and have the figs as part of appendix
define the notation close to where first used
add comment

@@
 Similarly, the function $\taddedge{\Et, \ell, r, p, c}$
-adds a new edge $(\ell, r, p, c)$ to the edge table $\Et$.
+adds a new edge $(\ell\mathrm{eft}, r\mathrm{ight}, p\mathrm{arent}, c\mathrm{hild})$ to the edge table $\Et$.
 The function $\tsimplify{P, \Nt, \Et}$ (described in section \ref{ss:simplify})
@@
 \end{figure}
+
+\krt{The following paragraph has a sentence mangling.  I think some comment is warrented about when it is useful to
+record selected mutations, like for ancient samples?}
 
@@
 
-\newpage
-\clearpage
-\section*{Supplementary Information}
-\renewcommand{\thefigure}{S\arabic{figure}}
+%% \newpage
+%% \clearpage
+%% \section*{Supplementary Information}
+\renewcommand{\thefigure}{A\arabic{figure}}
 \setcounter{figure}{0}

### 2017-12-27 forwards_paper.tex
minor rewording

@@
 while on the right
-is the minimal tree representing the ancestry of the currently alive population.
+is the minimal tree representing the ancestry of the current generation of the population.
 Clearly there is a great deal of redundancy in the topological

### 2017-12-27 forwards_paper.tex
add missing word

@@
 Conceptually, this works by
-(a) beginning by painting the chromosome each sample a distinct color;
+(a) beginning by painting the chromosome in each sample a distinct color;
 (b) moving back through history,

### 2017-12-27 forwards_paper.tex
add comment
add into para to run time section

@@
 
+\krt{Empirically, I did find that simplifying more often than every 1,000 generations gave a run-time hit of about 25
+percent???}
+
 %%%%%%%%%%%%
@@
 
-\plr{Refer to simulation results here.}
+The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
+pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
+near-linear in $rho$ (Figure \ref{fig:runtimes_selection}).  In this section, we explore the run-time complexity using a
+simplified Wright-Fisher model with recombination.
 

### 2017-12-27 forwards_paper.tex
clarify that rec breaks != edges

@@
 partially obscures an important observation that the run times appear approximately linear in region size when pedigree
-tracking.  The reason for this behavior is that the simulations are generating a constant number of new edges on average
-each generation as the number of new edges is Poisson distributed with expectation $\rho$.
+tracking.  The reason for this behavior is that the simulations are generating a constant number of new edges on
+average
+each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
 

### 2017-12-27 forwards_paper.tex
make clear we are approximating

@@
 of only the current generation of $2N$ individuals.
-This tree sequence has $4N-2$ edges to describe the leftmost tree;
+Ignoring the possibility of coalescent events involving more than two lineages, this tree sequence has $4N-2$ edges to describe the leftmost tree;
 and each time the marginal tree changes along the sequence,

### 2017-12-27 forwards_paper.tex
get rid of sentence starting with This
get rid of another sentence starting with This

@@
 \end{align}
-This holds up quite well in practice,
+These calculations hold up quite well in practice,
 as shown in Figure~\ref{fig:simplify_complexity}.
@@
 and so multiplies this by a constant factor.
-This implies that propagating neutral genotypes for $T$ generations has time complexity $O(\mu T N^2)$,
+These considerations imply that propagating neutral genotypes for $T$ generations has time complexity $O(\mu T N^2)$,
 while our implementation is only $O(T N \log(2N))$.

### 2017-12-27 forwards_paper.tex
get rid of yet another sentence starting with This

@@
 leading to advantages for downstream analysis as well.
-This is because many algorithms to compute statistics of interest for population genetics
+The dowstream advantages are because many algorithms to compute statistics of interest for population genetics
 are naturally expressed in terms of tree topologies,

### 2017-12-27 forwards_paper.tex
add comment
a bit more on parallellization

@@
 
+\krt{The following para contains numbers that are results, not discussion.}
+
 The tree sequences produced by default by this method
@@
 Recently, \citet{lawrie2017accelerating} used parallelization
-to improve performance for the ``Poisson Random Field'' family of models~\citep{Sawyer1992-jw},
+to improve performance for the ``Poisson Random Field'' (PRF) family of models~\citep{Sawyer1992-jw},
 in which mutations are unlinked and there is no effect of genetic background on fitness.
-However, thus far little attention has been paid to parallelization of simulations involving linkage.
+However, thus far little attention has been paid to parallelization of simulations involving linkage, which is much more
+complex than the PRF model.
 

### 2017-12-27 sims/plot_RAM.py
figs for RAM use

@@ -1,0 +1,112 @@
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+import pandas as pd
+
+spop = pd.read_csv("simupop_timings.txt", sep="\t")
+fp11 = pd.read_csv("cpp_benchmarks/cpp_timings.txt", sep="\t")
+
+spop['engine'] = ['simuPOP'] * len(spop.index)
+fp11['engine'] = ['fwdpy11'] * len(fp11.index)
+
+data = pd.concat([spop, fp11])
+
+cmap = matplotlib.cm.get_cmap('viridis')
+lcolors = [cmap(0), cmap(125), cmap(cmap.N-25)]
+colors = {}
+i = 0
+for x in sorted(data.N.unique()):
+    colors[x] = lcolors[i]
+    i = i + 1
+ptypes = ['o','^','x']
+points = {}
+i=0
+for x in sorted(data.N.unique()):
+    points[x] = ptypes[i]
+    i = i+1
+
+data_neut = data[data['sel'] == False].copy(deep=True)
+
+# Plot out the run times for the simulations with selection
+data = data[data['sel'] == True]
+data.sort_values(by='rho', inplace=True)
+groups = data.groupby(['engine', 'arg', 'queue', 'N'])
+fig, ((ax_fwdpp, ax_fwdpp_arg), (ax_simupop, ax_simupop_arg)) = plt.subplots(
+    2, 2, sharex=True, sharey=False)
+for name, group in groups:
+    lstyle = '-'
+    if name[0] == 'fwdpy11' and name[1] is False:
+        ax = ax_fwdpp
+    if name[0] == 'fwdpy11' and name[1] is True:
+        ax = ax_fwdpp_arg
+    m = points[name[3]]
+    if name[0] == 'simuPOP':
+        ax = ax_simupop
+        if name[1] is True:
+            ax = ax_simupop_arg
+    mfacecolor = colors[name[3]]
+    if name[2] is True:
+        lstyle = 'dashed'
+        mfacecolor = 'none'
+    popsize = int(name[3])
+    ax.plot(group.rho, group.mem / (1024.**2.), marker=m,
+            ms=4, linestyle=lstyle, label=r'$N = {:.0e}$'.format(popsize),
+            color=colors[name[3]],
+            markerfacecolor=mfacecolor)
+
+ax_fwdpp.legend(loc='best',frameon=False)
[…]

### 2017-12-27 sims/quick_plot.py
minor tweaks to main benchmark plots

@@ -54,7 +54,7 @@ for name, group in groups:
             color=colors[name[3]],
             markerfacecolor=mfacecolor)
 
-ax_fwdpp.legend(loc='best')
+ax_fwdpp.legend(loc='upper left',frameon=False)
 ax_fwdpp.set_title("fwdpy11 with neutral mutations", fontsize='medium')
 ax_fwdpp_arg.set_title("fwdpy11 with pedigree tracking", fontsize='medium')
 ax_simupop.set_ylabel("Run time (hours)")
@@ -100,7 +100,7 @@ ax_fwdpp.set_title("fwdpy11", fontsize='medium')
 ax_simupop.set_title("simuPOP", fontsize='medium')
 ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop.set_xscale('log')
-ax_fwdpp.legend(loc='best')
+ax_fwdpp.legend(loc='best',frameon=False)
 fig.tight_layout()
 plt.savefig("speedup.pdf")
 
@@ -131,7 +131,7 @@ for name, group in groups:
             color=colors[name[3]],
             markerfacecolor=mfacecolor)
 
-ax_fwdpp.legend(loc='upper left')
+ax_fwdpp.legend(loc='upper left',frameon=False)
 ax_fwdpp.set_title("fwdpy11 with neutral mutations", fontsize='medium')
 ax_fwdpp_arg.set_title("fwdpy11 with pedigree tracking", fontsize='medium')
 ax_simupop.set_ylabel("Run time (hours)")
@@ -177,6 +177,6 @@ ax_fwdpp.set_title("fwdpy11", fontsize='medium')
 ax_simupop.set_title("simuPOP", fontsize='medium')
 ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop.set_xscale('log')
-ax_fwdpp.legend(loc='best')
+ax_fwdpp.legend(loc='best',frameon=False)
 fig.tight_layout()
 plt.savefig("speedup_nosel.pdf")

### 2017-12-27 forwards_paper.tex
add figs on RAM use and discuss them

@@
 
-\krt{I should discuss RAM use here}
+The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel} aand for the
+simulations without selection in Figure~\ref{sfig:ramnosel}.  The general picture is that pedigree tracking requires
+more RAM.  With \fwdpy{}/\fwdpp{}, the reason for this behavior is clear.  When pedigree tracking, a non-recombinant
+individual still requires one new edge and two new nodes to be allocated.  Without pedigree tracking, no new memory is
+allocated for such an offspring.  Thus, pedigree tracking accumulates a large amount of data in RAM until
+simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extram RAM needed
+can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
+require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four epochs of unsimplified data are
+stored in RAM.
+
+\krt{I've avoided getting into how this much RAM means parallelization is required to make more progress.}
+
 % Comparison of simulation with/without msprime, using \simupop{}
@@
 
+\begin{figure}
+    \includegraphics[]{sims/memuse}
+    \caption{\label{sfig:ramsel}Peak RAM use in gigabytes (GB) for simulations with selection.  The plots are from the
+        same data as in Figure \ref{fig:runtimes_selection}.}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[]{sims/memuse_nosel}
+    \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
+    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
+\end{figure}
+
 \end{document}

### 2017-12-28 forwards_paper.tex
methods place holder
comment out bit about detailed timings
make clear what we are talking about re: performance

@@
 and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
-Our simulation scripts separately recorded the time spent in various steps of each run.
-
-\krt{currently, we are not doing anything with the detailed timings.  I think that is fine, and that we can delete last
-sentence of preceding paragraph.}
+%% Our simulation scripts separately recorded the time spent in various steps of each run.
+
+%% \krt{currently, we are not doing anything with the detailed timings.  I think that is fine, and that we can delete last
+%%sentence of preceding paragraph.}
 
@@
   Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
-partially obscures an important observation that the run times appear approximately linear in region size when pedigree
-tracking.  The reason for this behavior is that the simulations are generating a constant number of new edges on
+partially obscures an important observation that the run times appear approximately linear in region size when tracking
+pedigree.  The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on
 average
@@
 
+\krt{Are we doing a methods section?}
+
+\subsection{Reagents list for the C++ implementation}
+
+\begin{enumerate}
+    \item \fwdpy{} 0.13.a0
+    \item \fwdpp{} 0.5.7
+    \item GNU Scientific Library 1.16
+    \item pybind11 2.2.1
+    \item GCC 4.8.5
+\end{enumerate}
+
+All of the above installed via conda.
+
 %%%%%%%%%%%%%%%%%%%%%%

### 2017-12-28 forwards_paper.tex
clarify RAM use discussion

@@
 The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel} aand for the
-simulations without selection in Figure~\ref{sfig:ramnosel}.  The general picture is that pedigree tracking requires
-more RAM.  With \fwdpy{}/\fwdpp{}, the reason for this behavior is clear.  When pedigree tracking, a non-recombinant
+simulations without selection in Figure~\ref{sfig:ramnosel}.  For our parameters, which involved simplifying every
+$10^3$ generations,  pedigree tracking requires more RAM.  With \fwdpy{}/\fwdpp{}, the reason for the extran RAM use is clear.  When pedigree tracking, a non-recombinant
 individual still requires one new edge and two new nodes to be allocated.  Without pedigree tracking, no new memory is
@@
 require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four epochs of unsimplified data are
-stored in RAM.
-
-\krt{I've avoided getting into how this much RAM means parallelization is required to make more progress.}
+stored in RAM.  In practice, the RAM consumption can be minimized by simplifying more often.
+
+%% \krt{I've avoided getting into how this much RAM means parallelization is required to make more progress.}
 

### 2017-12-28 forwards_paper.tex
update to say current set of samples, as that allows for overlapping gens, ancient samples, and whatnot

@@
 while on the right
-is the minimal tree representing the ancestry of the current generation of the population.
+is the minimal tree representing the ancestry of the current set of samples.
 Clearly there is a great deal of redundancy in the topological

### 2017-12-28 forwards_paper.tex
update pseudo-methods

@@
 providing an easy, efficient way to specify prior history.
-A natural question is now: how often should simplification occur?
-As shown in the next section,
-there is no computational advantage to simplifying more often than is necessary
-to keep memory usage down.
-
-\krt{Empirically, I did find that simplifying more often than every 1,000 generations gave a run-time hit of about 25
-percent???}
+A natural question is now: how often should simplification occur?  We did some limited testing with the C++
+implementation of our simulations, and found that different simplification intervals affect run times by approximately
+25\%, with the lowest run time occuring when simplifying every $10^3$ generations.  Thus, there is a memory-versus-speed
+tradeoff as simplifying more often would keep fewer extinct nodes and edges in memory.  The run-time costs associated
+with simplifying have to do with preparing the data to be sent to the simplification algorithm, rather than with the
+algorithm itself, which we show in the next section has linear time complexity.
+
+% As shown in the next section,
+% there is no computational advantage to simplifying more often than is necessary
+% to keep memory usage down.
+% 
+% \krt{Empirically, I did find that simplifying more often than every 1,000 generations gave a run-time hit of about 25
+% percent???}
 
@@
     \item GCC 4.8.5
+    \item msprime master branch
 \end{enumerate}
 
-All of the above installed via conda.
+All of the above installed via conda, except for msprime, which must be installed from GitHub.
 

### 2017-12-28 forwards_paper.tex
fix typo
add comment
add comment that algos look good

@@
 a list of ancestors
-each of which is a collections of ancestral segments.
+each of which is a collection of ancestral segments.
 An ancestral segment is a triple $(\ell, r, x)$,
@@
 }
-
+\krt{We need a verbal definition of $Q$}
 \algstep{S1.}{Input parent loop.}{Set $Q \leftarrow \emptyset$.
@@
 \end{taocpalg}
+\krt{I think the algo discussion is  clear.  I've not tried to implement it, which would be the real test, but I find the description fine other than needing to define $Q$ when first used.}
 
@@
 \end{figure}
-
 \end{document}

### 2017-12-28 forwards_paper.tex
epochs -> sets

@@
 can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
-require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four epochs of unsimplified data are
+require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four sets of unsimplified data are
 stored in RAM.  In practice, the RAM consumption can be minimized by simplifying more often.

### 2017-12-28 forwards_paper.tex
update bit on parallelization

@@
 in which mutations are unlinked and there is no effect of genetic background on fitness.
-However, thus far little attention has been paid to parallelization of simulations involving linkage, which is much more
+However, thus far little attention has been paid to parallelization of simulations involving linkage, as our simulations
+where do, which is much more
 complex than the PRF model.

### 2017-12-28 forwards_paper.tex
not on y-axis scales

@@
     \caption{\label{sfig:ramsel}Peak RAM use in gigabytes (GB) for simulations with selection.  The plots are from the
-        same data as in Figure \ref{fig:runtimes_selection}.}
+        same data as in Figure \ref{fig:runtimes_selection}. Note that the y-axis scales differ considerably across
+    panels.}
 \end{figure}
@@
     \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
-    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
+    the same data as in Figure \ref{sfig:rawspeed_nosel}. Note that the y-axis scales differ considerably across panels.}
 \end{figure}

### 2017-12-29 forwards_paper.tex
oops--y axis same here

@@
 
-The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel} aand for the
+The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel} and for the
 simulations without selection in Figure~\ref{sfig:ramnosel}.  For our parameters, which involved simplifying every
-$10^3$ generations,  pedigree tracking requires more RAM.  With \fwdpy{}/\fwdpp{}, the reason for the extran RAM use is clear.  When pedigree tracking, a non-recombinant
+$10^3$ generations,  pedigree tracking requires more RAM.  With \fwdpy{}/\fwdpp{}, the reason for the extra RAM use is clear.  When pedigree tracking, a non-recombinant
 individual still requires one new edge and two new nodes to be allocated.  Without pedigree tracking, no new memory is
 allocated for such an offspring.  Thus, pedigree tracking accumulates a large amount of data in RAM until
-simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extram RAM needed
+simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extra RAM needed
 can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
@@
     \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
-    the same data as in Figure \ref{sfig:rawspeed_nosel}. Note that the y-axis scales differ considerably across panels.}
+    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
 \end{figure}

### 2017-12-30 forwards_paper.tex
Add simupop reqs

@@
     \item
-        In Python, using the \simupop{} library interfacing with a custom Python library (\ftprime{}),
+        In Python, using the \simupop{} library interfacing with Python code
+	(via our package \ftprime{}),
         using a discrete but large number of sites.
@@
         at each point in time,
-        in the example of Figure~\ref{fig:method_diagram}.
+        in the example of Figure~\ref{fig:method_diagram}A.
+        As in that figure, blue numbers denote the crossing over location in a meiosis --
+        for instance, $G$ and $H$ were parents to $I$,
+        who inherited the left 70\% of the chromosome from $G$ and the remainder from $H$.
         \label{fig:simplify_state}
@@
 from the memory of a producer without incurring the overhead of a copy.
+\jda{TODO citation missing}
 Our implementation uses the numpy C API \citep{numpy_C_api} to efficiently copy
@@
 such as HDF5 \citep{hdf5}, Dask, or Zarr.
+\jda{TODO citation missing}
 (For small scale data and debugging purposes, a simple text format is also supported.)
@@
 
+\subsection{Requirements for the \simupop{} implementation}
+
+The simulations are implemented the Python package \ftprime{} 0.0.6rc0
+which is available at \url{https://www.github.com/ashander/ftprime}.
+Its requirements are encoded in an \texttt{environment.yml} file, compatible
+with conda, but \msprime{} must be updated to the version listed below via pip
+from GitHub \jda{TODO add command}.
+In brief the requirements are:
+
+\begin{enumerate}
+    \item \msprime{} \jda{TODO add version from KRT's dev machine}
+    \item \simupop{} 1.1.8.3
+    \item python 3.5
+\end{enumerate}
+
 %%%%%%%%%%%%%%%%%%%%%%

### 2017-12-30 forwards_paper.tex
Revise intro aiming for clearer delination
between prior work, motivating concepts and our results

@@
 \newcommand{\Al}{\mathcal{A}}  % list of ancestors
-\newcommand{\taddnode}[1]{\mbox{addnode}(#1)}
-\newcommand{\taddedge}[1]{\mbox{addedge}(#1)}
-\newcommand{\tsimplify}[1]{\mbox{simplify}(#1)}
+% mbox to avoid hyphenating
+\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
+\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
+\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
 \DeclareMathOperator{\len}{length}
@@
     but saves substantially more information from the simulation
-    (all marginal genealogies).
+    (all marginal genealogies) than simply recording the final sequences of
+    simulated individuals.
+    \jda{ this is still only 120 words and we should say 'more than what'}
     % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
@@
 and (c) small sample size relative to the population size.
-The first two assumptions can be side-stepped to a limited extent,
+The first two assumptions can be side-stepped to a limited extent \jda{cite?},
 but it remains a challenge to map results of coalescent models
@@
 -- and often, the point of such simulations is to compare to real data --
-then such simulations must somehow produce at the end data for each individual on a genomic scale.
+then such simulations must record data for each individual genome.
 As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
@@
 a population of size $N$ must be simulated across many multiples of $N$ generations
-to produce stable genetic patterns.
-Because of this computational burden of neutral mutations, even the fastest simulation frameworks such as SLiM,
+to produce stable genetic patterns \jda {citation?}.
+Because of this computational burden, even the fastest simulation frameworks such as SLiM,
+\jda{We haven't motivatd that neutral muts are the burden yet}
 SLiM2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
@@
 at each position along the genome.
-In this paper, we describe how algorithmic tools and data structures developed for the
-coalescent simulator \msprime{}
-can be used to efficiently record, and later process, this history.
-In so doing we record the \emph{population pedigree} --
+This \emph{population pedigree} is 
 the complete history of parent-offspring relationships of an entire population
@@
 meiosis.
-This embellished pedigree contains all the information necessary
-to construct the genealogical tree that relates each individual to each other
-at each position on the genome, i.e., the \emph{tree sequence}~\citep{kelleher2016efficient}.
+The embellished pedigree contains all the information necessary
+to construct the genealogical tree that relates each individual to every other
+at each position on the genome,
+i.e., the \emph{tree sequence} implemented in the coalescent simulator \msprime{}~\citep{kelleher2016efficient}.
+\jda{Currently (even before my edit) we no longer mention the ARG equivalence *at all* in the intro which
+	seems wrong, but we might've hashed this over already? (and it's
+	mentioned later so maybe no matter)}
 Combined with ancestral genotypes and the origins of new mutations,
 it also completely specifies the genomic sequence of any individual in the population at any time.
-This is much more than we need to know, however,
-so we discard all information irrelevant to the genetic history
-of the extant population. We show that we need only $O(N \log N)$
-space to store the resulting tree sequence for a single crossover per
-generation under the Wright-Fisher model.
-
-In this paper we discuss storage methods for genealogies (and hence, genome sequence),
-an algorithm for simplifying these,
-and their use in forwards-time simulation.
+This is much much more than typically recorded in a forward simulation, however,
+so further savings can be had by discarding all information irrelevant to the genetic history
+of the extant population.
+\jda{The space complexity result seemed like a result in the intro; and maybe didn't include some detail
+	about assumed number of generations}
+
+In this paper, we describe storage methods for genealogies
+(and hence, genome sequence) and an algorithm for simplifying these,
+both implemented as improvements to the algorithmic tools and data structures in \msprime{}.
+We also describe a general method of using these tools to efficiently record,
+and later process, the population pedgiree in forwards-time simulation.
+While providing substantial savings in time and space, our methods provide much
+more information than simply simulating the genomes -- the tree sequence format
+encodes all marginal genealogies of individuals living at the end of the
+simulation.
 Although we were motivated by a need for more efficient genomic simulations,
@@
 \algstep{W1.}{Initialisation.}{
-Set $\Nt \leftarrow \mbox{NodeTable}()$, $\mathcal{E}
-\leftarrow \mbox{EdgeTable}()$, $t \leftarrow T$, and $j \leftarrow 0$.
+Set $\Nt \leftarrow \mbox{\texttt{NodeTable}}()$, $\mathcal{E}
+\leftarrow \mbox{\texttt{EdgeTable}}()$, $t \leftarrow T$, and $j \leftarrow 0$.
  For $0 \leq k < N$, set $P_k \leftarrow \taddnode{\Nt, T}$.

### 2017-12-30 forwards_paper.tex
Tweak time complexity statement per #37

@@
 \newcommand{\fwdpy}{\texttt{fwdpy11}}
+\newcommand{\cpp}{\texttt{C++}}
 \newcommand{\msprime}{\texttt{msprime}}
@@
 we implemented these simulations separately using two different simulation engines:
+
+\jda{Note/TODO: the varying references to c++, python, fwdpp, fwdpy11 and
+	simupop are probably confusing to a new reader. This is in part because
+	we refer to the python/simupop method in various places as one or the
+	other; the same goes for c++/fwdpy11/fwdpp. We should decide what to
+	call these and make our usage consistent}
+
 \begin{enumerate}
@@
     \item
-        In C++, using \fwdpp{} library functions and interface code (details below),
+        In \cpp{}, using \fwdpp{} library functions and interface code (details below),
         using a continuum-sites model for both mutation and recombination. Simulations were run using \fwdpy{}, a Python
         package based on \fwdpp.
-        % $\mu$ and $r$ were the means of Poisson processes,
-        % and mutations used the infinite-sites model \citep{Kimura1969-uc}.
+        % $\mu$ and $r$ were the means of Poisson processes, and mutations used the infinite-sites model \citep{Kimura1969-uc}.
         % We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.
@@
 providing an easy, efficient way to specify prior history.
-A natural question is now: how often should simplification occur?  We did some limited testing with the C++
+A natural question is now: how often should simplification occur?  We did some limited testing with the \cpp{}
 implementation of our simulations, and found that different simplification intervals affect run times by approximately
@@
 Simply because it must process each edge,
-the simplification algorithm is linear in the number of edges of the input tree sequence.
-This is seen in Figure~\ref{fig:simplify_complexity}X,
+% below per discussion
+% https://github.com/petrelharp/ftprime_ms/pull/37#discussion_r156617283
+the simplification algorithm is at least linear in the number of edges of the input tree sequence.
+Empirically the algorithm is exactly linear,
+as seen in Figure~\ref{fig:simplify_complexity}X,
 where we show shows the time required to simplify increasingly large subsets of a large tree sequence.
@@
 For example, in the benchmarking simulations above,
-we used the \fwdpp{} C++ API \citep{fwdpp}
+we used the \fwdpp{} \cpp{} API \citep{fwdpp}
 to fill arrays of simple structures representing the node and edge data.
@@
 Using \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}),
-the node/edge arrays on the C++ side are made visible, without copy, to Python as a multi-column numpy array,
+the node/edge arrays on the \cpp{} side are made visible, without copy, to Python as a multi-column numpy array,
 which is processed by the \msprime{} tables API.
@@
 % (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
-% API from a command-line C/C++ application via an embedded Python interpreter.)
+% API from a command-line C/\cpp{} application via an embedded Python interpreter.)
 
@@
 %%%%%%%%%%%%%%
-% \section*{Methods}
+\section*{Methods}
 
@@
 
-\subsection{Reagents list for the C++ implementation}
+\subsection{Requirements for the \cpp{} implementation}
 
@@
     \item GCC 4.8.5
-    \item msprime master branch
+    \item \msprime{} master branch \jda{which commit hash or short version number
+		    from pip freeze or {\tt msprime.\_\_version\_\_}?}
 \end{enumerate}

### 2017-12-30 forwards_paper.tex
Suggested tweaks to c++ methods section
& formatting of C++

@@
 we implemented these simulations separately using two different simulation engines:
-
-\jda{Note/TODO: the varying references to c++, python, fwdpp, fwdpy11 and
-	simupop are probably confusing to a new reader. This is in part because
-	we refer to the python/simupop method in various places as one or the
-	other; the same goes for c++/fwdpy11/fwdpp. We should decide what to
-	call these and make our usage consistent}
-
 \begin{enumerate}
@@
         package based on \fwdpp.
-        % $\mu$ and $r$ were the means of Poisson processes, and mutations used the infinite-sites model \citep{Kimura1969-uc}.
+        % $\mu$ and $r$ were the means of Poisson processes,
+        % and mutations used the infinite-sites model \citep{Kimura1969-uc}.
         % We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.
@@
 Simply because it must process each edge,
-% below per discussion
-% https://github.com/petrelharp/ftprime_ms/pull/37#discussion_r156617283
-the simplification algorithm is at least linear in the number of edges of the input tree sequence.
-Empirically the algorithm is exactly linear,
-as seen in Figure~\ref{fig:simplify_complexity}X,
+the simplification algorithm is linear in the number of edges of the input tree sequence.
+This is seen in Figure~\ref{fig:simplify_complexity}X,
 where we show shows the time required to simplify increasingly large subsets of a large tree sequence.

### 2017-12-30 forwards_paper.tex
Formatting tweak for msprime method/class names

--- 2017-12-30 forwards_paper.tex
@@
     but saves substantially more information from the simulation
-    (all marginal genealogies) than simply recording the final sequences of
-    simulated individuals.
-    \jda{ this is still only 120 words and we should say 'more than what'}
+    (all marginal genealogies).
     % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
@@
 and (c) small sample size relative to the population size.
-The first two assumptions can be side-stepped to a limited extent \jda{cite?},
+The first two assumptions can be side-stepped to a limited extent,
 but it remains a challenge to map results of coalescent models
@@
 -- and often, the point of such simulations is to compare to real data --
-then such simulations must record data for each individual genome.
+then such simulations must somehow produce at the end data for each individual on a genomic scale.
 As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
@@
 a population of size $N$ must be simulated across many multiples of $N$ generations
-to produce stable genetic patterns \jda {citation?}.
-Because of this computational burden, even the fastest simulation frameworks such as SLiM,
-\jda{We haven't motivatd that neutral muts are the burden yet}
+to produce stable genetic patterns.
+Because of this computational burden of neutral mutations, even the fastest simulation frameworks such as SLiM,
 SLiM2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
@@
 at each position along the genome.
-This \emph{population pedigree} is 
+In this paper, we describe how algorithmic tools and data structures developed for the
+coalescent simulator \msprime{}
+can be used to efficiently record, and later process, this history.
+In so doing we record the \emph{population pedigree} --
 the complete history of parent-offspring relationships of an entire population
@@
 meiosis.
-The embellished pedigree contains all the information necessary
-to construct the genealogical tree that relates each individual to every other
-at each position on the genome,
-i.e., the \emph{tree sequence} implemented in the coalescent simulator \msprime{}~\citep{kelleher2016efficient}.
-\jda{Currently (even before my edit) we no longer mention the ARG equivalence *at all* in the intro which
-	seems wrong, but we might've hashed this over already? (and it's
-	mentioned later so maybe no matter)}
+This embellished pedigree contains all the information necessary
+to construct the genealogical tree that relates each individual to each other
+at each position on the genome, i.e., the \emph{tree sequence}~\citep{kelleher2016efficient}.
 Combined with ancestral genotypes and the origins of new mutations,
 it also completely specifies the genomic sequence of any individual in the population at any time.
-This is much much more than typically recorded in a forward simulation, however,
-so further savings can be had by discarding all information irrelevant to the genetic history
-of the extant population.
-\jda{The space complexity result seemed like a result in the intro; and maybe didn't include some detail
-	about assumed number of generations}
-
-In this paper, we describe storage methods for genealogies
-(and hence, genome sequence) and an algorithm for simplifying these,
-both implemented as improvements to the algorithmic tools and data structures in \msprime{}.
-We also describe a general method of using these tools to efficiently record,
-and later process, the population pedgiree in forwards-time simulation.
-While providing substantial savings in time and space, our methods provide much
-more information than simply simulating the genomes -- the tree sequence format
-encodes all marginal genealogies of individuals living at the end of the
-simulation.
+This is much more than we need to know, however,
+so we discard all information irrelevant to the genetic history
+of the extant population. We show that we need only $O(N \log N)$
+space to store the resulting tree sequence for a single crossover per
+generation under the Wright-Fisher model.
+
+In this paper we discuss storage methods for genealogies (and hence, genome sequence),
+an algorithm for simplifying these,
+and their use in forwards-time simulation.
 Although we were motivated by a need for more efficient genomic simulations,

### 2017-12-30 forwards_paper.tex
Elaborate fig 5 caption

@@
 \newcommand{\Al}{\mathcal{A}}  % list of ancestors
-% mbox to avoid hyphenating
-\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
-\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
-\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
+\newcommand{\taddnode}[1]{\mbox{addnode}(#1)}
+\newcommand{\taddedge}[1]{\mbox{addedge}(#1)}
+\newcommand{\tsimplify}[1]{\mbox{simplify}(#1)}
 \DeclareMathOperator{\len}{length}
@@
     \item
-        In Python, using the \simupop{} library interfacing with Python code
-	(via our package \ftprime{}),
+        In Python, using the \simupop{} library interfacing with a custom Python library (\ftprime{}),
         using a discrete but large number of sites.
@@
 \algstep{W1.}{Initialisation.}{
-Set $\Nt \leftarrow \mbox{\texttt{NodeTable}}()$, $\mathcal{E}
-\leftarrow \mbox{\texttt{EdgeTable}}()$, $t \leftarrow T$, and $j \leftarrow 0$.
+Set $\Nt \leftarrow \mbox{NodeTable}()$, $\mathcal{E}
+\leftarrow \mbox{EdgeTable}()$, $t \leftarrow T$, and $j \leftarrow 0$.
  For $0 \leq k < N$, set $P_k \leftarrow \taddnode{\Nt, T}$.
@@
 
-\subsection{Requirements for the \simupop{} implementation}
-
-The simulations are implemented the Python package \ftprime{} 0.0.6rc0
-which is available at \url{https://www.github.com/ashander/ftprime}.
-Its requirements are encoded in an \texttt{environment.yml} file, compatible
-with conda, but \msprime{} must be updated to the version listed below via pip
-from GitHub \jda{TODO add command}.
-In brief the requirements are:
-
-\begin{enumerate}
-    \item \msprime{} \jda{TODO add version from KRT's dev machine}
-    \item \simupop{} 1.1.8.3
-    \item python 3.5
-\end{enumerate}
-
 %%%%%%%%%%%%%%%%%%%%%%

### 2017-12-30 forwards_paper.tex
Add note about consistent term for two implementations

--- 2017-12-30 forwards_paper.tex
@@
         package based on \fwdpp.
-        % $\mu$ and $r$ were the means of Poisson processes,
-        % and mutations used the infinite-sites model \citep{Kimura1969-uc}.
+        % $\mu$ and $r$ were the means of Poisson processes, and mutations used the infinite-sites model \citep{Kimura1969-uc}.
         % We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.

### 2017-12-30 forwards_paper.tex
Implement suggestion in #39 to put move fig details to appendix

@@
 
+Our method uses more RAM use than simple forward simulations (see~\ref{ss:memuse_nosel}),
+although in practice shorter simplification intervals may reduce RAM use
+(but at a slight cost to the speed up achieved, see~\ref{ss:seq_simp}).
+The benefits to our method are not limited to situations with selection:
+in a more limited set of neutral simulations we found the same qualitative
+behavior with a numerically larger benefit to pedigree tracking (see~\ref{ss:speedup_nosel}).
+
 \begin{figure}
@@
 \end{figure}
-
-The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel} and for the
-simulations without selection in Figure~\ref{sfig:ramnosel}.  For our parameters, which involved simplifying every
-$10^3$ generations,  pedigree tracking requires more RAM.  With \fwdpy{}/\fwdpp{}, the reason for the extra RAM use is clear.  When pedigree tracking, a non-recombinant
-individual still requires one new edge and two new nodes to be allocated.  Without pedigree tracking, no new memory is
-allocated for such an offspring.  Thus, pedigree tracking accumulates a large amount of data in RAM until
-simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extra RAM needed
-can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
-require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four sets of unsimplified data are
-stored in RAM.  In practice, the RAM consumption can be minimized by simplifying more often.
-
-%% \krt{I've avoided getting into how this much RAM means parallelization is required to make more progress.}
 
@@
 \subsubsection*{Sequential simplification and prior history}
+\label{ss:seq_simp}
 
@@
 %% \section*{Supplementary Information}
-\renewcommand{\thefigure}{A\arabic{figure}}
+
+\section{Memory usage with selection}
+\label{ss:memuse}
+\renewcommand{\thefigure}{B\arabic{figure}}
 \setcounter{figure}{0}
+
+The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel} and for the
+simulations without selection in Figure~\ref{sfig:ramnosel}.  For our parameters, which involved simplifying every
+$10^3$ generations,  pedigree tracking requires more RAM.  With \fwdpy{}/\fwdpp{}, the reason for the extra RAM use is clear.  When pedigree tracking, a non-recombinant
+individual still requires one new edge and two new nodes to be allocated.  Without pedigree tracking, no new memory is
+allocated for such an offspring.  Thus, pedigree tracking accumulates a large amount of data in RAM until
+simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extra RAM needed
+can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
+require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four sets of unsimplified data are
+stored in RAM.  In practice, the RAM consumption can be minimized by simplifying more often.
+
+%% \krt{I've avoided getting into how this much RAM means parallelization is required to make more progress.}
+
+\begin{figure}
+    \includegraphics[]{sims/memuse}
+    \caption{\label{sfig:ramsel}Peak RAM use in gigabytes (GB) for simulations with selection.  The plots are from the
+        same data as in Figure \ref{fig:runtimes_selection}. Note that the y-axis scales differ considerably across
+    panels.}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[]{sims/memuse_nosel}
+    \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
+    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
+\end{figure}
+
+\section{Timing with no selection}
+\label{ss:timing_nosel}
+\renewcommand{\thefigure}{C\arabic{figure}}
+\setcounter{figure}{0}
+
+We also performed a more limited set of simulations without natural selection.  The total run times are shown in
+Figure~\ref{sfig:rawspeed_nosel} and show the same qualitative behavior as simulations with selection
+(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to pedigree tracking is again substantial in these
+simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to pedigree tracking here with \fwdpy{}
+than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
+close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
+more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
+and~\ref{sfig:speedup_nosel}.
 
@@
 \end{figure}
-
-\begin{figure}
-    \includegraphics[]{sims/memuse}
-    \caption{\label{sfig:ramsel}Peak RAM use in gigabytes (GB) for simulations with selection.  The plots are from the
-        same data as in Figure \ref{fig:runtimes_selection}. Note that the y-axis scales differ considerably across
-    panels.}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[]{sims/memuse_nosel}
-    \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
-    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
-\end{figure}
 \end{document}

### 2017-12-30 forwards_paper.tex
Note missing cites

--- 2017-12-30 forwards_paper.tex
@@
         at each point in time,
-        in the example of Figure~\ref{fig:method_diagram}A.
-        As in that figure, blue numbers denote the crossing over location in a meiosis --
-        for instance, $G$ and $H$ were parents to $I$,
-        who inherited the left 70\% of the chromosome from $G$ and the remainder from $H$.
+        in the example of Figure~\ref{fig:method_diagram}.
         \label{fig:simplify_state}

### 2017-12-31 forwards_paper.tex
edit abstract

--- 2017-12-30 forwards_paper.tex
@@
     and show that not only does this speed up large simulations by one or two orders of magnitude,
-    but saves substantially more information from the simulation
-    (all marginal genealogies) than simply recording the final sequences of
-    simulated individuals.
-    \jda{ this is still only 120 words and we should say 'more than what'}
+    but saves not only full genotype sequences, but also all marginal
+    geneologies of the simulated individuals.
     % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.

### 2017-12-31 forwards_paper.tex
intro edit

@@
 and (c) small sample size relative to the population size.
-The first two assumptions can be side-stepped to a limited extent \jda{cite?},
+The first two assumptions can be side-stepped to a limited extent \citep{Neuhauser1997-nn, hudson1990gene},
 but it remains a challenge to map results of coalescent models
@@
 -- and often, the point of such simulations is to compare to real data --
-then such simulations must record data for each individual genome.
+then such simulations must record each individual's genome.
 As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
@@
 a population of size $N$ must be simulated across many multiples of $N$ generations
-to produce stable genetic patterns \jda {citation?}.
-Because of this computational burden, even the fastest simulation frameworks such as SLiM,
-\jda{We haven't motivatd that neutral muts are the burden yet}
+to produce stable genetic patterns \citep{wright1931evolution, wakeley2005coalescent}.
+Because of this computational burden, even the fastest simulation frameworks such as
 SLiM2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
@@
 i.e., the \emph{tree sequence} implemented in the coalescent simulator \msprime{}~\citep{kelleher2016efficient}.
-\jda{Currently (even before my edit) we no longer mention the ARG equivalence *at all* in the intro which
-	seems wrong, but we might've hashed this over already? (and it's
-	mentioned later so maybe no matter)}
 Combined with ancestral genotypes and the origins of new mutations,

### 2017-12-31 forwards_paper.tex
add comment on quantifying RAM use

@@
 
-Our method uses more RAM use than simple forward simulations (see~\ref{ss:memuse_nosel}),
+\jda{Maybe we should quantify the factor by which RAM use increased; it's not
+	easy to read off the figures}
+Our method uses more RAM (up to 2 orders of magnitude more) than simple forward simulations (see~\ref{ss:memuse_nosel}),
 although in practice shorter simplification intervals may reduce RAM use

### 2018-01-01 forwards_paper.tex
edit intro

@@
 at each position along the genome.
-This \emph{population pedigree} is 
+This is equivalent to knowing the \emph{population pedigree} --
 the complete history of parent-offspring relationships of an entire population
-going back to a remote time including the genetic outcomes of each ancestral
+going back to a remote time -- and the genetic outcomes of each ancestral
 meiosis.

### 2018-01-01 forwards_paper.tex

@@
 simulations without selection in Figure~\ref{sfig:ramnosel}.  For our parameters, which involved simplifying every
-$10^3$ generations,  pedigree tracking requires more RAM.  With \fwdpy{}/\fwdpp{}, the reason for the extra RAM use is clear.  When pedigree tracking, a non-recombinant
+$10^3$ generations,  pedigree tracking requires more RAM with \fwdpp{} and less
+RAM with \simupop{}. The reason for reduced RAM use with \simupop{} is unclear;
+with \fwdpy{}/\fwdpp{}, however the reason for the extra RAM use is clear.  When pedigree tracking, a non-recombinant
 individual still requires one new edge and two new nodes to be allocated.  Without pedigree tracking, no new memory is

### 2018-01-03 sims/gc_interval/plot.py
Add GC interval benchmarks

@@ -1,0 +1,97 @@
+import matplotlib
+import matplotlib.pyplot as plt
+import pandas as pd
+
+data = pd.read_csv("gc_interval_timings.txt", sep="\t")
+data['time'] /= (60.*60.)
+data['mem'] /= (1024.*1024.)
+
+gs = matplotlib.gridspec.GridSpec(1, 2)
+
+ax0 = plt.subplot(gs[0])
+ax1 = plt.subplot(gs[1])
+ax1.yaxis.tick_right()
+
+
+N1000 = data[data['N'] == 1000]
+
+N1000g = N1000.groupby('GC')
+
+X = 0
+for n, g in N1000g:
+    print(g)
+    ax0.scatter(g['size'], g['time'],
+                label=r'{}'.format(g['GC'].unique()[0]))
+    X += 1
+
+
+N10000 = data[data['N'] == 10000]
+
+N10000g = N10000.groupby('GC')
+
+for n, g in N10000g:
+    ax1.scatter(g['size'], g['time'])
+
+ax0.legend(loc='upper left', frameon = False,title="GC interval\n(generations)",fontsize='x-small')
+for ax in [ax0,ax1]:
+    print(sorted(data['size'].unique()))
+    ax.xaxis.set_ticks(sorted(data['size'].unique()))
+    ax.set_xscale('log')
+    ax.set_xlim((500,100000))
+    ax.set_xlabel(r'Region size ($\rho$ = 4Nr)')
+    ax.set_ylabel("Run time (hours)")
+
+
+ax0.set_title("N = 1,000",fontsize='medium')
+ax1.set_title("N = 10,000",fontsize='medium')
+ax1.yaxis.set_label_position("right")
+plt.tight_layout()
+
+plt.savefig("GCtime.pdf")
+
+plt.figure()
+
+gs = matplotlib.gridspec.GridSpec(1, 2)
+
+ax0 = plt.subplot(gs[0])
+ax1 = plt.subplot(gs[1])
+ax1.yaxis.tick_right()
+
+
+
+N1000 = data[data['N'] == 1000]
+
+N1000g = N1000.groupby('GC')
+
+X = 0
+for n, g in N1000g:
+    print(g)
+    ax0.scatter(g['size'], g['mem'],
+                label=r'{}'.format(g['GC'].unique()[0]))
+    X += 1
+
+
+N10000 = data[data['N'] == 10000]
+
+N10000g = N10000.groupby('GC')
+
+for n, g in N10000g:
[…]

### 2018-01-03 sims/gc_interval/plot.py
add lines to plots and autopep8 the plotting script

@@ -3,8 +3,8 @@ import matplotlib.pyplot as plt
 import pandas as pd
 
 data = pd.read_csv("gc_interval_timings.txt", sep="\t")
-data['time'] /= (60.*60.)
-data['mem'] /= (1024.*1024.)
+data['time'] /= (60. * 60.)
+data['mem'] /= (1024. * 1024.)
 
 gs = matplotlib.gridspec.GridSpec(1, 2)
 
@@ -20,8 +20,8 @@ N1000g = N1000.groupby('GC')
 X = 0
 for n, g in N1000g:
     print(g)
-    ax0.scatter(g['size'], g['time'],
-                label=r'{}'.format(g['GC'].unique()[0]))
+    ax0.plot(g['size'], g['time'], marker='.', ms=10,
+             label=r'{}'.format(g['GC'].unique()[0]))
     X += 1
 
 
@@ -30,20 +30,21 @@ N10000 = data[data['N'] == 10000]
 N10000g = N10000.groupby('GC')
 
 for n, g in N10000g:
-    ax1.scatter(g['size'], g['time'])
+    ax1.plot(g['size'], g['time'], marker='.', ms=10)
 
-ax0.legend(loc='upper left', frameon = False,title="GC interval\n(generations)",fontsize='x-small')
-for ax in [ax0,ax1]:
+ax0.legend(loc='upper left', frameon=False,
+           title="GC interval\n(generations)", fontsize='x-small')
+for ax in [ax0, ax1]:
     print(sorted(data['size'].unique()))
     ax.xaxis.set_ticks(sorted(data['size'].unique()))
     ax.set_xscale('log')
-    ax.set_xlim((500,100000))
+    ax.set_xlim((500, 100000))
     ax.set_xlabel(r'Region size ($\rho$ = 4Nr)')
     ax.set_ylabel("Run time (hours)")
 
 
-ax0.set_title("N = 1,000",fontsize='medium')
-ax1.set_title("N = 10,000",fontsize='medium')
+ax0.set_title("N = 1,000", fontsize='medium')
+ax1.set_title("N = 10,000", fontsize='medium')
 ax1.yaxis.set_label_position("right")
 plt.tight_layout()
 
@@ -58,7 +59,6 @@ ax1 = plt.subplot(gs[1])
 ax1.yaxis.tick_right()
 
 
-
 N1000 = data[data['N'] == 1000]
 
 N1000g = N1000.groupby('GC')
[…]

### 2018-01-03 forwards_paper.tex
add GC interval text to paper

@@
 	easy to read off the figures}
-Our method uses more RAM (up to 2 orders of magnitude more) than simple forward simulations (see~\ref{ss:memuse_nosel}),
+Our method uses more RAM (up to 2 orders of magnitude more) than simple forward simulations (see~\ref{ss:memuse}),
 although in practice shorter simplification intervals may reduce RAM use
-(but at a slight cost to the speed up achieved, see~\ref{ss:seq_simp}).
+(but at a slight cost to the speed up achieved, see~\ref{ss:gcinterval}).
 The benefits to our method are not limited to situations with selection:
 in a more limited set of neutral simulations we found the same qualitative
-behavior with a numerically larger benefit to pedigree tracking (see~\ref{ss:speedup_nosel}).
+behavior with a numerically larger benefit to pedigree tracking (see~\ref{ss:timing_nosel}).
 
@@
 
-\section{Timing with no selection}
+\newpage
+
+\section{Effect of simplification interval on run time and memory use}
+\label{ss:gcinterval}
+
+We performed a limited set of simulations to explore the effect of the simplification interval on run times and memory
+use. Figure~\ref{sfig:gctime} shows that run times increase for simplification intervals less than 100 generations.  For
+intervals $\geq 100$ generations, run times are very similar, with no appreciable difference between $10^3$ and $10^4$
+generations.  As expected, the simplification interval has a near-linear effect on peak RAM use
+(Figure~\ref{sfig:gcmem}).  Taken together Figures~\ref{sfig:gctime} and~\ref{sfig:gcmem} suggest that one can tune the
+simplification interval to available RAM with little effect on run times provided that simplification doesn't occur too
+often.
+
+\begin{figure}[!h]
+    \includegraphics{sims/gc_interval/GCtime}
+    \caption{\label{sfig:gctime}The effect of simplification interval on run times.  A single replicate was done for the
+        parameters shown in the figure.  These simulations were done with the same selection parameters as in
+    Figure~\ref{fig:runtimes_selection}.}
+\end{figure}
+
+\begin{figure}[!h]
+    \includegraphics{sims/gc_interval/GCmem}
+    \caption{\label{sfig:gcmem}Peak RAM use for the simulations shown in Figure~\ref{sfig:gctime}.}
+\end{figure}
+
+\newpage
+\section{Timing and memory use with no selection}
 \label{ss:timing_nosel}

### 2018-01-03 sims/gc_interval/plot.py
GC -> simplification in plot legend

@@ -33,7 +33,7 @@ for n, g in N10000g:
     ax1.plot(g['size'], g['time'], marker='.', ms=10)
 
 ax0.legend(loc='upper left', frameon=False,
-           title="GC interval\n(generations)", fontsize='x-small')
+           title="Simplification interval\n(generations)", fontsize='xx-small')
 for ax in [ax0, ax1]:
     print(sorted(data['size'].unique()))
     ax.xaxis.set_ticks(sorted(data['size'].unique()))
@@ -78,8 +78,8 @@ N10000g = N10000.groupby('GC')
 for n, g in N10000g:
     ax1.plot(g['size'], g['mem'], marker='.', ms = 10)
 
-ax0.legend(loc='upper left', frameon=False,
-           title="GC interval\n(generations)", fontsize='x-small')
+ax0.legend(loc="upper left", frameon=False,
+           title="Simplification\ninterval (generations)", fontsize='xx-small')
 for ax in [ax0, ax1]:
     print(sorted(data['size'].unique()))
     ax.xaxis.set_ticks(sorted(data['size'].unique()))

### 2018-01-03 forwards_paper.tex
add comment

@@
 	easy to read off the figures}
+\krt{I'm not sure it really matters.  It goes from a little to a lot.  And if we do relative, then we lose a lot of
+    points again}
 Our method uses more RAM (up to 2 orders of magnitude more) than simple forward simulations (see~\ref{ss:memuse}),

### 2018-01-04 forwards_paper.tex
Address some comments on the PR
added parens, closes #48

@@
 Roughly, the length of time for which a marginal tree has $k$ tips is
-$2N/k(k-1) = 2N(1/(k-1) - 1/k)$,
+$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$,
 and so the time to go from $N$ to $n$ tips is $2N(1/(n-1) - 1/(N-1))$.
@@
 \label{ss:gcinterval}
-
-We performed a limited set of simulations to explore the effect of the simplification interval on run times and memory
+\renewcommand{\thefigure}{C\arabic{figure}}
+
+We performed a limited set of simulations with \fwdpy{} to explore the effect of the simplification interval on run times and memory
 use. Figure~\ref{sfig:gctime} shows that run times increase for simplification intervals less than 100 generations.  For
@@
 \label{ss:timing_nosel}
-\renewcommand{\thefigure}{C\arabic{figure}}
+\renewcommand{\thefigure}{D\arabic{figure}}
 \setcounter{figure}{0}

### 2018-01-04 forwards_paper.tex
SLiM2 -> SLiM 2

@@
 Because of this computational burden, even the fastest simulation frameworks such as
-SLiM2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
+SLiM 2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
 can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals

### 2018-01-04 forwards_paper.tex
clarified horiz lines in fig 5

@@
         A simple example of the simplify method.
-        \textbf{Top:} the diagram on the left relates ten haploid individuals to each other.
-        It has 10 node records (one for each individual)
+        \textbf{Top:} the augmented pedigree diagram on the left 
+        relates ten haploid ``individuals'' (i.e., parental chromosomes) to each other;
+        dotted lines connect individuals who mate, and solid lines lead to their offspring.
+        The corresponding tables (right) have 10 node records (one for each individual)
         and 14 edge records (one for each distinctly inherited segment).
-        Blue numbers denote crossing over locations in each meiosis --
+        Blue numbers denote crossing over locations --
         for instance, $C$ and $D$ were parents to $F$,
         who inherited the left 70\% of the chromosome from $D$ and the remainder from $C$.
-        The individuals $B$, $C$, and $D$ inherit clonally from $A$.
-        \textbf{Center:} the five distinct trees relating all individuals to each other
+        Individuals $B$, $C$, and $D$ inherit clonally from $A$.
+        \textbf{Center:} the five distinct trees
         found across the chromosome (blue numbers denote locations on the chromosome).

### 2018-01-06 forwards_paper.tex
added points to the abstract, re: #47

@@
     We implement the method in two popular forwards-time simulation frameworks,
-    and show that not only does this speed up large simulations by one or two orders of magnitude,
-    but saves not only full genotype sequences, but also all marginal
-    geneologies of the simulated individuals.
-    % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
+    which speeds up large, whole-genome simulations by one or two orders of magnitude.
+    In addition to speed, we show that the method has several advantages:
+    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
+    (2) A population of $N$ individuals with $M$ polymorphic sites
+    can be stored on $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation
+    rather than only a subset.
+    (3) A simulation can easily be initialized with prior history produced by a more efficient coalescent simulation of deep history.
+    To make these possible, 
+    we introduce the \msprime{} Tables API, which allows for the efficient interchange of genealogical data between modular program components.
 \end{abstract}

### 2018-01-06 forwards_paper.tex
edited RAM section re: #44

@@
 partially obscures an important observation that the run times appear approximately linear in region size when tracking
-pedigree.  The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on
-average
+pedigree.  The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
 each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
-
-\jda{Maybe we should quantify the factor by which RAM use increased; it's not
-	easy to read off the figures}
-\krt{I'm not sure it really matters.  It goes from a little to a lot.  And if we do relative, then we lose a lot of
-    points again}
-Our method uses more RAM (up to 2 orders of magnitude more) than simple forward simulations (see~\ref{ss:memuse}),
-although in practice shorter simplification intervals may reduce RAM use
-(but at a slight cost to the speed up achieved, see~\ref{ss:gcinterval}).
-The benefits to our method are not limited to situations with selection:
-in a more limited set of neutral simulations we found the same qualitative
+In a more limited set of neutral simulations we found the same qualitative
 behavior with a numerically larger benefit to pedigree tracking (see~\ref{ss:timing_nosel}).
+
+In our test runs, our method used substantially more RAM than simple forward simulations (see~\ref{ss:memuse}).
+This is unsurprising:
+unsimplified tree sequences grow quickly, and so storing history can use arbitrarily much memory. 
+However, this is not a requirement of the method, only a straightforwards consequence of a speed--memory tradeoff:
+the amount of required memory is mostly determined by the interval between simplification steps,
+but less frequent simplification reduces overall computation time (see~\ref{ss:gcinterval}). 
+In fact, our method could in some situations \emph{reduce} the amount of memory required, 
+if memory usage in the forwards simulation was dominated by the cost of maintaining neutral genetic variants.
 

### 2018-01-06 forwards_paper.tex
rejiggered figures; as in #45 and #46

@@
     \caption{
-        A simple example of the simplify method.
+        An example of tree sequence simplification.
         \textbf{Top:} the augmented pedigree diagram on the left 
-        relates ten haploid ``individuals'' (i.e., parental chromosomes) to each other;
-        dotted lines connect individuals who mate, and solid lines lead to their offspring.
-        The corresponding tables (right) have 10 node records (one for each individual)
+        relates the ten homologous chromosomes of five diploid individuals (BC, DE, FG, HI, and JK) 
+        to each other and to a common ancestral chromosome (A);
+        dotted lines connect the two chromosomes of each individual, and solid lines lead to the products of their meioses.
+        The corresponding tables (right) have 10 node records (one for each chromosome)
         and 14 edge records (one for each distinctly inherited segment).
         Blue numbers denote crossing over locations --
-        for instance, $C$ and $D$ were parents to $F$,
-        who inherited the left 70\% of the chromosome from $D$ and the remainder from $C$.
-        Individuals $B$, $C$, and $D$ inherit clonally from $A$.
+        for instance, $D$ and $E$ were parents to $F$,
+        who inherited the left 70\% of the chromosome from $E$ and the remainder from $D$.
+        $B$, $C$, $D$, and $E$ inherit clonally from $A$.
         \textbf{Center:} the five distinct trees
@@
         in the example of Figure~\ref{fig:method_diagram}A.
-        As in that figure, blue numbers denote the crossing over location in a meiosis --
-        for instance, $G$ and $H$ were parents to $I$,
-        who inherited the left 70\% of the chromosome from $G$ and the remainder from $H$.
+        Ancestral material from J (red) and K (blue) are traced up through the pedigree
+        until they coalesce;
+        the smaller colored chromosomes on either side of each solid arrow show the bits inherited from each of the two parental chromosomes;
+        each time two colors overlap, a coalescence occurs, and two edges are output.
+        For instance, both J and K inherit from H beteen 0.5 and 0.9,
+        which resulted in the first two edges of the simplified table of Figure~\ref{fig:method_diagram}C.
         \label{fig:simplify_state}

### 2018-01-08 peq_abstract.tex
add peq abstract draft

\documentclass{article}
\usepackage{fullpage}
\begin{document}

We present a new approach to recording the entire history of a population during forward-time, individual-based
population genetics simulation.  Our approach uses extensions of the data structures in the \emph{mprime} software to
record the marginal genealogies of all individuals (including ancient samples when needed), eliminating the need to
expicitly simulate mutations not affecting fitness.  We show that recording histories substantially reduces the run-time
complexity of two different forward simulation engines, providing run time improvements of at least an order of
magnitude and the possibility of substantial reductions in memory requirements.  We will also describe more general
features of the new \emph{msprime} API that apply both to simulation methods and more generally to storage of
population-genomic information.

\end{document}

### 2018-01-08 forwards_paper.tex
labeling bug in #45

@@
         Blue numbers denote crossing over locations --
-        for instance, $D$ and $E$ were parents to $F$,
+        for instance, $D$ and $E$ were parents to $G$,
         who inherited the left 70\% of the chromosome from $E$ and the remainder from $D$.
@@
         \textbf{Bottom:} tables recording the tree sequence after simplification
-        with nodes $I$ and $J$ as samples.
+        with nodes $J$ and $K$ as samples.
         The mapping from labels in the forwards time simulation to nodes in the tree sequence

### 2018-01-09 peq_abstract.tex
fix typo

@@
 We present a new approach to recording the entire history of a population during forward-time, individual-based
-population genetics simulation.  Our approach uses extensions of the data structures in the \emph{mprime} software to
+population genetics simulation.  Our approach uses extensions of the data structures in the \emph{msprime} software to
 record the marginal genealogies of all individuals (including ancient samples when needed), eliminating the need to

### 2018-01-09 peq_abstract.tex
revise last sentence as per Jaime's suggestion

@@
 complexity of two different forward simulation engines, providing run time improvements of at least an order of
-magnitude and the possibility of substantial reductions in memory requirements.  We will also describe more general
-features of the new \emph{msprime} API that apply both to simulation methods and more generally to storage of
-population-genomic information.
+magnitude and the possibility of substantial reductions in memory requirements.  We will also describe the API, new in
+\emph{msprime} 0.5.0, that enables our simulation method.  These new features generally enable storage of
+population-genomic information within the \emph{msprime} data structures from third-party code.
 

### 2018-01-09 examples/fig5.py
Fig 5 simplified, no metadata

@@ -1,0 +1,6 @@
+# coding: utf-8
+import msprime
+print("Simplifed trees")
+ts = msprime.load_text(edges=open('edges.txt'), nodes=open('nodes.txt'))
+for t in ts.trees():
+    print(t.draw(format='unicode'))

### 2018-01-09 examples/fig5.py
Remove metadata from full_nodes, add full to script
output from python fig5.py included

@@ -1,6 +1,16 @@
 # coding: utf-8
 import msprime
-print("Simplifed trees")
-ts = msprime.load_text(edges=open('edges.txt'), nodes=open('nodes.txt'))
+
+ts = msprime.load_text(edges=open('edges_full.txt'),
+                       nodes=open('nodes_full.txt'))
+full_node_map = {0: 'A', 1: 'B', 2: 'C', 3: 'D', 4: 'E', 5: 'F', 6: 'G', 7: 'H',
+                 8: 'I', 9: 'J', 10: 'K'}
+
+print("Full trees")
 for t in ts.trees():
+    print(t.draw(format='unicode', node_label_text=full_node_map))
+
+print("Simplifed trees")
+tss = msprime.load_text(edges=open('edges.txt'), nodes=open('nodes.txt'))
+for t in tss.trees():
     print(t.draw(format='unicode'))

### 2018-01-09 examples/fig5.py
Add output from ts.simplify to compare

@@ -10,7 +10,21 @@ print("Full trees")
 for t in ts.trees():
     print(t.draw(format='unicode', node_label_text=full_node_map))
 
-print("Simplifed trees")
-tss = msprime.load_text(edges=open('edges.txt'), nodes=open('nodes.txt'))
+
+print("Simplifed trees with to J (9), K (10) with `ts.simplify()`)")
+tss = ts.simplify(samples=[9, 10])
 for t in tss.trees():
     print(t.draw(format='unicode'))
+
+
+print("Simplifed trees from tables in fig 5C",
+      "with J (0) and K(1) marked as samples.")
+tss2 = msprime.load_text(edges=open('edges.txt'), nodes=open('nodes.txt'))
+for t in tss2.trees():
+    print(t.draw(format='unicode'))
+
+print("note the nodes.flags differ between the output of `ts.simplify()`:")
+print(tss.dump_tables())
+print("\n\n\n...and the the trees loaded from the raw table:")
+print(tss2.dump_tables())
+

### 2018-01-09 examples/fig5.py
fixup! fixup! fixup! with is_sample only true for J,K

@@ -23,8 +23,7 @@ tss2 = msprime.load_text(edges=open('edges.txt'), nodes=open('nodes.txt'))
 for t in tss2.trees():
     print(t.draw(format='unicode'))
 
-print("note the nodes.flags differ between the output of `ts.simplify()`:")
+print("Raw tables from `ts.simplify()`:")
 print(tss.dump_tables())
-print("\n\n\n...and the the trees loaded from the raw table:")
+print("\n\n\n...and from the trees loaded from text tables:")
 print(tss2.dump_tables())
-

### 2018-01-09 forwards_paper.tex
Light editing pass + added some references.

@@
     during a forwards-time, individual-based population genetics simulation.
-    This allows us to dramatically reduce the computational burden of tracking individual genomes
+    This dramatically reduces the computational burden of tracking individual genomes
     by simulating only those loci that may affect reproduction (those having non-neutral variants),
@@
 and (c) small sample size relative to the population size.
-The first two assumptions can be side-stepped to a limited extent \citep{Neuhauser1997-nn, hudson1990gene},
+The first two assumptions can be side-stepped to a limited extent \citep{hudson1990gene, Neuhauser1997-nn},
 but it remains a challenge to map results of coalescent models
-onto species that are distributed across continuous geographical space
+onto species that are distributed across continuous geographical
+space~\citep{barton2010new,kelleher2014coalescent}
 and/or have large numbers of loci under various sorts of selection.
@@
 simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
-Several studies have now shown that in samples of size approaching that of the population,
+Several studies have now shown that in samples approaching the size of the population,
 genealogical properties may be distorted relative to the coalescent expectation
@@
 
-We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 Gigahertz Intel E5-2650 CPU with
+We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
 hyperthreading enabled.
@@
 The simulations above recorded information about each new individual
-in collection of tables that together encode a \emph{tree sequence}.
+in a collection of tables that together define a \emph{tree sequence}.
 A {tree sequence} is an encoding for a sequence of correlated trees,
@@
 and the identities $p$ and $c$ of the parent and child nodes
-of a single branch that occurs in all trees covering this interval.
+of a single branch that occurs in all trees in this interval.
 The spatial extent of the edges defining the topology of Figure~\ref{fig:example_tree_sequence}
@@
 It is this method of capturing the shared structure between adjacent trees that makes the
-tree sequence encoding very compact and algorithmically efficient.
+tree sequence encoding compact and algorithmically efficient.
 
@@
 in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
-2 with ancestral state `A' and the other at position 7 with ancestral state `G'. If
+2.5 with ancestral state `A' and the other at position 7.5 with ancestral state `G'. If
 no mutations occur, all samples inherit the ancestral state at a given site.
@@
 Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
-illustrated by red hexagons.
+illustrated by red stars.
 The first site, in the left-hand tree,
@@
 
-\algstep{W4.}{Record edges}{
+\algstep{W4.}{Record edges.}{
 Call $\taddedge{\Et, 0, x, P_a, u}$ and $\taddedge{\Et, x, r, P_b, u}$.
@@
 
-\algstep{W5.}{Individual loop}{Set $j \leftarrow j + 1$. If $j < N$ go to \algref{W2}.
+\algstep{W5.}{Individual loop.}{Set $j \leftarrow j + 1$. If $j < N$ go to \algref{W2}.
 Otherwise, if $t\bmod s \neq 0$ go to \algref{W7}.
@@
 This redundancy comes from two sources.
-First, there are a large number of nodes in tree that have only one child.
+First, there are a large number of nodes in the tree that have only one child.
 In Algorithm~\algref{W} we do not attempt to identify coalescence events,
@@
 and only some small subset will correspond to coalescences within the marginal
-trees. The second source of redundancy in the output of Algorithm~\algref{W}
+trees. The second source of redundancy in the (unsimplified) output of Algorithm~\algref{W}
 is due to the fact that lineages die out: a large number of
@@
 Any simulation scheme that records data into tables,
-as does Algorithm~\algref{W},
+as Algorithm~\algref{W} does,
 has its genealogical history available at any time as a tree sequence.
@@
 pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
-near-linear in $rho$ (Figure \ref{fig:runtimes_selection}).  In this section, we explore the run-time complexity using a
+near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).  In this section, we explore the run-time complexity using a
 simplified Wright-Fisher model with recombination.
@@
 data between programs or between different modules of the same program.
-% we are capitalizing Tables API but not the more generic "tables".
-% \jda{should capitalize Tables, or not, as above}
 Following current best practises
@@
 \jda{TODO citation missing}
-Our implementation uses the numpy C API \citep{numpy_C_api} to efficiently copy
+Our implementation uses the NumPy C API~\citep{walt2011numpy} to efficiently copy
 data from Python into the low-level C library used to manipulate tree sequences.
@@
 
-This also provides basic input and output operations via numpy's
+This also provides basic input and output operations via NumPy's
 array interface, which provides a great deal of flexibility and efficiency,
 and makes it straightforward to transfer data from sources
-such as HDF5 \citep{hdf5}, Dask, or Zarr.
-\jda{TODO citation missing}
-(For small scale data and debugging purposes, a simple text format is also supported.)
+such as HDF5 \citep{hdf5}or Dask~\citep{dask}.
+For small scale data and debugging purposes, a simple text format is also supported.
+
 % PLR: this was mentioned above, and belongs in the documentation, I think.

### 2018-01-10 forwards_paper.tex
Add detailed methods for simupop

@@
 \newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
+\newcommand{\tsort}[1]{\mbox{\texttt{sort}}(#1)}
+\newcommand{\tnodetable}[1]{\mbox{\texttt{NodeTable}}(#1)}
+\newcommand{\tedgetable}[1]{\mbox{\texttt{EdgeTable}}(#1)}
 \DeclareMathOperator{\len}{length}
@@
 \algstep{W1.}{Initialisation.}{
-Set $\Nt \leftarrow \mbox{\texttt{NodeTable}}()$, $\mathcal{E}
+Set $\Nt \leftarrow \tnodetable()$, $\mathcal{E}
 \leftarrow \mbox{\texttt{EdgeTable}}()$, $t \leftarrow T$, and $j \leftarrow 0$.
@@
 
-\krt{Are we doing a methods section?}
-
-\subsection{Requirements for the \cpp{} implementation}
+\subsection*{The \cpp{} implementation}
+
+\paragraph{Requirements.}
+The \cpp{} implementation uses several packages installed via conda,
+except for \msprime{}, which must be installed from GitHub.
 
@@
     \item GCC 4.8.5
-    \item \msprime{} master branch \jda{which commit hash or short version number
-		    from pip freeze or {\tt msprime.\_\_version\_\_}?}
+    \item \msprime{} 0.5.0 \jda{Changed these but the actual sims were not run
+		    with this version...}
 \end{enumerate}
 
-All of the above installed via conda, except for msprime, which must be installed from GitHub.
-
-\subsection{Requirements for the \simupop{} implementation}
-
-The simulations are implemented the Python package \ftprime{} 0.0.6rc0
-which is available at \url{https://www.github.com/ashander/ftprime}.
+\subsection*{The \simupop{} implementation}
+
+We implemented simulations in a Python script, usable as a command-line utility,
+taking input parameters for population size $N$ and scaled recombination $\rho$.
+As described in 'Simulation benchmarks', we set scaled mutation $\theta=\rho$,
+we used simplify interval of 1000 and simulated for $T=10N$ generations in each
+simulation, introducing deleterious mutations at a rate of $\rho/100$ per
+generation.
+Because \simupop{} requires explicit locations for loci, we set the number of
+loci $n=\rho / (4Nr)$ and marked
+$k={\rm round}(\mu / (100 (\mu + \mu / 100))) = 0$ of these as selected loci.
+\jda{unfortunately probably what we did...}
+The selected loci were chosen as a random sample of all the loci.
+
+We coupled \simupop{} to \msprime{} using a Python class \texttt{RecombCollector} 1)
+tracks time, and 2) collects and parses recombination information provided by
+the \texttt{Recombinator} class of \simupop{}, adding them to a \tnodetable{} and
+\tedgetable{} upon which simplification is performed every 1000 years (using
+\texttt{msprime.sort\_tables()} and \texttt{msprime.simplify\_tables}).
+The tables are initialized using the \texttt{TreeSequence} returned from
+\texttt{msprime.simulate($2N$, Ne=$N$, recombination\_rate=$r/2$, length=$n$)}.
+The \texttt{simuPOP.Population()} is initialized and tagged with individual IDs,
+a mapping of these IDs to the \tnodetable IDs is passed to
+an instance of \texttt{RecombCollector} and used internally to update the tables.
+The \texttt{simuPOP.Population()} is then evolved.
+Before mating: selected mutations are created (using
+\texttt{simuPOP.SNPMutator}) and multiplicative selection occurs (using
+\texttt{simuPOP.PyMlSelector}) only on the selected loci;
+the internal time of the \texttt{RecombCollector} instance is updated.
+During mating: male and female parents are drawn randomly, recombination data is
+output by \simupop{} to the \texttt{RecombCollector} instance.
+After mating: nothing happens 999/1000 years, every 1000 years the tables are
+simplifed.
+
+\texttt{RecombCollector} uses the Python class \texttt{ARGrecorder} internally
+to add data to the tables and perform simplification.
+This class contains the \tnodetable and \tedgetable and also the mapping from
+(haploid) ids to ids in the tables.
+It also provides methods (\texttt{add\_individual} and \texttt{add\_record})
+to append data to the tables.
+The only functions of
+\texttt{RecombCollector} not delegated to this internal class are computing
+haploid IDs from \simupop{}'s diploid IDs and parsing recomination data into
+data suitible for use with use with \texttt{add\_individual} and
+\texttt{add\_record}; these functions are used to stream data into the tables as
+it is received by an instance of \texttt{RecombCollector}.
+
+\paragraph{Requirements.}
+The simulations as implemented depend on the class \texttt{RecombCollector} provided by
+the Python package \ftprime{} 0.0.6rc0,
+available at \url{https://www.github.com/ashander/ftprime}.
 Its requirements are encoded in an \texttt{environment.yml} file, compatible
 with conda, but \msprime{} must be updated to the version listed below via pip
-from GitHub \jda{TODO add command}.
+from GitHub \jda{TODO add command?}.
+
 In brief the requirements are:
@@
 \begin{enumerate}
-    \item \msprime{} \jda{TODO add version from KRT's dev machine}
+    \item \msprime{} 0.5.0
     \item \simupop{} 1.1.8.3

### 2018-01-10 peq_abstract.tex
edits

@@
 
+\textbf{KT:}
+
 We present a new approach to recording the entire history of a population during forward-time, individual-based
@@
 record the marginal genealogies of all individuals (including ancient samples when needed), eliminating the need to
-expicitly simulate mutations not affecting fitness.  We show that recording histories substantially reduces the run-time
+expicitly simulate mutations not affecting fitness.  
+As another benefit, it is easy with these tools to initialize a forward-time simulation
+with prior history produced by efficient coalescent simulation.
+We show that recording histories substantially reduces the run-time
 complexity of two different forward simulation engines, providing run time improvements of at least an order of
@@
 \emph{msprime} 0.5.0, that enables our simulation method.  These new features generally enable storage of
-population-genomic information within the \emph{msprime} data structures from third-party code.
+population-genomic information within the \emph{msprime} data structures from third-party code. The method
+makes it possible to not only run substantially larger simulations (whole chromosomes with XX individuals in X hours),
+but also to record and analyze the resulting whole-population genealogies across entire genomes.
 

### 2018-01-10 forwards_paper.tex
Formatting fix

@@
 The \texttt{simuPOP.Population()} is initialized and tagged with individual IDs,
-a mapping of these IDs to the \tnodetable IDs is passed to
+a mapping of these IDs to the \tnodetable{} IDs is passed to
 an instance of \texttt{RecombCollector} and used internally to update the tables.
@@
 to add data to the tables and perform simplification.
-This class contains the \tnodetable and \tedgetable and also the mapping from
+This class contains the \tnodetable{} and \tedgetable{} and also the mapping from
 (haploid) ids to ids in the tables.

### 2018-01-10 forwards_paper.tex
Edits from jk

@@
 We implemented simulations in a Python script, usable as a command-line utility,
-taking input parameters for population size $N$ and scaled recombination $\rho$.
-As described in 'Simulation benchmarks', we set scaled mutation $\theta=\rho$,
+taking input parameters for population size $N$ and scaled recombination rate $\rho$.
+As described in 'Simulation benchmarks', we set scaled mutation rate $\theta=\rho$,
 we used simplify interval of 1000 and simulated for $T=10N$ generations in each
@@
 
-We coupled \simupop{} to \msprime{} using a Python class \texttt{RecombCollector} 1)
+We coupled \simupop{} to \msprime{} using a Python class \texttt{RecombCollector} that 1)
 tracks time, and 2) collects and parses recombination information provided by
 the \texttt{Recombinator} class of \simupop{}, adding them to a \tnodetable{} and
-\tedgetable{} upon which simplification is performed every 1000 years (using
+\tedgetable{} upon which simplification is performed every 1000 generations (using
 \texttt{msprime.sort\_tables()} and \texttt{msprime.simplify\_tables}).
@@
 output by \simupop{} to the \texttt{RecombCollector} instance.
-After mating: nothing happens 999/1000 years, every 1000 years the tables are
+After mating: nothing happens 999/1000 generations, every 1000 generations the tables are
 simplifed.

### 2018-01-10 sims/plot_RAM.py
remove simupop from performance figures

@@ -31,8 +31,8 @@ data_neut = data[data['sel'] == False].copy(deep=True)
 data = data[data['sel'] == True]
 data.sort_values(by='rho', inplace=True)
 groups = data.groupby(['engine', 'arg', 'queue', 'N'])
-fig, ((ax_fwdpp, ax_fwdpp_arg), (ax_simupop, ax_simupop_arg)) = plt.subplots(
-    2, 2, sharex=True, sharey=False)
+fig, (ax_fwdpp, ax_fwdpp_arg) = plt.subplots(
+    1, 2, sharex=True, sharey=False)
 for name, group in groups:
     lstyle = '-'
     if name[0] == 'fwdpy11' and name[1] is False:
@@ -41,6 +41,7 @@ for name, group in groups:
         ax = ax_fwdpp_arg
     m = points[name[3]]
     if name[0] == 'simuPOP':
+        continue
         ax = ax_simupop
         if name[1] is True:
             ax = ax_simupop_arg
@@ -57,14 +58,12 @@ for name, group in groups:
 ax_fwdpp.legend(loc='best',frameon=False)
 ax_fwdpp.set_title("fwdpy11 with neutral mutations", fontsize='medium')
 ax_fwdpp_arg.set_title("fwdpy11 with pedigree tracking", fontsize='medium')
-ax_simupop.set_ylabel("Peak RAM use (GB)")
 ax_fwdpp.set_ylabel("Peak RAM use (GB)")
-ax_simupop.set_title("simuPOP with neutral mutations", fontsize='medium')
-ax_simupop_arg.set_title("simuPOP with pedigree tracking", fontsize='medium')
-ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
-ax_simupop_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
-ax_simupop_arg.set_xticks([1e3, 1e4, 1e5])
-for ax in (ax_simupop, ax_simupop_arg):
+ax_fwdpp.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
+ax_fwdpp_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
+ax_fwdpp_arg.set_xticks([1e3, 1e4, 1e5])
+ax_fwdpp.set_xticks([1e3, 1e4, 1e5])
[…]

### 2018-01-10 sims/quick_plot.py

@@ -31,8 +31,10 @@ data_neut = data[data['sel'] == False].copy(deep=True)
 data = data[data['sel'] == True]
 data.sort_values(by='rho', inplace=True)
 groups = data.groupby(['engine', 'arg', 'queue', 'N'])
-fig, ((ax_fwdpp, ax_fwdpp_arg), (ax_simupop, ax_simupop_arg)) = plt.subplots(
-    2, 2, sharex=True, sharey=True)
+fig, (ax_fwdpp, ax_fwdpp_arg) = plt.subplots(
+    1, 2, sharex=True, sharey=True)
+# fig, ((ax_fwdpp, ax_fwdpp_arg), (ax_simupop, ax_simupop_arg)) = plt.subplots(
+#     2, 2, sharex=True, sharey=True)
 for name, group in groups:
     lstyle = '-'
     if name[0] == 'fwdpy11' and name[1] is False:
@@ -41,9 +43,10 @@ for name, group in groups:
         ax = ax_fwdpp_arg
     m = points[name[3]]
     if name[0] == 'simuPOP':
-        ax = ax_simupop
-        if name[1] is True:
-            ax = ax_simupop_arg
+        continue
+    #     ax = ax_simupop
+    #     if name[1] is True:
+    #         ax = ax_simupop_arg
     mfacecolor = colors[name[3]]
     if name[2] is True:
         lstyle = 'dashed'
@@ -57,15 +60,19 @@ for name, group in groups:
 ax_fwdpp.legend(loc='upper left',frameon=False)
 ax_fwdpp.set_title("fwdpy11 with neutral mutations", fontsize='medium')
 ax_fwdpp_arg.set_title("fwdpy11 with pedigree tracking", fontsize='medium')
-ax_simupop.set_ylabel("Run time (hours)")
+# ax_fwdpp_arg.set_ylabel("Run time (hours)")
 ax_fwdpp.set_ylabel("Run time (hours)")
-ax_simupop.set_title("simuPOP with neutral mutations", fontsize='medium')
-ax_simupop_arg.set_title("simuPOP with pedigree tracking", fontsize='medium')
-ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
-ax_simupop_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
-ax_simupop_arg.set_xticks([1e3, 1e4, 1e5])
[…]

### 2018-01-10 forwards_paper.tex
Refocussed the API overview section.

--- 2018-01-10 forwards_paper.tex
@@
 \newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
-\newcommand{\tsort}[1]{\mbox{\texttt{sort}}(#1)}
-\newcommand{\tnodetable}[1]{\mbox{\texttt{NodeTable}}(#1)}
-\newcommand{\tedgetable}[1]{\mbox{\texttt{EdgeTable}}(#1)}
 \DeclareMathOperator{\len}{length}
@@
 \algstep{W1.}{Initialisation.}{
-Set $\Nt \leftarrow \tnodetable()$, $\mathcal{E}
+Set $\Nt \leftarrow \mbox{\texttt{NodeTable}}()$, $\mathcal{E}
 \leftarrow \mbox{\texttt{EdgeTable}}()$, $t \leftarrow T$, and $j \leftarrow 0$.
@@
 %%%%%%
-\subsection*{Overview of the API}
-
-These tools are implemented through the \msprime{} Python API,
-which provides a powerful platform for working with tree topology and mutation data.
-The new portions of \msprime{} that we cover here
-are dedicated to tree sequence input and output using simple tables of data,
-so we refer to this as the `Tables API'.
-The four key tables: nodes, edges, sites and mutations, are described above.
+\subsection*{Implementation overview}
+
+The facilities for working with succinct tree sequences are implemented as part
+of the \msprime\ Python API, which provides a powerful platform for processing
+tree topology and mutation data. The new portions of \msprime{} that we discuss
+here are dedicated to tree sequence input and output using simple tables of
+data, so we refer to this as the `Tables API'. The four key tables (nodes,
+edges, sites and mutations) are described above.
 
 The Tables API is primarily designed to facilitate efficient interchange of
-data between programs or between different modules of the same program.
-Following current best practises
-\plr{citations: Apache arrow, etc},
-data is stored in a columnar format,
-which allows for very efficient interchange of large amounts of numerical data.
-In principle, this enables zero-copy semantics,
-where a data consumer can read the information directly
-from the memory of a producer without incurring the overhead of a copy.
-\jda{TODO citation missing}
+data between programs or between different modules of the same program. We
+adopted a `columnar' design, i.e., where all the values for a
+particular column are stored in adjacent memory locations.
+There are many advantages to columnar storage---for example, since adjacent
+values in memory are from the same column, they tend to compress well,
+and suitable encodings can be chosen on a per-column basis~\citep{abadi2006integrating}.
+A particular advantage of this approach is that it enables very
+efficient copying of data, and in principle zero-copy data access
+(where a data consumer reads directly from the memory of a producer) is possible.
 Our implementation uses the NumPy C API~\citep{walt2011numpy} to efficiently copy
 data from Python into the low-level C library used to manipulate tree sequences.
-Since tables are encoded as contiguous arrays of numeric data,
-this achieves data transfer rates that would be impossible under
-any text-based approach while retaining excellent portability.
-
-This also provides basic input and output operations via NumPy's
-array interface, which provides a great deal of flexibility and efficiency,
+This architecture allows for data transfer rates of gigabytes per second
+(impossible under any text-based approach), while retaining excellent portability.
+NumPy's array interface provides a great deal of flexibility and efficiency,
 and makes it straightforward to transfer data from sources
 such as HDF5 \citep{hdf5}or Dask~\citep{dask}.
-For small scale data and debugging purposes, a simple text format is also supported.
-
-% PLR: this was mentioned above, and belongs in the documentation, I think.
-% The API also provides functions to sort tables
-% to ensure that the records are in the form required to input
-% into an \msprime{} tree sequence object. (Simplification may also be required.)
-
-The \msprime{} Tables API is designed to be easily ``plugged'' into any simulation engine.
-One simply needs to fill the relevant columns,
-and to send them to \msprime{} at regular intervals for simplification.
-For example, in the benchmarking simulations above,
-we used the \fwdpp{} \cpp{} API \citep{fwdpp}
-to fill arrays of simple structures representing the node and edge data.
-Doing so requires no modification of the \fwdpp{} code base.
-Rather, we simply need to bookkeep parent/offspring labels,
-and perform simple processing of the recombination breakpoints from each mating event.
-Using \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}),
-the node/edge arrays on the \cpp{} side are made visible, without copy, to Python as a multi-column numpy array,
-which is processed by the \msprime{} tables API.
-Our implementation also uses \texttt{pybind11}
-to create a Python package extending \texttt{fwdpy11}
-(a Python package based on \fwdpp{}; \url{https://github.com/molpopgen/fwdpy11}),
-% (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
-% API from a command-line C/\cpp{} application via an embedded Python interpreter.)
+For small scale data and debugging purposes, a simple text based format is also supported.
+
+The \msprime\ Python Tables API provides a general purpose toolkit for importing
+and processing succinct tree sequences. Interoperation with Python simulators
+is then straightforward. For the proof-of-concept implementations discussed
+here, we took two different approaches to doing this. In the first method,
+based on \simupop{}, the simulation engine invokes callbacks at critical points
+of the simulation, and we infer nodes and edges from the information that is
+provided. Rows are appended to the tables one-by-one, and the tables are
+periodically sorted and simplified to control memory usage. The second method we developed uses
+\texttt{pybind11} (\url{https://github.com/pybind/pybind11/}) to inferface
+with the \fwdpp{} \cpp{} API \citep{fwdpp}. No modifications were
+required to the \fwdpp{} code base; rather, we simply need to bookkeep parent/offspring labels,
+and perform simple processing of the recombination breakpoints from each mating
+event to generate node and edge data. This information is then periodically copied
+to the \msprime\ Tables API, where it is sorted and simplified.
+
+While these methods work well, and indeed result in significant performance
+gains over non-ancestry based simulation, they are far from optimal. Because
+forwards time simulation is highly CPU intensive, most simulation engines
+are written in C or C++ out of necessity. Therefore, the simplest and most
+efficient approach would be to call the \msprime\ C library directly as part of
+the core simulation logic, in the manner outlined in Algorithm W. If designed
+in this way, the simulation engine would not need to keep copies of the node
+and edge information and would not need to infer edges;
+even greater gains in computational efficiency could then be obtained. To
+facilitate this usage, a standalone, embeddable C library encompassing the core
+tree sequence functionality is planned, which should make recording and
+outputting tree sequences from any C/C++ code straightforward.
+
+% is designed to be easily ``plugged'' into any simulation engine.
+% One simply needs to fill the relevant columns,
+% and to send them to \msprime{} at regular intervals for simplification.
+% For example, in the benchmarking simulations above,
+% we used the \fwdpp{} \cpp{} API \citep{fwdpp}
+% to fill arrays of simple structures representing the node and edge data.
+% Doing so requires no modification of the \fwdpp{} code base.
+% Rather, we simply need to bookkeep parent/offspring labels,
+% and perform simple processing of the recombination breakpoints from each mating event.
+% Using \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}),
+% the node/edge arrays on the \cpp{} side are made visible, without copy, to Python as a multi-column numpy array,
+% which is processed by the \msprime{} tables API.
+% Our implementation also uses \texttt{pybind11}
+% to create a Python package extending \texttt{fwdpy11}
+% (a Python package based on \fwdpp{}; \url{https://github.com/molpopgen/fwdpy11}),
+% % (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
+% % API from a command-line C/\cpp{} application via an embedded Python interpreter.)
 
@@
 
-\subsection*{The \cpp{} implementation}
-
-\paragraph{Requirements.}
-The \cpp{} implementation uses several packages installed via conda,
-except for \msprime{}, which must be installed from GitHub.
+\krt{Are we doing a methods section?}
+
+\subsection{Requirements for the \cpp{} implementation}
 
@@
     \item GCC 4.8.5
-    \item \msprime{} 0.5.0 \jda{Changed these but the actual sims were not run
-		    with this version...}
+    \item \msprime{} master branch \jda{which commit hash or short version number
+		    from pip freeze or {\tt msprime.\_\_version\_\_}?}
 \end{enumerate}
 
-\subsection*{The \simupop{} implementation}
-
-We implemented simulations in a Python script, usable as a command-line utility,
-taking input parameters for population size $N$ and scaled recombination $\rho$.
-As described in 'Simulation benchmarks', we set scaled mutation $\theta=\rho$,
-we used simplify interval of 1000 and simulated for $T=10N$ generations in each
-simulation, introducing deleterious mutations at a rate of $\rho/100$ per
-generation.
-Because \simupop{} requires explicit locations for loci, we set the number of
-loci $n=\rho / (4Nr)$ and marked
-$k={\rm round}(\mu / (100 (\mu + \mu / 100))) = 0$ of these as selected loci.
-\jda{unfortunately probably what we did...}
-The selected loci were chosen as a random sample of all the loci.
-
-We coupled \simupop{} to \msprime{} using a Python class \texttt{RecombCollector} 1)
-tracks time, and 2) collects and parses recombination information provided by
-the \texttt{Recombinator} class of \simupop{}, adding them to a \tnodetable{} and
-\tedgetable{} upon which simplification is performed every 1000 years (using
-\texttt{msprime.sort\_tables()} and \texttt{msprime.simplify\_tables}).
-The tables are initialized using the \texttt{TreeSequence} returned from
-\texttt{msprime.simulate($2N$, Ne=$N$, recombination\_rate=$r/2$, length=$n$)}.
-The \texttt{simuPOP.Population()} is initialized and tagged with individual IDs,
-a mapping of these IDs to the \tnodetable IDs is passed to
-an instance of \texttt{RecombCollector} and used internally to update the tables.
-The \texttt{simuPOP.Population()} is then evolved.
-Before mating: selected mutations are created (using
-\texttt{simuPOP.SNPMutator}) and multiplicative selection occurs (using
-\texttt{simuPOP.PyMlSelector}) only on the selected loci;
-the internal time of the \texttt{RecombCollector} instance is updated.
-During mating: male and female parents are drawn randomly, recombination data is
-output by \simupop{} to the \texttt{RecombCollector} instance.
-After mating: nothing happens 999/1000 years, every 1000 years the tables are
-simplifed.
-
-\texttt{RecombCollector} uses the Python class \texttt{ARGrecorder} internally
-to add data to the tables and perform simplification.
-This class contains the \tnodetable and \tedgetable and also the mapping from
-(haploid) ids to ids in the tables.
-It also provides methods (\texttt{add\_individual} and \texttt{add\_record})
-to append data to the tables.
-The only functions of
-\texttt{RecombCollector} not delegated to this internal class are computing
-haploid IDs from \simupop{}'s diploid IDs and parsing recomination data into
-data suitible for use with use with \texttt{add\_individual} and
-\texttt{add\_record}; these functions are used to stream data into the tables as
-it is received by an instance of \texttt{RecombCollector}.
-
-\paragraph{Requirements.}
-The simulations as implemented depend on the class \texttt{RecombCollector} provided by
-the Python package \ftprime{} 0.0.6rc0,
-available at \url{https://www.github.com/ashander/ftprime}.
+All of the above installed via conda, except for msprime, which must be installed from GitHub.
+
+\subsection{Requirements for the \simupop{} implementation}
+
+The simulations are implemented the Python package \ftprime{} 0.0.6rc0
+which is available at \url{https://www.github.com/ashander/ftprime}.
 Its requirements are encoded in an \texttt{environment.yml} file, compatible
 with conda, but \msprime{} must be updated to the version listed below via pip
-from GitHub \jda{TODO add command?}.
-
+from GitHub \jda{TODO add command}.
 In brief the requirements are:
@@
 \begin{enumerate}
-    \item \msprime{} 0.5.0
+    \item \msprime{} \jda{TODO add version from KRT's dev machine}
     \item \simupop{} 1.1.8.3

### 2018-01-10 sims/quick_plot.py
remove simupop from fig w/selection

@@ -31,8 +31,8 @@ data_neut = data[data['sel'] == False].copy(deep=True)
 data = data[data['sel'] == True]
 data.sort_values(by='rho', inplace=True)
 groups = data.groupby(['engine', 'arg', 'queue', 'N'])
-fig, ((ax_fwdpp, ax_fwdpp_arg), (ax_simupop, ax_simupop_arg)) = plt.subplots(
-    2, 2, sharex=True, sharey=True)
+fig, (ax_fwdpp, ax_fwdpp_arg) = plt.subplots(
+    1, 2, sharex=True, sharey=True)
 for name, group in groups:
     lstyle = '-'
     if name[0] == 'fwdpy11' and name[1] is False:
@@ -41,6 +41,7 @@ for name, group in groups:
         ax = ax_fwdpp_arg
     m = points[name[3]]
     if name[0] == 'simuPOP':
+        continue
         ax = ax_simupop
         if name[1] is True:
             ax = ax_simupop_arg
@@ -55,16 +56,13 @@ for name, group in groups:
             markerfacecolor=mfacecolor)
 
 ax_fwdpp.legend(loc='upper left',frameon=False)
-ax_fwdpp.set_title("fwdpy11 with neutral mutations", fontsize='medium')
-ax_fwdpp_arg.set_title("fwdpy11 with pedigree tracking", fontsize='medium')
-ax_simupop.set_ylabel("Run time (hours)")
+ax_fwdpp.set_title("With neutral mutations", fontsize='medium')
+ax_fwdpp_arg.set_title("With pedigree tracking", fontsize='medium')
 ax_fwdpp.set_ylabel("Run time (hours)")
-ax_simupop.set_title("simuPOP with neutral mutations", fontsize='medium')
-ax_simupop_arg.set_title("simuPOP with pedigree tracking", fontsize='medium')
-ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
-ax_simupop_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
-ax_simupop_arg.set_xticks([1e3, 1e4, 1e5])
-for ax in (ax_simupop, ax_simupop_arg):
+ax_fwdpp.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
[…]

### 2018-01-10 sims/quick_plot.py
remove simupop from speedup fig w/selection

@@ -76,14 +76,15 @@ joined = data_arg.merge(data_noarg, on=[
                         'engine', 'N', 'rho'], how='outer', suffixes=('_arg', '_noarg')).dropna()
 joined['speedup'] = joined['time_noarg'] / joined['time_arg']
 groups = joined.groupby(['engine', 'N'])
-fig, (ax_fwdpp, ax_simupop) = plt.subplots(
-    2,  sharex=True, sharey=False)
+fig, (ax_fwdpp) = plt.subplots(
+    1,  sharex=True, sharey=False)
 for name, group in groups:
     lstyle = 'solid'
     if name[0] == 'fwdpy11':
         ax = ax_fwdpp
     m = points[name[1]]
     if name[0] == 'simuPOP':
+        continue
         ax = ax_simupop
     mfacecolor = colors[name[1]]
     popsize = int(name[1])
@@ -92,12 +93,10 @@ for name, group in groups:
             color=colors[name[1]],
             markerfacecolor=mfacecolor)
 
-for ax in (ax_fwdpp, ax_simupop):
+for ax in (ax_fwdpp,):
     ax.set_ylabel("Speedup due to\npedigree recording")
-ax_fwdpp.set_title("fwdpy11", fontsize='medium')
-ax_simupop.set_title("simuPOP", fontsize='medium')
-ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
-ax_simupop.set_xscale('log')
+ax_fwdpp.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
+ax_fwdpp.set_xscale('log')
 ax_fwdpp.legend(loc='best',frameon=False)
 fig.tight_layout()
 plt.savefig("speedup.pdf")

### 2018-01-10 sims/plot_RAM.py
remove simupop from memuse plots w/selection

@@ -31,8 +31,8 @@ data_neut = data[data['sel'] == False].copy(deep=True)
 data = data[data['sel'] == True]
 data.sort_values(by='rho', inplace=True)
 groups = data.groupby(['engine', 'arg', 'queue', 'N'])
-fig, ((ax_fwdpp, ax_fwdpp_arg), (ax_simupop, ax_simupop_arg)) = plt.subplots(
-    2, 2, sharex=True, sharey=False)
+fig, (ax_fwdpp, ax_fwdpp_arg) = plt.subplots(
+    1, 2, sharex=True, sharey=False)
 for name, group in groups:
     lstyle = '-'
     if name[0] == 'fwdpy11' and name[1] is False:
@@ -41,6 +41,7 @@ for name, group in groups:
         ax = ax_fwdpp_arg
     m = points[name[3]]
     if name[0] == 'simuPOP':
+        continue
         ax = ax_simupop
         if name[1] is True:
             ax = ax_simupop_arg
@@ -55,16 +56,13 @@ for name, group in groups:
             markerfacecolor=mfacecolor)
 
 ax_fwdpp.legend(loc='best',frameon=False)
-ax_fwdpp.set_title("fwdpy11 with neutral mutations", fontsize='medium')
-ax_fwdpp_arg.set_title("fwdpy11 with pedigree tracking", fontsize='medium')
-ax_simupop.set_ylabel("Peak RAM use (GB)")
+ax_fwdpp.set_title("With neutral mutations", fontsize='medium')
+ax_fwdpp_arg.set_title("With pedigree tracking", fontsize='medium')
 ax_fwdpp.set_ylabel("Peak RAM use (GB)")
-ax_simupop.set_title("simuPOP with neutral mutations", fontsize='medium')
-ax_simupop_arg.set_title("simuPOP with pedigree tracking", fontsize='medium')
-ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
-ax_simupop_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
-ax_simupop_arg.set_xticks([1e3, 1e4, 1e5])
-for ax in (ax_simupop, ax_simupop_arg):
+ax_fwdpp.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
[…]

### 2018-01-11 forwards_paper.tex

@@
     \item GCC 4.8.5
-    \item \msprime{} master branch \jda{which commit hash or short version number
-		    from pip freeze or {\tt msprime.\_\_version\_\_}?}
+    \item \msprime{} 0.5.0 \jda{Changed these but the actual sims were not run
+		    with this version...}
 \end{enumerate}
 
-All of the above installed via conda, except for msprime, which must be installed from GitHub.
-
-\subsection{Requirements for the \simupop{} implementation}
-
-The simulations are implemented the Python package \ftprime{} 0.0.6rc0
-which is available at \url{https://www.github.com/ashander/ftprime}.
+\subsection*{The \simupop{} implementation}
+
+We implemented simulations in a Python script, usable as a command-line utility,
+taking input parameters for population size $N$ and scaled recombination rate $\rho$.
+As described in 'Simulation benchmarks', we set scaled mutation rate $\theta=\rho$,
+we used simplify interval of 1000 and simulated for $T=10N$ generations in each
+simulation, introducing deleterious mutations at a rate of $\rho/100$ per
+generation.
+Because \simupop{} requires explicit locations for loci, we set the number of
+loci $n=\rho / (4Nr)$ and marked
+$k={\rm round}(\mu / (100 (\mu + \mu / 100))) = 0$ of these as selected loci.
+\jda{unfortunately probably what we did...}
+The selected loci were chosen as a random sample of all the loci.
+
+We coupled \simupop{} to \msprime{} using a Python class \texttt{RecombCollector} that 1)
+tracks time, and 2) collects and parses recombination information provided by
+the \texttt{Recombinator} class of \simupop{}, adding them to a \tnodetable{} and
+\tedgetable{} upon which simplification is performed every 1000 generations (using
+\texttt{msprime.sort\_tables()} and \texttt{msprime.simplify\_tables}).
+The tables are initialized using the \texttt{TreeSequence} returned from
+\texttt{msprime.simulate($2N$, Ne=$N$, recombination\_rate=$r/2$, length=$n$)}.
+The \texttt{simuPOP.Population()} is initialized and tagged with individual IDs,
+a mapping of these IDs to the \tnodetable{} IDs is passed to
+an instance of \texttt{RecombCollector} and used internally to update the tables.
+The \texttt{simuPOP.Population()} is then evolved.
+Before mating: selected mutations are created (using
+\texttt{simuPOP.SNPMutator}) and multiplicative selection occurs (using
+\texttt{simuPOP.PyMlSelector}) only on the selected loci;
+the internal time of the \texttt{RecombCollector} instance is updated.
+During mating: male and female parents are drawn randomly, recombination data is
+output by \simupop{} to the \texttt{RecombCollector} instance.
+After mating: nothing happens 999/1000 generations, every 1000 generations the tables are
+simplifed.
+
+\texttt{RecombCollector} uses the Python class \texttt{ARGrecorder} internally
+to add data to the tables and perform simplification.
+This class contains the \tnodetable{} and \tedgetable{} and also the mapping from
+(haploid) ids to ids in the tables.
+It also provides methods (\texttt{add\_individual} and \texttt{add\_record})
+to append data to the tables.
+The only functions of
+\texttt{RecombCollector} not delegated to this internal class are computing
+haploid IDs from \simupop{}'s diploid IDs and parsing recomination data into
+data suitible for use with use with \texttt{add\_individual} and
+\texttt{add\_record}; these functions are used to stream data into the tables as
+it is received by an instance of \texttt{RecombCollector}.
+
+\paragraph{Requirements.}
+The simulations as implemented depend on the class \texttt{RecombCollector} provided by
+the Python package \ftprime{} 0.0.6rc0,
+available at \url{https://www.github.com/ashander/ftprime}.
 Its requirements are encoded in an \texttt{environment.yml} file, compatible
 with conda, but \msprime{} must be updated to the version listed below via pip
-from GitHub \jda{TODO add command}.
+from GitHub \jda{TODO add command?}.
+
 In brief the requirements are:

### 2018-01-11 forwards_paper.tex
First pass moving simupop

@@
 We begin by benchmarking the performance improvement achieved by connecting
-two previously published forwards-time simulation libraries,
-\simupop{} \citep{peng2005simupop} and \fwdpp{} \citep{fwdpp}
+a previously published forwards-time simulation library,
+\fwdpp{} \citep{fwdpp},
 to the tools we describe here, as implemented in \msprime{}.
@@
 
-To demonstrate the method's flexibility and to test it in different contexts,
-we implemented these simulations separately using two different simulation engines:
-
-\jda{Note/TODO: the varying references to c++, python, fwdpp, fwdpy11 and
-	simupop are probably confusing to a new reader. This is in part because
-	we refer to the python/simupop method in various places as one or the
-	other; the same goes for c++/fwdpy11/fwdpp. We should decide what to
-	call these and make our usage consistent}
-
-\begin{enumerate}
-
-    \item
-        In \cpp{}, using \fwdpp{} library functions and interface code (details below),
-        using a continuum-sites model for both mutation and recombination. Simulations were run using \fwdpy{}, a Python
-        package based on \fwdpp.
+We implemented our method in \cpp{}, using \fwdpp{} library functions and
+interface code (details below), using a continuum-sites model for both mutation
+and recombination. Simulations were run using \fwdpy{}, a Python package based
+on \fwdpp.
 % $\mu$ and $r$ were the means of Poisson processes, and mutations used the infinite-sites model \citep{Kimura1969-uc}.
 % We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.
-
-    \item
-        In Python, using the \simupop{} library interfacing with Python code
-	(via our package \ftprime{}),
-        using a discrete but large number of sites.
-        % $\mu$ and $r$ define per-base-pair mutation rate and we
-        % defined the length of the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.
-        % We varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.
-
-\end{enumerate}
-
 We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
@@
 neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for
-these simulations).  With \fwdpy{}, simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
-With \simupop{}, we were only able to run simulations with $N=10^3$ and region sizes up to $10^4$.  Pedigree tracking
-dramatically improves run times for both \fwdpy{} and \simupop{} (right column of Figure~\ref{fig:runtimes_selection}).
-For parameters that ran to completion when tracking neutral mutations and when tracking
+these simulations).  Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
+Pedigree tracking
+dramatically improves run times (right column of Figure~\ref{fig:runtimes_selection}).  For parameters that ran to completion when tracking neutral mutations and when tracking
 the pedigree, the relative speedup due to pedigree tracking is up to $\approx 50$ fold
-(Figure~\ref{fig:relative_speedup_selection}).  For \fwdpy{}, pedigree tracking reduces runtimes more for larger
+(Figure~\ref{fig:relative_speedup_selection}).  Pedigree tracking reduces runtimes more for larger
 population sizes and we observe increasing relative speedups as $4Nr$ increases for a given $N$
-(Figure~\ref{fig:relative_speedup_selection}).  We saw a similar trend with $4Nr$ for \simupop{}, but the effect of $N$
-could not be explored because the simulations tended to run out of available memory.
+(Figure~\ref{fig:relative_speedup_selection}).
 Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
@@
 In a more limited set of neutral simulations we found the same qualitative
-behavior with a numerically larger benefit to pedigree tracking (see~\ref{ss:timing_nosel}).
-
-In our test runs, our method used substantially more RAM than simple forward simulations (see~\ref{ss:memuse}).
+behavior with a numerically larger benefit to pedigree tracking (see Appendix~\ref{ss:timing_nosel}).
+
+In our test runs, our method used substantially more RAM than simple forward simulations (see Appendix~\ref{ss:memuse}).
 This is unsurprising:
@@
 the amount of required memory is mostly determined by the interval between simplification steps,
-but less frequent simplification reduces overall computation time (see~\ref{ss:gcinterval}).
+but less frequent simplification reduces overall computation time (see Appendix~\ref{ss:gcinterval}).
 In fact, our method could in some situations \emph{reduce} the amount of memory required,
@@
         different diploid population sizes ($N$). The left column shows run times for
-        simulations including the tracking of neutral mutations.  Missing data points are due to either simulations
-    being killed at 72 hours or running out of memory (which only happened with \simupop{}).  The right column of
+        simulations including the tracking of neutral mutations. The right column of
 panels shows run times tracking the pedigree instead of the neutral mutations. The dashed line in the upper right panel shows
@@
 
-% Comparison of simulation with/without msprime, using \simupop{}
-% or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
-%
 % Maybe an estimate of how long \emph{just} the pedigree recording and simplification takes,
-% so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
-% in a day.
+% so that then we can say how fast the simulator would have to be to do $10^6$
+% whole chromosomes for $10^7$ generations in a day.
 
@@
 %%%%%%
-\subsection*{Implementation overview}
+\subsection*{Overview of \msprime\ APIs that enable the method}
 
@@
 and processing succinct tree sequences. Interoperation with Python simulators
-is then straightforward. For the proof-of-concept implementations discussed
-here, we took two different approaches to doing this. In the first method,
-based on \simupop{}, the simulation engine invokes callbacks at critical points
-of the simulation, and we infer nodes and edges from the information that is
-provided. Rows are appended to the tables one-by-one, and the tables are
-periodically sorted and simplified to control memory usage. The second method we developed uses
+is then straightforward.  The implementation we benchmark here uses
 \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}) to inferface
@@
 
+\paragraph{Flexibility.}
+To demonstrate the flexibility provided by the Tables API and provide an
+implementation that decouples forward simulation internals from transfer of data
+to \msprime\, we implemented a version of the simulations described in
+``Simulation benchmarks'' separately in Python (see Appendix~\ref{ss:simupop}).
+For this proof-of-concept implementation, the simulation engine invokes
+callbacks at critical points of the simulation, and we infer nodes and edges
+from the information that is provided. Rows are appended to the tables
+one-by-one, and the tables are periodically sorted and simplified to control
+memory usage.
+It uses the \simupop{} library, interfacing with Python code (via our
+package \ftprime{}) to simulate a discrete but large number of sites instead of the
+infinite sites model; we provide results from this implementation alongside the
+results from the \cpp implementation for simulations without selection  (see
+Appendix~\ref{ss:timing_nosel}).
+% $\mu$ and $r$ define per-base-pair mutation rate a
+% the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.  We
+% varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.
+
+\paragraph{Next steps.}
 While these methods work well, and indeed result in significant performance
@@
 
-\subsection*{The \cpp{} implementation}
-
 \paragraph{Requirements.}
@@
 
-\subsection*{The \simupop{} implementation}
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgements}
+Thanks to Brad Shaffer and Evan McCartney--Melstad for useful discussions.
+Funding for this project came from XXX
+
+\bibliography{references}
+
+\appendix
+
+%%%%%%%%%%%%%%
+\section{Simplification algorithm}
+\label{ss:simplify_algorithm}
+
+\plr{Replace this with Jerome's interval-tree implementation?}
+
+Here we describe the simplification algorithm.
+The flow is close to our implementation,
+but the description below uses set operations,
+while in the implementation,
+all ancestors are maintained as linked lists of segments
+ordered by left endpoint.
+As it goes, the algorithm
+records a list $M$ that gives the mapping
+from input nodes to output nodes,
+so that if $M[p] = u$ then the $p^\text{th}$ node in the input
+is the same as the $u^\text{th}$ node in the output.
+The current state of the algorithm is recorded in $\Al$,
+a list of ancestors
+each of which is a collection of ancestral segments.
+An ancestral segment is a triple $(\ell, r, x)$,
+indicating that the output node $x$ inherits from this ancestor
+on the genomic segment $[\ell, r)$.
+
+\begin{taocpalg}{S}{Simplify a tree sequence}
+{Input consists of
+    a list $S$ of sample IDs,
+    a list $\Nt_I$ of nodes (ordered by birth time),
+    a list $\Et_I$ of edges,
+    and the genome length, $L$.
+    The output is a
+    nodes $\Nt_O$,
+    and list of edges $\Et_O$.
+}
+
+\algstep{S0.}{Initialize.}{
+    Set $\Al[j] \leftarrow (0, L, j)$
+    and $\Nt_O[j] \leftarrow \Nt_I[S[j]]$ for $0 \le j < \len(S)$.
+    Set the current input parent index to $p=0$
+    and $M[u] \leftarrow -1$ for $0 \le u < \len(\Nt_I)$.
+}
+\krt{We need a verbal definition of $Q$}
+\algstep{S1.}{Input parent loop.}{Set $Q \leftarrow \emptyset$.
+}
+
+\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} on each edge $(\ell, r, p, c)$ in $\Et_I$
+    corresponding to parent $p$ (which stores the resulting segments in the merge queue $Q$).
+}
+
+\algstep{S3.}{Merge ancestors.}{Call Algorithm \algref{M} on $Q$
+    (which adds to $\Nt_O$ and $\Et_O$).
+}
+
+\algstep{S4.}{Next parent.}{Set $p \leftarrow p+1$;
+    if $p = \len(\Nt_I)$, terminate and return $\Nt_O$ and $\Et_O$;
+    otherwise, return to \algref{S1}.
+}
+
+\end{taocpalg}
+
+The algorithm steps back up through the pedigree,
+since nodes are ordered by time-since-birth.
+First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
+so, for every edge $(\ell, r, p, c)$,
+Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
+with $[u,v) \setdiff [\ell, r)$
+(deleting the segment or splitting in two if necessary),
+and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:
+
+\begin{taocpalg}{R}{Remove ancestry}
+    {Given an edge $(\ell, r, p, c)$,
+    and a list $\Al[c]$ of the $m$ ancestral segments corresponding to $c$,
+    remove segments overlapping $[\ell, r)$ from $\Al[c]$,
+    and add those overlapping segments to $Q$.
+    }
+
+    \algstep{R0.}{Initialize.}{
+        Set $j \leftarrow 0$.
+    }
+
+    \algstep{R1.}{Segment loop.}{
+    If $j=m$, terminate; else let $(u, v, x) = \Al[c][j]$.
+    (Here $x$ is the output node that inherits from this segment.)
+    }
+
+    \algstep{R2.}{Add to merge queue.}{If $[u,v) \cap [\ell,r) = \emptyset$,
+    skip to \algref{R4}.
+    Otherwise, append $(\max(u,\ell), \min(v,r), x)$ to $Q$.
+    }
+
+    \algstep{R3.}{Remove ancestry.}{
+    Delete $\Al[c][j]$ and insert in its place
+    $(\min(u,\ell), \max(v,\ell), x)$ (if $u<\ell$)
+    and(/or) $(\min(v,r), \max(v,r), x)$ (if $r<v$).
+    }
+
+    \algstep{R4.}{Iterate.}{Set $j \leftarrow j+1$ and return to \algref{R1}.
+    }
+
+\end{taocpalg}
+
+Next we must merge the ancestry segments in $Q$,
+creating the new ancestor corresponding to input node $p$,
+adding a node and edges to the output if any segments overlap,
+which implies that an output coalescence has occurred.
+Below we index elements of an ancestral segment $z = (\ell, r, u)$
+by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
+and $z.\lid = u$ for the output node ID.
+
+\begin{taocpalg}{M}{Merge ancestors}
+    {Merge the ancestral segments in $Q$,
+    and create the new ancestor $\Al[p]$.
+    This algorithm also appends to $\Nt_O$ and $\Et_O$,
+    and needs to know about $S$,
+    to check if input nodes are samples.
+    }
+
+    \algstep{M0.}{Initialize.}{
+        Set $z \leftarrow (0, 0, -1)$ and $d = \text{False}$.
+    }
+
+    \algstep{M1.}{Segment loop.}{If $\len(Q) = 0$, go to \algref{M9}.}
+
+    \algstep{M2.}{Find next segment overlap boundaries}{
+        Let $\ell_0 \leftarrow \min\{x.\lleft : x \in Q\}$
+        and $H \leftarrow \{x \in Q : x.\lleft = \ell_0\}$.
+        Also let $\ell_+ \leftarrow \min\{x.\lleft : x \in Q \setdiff H\}$
+        and $r = \min(\ell_+, \min\{x.\lright : x \in H\})$.
+    }
+
+    \algstep{M3.}{Copy single segment.}{
+        If $\len(H) = 1$,
+        set $\alpha \leftarrow (\ell, r, H[0].\lid)$.
+        If also $p \in S$, then
+        set $\alpha.\lid \leftarrow M[p]$ and
+        $\add(\Et_O, (\alpha.\lleft, \alpha.\lright, \Nt[p], \alpha.\lid))$.
+        Go to \algref{M7}.
+    }
+
+    \algstep{M4.}{New output node.}{
+        If $M[p] = -1$ set $M[p] \leftarrow \add(\Nt_O, \Nt_I[p])$
+        Set $\alpha \leftarrow (\ell, r, M[p])$.
+    }
+
+    \algstep{M6.}{Record edges and update $Q$.}{
+        For each $x \in H$, do $\add(\Et_O, (\ell, r, M[p], x.\lid))$.
+    }
+
+    \algstep{M7.}{Update $Q$.}{
+        Remove from $Q$ any segment $x \in H$ with $x.\lright = r$
+        and then set $x.\lleft = \ell$ for all remaining $x \in H$.
+    }
+
+    \algstep{M8.}{Add segment to ancestor.}{
+        If $z.\lright = \alpha.\lleft$ and $z.\lid = \alpha.\lid$ set $d = \text{True}$.
+        Append $\alpha$ to $\Al[p]$, set $z \leftarrow \alpha$, and return to \algref{M1}.
+    }
+
+    \algstep{M9.}{Defragment.}{
+        If $d$ is True, find and merge any adjacent segments in $\Al[p]$
+        with the same $\lid$.  Terminate.
+    }
+
+\end{taocpalg}
+\krt{I think the algo discussion is  clear.  I've not tried to implement it, which would be the real test, but I find the description fine other than needing to define $Q$ when first used.}
+
+%% \newpage
+%% \clearpage
+%% \section*{Supplementary Information}
+
+\section{Memory usage with selection}
+\label{ss:memuse}
+\renewcommand{\thefigure}{B\arabic{figure}}
+\setcounter{figure}{0}
+
+The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel}.  For our parameters, which involved simplifying every
+$10^3$ generations,  pedigree tracking requires more RAM.  This is because with pedigree tracking, a non-recombinant
+individual still requires one new edge and two new nodes to be allocated.  Without pedigree tracking, no new memory is
+allocated for such an offspring.  Thus, pedigree tracking accumulates a large amount of data in RAM until
+simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extra RAM needed
+can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
+require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four sets of unsimplified data are
+stored in RAM.  In practice, the RAM consumption can be minimized by simplifying more often.
+
+%% \krt{I've avoided getting into how this much RAM means parallelization is required to make more progress.}
+
+\begin{figure}
+    \includegraphics[]{sims/memuse}
+    \caption{\label{sfig:ramsel}Peak RAM use in gigabytes (GB) for simulations with selection.  The plots are from the
+        same data as in Figure \ref{fig:runtimes_selection}. Note that the y-axis scales differ considerably across
+    panels.}
+\end{figure}
+
+\newpage
+
+\section{Effect of simplification interval on run time and memory use}
+\label{ss:gcinterval}
+\renewcommand{\thefigure}{C\arabic{figure}}
+
+We performed a limited set of simulations  to explore the effect of the simplification interval on run times and memory
+use. Figure~\ref{sfig:gctime} shows that run times increase for simplification intervals less than 100 generations.  For
+intervals $\geq 100$ generations, run times are very similar, with no appreciable difference between $10^3$ and $10^4$
+generations.  As expected, the simplification interval has a near-linear effect on peak RAM use
+(Figure~\ref{sfig:gcmem}).  Taken together Figures~\ref{sfig:gctime} and~\ref{sfig:gcmem} suggest that one can tune the
+simplification interval to available RAM with little effect on run times provided that simplification doesn't occur too
+often.
+
+\begin{figure}[!h]
+    \includegraphics{sims/gc_interval/GCtime}
+    \caption{\label{sfig:gctime}The effect of simplification interval on run times.  A single replicate was done for the
+        parameters shown in the figure.  These simulations were done with the same selection parameters as in
+    Figure~\ref{fig:runtimes_selection}.}
+\end{figure}
+
+\begin{figure}[!h]
+    \includegraphics{sims/gc_interval/GCmem}
+    \caption{\label{sfig:gcmem}Peak RAM use for the simulations shown in Figure~\ref{sfig:gctime}.}
+\end{figure}
+
+\newpage
+\section{Timing and memory use with no selection}
+\label{ss:timing_nosel}
+\renewcommand{\thefigure}{D\arabic{figure}}
+\setcounter{figure}{0}
+
+We also performed a more limited set of simulations without natural selection.
+For these we used \cpp{} as in the main text, but also provide results from our
+proof-of-concept implementation that uses \simupop{} (see
+Appendix~\ref{ss:simupop}).
+
+The total run times are shown in
+Figure~\ref{sfig:rawspeed_nosel} and for \cpp{} show the same qualitative behavior as simulations with selection
+(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to pedigree tracking is again substantial in these
+simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to pedigree tracking here with \fwdpy{}
+than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
+close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
+more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
+and~\ref{sfig:speedup_nosel}.
+For \simupop{}, the results show a qualitatively similar pattern but 1) the
+simulations for $N = 10000$ or greater did no finish in the 72 hour limit or
+exceed memory limits (more than 500 GB), and 2) the magnitude of the speedup is
+much less.
+
+The total RAM use for the
+simulations without selection in Figure~\ref{sfig:ramnosel}.
+Pedigree tracking used more RAM with \fwdpp{} and less
+RAM with \simupop{}. The reason for reduced RAM use with \simupop{} is unclear;
+with \fwdpy{}/\fwdpp{}, however the reason for the extra RAM use is as in
+Appendix~\ref{ss:memuse}.
+
+\begin{figure}[!h]
+    \includegraphics[]{sims/rawspeed_nosel}
+    \caption{\label{sfig:rawspeed_nosel}Total run time for a single simulation replicate as a function of region
+        length, measured as the scaled recombination parameter $\rho = 4Nr$.  The simulations here involve no natural
+    selection.}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[]{sims/speedup_nosel}
+    \caption{\label{sfig:speedup_nosel}Relative performance improvement due to pedigree tracking for simulations without
+    selection.  Data points are taken from Figure~\ref{sfig:rawspeed_nosel} and show the ratio of run times tracking
+neutral variation over the run times tracking the pedigree.}
+\end{figure}
+\begin{figure}
+    \includegraphics[]{sims/memuse_nosel}
+    \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
+    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
+\end{figure}
+
+\section{The \simupop{} implementation}
+\label{ss:simupop}
 
@@
 Because \simupop{} requires explicit locations for loci, we set the number of
-loci $n=\rho / (4Nr)$ and marked
-$k={\rm round}(\mu / (100 (\mu + \mu / 100))) = 0$ of these as selected loci.
-\jda{unfortunately probably what we did...}
-The selected loci were chosen as a random sample of all the loci.
+loci $n=\rho / (4Nr)$ and for these simulations without selection marked $0$ of
+these as selected loci.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements}
-Thanks to Brad Shaffer and Evan McCartney--Melstad for useful discussions.
-Funding for this project came from XXX
-
-\bibliography{references}
-
-\appendix
-
-%%%%%%%%%%%%%%
-\section{Simplification algorithm}
-\label{ss:simplify_algorithm}
-
-\plr{Replace this with Jerome's interval-tree implementation?}
-
-Here we describe the simplification algorithm.
-The flow is close to our implementation,
-but the description below uses set operations,
-while in the implementation,
-all ancestors are maintained as linked lists of segments
-ordered by left endpoint.
-As it goes, the algorithm
-records a list $M$ that gives the mapping
-from input nodes to output nodes,
-so that if $M[p] = u$ then the $p^\text{th}$ node in the input
-is the same as the $u^\text{th}$ node in the output.
-The current state of the algorithm is recorded in $\Al$,
-a list of ancestors
-each of which is a collection of ancestral segments.
-An ancestral segment is a triple $(\ell, r, x)$,
-indicating that the output node $x$ inherits from this ancestor
-on the genomic segment $[\ell, r)$.
-
-\begin{taocpalg}{S}{Simplify a tree sequence}
-{Input consists of
-    a list $S$ of sample IDs,
-    a list $\Nt_I$ of nodes (ordered by birth time),
-    a list $\Et_I$ of edges,
-    and the genome length, $L$.
-    The output is a
-    nodes $\Nt_O$,
-    and list of edges $\Et_O$.
-}
-
-\algstep{S0.}{Initialize.}{
-    Set $\Al[j] \leftarrow (0, L, j)$
-    and $\Nt_O[j] \leftarrow \Nt_I[S[j]]$ for $0 \le j < \len(S)$.
-    Set the current input parent index to $p=0$
-    and $M[u] \leftarrow -1$ for $0 \le u < \len(\Nt_I)$.
-}
-\krt{We need a verbal definition of $Q$}
-\algstep{S1.}{Input parent loop.}{Set $Q \leftarrow \emptyset$.
-}
-
-\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} on each edge $(\ell, r, p, c)$ in $\Et_I$
-    corresponding to parent $p$ (which stores the resulting segments in the merge queue $Q$).
-}
-
-\algstep{S3.}{Merge ancestors.}{Call Algorithm \algref{M} on $Q$
-    (which adds to $\Nt_O$ and $\Et_O$).
-}
-
-\algstep{S4.}{Next parent.}{Set $p \leftarrow p+1$;
-    if $p = \len(\Nt_I)$, terminate and return $\Nt_O$ and $\Et_O$;
-    otherwise, return to \algref{S1}.
-}
-
-\end{taocpalg}
-
-The algorithm steps back up through the pedigree,
-since nodes are ordered by time-since-birth.
-First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
-so, for every edge $(\ell, r, p, c)$,
-Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
-with $[u,v) \setdiff [\ell, r)$
-(deleting the segment or splitting in two if necessary),
-and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:
-
-\begin{taocpalg}{R}{Remove ancestry}
-    {Given an edge $(\ell, r, p, c)$,
-    and a list $\Al[c]$ of the $m$ ancestral segments corresponding to $c$,
-    remove segments overlapping $[\ell, r)$ from $\Al[c]$,
-    and add those overlapping segments to $Q$.
-    }
-
-    \algstep{R0.}{Initialize.}{
-        Set $j \leftarrow 0$.
-    }
-
-    \algstep{R1.}{Segment loop.}{
-    If $j=m$, terminate; else let $(u, v, x) = \Al[c][j]$.
-    (Here $x$ is the output node that inherits from this segment.)
-    }
-
-    \algstep{R2.}{Add to merge queue.}{If $[u,v) \cap [\ell,r) = \emptyset$,
-    skip to \algref{R4}.
-    Otherwise, append $(\max(u,\ell), \min(v,r), x)$ to $Q$.
-    }
-
-    \algstep{R3.}{Remove ancestry.}{
-    Delete $\Al[c][j]$ and insert in its place
-    $(\min(u,\ell), \max(v,\ell), x)$ (if $u<\ell$)
-    and(/or) $(\min(v,r), \max(v,r), x)$ (if $r<v$).
-    }
-
-    \algstep{R4.}{Iterate.}{Set $j \leftarrow j+1$ and return to \algref{R1}.
-    }
-
-\end{taocpalg}
-
-Next we must merge the ancestry segments in $Q$,
-creating the new ancestor corresponding to input node $p$,
-adding a node and edges to the output if any segments overlap,
-which implies that an output coalescence has occurred.
-Below we index elements of an ancestral segment $z = (\ell, r, u)$
-by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
-and $z.\lid = u$ for the output node ID.
-
-\begin{taocpalg}{M}{Merge ancestors}
-    {Merge the ancestral segments in $Q$,
-    and create the new ancestor $\Al[p]$.
-    This algorithm also appends to $\Nt_O$ and $\Et_O$,
-    and needs to know about $S$,
-    to check if input nodes are samples.
-    }
-
-    \algstep{M0.}{Initialize.}{
-        Set $z \leftarrow (0, 0, -1)$ and $d = \text{False}$.
-    }
-
-    \algstep{M1.}{Segment loop.}{If $\len(Q) = 0$, go to \algref{M9}.}
-
-    \algstep{M2.}{Find next segment overlap boundaries}{
-        Let $\ell_0 \leftarrow \min\{x.\lleft : x \in Q\}$
-        and $H \leftarrow \{x \in Q : x.\lleft = \ell_0\}$.
-        Also let $\ell_+ \leftarrow \min\{x.\lleft : x \in Q \setdiff H\}$
-        and $r = \min(\ell_+, \min\{x.\lright : x \in H\})$.
-    }
-
-    \algstep{M3.}{Copy single segment.}{
-        If $\len(H) = 1$,
-        set $\alpha \leftarrow (\ell, r, H[0].\lid)$.
-        If also $p \in S$, then
-        set $\alpha.\lid \leftarrow M[p]$ and
-        $\add(\Et_O, (\alpha.\lleft, \alpha.\lright, \Nt[p], \alpha.\lid))$.
-        Go to \algref{M7}.
-    }
-
-    \algstep{M4.}{New output node.}{
-        If $M[p] = -1$ set $M[p] \leftarrow \add(\Nt_O, \Nt_I[p])$
-        Set $\alpha \leftarrow (\ell, r, M[p])$.
-    }
-
-    \algstep{M6.}{Record edges and update $Q$.}{
-        For each $x \in H$, do $\add(\Et_O, (\ell, r, M[p], x.\lid))$.
-    }
-
-    \algstep{M7.}{Update $Q$.}{
-        Remove from $Q$ any segment $x \in H$ with $x.\lright = r$
-        and then set $x.\lleft = \ell$ for all remaining $x \in H$.
-    }
-
-    \algstep{M8.}{Add segment to ancestor.}{
-        If $z.\lright = \alpha.\lleft$ and $z.\lid = \alpha.\lid$ set $d = \text{True}$.
-        Append $\alpha$ to $\Al[p]$, set $z \leftarrow \alpha$, and return to \algref{M1}.
-    }
-
-    \algstep{M9.}{Defragment.}{
-        If $d$ is True, find and merge any adjacent segments in $\Al[p]$
-        with the same $\lid$.  Terminate.
-    }
-
-\end{taocpalg}
-\krt{I think the algo discussion is  clear.  I've not tried to implement it, which would be the real test, but I find the description fine other than needing to define $Q$ when first used.}
-
-%% \newpage
-%% \clearpage
-%% \section*{Supplementary Information}
-
-\section{Memory usage with selection}
-\label{ss:memuse}
-\renewcommand{\thefigure}{B\arabic{figure}}
-\setcounter{figure}{0}
-
-The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel} and for the
-simulations without selection in Figure~\ref{sfig:ramnosel}.  For our parameters, which involved simplifying every
-$10^3$ generations,  pedigree tracking requires more RAM with \fwdpp{} and less
-RAM with \simupop{}. The reason for reduced RAM use with \simupop{} is unclear;
-with \fwdpy{}/\fwdpp{}, however the reason for the extra RAM use is clear.  When pedigree tracking, a non-recombinant
-individual still requires one new edge and two new nodes to be allocated.  Without pedigree tracking, no new memory is
-allocated for such an offspring.  Thus, pedigree tracking accumulates a large amount of data in RAM until
-simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extra RAM needed
-can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
-require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four sets of unsimplified data are
-stored in RAM.  In practice, the RAM consumption can be minimized by simplifying more often.
-
-%% \krt{I've avoided getting into how this much RAM means parallelization is required to make more progress.}
-
-\begin{figure}
-    \includegraphics[]{sims/memuse}
-    \caption{\label{sfig:ramsel}Peak RAM use in gigabytes (GB) for simulations with selection.  The plots are from the
-        same data as in Figure \ref{fig:runtimes_selection}. Note that the y-axis scales differ considerably across
-    panels.}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[]{sims/memuse_nosel}
-    \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
-    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
-\end{figure}
-
-\newpage
-
-\section{Effect of simplification interval on run time and memory use}
-\label{ss:gcinterval}
-\renewcommand{\thefigure}{C\arabic{figure}}
-
-We performed a limited set of simulations with \fwdpy{} to explore the effect of the simplification interval on run times and memory
-use. Figure~\ref{sfig:gctime} shows that run times increase for simplification intervals less than 100 generations.  For
-intervals $\geq 100$ generations, run times are very similar, with no appreciable difference between $10^3$ and $10^4$
-generations.  As expected, the simplification interval has a near-linear effect on peak RAM use
-(Figure~\ref{sfig:gcmem}).  Taken together Figures~\ref{sfig:gctime} and~\ref{sfig:gcmem} suggest that one can tune the
-simplification interval to available RAM with little effect on run times provided that simplification doesn't occur too
-often.
-
-\begin{figure}[!h]
-    \includegraphics{sims/gc_interval/GCtime}
-    \caption{\label{sfig:gctime}The effect of simplification interval on run times.  A single replicate was done for the
-        parameters shown in the figure.  These simulations were done with the same selection parameters as in
-    Figure~\ref{fig:runtimes_selection}.}
-\end{figure}
-
-\begin{figure}[!h]
-    \includegraphics{sims/gc_interval/GCmem}
-    \caption{\label{sfig:gcmem}Peak RAM use for the simulations shown in Figure~\ref{sfig:gctime}.}
-\end{figure}
-
-\newpage
-\section{Timing and memory use with no selection}
-\label{ss:timing_nosel}
-\renewcommand{\thefigure}{D\arabic{figure}}
-\setcounter{figure}{0}
-
-We also performed a more limited set of simulations without natural selection.  The total run times are shown in
-Figure~\ref{sfig:rawspeed_nosel} and show the same qualitative behavior as simulations with selection
-(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to pedigree tracking is again substantial in these
-simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to pedigree tracking here with \fwdpy{}
-than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
-close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
-more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
-and~\ref{sfig:speedup_nosel}.
-
-\begin{figure}[!h]
-    \includegraphics[]{sims/rawspeed_nosel}
-    \caption{\label{sfig:rawspeed_nosel}Total run time for a single simulation replicate as a function of region
-        length, measured as the scaled recombination parameter $\rho = 4Nr$.  The simulations here involve no natural
-    selection.}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[]{sims/speedup_nosel}
-    \caption{\label{sfig:speedup_nosel}Relative performance improvement due to pedigree tracking for simulations without
-    selection.  Data points are taken from Figure~\ref{sfig:rawspeed_nosel} and show the ratio of run times tracking
-neutral variation over the run times tracking the pedigree.}
-\end{figure}
 \end{document}

### 2018-01-11 forwards_paper.tex
fix 'infer edges' mention

@@
 in this way, the simulation engine would not need to keep copies of the node
-and edge information and would not need to infer edges;
+and edge information (and in the more flexible \simupop{} implementation, would
+not need to infer edges);
 even greater gains in computational efficiency could then be obtained. To

### 2018-01-11 forwards_paper.tex
Updated numpy pi example.

--- 2018-01-11 forwards_paper.tex
@@
 The \msprime{} API provides a method to compute $\pi$ among arbitrary subsets of the
-samples in a tree sequence, which took about 1.2 seconds
-applied to the example tree sequence above with 1.1 million mutations
-in 200 megabases for 500,000 samples.
-Calculating $\pi$ from the numeric genotype matrix using numpy took \plr{HOW MANY} seconds.
+samples in a tree sequence, which took about 0.7 seconds when
+applied to an example simulation of 100 megabases of human-like
+sequence for 200,000 samples ($~\sim$ 500K sites). The correspoding
+numeric genotype matrix required about 95GiB of RAM, and
+calculating $\pi$ took about 66 seconds with NumPy.
+
+% This is done by running $ python msprime-examples.py benchmark-pi
 

### 2018-01-11 forwards_paper.tex
Move 'next steps' to discussion

--- 2018-01-11 forwards_paper.tex
@@
 Checking this is more difficult, but is implemented in \msprime{}'s API.
-% \jda{TODO add function name}  PLR: not sure this is necessary.
 For efficiency, \msprime{} makes several other sortedness requirements on the tables,
@@
 infinite sites model; we provide results from this implementation alongside the
-results from the \cpp implementation for simulations without selection  (see
+results from the \cpp\ implementation for simulations without selection  (see
 Appendix~\ref{ss:timing_nosel}).
@@
 % varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.
-
-\paragraph{Next steps.}
-While these methods work well, and indeed result in significant performance
-gains over non-ancestry based simulation, they are far from optimal. Because
-forwards time simulation is highly CPU intensive, most simulation engines
-are written in C or C++ out of necessity. Therefore, the simplest and most
-efficient approach would be to call the \msprime\ C library directly as part of
-the core simulation logic, in the manner outlined in Algorithm W. If designed
-in this way, the simulation engine would not need to keep copies of the node
-and edge information (and in the more flexible \simupop{} implementation, would
-not need to infer edges);
-even greater gains in computational efficiency could then be obtained. To
-facilitate this usage, a standalone, embeddable C library encompassing the core
-tree sequence functionality is planned, which should make recording and
-outputting tree sequences from any C/C++ code straightforward.
 
@@
 
-A final note:
+\paragraph{Next steps.}
+While our methods work well, and indeed result in significant performance
+gains over non-ancestry based simulation, they are far from optimal. Because
+forwards time simulation is highly CPU intensive, most simulation engines
+are written in C or C++ out of necessity. Therefore, the simplest and most
+efficient approach would be to call the \msprime\ C library directly as part of
+the core simulation logic, in the manner outlined in Algorithm W. If designed
+in this way, the simulation engine would not need to keep copies of the node
+and edge information (and in the more flexible \simupop{} implementation, would
+not need to infer edges);
+even greater gains in computational efficiency could then be obtained. To
+facilitate this usage, a standalone, embeddable C library encompassing the core
+tree sequence functionality is planned, which should make recording and
+outputting tree sequences from any C/C++ code straightforward.
+
+\paragraph{A final note:}
 in preparing this manuscript,

### 2018-01-11 forwards_paper.tex
Reposition msprime api section to just after 'working with tables'

@@
 %%%%%%
+\subsection*{The \msprime\ Tables API}
+
+The facilities for working with succinct tree sequences are implemented as part
+of the \msprime\ Python API, which provides a powerful platform for processing
+tree topology and mutation data. The new portions of \msprime{} that we discuss
+here are dedicated to tree sequence input and output using simple tables of
+data, so we refer to this as the `Tables API'. The four key tables (nodes,
+edges, sites and mutations) are described above.
+
+The Tables API is primarily designed to facilitate efficient interchange of
+data between programs or between different modules of the same program. We
+adopted a `columnar' design, i.e., where all the values for a
+particular column are stored in adjacent memory locations.
+There are many advantages to columnar storage---for example, since adjacent
+values in memory are from the same column, they tend to compress well,
+and suitable encodings can be chosen on a per-column basis~\citep{abadi2006integrating}.
+A particular advantage of this approach is that it enables very
+efficient copying of data, and in principle zero-copy data access
+(where a data consumer reads directly from the memory of a producer) is possible.
+Our implementation uses the NumPy C API~\citep{walt2011numpy} to efficiently copy
+data from Python into the low-level C library used to manipulate tree sequences.
+This architecture allows for data transfer rates of gigabytes per second
+(impossible under any text-based approach), while retaining excellent portability.
+NumPy's array interface provides a great deal of flexibility and efficiency,
+and makes it straightforward to transfer data from sources
+such as HDF5 \citep{hdf5}or Dask~\citep{dask}.
+For small scale data and debugging purposes, a simple text based format is also supported.
+
+The \msprime\ Python Tables API provides a general purpose toolkit for importing
+and processing succinct tree sequences. Interoperation with Python simulators
+is then straightforward.  The implementation we benchmark here uses
+\texttt{pybind11} (\url{https://github.com/pybind/pybind11/}) to inferface
+with the \fwdpp{} \cpp{} API \citep{fwdpp}. No modifications were
+required to the \fwdpp{} code base; rather, we simply need to bookkeep parent/offspring labels,
+and perform simple processing of the recombination breakpoints from each mating
+event to generate node and edge data. This information is then periodically copied
+to the \msprime\ Tables API, where it is sorted and simplified.
+
+\paragraph{Flexibility.}
+To demonstrate the flexibility provided by the Tables API and provide an
+implementation that decouples forward simulation internals from transfer of data
+to \msprime\, we implemented a version of the simulations described in
+``Simulation benchmarks'' separately in Python (see Appendix~\ref{ss:simupop}).
+For this proof-of-concept implementation, the simulation engine invokes
+callbacks at critical points of the simulation, and we infer nodes and edges
+from the information that is provided. Rows are appended to the tables
+one-by-one, and the tables are periodically sorted and simplified to control
+memory usage.
+It uses the \simupop{} library, interfacing with Python code to simulate a
+discrete but large number of sites instead of the infinite sites model; we
+provide results from this implementation alongside the results from the \cpp\
+implementation for simulations without selection  (see
+Appendix~\ref{ss:timing_nosel}).
+% $\mu$ and $r$ define per-base-pair mutation rate a
+% the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.  We
+% varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.
+
+% is designed to be easily ``plugged'' into any simulation engine.
+% One simply needs to fill the relevant columns,
+% and to send them to \msprime{} at regular intervals for simplification.
+% For example, in the benchmarking simulations above,
+% we used the \fwdpp{} \cpp{} API \citep{fwdpp}
+% to fill arrays of simple structures representing the node and edge data.
+% Doing so requires no modification of the \fwdpp{} code base.
+% Rather, we simply need to bookkeep parent/offspring labels,
+% and perform simple processing of the recombination breakpoints from each mating event.
+% Using \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}),
+% the node/edge arrays on the \cpp{} side are made visible, without copy, to Python as a multi-column numpy array,
+% which is processed by the \msprime{} tables API.
+% Our implementation also uses \texttt{pybind11}
+% to create a Python package extending \texttt{fwdpy11}
+% (a Python package based on \fwdpp{}; \url{https://github.com/molpopgen/fwdpy11}),
+% % (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
+% % API from a command-line C/\cpp{} application via an embedded Python interpreter.)
+
+%%%%%%
 \subsection*{Recording the pedigree in forwards time}
@@
 
-%%%%%%
-\subsection*{Overview of \msprime\ APIs that enable the method}
-
-The facilities for working with succinct tree sequences are implemented as part
-of the \msprime\ Python API, which provides a powerful platform for processing
-tree topology and mutation data. The new portions of \msprime{} that we discuss
-here are dedicated to tree sequence input and output using simple tables of
-data, so we refer to this as the `Tables API'. The four key tables (nodes,
-edges, sites and mutations) are described above.
-
-The Tables API is primarily designed to facilitate efficient interchange of
-data between programs or between different modules of the same program. We
-adopted a `columnar' design, i.e., where all the values for a
-particular column are stored in adjacent memory locations.
-There are many advantages to columnar storage---for example, since adjacent
-values in memory are from the same column, they tend to compress well,
-and suitable encodings can be chosen on a per-column basis~\citep{abadi2006integrating}.
-A particular advantage of this approach is that it enables very
-efficient copying of data, and in principle zero-copy data access
-(where a data consumer reads directly from the memory of a producer) is possible.
-Our implementation uses the NumPy C API~\citep{walt2011numpy} to efficiently copy
-data from Python into the low-level C library used to manipulate tree sequences.
-This architecture allows for data transfer rates of gigabytes per second
-(impossible under any text-based approach), while retaining excellent portability.
-NumPy's array interface provides a great deal of flexibility and efficiency,
-and makes it straightforward to transfer data from sources
-such as HDF5 \citep{hdf5}or Dask~\citep{dask}.
-For small scale data and debugging purposes, a simple text based format is also supported.
-
-The \msprime\ Python Tables API provides a general purpose toolkit for importing
-and processing succinct tree sequences. Interoperation with Python simulators
-is then straightforward.  The implementation we benchmark here uses
-\texttt{pybind11} (\url{https://github.com/pybind/pybind11/}) to inferface
-with the \fwdpp{} \cpp{} API \citep{fwdpp}. No modifications were
-required to the \fwdpp{} code base; rather, we simply need to bookkeep parent/offspring labels,
-and perform simple processing of the recombination breakpoints from each mating
-event to generate node and edge data. This information is then periodically copied
-to the \msprime\ Tables API, where it is sorted and simplified.
-
-\paragraph{Flexibility.}
-To demonstrate the flexibility provided by the Tables API and provide an
-implementation that decouples forward simulation internals from transfer of data
-to \msprime\, we implemented a version of the simulations described in
-``Simulation benchmarks'' separately in Python (see Appendix~\ref{ss:simupop}).
-For this proof-of-concept implementation, the simulation engine invokes
-callbacks at critical points of the simulation, and we infer nodes and edges
-from the information that is provided. Rows are appended to the tables
-one-by-one, and the tables are periodically sorted and simplified to control
-memory usage.
-It uses the \simupop{} library, interfacing with Python code (via our
-package \ftprime{}) to simulate a discrete but large number of sites instead of the
-infinite sites model; we provide results from this implementation alongside the
-results from the \cpp\ implementation for simulations without selection  (see
-Appendix~\ref{ss:timing_nosel}).
-% $\mu$ and $r$ define per-base-pair mutation rate a
-% the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.  We
-% varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.
-
-% is designed to be easily ``plugged'' into any simulation engine.
-% One simply needs to fill the relevant columns,
-% and to send them to \msprime{} at regular intervals for simplification.
-% For example, in the benchmarking simulations above,
-% we used the \fwdpp{} \cpp{} API \citep{fwdpp}
-% to fill arrays of simple structures representing the node and edge data.
-% Doing so requires no modification of the \fwdpp{} code base.
-% Rather, we simply need to bookkeep parent/offspring labels,
-% and perform simple processing of the recombination breakpoints from each mating event.
-% Using \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}),
-% the node/edge arrays on the \cpp{} side are made visible, without copy, to Python as a multi-column numpy array,
-% which is processed by the \msprime{} tables API.
-% Our implementation also uses \texttt{pybind11}
-% to create a Python package extending \texttt{fwdpy11}
-% (a Python package based on \fwdpp{}; \url{https://github.com/molpopgen/fwdpy11}),
-% % (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
-% % API from a command-line C/\cpp{} application via an embedded Python interpreter.)
-
 %%%%%%%%%%%%%%%%%%%%%%

### 2018-01-11 sims/simplify-alg.py
Partial implementatin of algorithm.

@@ -2,12 +2,117 @@
 Set-theoretic algorithm implementation of Algorithm S.
 """
 import numpy as np
+import heapq
 
 import msprime
 import intervaltree
 
 
+class Segment(object):
+    def __init__(self, left, right, id_):
+        if left >= right:
+            print("ERROR", left ,right)
+        assert left < right
+        self.left = left
+        self.right = right
+        self.id = id_
+
+    def __lt__(self, other):
+        return (self.left, self.right, self.id) < (other.left, other.right, other.id)
+
+    def __repr__(self):
+        return repr((self.left, self.right, self.id))
+
 def simplify(S, Ni, Ei, L):
+    No = msprime.NodeTable()
+    Eo = msprime.EdgeTable()
+    A = [[] for _ in range(len(Ni))]
+    for input_id in S:
+        output_id = No.add_row(time=Ni.time[input_id], flags=1)
+        A[output_id] = [Segment(0, L, output_id)]
+
+    for j in range(len(Ni)):
+        print(j, "\t", A[j])
+
+    for input_parent in range(len(Ni)):
+        print("input_parent = ", input_parent)
+        Q = []
+        for edge in [e for e in Ei if e.parent == input_parent]:
+            for seg in A[edge.child]:
+                if not (seg.right <= edge.left or edge.right <= seg.left):
+                    y = Segment(
+                            max(seg.left, edge.left),
+                            min(seg.right, edge.right), seg.id)
+                    heapq.heappush(Q, y)
+        output_parent = -1
+        while len(Q) > 0:
+            l0 = Q[0].left
+            print("l0 = ", l0)
+            H = []
+            while len(Q) > 0 and Q[0].left == l0:
+                H.append(heapq.heappop(Q))
+            r = min(x.right for x in H)
+            lp = r
+            if len(Q) > 0:
+                lp = Q[0].left
+            if len(H) == 1:
+                A[input_parent].append(H[0])
+            else:
+                if output_parent == -1:
+                    output_parent = No.add_row(time=Ni.time[input_parent])
+                for j, x in enumerate(H):
+                    print("COAL")
+                    print(A[input_parent])
+                    Eo.add_row(x.left, r, output_parent, x.id)
+                    if x.right != r:
+                        x.left = lp
+                        assert x.left != x.right
[…]

### 2018-01-11 forwards_paper.tex

--- 2018-01-11 forwards_paper.tex
@@
     \begin{center}
-        \includegraphics{sims/num_edges_50}
-        \includegraphics{sims/simplify_timings}
+        \includegraphics{sims/simplify-results}
     \end{center}
     \caption{
+        Time and space complexity of simplify.
         \textbf{(A)}
         Number of edges in the simplified tree sequence
-        for 50 Wright--Fisher simulations begun with no prior history,
-        as a function of number of generations.
-        Each line is one simulation, the black line gives the average,
-        and the dotted line is the upper bound of equation \eqref{eqn:edge_bound}.
-        \plr{Propose to add other figures as subfigures here; but better than this.}
+        for 10 replicate Wright--Fisher simulations with $N=50$ as a function
+        of number of generations.
+        Each line is one simulation, the heavy blue line gives the average,
+        and the dashed line is the upper bound of equation \eqref{eqn:edge_bound}.
         \textbf{(B)}
-        Time required to simplify the first $k$ edges
-        of \plr{what tree sequence?},
-        plotted against $k$.
+        Time required to simplify the first $k$ edges of a large (4.2GiB)
+        unsimplified tree sequence produced by a forwards-time simulation plotted
+        against $k$. The time scales linearly with the number of input edges.
         \textbf{(C)}
-        Time required to simplify
-        \plr{what tree sequence?}
-        down to a sample of size $n$,
-        plotted against $n$
-        (note the log scale;
-        the time scales logarithmically with $n$).
+        Time required to simplify the tree sequence resulting from a coalescent
+        simulation of 500K samples of a 200 megabase human chromosome
+        for a random sample of $n$ samples, plotted against $n$
+        (note the log scale; the time scales logarithmically with $n$).
         \label{fig:simplify_complexity}

### 2018-01-12 forwards_paper.tex
succinct ts as in #54

@@
     by simulating only those loci that may affect reproduction (those having non-neutral variants),
-    and recording history in the \emph{tree sequence} format introduced in the software package \msprime,
+    and recording history as a \emph{succinct tree sequence} as introduced in the software package \msprime,
     on which neutral mutations can be quickly placed afterwards.
@@
 the complete history of parent-offspring relationships of an entire population
-going back to a remote time -- and the genetic outcomes of each ancestral
-meiosis.
+going back to a remote time -- and the genetic outcomes of each ancestral meiosis.
 The embellished pedigree contains all the information necessary
 to construct the genealogical tree that relates each individual to every other
-at each position on the genome,
-i.e., the \emph{tree sequence} implemented in the coalescent simulator \msprime{}~\citep{kelleher2016efficient}.
+at each position on the genome.
 Combined with ancestral genotypes and the origins of new mutations,
@@
 of the extant population.
-\jda{The space complexity result seemed like a result in the intro; and maybe didn't include some detail
-	about assumed number of generations}
-
-In this paper, we describe storage methods for genealogies
-(and hence, genome sequence) and an algorithm for simplifying these,
-both implemented as improvements to the algorithmic tools and data structures in \msprime{}.
+
+In this paper, we describe a method for storing a \emph{succinct tree sequence} 
+(or, for brevity, ``tree sequence''),
+and, hence, genome sequence, as well as an algorithm for simplifying these.
+Both are implemented as improvements to the algorithmic tools and data structures in \msprime{}.
 We also describe a general method of using these tools to efficiently record,
@@
 While providing substantial savings in time and space, our methods provide much
-more information than simply simulating the genomes -- the tree sequence format
-encodes all marginal genealogies of individuals living at the end of the
-simulation.
+more information than simply simulating the genomes -- the tree sequence
+encodes all marginal genealogies of individuals living at the end of the simulation.
 Although we were motivated by a need for more efficient genomic simulations,

### 2018-01-12 forwards_paper.tex
numpy ref #57

@@
 (where a data consumer reads directly from the memory of a producer) is possible.
-Our implementation uses the NumPy C API~\citep{walt2011numpy} to efficiently copy
-data from Python into the low-level C library used to manipulate tree sequences.
+Our implementation 
+% uses the NumPy C API~\citep{walt2011numpy} to efficiently copy
+efficiently copies data from Python as a NumPy array \citep{walt2011numpy} 
+into the low-level C library used to manipulate tree sequences.
 This architecture allows for data transfer rates of gigabytes per second

### 2018-01-12 forwards_paper.tex
abstract

@@
     We implement the method in two popular forwards-time simulation frameworks,
-    which speeds up large, whole-genome simulations by one or two orders of magnitude.
-    In addition to speed, we show that the method has several advantages:
+    which speeds up large, whole-genome simulations by one to orders of magnitude.
+    In addition to speed, the method has several advantages:
     (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
     (2) A population of $N$ individuals with $M$ polymorphic sites
-    can be stored on $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation
+    can be stored in $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation (and its' history)
     rather than only a subset.
-    (3) A simulation can easily be initialized with prior history produced by a more efficient coalescent simulation of deep history.
+    (3) A simulation can easily be initialized with a more efficient coalescent simulation of deep history.
     To make these possible,
-    we introduce the \msprime{} Tables API, which allows for the efficient interchange of genealogical data between modular program components.
+    we introduce the \msprime{} Tables API, which allows for the efficient interchange 
+    of tree sequences between modular program components,
+    and show how to quickly simplify a tree sequence by removing history irrelevant to a given set of genomes.
 \end{abstract}

### 2018-01-12 forwards_paper.tex
edited through

@@
 \newcommand{\fwdpy}{\texttt{fwdpy11}}
+\newcommand{\justc}{\texttt{C}}
 \newcommand{\cpp}{\texttt{C++}}
@@
 \cite{Sanjak2017-ko} used
-fwdpp \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
+\fwdpp{} \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
 population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
@@
 As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
-naively carrying full genotypes for even modest numbers of individuals through a simulation
-becomes quickly prohibitive.
+carrying full genotypes for even modest numbers of individuals through a simulation
+can quickly become prohibitive.
 To make matters worse,
@@
 can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
-for tens of thousands of generations, and there are still near-order-of-magnitude performance differences between these
-implementations~\citep{haller2017flexible}.
+for tens of thousands of generations.
+%% plr: not sure why we need to say these things here?
+% and there are still near-order-of-magnitude performance differences between these implementations~\citep{haller2017flexible}.
 In practice, current state-of-the-art simulation software may take on the order of
@@
 However, it is thought that most genetic variation is selectively neutral (or nearly so).
-By definition, the neutral alleles carried by individuals in a population
+By definition, neutral alleles carried by individuals in a population
 do not affect the population process.
@@
 simply laying down neutral mutations on top of that history afterwards
-is equivalent to having generated them during the simulation.
-Precisely, we would need to know the genealogical tree relating all sampled individuals
+is equivalent to having generated them during the simulation:
+it does not matter if we generate each generation's mutations during the simulation, or afterwards.
+Precisely, we need to know the genealogical trees relating all sampled individuals
 at each position along the genome.
-This is equivalent to knowing the \emph{population pedigree} --
+Combined with ancestral genotypes and the origins of new mutations,
+these trees completely specify the genomic sequence of any individual in the population at any time.
+To obtain this information, we record from forward simulation the \emph{population pedigree} --
 the complete history of parent-offspring relationships of an entire population
-going back to a remote time -- and the genetic outcomes of each ancestral meiosis.
-The embellished pedigree contains all the information necessary
+going back to a remote time -- and the genetic outcomes of each ancestral meiosis,
+periodically discarding all information irrelevant to the genetic history
+of the extant population.
+The information in this embellished pedigree is stored as a \emph{succinct tree sequence} 
+(or, for brevity, ``tree sequence''),
+which contains all the information necessary
 to construct the genealogical tree that relates each individual to every other
 at each position on the genome.
-Combined with ancestral genotypes and the origins of new mutations,
-it also completely specifies the genomic sequence of any individual in the population at any time.
-This is much much more than typically recorded in a forward simulation, however,
-so further savings can be had by discarding all information irrelevant to the genetic history
-of the extant population.
-
-In this paper, we describe a method for storing a \emph{succinct tree sequence} 
-(or, for brevity, ``tree sequence''),
+
+In this paper, we describe a storage method for \emph{succinct tree sequences} 
 and, hence, genome sequence, as well as an algorithm for simplifying these.
+The data structure is \emph{succinct} in the sense that its' space usage is close to optimal,
+while still allowing efficient retrieval of information.
 Both are implemented as improvements to the algorithmic tools and data structures in \msprime{}.
-We also describe a general method of using these tools to efficiently record,
-and later process, the population pedgiree in forwards-time simulation.
-While providing substantial savings in time and space, our methods provide much
+We also describe how these tools can efficiently record,
+and later process, the embellished population pedigree from a forwards-time simulation.
+While providing substantial savings in computational time and space, our methods provide much
 more information than simply simulating the genomes -- the tree sequence
@@
 
-We begin by benchmarking the performance improvement achieved by connecting
-a previously published forwards-time simulation library,
-\fwdpp{} \citep{fwdpp},
-to the tools we describe here, as implemented in \msprime{}.
+The strategy described above is only of interest if it is computationally feasible.
+Therefore,
+we begin by benchmarking the performance improvement achieved by this method,
+implemented using the forwards-time simulation library \fwdpp{} \citep{fwdpp}
+and tree sequence tools implemented in \msprime{}.
 Then, we describe the conceptual and algorithmic foundations for the method:
 (a) a format, implemented in the \msprime{} Python API,
-for recording tree sequences in several \emph{tables};
-(b) an algorithm to record tree sequence information into these tables on the fly
-    from a forwards-time simulation;
+for recording tree sequences succinctly in several \emph{tables};
+(b) an algorithm to record these tables on the fly during a forwards-time simulation;
 and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
@@
 
-To measure the performance gains from recording the pedigree we ran simulations both with and without recording.
-All simulations implemented a discrete-time Wright-Fisher population of diploid
-population size $N$, simulated for $10N$ generations.
-The simulations without pedigree recording introduced neutral mutations at a rate
+To measure the performance gains from recording the pedigree we ran simulations 
+both with and without recording.
+(Although we record more than just the parent--offspring relationships of the pedigree,
+for brevity we refer to the method as ``pedigree recording''.)
+All simulations used \fwdpp{} to implement a discrete-time Wright-Fisher population of $N$ diploid individuals,
+simulated for $10N$ generations (details below).
+Simulations without pedigree recording introduced neutral mutations at a rate
 equal to the recombination rate,
-so $\mu = r$, where $\mu$ and $r$ are the expected number of mutations per gamete and recombination breakpoints per
-diploid,
-respectively, each generation.
+so $\mu = r$, where $\mu$ and $r$ are the expected per-generation number of mutations per gamete 
+and recombination breakpoints per diploid, respectively.
+Simulations with pedigree recording introduced neutral mutations at the same rate
+retrospectively, as described below, resulting in statistically identical simulation results.
 % Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,
 % and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
-We ran simulations with different values of $N$ and varied the size of the region according to the scaled recombination
-parameter $\rho = 4Nr$.  When simulating neutral mutations, we kept the scaled mutation parameter, $\theta = 4N\mu$
-equal to $\rho$.
+We ran simulations with different values of $N$ and varied the size of the genomic region according to the scaled recombination
+parameter $\rho = 4Nr$.  
+%% this was said already:
+% When simulating neutral mutations, we kept the scaled mutation parameter, $\theta = 4N\mu$, equal to $\rho$.
 
@@
 
-We implemented our method in \cpp{}, using \fwdpp{} library functions and
-interface code (details below), using a continuum-sites model for both mutation
-and recombination. Simulations were run using \fwdpy{}, a Python package based
-on \fwdpp.
-% $\mu$ and $r$ were the means of Poisson processes, and mutations used the infinite-sites model \citep{Kimura1969-uc}.
-% We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.
-We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
-hyperthreading enabled.
-We ran one simulation at a time and the machine was under minimal load otherwise.
-We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
-and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
-%% Our simulation scripts separately recorded the time spent in various steps of each run.
-
-%% \krt{currently, we are not doing anything with the detailed timings.  I think that is fine, and that we can delete last
-%%sentence of preceding paragraph.}
-
-The total run times for simulations with selection are shown in Figure~\ref{fig:runtimes_selection}.  When tracking
-neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for
-these simulations).  Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
-Pedigree tracking
-dramatically improves run times (right column of Figure~\ref{fig:runtimes_selection}).  For parameters that ran to completion when tracking neutral mutations and when tracking
-the pedigree, the relative speedup due to pedigree tracking is up to $\approx 50$ fold
-(Figure~\ref{fig:relative_speedup_selection}).  Pedigree tracking reduces runtimes more for larger
-population sizes and we observe increasing relative speedups as $4Nr$ increases for a given $N$
+% When tracking neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for these simulations).  
+Pedigree tracking dramatically reduced runtimes, as shown in Figure~\ref{fig:runtimes_selection},
+producing a relative speedup of up to $\approx 50$ fold relative to standard simulations that track neutral mutations
 (Figure~\ref{fig:relative_speedup_selection}). 
-Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
-partially obscures an important observation that the run times appear approximately linear in region size when tracking
-pedigree.  The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
-each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
-In a more limited set of neutral simulations we found the same qualitative
-behavior with a numerically larger benefit to pedigree tracking (see Appendix~\ref{ss:timing_nosel}).
-
-In our test runs, our method used substantially more RAM than simple forward simulations (see Appendix~\ref{ss:memuse}).
+Pedigree tracking results in greater relative speedups for larger $N$
+and we observe increasing relative speedups as $4Nr$ increases for a given $N$
+(Figure~\ref{fig:relative_speedup_selection}).
+Importantly, runtimes are approximately linear in region size $\rho$ when pedigree tracking
+(partially obscured by the log scale of the horizontal axis in Figure~\ref{fig:runtimes_selection}).
+%% don't know what an 'edge' is yet
+% The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
+% each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
+In a more limited set of neutral simulations we found the same qualitative behavior,
+and a numerically larger speedup to pedigree tracking (see Appendix~\ref{ss:timing_nosel}).
+
+In our implementation, simulations with pedigree recording 
+used substantially more RAM than simple forward simulations (see Appendix~\ref{ss:memuse}).
 This is unsurprising:
@@
     \includegraphics[]{sims/rawspeed}
-    \caption{\label{fig:runtimes_selection}Total run time for a single simulation replicate as a function of region
-        length, measured as the scaled recombination parameter $\rho = 4Nr$. The different colored lines represent
-        different diploid population sizes ($N$). The left column shows run times for
-        simulations including the tracking of neutral mutations. The right column of
-panels shows run times tracking the pedigree instead of the neutral mutations. The dashed line in the upper right panel shows
-the results for an implementation using \fwdpy{} where the pedigree simplification steps were handled in a separate thread of
-execution and fitness calculations were parallelized.}
+    \caption{
+        \label{fig:runtimes_selection}
+    Total run time per single simulation replicate as a function of region length.
+    % measured as the scaled recombination parameter $\rho = 4Nr$. 
+    Line color represents different diploid population sizes ($N$). 
+    The left figure shows run times for standard simulations including neutral mutations. 
+    The right column of panels shows run times of simulations that recorded the pedigree
+    and added neutral mutations afterwards.
+    The dashed line in the right panel shows results for an implementation using \fwdpy{} 
+    where the pedigree simplification steps were handled in a separate thread of execution 
+    and fitness calculations were parallelized.
+    Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
+    }
 \end{figure}
@@
     \includegraphics[]{sims/speedup}
-    \caption{\label{fig:relative_speedup_selection}Relative performance improvement due to pedigree tracking. This figure
-    shows the ratio of the run time tracking neutral mutations over the run time tracking the pedigree. Data points are taken
-from Figure~\ref{fig:runtimes_selection} for simulations that ran to completion using both methods.}
+    \caption{
+        \label{fig:relative_speedup_selection}
+        Relative speedup of simulations due to pedigree recording. 
+        Each line shows the ratio of total run times of standard simulations to 
+        those of simulations with pedigree recording. 
+        Data points are taken from Figure~\ref{fig:runtimes_selection} 
+        for simulations that ran to completion in both cases.
+    }
 \end{figure}
@@
 %%%%%%
-\subsection*{Working with a tree sequence through tables}
-
-The simulations above recorded information about each new individual
-in a collection of tables that together define a \emph{tree sequence}.
+\subsection*{Tables for succinct tree sequences}
+
+We now explain what we actually did to achieve a $50\times$ speedup.
+The ``pedigree recording'' simulations above recorded information about each new individual
+in a collection of tables that together define a \emph{succinct tree sequence}
+(or, simply ``tree sequence'').
 A {tree sequence} is an encoding for a sequence of correlated trees,
 such as those describing the history of a sexual population.
-A closely related structure is known as the \emph{ancestral recombination graph} --
-the {ARG} \citep{griffiths1991two,griffiths1997ancestral},
-where nodes are common ancestor or recombination events --
-which has been the subject of substantial study
-under the assumptions of coalescent
-theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
-However, the properties of the ARG as a computational structure have not
-been studied and, despite several efforts to standardise a common
-format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper.
-ARGs are rarely used
-in practise. In contrast, the algorithmic properties of tree sequence
-algorithms have been explored in detail~\citep{kelleher2016efficient},
-contributing substantially to the efficiency of the \msprime\ coalescent simulator.
-
 Tree sequences are efficient because branches that are shared by adjacent trees are stored once,
@@
 The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
-while information about variants are recorded as \emph{sites} and \emph{mutations}.
-The format for these four tables is given in more detail in the Methods;
-here we explain conceptually,
-using the example of Figure~\ref{fig:example_tree_sequence}. This formulation
-is derived from the ``coalescence records'' encoding of tree
+while information about variants is recorded as \emph{sites} and \emph{mutations};
+we give an example in Figure~\ref{fig:example_tree_sequence}. 
+This formulation is derived from the ``coalescence records'' encoding of tree
 sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
-and generalised to include a much larger class of tree topology.
+and generalised to include a more general class of tree topologies.
 
 The \emph{nodes} of a tree sequence
-correspond to the vertices in the individual genealogies along the sequence,
-and each node refers to a distinct ancestor.
-Since each node represents a specific ancestor, it has a unique ``time'',
-thought of as her birth time, which determines the height of any vertices
-she is associated with.
-Note that since each node time is equal to the amount of time since the {birth} of the
-corresponding parent, time is measured in clock time, not in meioses.
+correspond to the vertices in the individual genealogies along the sequence.
+Each node refers to a specific, distinct ancestor,
+and so has a unique ``time'',
+thought of as the node's birth time, which determines the height of any vertices
+the node is associated with.
+(Note that since each node time is equal to the amount of time since the {birth} of the
+corresponding parent, time is measured in clock time, not in meioses.)
 The example of Figure~\ref{fig:example_tree_sequence} has five nodes:
@@
 The \emph{edges} define how nodes relate to each other over specific genomic intervals.
-Each edge records:
+Each edge records
 % is a tuple $(\ell, r, p, c)$, where
@@
 Recovering the sequence of trees from this information is straightforward:
-each point along the genome that the tree topology changes
+each point along the genome at which the tree topology changes
 is accompanied by the end of some {edges} and the beginning of others.
@@
 encode the sequence information for each sample in an efficient way. Each site
-is associated with a position on the genome and an ancestral state. For example,
+records two things: its position on the genome and an ancestral state. 
+For example,
 in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
-2.5 with ancestral state `A' and the other at position 7.5 with ancestral state `G'. If
-no mutations occur, all samples inherit the ancestral state at a given site.
-Each mutation occurs above a specific node at a given site,
-and results in a specific derived state.
-Thus, all samples below the mutation's node in the tree will inherit this state,
+2.5 with ancestral state `A' and the other at position 7.5 with ancestral state `G'. 
+If no mutations occur at a given site, all nodes inherit the ancestral state.
+
+Each mutation records three things: the site at which it occurs,
+the first node to inherit the mutation, and the derived state.
+Thus, all nodes below the mutation's node in the tree will inherit this state,
 unless further mutations are encountered.
@@
 This encoding of a sequence of trees and accompanying mutational information is
-very concise. To illustrate this, we ran a simulation of 500,000 samples of a
-$200$ megabase human-like chromosome ($N_e=10^4$ and per-base mutation and
-recombination rates of $10^{-8}$ per generation) using \msprime. This resulted
+very concise. To illustrate this, we used \msprime{} to simulate 500,000 samples of a
+$200$ megabase chromosome with human-like parameters: $N_e=10^4$ and per-base mutation and
+recombination rates of $10^{-8}$ per generation. This resulted
 in about 1 million distinct marginal trees and $1.1$ million infinite-sites
-mutations. The HDF5 file encoding the nodes, edges, sites and mutations (as
+mutations. The HDF5 file encoding the node, edge, site and mutation tables (as
 described above) for this simulation consumed 157MiB of storage space. Using
@@
 here are dedicated to tree sequence input and output using simple tables of
-data, so we refer to this as the `Tables API'. The four key tables (nodes,
-edges, sites and mutations) are described above.
+data, as described above, so we refer to this as the ``Tables API''.
 
@@
 data between programs or between different modules of the same program. We
-adopted a `columnar' design, i.e., where all the values for a
+adopted a `columnar' design, where all the values for a
 particular column are stored in adjacent memory locations.
-There are many advantages to columnar storage---for example, since adjacent
+There are many advantages to columnar storage -- for example, since adjacent
 values in memory are from the same column, they tend to compress well,
@@
 efficient copying of data, and in principle zero-copy data access
-(where a data consumer reads directly from the memory of a producer) is possible.
+(where a data consumer reads directly from the memory of a producer).
 Our implementation 
@@
 and makes it straightforward to transfer data from sources
-such as HDF5 \citep{hdf5}or Dask~\citep{dask}.
+such as HDF5 \citep{hdf5} or Dask~\citep{dask}.
 For small scale data and debugging purposes, a simple text based format is also supported.
@@
 is then straightforward.  The implementation we benchmark here uses
-\texttt{pybind11} (\url{https://github.com/pybind/pybind11/}) to inferface
+\texttt{pybind11} (\url{https://github.com/pybind/pybind11/}) to interface
 with the \fwdpp{} \cpp{} API \citep{fwdpp}. No modifications were
@@
 implementation that decouples forward simulation internals from transfer of data
-to \msprime\, we implemented a version of the simulations described in
-``Simulation benchmarks'' separately in Python (see Appendix~\ref{ss:simupop}).
-For this proof-of-concept implementation, the simulation engine invokes
-callbacks at critical points of the simulation, and we infer nodes and edges
+to \msprime, we also implemented a version of the simulations described in
+``Simulation benchmarks'' separately in Python, described in Appendix~\ref{ss:simupop}.
+This proof-of-concept implementation, the simulation engine (we use \simupop{}, \citet{simupop}) 
+invokes callbacks at critical points of the simulation, and we infer nodes and edges
 from the information that is provided. Rows are appended to the tables
@@
 memory usage.
-It uses the \simupop{} library, interfacing with Python code to simulate a
-discrete but large number of sites instead of the infinite sites model; we
-provide results from this implementation alongside the results from the \cpp\
-implementation for simulations without selection  (see
-Appendix~\ref{ss:timing_nosel}).
-% $\mu$ and $r$ define per-base-pair mutation rate a
-% the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.  We
-% varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.
-
-% is designed to be easily ``plugged'' into any simulation engine.
-% One simply needs to fill the relevant columns,
-% and to send them to \msprime{} at regular intervals for simplification.
-% For example, in the benchmarking simulations above,
-% we used the \fwdpp{} \cpp{} API \citep{fwdpp}
-% to fill arrays of simple structures representing the node and edge data.
-% Doing so requires no modification of the \fwdpp{} code base.
-% Rather, we simply need to bookkeep parent/offspring labels,
-% and perform simple processing of the recombination breakpoints from each mating event.
-% Using \texttt{pybind11} (\url{https://github.com/pybind/pybind11/}),
-% the node/edge arrays on the \cpp{} side are made visible, without copy, to Python as a multi-column numpy array,
-% which is processed by the \msprime{} tables API.
-% Our implementation also uses \texttt{pybind11}
-% to create a Python package extending \texttt{fwdpy11}
-% (a Python package based on \fwdpp{}; \url{https://github.com/molpopgen/fwdpy11}),
-% % (A neat aside, that we may wish to skip here, is that it may be possible to use the tables
-% % API from a command-line C/\cpp{} application via an embedded Python interpreter.)
+Benchmarking results from this implementation are shown (alongside results from \fwdpp{})
+for simulations without selection in Appendix~\ref{ss:timing_nosel}:
+a relatively modest speedup of around $5 \times$ is achieved, likely due to increased overhead.
 
@@
 
-To record the genealogical history of a forwards in time simulation
+To record the genealogical history of a forwards time simulation,
 we need to record two things for each new chromosome:
-the birth time, and the endpoints and parental IDs of each distinctly inherited segment,
-which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
-For concreteness, we write out in pseudocode how to run a neutral Wright--Fisher simulation
+the birth time; and the endpoints and parental IDs of each distinctly inherited segment.
+These are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
+To demonstrate the idea, we write out in pseudocode how to run a neutral Wright--Fisher simulation
 that records genealogical history in this way.
@@
 adds a new edge $(\ell\mathrm{eft}, r\mathrm{ight}, p\mathrm{arent}, c\mathrm{hild})$ to the edge table $\Et$.
-The function $\tsimplify{P, \Nt, \Et}$ (described in section \ref{ss:simplify})
+The function $\tsimplify{P, \Nt, \Et}$ (described below) % in section \ref{ss:simplify} 
 simplifies the history stored in the tables $\Nt$ and $\Et$
@@
 The tables are simplified every $s$ generations, removing genealogical
-information from $\Nt$ and $\Et$ unreachable from the current population $P$.
+information from $\Nt$ and $\Et$ irrelevant to the current population $P$.
 }
@@
 \algstep{W1.}{Initialisation.}{
-Set $\Nt \leftarrow \tnodetable()$, $\mathcal{E}
-\leftarrow \mbox{\texttt{EdgeTable}}()$, $t \leftarrow T$, and $j \leftarrow 0$.
+    Set $\Nt \leftarrow \tnodetable{}$, $\mathcal{E}
+    \leftarrow \tedgetable{}$, $t \leftarrow T$, and $j \leftarrow 0$.
  For $0 \leq k < N$, set $P_k \leftarrow \taddnode{\Nt, T}$.
@@
 \algstep{W4.}{Record edges.}{
-Call $\taddedge{\Et, 0, x, P_a, u}$ and $\taddedge{\Et, x, r, P_b, u}$.
+Call $\taddedge{\Et, 0, x, P_a, u}$ and $\taddedge{\Et, x, L, P_b, u}$.
 }
@@
 \algstep{W6.}{Simplify.}{Call $\tsimplify{P', \Nt, \Et}$, and set $P'_k
-\leftarrow k $ for $0 \leq k < N$. (Tables may need to be sorted.)}
+    \leftarrow k $ for $0 \leq k < N$. } %% too cryptic: (Tables may need to be sorted.)
 
@@
 ago. We also initialise our generation clock $t$ and individual index $j$.
-Steps~\algref{W2}--\algref{W4} replace the individual with node ID
-$P_j$ by first creating a new node with birth time $t$ (and ID $u$).
-In step~\algref{W3} we choose two indexes $a$ and $b$ uniformly,
+Step~\algref{W2} replaces the $j^\text{th}$ individual (with node ID $P_j$) 
+by creating a new node with birth time $t$ (and ID $u$).
+In step~\algref{W3} we determine the new node's ancestry 
+by choosing two indexes $a$ and $b$ uniformly,
 giving us parental IDs $P_a$ and $P_b$,
@@
 At the end of a generation, we then check
-if we need to simplify (which is done every $s$ generations).
+if we need to simplify (done every $s$ generations).
 If simplification is required, we do this in step \algref{W6} by calling the simplify function
@@
 
-\krt{The following paragraph has a sentence mangling.  I think some comment is warrented about when it is useful to
-record selected mutations, like for ancient samples?}
-
 This algorithm records only topological information about the simulated genealogies,
 but it is straightforward to add mutational information.
-This can be done in two different ways.
-We can record mutations that occur during the simulation by simply
-randomly sampling new mutations after recording the edges joining the new node $u$ to its parents
-$P_a$ and $P_b$ in step \algref{W7}.
-It is straightforward to record this information during the simulation, but it
-Alternatively, we can generate these mutations after the simulation has completed, thus
+Mutations that occur during the simulation can be recorded 
+by simply recording the node in which they first occur, the derived state,
+and (if not already present) the genomic position of the site at which it occurs.
+This allows selected mutations, that the forwards time simulation must generate,
+to be recorded in the tree sequence.
+Neutral mutations can be generated after the simulation has completed, thus
 avoiding the cost of generating the many mutations that are lost in the population.
@@
 ancestors in the current population, and so the entire path tracing back to
-the root is redundant.
+its common ancestor with 27 is redundant.
 
@@
 and wish to reduce this input tree sequence
-into the minimum representation of the topologies that include the specified samples.
-The output tree sequence must have the following properties:
+to the smallest one that still completely describes the history of the specified samples,
+having the following properties:
 \begin{enumerate}
-\item All marginal trees must match the subtree induced by the samples of the corresponding tree in the input tree sequence.
-\item Within the marginal trees, all non-sample vertices must has at least
+
+\item All marginal trees must match the subtree 
+        of the corresponding tree in the input tree sequence
+        that is induced by the samples.
+
+\item Within the marginal trees, all non-sample vertices must have at least
         two children (i.e., unary tree vertices are removed).
+
 \item Any nodes and edges not ancestral to any of the sampled nodes are removed.
+
 \item There are no adjacent redundant edges, i.e., pairs of edges
@@
     $(\ell, r, p, c)$.
+
 \end{enumerate}
@@
         An example of tree sequence simplification.
-        \textbf{Top:} the augmented pedigree diagram on the left
+        \textbf{(A)} The augmented pedigree diagram on the left
         relates the ten homologous chromosomes of five diploid individuals (BC, DE, FG, HI, and JK)
         to each other and to a common ancestral chromosome (A);
-        dotted lines connect the two chromosomes of each individual, and solid lines lead to the products of their meioses.
-        The corresponding tables (right) have 10 node records (one for each chromosome)
-        and 14 edge records (one for each distinctly inherited segment).
+        dotted lines connect the two chromosomes of each individual, 
+        and solid lines lead to the products of their meioses.
+        The corresponding tables (right) have 11 node records (one for each chromosome)
+        and 15 edge records (one for each distinctly inherited segment).
         Blue numbers denote crossing over locations --
@@
         $B$, $C$, $D$, and $E$ inherit clonally from $A$.
-        \textbf{Center:} the five distinct trees
+        \textbf{(B)} The five distinct trees
         found across the chromosome (blue numbers denote locations on the chromosome).
         Labels after simplification are shown in red.
-        \textbf{Bottom:} tables recording the tree sequence after simplification
+        \textbf{(C)} Tables recording the tree sequence after simplification
         with nodes $J$ and $K$ as samples.
@@
         each time two colors overlap, a coalescence occurs, and two edges are output.
-        For instance, both J and K inherit from H beteen 0.5 and 0.9,
+        For instance, both J and K inherit from H between 0.5 and 0.9,
         which resulted in the first two edges of the simplified table of Figure~\ref{fig:method_diagram}C.
@@
 % one list of segments for each ancestral lineage present at that time.
-We also maintain a map $A$ such that $A_j$ is the lineage
-corresponding to node $j$ in the input tree sequence.
+We also maintain a map from input nodes to lineages. % called $A_j$
 Crucially, the time required to run the algorithm is
@@
 providing an easy, efficient way to specify prior history.
-A natural question is now: how often should simplification occur?  We did some limited testing with the \cpp{}
-implementation of our simulations, and found that different simplification intervals affect run times by approximately
-25\%, with the lowest run time occuring when simplifying every $10^3$ generations.  Thus, there is a memory-versus-speed
-tradeoff as simplifying more often would keep fewer extinct nodes and edges in memory.  The run-time costs associated
+A natural question is now: how often should simplification occur?  
+Limited testing (described in Appendix~\ref{ss:gcinterval})
+found that different simplification intervals affect run times by approximately
+25\%, with the lowest run time occurring when simplifying every $10^3$ generations.  
+Thus, there is a memory-versus-speed tradeoff 
+-- simplifying more often would keep fewer extinct nodes and edges in memory.  
+The run-time costs associated
 with simplifying have to do with preparing the data to be sent to the simplification algorithm, rather than with the
@@
 Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
-($T \sim 20N$, say).
+($T = 20N$, say).
 After simplification, we are left with the tree sequence describing the history
@@
 Roughly, the length of time for which a marginal tree has $k$ tips is
-$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$,
+$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations,
 and so the time to go from $N$ to $n$ tips is $2N(1/(n-1) - 1/(N-1))$.
@@
         Time required to simplify the tree sequence resulting from a coalescent
-        simulation of 500K samples of a 200 megabase human chromosome
-        for a random sample of $n$ samples, plotted against $n$
+        simulation of 500,000 samples of a 200 megabase human chromosome
+        to a random subsample of $n$ samples, plotted against $n$
         (note the log scale; the time scales logarithmically with $n$).
@@
 %% simplify run time
-Above we have described memory requirements;
-how does the computation time required for simplification scale?
+How does the computation \emph{time} required for simplification scale?
 Simply because it must process each edge,
@@
 Empirically the algorithm is exactly linear,
-as seen in Figure~\ref{fig:simplify_complexity}X,
+as seen in Figure~\ref{fig:simplify_complexity}B,
 where we show shows the time required to simplify increasingly large subsets of a large tree sequence.
@@
 In this case, the computation is dominated by the size of the output tree sequence,
-which grows with $\log(n)$.
-this is also seen in Figure~\ref{fig:simplify_complexity}X,
-showing the time required to simplify progressively larger subsets of a single large tree sequence.
-
-Our method stores genealogies, and so records substantially more information
-than would a method only recording genotypes.
-However, since the simplification algorithm requires computational effort,
-it is informative to compare computational complexity of the algorithm
-to one that propagates neutral genotypes.
-A typical individual differs at around $2 N \mu$ sites from the population's consensus sequence,
-so propagating these to offspring by simple copying will take $4 N^2 \mu$ operations per generation.
-On the other hand,
-in our scheme we must store $13N$ quantities per generation (two edges and one node per individual).
-\plr{XXX}
-The simplification algorithm is linear in the number of edges of the input tree sequence
-(because it must process each edge),
-and so multiplies this by a constant factor.
-These considerations imply that propagating neutral genotypes for $T$ generations has time complexity $O(\mu T N^2)$,
-while our implementation is only $O(T N \log(2N))$.
+which grows with $\log(n)$, as shown in Figure~\ref{fig:simplify_complexity}C,
+
+%% the below assumes naive propagation of genotypes
+% Our method stores genealogies, and so records substantially more information
+% than would a method only recording genotypes.
+% However, since the simplification algorithm requires computational effort,
+% it is informative to compare time complexity of the algorithm
+% to one that propagates neutral genotypes.
+% A typical individual differs at around $2 N \mu$ sites from the population's consensus sequence,
+% so propagating these to offspring by simple copying will take $4 N^2 \mu$ operations per generation.
+% On the other hand,
+% in our scheme we must store $13N$ quantities per generation (two edges and one node per individual).
+% The simplification algorithm is linear in the number of edges of the input tree sequence
+% (because it must process each edge),
+% and so multiplies this by a constant factor.
+% These considerations imply that propagating neutral genotypes for $T$ generations has time complexity $O(\mu T N^2)$,
+% while our implementation is only $O(T N \log(2N))$.
 
@@
 A similar but much more limited method for
-discarding this information does appear in \citet{padhukasahasram2008exploring}.
-
-\krt{The following para contains numbers that are results, not discussion.}
+discarding this information also appears in \citet{padhukasahasram2008exploring}.
 
@@
 in a small fraction of the space taken by a compressed VCF file.
-The format is also highly efficient to process,
-leading to advantages for downstream analysis as well.
-The dowstream advantages are because many algorithms to compute statistics of interest for population genetics
+The format also allows highly efficient processing for downstream analysis.
+Efficient processing is possible because many statistics of interest for population genetics
 are naturally expressed in terms of tree topologies,
@@
 applied to an example simulation of 100 megabases of human-like
-sequence for 200,000 samples ($~\sim$ 500K sites). The correspoding
+sequence for 200,000 samples (about 500K sites). The corresponding
 numeric genotype matrix required about 95GiB of RAM, and
@@
 Recently, \citet{lawrie2017accelerating} used parallelization
-to improve performance for the ``Poisson Random Field'' (PRF) family of models~\citep{Sawyer1992-jw},
+to improve performance for the ``Poisson Random Field'' family of models~\citep{Sawyer1992-jw},
 in which mutations are unlinked and there is no effect of genetic background on fitness.
-However, thus far little attention has been paid to parallelization of simulations involving linkage, as our simulations
-where do, which is much more
-complex than the PRF model.
+However, thus far little attention has been paid to parallelization of more complex simulations.
+
+\paragraph{Relationship to the ARG}
+The idea of a tree sequence is closely related to the \emph{ancestral recombination graph}, 
+or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
+which also describes the embellished pedigree.
+The ARG has been the subject of substantial study
+under the assumptions of coalescent
+theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
+However, the properties of the ARG as a computational structure have not
+been studied and, despite several efforts to standardise a common
+format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper
+ARGs are rarely used in practise. 
+In contrast, the algorithmic properties of tree sequence
+algorithms have been explored in detail~\citep{kelleher2016efficient},
+contributing substantially to the efficiency of the \msprime{} coalescent simulator.
+
+\paragraph{Bayesian phylogenetics}
+The methods described here for efficiently storing tree sequences may prove useful in other fields.
+For instance, the large numbers of large, correlated trees produced by 
+MCMC samplers used in Bayesian phylogenetics \citep[e.g.,][]{drummond2012bayesian}
+might be compactly stored as a tree sequence,
+which would allow highly efficient computation of properties of the posterior distribution.
 
 \paragraph{Next steps.}
-While our methods work well, and indeed result in significant performance
-gains over non-ancestry based simulation, they are far from optimal. Because
+While our methods result in significant performance
+gains over current simulation methods, they are far from optimal. Because
 forwards time simulation is highly CPU intensive, most simulation engines
-are written in C or C++ out of necessity. Therefore, the simplest and most
-efficient approach would be to call the \msprime\ C library directly as part of
-the core simulation logic, in the manner outlined in Algorithm W. If designed
-in this way, the simulation engine would not need to keep copies of the node
-and edge information (and in the more flexible \simupop{} implementation, would
-not need to infer edges);
-even greater gains in computational efficiency could then be obtained. To
-facilitate this usage, a standalone, embeddable C library encompassing the core
-tree sequence functionality is planned, which should make recording and
-outputting tree sequences from any C/C++ code straightforward.
+are written in \justc{} or \cpp{} out of necessity. Therefore, the simplest and most
+efficient approach would be to call the \msprime{} \justc{} library directly as part of
+the core simulation logic, in the manner outlined in Algorithm \algref{W}. If designed
+in this way, the simulation engine would not need to keep copies of the node and edge information.
+To facilitate this usage, we plan to release a standalone, embeddable \justc{} library encompassing the core
+tree sequence functionality, which should make recording and
+outputting tree sequences from any \justc{}/\cpp{} code straightforward.
 
@@
 
-\paragraph{Requirements.}
-The \cpp{} implementation uses several packages installed via conda,
-except for \msprime{}, which must be installed from GitHub.
-
-\begin{enumerate}
-    \item \fwdpy{} 0.13.a0
-    \item \fwdpp{} 0.5.7
-    \item GNU Scientific Library 1.16
-    \item pybind11 2.2.1
-    \item GCC 4.8.5
-    \item \msprime{} 0.5.0 \jda{Changed these but the actual sims were not run
-		    with this version...}
-\end{enumerate}
+We implemented simulations and the connection to \msprime{} in \cpp{}, using \fwdpp{} library functions and
+interface code using a continuum-sites model for both mutation
+and recombination. Simulations were run using \fwdpy{} (version 0.13.a0), a Python package based
+on \fwdpp{} (version 0.5.7).
+Code for all simulations and figures is available at \url{https://github.com/petrelharp/ftprime_ms}.
+These made use of the GNU Scientific Library \citep[version 1.16][]{galassi2018scientific},
+pybind11 \citep[version 2.2.1][]{pybind11},
+and GCC (version 4.8.5).
+We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
+hyperthreading enabled.
+We ran one simulation at a time and the machine was under minimal load otherwise.
+We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
+and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
+%% Our simulation scripts separately recorded the time spent in various steps of each run.
 
@@
 \section*{Acknowledgements}
-Thanks to Brad Shaffer and Evan McCartney--Melstad for useful discussions.
-Funding for this project came from XXX
+Thanks to Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
+Work on this project was supported by funding from 
+the Sloan Foundation and the NSF (under DBI-1262645) to PR;
+and XXX.
 
@@
 
-%% \newpage
-%% \clearpage
-%% \section*{Supplementary Information}
-
 \section{Memory usage with selection}
@@
 The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel}.  For our parameters, which involved simplifying every
-$10^3$ generations,  pedigree tracking requires more RAM.  This is because with pedigree tracking, a non-recombinant
-individual still requires one new edge and two new nodes to be allocated.  Without pedigree tracking, no new memory is
-allocated for such an offspring.  Thus, pedigree tracking accumulates a large amount of data in RAM until
+$10^3$ generations, simulation while recording pedigrees uses more RAM than a standard simulation. 
+This is because with pedigree recording, a non-recombinant
+individual still requires one new edge and two new nodes to be allocated.  Without pedigree recording, no new memory is allocated for such an offspring by \fwdpp{}.
+Thus, pedigree recording accumulates a large amount of data in RAM until
 simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extra RAM needed
@@
 We performed a limited set of simulations to explore the effect of the simplification interval on run times and memory
-use. Figure~\ref{sfig:gctime} shows that run times increase for simplification intervals less than 100 generations.  For
-intervals $\geq 100$ generations, run times are very similar, with no appreciable difference between $10^3$ and $10^4$
+use. Figure~\ref{sfig:gctime} shows that run times increase for simplification intervals less than 100 generations.  
+For intervals longer than 100 generations, run times are very similar, with no appreciable difference between $10^3$ and $10^4$
 generations.  As expected, the simplification interval has a near-linear effect on peak RAM use
@@
 
-\newpage
 \section{Timing and memory use with no selection}
@@
 and~\ref{sfig:speedup_nosel}.
-For \simupop{}, the results show a qualitatively similar pattern but 1) the
+For \simupop{}, the results show a qualitatively similar pattern but (1) the
 simulations for $N = 10000$ or greater did no finish in the 72 hour limit or
-exceed memory limits (more than 500 GB), and 2) the magnitude of the speedup is
-much less.
-
-The total RAM use for the
-simulations without selection in Figure~\ref{sfig:ramnosel}.
-Pedigree tracking used more RAM with \fwdpp{} and less
-RAM with \simupop{}. The reason for reduced RAM use with \simupop{} is unclear;
-with \fwdpy{}/\fwdpp{}, however the reason for the extra RAM use is as in
-Appendix~\ref{ss:memuse}.
+exceed memory limits (more than 500 GB), and (2) the magnitude of the speedup is much less.
+The total RAM use for 
+simulations without selection is shown in Figure~\ref{sfig:ramnosel}.
+% Pedigree recording again used more RAM with \fwdpp{} and less
+% RAM with \simupop{}. The reason for reduced RAM use with \simupop{} is unclear;
+% with \fwdpy{}/\fwdpp{}, however the reason for the extra RAM use is as in
+% Appendix~\ref{ss:memuse}.
 
@@
 
+To use the methods with \simupop{}, we created a python module
+(available at \url{https://github.com/ashander/ftprime})
+that manages communication between the forward simulator and \msprime{}.
 We implemented simulations in a Python script, usable as a command-line utility,
 taking input parameters for population size $N$ and scaled recombination rate $\rho$.
-As described in 'Simulation benchmarks', we set scaled mutation rate $\theta=\rho$,
-we used simplify interval of 1000 and simulated for $T=10N$ generations in each
-simulation, introducing deleterious mutations at a rate of $\rho/100$ per
-generation.
-Because \simupop{} requires explicit locations for loci, we set the number of
-loci $n=\rho / (4Nr)$ and for these simulations without selection marked $0$ of
-these as selected loci.
+As for the \fwdpp{} simulations above, we set scaled mutation rate $\theta=\rho$,
+used a simplify interval of 1000 generations and simulated for $T=10N$ generations in each
+simulation. 
+% introducing deleterious mutations at a rate of $\rho/100$ per generation.
+% Because \simupop{} requires explicit locations for loci, we set the number of
+% loci $n=\rho / (4Nr)$ and for these simulations without selection marked $0$ of
+% these as selected loci.
 
@@
 The \texttt{simuPOP.Population()} is then evolved.
-Before mating: selected mutations are created (using
-\texttt{simuPOP.SNPMutator}) and multiplicative selection occurs (using
-\texttt{simuPOP.PyMlSelector}) only on the selected loci;
+Before mating:
+% selected mutations are created (using
+% \texttt{simuPOP.SNPMutator}) and multiplicative selection occurs (using
+% \texttt{simuPOP.PyMlSelector}) only on the selected loci;
 the internal time of the \texttt{RecombCollector} instance is updated.
@@
 output by \simupop{} to the \texttt{RecombCollector} instance.
-After mating: nothing happens 999/1000 generations, every 1000 generations the tables are
-simplifed.
+After mating: every 1000 generations the tables are simplified.
 
@@
 \texttt{RecombCollector} not delegated to this internal class are computing
-haploid IDs from \simupop{}'s diploid IDs and parsing recomination data into
-data suitible for use with use with \texttt{add\_individual} and
+haploid IDs from \simupop{}'s diploid IDs and parsing recombination data into
+data suitable for use with use with \texttt{add\_individual} and
 \texttt{add\_record}; these functions are used to stream data into the tables as
@@
 
-\paragraph{Requirements.}
-The simulations as implemented depend on the class \texttt{RecombCollector} provided by
-the Python package \ftprime{} 0.0.6rc0,
-available at \url{https://www.github.com/ashander/ftprime}.
-Its requirements are encoded in an \texttt{environment.yml} file, compatible
-with conda, but \msprime{} must be updated to the version listed below via pip
-from GitHub \jda{TODO add command?}.
-
-In brief the requirements are:
-
-\begin{enumerate}
-    \item \msprime{} 0.5.0
-    \item \simupop{} 1.1.8.3
-    \item python 3.5
-\end{enumerate}
+% \paragraph{Requirements.}
+% The simulations as implemented depend on the class \texttt{RecombCollector} provided by
+% the Python package \ftprime{} 0.0.6rc0,
+% available at \url{https://www.github.com/ashander/ftprime}.
+% Its requirements are encoded in an \texttt{environment.yml} file, compatible
+% with conda, but \msprime{} must be updated to the version listed below via pip
+% from GitHub \jda{TODO add command?}.
+% 
+% In brief the requirements are:
+% 
+% \begin{enumerate}
+%     \item \msprime{} 0.5.0
+%     \item \simupop{} 1.1.8.3
+%     \item python 3.5
+% \end{enumerate}
 

### 2018-01-12 sims/quick_plot.py

@@ -57,7 +57,7 @@ for name, group in groups:
 
 ax_fwdpp.legend(loc='upper left',frameon=False)
 ax_fwdpp.set_title("With neutral mutations", fontsize='medium')
-ax_fwdpp_arg.set_title("With pedigree tracking", fontsize='medium')
+ax_fwdpp_arg.set_title("With pedigree recording", fontsize='medium')
 ax_fwdpp.set_ylabel("Run time (hours)")
 ax_fwdpp.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_fwdpp_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
@@ -130,11 +130,11 @@ for name, group in groups:
 
 ax_fwdpp.legend(loc='upper left',frameon=False)
 ax_fwdpp.set_title("fwdpy11 with neutral mutations", fontsize='medium')
-ax_fwdpp_arg.set_title("fwdpy11 with pedigree tracking", fontsize='medium')
+ax_fwdpp_arg.set_title("fwdpy11 with pedigree recording", fontsize='medium')
 ax_simupop.set_ylabel("Run time (hours)")
 ax_fwdpp.set_ylabel("Run time (hours)")
 ax_simupop.set_title("simuPOP with neutral mutations", fontsize='medium')
-ax_simupop_arg.set_title("simuPOP with pedigree tracking", fontsize='medium')
+ax_simupop_arg.set_title("simuPOP with pedigree recording", fontsize='medium')
 ax_simupop.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
 ax_simupop_arg.set_xticks([1e3, 1e4, 1e5])

### 2018-01-12 forwards_paper.tex
Minor updates (+ remove trailing whitespace).

@@
     We implement the method in two popular forwards-time simulation frameworks,
-    which speeds up large, whole-genome simulations by one to orders of magnitude.
+    which speeds up large, whole-genome simulations by one to two orders of magnitude.
     In addition to speed, the method has several advantages:
@@
 the Sloan Foundation and the NSF (under DBI-1262645) to PR;
+the Wellcome Trust (grant 100956/Z/13/Z to Gil McVean);
 and XXX.

### 2018-01-12 forwards_paper.tex
Added SDSL reference for succinct data structures.

@@
 The data structure is \emph{succinct} in the sense that its' space usage is close to optimal,
-while still allowing efficient retrieval of information.
+while still allowing efficient retrieval of information (see, e.g.,
+\citet{gog2014theory}).
 Both are implemented as improvements to the algorithmic tools and data structures in \msprime{}.

### 2018-01-12 sims/simplify-alg.py
Working version of simplify. Yay!

@@ -10,8 +10,6 @@ import intervaltree
 
 class Segment(object):
     def __init__(self, left, right, id_):
-        if left >= right:
-            print("ERROR", left ,right)
         assert left < right
         self.left = left
         self.right = right
@@ -31,11 +29,7 @@ def simplify(S, Ni, Ei, L):
         output_id = No.add_row(time=Ni.time[input_id], flags=1)
         A[output_id] = [Segment(0, L, output_id)]
 
-    for j in range(len(Ni)):
-        print(j, "\t", A[j])
-
     for input_parent in range(len(Ni)):
-        print("input_parent = ", input_parent)
         Q = []
         for edge in [e for e in Ei if e.parent == input_parent]:
             for seg in A[edge.child]:
@@ -46,46 +40,34 @@ def simplify(S, Ni, Ei, L):
                     heapq.heappush(Q, y)
         output_parent = -1
         while len(Q) > 0:
-            l0 = Q[0].left
-            print("l0 = ", l0)
-            H = []
-            while len(Q) > 0 and Q[0].left == l0:
-                H.append(heapq.heappop(Q))
-            r = min(x.right for x in H)
-            lp = r
+            l = Q[0].left
+            r = L
+            X = []
+            while len(Q) > 0 and Q[0].left == l:
+                x = heapq.heappop(Q)
+                X.append(x)
+                r = min(r, x.right)
             if len(Q) > 0:
-                lp = Q[0].left
-            if len(H) == 1:
-                A[input_parent].append(H[0])
+                r = min(r, Q[0].left)
+
+            if len(X) == 1:
+                x = X[0]
+                alpha = x
+                if len(Q) > 0 and Q[0].left < x.right:
+                    alpha = Segment(x.left, Q[0].left, x.id)
+                    x.left = Q[0].left
+                    heapq.heappush(Q, x)
             else:
                 if output_parent == -1:
                     output_parent = No.add_row(time=Ni.time[input_parent])
-                for j, x in enumerate(H):
-                    print("COAL")
-                    print(A[input_parent])
-                    Eo.add_row(x.left, r, output_parent, x.id)
-                    if x.right != r:
-                        x.left = lp
-                        assert x.left != x.right
+                alpha = Segment(l, r, output_parent)
+                for x in X:
+                    Eo.add_row(l, r, output_parent, x.id)
+                    if x.right > r:
+                        x.left = r
                         heapq.heappush(Q, x)
-                    if j == 0:
[…]

### 2018-01-12 forwards_paper.tex
First draft of single simplify algorithm.

@@
 \newcommand{\Al}{\mathcal{A}}  % list of ancestors
+\newcommand{\priorityq}[1]{\mbox{PriorityQueue}({#1})}
+\newcommand{\listnew}[1]{\mbox{List}(#1)}
+\newcommand{\pqpush}[2]{#1.\mbox{push}(#2)}
+\newcommand{\pqpop}[1]{#1.\mbox{pop}()}
+\newcommand{\listappend}[2]{#1.\mbox{append}(#2)}
+\newcommand{\ancsegment}[1]{\mbox{Segment}(#1)}
+\newcommand{\taddrow}[2]{#1.\mbox{addrow}(#2)}
+\newcommand{\attrparent}[1]{#1.\mbox{parent}}
+\newcommand{\attrchild}[1]{#1.\mbox{child}}
+\newcommand{\attrleft}[1]{#1.\mbox{left}}
+\newcommand{\attrright}[1]{#1.\mbox{right}}
+\newcommand{\attrid}[1]{#1.\mbox{id}}
 % mbox to avoid hyphenating
@@
 \algstep{S0.}{Initialize.}{
-    Set $\Al[j] \leftarrow (0, L, j)$
-    and $\Nt_O[j] \leftarrow \Nt_I[S[j]]$ for $0 \le j < \len(S)$.
-    Set the current input parent index to $p=0$
-    and $M[u] \leftarrow -1$ for $0 \le u < \len(\Nt_I)$.
-}
-\krt{We need a verbal definition of $Q$}
-\algstep{S1.}{Input parent loop.}{Set $Q \leftarrow \emptyset$.
-}
-
-\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} on each edge $(\ell, r, p, c)$ in $\Et_I$
-    corresponding to parent $p$ (which stores the resulting segments in the merge queue $Q$).
-}
-
-\algstep{S3.}{Merge ancestors.}{Call Algorithm \algref{M} on $Q$
-    (which adds to $\Nt_O$ and $\Et_O$).
-}
-
-\algstep{S4.}{Next parent.}{Set $p \leftarrow p+1$;
-    if $p = \len(\Nt_I)$, terminate and return $\Nt_O$ and $\Et_O$;
-    otherwise, return to \algref{S1}.
+    Set $\Al_u \leftarrow \listnew{}$ for $0 \leq u < |\Nt_I|$. Then,
+    for each $u$ in $S$, set $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ and
+    call $\listappend{\Al[u]}{\ancsegment{0, L, v}}$. Then set $Q \leftarrow \priorityq{}$
+    and $u \leftarrow 0$.
+}
+
+\algstep{S1.}{Select parent edges.}{
+    Set $E \leftarrow \{e: \attrparent{e} = u \mbox{ and } e \in \Et_I\}$.
+    Set $v \leftarrow -1$.
+}
+
+\algstep{S2.}{Insert edge/ancestry intersections.}{
+    For each $e\in E$ and each $x \in \Al[\attrchild{e}]$,
+    if $\attrright{x} > \attrleft{e}$ and $\attrright{e} > \attrleft{x}$,
+    set $\ell \leftarrow \max(\attrleft{x}, \attrright{e})$,
+    $r \leftarrow \min(\attrright{x}, \attrright{e})$,
+    $y \leftarrow \ancsegment{\ell, r, \attrid{x}}$ and call
+    $\pqpush{Q}{y}$.
+}
+
+\algstep{S3.}{Find segments with minimum left coordinate.}{
+    Set $\ell \leftarrow \attrleft{Q[0]}$, $r \leftarrow L$ and $X \leftarrow \mbox{List}()$.
+    Then, while $|Q| > 0$ and $\attrleft{Q[0]} = \ell$, set $x \leftarrow \pqpop{Q}$,
+    $r \leftarrow \min(r, \attrright{x})$ and call $\listappend{X}{x}$.
+    Afterwards, if $|Q| > 0$, set $r \leftarrow \min(r, \attrleft{Q[0]})$. If
+    $|X| > 1$ go to \algref{S5}.
+}
+
+\algstep{S4.}{No overlapping segments.}{
+    Set $x \leftarrow X[0]$ and $\alpha \leftarrow x$. If $|Q| > 0$ and
+    $\attrleft{Q[0]} < \attrright{x}$, set $\alpha \leftarrow
+    \ancsegment{\attrleft{x}, \attrright{Q[0]}, \attrid{x}}$,
+    $\attrleft{x} \leftarrow \attrleft{Q[0]}$ and call $\pqpush{Q}{x}$.
+    Go to \algref{S7}.
+}
+
+\algstep{S5.}{New output node.}{
+    If $v = -1$, set $v \leftarrow \taddrow{\Nt_0}{\Nt_I[u]}$.
+}
+
+\algstep{S6.}{Record edges.}{
+    Set $\alpha \leftarrow \ancsegment{\ell, r, v}$. Then for each $x \in X$
+    call $\taddrow{\Et_O}{l, r, v, \attrid{x}}$ and, if $\attrright{x} > r$,
+    set $\attrleft{x} \leftarrow r$ and call $\pqpush{Q}{x}$.
+}
+
+\algstep{S7.}{Left coordinate loop.}{
+    Call $\listappend{\Al}{\alpha}$. If $|Q| > 0$ return to \algref{S3}.
+}
+
+\algstep{S8.}{Parent loop.}{
+    Set $u \leftarrow u + 1$. If $u < |\Nt_I|$, go to \algref{S1}.
 }
@@
 
-\begin{taocpalg}{R}{Remove ancestry}
-    {Given an edge $(\ell, r, p, c)$,
-    and a list $\Al[c]$ of the $m$ ancestral segments corresponding to $c$,
-    remove segments overlapping $[\ell, r)$ from $\Al[c]$,
-    and add those overlapping segments to $Q$.
-    }
-
-    \algstep{R0.}{Initialize.}{
-        Set $j \leftarrow 0$.
-    }
-
-    \algstep{R1.}{Segment loop.}{
-    If $j=m$, terminate; else let $(u, v, x) = \Al[c][j]$.
-    (Here $x$ is the output node that inherits from this segment.)
-    }
-
-    \algstep{R2.}{Add to merge queue.}{If $[u,v) \cap [\ell,r) = \emptyset$,
-    skip to \algref{R4}.
-    Otherwise, append $(\max(u,\ell), \min(v,r), x)$ to $Q$.
-    }
-
-    \algstep{R3.}{Remove ancestry.}{
-    Delete $\Al[c][j]$ and insert in its place
-    $(\min(u,\ell), \max(v,\ell), x)$ (if $u<\ell$)
-    and(/or) $(\min(v,r), \max(v,r), x)$ (if $r<v$).
-    }
-
-    \algstep{R4.}{Iterate.}{Set $j \leftarrow j+1$ and return to \algref{R1}.
-    }
-
-\end{taocpalg}
-
 Next we must merge the ancestry segments in $Q$,
@@
 and $z.\lid = u$ for the output node ID.
-
-\begin{taocpalg}{M}{Merge ancestors}
-    {Merge the ancestral segments in $Q$,
-    and create the new ancestor $\Al[p]$.
-    This algorithm also appends to $\Nt_O$ and $\Et_O$,
-    and needs to know about $S$,
-    to check if input nodes are samples.
-    }
-
-    \algstep{M0.}{Initialize.}{
-        Set $z \leftarrow (0, 0, -1)$ and $d = \text{False}$.
-    }
-
-    \algstep{M1.}{Segment loop.}{If $\len(Q) = 0$, go to \algref{M9}.}
-
-    \algstep{M2.}{Find next segment overlap boundaries}{
-        Let $\ell_0 \leftarrow \min\{x.\lleft : x \in Q\}$
-        and $H \leftarrow \{x \in Q : x.\lleft = \ell_0\}$.
-        Also let $\ell_+ \leftarrow \min\{x.\lleft : x \in Q \setdiff H\}$
-        and $r = \min(\ell_+, \min\{x.\lright : x \in H\})$.
-    }
-
-    \algstep{M3.}{Copy single segment.}{
-        If $\len(H) = 1$,
-        set $\alpha \leftarrow (\ell, r, H[0].\lid)$.
-        If also $p \in S$, then
-        set $\alpha.\lid \leftarrow M[p]$ and
-        $\add(\Et_O, (\alpha.\lleft, \alpha.\lright, \Nt[p], \alpha.\lid))$.
-        Go to \algref{M7}.
-    }
-
-    \algstep{M4.}{New output node.}{
-        If $M[p] = -1$ set $M[p] \leftarrow \add(\Nt_O, \Nt_I[p])$
-        Set $\alpha \leftarrow (\ell, r, M[p])$.
-    }
-
-    \algstep{M6.}{Record edges and update $Q$.}{
-        For each $x \in H$, do $\add(\Et_O, (\ell, r, M[p], x.\lid))$.
-    }
-
-    \algstep{M7.}{Update $Q$.}{
-        Remove from $Q$ any segment $x \in H$ with $x.\lright = r$
-        and then set $x.\lleft = \ell$ for all remaining $x \in H$.
-    }
-
-    \algstep{M8.}{Add segment to ancestor.}{
-        If $z.\lright = \alpha.\lleft$ and $z.\lid = \alpha.\lid$ set $d = \text{True}$.
-        Append $\alpha$ to $\Al[p]$, set $z \leftarrow \alpha$, and return to \algref{M1}.
-    }
-
-    \algstep{M9.}{Defragment.}{
-        If $d$ is True, find and merge any adjacent segments in $\Al[p]$
-        with the same $\lid$.  Terminate.
-    }
-
-\end{taocpalg}
-\krt{I think the algo discussion is  clear.  I've not tried to implement it, which would be the real test, but I find the description fine other than needing to define $Q$ when first used.}
 

### 2018-01-12 sims/simplify-alg.py

@@ -25,20 +25,19 @@ def simplify(S, Ni, Ei, L):
     No = msprime.NodeTable()
     Eo = msprime.EdgeTable()
     A = [[] for _ in range(len(Ni))]
-    for input_id in S:
-        output_id = No.add_row(time=Ni.time[input_id], flags=1)
-        A[output_id] = [Segment(0, L, output_id)]
+    Q = []
 
-    for input_parent in range(len(Ni)):
-        Q = []
-        for edge in [e for e in Ei if e.parent == input_parent]:
-            for seg in A[edge.child]:
-                if not (seg.right <= edge.left or edge.right <= seg.left):
-                    y = Segment(
-                            max(seg.left, edge.left),
-                            min(seg.right, edge.right), seg.id)
+    for u in S:
+        output_id = No.add_row(time=Ni.time[u], flags=1)
+        A[u] = [Segment(0, L, output_id)]
+
+    for u in range(len(Ni)):
+        for e in [e for e in Ei if e.parent == u]:
+            for x in A[e.child]:
+                if x.right > e.left and e.right > x.left:
+                    y = Segment(max(x.left, e.left), min(x.right, e.right), x.id)
                     heapq.heappush(Q, y)
-        output_parent = -1
+        v = -1
         while len(Q) > 0:
             l = Q[0].left
             r = L
@@ -58,16 +57,16 @@ def simplify(S, Ni, Ei, L):
                     x.left = Q[0].left
                     heapq.heappush(Q, x)
             else:
-                if output_parent == -1:
-                    output_parent = No.add_row(time=Ni.time[input_parent])
-                alpha = Segment(l, r, output_parent)
+                if v == -1:
+                    v = No.add_row(time=Ni.time[u])
+                alpha = Segment(l, r, v)
                 for x in X:
-                    Eo.add_row(l, r, output_parent, x.id)
+                    Eo.add_row(l, r, v, x.id)
                     if x.right > r:
                         x.left = r
                         heapq.heappush(Q, x)
 
-            A[input_parent].append(alpha)
+            A[u].append(alpha)
 
     # Sort the output edges and compact them as much as possible into
     # the output table. We can probably skip this for the algorithm listing as

### 2018-01-12 forwards_paper.tex
Changed function notation slightly.

@@
 \newcommand{\Al}{\mathcal{A}}  % list of ancestors
-\newcommand{\priorityq}[1]{\mbox{PriorityQueue}({#1})}
-\newcommand{\listnew}[1]{\mbox{List}(#1)}
-\newcommand{\pqpush}[2]{#1.\mbox{push}(#2)}
-\newcommand{\pqpop}[1]{#1.\mbox{pop}()}
-\newcommand{\listappend}[2]{#1.\mbox{append}(#2)}
-\newcommand{\ancsegment}[1]{\mbox{Segment}(#1)}
-\newcommand{\taddrow}[2]{#1.\mbox{addrow}(#2)}
+\newcommand{\priorityq}[1]{\mbox{\textbf{PriorityQueue}}({#1})}
+\newcommand{\listnew}[1]{\mbox{\textbf{List}}(#1)}
+\newcommand{\pqpush}[2]{#1.\mbox{\textbf{push}}(#2)}
+\newcommand{\pqpop}[1]{#1.\mbox{\textbf{pop}}()}
+\newcommand{\listappend}[2]{#1.\mbox{\textbf{append}}(#2)}
+\newcommand{\ancsegment}[1]{\mbox{\textbf{Segment}}(#1)}
+\newcommand{\taddrow}[2]{#1.\mbox{\textbf{addrow}}(#2)}
 \newcommand{\attrparent}[1]{#1.\mbox{parent}}
@@
 % mbox to avoid hyphenating
-\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
-\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
-\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
-\newcommand{\tsort}[1]{\mbox{\texttt{sort}}(#1)}
-\newcommand{\tnodetable}[1]{\mbox{\texttt{NodeTable}}(#1)}
-\newcommand{\tedgetable}[1]{\mbox{\texttt{EdgeTable}}(#1)}
-\DeclareMathOperator{\len}{length}
-\DeclareMathOperator{\add}{add} % for adding to tables
-\newcommand{\lleft}{\mathrm{left}}  % left label
-\newcommand{\lright}{\mathrm{right}}
-\newcommand{\lid}{\mathrm{id}}
+\newcommand{\tsimplify}[1]{\mbox{\textbf{simplify}}(#1)}
+\newcommand{\tsort}[1]{\mbox{\textbf{sort}}(#1)}
+\newcommand{\tnodetable}[1]{\mbox{\textbf{NodeTable}}(#1)}
+\newcommand{\tedgetable}[1]{\mbox{\textbf{EdgeTable}}(#1)}
 
@@
 (and all such instances are independent).
-Given a node table $\Nt$, the function $\taddnode{\Nt, t}$
+Given a node table $\Nt$, the function $\taddrow{\Nt}{t}$
 adds a new node to the table $\Nt$ with time $t$
 and returns the ID of this new node.
-Similarly, the function $\taddedge{\Et, \ell, r, p, c}$
-adds a new edge $(\ell\mathrm{eft}, r\mathrm{ight}, p\mathrm{arent}, c\mathrm{hild})$ to the edge table $\Et$.
+Similarly, the function $\taddrow{\Et}{\ell, r, p, c}$
+adds a new edge $(\ell\mbox{eft}, r\mathrm{ight}, p\mathrm{arent}, c\mathrm{hild})$ to the edge table $\Et$.
 The function $\tsimplify{P, \Nt, \Et}$ (described below) % in section \ref{ss:simplify}
@@
     \leftarrow \tedgetable{}$, $t \leftarrow T$, and $j \leftarrow 0$.
- For $0 \leq k < N$, set $P_k \leftarrow \taddnode{\Nt, T}$.
-}
-
-\algstep{W2.}{Generation loop head: new node.}{Set $u \leftarrow \taddnode{\Nt, t}$ and $P'_j \leftarrow u$.
+ For $0 \leq k < N$, set $P_k \leftarrow \taddrow{\Nt}{T}$.
+}
+
+\algstep{W2.}{Generation loop head: new node.}{Set $u \leftarrow \taddrow{\Nt}{t}$ and $P'_j \leftarrow u$.
 }
@@
 \algstep{W4.}{Record edges.}{
-Call $\taddedge{\Et, 0, x, P_a, u}$ and $\taddedge{\Et, x, L, P_b, u}$.
+Call $\taddrow{\Et}{0, x, P_a, u}$ and $\taddrow{\Et}{x, L, P_b, u}$.
 }
@@
 
-The algorithm steps back up through the pedigree,
-since nodes are ordered by time-since-birth.
-First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
-so, for every edge $(\ell, r, p, c)$,
-Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
-with $[u,v) \setdiff [\ell, r)$
-(deleting the segment or splitting in two if necessary),
-and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:
-
-Next we must merge the ancestry segments in $Q$,
-creating the new ancestor corresponding to input node $p$,
-adding a node and edges to the output if any segments overlap,
-which implies that an output coalescence has occurred.
-Below we index elements of an ancestral segment $z = (\ell, r, u)$
-by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
-and $z.\lid = u$ for the output node ID.
+% The algorithm steps back up through the pedigree,
+% since nodes are ordered by time-since-birth.
+% First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
+% so, for every edge $(\ell, r, p, c)$,
+% Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
+% with $[u,v) \setdiff [\ell, r)$
+% (deleting the segment or splitting in two if necessary),
+% and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:
+
+% Next we must merge the ancestry segments in $Q$,
+% creating the new ancestor corresponding to input node $p$,
+% adding a node and edges to the output if any segments overlap,
+% which implies that an output coalescence has occurred.
+% Below we index elements of an ancestral segment $z = (\ell, r, u)$
+% by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
+% and $z.\lid = u$ for the output node ID.
 

### 2018-01-12 forwards_paper.tex
Updated description of alg S.

@@
 \newcommand{\attrright}[1]{#1.\mbox{right}}
-\newcommand{\attrid}[1]{#1.\mbox{id}}
+\newcommand{\attrnode}[1]{#1.\mbox{node}}
 % mbox to avoid hyphenating
@@
 
-\plr{Replace this with Jerome's interval-tree implementation?}
-
-Here we describe the simplification algorithm.
-The flow is close to our implementation,
-but the description below uses set operations,
-while in the implementation,
-all ancestors are maintained as linked lists of segments
-ordered by left endpoint.
-As it goes, the algorithm
-records a list $M$ that gives the mapping
-from input nodes to output nodes,
-so that if $M[p] = u$ then the $p^\text{th}$ node in the input
-is the same as the $u^\text{th}$ node in the output.
-The current state of the algorithm is recorded in $\Al$,
-a list of ancestors
-each of which is a collection of ancestral segments.
-An ancestral segment is a triple $(\ell, r, x)$,
-indicating that the output node $x$ inherits from this ancestor
-on the genomic segment $[\ell, r)$.
+Here we describe the simplification algorithm. The algorithm uses a few simple structures
+besides the node and edge tables discussed earlier. As the algorithm needs to track the mapping
+between input and output node IDs over specific genomic intervals, we use a $\ancsegment$
+type to track these individual mappings. Thus, a segment $x$ has three attributes:
+$\attrleft{x}$ and $\attrright{x}$ represents the genomic interval in the usual way,
+and $\attrnode{x}$ represents the output node ID that is assigned to this interval. The per-interval
+segments are assigned to input IDs via the $\Al$ mapping, which is the main state required
+by the algorithm. Each element $\Al[u]$
+is a list of ancestral segments, describing the output node ID mapped to input
+node ID $u$ over a set of intervals. For each input node, we need to find the process
+the resulting set of ancestral segments to find overlaps (and hence coalescences). We
+do this using a $\priorityq{}$ type. This priority queue maintains a list of ancestral
+segments in sorted order, such that the segements with the smallest left coordinate
+are at the start of the list. Given a priority queue $Q$, the operation
+$\pqpush{Q}{x}$ inserts the segment $x$ into the queue, and $\pqpop{Q}{x}$
+returns (and removes) the segment in $Q$ with the smallest left coordinate.
+To obtain the segment currently in the queue with the smallest left value we
+write $Q[0]$.
+
+A step-by-step explanation of the algorithm follows the algorithm listing.
 
@@
 
-\algstep{S0.}{Initialize.}{
+\algstep{S0.}{Initialisation.}{
     Set $\Al_u \leftarrow \listnew{}$ for $0 \leq u < |\Nt_I|$. Then,
@@
     $r \leftarrow \min(\attrright{x}, \attrright{e})$,
-    $y \leftarrow \ancsegment{\ell, r, \attrid{x}}$ and call
+    $y \leftarrow \ancsegment{\ell, r, \attrnode{x}}$ and call
     $\pqpush{Q}{y}$.
@@
     $\attrleft{Q[0]} < \attrright{x}$, set $\alpha \leftarrow
-    \ancsegment{\attrleft{x}, \attrright{Q[0]}, \attrid{x}}$,
+    \ancsegment{\attrleft{x}, \attrright{Q[0]}, \attrnode{x}}$,
     $\attrleft{x} \leftarrow \attrleft{Q[0]}$ and call $\pqpush{Q}{x}$.
@@
     Set $\alpha \leftarrow \ancsegment{\ell, r, v}$. Then for each $x \in X$
-    call $\taddrow{\Et_O}{l, r, v, \attrid{x}}$ and, if $\attrright{x} > r$,
+    call $\taddrow{\Et_O}{l, r, v, \attrnode{x}}$ and, if $\attrright{x} > r$,
     set $\attrleft{x} \leftarrow r$ and call $\pqpush{Q}{x}$.
@@
 \algstep{S7.}{Left coordinate loop.}{
-    Call $\listappend{\Al}{\alpha}$. If $|Q| > 0$ return to \algref{S3}.
+    Call $\listappend{\Al[u]}{\alpha}$. If $|Q| > 0$ return to \algref{S3}.
 }
@@
 
-% The algorithm steps back up through the pedigree,
-% since nodes are ordered by time-since-birth.
-% First, for each input node $p$, we must find all ancestral segments inheriting from $p$ --
-% so, for every edge $(\ell, r, p, c)$,
-% Algorithm \algref{R} replaces any ancestry segment $[u, v)$ held by $c$
-% with $[u,v) \setdiff [\ell, r)$
-% (deleting the segment or splitting in two if necessary),
-% and adding the removed segment $[u,v) \cap [\ell, r)$ to the merge queue $Q$:
-
-% Next we must merge the ancestry segments in $Q$,
-% creating the new ancestor corresponding to input node $p$,
-% adding a node and edges to the output if any segments overlap,
-% which implies that an output coalescence has occurred.
-% Below we index elements of an ancestral segment $z = (\ell, r, u)$
-% by writing $z.\lleft = \ell$ and $z.\lright = r$ for the endpoints,
-% and $z.\lid = u$ for the output node ID.
+The algorithm begins in \algref{S1} by allocating an empty list to store ancestral
+segments for each of the $|\Nt_I|$ input nodes, and then creating the initial
+state for the samples in $S$. For each input sample with node ID $u$, we
+add a row in the output node table $\Nt_O$, and create a mapping for this
+new node $v$. After this initial state has been created, we then allocate our
+priority queue $Q$, and set the current input ID $u$ to $0$.
+
+The algorithm steps back up through the pedigree and steps \algref{S1}--\algref{S8}
+consider each node $u$ in turn (note we are assuming that nodes are ordered by time-since-birth).
+In \algref{S1} we first find all edges in the input where the parent is equal to
+our current input ID $u$, and then set the corresponding output ID $v$ to $-1$.
+Then, in \algref{S2} we find all intersections between these edges and the
+mapped ancestry segments for their child nodes. For each of these intersections we
+create a new segment $y$ and insert it into the priority queue.
+
+After finding all overlaps and filling the priority queue with the resulting ancestry
+segments, we then need to process these segments. Steps \algref{S3}--\algref{S7}
+loop over the segments in the queue, considering each in turn until the queue is empty.
+Step \algref{S3} builds a list $X$ of all the segments that have the
+minimum left coordinate (along with some bookkeeping to keep track of where our next
+right coordinate $r$ is), removing them from $Q$. If there is only one element in $X$
+this means that we have no overlapping ancestral segments at this coordinate and
+we proceed to \algref{S4}.
+
+As no overlaps occur, there is no change in the mapping between input and output IDs
+and so the ancestry segment $\alpha$ is not changed (this is assigned to $\Al_u$ in \algref{S7}).
+However, there is a complication caused by other segments which may intersect with $x$,
+which have left coordinate greater then $\attrleft{x}$. If $x$ intersects with this value,
+then there are other overlaps we must consider. Therefore, we change the left coordinate
+of $x$ to this coordinate and reinsert it into $Q$ where it can be processed again
+later. After doing this, we proceed to \algref{S7}, update the ancestry mapping for
+the input node $u$, and loop back to \algref{S3} if $Q$ is not empty.
+
+On the other hand, if there is more than one segment in $X$, we know that an
+overlap among the ancestral segments has occured and we must update the output
+nodes and edges accordingly. In step \algref{S5} we first allocate a new output
+node $v$ for $u$, if it has not been done already (we may have many overlapping
+segments along the genome). Step \algref{S6} then continues by creating a new
+ancestry segment $\alpha$ mapping the current interval to the output node $v$.
+We then iterate over all segments in $X$, adding the appropriate output edges
+and editing and reinserting the ancestry segment $x$ into the priority queue if
+necessary, as before.
 

### 2018-01-12 forwards_paper.tex
method for parallel

--- 2018-01-12 forwards_paper.tex
@@
 \newcommand{\Al}{\mathcal{A}}  % list of ancestors
-\newcommand{\priorityq}[1]{\mbox{PriorityQueue}({#1})}
-\newcommand{\listnew}[1]{\mbox{List}(#1)}
-\newcommand{\pqpush}[2]{#1.\mbox{push}(#2)}
-\newcommand{\pqpop}[1]{#1.\mbox{pop}()}
-\newcommand{\listappend}[2]{#1.\mbox{append}(#2)}
-\newcommand{\ancsegment}[1]{\mbox{Segment}(#1)}
-\newcommand{\taddrow}[2]{#1.\mbox{addrow}(#2)}
-\newcommand{\attrparent}[1]{#1.\mbox{parent}}
-\newcommand{\attrchild}[1]{#1.\mbox{child}}
-\newcommand{\attrleft}[1]{#1.\mbox{left}}
-\newcommand{\attrright}[1]{#1.\mbox{right}}
-\newcommand{\attrid}[1]{#1.\mbox{id}}
 % mbox to avoid hyphenating
@@
     We implement the method in two popular forwards-time simulation frameworks,
-    which speeds up large, whole-genome simulations by one to two orders of magnitude.
+    which speeds up large, whole-genome simulations by one to orders of magnitude.
     In addition to speed, the method has several advantages:
@@
 The data structure is \emph{succinct} in the sense that its' space usage is close to optimal,
-while still allowing efficient retrieval of information (see, e.g.,
-\citet{gog2014theory}).
+while still allowing efficient retrieval of information.
 Both are implemented as improvements to the algorithmic tools and data structures in \msprime{}.
@@
 and recombination. Simulations were run using \fwdpy{} (version 0.13.a0), a Python package based
-on \fwdpp{} (version 0.5.7).
+on \fwdpp{} (version 0.5.7).  The majority of results are presented based on a single-threaded implementation.  However,
+we also implemented a parallelized version using Python's \texttt{queue.Queue} to run the simplification step in a separate Python
+thread. Our implementation allows a maxmimum of four simplification intervals to be in the queue at once. This
+parallelized version also performed fitness calculation in parallel using two threads of execution in C++.
+
 Code for all simulations and figures is available at \url{https://github.com/petrelharp/ftprime_ms}.
@@
 the Sloan Foundation and the NSF (under DBI-1262645) to PR;
-the Wellcome Trust (grant 100956/Z/13/Z to Gil McVean);
 and XXX.
@@
 \algstep{S0.}{Initialize.}{
-    Set $\Al_u \leftarrow \listnew{}$ for $0 \leq u < |\Nt_I|$. Then,
-    for each $u$ in $S$, set $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ and
-    call $\listappend{\Al[u]}{\ancsegment{0, L, v}}$. Then set $Q \leftarrow \priorityq{}$
-    and $u \leftarrow 0$.
-}
-
-\algstep{S1.}{Select parent edges.}{
-    Set $E \leftarrow \{e: \attrparent{e} = u \mbox{ and } e \in \Et_I\}$.
-    Set $v \leftarrow -1$.
-}
-
-\algstep{S2.}{Insert edge/ancestry intersections.}{
-    For each $e\in E$ and each $x \in \Al[\attrchild{e}]$,
-    if $\attrright{x} > \attrleft{e}$ and $\attrright{e} > \attrleft{x}$,
-    set $\ell \leftarrow \max(\attrleft{x}, \attrright{e})$,
-    $r \leftarrow \min(\attrright{x}, \attrright{e})$,
-    $y \leftarrow \ancsegment{\ell, r, \attrid{x}}$ and call
-    $\pqpush{Q}{y}$.
-}
-
-\algstep{S3.}{Find segments with minimum left coordinate.}{
-    Set $\ell \leftarrow \attrleft{Q[0]}$, $r \leftarrow L$ and $X \leftarrow \mbox{List}()$.
-    Then, while $|Q| > 0$ and $\attrleft{Q[0]} = \ell$, set $x \leftarrow \pqpop{Q}$,
-    $r \leftarrow \min(r, \attrright{x})$ and call $\listappend{X}{x}$.
-    Afterwards, if $|Q| > 0$, set $r \leftarrow \min(r, \attrleft{Q[0]})$. If
-    $|X| > 1$ go to \algref{S5}.
-}
-
-\algstep{S4.}{No overlapping segments.}{
-    Set $x \leftarrow X[0]$ and $\alpha \leftarrow x$. If $|Q| > 0$ and
-    $\attrleft{Q[0]} < \attrright{x}$, set $\alpha \leftarrow
-    \ancsegment{\attrleft{x}, \attrright{Q[0]}, \attrid{x}}$,
-    $\attrleft{x} \leftarrow \attrleft{Q[0]}$ and call $\pqpush{Q}{x}$.
-    Go to \algref{S7}.
-}
-
-\algstep{S5.}{New output node.}{
-    If $v = -1$, set $v \leftarrow \taddrow{\Nt_0}{\Nt_I[u]}$.
-}
-
-\algstep{S6.}{Record edges.}{
-    Set $\alpha \leftarrow \ancsegment{\ell, r, v}$. Then for each $x \in X$
-    call $\taddrow{\Et_O}{l, r, v, \attrid{x}}$ and, if $\attrright{x} > r$,
-    set $\attrleft{x} \leftarrow r$ and call $\pqpush{Q}{x}$.
-}
-
-\algstep{S7.}{Left coordinate loop.}{
-    Call $\listappend{\Al}{\alpha}$. If $|Q| > 0$ return to \algref{S3}.
-}
-
-\algstep{S8.}{Parent loop.}{
-    Set $u \leftarrow u + 1$. If $u < |\Nt_I|$, go to \algref{S1}.
+    Set $\Al[j] \leftarrow (0, L, j)$
+    and $\Nt_O[j] \leftarrow \Nt_I[S[j]]$ for $0 \le j < \len(S)$.
+    Set the current input parent index to $p=0$
+    and $M[u] \leftarrow -1$ for $0 \le u < \len(\Nt_I)$.
+}
+\krt{We need a verbal definition of $Q$}
+\algstep{S1.}{Input parent loop.}{Set $Q \leftarrow \emptyset$.
+}
+
+\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} on each edge $(\ell, r, p, c)$ in $\Et_I$
+    corresponding to parent $p$ (which stores the resulting segments in the merge queue $Q$).
+}
+
+\algstep{S3.}{Merge ancestors.}{Call Algorithm \algref{M} on $Q$
+    (which adds to $\Nt_O$ and $\Et_O$).
+}
+
+\algstep{S4.}{Next parent.}{Set $p \leftarrow p+1$;
+    if $p = \len(\Nt_I)$, terminate and return $\Nt_O$ and $\Et_O$;
+    otherwise, return to \algref{S1}.
 }
@@
 
+\begin{taocpalg}{R}{Remove ancestry}
+    {Given an edge $(\ell, r, p, c)$,
+    and a list $\Al[c]$ of the $m$ ancestral segments corresponding to $c$,
+    remove segments overlapping $[\ell, r)$ from $\Al[c]$,
+    and add those overlapping segments to $Q$.
+    }
+
+    \algstep{R0.}{Initialize.}{
+        Set $j \leftarrow 0$.
+    }
+
+    \algstep{R1.}{Segment loop.}{
+    If $j=m$, terminate; else let $(u, v, x) = \Al[c][j]$.
+    (Here $x$ is the output node that inherits from this segment.)
+    }
+
+    \algstep{R2.}{Add to merge queue.}{If $[u,v) \cap [\ell,r) = \emptyset$,
+    skip to \algref{R4}.
+    Otherwise, append $(\max(u,\ell), \min(v,r), x)$ to $Q$.
+    }
+
+    \algstep{R3.}{Remove ancestry.}{
+    Delete $\Al[c][j]$ and insert in its place
+    $(\min(u,\ell), \max(v,\ell), x)$ (if $u<\ell$)
+    and(/or) $(\min(v,r), \max(v,r), x)$ (if $r<v$).
+    }
+
+    \algstep{R4.}{Iterate.}{Set $j \leftarrow j+1$ and return to \algref{R1}.
+    }
+
+\end{taocpalg}
+
 Next we must merge the ancestry segments in $Q$,
@@
 and $z.\lid = u$ for the output node ID.
+
+\begin{taocpalg}{M}{Merge ancestors}
+    {Merge the ancestral segments in $Q$,
+    and create the new ancestor $\Al[p]$.
+    This algorithm also appends to $\Nt_O$ and $\Et_O$,
+    and needs to know about $S$,
+    to check if input nodes are samples.
+    }
+
+    \algstep{M0.}{Initialize.}{
+        Set $z \leftarrow (0, 0, -1)$ and $d = \text{False}$.
+    }
+
+    \algstep{M1.}{Segment loop.}{If $\len(Q) = 0$, go to \algref{M9}.}
+
+    \algstep{M2.}{Find next segment overlap boundaries}{
+        Let $\ell_0 \leftarrow \min\{x.\lleft : x \in Q\}$
+        and $H \leftarrow \{x \in Q : x.\lleft = \ell_0\}$.
+        Also let $\ell_+ \leftarrow \min\{x.\lleft : x \in Q \setdiff H\}$
+        and $r = \min(\ell_+, \min\{x.\lright : x \in H\})$.
+    }
+
+    \algstep{M3.}{Copy single segment.}{
+        If $\len(H) = 1$,
+        set $\alpha \leftarrow (\ell, r, H[0].\lid)$.
+        If also $p \in S$, then
+        set $\alpha.\lid \leftarrow M[p]$ and
+        $\add(\Et_O, (\alpha.\lleft, \alpha.\lright, \Nt[p], \alpha.\lid))$.
+        Go to \algref{M7}.
+    }
+
+    \algstep{M4.}{New output node.}{
+        If $M[p] = -1$ set $M[p] \leftarrow \add(\Nt_O, \Nt_I[p])$
+        Set $\alpha \leftarrow (\ell, r, M[p])$.
+    }
+
+    \algstep{M6.}{Record edges and update $Q$.}{
+        For each $x \in H$, do $\add(\Et_O, (\ell, r, M[p], x.\lid))$.
+    }
+
+    \algstep{M7.}{Update $Q$.}{
+        Remove from $Q$ any segment $x \in H$ with $x.\lright = r$
+        and then set $x.\lleft = \ell$ for all remaining $x \in H$.
+    }
+
+    \algstep{M8.}{Add segment to ancestor.}{
+        If $z.\lright = \alpha.\lleft$ and $z.\lid = \alpha.\lid$ set $d = \text{True}$.
+        Append $\alpha$ to $\Al[p]$, set $z \leftarrow \alpha$, and return to \algref{M1}.
+    }
+
+    \algstep{M9.}{Defragment.}{
+        If $d$ is True, find and merge any adjacent segments in $\Al[p]$
+        with the same $\lid$.  Terminate.
+    }
+
+\end{taocpalg}
+\krt{I think the algo discussion is  clear.  I've not tried to implement it, which would be the real test, but I find the description fine other than needing to define $Q$ when first used.}
 

### 2018-01-12 sims/simplify-alg.py
add comments to example

@@ -1,5 +1,10 @@
 """
 Set-theoretic algorithm implementation of Algorithm S.
+
+The example works by generating an initial TreeSequence
+for a sample of 10 haplotypes using msprime.  We then
+simplify the node/edge table in that TreeSequence with
+respect to the first three samples.
 """
 import numpy as np
 
@@ -194,6 +199,7 @@ def verify():
 if __name__ == "__main__":
     verify()
 
+    # Generate initial TreeSequence
     ts = msprime.simulate(10, recombination_rate=2, random_seed=1)
     nodes= ts.tables.nodes
     edges = ts.tables.edges
@@ -207,6 +213,9 @@ if __name__ == "__main__":
     #     edges.add_row(
     #         breakpoint_map[e.left], breakpoint_map[e.right], e.parent, e.child)
 
+    # Simplify nodes, edges 
+    # with respect to
+    # the following samples:
     sample = [0, 1, 2]
     ts1 = simplify(sample, nodes, edges, ts.sequence_length)
     # ts1 = simplify_loci(sample, nodes, edges, len(breakpoint_map) - 1)

### 2018-01-12 forwards_paper.tex
Light editing pass.

@@
     of tree sequences between modular program components,
-    and show how to quickly simplify a tree sequence by removing history irrelevant to a given set of genomes.
+    and provide an algorithm to quickly simplify a tree sequence by removing history irrelevant to a given set of genomes.
 \end{abstract}
@@
 In this paper, we describe a storage method for \emph{succinct tree sequences}
-and, hence, genome sequence, as well as an algorithm for simplifying these.
+(and hence, genome sequence) as well as an algorithm for simplifying these.
 The data structure is \emph{succinct} in the sense that its' space usage is close to optimal,
@@
 \citet{gog2014theory}).
-Both are implemented as improvements to the algorithmic tools and data structures in \msprime{}.
+The storage method and simplification algorithm are both implemented
+as improvements to the algorithmic tools and data structures in \msprime{}.
 We also describe how these tools can efficiently record,
@@
 (a) a format, implemented in the \msprime{} Python API,
-for recording tree sequences succinctly in several \emph{tables};
-(b) an algorithm to record these tables on the fly during a forwards-time simulation;
+for recording tree sequences efficiently in several \emph{tables};
+(b) an algorithm to record these tables during a forwards-time simulation;
 and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
@@
 In a more limited set of neutral simulations we found the same qualitative behavior,
-and a numerically larger speedup to pedigree tracking (see Appendix~\ref{ss:timing_nosel}).
+and a numerically larger speedup by using pedigree tracking (see Appendix~\ref{ss:timing_nosel}).
 
@@
 and generalised to include a more general class of tree topologies.
+
+The idea of a tree sequence is closely related to the \emph{ancestral recombination graph},
+or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
+which also describes the embellished pedigree.
+The ARG has been the subject of substantial study
+under the assumptions of coalescent
+theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
+However, the properties of the ARG as a computational structure have not
+been studied and, despite several efforts to standardise a common
+format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper
+ARGs are rarely used in practise.
+In contrast, the algorithmic properties of tree sequence
+algorithms have been explored in detail~\citep{kelleher2016efficient},
+contributing substantially to the efficiency of the \msprime{} coalescent simulator.
 
@@
 If no mutations occur at a given site, all nodes inherit the ancestral state.
-
 Each mutation records three things: the site at which it occurs,
@@
 ``Simulation benchmarks'' separately in Python, described in Appendix~\ref{ss:simupop}.
-This proof-of-concept implementation, the simulation engine (we use \simupop{}, \citet{simupop})
+In this proof-of-concept implementation, the simulation engine (we use \simupop{}, \citet{simupop})
 invokes callbacks at critical points of the simulation, and we infer nodes and edges
@@
 \algstep{W3.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
-    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
+    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform((0, L))$.
 }
@@
 Mutations that occur during the simulation can be recorded
-by simply recording the node in which they first occur, the derived state,
+by simply storing the node in which they first occur, the derived state,
 and (if not already present) the genomic position of the site at which it occurs.
@@
 
-It is desirable for many reasons to remove redundant information from a tree sequence,
-such as those produced by a forwards-time simulation.
+It is desirable for many reasons to remove redundant information from a tree sequence.
 To formalize this:
@@
 -- simplifying more often would keep fewer extinct nodes and edges in memory.
-The run-time costs associated
-with simplifying have to do with preparing the data to be sent to the simplification algorithm, rather than with the
-algorithm itself, which we show in the next section has linear time complexity.
 
@@
 sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
-can do this in roughly $O(n + m \log n)$ operations.
+can do this in roughly $O(n + m + t)$ operations, where
+$t$ is the number of distinct trees.
 The \msprime{} API provides a method to compute $\pi$ among arbitrary subsets of the
@@
 
-\paragraph{Relationship to the ARG}
-The idea of a tree sequence is closely related to the \emph{ancestral recombination graph},
-or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
-which also describes the embellished pedigree.
-The ARG has been the subject of substantial study
-under the assumptions of coalescent
-theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
-However, the properties of the ARG as a computational structure have not
-been studied and, despite several efforts to standardise a common
-format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper
-ARGs are rarely used in practise.
-In contrast, the algorithmic properties of tree sequence
-algorithms have been explored in detail~\citep{kelleher2016efficient},
-contributing substantially to the efficiency of the \msprime{} coalescent simulator.
-
 \paragraph{Bayesian phylogenetics}
 The methods described here for efficiently storing tree sequences may prove useful in other fields.
+We have focused on the interpretation of tree sequences as the outcome of the process
+of recombination, but in principle, we can efficiently encode any
+sequence of trees which differ by subtree-prune-and-regraft
+(SPR) operations. Since each SPR operation requires a constant amount of space to encode, the total
+space required is $O(n + t)$ for $t$ trees with $n$ leaves~\citep{kelleher2016efficient}.
 For instance, the large numbers of large, correlated trees produced by
@@
 might be compactly stored as a tree sequence,
-which would allow highly efficient computation of properties of the posterior distribution.
-
-\paragraph{Next steps.}
-While our methods result in significant performance
-gains over current simulation methods, they are far from optimal. Because
-forwards time simulation is highly CPU intensive, most simulation engines
-are written in \justc{} or \cpp{} out of necessity. Therefore, the simplest and most
-efficient approach would be to call the \msprime{} \justc{} library directly as part of
-the core simulation logic, in the manner outlined in Algorithm \algref{W}. If designed
-in this way, the simulation engine would not need to keep copies of the node and edge information.
-To facilitate this usage, we plan to release a standalone, embeddable \justc{} library encompassing the core
-tree sequence functionality, which should make recording and
-outputting tree sequences from any \justc{}/\cpp{} code straightforward.
+which would then allow highly efficient computation of properties of the posterior distribution.
+
+% JK: This needs more context. Probably not worth putting in here.
+% \paragraph{Next steps.}
+% While our methods result in significant performance
+% gains over current simulation methods, they are far from optimal. Because
+% forwards time simulation is highly CPU intensive, most simulation engines
+% are written in \justc{} or \cpp{} out of necessity. Therefore, the simplest and most
+% efficient approach would be to call the \msprime{} \justc{} library directly as part of
+% the core simulation logic, in the manner outlined in Algorithm \algref{W}. If designed
+% in this way, the simulation engine would not need to keep copies of the node and edge information.
+% To facilitate this usage, we plan to release a standalone, embeddable \justc{} library encompassing the core
+% tree sequence functionality, which should make recording and
+% outputting tree sequences from any \justc{}/\cpp{} code straightforward.
 

### 2018-01-12 forwards_paper.tex
edited simplify alg

--- 2018-01-12 forwards_paper.tex
@@
 Similarly, the function $\taddrow{\Et}{\ell, r, p, c}$
-adds a new edge $(\ell\mbox{eft}, r\mathrm{ight}, p\mathrm{arent}, c\mathrm{hild})$ to the edge table $\Et$.
+adds a new edge $(\ell\text{eft}, r\text{ight}, p\text{arent}, c\text{hild})$ to the edge table $\Et$.
 The function $\tsimplify{P, \Nt, \Et}$ (described below) % in section \ref{ss:simplify}
@@
 to the minimal information required to represent the genealogies of the list of node IDs $P$;
-after simplification the nodes appearing in $P$ are relabeled $(0, 1, \ldots, |P|)$.
+after simplification the nodes appearing in $P$ are relabeled $(0, 1, \ldots, |P|-1)$.
 A step-by-step explanation follows the pseudocode.
@@
 and recombination. Simulations were run using \fwdpy{} (version 0.13.a0), a Python package based
-on \fwdpp{} (version 0.5.7).  The majority of results are presented based on a single-threaded implementation.  However,
-we also implemented a parallelized version using Python's \texttt{queue.Queue} to run the simplification step in a separate Python
-thread. Our implementation allows a maxmimum of four simplification intervals to be in the queue at once. This
-parallelized version also performed fitness calculation in parallel using two threads of execution in C++.
-
+on \fwdpp{} (version 0.5.7).
 Code for all simulations and figures is available at \url{https://github.com/petrelharp/ftprime_ms}.
@@
 
-Here we describe the simplification algorithm. The algorithm uses a few simple structures
+Here we provide a concrete implementation of the simplification algorithm.
+The algorithm uses a few simple structures
 besides the node and edge tables discussed earlier. As the algorithm needs to track the mapping
@@
 type to track these individual mappings. Thus, a segment $x$ has three attributes:
-$\attrleft{x}$ and $\attrright{x}$ represents the genomic interval in the usual way,
-and $\attrnode{x}$ represents the output node ID that is assigned to this interval. The per-interval
-segments are assigned to input IDs via the $\Al$ mapping, which is the main state required
-by the algorithm. Each element $\Al[u]$
-is a list of ancestral segments, describing the output node ID mapped to input
-node ID $u$ over a set of intervals. For each input node, we need to find the process
-the resulting set of ancestral segments to find overlaps (and hence coalescences). We
-do this using a $\priorityq{}$ type. This priority queue maintains a list of ancestral
-segments in sorted order, such that the segements with the smallest left coordinate
-are at the start of the list. Given a priority queue $Q$, the operation
-$\pqpush{Q}{x}$ inserts the segment $x$ into the queue, and $\pqpop{Q}{x}$
-returns (and removes) the segment in $Q$ with the smallest left coordinate.
+$\attrleft{x}$ and $\attrright{x}$ represent the genomic interval in the usual way,
+and $\attrnode{x}$ represents the output node ID that is assigned to this interval.
+These output node IDs correspond to colors in the ``paint pot'' analogy.
+The per-interval segments are assigned to input IDs via the $\Al$ mapping,
+which is the main state required by the algorithm. Each element $\Al[u]$
+is a list of ancestral segments, describing the output node IDs mapped to input
+node ID $u$ over a set of intervals.
+For each input node, we need to process the ancestral segments
+to find overlaps (and hence coalescences), and to update the $\Al$.
+We do this using a $\priorityq{}$ type, which maintains a list of ancestral
+segments sorted by left coordinate. For a priority queue $Q$, the operation
+$\pqpush{Q}{x}$ inserts the segment $x$ into the queue, and $\pqpop{Q}$
+removes and returns the segment in $Q$ with the smallest left coordinate.
 To obtain the segment currently in the queue with the smallest left value we
 write $Q[0]$.
-
-A step-by-step explanation of the algorithm follows the algorithm listing.
 
@@
 \algstep{S1.}{Select parent edges.}{
-    Set $E \leftarrow \{e: \attrparent{e} = u \mbox{ and } e \in \Et_I\}$.
-    Set $v \leftarrow -1$.
+    Set $E \leftarrow \{e \in \Et_I : \attrparent{e} = u \}$.
+    Set $v \leftarrow -1$.  If $|E| = 0$, go to \algref{S8}.
 }
@@
     if $\attrright{x} > \attrleft{e}$ and $\attrright{e} > \attrleft{x}$,
-    set $\ell \leftarrow \max(\attrleft{x}, \attrright{e})$,
+    set $\ell \leftarrow \max(\attrleft{x}, \attrleft{e})$,
     $r \leftarrow \min(\attrright{x}, \attrright{e})$,
@@
 
-\algstep{S5.}{New output node.}{
+\algstep{S5.}{Overlap; new output node.}{
     If $v = -1$, set $v \leftarrow \taddrow{\Nt_0}{\Nt_I[u]}$.
@@
 
-The algorithm begins in \algref{S1} by allocating an empty list to store ancestral
+The algorithm begins in \algref{S0} by allocating an empty list to store ancestral
 segments for each of the $|\Nt_I|$ input nodes, and then creating the initial
@@
 priority queue $Q$, and set the current input ID $u$ to $0$.
-
-The algorithm steps back up through the pedigree and steps \algref{S1}--\algref{S8}
+Recall that the function $\taddrow{\Nt}{t}$ adds a new node to the node table $\Nt$
+with birth time $t$, and returns the ID (i.e., index) of that new node in $\Nt$.
+So, the operation $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ looks up the birth time
+for input node ID $u$ in $\Nt_I$, adds a new node with this time to $\Nt_O$,
+and stores the resulting (output) node ID in $v$.
+
+To step back up through the pedigree, steps \algref{S1}--\algref{S8}
 consider each node $u$ in turn (note we are assuming that nodes are ordered by time-since-birth).
@@
 our current input ID $u$, and then set the corresponding output ID $v$ to $-1$.
-Then, in \algref{S2} we find all intersections between these edges and the
-mapped ancestry segments for their child nodes. For each of these intersections we
+Then, in \algref{S2} we find all intersections between these input edges and the
+mapped ancestry segments of their child nodes.
+For each of these intersections we
 create a new segment $y$ and insert it into the priority queue.
+(In the paint pot analogy,
+this stores in $Q$ all segments of color that must be transferred to the ancestor
+with input ID $u$.)
 
@@
 loop over the segments in the queue, considering each in turn until the queue is empty.
-Step \algref{S3} builds a list $X$ of all the segments that have the
-minimum left coordinate (along with some bookkeeping to keep track of where our next
+Step \algref{S3} builds a list $X$ of all segments covering the current left-most coordinate
+(along with some bookkeeping to keep track of where our next
 right coordinate $r$ is), removing them from $Q$. If there is only one element in $X$
@@
 
-As no overlaps occur, there is no change in the mapping between input and output IDs
-and so the ancestry segment $\alpha$ is not changed (this is assigned to $\Al_u$ in \algref{S7}).
-However, there is a complication caused by other segments which may intersect with $x$,
-which have left coordinate greater then $\attrleft{x}$. If $x$ intersects with this value,
-then there are other overlaps we must consider. Therefore, we change the left coordinate
-of $x$ to this coordinate and reinsert it into $Q$ where it can be processed again
-later. After doing this, we proceed to \algref{S7}, update the ancestry mapping for
+In this case, only a single segment covers the leftmost region of the segments,
+so there is no change in the mapping between input and output IDs,
+and the ancestry segment $\alpha$ is not changed (this is assigned to $\Al_u$ in \algref{S7}).
+(The color in this segment of the offspring can simply be transferred to the parent.)
+However, if other segments in $Q$ intersect with $x$
+(since $x$ is the current left-most segment, these have left coordinate less then $\attrright{x}$),
+then we can only transfer the portion of $x$ up until it overlaps with other segments.
+Therefore, we change the left coordinate
+of $x$ to this coordinate and reinsert it into $Q$ where it will be processed again
+later, effectively cutting off (and propagating) the leftmost segment of $x$.
+After doing this, we proceed to \algref{S7}, update the ancestry mapping for
 the input node $u$, and loop back to \algref{S3} if $Q$ is not empty.
@@
 On the other hand, if there is more than one segment in $X$, we know that an
-overlap among the ancestral segments has occured and we must update the output
+overlap among the ancestral segments has occurred and we must update the output
 nodes and edges accordingly. In step \algref{S5} we first allocate a new output
-node $v$ for $u$, if it has not been done already (we may have many overlapping
+node $v$ for input node $u$, if it has not been done already (we may have many overlapping
 segments along the genome). Step \algref{S6} then continues by creating a new

### 2018-01-12 sims/simplify-alg.py
more comments in python

@@ -27,6 +27,10 @@ class Segment(object):
         return repr((self.left, self.right, self.id))
 
 def simplify(S, Ni, Ei, L):
+    '''
+    This is an implementation of the simplify algorithm described in Appendix A
+    of the paper.
+    '''
     No = msprime.NodeTable()
     Eo = msprime.EdgeTable()
     A = [[] for _ in range(len(Ni))]
@@ -99,6 +103,10 @@ def simplify(S, Ni, Ei, L):
 
 
 def simplify_interval_tree(S, Ni, Ei, L):
+    '''
+    This is an alternative implementation of the simplify algorithm, relying on
+    operations on collections of intervals provided by IntervalTree objects.
+    '''
     No = msprime.NodeTable()
     Eo = msprime.EdgeTable()
     left = Ei.left
@@ -177,6 +185,10 @@ class Edge(object):
 
 
 def simplify_loci(S, Ni, Ei, L):
+    '''
+    This is an implementation of the simplify algorithm that works with a discrete set
+    of loci, rather than continuous intervals.
+    '''
     No = msprime.NodeTable()
     Eo = msprime.EdgeTable()
     left = Ei.left.astype(int)
@@ -241,6 +253,8 @@ def simplify_loci(S, Ni, Ei, L):
     return msprime.load_tables(nodes=No, edges=Eo)
 
 
+# Checks that simplify() does the right thing, by comparing to the implementation
+# in msprime.
 
 def verify():
     for n in [10, 100, 1000]:

### 2018-01-12 forwards_paper.tex
minor fixen

@@
 For each input node, we need to process the ancestral segments
-to find overlaps (and hence coalescences), and to update the $\Al$.
+to find overlaps (and hence coalescences), and to update $\Al$.
 We do this using a $\priorityq{}$ type, which maintains a list of ancestral
@@
 
-\algstep{S0.}{Initialisation.}{
+\algstep{S1.}{Initialisation.}{
     Set $\Al_u \leftarrow \listnew{}$ for $0 \leq u < |\Nt_I|$. Then,
@@
 
-\algstep{S1.}{Select parent edges.}{
+\algstep{S2.}{Select parent edges.}{
     Set $E \leftarrow \{e \in \Et_I : \attrparent{e} = u \}$.
-    Set $v \leftarrow -1$.  If $|E| = 0$, go to \algref{S8}.
-}
-
-\algstep{S2.}{Insert edge/ancestry intersections.}{
+    Set $v \leftarrow -1$.  If $|E| = 0$, go to \algref{S9}.
+}
+
+\algstep{S3.}{Insert edge/ancestry intersections.}{
     For each $e\in E$ and each $x \in \Al[\attrchild{e}]$,
@@
 
-\algstep{S3.}{Find segments with minimum left coordinate.}{
+\algstep{S4.}{Find segments with minimum left coordinate.}{
     Set $\ell \leftarrow \attrleft{Q[0]}$, $r \leftarrow L$ and $X \leftarrow \mbox{List}()$.
@@
     Afterwards, if $|Q| > 0$, set $r \leftarrow \min(r, \attrleft{Q[0]})$. If
-    $|X| > 1$ go to \algref{S5}.
-}
-
-\algstep{S4.}{No overlapping segments.}{
+    $|X| > 1$ go to \algref{S6}.
+}
+
+\algstep{S5.}{No overlapping segments.}{
     Set $x \leftarrow X[0]$ and $\alpha \leftarrow x$. If $|Q| > 0$ and
@@
     $\attrleft{x} \leftarrow \attrleft{Q[0]}$ and call $\pqpush{Q}{x}$.
-    Go to \algref{S7}.
-}
-
-\algstep{S5.}{Overlap; new output node.}{
+    Go to \algref{S8}.
+}
+
+\algstep{S6.}{Overlap; new output node.}{
     If $v = -1$, set $v \leftarrow \taddrow{\Nt_0}{\Nt_I[u]}$.
@@
 
-\algstep{S6.}{Record edges.}{
+\algstep{S7.}{Record edges.}{
     Set $\alpha \leftarrow \ancsegment{\ell, r, v}$. Then for each $x \in X$
@@
 
-\algstep{S7.}{Left coordinate loop.}{
-    Call $\listappend{\Al[u]}{\alpha}$. If $|Q| > 0$ return to \algref{S3}.
-}
-
-\algstep{S8.}{Parent loop.}{
-    Set $u \leftarrow u + 1$. If $u < |\Nt_I|$, go to \algref{S1}.
+\algstep{S8.}{Left coordinate loop.}{
+    Call $\listappend{\Al[u]}{\alpha}$. If $|Q| > 0$ return to \algref{S4}.
+}
+
+\algstep{S9.}{Parent loop.}{
+    Set $u \leftarrow u + 1$. If $u < |\Nt_I|$, go to \algref{S2}.
 }
@@
 
-The algorithm begins in \algref{S0} by allocating an empty list to store ancestral
+The algorithm begins in \algref{S1} by allocating an empty list to store ancestral
 segments for each of the $|\Nt_I|$ input nodes, and then creating the initial
@@
 
-To step back up through the pedigree, steps \algref{S1}--\algref{S8}
+To process ancestral information moving back up through the pedigree, steps \algref{S2}--\algref{S9}
 consider each node $u$ in turn (note we are assuming that nodes are ordered by time-since-birth).
-In \algref{S1} we first find all edges in the input where the parent is equal to
+In \algref{S2} we first find all edges in the input where the parent is equal to
 our current input ID $u$, and then set the corresponding output ID $v$ to $-1$.
-Then, in \algref{S2} we find all intersections between these input edges and the
+Then, in \algref{S3} we find all intersections between these input edges and the
 mapped ancestry segments of their child nodes.
@@
 After finding all overlaps and filling the priority queue with the resulting ancestry
-segments, we then need to process these segments. Steps \algref{S3}--\algref{S7}
+segments, we then need to process these segments. Steps \algref{S4}--\algref{S8}
 loop over the segments in the queue, considering each in turn until the queue is empty.
-Step \algref{S3} builds a list $X$ of all segments covering the current left-most coordinate
+Step \algref{S4} builds a list $X$ of all segments covering the current left-most coordinate
 (along with some bookkeeping to keep track of where our next
@@
 this means that we have no overlapping ancestral segments at this coordinate and
-we proceed to \algref{S4}.
+we proceed to \algref{S5}.
 
@@
 so there is no change in the mapping between input and output IDs,
-and the ancestry segment $\alpha$ is not changed (this is assigned to $\Al_u$ in \algref{S7}).
+and the ancestry segment $\alpha$ is not changed (this is assigned to $\Al_u$ in \algref{S8}).
 (The color in this segment of the offspring can simply be transferred to the parent.)
@@
 later, effectively cutting off (and propagating) the leftmost segment of $x$.
-After doing this, we proceed to \algref{S7}, update the ancestry mapping for
-the input node $u$, and loop back to \algref{S3} if $Q$ is not empty.
+After doing this, we proceed to \algref{S8}, update the ancestry mapping for
+the input node $u$, and loop back to \algref{S4} if $Q$ is not empty.
 
@@
 overlap among the ancestral segments has occurred and we must update the output
-nodes and edges accordingly. In step \algref{S5} we first allocate a new output
+nodes and edges accordingly. In step \algref{S6} we first allocate a new output
 node $v$ for input node $u$, if it has not been done already (we may have many overlapping
-segments along the genome). Step \algref{S6} then continues by creating a new
+segments along the genome). Step \algref{S7} then continues by creating a new
 ancestry segment $\alpha$ mapping the current interval to the output node $v$.

### 2018-01-12 forwards_paper.tex
added log n

@@
 sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
-can do this in roughly $O(n + m + t)$ operations, where
+can do this in roughly $O(n + m + t \log n)$ operations, where
 $t$ is the number of distinct trees.

### 2018-01-12 forwards_paper.tex
rearrangement

@@
 it does not matter if we generate each generation's mutations during the simulation, or afterwards.
-Precisely, we need to know the genealogical trees relating all sampled individuals
+To add mutations after the fact, we need to know the genealogical trees relating all sampled individuals
 at each position along the genome.
@@
 at each position on the genome.
+
+The idea of storing genealogical information to speed up simulations is not new.
+It was implemented completely in AnA-FiTS~\citep{aberer2013rapid},
+but without the critical step of discarding irrelevant genealogical information.
+\citet{padhukasahasram2008exploring} obtained impressive speedups for a Wright--Fisher simulation
+by keeping track of genealogies over the preceeding 8 generations
+and only tracking neutral genotypes for those segments having descendants across this window.
+Our approach is similar, but uses genealogies across the entire duration of the simulation.
+The embellished pedigree is equivalent to the \emph{ancestral recombination graph},
+or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
+which has been the subject of substantial study
+% under the assumptions of coalescent theory
+\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
+However, it is unclear if an ARG-based approach would share
+the computational advantages of the data structures we use here \citet{kelleher2016efficient}.
+
+% However, the properties of the ARG as a computational structure have not
+% been studied and, despite several efforts to standardise a common
+% format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper
+% ARGs are rarely used in practise.
+% In contrast, the algorithmic properties of tree sequence
+% algorithms have been explored in detail~\citep{kelleher2016efficient},
+% contributing substantially to the efficiency of the \msprime{} coalescent simulator.
 
@@
 
-We now explain what we actually did to achieve a $50\times$ speedup.
+We now explain what we actually did to achieve this $50\times$ speedup.
 The ``pedigree recording'' simulations above recorded information about each new individual
@@
 and generalised to include a more general class of tree topologies.
-
-The idea of a tree sequence is closely related to the \emph{ancestral recombination graph},
-or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
-which also describes the embellished pedigree.
-The ARG has been the subject of substantial study
-under the assumptions of coalescent
-theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
-However, the properties of the ARG as a computational structure have not
-been studied and, despite several efforts to standardise a common
-format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper
-ARGs are rarely used in practise.
-In contrast, the algorithmic properties of tree sequence
-algorithms have been explored in detail~\citep{kelleher2016efficient},
-contributing substantially to the efficiency of the \msprime{} coalescent simulator.
 
@@
 
-The idea of storing genealogical information to speed up simulations is not new.
-It was implemented completely in AnA-FiTS~\citep{aberer2013rapid},
-but without the critical step of discarding irrelevant genealogical information.
-A similar but much more limited method for
-discarding this information also appears in \citet{padhukasahasram2008exploring}.
-
 The tree sequences produced by default by this method
@@
 
-Simulating very large populations and entire genomes will likely require parallelization.
-The one-way nature of information flow here makes our scheme relatively easy to incorporate
-into a parallel algorithm.
-Recently, \citet{lawrie2017accelerating} used parallelization
-to improve performance for the ``Poisson Random Field'' family of models~\citep{Sawyer1992-jw},
-in which mutations are unlinked and there is no effect of genetic background on fitness.
-However, thus far little attention has been paid to parallelization of more complex simulations.
-
-\paragraph{Bayesian phylogenetics}
+% Simulating very large populations and entire genomes will likely require parallelization.
+% The one-way nature of information flow here makes our scheme relatively easy to incorporate
+% into a parallel algorithm.
+% Recently, \citet{lawrie2017accelerating} used parallelization
+% to improve performance for the ``Poisson Random Field'' family of models~\citep{Sawyer1992-jw},
+% in which mutations are unlinked and there is no effect of genetic background on fitness.
+% However, thus far little attention has been paid to parallelization of more complex simulations.
+
+\paragraph{Other applications}
 The methods described here for efficiently storing tree sequences may prove useful in other fields.
@@
 sequence of trees which differ by subtree-prune-and-regraft
-(SPR) operations. Since each SPR operation requires a constant amount of space to encode, the total
+operations. Since each such operation requires a constant amount of space to encode, the total
 space required is $O(n + t)$ for $t$ trees with $n$ leaves~\citep{kelleher2016efficient}.

### 2018-01-12 forwards_paper.tex
edits for specific comments

--- 2018-01-12 forwards_paper.tex
@@
 %%%%%%%%%%%%
-\subsubsection*{Estimates of run-time complexity}
-
-The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
-pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
-near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).  In this section, we explore the run-time complexity using a
-simplified Wright-Fisher model with recombination.
+\subsubsection*{Computational complexity}
+
+% I may need to cite my student's t-shirt for the following line.
+Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that this method
+can dramatically improve simulation performance in practice --
+but, how does it perform in theory?
+% The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
+% pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
+% near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).  
+In this section, we explore the how required computational effort and storage space scale
+in a Wright-Fisher model with recombination.
 
@@
 %% Progressive memory reduction as a function of T
-To get an idea of how required space depends on the length of time between simplification steps,
-suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
+\citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
+by retaining only 8 generations of history.
+This begs the question: how fast does the information in the pedigree become irrelevant?
+Suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
 begun with no prior history.
+One way to answer this question is to ask:
 How many of the resulting $4NT$ edges are required after simplification?
@@
 Since mutations fall as recombinations do on the marginal trees,
-adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations,
+the number of mutations is proportional to the total length of all the marginal trees,
+and so adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations,
 and the same savings as we obtained from edges above,
@@
 computational burden of propagating mutations forwards across generations.
+
+Since each mutation is stored only as a single row in the mutation table,
+and at most one row in the site table, the space required for $M$ mutations is $O(M)$;
+combined with the $O(N \log N)$ storage for edges and nodes of a tree sequence above,
+this implies that the full results of a simulation of $N$ individuals
+having $M$ mutations can be stored in $O(N \log N + M)$ space.
+In simulations of whole chromosomes, the number of mutations per chromosome per generation is $O(1)$,
+so $M$ will typically be $O(N \log N)$ as well.
 

### 2018-01-12 forwards_paper.tex
added details to complexity section

@@
 of only the current generation of $2N$ individuals.
-Ignoring the possibility of coalescent events involving more than two lineages, this tree sequence has $4N-2$ edges to describe the leftmost tree;
+If no coalescent events involve more than two lineages, this tree sequence has $4N-2$ edges to describe the leftmost tree;
 and each time the marginal tree changes along the sequence,
@@
 Coalescent theory tells us that
-the expected total length of the edges of a marginal tree is approximately $4N\log(2N)$,
-which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
-since we have taken one crossover per individual per generation.
+the expected total length of the edges of a marginal tree is approximately $4N\log(2N)$.
+Since we have taken one crossover per individual per generation (as is typical for a single chromosome),
+and we are measuring tree lengths in units of generations,
+the mean number of ancestral recombination events that occur somewhere on the marginal trees
+per unit of genetic length.
+Any such ancestral recombination might change the marginal gene tree;
+if it does, it will add four new edges.
 Not all such recombinations actually change the tree topology,
-so four times this gives us an upper bound on the expected number of additional edges.
+so at most we expect $16N\log(2N)$ new edges after those required to describe the first tree.
 Similarly, not every new edge derives from a never-before-seen node,
 but the number of nodes is at most equal to the number of edges plus the sample size.
-With $T=20N$, this reduces the initial storage of $120 N^2$ items to
+This reduces the initial storage of $120 N^2$ items to
 % 4N - 2 + 2N + 4N log(2N) * 4 * 2 = 6N + 32 N log(2N)
-$6N + 32 N \log(2N)$;
-at $N=10^4$, a factor of 3,700.
+the space required after simplification of roughly
+\begin{align}
+    6N + 32 N \log(2N) .
+\end{align}
+At $N=10^4$, this reduces the space required to store the history of the entire population
+by a factor of 3,700.
 
@@
 If we were to add mutations in forwards time
-under the infinite-sites model with total mutation rate per generation $\mu$,
-there would also be around $\mu 2NT$ mutations (and the same number of sites).
+under the infinite-sites model with total mutation rate per generation $\mu$ (i.e., per meiosis),
+there would also be around $\mu 2NT$ mutations, and the same number of sites,
+simply because there was a total of $2NT$ meioses.
 Since mutations fall as recombinations do on the marginal trees,
 the number of mutations is proportional to the total length of all the marginal trees,
-and so adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations,
-and the same savings as we obtained from edges above,
+and so by the same argument as for the number of edges,
+adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations.
+With $N=10^4$ and $T=20N$ as above,
+this implies that adding neutral mutations afterwards reduces the number of mutations that must be generated
+by a factor of 10,000.
+This could result in substantial time savings,
 even without considering the
@@
 and at most one row in the site table, the space required for $M$ mutations is $O(M)$;
-combined with the $O(N \log N)$ storage for edges and nodes of a tree sequence above,
+combined with the $O(N \log N)$ storage for edges and nodes of a simplified tree sequence above,
 this implies that the full results of a simulation of $N$ individuals
@@
 as seen in Figure~\ref{fig:simplify_complexity}B,
-where we show shows the time required to simplify increasingly large subsets of a large tree sequence.
+which shows the time required to simplify increasingly large subsets of a large tree sequence.
 When simplifying the result of a forwards-time sequence, the number of edges is the main contributing factor.

### 2018-01-12 sims/simplify-alg.py
removed other python simplify implementations

@@ -100,159 +100,6 @@ def simplify(S, Ni, Ei, L):
 
 
 
-
-
-def simplify_interval_tree(S, Ni, Ei, L):
-    '''
-    This is an alternative implementation of the simplify algorithm, relying on
-    operations on collections of intervals provided by IntervalTree objects.
-    '''
-    No = msprime.NodeTable()
-    Eo = msprime.EdgeTable()
-    left = Ei.left
-    right = Ei.right
-    child = Ei.child
-    parent = Ei.parent
-    time = Ni.time
-    # TODO Should be an interval tree for each ID.
-    A = [intervaltree.IntervalTree() for _ in range(len(Ni))]
-    for input_id in S:
-        output_id = No.add_row(time=time[input_id], flags=1)
-        A[output_id] = intervaltree.IntervalTree([intervaltree.Interval(0, L, input_id)])
-
-    E = []
-    for input_parent in range(len(Ni)):
-        index = parent == input_parent
-        new_output_id = -1
-        P = intervaltree.IntervalTree()
-        for l, r, c in zip(left[index], right[index], child[index]):
-            # We have to slice the intervals to make sure we only return the bits
-            # we're interested in.
-            A[c].slice(l)
-            A[c].slice(r)
-            # Note there's no need to actually remove the ancestral material. We just
-            # leave it in there because it can't affect anything later.
-            P = P.union(A[c][l: r])
-        P.split_overlaps()
-        for l, r, output_id in  P:
-            # We look at each distinct segment in turn. If there are coalescences, there
-            # will be segments that overlap each other which we detect by the effects
-            # on A[input_parent]
-            A[input_parent].slice(l)
-            A[input_parent].slice(r)
-            if len(A[input_parent][l: r]) == 0:
-                A[input_parent][l: r] = output_id
-            else:
-                if new_output_id == -1:
-                    new_output_id = No.add_row(time=time[input_parent], flags=0)
-                child_output_id = list(A[input_parent][l:r])[0].data
-                if child_output_id != new_output_id:
-                    E.append(Edge(l, r, new_output_id, child_output_id))
-                E.append(Edge(l, r, new_output_id, output_id))
-                A[input_parent].remove_overlap(l, r)
-                A[input_parent][l: r] = new_output_id
-
-    # Sort the output locus-wise edges and compact them as much as possible into
[…]

### 2018-01-12 forwards_paper.tex
removed other simplify implementations and added python code to the end of the paper

--- 2018-01-12 forwards_paper.tex
@@
 paralleling the implementation in \msprime{} \citep{kelleher2016efficient};
-an implementation in pseudocode is provided in Appendix~\ref{ss:simplify_algorithm}.
+an implementation in pseudocode is provided in Appendix~\ref{ss:simplify_algorithm},
+and a python implementation as supplementary information.
 Conceptually, this works by
@@
 %%%%%%%%%%%%
-\subsubsection*{Computational complexity}
-
-% I may need to cite my student's t-shirt for the following line.
-Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that this method
-can dramatically improve simulation performance in practice --
-but, how does it perform in theory?
-% The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
-% pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
-% near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).  
-In this section, we explore the how required computational effort and storage space scale
-in a Wright-Fisher model with recombination.
+\subsubsection*{Estimates of run-time complexity}
+
+The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
+pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
+near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).  In this section, we explore the run-time complexity using a
+simplified Wright-Fisher model with recombination.
 
@@
 %% Progressive memory reduction as a function of T
-\citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
-by retaining only 8 generations of history.
-This begs the question: how fast does the information in the pedigree become irrelevant?
-Suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
+To get an idea of how required space depends on the length of time between simplification steps,
+suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
 begun with no prior history.
-One way to answer this question is to ask:
 How many of the resulting $4NT$ edges are required after simplification?
@@
 Since mutations fall as recombinations do on the marginal trees,
-the number of mutations is proportional to the total length of all the marginal trees,
-and so adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations,
+adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations,
 and the same savings as we obtained from edges above,
@@
 computational burden of propagating mutations forwards across generations.
-
-Since each mutation is stored only as a single row in the mutation table,
-and at most one row in the site table, the space required for $M$ mutations is $O(M)$;
-combined with the $O(N \log N)$ storage for edges and nodes of a tree sequence above,
-this implies that the full results of a simulation of $N$ individuals
-having $M$ mutations can be stored in $O(N \log N + M)$ space.
-In simulations of whole chromosomes, the number of mutations per chromosome per generation is $O(1)$,
-so $M$ will typically be $O(N \log N)$ as well.
 
@@
 
+\newpage
+
+\section*{Python implementation of simplify (supplementary information)}
+
+\lstinputlisting[language=python]{sims/simplify-alg.py}
+
 \end{document}

### 2018-01-13 sims/simplify-alg.py
Cleaned up the Python implementation of simplify.

@@ -1,50 +1,56 @@
 """
-Set-theoretic algorithm implementation of Algorithm S.
+Python implementation of Algorithm S.
 
 The example works by generating an initial TreeSequence
 for a sample of 10 haplotypes using msprime.  We then
 simplify the node/edge table in that TreeSequence with
 respect to the first three samples.
 """
-import numpy as np
 import heapq
+import numpy as np
 
 import msprime
-import intervaltree
 
 
 class Segment(object):
-    def __init__(self, left, right, id_):
+    """
+    An ancestral segment mapping a given node ID to a half-open genomic
+    interval [left, right).
+    """
+    def __init__(self, left, right, node):
         assert left < right
         self.left = left
         self.right = right
-        self.id = id_
+        self.node = node
 
     def __lt__(self, other):
-        return (self.left, self.right, self.id) < (other.left, other.right, other.id)
+        # We implement the < operator so that we can use the heapq directly.
+        # We are only interested in the left coordinate.
+        return self.left < other.left
 
     def __repr__(self):
-        return repr((self.left, self.right, self.id))
+        return repr((self.left, self.right, self.node))
+
 
 def simplify(S, Ni, Ei, L):
-    '''
+    """
     This is an implementation of the simplify algorithm described in Appendix A
     of the paper.
-    '''
+    """
     No = msprime.NodeTable()
     Eo = msprime.EdgeTable()
     A = [[] for _ in range(len(Ni))]
     Q = []
 
     for u in S:
-        output_id = No.add_row(time=Ni.time[u], flags=1)
-        A[u] = [Segment(0, L, output_id)]
+        v = No.add_row(time=Ni.time[u], flags=1)
+        A[u] = [Segment(0, L, v)]
 
     for u in range(len(Ni)):
         for e in [e for e in Ei if e.parent == u]:
             for x in A[e.child]:
                 if x.right > e.left and e.right > x.left:
-                    y = Segment(max(x.left, e.left), min(x.right, e.right), x.id)
+                    y = Segment(max(x.left, e.left), min(x.right, e.right), x.node)
                     heapq.heappush(Q, y)
         v = -1
         while len(Q) > 0:
@@ -62,7 +68,7 @@ def simplify(S, Ni, Ei, L):
                 x = X[0]
                 alpha = x
                 if len(Q) > 0 and Q[0].left < x.right:
[…]

### 2018-01-13 forwards_paper.tex
Cleaned up the Python implementation of simplify.
Minor updates to the calculations.

@@
 The idea of storing genealogical information to speed up simulations is not new.
-It was implemented completely in AnA-FiTS~\citep{aberer2013rapid},
+It was implemented in AnA-FiTS~\citep{aberer2013rapid},
 but without the critical step of discarding irrelevant genealogical information.
@@
 However, it is unclear if an ARG-based approach would share
-the computational advantages of the data structures we use here \citet{kelleher2016efficient}.
+the computational advantages of the data structures we use here~\citep{kelleher2016efficient}.
 
@@
 % near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).
-In this section, we explore the how required computational effort and storage space scale
+In this section, we explore how the required computational effort and storage space scale
 in a Wright-Fisher model with recombination.
@@
 %% Memory reduction thanks to simplification
-Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
+Consider a simulation of a Wright-Fisher population of $N$ haploid individuals
 using Algorithm~\algref{W} for $T$ generations,
@@
 without simplification this would produce tables of
-$2NT$ nodes and
-$4NT$ edges.
-Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
-($T = 20N$, say).
+$NT$ nodes and
+$2NT$ edges.
+Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability
+($T = 20N$, say, giving a total space requirement of $60N^2$).
 After simplification, we are left with the tree sequence describing the history
-of only the current generation of $2N$ individuals.
-If no coalescent events involve more than two lineages, this tree sequence has $4N-2$ edges to describe the leftmost tree;
+of only the current generation of $N$ individuals.
+If no coalescent events involve more than two lineages, this tree sequence has $2N-2$ edges to describe the leftmost tree;
 and each time the marginal tree changes along the sequence,
@@
 Coalescent theory tells us that
-the expected total length of the edges of a marginal tree is approximately $4N\log(2N)$.
+the expected total length of the edges of a marginal tree is approximately $2N\log(N)$.
 Since we have taken one crossover per individual per generation (as is typical for a single chromosome),
 and we are measuring tree lengths in units of generations,
-the mean number of ancestral recombination events that occur somewhere on the marginal trees
-per unit of genetic length.
+this is the mean number of ancestral recombination events that occur somewhere on the marginal trees.
 Any such ancestral recombination might change the marginal gene tree;
@@
 Not all such recombinations actually change the tree topology,
-so at most we expect $16N\log(2N)$ new edges after those required to describe the first tree.
+so at most we expect $8N\log(N)$ new edges after those required to describe the first tree.
 Similarly, not every new edge derives from a never-before-seen node,
 but the number of nodes is at most equal to the number of edges plus the sample size.
-This reduces the initial storage of $120 N^2$ items to
-% 4N - 2 + 2N + 4N log(2N) * 4 * 2 = 6N + 32 N log(2N)
-the space required after simplification of roughly
-\begin{align}
-    6N + 32 N \log(2N) .
-\end{align}
-At $N=10^4$, this reduces the space required to store the history of the entire population
-by a factor of 3,700.
+The space complexity after simplification is therefore $O(N + N\log N)$, since
+we need $O(N)$ space to store the first marginal tree, and $O(1)$ space
+for each of the $O(N \log N)$ trees that follow. Thus, we need $O(N^2)$ space
+to store the complete history of the simulation, but only $O(N + N \log N)$
+space to store the history that is relevant to the extant population.
 

### 2018-01-13 sims/simplify-alg.py
delete import of intervaltree, as it is not used.  May cause import error for no reason.

@@ -10,7 +10,6 @@ import numpy as np
 import heapq
 
 import msprime
-import intervaltree
 
 
 class Segment(object):

### 2018-01-14 forwards_paper.tex

--- 2018-01-12 forwards_paper.tex
@@
 
+In this article, we applied our methods for storing trees to the problem of pedigree recording in a forward-time
+simulation.  However, the method applies to any simulation scheme generating nodes and edges.  For example, one could
+use the methods described here to generate succinct tree sequences under coalescent processes not currently implemented
+in \msprime{}, such as the coalescent with gene conversion \citep{Wiuf2000-rc} or using the structured coalescent to
+model various forms of natural selection \citep{Kaplan1988-in,Kaplan1989-rt,Braverman1995-gn}.  For such models, one
+could in principle generate tables of nodes and edges to be simpified in \msprime{}.  The resulting succinct tree
+sequence object would be in the same format as those generated by \msprime{}'s built-in simulate function, and therefore
+compatible with exising methods for downstream analyses.
+
 % JK: This needs more context. Probably not worth putting in here.

### 2018-01-14 forwards_paper.tex
more explanation in complexity section

@@
 in a Wright-Fisher model with recombination.
+Each edge in a tree sequence represents a segment of genome
+that has been inherited across a certain number of generations, given by the length of the edge.
+Each inheritance of each segment of genome across a generation
+provides an opportunity for mutation and for recombination (which changes the tree).
+For this reason, the key quantity we use is the total ``area'' of the tree sequence,
+i.e., the product of segment length (left minus right coordinates) 
+and edge length (difference in birth times between parent and child),
+summed across all edges.
+This area is also equal to the sum of the total lengths of all marginal trees
+(i.e., the trees describing inheritance at each position on the genome),
+so we can compute total area as sequence length
+multiplied by the mean area of a given marginal tree,
+which is well-described by coalescent theory.
+Similar arguments to those below go back to \citet{watterson1975number},
+who explicitly addresses smaller order corrections relevant to whole-population genealogies
+of the Wright-Fisher model.
+For more on coalescent theory, see \citet{wakeley2005coalescent}.
 
@@
 without simplification this would produce tables of
-$NT$ nodes and
-$2NT$ edges.
+$NT$ nodes and $2NT$ edges.
 Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability
@@
 The space complexity after simplification is therefore $O(N + N\log N)$, since
-we need $O(N)$ space to store the first marginal tree, and $O(1)$ space
+we need $O(N)$ space to store the first marginal tree, and a constant amount of space
 for each of the $O(N \log N)$ trees that follow. Thus, we need $O(N^2)$ space
-to store the complete history of the simulation, but only $O(N + N \log N)$
+to store the complete history of the simulation, but only $O(N \log N)$ % $O(N + N \log N)$
 space to store the history that is relevant to the extant population.
@@
 How many of the resulting $4NT$ edges are required after simplification?
-As above, the expected number of edges is bounded by $2N-2$ plus four times the mean marginal tree length.
-Roughly, the length of time for which a marginal tree has $k$ tips is
-$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations,
-and so the time to go from $N$ to $n$ tips is $2N(1/(n-1) - 1/(N-1))$.
-This implies that after running for time $T$ we expect to
-have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
+As above, the expected number of edges is bounded by $2N-2$
+plus four times the sum of the lengths of all marginal trees,
+because new edges appear with most recombinations that occur on any marginal tree.
+Therefore, we need to compute the mean total length of all branches in a coalescent tree
+up until time $T$ (or the common ancestor, whichever comes first).
+Since $k$ lineages coalesce at rate $\binom{k}{2}/N$,
+the expected length of time for which a marginal tree has $k$ branches is
+$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
+Since there are $k$ branches for this time,
+this period is expected to contribute $2N/(k-1)$ generations to the total tree length.
+By summing over $n < k \le N$,
+the $N$ tips of a tree are expected to descend from only $n$ lineages
+around $2N(1/(n-1) - 1/(N-1))$ generations ago.
+Inverting this relationship between time and number of roots
+implies that a marginal tree cut $T$ units of time ago
+is expected to have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
 The total tree length over this time is
@@
 having $M$ mutations can be stored in $O(N \log N + M)$ space.
-In simulations of whole chromosomes, the number of mutations per chromosome per generation is $O(1)$,
-so $M$ will typically be $O(N \log N)$ as well.
+In simulations of whole chromosomes, there are typically a small, bounded number of mutations per chromosome per generation,
+so $M$ will be $O(N \log N)$ as well.
 
@@
 use the methods described here to generate succinct tree sequences under coalescent processes not currently implemented
-in \msprime{}, such as the coalescent with gene conversion \citep{Wiuf2000-rc} or using the structured coalescent to
-model various forms of natural selection \citep{Kaplan1988-in,Kaplan1989-rt,Braverman1995-gn}.  For such models, one
+in \msprime{}, such as the coalescent with gene conversion \citep{Wiuf2000-rc}, using the structured coalescent to
+model various forms of natural selection \citep{Kaplan1988-in,Kaplan1989-rt,Braverman1995-gn},
+or the coalescent within a known pedigree.  For such models, one
 could in principle generate tables of nodes and edges to be simpified in \msprime{}.  The resulting succinct tree

### 2018-01-14 forwards_paper.tex
factors of 2

@@
 This begs the question: how fast does the information in the pedigree become irrelevant?
-Suppose instead that we have run a simulation of $2N$ individuals for $T$ generations,
+Suppose instead that we have run a simulation of $N$ individuals for $T$ generations,
 begun with no prior history.
 One way to answer this question is to ask:
-How many of the resulting $4NT$ edges are required after simplification?
-As above, the expected number of edges is bounded by $2N-2$
+How many of the resulting $2NT$ edges are required after simplification?
+As above, the expected number of edges is bounded by $N-2$
 plus four times the sum of the lengths of all marginal trees,
@@
 the expected length of time for which a marginal tree has $k$ branches is
-$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
+$N/(k(k-1)) = N(1/(k-1) - 1/k)$ generations.
 Since there are $k$ branches for this time,
-this period is expected to contribute $2N/(k-1)$ generations to the total tree length.
+this period is expected to contribute $N/(k-1)$ generations to the total tree length.
 By summing over $n < k \le N$,
 the $N$ tips of a tree are expected to descend from only $n$ lineages
-around $2N(1/(n-1) - 1/(N-1))$ generations ago.
+around $N(1/(n-1) - 1/(N-1))$ generations ago.
 Inverting this relationship between time and number of roots
 implies that a marginal tree cut $T$ units of time ago
-is expected to have around $n(T)$ roots, where $n(T) \approx 1 + 2N/(T+2)$.
+is expected to have around $n(T)$ roots, where $n(T) \approx 1 + N/(T+2)$.
 The total tree length over this time is
-$2 N \sum_{k=n(T)+1}^{N-1} 1/k$, which
+$N \sum_{k=n(T)+1}^{N-1} 1/k$, which
 % is approximately
 % \begin{align*}
-%     2 N \log\left( \frac{N}{1 + 2 N/(T+2)} \right) .
+%     N \log\left( \frac{N/2}{1 + N/(T+2)} \right) .
 % \end{align*}
@@
     \label{eqn:edge_bound}
-    2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
+    N \left( 1 + 4 \log\left( \frac{NT}{2(T + N)} \right)\right) .
 \end{align}
@@
         Number of edges in the simplified tree sequence
-        for 10 replicate Wright--Fisher simulations with $N=50$ as a function
+        for 10 replicate Wright--Fisher simulations with $N=100$ as a function
         of number of generations.
@@
 under the infinite-sites model with total mutation rate per generation $\mu$ (i.e., per meiosis),
-there would also be around $\mu 2NT$ mutations, and the same number of sites,
-simply because there was a total of $2NT$ meioses.
+there would also be around $\mu NT$ mutations, and the same number of sites,
+simply because there was a total of $NT$ meioses.
 Since mutations fall as recombinations do on the marginal trees,
@@
 and so by the same argument as for the number of edges,
-adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations.
-With $N=10^4$ and $T=20N$ as above,
+adding them after simplification results in only around $\mu 2 N \log(N)$ mutations.
+With $N=2 \times 10^4$ and $T=10N$ as above,
 this implies that adding neutral mutations afterwards reduces the number of mutations that must be generated

### 2018-01-14 forwards_paper.tex
para on known pedigress

--- 2018-01-14 forwards_paper.tex
@@
 
+Another application of our methods would be the case of simulating coalescent histories conditional on known pedigrees.
+The standard description of the Wright-Fisher coalescent averages over pedigrees.  However, conditional on a realized
+pedigree, the distribution of coalescent times in the recent past differs from that of the unconditional coalescent
+\citep{Wakeley2012-kw}.  For populations with known pedigrees \citep[\textit{e.g.}][]{Aguillon2017-ac}, it may be of use to
+simulate transmission along such pedigrees for the purpose of inference.
+
 % JK: This needs more context. Probably not worth putting in here.

### 2018-01-14 forwards_paper.tex
in progress

--- 2018-01-14 forwards_paper.tex
@@
 Since $k$ lineages coalesce at rate $\binom{k}{2}/N$,
-the expected length of time for which a marginal tree has $k$ branches is
-$N/(k(k-1)) = N(1/(k-1) - 1/k)$ generations.
-Since there are $k$ branches for this time,
-this period is expected to contribute $N/(k-1)$ generations to the total tree length.
+the expected length of time for which the coalescent process has $k$ lineages is
+$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
+Since this is the time period over which the tree has $k$ branches,
+this period is expected to contribute $2N/(k-1)$ generations to the total tree length.
 By summing over $n < k \le N$,
 the $N$ tips of a tree are expected to descend from only $n$ lineages
-around $N(1/(n-1) - 1/(N-1))$ generations ago.
+around $2N(1/n - 1/N)$ generations ago.
 Inverting this relationship between time and number of roots
 implies that a marginal tree cut $T$ units of time ago
-is expected to have around $n(T)$ roots, where $n(T) \approx 1 + N/(T+2)$.
+% since 2N(1/n - 1/N) = 2(N/n-1) = T implies
+%    N/n = (T+2)/2 and so n = 2N/(T+2)
+is expected to have around $n(T)$ roots, where $n(T) \approx 2N/(T+2)$.
 The total tree length over this time is
-$N \sum_{k=n(T)+1}^{N-1} 1/k$, which
+$2N \sum_{k=N}^{n(T)+1} 1/(k-1)$, which
 % is approximately
 % \begin{align*}
-%     N \log\left( \frac{N/2}{1 + N/(T+2)} \right) .
+%     2 N \left\{ \log(N-1) - \log(n(T)) \} = 2 N \log( ((T+2)(N-1))/2N ) \approx 2 N \log( (T+2)/2 )
 % \end{align*}

### 2018-01-14 sims/space-complexity.py

@@ -63,7 +63,8 @@ def total_length(nodes, edges):
 def ub(T, N):
     # 2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
     # return 2 * (N - 1) + 8 * N * np.log(N/(1+2*N/(T+2)))
-    return 2 * N * (1 + 4 * np.log(N * T / (T + 2 * N)))
+    # return 2 * N * (1 + 4 * np.log(N * T / (T + 2 * N)))
+    return 2 * N * np.log( (T + 2) / 2 )
 
 
 def get_mean_edges_per_transition(ts):

### 2018-01-14 forwards_paper.tex

--- 2018-01-14 forwards_paper.tex
@@
 
-Another application of our methods would be the case of simulating coalescent histories conditional on known pedigrees.
-The standard description of the Wright-Fisher coalescent averages over pedigrees.  However, conditional on a realized
-pedigree, the distribution of coalescent times in the recent past differs from that of the unconditional coalescent
-\citep{Wakeley2012-kw}.  For populations with known pedigrees \citep[\textit{e.g.}][]{Aguillon2017-ac}, it may be of use to
-simulate transmission along such pedigrees for the purpose of inference.
-
 % JK: This needs more context. Probably not worth putting in here.
@@
 the Wellcome Trust (grant 100956/Z/13/Z to Gil McVean);
-and XXX.
+and NIH grant R01GM115564 to KRT.
 

### 2018-01-15 forwards_paper.tex
rewrite of coal stuff

--- 2018-01-14 forwards_paper.tex
@@
 
-% I may need to cite my student's t-shirt for the following line.
+% I may need to cite my student's t-shirt for the following line:
+%  "that's all very good in practice, but how does it hold up in theory?"
 Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that this method
@@
 % near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).
-In this section, we explore how the required computational effort and storage space scale
-in a Wright-Fisher model with recombination.
-Each edge in a tree sequence represents a segment of genome
-that has been inherited across a certain number of generations, given by the length of the edge.
-Each inheritance of each segment of genome across a generation
-provides an opportunity for mutation and for recombination (which changes the tree).
-For this reason, the key quantity we use is the total ``area'' of the tree sequence,
+Both computational time and storage space are depend mostly
+on the number of \emph{mutations} and \emph{edges} in the tree sequence.
+% Each edge represents a segment of genome
+% that has been inherited across a number of generations equal to the length of the edge.
+% Each of these generations provided
+% an opportunity for mutation and for recombination
+% (which changes the tree, thus usually introducing a new edge).
+% For this reason, 
+The key quantity to understand for this will be the total ``area'' of the tree sequence,
 i.e., the product of segment length (left minus right coordinates)
@@
 This area is also equal to the sum of the total lengths of all marginal trees
-(i.e., the trees describing inheritance at each position on the genome),
-so we can compute total area as sequence length
-multiplied by the mean area of a given marginal tree,
-which is well-described by coalescent theory.
-Similar arguments to those below go back to \citet{watterson1975number},
-who explicitly addresses smaller order corrections relevant to whole-population genealogies
-of the Wright-Fisher model.
-For more on coalescent theory, see \citet{wakeley2005coalescent}.
+(i.e., the trees describing inheritance at each position on the genome).
+Therefore, the total area of a tree sequence is equal to the sequence length
+multiplied by the mean area of a marginal tree.
+Since we analyze tree sequences arising from a Wright--Fisher model,
+statistical properties of a marginal tree are well-described by coalescent theory.
+Similar arguments to those below go back at least to \citet{watterson1975number},
+who also explicitly computed smaller order corrections relevant to whole-population genealogies
+of the Wright--Fisher model.
+For an introduction to coalescent theory, see \citet{wakeley2005coalescent}.
 
 %% Memory reduction thanks to simplification
-Consider a simulation of a Wright-Fisher population of $N$ haploid individuals
-using Algorithm~\algref{W} for $T$ generations,
-with simplify interval $s = 1$ so that redundant information is removed after every generation.
+First: how much memory do simplified tree sequences require?
+Consider a simulation of a Wright--Fisher population of $N$ haploid individuals
+using Algorithm~\algref{W} for $T$ generations.
 Since every chromosome inherits material from both parents,
@@
 $NT$ nodes and $2NT$ edges.
-Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability
-($T = 20N$, say, giving a total space requirement of $60N^2$).
 After simplification, we are left with the tree sequence describing the history
-of only the current generation of $N$ individuals.
-If no coalescent events involve more than two lineages, this tree sequence has $2N-2$ edges to describe the leftmost tree;
-and each time the marginal tree changes along the sequence,
+of only the current generation of $N$ individuals, back to either $T$ generations ago,
+or their common ancestor, whichever comes first.
+The tree sequence must store edges describing the leftmost marginal tree,
+which requires at most $N-2$ edges.
+Then, each time the marginal tree changes along the sequence,
 four edges end and four new edges begin
 (except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
-Coalescent theory tells us that
-the expected total length of the edges of a marginal tree is approximately $2N\log(N)$.
-Since we have taken one crossover per individual per generation (as is typical for a single chromosome),
-and we are measuring tree lengths in units of generations,
-this is the mean number of ancestral recombination events that occur somewhere on the marginal trees.
-Any such ancestral recombination might change the marginal gene tree;
-if it does, it will add four new edges.
-Not all such recombinations actually change the tree topology,
-so at most we expect $8N\log(N)$ new edges after those required to describe the first tree.
-Similarly, not every new edge derives from a never-before-seen node,
+Suppose that $T_x$ is the marginal tree describing how each of the present-day population
+descends from their most recent common ancestor
+at a particular genomic position $x$ on the chromosome,
+and write $|T_x|$ for the total length of the tree,
+i.e., the number of ancestors 
+from which some but not all of the present-day population has inherited at position $x$
+(including the population itself).
+For the tree at nearby position $x + dx$ to be different,
+there must have been a crossing-over between $x$ and $x + dx$
+in one of the $|T_x|$ meioses that gave birth to those individuals.
+(Recall that these ``individuals'' are haploid.)
+Since we measure distance along the genome
+so that length is equal to the expected number of crossing-overs per generation,
+the expected distance until the next crossing-over is $1/|T_x|$.
+If every crossing-over changed the marginal tree,
+then this would imply, for consistency, 
+that the expected total number of times that the marginal tree changes along the genome
+is equal to the total area of the tree sequence (as defined above).
+Not every such crossing-over changes the marginal tree, but most do,
+so the total area of the tree sequence,
+multiplied by four,
+gives an upper bound on the expected number of edges in the tree sequence 
+beyond those describing the leftmost tree.
+Since we are considering a chromosome of length 1,
+the expected total area is equal to the mean marginal tree length, as above.
+
+If $T$ is large enough that all samples coalesce within the simulation with high probability
+-- $T = 20N$, say --
+then coalescent theory tells us that
+the expected total length of the branches of a marginal tree back to the most recent common ancestor
+is approximately $2N\log(N)$.
+Therefore, the tree sequence describing the entire population
+is expected to need no more than $2N + 8N\log(N)$ edges.
+Not every new edge derives from a never-before-seen node,
 but the number of nodes is at most equal to the number of edges plus the sample size.
-The space complexity after simplification is therefore $O(N + N\log N)$, since
-we need $O(N)$ space to store the first marginal tree, and a constant amount of space
-for each of the $O(N \log N)$ trees that follow. Thus, we need $O(N^2)$ space
-to store the complete history of the simulation, but only $O(N \log N)$ % $O(N + N \log N)$
-space to store the history that is relevant to the extant population.
+Therefore, 
+we would need $O(N^2)$ space to store the complete history of the simulation, 
+but only $O(N \log N)$ % $O(N + N \log N)$
+to store the history that is relevant to the extant population.
 
 %% Progressive memory reduction as a function of T
-\citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
-by retaining only 8 generations of history.
-This begs the question: how fast does the information in the pedigree become irrelevant?
-Suppose instead that we have run a simulation of $N$ individuals for $T$ generations,
-begun with no prior history.
-One way to answer this question is to ask:
-How many of the resulting $2NT$ edges are required after simplification?
-As above, the expected number of edges is bounded by $N-2$
-plus four times the sum of the lengths of all marginal trees,
-because new edges appear with most recombinations that occur on any marginal tree.
-Therefore, we need to compute the mean total length of all branches in a coalescent tree
+What if $T$ is smaller: 
+how many of the resulting $2NT$ edges are required after simplification?
+In other words, how fast does the information in the pedigree become irrelevant?
+% \citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
+% by retaining only 8 generations of history.
+% This begs the question: how fast does the information in the pedigree become irrelevant?
+Now, we need to compute the expected total length of all branches in a coalescent tree
 up until time $T$ (or the common ancestor, whichever comes first).
-Since $k$ lineages coalesce at rate $\binom{k}{2}/N$,
-the expected length of time for which the coalescent process has $k$ lineages is
-$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
-Since this is the time period over which the tree has $k$ branches,
-this period is expected to contribute $2N/(k-1)$ generations to the total tree length.
+Again, coalescent theory tells us that
+the expected length of time for which a coalescent tree has $k$ lineages 
+is $2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
+Since the tree has $k$ branches over this period,
+it is expected to contribute $2N/(k-1)$ to the total tree length.
 By summing over $n < k \le N$,
 the $N$ tips of a tree are expected to descend from only $n$ lineages
-around $2N(1/n - 1/N)$ generations ago.
+around $2(N/n - 1)$ generations ago.
 Inverting this relationship between time and number of roots
@@
 %    N/n = (T+2)/2 and so n = 2N/(T+2)
-is expected to have around $n(T)$ roots, where $n(T) \approx 2N/(T+2)$.
+is expected to have around $r(T)$ roots, where $r(T) \approx 2N/(T+2)$.
 The total tree length over this time is
-$2N \sum_{k=N}^{n(T)+1} 1/(k-1)$, which
-% is approximately
+$2N \sum_{k=N}^{r(T)+1} 1/(k-1) \approx 2 N \log(N/r(T))$.
+This is a crude estimate for several reasons:
+first, we should not count the branch leading to the root of the tree (i.e., when $r(T)=1$),
+and second, this does not account for the discrete nature of the Wright--Fisher model.
+% which is approximately
 % \begin{align*}
-%     2 N \left\{ \log(N-1) - \log(n(T)) \} = 2 N \log( ((T+2)(N-1))/2N ) \approx 2 N \log( (T+2)/2 )
+%     2N - 2 + 8 N \left\{ \log(N-1) - \log(r(T)) \} 
+%        = 2N - 2 + 8 N \log( (T+2)(N-1)/2N ) 
+%        \approx 2 N (1 + 4 \log( (T+2)/2 ))
 % \end{align*}
-leads to an upper bound on the number of edges of
+Nonetheless, this leads as above to an upper bound on the number of edges of
 \begin{align}
     \label{eqn:edge_bound}
-    N \left( 1 + 4 \log\left( \frac{NT}{2(T + N)} \right)\right) .
+    2 N \left( 1 + 4 \log\left( \min\left(N, \frac{T+2}{2}\right) \right)\right) .
 \end{align}
-These calculations hold up quite well in practice,
+This implies that the number of edges required to store the last $T$ generations
+of history for the entire chromosome of a population of size $N$
+grows as $O(N \log T)$ -- proportionally to $N$ at first but rapidly tapering off.
+The bound holds up reasonably well in practice,
 as shown in Figure~\ref{fig:simplify_complexity}.
-% by comparison to
-% the number of edges actually required for 50 independent Wright--Fisher simulations,
-% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
-% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.
 
@@
 %% Above was edges; same holds for memory usage savings due to mutations
+What about mutations?
 If we were to add mutations in forwards time
-under the infinite-sites model with total mutation rate per generation $\mu$ (i.e., per meiosis),
-there would also be around $\mu NT$ mutations, and the same number of sites,
-simply because there was a total of $NT$ meioses.
+under the infinite-sites model with total mutation rate per generation $\mu$,
+there would also be around $\mu N T$ mutations, and the same number of sites.
+% simply because there was a total of $N T$ meioses.
 Since mutations fall as recombinations do on the marginal trees,
@@
 adding them after simplification results in only around $\mu 2 N \log(N)$ mutations.
-With $N=2 \times 10^4$ and $T=10N$ as above,
+With $N=2 \times 10^4$ and $T=10N$,
 this implies that adding neutral mutations afterwards reduces the number of mutations that must be generated

### 2018-01-15 sims/space-complexity.py

@@ -64,7 +64,8 @@ def ub(T, N):
     # 2 N \left( 1 + 4 \log\left( \frac{NT}{T + 2 N} \right)\right) .
     # return 2 * (N - 1) + 8 * N * np.log(N/(1+2*N/(T+2)))
     # return 2 * N * (1 + 4 * np.log(N * T / (T + 2 * N)))
-    return 2 * N * np.log( (T + 2) / 2 )
+    # return 2 * N * (1 + 4 * np.log(N * T / (2 * T + 2 * N)))
+    return 2 * N * (1 - (1 / N) + 4 * np.log(min(N,  (T + 2) / 2 )))
 
 
 def get_mean_edges_per_transition(ts):

### 2018-01-15 forwards_paper.tex
minor smoothing

@@
 The key quantity to understand for this will be the total ``area'' of the tree sequence,
-i.e., the product of segment length (left minus right coordinates)
+which is the sum of the lengths of all ancestral genomic segments that some, but not all,
+of the present population has inherited.
+It can be found by summing the product of segment length (left minus right coordinates)
 and edge length (difference in birth times between parent and child),
-summed across all edges.
+across all edges.
 This area is also equal to the sum of the total lengths of all marginal trees
-(i.e., the trees describing inheritance at each position on the genome).
-Therefore, the total area of a tree sequence is equal to the sequence length
-multiplied by the mean area of a marginal tree.
+(i.e., the trees describing inheritance at each position on the genome),
+so can be computed as the sequence length multiplied by the mean marginal tree length.
 Since we analyze tree sequences arising from a Wright--Fisher model,
-statistical properties of a marginal tree are well-described by coalescent theory.
+statistical properties of a marginal tree are fairly well-described by coalescent theory.
 Similar arguments to those below go back at least to \citet{watterson1975number},
@@
 (except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
-Suppose that $T_x$ is the marginal tree describing how each of the present-day population
-descends from their most recent common ancestor
-at a particular genomic position $x$ on the chromosome,
-and write $|T_x|$ for the total length of the tree,
-i.e., the number of ancestors 
-from which some but not all of the present-day population has inherited at position $x$
-(including the population itself).
+Suppose that $\mathcal{T}_x$ is the marginal tree at genomic position $x$,
+and write $|\mathcal{T}_x|$ for the total length of the tree.
 For the tree at nearby position $x + dx$ to be different,
 there must have been a crossing-over between $x$ and $x + dx$
-in one of the $|T_x|$ meioses that gave birth to those individuals.
+in one of the $|\mathcal{T}_x|$ meioses that gave birth to those individuals.
 (Recall that these ``individuals'' are haploid.)
@@
 so that length is equal to the expected number of crossing-overs per generation,
-the expected distance until the next crossing-over is $1/|T_x|$.
+the expected distance until the next crossing-over is $1/|\mathcal{T}_x|$.
 If every crossing-over changed the marginal tree,
@@
 
-If $T$ is large enough that all samples coalesce within the simulation with high probability
--- $T = 20N$, say --
+If $T$ is large relative to $N$,
+so that all marginal trees have a single root with high probability,
 then coalescent theory tells us that
@@
 What about mutations?
-If we were to add mutations in forwards time
-under the infinite-sites model with total mutation rate per generation $\mu$,
-there would also be around $\mu N T$ mutations, and the same number of sites.
-% simply because there was a total of $N T$ meioses.
-Since mutations fall as recombinations do on the marginal trees,
-the number of mutations is proportional to the total length of all the marginal trees,
-and so by the same argument as for the number of edges,
-adding them after simplification results in only around $\mu 2 N \log(N)$ mutations.
+Forwards-time generation of infinite-sites mutations with total mutation rate per generation $\mu$
+would produce around $\mu N T$ mutations (and the same number of sites),
+simply because there was a total of $N T$ meioses.
+Since mutations that are retained after simplification
+are precisely those that fall on the marginal trees,
+the number of mutations is proportional to the total area of the tree sequence.
+By the same argument as for the number of edges,
+this will result in around $\mu 2 N \log(N)$ mutations.
 With $N=2 \times 10^4$ and $T=10N$,
-this implies that adding neutral mutations afterwards reduces the number of mutations that must be generated
+this implies that adding neutral mutations
+to the simplified tree sequence reduces the number of mutations that must be generated
 by a factor of 10,000.
@@
 and at most one row in the site table, the space required for $M$ mutations is $O(M)$;
-combined with the $O(N \log N)$ storage for edges and nodes of a simplified tree sequence above,
+combined with the $O(N \log N)$ storage for edges and nodes of a simplified tree sequence,
 this implies that the full results of a simulation of $N$ individuals

### 2018-01-15 forwards_paper.tex
merged disco edits; spelling

@@
 \citet{padhukasahasram2008exploring} obtained impressive speedups for a Wright--Fisher simulation
-by keeping track of genealogies over the preceeding 8 generations
+by keeping track of genealogies over the preceding 8 generations
 and only tracking neutral genotypes for those segments having descendants across this window.
@@
 or the coalescent within a known pedigree.  For such models, one
-could in principle generate tables of nodes and edges to be simpified in \msprime{}.  The resulting succinct tree
+could in principle generate tables of nodes and edges to be simplified in \msprime{}.  The resulting succinct tree
 sequence object would be in the same format as those generated by \msprime{}'s built-in simulate function, and therefore
-compatible with exising methods for downstream analyses.
+compatible with existing methods for downstream analyses.
 
@@
 we also implemented a parallelized version using Python's \texttt{queue.Queue} to run the simplification step in a separate Python
-thread. Our implementation allows a maxmimum of four simplification intervals to be in the queue at once. This
+thread. Our implementation allows a maximum of four simplification intervals to be in the queue at once. This
 parallelized version also performed fitness calculation in parallel using two threads of execution in C++.
@@
 the Wellcome Trust (grant 100956/Z/13/Z to Gil McVean);
-and XXX.
+the NIH (R01GM115564) to KRT.
 

### 2018-01-15 forwards_paper.tex
minor fixen

@@
 The tree sequence must store edges describing the leftmost marginal tree,
-which requires at most $N-2$ edges.
+which requires at most $2N-2$ edges.
 Then, each time the marginal tree changes along the sequence,
@@
 %    N/n = (T+2)/2 and so n = 2N/(T+2)
-is expected to have around $r(T)$ roots, where $r(T) \approx 2N/(T+2)$.
+is expected to have around $r(T)$ roots, where $r(T) = 2N/(T+2)$.
 The total tree length over this time is
-$2N \sum_{k=N}^{r(T)+1} 1/(k-1) \approx 2 N \log(N/r(T))$.
+$2N \sum_{k=r(T)+1}^N 1/(k-1) \approx 2 N \log(N/r(T))$.
 This is a crude estimate for several reasons:

### 2018-01-15 sims/space-complexity.py

@@ -65,7 +65,7 @@ def ub(T, N):
     # return 2 * (N - 1) + 8 * N * np.log(N/(1+2*N/(T+2)))
     # return 2 * N * (1 + 4 * np.log(N * T / (T + 2 * N)))
     # return 2 * N * (1 + 4 * np.log(N * T / (2 * T + 2 * N)))
-    return 2 * N * (1 - (1 / N) + 4 * np.log(min(N,  (T + 2) / 2 )))
+    return 2 * N * (1 + 4 * np.log(min(N,  (T + 2) / 2 )))
 
 
 def get_mean_edges_per_transition(ts):

### 2018-01-17 forwards_paper.tex
note on harmonic series
missing 1/k
Fixed parathesis in grant ACK.

@@
 The total tree length over this time is
-$2N \sum_{k=r(T)+1}^N 1/(k-1) \approx 2 N \log(N/r(T))$.
+$2N \sum_{k=r(T)+1}^N 1/(k-1) \approx 2 N \log(N/r(T))$,
+since $\sum_{k=1}^x 1/k \approx \log(x) + \gamma$.
 This is a crude estimate for several reasons:
@@
 the Sloan Foundation and the NSF (under DBI-1262645) to PR;
-the Wellcome Trust (grant 100956/Z/13/Z to Gil McVean);
+the Wellcome Trust (grant 100956/Z/13/Z) to Gil McVean;
 the NIH (R01GM115564) to KRT.

### 2018-01-21 forwards_paper.tex
Added Gil to the ACK list.
edits for AKs comments

@@
         An example tree sequence with three samples over a chromosome of length 10.
-        The left-hand panel shows the tree sequence pictorially in two different ways:
-        (top) as a sequence of tree topologies
-        and (bottom) the spatial extent of the edges that define these topologies.
+        The left-hand panels show the tree sequence pictorially in two different ways:
+        (top) a sequence of tree topologies; the first tree extends from genomic position 0 to 5,
+        and the second from 5 to 10; and
+        (bottom) the edges that define these topologies,
+        displayed over their corresponding genomic segment
+        (for instance, the edge from node 2 to node 4 is present only on the interval from 0 to 5).
         The right-hand panels show the specific encoding
@@
 and edge tables ($\Nt$ and $\Et$) recording the simulated history.
+In each generation, the current population is stored in $P$,
+while produced offspring are placed in $P'$.
 The tables are simplified every $s$ generations, removing genealogical
-information from $\Nt$ and $\Et$ irrelevant to the current population $P$.
+information from $\Nt$ and $\Et$ irrelevant to the current population.
 }
@@
 
+\paragraph{Essential steps in tree sequence recording}
+To record a tree sequence, including mutations,
+a simulator must record:
+\begin{itemize}
+    \item the birth time of each individual in the Node Table,
+    \item the genetic segments each individual inherits from her parents in the Edge Table, and
+    \item the individual each new mutation occurs in, and the derived state in the Mutation Table,
+        adding also to the Site Table if the site is previously unmutated.
+\end{itemize}
+Recall that here we mean ``\emph{haploid} individual'', i.e., gamete.
+Before simplification, time must be translated to ``time ago'' if it is not already,
+and the tables must be sorted.
+Algorithm~\algref{W} avoided both of these details since $t$ denotes 
+``time until the end of the simulation'',
+and nonoverlapping generations ensure that the tables will be already sorted.
+Finally, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
+being relabeled in \msprime{}'s tables as $0, 1, \ldots, N-1$,
+there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
+with the node IDs recorded in the tables.
+
 %%%%%%
@@
         found across the chromosome (blue numbers denote locations on the chromosome).
-        Labels after simplification are shown in red.
+        Labels after simplification are shown in red (see text).
         \textbf{(C)} Tables recording the tree sequence after simplification
@@
 during the process of simplifying this tree sequence.
+Since the method begins with the samples and moves back through time,
+in the output tree sequence, the $n$ samples will be numbered $0, 1, \ldots, n-1$
+and subsequent nodes will be ordered by time since birth.
+This is seen in the red labels of Figure~\ref{fig:method_diagram}.
 
@@
 \section*{Acknowledgements}
-Thanks to Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
+Thanks to Gil McVean, Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
 Work on this project was supported by funding from

### 2018-01-21 forwards_paper.tex
a few more words

@@
         A depiction of the state of the simplification algorithm
-        at each point in time,
+        as it moves up through the embellished pedigree
         in the example of Figure~\ref{fig:method_diagram}A.
-        Ancestral material from J (red) and K (blue) are traced up through the pedigree
-        until they coalesce;
-        the smaller colored chromosomes on either side of each solid arrow show the bits inherited from each of the two parental chromosomes;
-        each time two colors overlap, a coalescence occurs, and two edges are output.
+        Following the ``paint pot'' description in the text,
+        we begin by coloring J and K's genomes in red and blue respectively,
+        then trace how these colors were inherited back up through the pedigree
+        until they coalesce.
+        To aid in this,
+        the smaller colored chromosomes on either side of each solid arrow
+        show the bits inherited from each of the two parental chromosomes,
+        with genomic position 0.0 on the bottom and 1.0 at the top.
+        Each time a red and a blue segment overlap, a coalescence occurs, two edges are output,
+        and we stop propagating that segment.
         For instance, both J and K inherit from H between 0.5 and 0.9,
         which resulted in the first two edges of the simplified table of Figure~\ref{fig:method_diagram}C.
+        Later, both inherit from E between 0.2 and 0.5,
+        along the paths J-H-G-E and K-I-E respectively,
+        resulting in the next two edges.
         \label{fig:simplify_state}

### 2018-01-22 forwards_paper.tex

@@
 Recall that here we mean ``\emph{haploid} individual'', i.e., gamete.
-Before simplification, time must be translated to ``time ago'' if it is not already,
-and the tables must be sorted.
-Algorithm~\algref{W} avoided both of these details since $t$ denotes
-``time until the end of the simulation'',
-and nonoverlapping generations ensure that the tables will be already sorted.
-Finally, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
+Before simplification, time must be translated to ``time ago'' if it is not already.
+which was avoided in Algorithm~\algref{W} since there $t$ denoted
+``time until the end of the simulation''.
+Furthermore, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
 being relabeled in \msprime{}'s tables as $0, 1, \ldots, N-1$,
@@
 with the node IDs recorded in the tables.
+% Finally, the first step in simplification is to \emph{sort} the tables
+% according to several criteria, for algorithmic efficiency,
+% as described in the documentation.
 

### 2018-01-23 forwards_paper.tex
added note about sortedness
minor changes re jda comments

@@
         An example tree sequence with three samples over a chromosome of length 10.
-        The left-hand panels show the tree sequence pictorially in two different ways:
+        The leftmost panels show the tree sequence pictorially in two different ways:
         (top) a sequence of tree topologies; the first tree extends from genomic position 0 to 5,
@@
         (for instance, the edge from node 2 to node 4 is present only on the interval from 0 to 5).
-        The right-hand panels show the specific encoding
+        The remaining panels show the specific encoding
         of this tree sequence in the four tables (nodes, edges, sites and mutations)
@@
     \item the birth time of each individual in the Node Table,
-    \item the genetic segments each individual inherits from her parents in the Edge Table, and
+    \item the genetic segments each individual inherits from its parents in the Edge Table, and
     \item the individual each new mutation occurs in, and the derived state in the Mutation Table,
@@
 Before simplification, time must be translated to ``time ago'' if it is not already.
-which was avoided in Algorithm~\algref{W} since there $t$ denoted
-``time until the end of the simulation''.
+(This was avoided in Algorithm~\algref{W} since there $t$ denoted
+``time until the end of the simulation''.)
 Furthermore, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
@@
     \caption{
-        A depiction of the state of the simplification algorithm
+        A depiction of each state of the simplification algorithm
         as it moves up through the embellished pedigree
@@
 the Wellcome Trust (grant 100956/Z/13/Z) to Gil McVean;
-the NIH (R01GM115564) to KRT.
+the NIH (R01GM115564) to KRT;
+and the USF\&WS to H.\ Bradley Shaffer.
 
@@
 write $Q[0]$.
+Finally, note that to keep the exposition simple,
+the algorithm described here requires that the input node table
+is sorted by time since the parent's birth;
+but the version implemented in \msprime{} makes different sortedness assumptions
+(described in the documentation).
 

### 2018-01-26 forwards_paper.tex
author summary

@@
 % Please excuse the ragged right margins in the source.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section*{Author Summary}
+
+Sexually reproducing organisms are related to the others in their species
+by the complex web of parent-offspring relationships that constitute the pedigree.
+In this paper, we describe a way to record all of these relationships,
+as well as how genetic material is passed down through the pedigree,
+during a forwards-time population genetic simulation.
+To make effective use of this information,
+we describe both efficient storage methods for this embellished pedigree
+as well as a way to remove all information 
+that is irrelevant to the genetic history of a given set of individuals,
+which dramatically reduces the required amount of storage space.
+Storing this information allows us to produce whole-genome sequence
+from simulations of large populations in which we have not explicitly recorded new genomic mutations;
+we find that this results in computational run times of up to 50 times faster
+than simulations forced to explicitly carry along that information.
 

### 2018-01-26 forwards_paper.tex
author affils

@@
 \title{Efficient pedigree recording for fast population genetics simulation}
-\author{Jerome Kelleher,
-        Kevin R. Thornton,
-        Jaime Ashander, and
-        Peter L. Ralph}
+\author{Jerome Kelleher\footnote{Big Data Institute, University of Oxford},
+        Kevin R. Thornton\footnote{Ecology and Evolutionary Biology, University of California at Irvine},
+        Jaime Ashander\footnote{Ecology and Evolutionary Biology, University of California at Los Angeles}, and
+        Peter L. Ralph\footnote{Institute for Ecology and Evolution, University of Oregon}}
 \maketitle

### 2018-03-10 talks/make_figs.py
draft

@@ -1,0 +1,11 @@
+import msprime
+import os
+
+if not os.path.exists("sim_ts"):
+    os.makedirs("sim_ts")
+
+ts = msprime.simulate(5, length=1, recombination_rate=5, random_seed=23)
+
+for k, t in enumerate(ts.trees()):
+    t.draw(path="sim_ts/sim_ts.{0:03d}.svg".format(k))
+

### 2018-03-10 talks/make_figs.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+
+python3 make_figs.py
+for x in sim_ts/*.svg; do make ${x%%svg}gif; done
+make sim_ts.anim.gif

### 2018-04-12 talks/poster/make_example.R
poster

@@ -1,0 +1,18 @@
+#!/usr/bin/Rscript
+
+pycode <- '
+import msprime
+ts = msprime.simulate(8, recombination_rate=2.0)
+for j,t in enumerate(ts.trees()):
+   t.draw("tree{}.svg".format(j))
+
+ts.dump_text(nodes=open("nodes.txt", "w"), edges=open("edges.txt", "w"))
+'
+
+nodes <- read.table("nodes.txt", header=TRUE, fill=TRUE)
+edges <- read.table("edges.txt", header=TRUE)
+
+library(xtable)
+
+print(xtable(nodes[,c("id","time")], digits=2), include.rownames=FALSE)
+print(xtable(edges, digits=2), include.rownames=FALSE)

### 2018-04-12 talks/poster/evo-wibo-2018-poster.tex
poster

\documentclass[fontscale=0.38,a0paper]{baposter}

\tracingstats=2

\usepackage{times}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{relsize}
\usepackage{multirow}
\usepackage{bm}
\usepackage{caption}
\usepackage[authoryear,round]{natbib}
\usepackage[hidelinks]{hyperref}

\usepackage{tabularx}
\usepackage{booktabs}

\usepackage{colortbl}

\usepackage{multicol}

%\usepackage{pgfbaselayers}
%\pgfdeclarelayer{background}
%\pgfdeclarelayer{foreground}
%\pgfsetlayers{background,main,foreground}

%\usepackage{tikz}
%\usetikzlibrary{arrows,backgrounds,patterns,matrix,shapes,fit,calc,shadows,plotmarks,decorations.pathmorphing,positioning,trees}

\renewcommand{\familydefault}{\sfdefault}
\usepackage{helvet}
\usepackage{bookman}
\usepackage{palatino}

% My stuff
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
% end my stuff

\selectcolormodel{cmyk}

% \graphicspath{{images/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Some symbols used in the text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\newcommand{\sub}[2]{\ensuremath{#1_{\textrm{#2}}}}
%\newcommand{\gen}[2]{\ensuremath{\textrm{#1}_{\textrm{#2}}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Multicol Settings
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setlength{\columnsep}{1em}
\setlength{\columnseprule}{0mm} % no vertical line

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Save space in lists. Use this after the opening of the list
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\compresslist}{
\setlength{\itemsep}{1pt}
\setlength{\parskip}{0pt}
\setlength{\parsep}{0pt}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Other Macros
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \renewcommand{\labelenumi}{\alph{enumi})}
% \renewcommand{\theenumi}{\alph{enumi})}
\newcommand{\postercaption}[1]{\begin{minipage}{\linewidth}\center\smaller
  {#1}\end{minipage}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Begin of Document
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Here starts the poster
%%%---------------------------------------------------------------------------
%%% Format it to your taste with the options
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Define some colors
\definecolor{silver}{cmyk}{0,0,0,0.3}
\definecolor{black}{cmyk}{0,0,0.0,1.0}
\definecolor{white}{cmyk}{0,0,0.0,0.0}
\definecolor{darkSilver}{cmyk}{0,0,0,0.1}

\definecolor{darkYellow}{cmyk}{0,0,1.0,0.5}
\definecolor{yellow}{cmyk}{0,0,0.9,0.0}
\definecolor{reddishyellow}{cmyk}{0,0.22,1.0,0.0}
\definecolor{lightyellow}{cmyk}{0,0,0.3,0.0}
\definecolor{lighteryellow}{cmyk}{0,0,0.1,0.0}
\definecolor{lightestyellow}{cmyk}{0,0,0.05,0.0}

\definecolor{DavisBlue}{cmyk}{1,0.56,0,0.34}
%    C = 100  M = 56  Y = 0 K = 34

\definecolor{HohenheimBlue}{cmyk}{1,0.5,0,0.45}
\definecolor{HohenheimLightLightBlue}{RGB}{243,250,255}
\definecolor{HohenheimLightBlue}{RGB}{215,221,235}
\definecolor{HohenheimDarkBlue}{RGB}{52,104,152}
\definecolor{darkCyan}{cmyk}{1,1,0,0.5}
\definecolor{cyan}{cmyk}{1,0,0,0.0}
\definecolor{lightcyan}{cmyk}{0.3,0,0,0.0}
\definecolor{lightercyan}{cmyk}{0.1,0,0,0.0}
\definecolor{lightestcyan}{cmyk}{0.05,0,0,0.0}

\typeout{Poster Starts}
%Define custom background
\background{
  \begin{tikzpicture}[remember picture,overlay]%
    \draw (current page.north west)+(-2em,2em) node[anchor=north west] {\includegraphics[height=\textheight]{silhouettes_background}};
  \end{tikzpicture}%
}

\newlength{\leftimgwidth}
\begin{poster}%
  % Poster Options
  {
  % Columns and Column spacing
  columns=3,
  colspacing=1.5em,
  % Background
  % background=user,
  background=shade-tb,
  % Poster header options
  eyecatcher=yes,
  posterheaderheight=0.15\textheight,
  % Format of textbox header
  headerborder=open,
  % headershape=small-rounded,
  headershape=roundedright,
  headershade=shade-tb,
  % headershade=plain,
  headerfont=\Large\textsf, %Sans Serif
  % Format of textbox
  boxborder=small-rounded,
  % boxborder=roundedleft,    
  boxshade=plain,
  linewidth=0.5pt,
  % Color style
  % bgColorOne=lightestcyan,
  % bgColorTwo=lightcyan,
  bgColorOne=white,
  bgColorTwo=white,
  % borderColor=cyan,
  borderColor=black,
  headerColorOne=DavisBlue,
  % headerColorOne=cyan,
  headerColorTwo=black,
  % headerFontColor=black,
  headerFontColor=white,
  % boxColorOne=lightercyan,
  boxColorOne=HohenheimLightLightBlue,
  boxColorTwo=white,
  % Show grid to help with alignment
  grid=no
  % grid=yes
  }
  % Eye Catcher
  {
  \makebox[15em][r]{%
    \begin{minipage}{15em}
       \hfill
       \includegraphics[width=15em]{tskit_logo} \\
    \end{minipage}
    }
  }
  % Title
  {\sf %Sans Serif
  %\bf% %bold  
  \vspace{0.5em}
     \textbf{\textcolor{DavisBlue}{Recording pedigrees for fast genome simulation and analysis}}\vspace{0.5em}}
  % Authors
  {\sf %Sans Serif
    Peter Ralph$^{\ddagger}$ \\  \vspace{-1.0mm}
    {\small \textit{$\ddagger$ Mathematics and Biology, University of Oregon} }\\
    {\textbf{code:}  \url{https://github.com/tskit-dev} }\\
    {\textbf{preprint:}  \url{https://www.biorxiv.org/content/early/2018/01/26/248500} }\\
  }
  % Project logo
  {
    \makebox[15em][r]{%
      \begin{minipage}{12em}
        \hfill
          \includegraphics[width=12em]{UOSignature-STK-BLK} \\
      \end{minipage}
    }
  }

  % Width of left inset image
%  \setlength{\leftimgwidth}{0.78em+0.0em}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Now define the boxes that make up the poster
%%%---------------------------------------------------------------------------
%%% Each box has a name and can be placed absolutely or relatively.
%%% The only inconvenience is that you can only specify a relative position 
%%% towards an already declared box. So if you have a box attached to the 
%%% bottom, one to the top and a third one which should be in between, you 
%%% have to specify the top and bottom boxes before you specify the middle 
%%% box.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Tree sequences: all the genealogies}{name=intro,column=0,row=0,span=1}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For a set of sampled chromosomes,
at each position along the genome there is a genealogical tree
that says how they are related.

A \textbf{tree sequence} describes this sequence of trees.

\emph{Observations:}

\begin{itemize}

    \item The \emph{pedigree} (parental relationships) plus crossover locations
        would give us the tree sequence for \emph{everyone, ever}.

    \item Much less can fully describe the history relevant to a \emph{sample} of genomes.

    \item This information is equivalent to the Ancestral Recombination Graph (ARG).

\end{itemize}

\citep{kelleher2016efficient}
introduced the \textbf{tree sequence} data structure
for a fast coalescent simulator, \texttt{msprime}.

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Tables: a data structure for tree sequences}{name=tables,column=0,span=1,below=intro}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \begin{itemize}

        \item \textbf{Edges:}   Who inherits from who; only \emph{necessary} for coalescent events.
            \emph{Records:} interval (left, right); parent node; child node.

        \item \textbf{Nodes:}   The ancestors those happen in.
            \emph{Records:} time ago (of birth); ID (implicit).

        \item \textbf{Mutations:}   When state changes along the tree.
            \emph{Records:} site it occurred at; node it occurred in; derived state.

        \item \textbf{Sites:}   Where mutations fall on the genome.
            \emph{Records:} genomic position; ancestral (root) state; ID (implicit).

    \end{itemize}

    \begin{center}
        \includegraphics[width=0.75\textwidth]{../file_size}
    \end{center}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Example}{name=ex,column=0,below=tables}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \includegraphics[width=\textwidth]{example_tree_sequence}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Another example}{name=ex2,column=0,below=ex}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \noindent
    \tiny
\hfill
\begin{tabular}{|rr|}
  \hline
    \textbf{Nodes:} & \\
id & time \\ 
  \hline
  0 & 0.00 \\ 
    1 & 0.00 \\ 
    2 & 0.00 \\ 
    3 & 0.00 \\ 
    4 & 0.00 \\ 
    5 & 0.00 \\ 
    6 & 0.00 \\ 
    7 & 0.00 \\ 
    8 & 0.09 \\ 
    9 & 0.31 \\ 
   10 & 0.39 \\ 
   11 & 0.41 \\ 
   12 & 0.43 \\ 
   13 & 0.97 \\ 
   14 & 1.04 \\ 
   15 & 1.21 \\ 
   16 & 1.36 \\ 
   17 & 1.45 \\ 
   18 & 2.41 \\ 
   19 & 2.46 \\ 
   20 & 2.85 \\ 
   21 & 3.84 \\ 
   22 & 4.45 \\ 
   \hline
\end{tabular}
\hfill
\begin{tabular}{|rrrr|}
  \hline
    \textbf{Edges:} & & & \\
left & right & parent & child \\ 
  \hline
0.00 & 1.00 &   8 &   2 \\ 
  0.00 & 1.00 &   8 &   6 \\ 
  0.00 & 1.00 &   9 &   5 \\ 
  0.00 & 1.00 &   9 &   7 \\ 
  0.00 & 1.00 &  10 &   3 \\ 
  0.00 & 1.00 &  10 &   8 \\ 
  0.00 & 1.00 &  11 &   0 \\ 
  0.00 & 1.00 &  11 &   1 \\ 
  0.00 & 1.00 &  12 &   9 \\ 
  0.00 & 1.00 &  12 &  10 \\ 
  0.99 & 1.00 &  13 &   4 \\ 
  0.99 & 1.00 &  13 &  12 \\ 
  0.00 & 0.19 &  14 &  11 \\ 
  0.00 & 0.19 &  14 &  12 \\ 
  0.93 & 0.99 &  15 &   4 \\ 
  0.93 & 1.00 &  15 &  11 \\ 
  0.99 & 1.00 &  15 &  13 \\ 
  0.30 & 0.54 &  16 &   4 \\ 
  0.91 & 0.93 &  16 &   4 \\ 
  0.30 & 0.54 &  16 &  11 \\ 
  0.91 & 0.93 &  16 &  11 \\ 
  0.54 & 0.69 &  17 &   4 \\ 
   \hline
\end{tabular}
\begin{tabular}{|rrrr|}
  \hline
    \textbf{cont'd} & ... &  ... & ... \\
left & right & parent & child \\ 
  \hline
  0.54 & 0.69 &  17 &  12 \\ 
  0.69 & 0.91 &  18 &   4 \\ 
  0.69 & 0.91 &  18 &  11 \\ 
  0.00 & 0.30 &  19 &   4 \\ 
  0.69 & 0.69 &  19 &  11 \\ 
  0.19 & 0.46 &  19 &  12 \\ 
  0.69 & 0.99 &  19 &  12 \\ 
  0.00 & 0.19 &  19 &  14 \\ 
  0.93 & 0.99 &  19 &  15 \\ 
  0.30 & 0.46 &  19 &  16 \\ 
  0.91 & 0.93 &  19 &  16 \\ 
  0.69 & 0.69 &  19 &  17 \\ 
  0.69 & 0.91 &  19 &  18 \\ 
  0.19 & 0.30 &  20 &  11 \\ 
  0.19 & 0.30 &  20 &  19 \\ 
  0.60 & 0.69 &  21 &  11 \\ 
  0.60 & 0.69 &  21 &  17 \\ 
  0.54 & 0.60 &  22 &  11 \\ 
  0.46 & 0.54 &  22 &  12 \\ 
  0.46 & 0.54 &  22 &  16 \\ 
  0.54 & 0.60 &  22 &  17 \\ 
   \hline
\end{tabular}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Contributions}{name=contrib,column=1,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \begin{itemize}
        \compresslist
    \item Jerome Kelleher -- algorithms, data structures, and core development
    \item Kevin Thornton -- implementation, profiling with \texttt{fwdpp},
    \item Jaime Ashander -- implementation, profiling with \texttt{simuPOP}
    \item Ben Haller -- implementation in \texttt{SLiM}
    \item Jared Galloway -- implementation in \texttt{SLiM}
    \end{itemize}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Record history in forwards simulations?}{name=forwards,column=1,below=contrib}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \begin{itemize}

        \item Coalescent simulations are \emph{much faster}
            than forwards-time, individual-based simulations
            because they don't have to keep track of \emph{everyone},
            only the ancestors of your sample.

        \item \textbf{But:} selection, or sufficient geographic structure,
            break the assumptions of coalescent theory.

        \item If we \emph{record the tree sequence}
            that relates everyone to everyone else,
            after the simulation is over we can put neutral mutations down on the trees.

        \item Since neutral mutations don't affect demography,
            this is \emph{equivalent} to having kept track of them throughout.

        \item This means recording the entire genetic history of \textbf{everyone} in the population, \textbf{ever}.

    \end{itemize}

\includegraphics[width=7em]{../finger_right} {\Large \emph{This may be a bad idea.}}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{How to do it}{name=howfwd,column=1,below=forwards}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \begin{enumerate}

        \item add each gamete to the Node Table,
        \item add entries to the Edge Table
            recording which parent each gamete inherited each bit of genome from, and
        \item add any new selected mutations to the Mutation Table 
            and (if necessary) their locations to the Site Table.

    \end{enumerate}

\textbf{But,} we won't end up needing the entire history of everyone ever.

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Simplification}{name=simplify,column=1,below=howfwd}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    Given an input tree sequence
    and a subset of its nodes (the \emph{samples}),
    we want a new tree sequence for which:

    \begin{enumerate}
        \item All marginal trees match the corresponding subtree 
            in the input tree sequence.

        \item Every non-sample node in marginal trees has at least two children.

        \item All nodes and edges ancestral to at least one sample.

        \item No adjacent redundant edges \\
            (e.g., $(\ell, x, p, c) + (x, r, p, c) \rightarrow (\ell, r, p, c)$).
    \end{enumerate}

    To simplify a tree sequence
    to the history of the \emph{samples},
    we:

    \begin{itemize}
        \item Paint each \emph{sampled} chromosome a distinct color.

        \item Moving back up the tree sequence,
            copy colors of each chromosome to the parental chromosomes
            they inherited from.

        \item If two colors go in the same spot (\emph{coalescence}),
            replace with a new color (unique to that ancestor).
            Output a node for the ancestor and an edge for the coalescence.

        \item Once all colors have coalesced in a given segment,
            stop propagating it.
    \end{itemize}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Simulations, $50\times$ faster}{name=fast,column=2,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    Recording, simplifying, and output of tables, was done with \texttt{C} code in \texttt{tskit}.
    The simulation was done in \texttt{fwdpp} (C++ by Kevin Thornton),
    connected with \texttt{pybind11} and \texttt{numpy}.
    \textit{Machine:} Ubuntu / 2x 2.6 GHz Intel E5-2650 CPU.

    \emph{Simulation parameters:}
    \begin{itemize}
            \compresslist
        \item  Wright-Fisher population of size $N$, for $10N$ generations
        \item  neutral mutation rate $\mu$ equal to recombination rate $r$ per gamete
        \item  many, weakly deleterious mutations: rate $\mu/100$ with
            $s$ exponentially distributed with mean $2.5/N$.
    \end{itemize}

    \emph{Note:}
    simulations that recorded tree sequences ("pedigree recording")
    had neutral mutation rate was zero,
    but neutral mutations were added \emph{afterwards}.

    \begin{center}
    \includegraphics[width=\textwidth]{rawspeed}

    \includegraphics[width=0.75\textwidth]{speedup}
    \end{center}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Fast statistics computation}{name=stats,column=2,below=fast}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \textbf{A general class of statistics:}
    any polynomial $f(p_1, \ldots, p_k)$ gives a statistic on $k$ populations,
    by setting $p_i$ to the allele frequency in population $i$ and averaging $f$ across loci.
    \emph{(Example: divergence is $p_1(1-p-2) + p-2(1-p_1)$.)}

    The polynomial defines a weighting function on edges of the trees in the tree sequence,
    and the statistic can be calculated by summing across mutations
    weighted by the edge they occur on.
    \emph{(Example: proportion of pairs from the two populations on opposite sides of the edge.)}

    \textbf{Computation on a tree sequence:}
    If we compute weights on the first tree, then we only need update those weights above nodes that change
    in the second tree.

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{Summary}{name=summary,column=2,below=stats}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    Tree sequences 
    \begin{itemize}
        \item \ldots can make your simulations much faster
        \item \emph{because} they record genealogical history.
        \item (And, then you have genealogies!)
        \item \ldots can store genomic data \emph{very} efficiently,
        \item and compute with genomic data \emph{very} quickly.
    \end{itemize}

    \emph{Why} are they so useful?
    It is a compression scheme, provided by the process itself that produced the data.
    \emph{(think: optimal compression into shared haplotypes)}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{}{name=references,column=2,below=summary}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

  \scriptsize
  \renewcommand{\section}[2]{\vskip 0.0em}
  \bibliographystyle{abbrvnat}
  \setlength{\bibsep}{0.0pt}
  \bibliography{../../references}
  }
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\headerbox{}{name=trees,column=0,above=bottom,span=3}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \includegraphics[width=0.085\textwidth]{tree0}
    \includegraphics[width=0.085\textwidth]{tree1}
    \includegraphics[width=0.085\textwidth]{tree2}
    \includegraphics[width=0.085\textwidth]{tree3}
    \includegraphics[width=0.085\textwidth]{tree4}
    \includegraphics[width=0.085\textwidth]{tree5}
    \includegraphics[width=0.085\textwidth]{tree6}
    \includegraphics[width=0.085\textwidth]{tree7}
    \includegraphics[width=0.085\textwidth]{tree8}
    \includegraphics[width=0.085\textwidth]{tree9}
    \includegraphics[width=0.085\textwidth]{tree10}

}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{poster}%
\end{document}

### 2018-06-03 review-responses.tex
responses

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it PLoS Computational Biology}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are pleased to submit a revision of our manuscript,

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
    Jerome Kelleher, Kevin Thornton, \\
    Jaime Ashander \emph{and} Peter Ralph \\
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
    Both reviewers are clearly very enthusiastic about this manuscript, but both
    have made suggestions that require major revisions to address. Please be
    very careful in addressing the following points exhaustively:
    \begin{itemize}
        \item  Discuss the limitations imposed by Wright-Fisher assumptions (Rev 1) 
            and how the proposed approach may/may not be extended 
            to other demographic and population scenarios.

        \item  Improve the readability of the text and provide necessary background information (Rev 1)

        \item  Discuss implementation availability (Rev 2)
    \end{itemize}
\end{quote}

We thank the reviewers for helpful feedback, and respond in detail to their points below.
Something we did that should help address all these points
was to make more clear the main purpose of this paper:
to describe algorithms and data structures for tree sequences,
and how these are useful in simulation.
Our main goal is not to write a simulation package.
This is useful and important, because: 
(a) we provide a fully functional and flexible, well-documented API
so that developers can include this idea in their own simulator
(with no restrictions beyond conserved synteny of chromosomes); and
(b) we explain the data structures,
which are closely tied to the idea of a tree sequence,
and therefore important for even end users of a simulator to understand.
We describe and implement Wright--Fisher simulations not because of any limitation of our method,
but to keep things as simple as possible, only focusing on those aspects necessary 
for recording the tree sequence.

TODO: mention the tutorials?

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    Kelleher and colleagues present a novel method to efficiently record pedigree
    information in forward-time individual-based simulations of the Wright-Fisher
    process in a single large population of haploids. The method shows a large
    performance improvement compared to forward-time simulations of the same
    Wright-Fisher model without pedigree recording. The gain obtained with pedigree
    recording is due to the fact that mutations are generated post-hoc by placing
    them on the gene trees that resulted from the forward-time demographic process
    recorded as pedigree and cross-over information. The method for storing
    "succinct tree sequences" has been implemented in the coalescent-simulation
    software msprime and the authors here provide a Python API to use the data
    format previously developed in msprime in forward-time simulators. This
    represents a neat addition to the tool kit of population genetics simulations.
\end{quote}

We'd like to emphasize that although the tools for dealing with tree sequences
are currently bundled with the coalescent simulator, \msprime{},
this is for historical reasons only, 
and we are in the process of separating out these tools into a separate package,
to be called \tskit.
The format for storing tree sequences was developed for the coalescent simulator,
but the work we describe here generalized the data structure and algorithms
to allow it to be used in any simulator.
We have stated this more clearly in the text at \llname{msprime_rel}.

TODO: think about switching to \tskit{} in the paper?

\begin{point}{}
    Overall, I find the method very clever and appealing, and I am convinced that
    it will be well received by the community. However, nowadays, researchers
    generate and use genomic data in organisms other than humans and from wild
    populations that by far are not Wright-Fisher populations. How are you selling
    your method to that broader category of researchers? Compared to SPLATCHE,
    which simulates structured populations with gene flow, it may look like a
    regression (quantiNemo implemented something similar, but I don't think it was
    ever published).
    \emph{[\ldots]}
    As a developer, I am interested in the limitations and the constraints that may
    prevent me to implement your method to model populations that are not W-F.
    populations. In particular, you didn't address the question of structured
    populations with gene flow, or the effect of overlapping generations. Although
    developing an approach that is efficient in structured populations is probably
    out of scope here, giving a minimum of guidelines or just discussing the issue
    would greatly improve the scope of the manuscript.
\end{point}

\reply{
    We're sorry to have not made this more clear,
    but our goal was \emph{not} to make a faster WF simulator,
    but rather to describe a scheme that can be easily applied to \emph{any} simulator
    whatsoever, as long as it has the notion of a recombining chromosome.
    We have added a note to this effect before the proof-of-concept WF algorithm \llname{ll:not_wf};
    and substantially expanded the discussion of what steps need to be taken in general \revref.
    Also see the summarizing paragraph at \llname{ll:other_words}.
    The only thing that should stand in your way as a developer
    to implementing this is bookkeeping;
    and we've provided tools to make this bookkeeping easier
    (like the \texttt{sort tables} functionality,
    so that the simulator does not have to guarentee sortedness of the tables itself).
}

\begin{point}{}
    More generally, I found the ms very technical and addressed to a specialized
    audience. I would not assume, as you do, that readers are necessarily
    proficient in coalescent theory. Therefore, please improve the introduction by
    providing more information on: \emph{(points below)}
\end{point}

\reply{
    Thanks for the suggestion;
    since this paper is entirely about forwards simulation,
    we hope not to depend on coalescent theory at all in any crucial way.
    We have added some more explanation of the relationship to simplification
    at \revref;
    and have tried to better signpost the technical arguments at \llname{ll:coal_signpost}.
}

\begin{point}{}
    The idea of running simulations forward in time for the demography and then
    backward in time for the neutral genetics is by far not new and was already
    implemented in the software SPLATCHE (Ray, Currat, Foll, Excoffier,
    Bioinformatics 2010, 26: 2993-2994), although probably not as efficiently. The
    reference should be added.
\end{point}

\reply{
    We are assuming that you are drawing the analogy of tree sequence simplification 
    to the forward-backward scheme used by SPLATCHE.
    This is a good analogy, since simplification is tracing ancestry back through time
    as does the coalescent.
    However, we think it's fair to say that it differs fundamentally:
    the coalescent is a stochastic process that is dual to the forwards-time process,
    and relies on certain assumptions for correctness;
    whereas tree sequence simplification is deterministic (just ``bookkeeping'').
    We have added something about SPLATCHE2 to the introduction \revref.
}

\begin{point}{}
  What is the use of gene trees? What can be inferred/calculated from them that
  we cannot get from the endpoint genetic polymorphism data? (An example comes
very late in the ms, should be moved up.)
\end{point}

\reply{
    We agree this makes nice motivation,
    but think it is somewhat outside of the scope of this paper:
    for our purposes here, it suffices to explain that they help everything go faster.
    We've expanded a bit \revref.
}

\begin{point}{}
  Spell out once what is meant with $\theta = 4N_e\mu$ (and make clear that $\mu$ is
  the num loci $\times$ mutation rate)
\end{point}

\reply{
    We agree this was out of place; but rather than explain, we have removed this. \revref
}

\begin{point}{}
  Make clear how to transfer from haploid to diploid case (I guess by having
  about twice the data recorded).
\end{point}

\reply{
    We've tried to make this more clear. \revref
}

\begin{point}{}
  Figure 3 doesn't show the red stars mentioned in the main text (p8).
\end{point}

\reply{
    These are visible for us; there may have been a problem with PDF transparency;
    we've set opacity on these to 100\%; let us know if they are still invisible?
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    This manuscript presents algorithms that could very well change the way
    population genetic simulations are done. The current methods are either
    coalescent simulations (which build trees backwards in time, but for all
    intents and purposes do not allow selection) or forward simulations (which are
    as flexible as the user desires, but comes at the cost of being slow). The
    algorithms here are super creative solutions that allow the user to simulate a
    sequence tree forward in time. This approach can dramatically shrink the
    computational time needed to do large-scale simulations. The API proposed goes
    one step further. It is based on the msprime architecture, so a user could
    ostensibly seamlessly switch from coalescent to forward simulations very easily
    (or stitch the two together to generate ancient history using the coalescent
    and then forward simulate based on the output). This really could be a game
    changing paper for the population genetics world (as well as possibly others,
    as noted in the manuscript).

    I've thought quite a bit about how to record trees in a forward simulation, and
    could only come up with ways that drastically exploded the run time. The
    approach presented here is very creative.

    In addition, the use of the compressed format of the sequence tree is quite
    amazing. The 144k over the VCF equivalent is unbelievable.
\end{quote}

Thanks!

\begin{point}{}
    I only see one major criticism: accessibility. The authors do not provide a
    clear overview of how to obtain/install/use the algorithms presented. A
    ``proof-of-concept implementation'' is really not sufficient. In contrast,
    msprime has great documentation, but the current website does not seem to
    describe the forward simulation model (in contrast to what is suggested by the
    manuscript). \url{https://github.com/ashander/ftprime} does have installation
    instructions, but seems to be entirely geared toward integration with simuPOP.
    Most of the figures are generated with fwdpp, which are supposedly included
    here (\url{https://github.com/petrelharp/ftprime_ms}), but it is not at all clear how
    one would access the code given the current organization of this repository. It
    is unrealistic to expect all downstream users to code up the algorithms
    themselves (and would result in code that is not as well tested). The authors
    should have a more dedicated repository for users to access the software and
    start running their own simulations.
\end{point}

\reply{
    As the reviewer suggests, our main goal here is to provide easy access
    to the \emph{algorithm}, not to a forward simulation model --
    we aren't describing a forward simulation model, 
    but rather a way to speed up and get more information out of an existing one.
    But, is the algorithm accessible?
    An interface to our algorithms is available at the python level or at the C level.
    The python interface is documented through the \msprime{} documentation
    (but you need to select ``latest'': \url{https://msprime.readthedocs.io/en/latest/});
    and closely mirrors the underlying C library.
    We have also written documentation at \plrXXX LINK TO TUTORIALS}.
}

\begin{point}{}
Figure 1 shows that pedigree recording seems to scale poorly in N. Going from
N=1e3 to N=1e4 results in what seems like a $\sim 5 \times$ increase in runtime, but going
up to N=5e4 results in a $> 10 \times$ increase in runtime. Given your attention to
scaling, is this expected?
\end{point}

\reply{
    This is actually the fault of a quadratic algorithm for counting mutations
    under the hood in \fwdpp{}.
    In fact, some of us (KRT) are currently using ideas from tree sequences to fix this bottleneck.
}

%%%  Minor comments:

\begin{point}{}
The word "thanks” in the discussion of coalescent caveats is awkward.
\end{point}

\reply{
    Thanks; fixed. \revref
}

\begin{point}{}
The Appendices seem to be out of order with the text. The current order does
not seem any more logical.
\end{point}

\reply{
    Good point. Fixed.
}

\begin{point}{}
In Figure 1, it would potentially be helpful to see y-axis on a log scale so
that we could see scaling on the low end. Legend for this figure refers to
``right column of panels''. How many threads were used to obtain the dashed line?
\end{point}

\reply{
    Thanks for the suggestions.
    \plr{TODO: I've stuck in the "log" figure. What do you think?}
    \plr{Kevin: how many threads?}
}

\begin{point}{}
"its'" on page 11.
\end{point}

\reply{
    Thanks. Fixed, in two places.
}

\begin{point}{}
I appreciate the ``how does it perform in theory?'' section, but the final note
on the use of ``nedigree'' seems unnecessary. If you are going to use it (which I
do not recommend), you can justify it, otherwise drop it.
\end{point}

\reply{
    We have removed this section.
}

\begin{point}{}
What characteristics of each simulated mutation are stored? On page 10, it
suggests you store the node in which they first occur, the derived state, and
the genomic position. However, the selection coefficient and dominance
coefficient? Are there other features that can be stored (e.g., nucleotide
context)?
\end{point}

\reply{
    Good question;
    the answer is that anything can be stored (as arbitrary bytes);
    we've added a note about this. \revref
}

\begin{point}{}
The code for the Python implementation of simplify is included as supplementary
information. This is entirely unnecessary, as this code should be easily
accessible from the appropriate github repository.
\end{point}

\reply{
    Perhaps, but since github does not provide a permanent home,
    and a working implementation is, arguably, the most clear explanation of an algorithm,
    we'd prefer to keep this in as supplementary material.
}

### 2018-06-03 forwards_paper.tex
responses

@@
 \usepackage[utf8]{inputenc}
-\usepackage{natbib}
 \usepackage{graphicx}
@@
 \usepackage{listings}
-\bibliographystyle{plainnat}
 % \usepackage[T1]{fontenc}
@@
 \newcommand{\msprime}{\texttt{msprime}}
+\newcommand{\tskit}{\texttt{tskit}}
 \newcommand{\ftprime}{\texttt{ftprime}}
@@
 
+%% for submitting with responses to reviews
+\newif\ifsubmission
+% \submissiontrue
+\submissionfalse
+\ifsubmission
+    % % stuff for submission
+    \usepackage[nofiglist,nomarkers,figuresonly]{endfloat}
+    \usepackage{setspace}
+    % \doublespacing
+    \bibliographystyle{plos2015}
+    \newcommand\citet{\cite}
+    \newcommand\citep{\cite}
+    \renewcommand\includegraphics{\relax}
+    \renewcommand{\includegraphics}[2][]{\relax}
+    \renewcommand{\figurename}{Fig.}
+    \newcommand{\Figure}{Fig }
+    \newcommand{\Figures}{Fig }
+\else
+    % for the nice version
+    \usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
+    \bibliographystyle{plainnat}
+    \newcommand{\Figure}{{Figure }}
+    \newcommand{\Figures}{{Figures }}
+\fi
+% grrr
+\newcommand{\citetx}[2][]{\ifsubmission#1 \cite{#2}\else\citet{#2}\fi}
+
+% Responses to reviews
+\input{review-response-commands}
+% set this to show line numbers and include responses to reviews or not
+\newif\ifreviewresponses
+\reviewresponsestrue  % include them
+% \reviewresponsesfalse  % don't include them
+\newcommand{\responsefile}{review-responses}  % name of the review responses file
+
 \begin{document}
@@
 \begin{abstract}
-    In this paper we describe how to
-    efficiently record the entire genetic history of a population
-    during a forwards-time, individual-based population genetics simulation.
-    This dramatically reduces the computational burden of tracking individual genomes
-    by simulating only those loci that may affect reproduction (those having non-neutral variants),
-    and recording history as a \emph{succinct tree sequence} as introduced in the software package \msprime,
+    In this paper we describe how to efficiently record the entire genetic history of a population
+    in forwards-time, individual-based population genetics simulations with arbitrary breeding models,
+    population structure and demography.
+    This approach dramatically reduces the computational burden of tracking individual genomes
+    by allowing us to simulate only those loci that may affect reproduction
+    (those having non-neutral variants).
+    The genetic history of the population is recorded as a succinct tree sequence
+    as introduced in the software package \msprime,
     on which neutral mutations can be quickly placed afterwards.
-    We implement the method in two popular forwards-time simulation frameworks,
-    which speeds up large, whole-genome simulations by one to two orders of magnitude.
-    In addition to speed, the method has several advantages:
+    Recording the results of each breeding event
+    results in a linear growth in the amount of information that must be stored.
+    However, there is a great deal of redundancy in this information,
+    since the majority of individuals simulated over the course of the simulation
+    will leave no descendents in the eventual sample.
+    We solve this storage problem by providing an algorithm to quickly `simplify'
+    a tree sequence by removing this irrelevant history for a given set of genomes.
+    By periodically simplifying the history with respect to the extant population,
+    we show that the total storage space required is modest
+    and overall large efficiency gains can be made over classical forward-time simulations.
+    We implement a general-purpose framework for recording and simplifying genealogical data,
+    which can be used to make simulations of any population model more efficient.
+    We modify two popular forwards-time simulation frameworks to use this new approach
+    and observe efficiency gains in large, whole-genome simulations of one to two orders of magnitude.
+    In addition to speed, our method for recording pedigrees has several advantages:
     (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
     (2) A population of $N$ individuals with $M$ polymorphic sites
-    can be stored in $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation (and its' history)
-    rather than only a subset.
-    (3) A simulation can easily be initialized with a more efficient coalescent simulation of deep history.
-    To make these possible,
-    we introduce the \msprime{} Tables API, which allows for the efficient interchange
-    of tree sequences between modular program components,
-    and provide an algorithm to quickly simplify a tree sequence by removing history irrelevant to a given set of genomes.
+        can be stored in $O(N log N + M)$ space,
+        making it feasible to store a simulation's entire final generation
+        (and its history) rather than only a subset.
+    (3) A simulation can easily be initialized with a more efficient coalescent simulation 
+        of deep history.
 \end{abstract}
@@
 Coalescent theory thus had three transformative effects on population genetics:
-first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
-second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
-and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
+first, giving researchers better conceptual tools to describe \emph{gene trees}
+and thus bringing within-population trees into better focus;
+second,
+producing analytical methods to estimate parameters of interest from genetic data; \revpoint{1}{5}
+and finally, providing a computationally feasible method
+to produce computer simulations of population genetics processes.
 However, these powerful advances came with substantial caveats:
@@
 are only \emph{Markovian}, and thus feasible to work with,
-thanks to the important assumptions of
+under the following important assumptions: \revpoint{2}{3}
 (a) random mating,
 (b) neutrality,
-and (c) small sample size relative to the population size.
+(c) large population size,
+and (d) small sample size relative to the population size.
 The first two assumptions can be side-stepped to a limited extent \citep{hudson1990gene, Neuhauser1997-nn},
@@
 and/or have large numbers of loci under various sorts of selection.
-Furthermore, the relationship between the life history of a species --
+Usually, the relationship between the life history of a species --
 fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
 are all absorbed into a single compound parameter, the coalescence rate.
-The third assumption is no longer safe, either --
+More mechanistic models are possible using ``forwards--backwards'' simulations,
+that first simulate population size changes forwards in time
+and then thread a coalescent backwards \citep{ray2010splatche2}, \revpoint{1}{3}
+but these still require the assumptions above to be met for each subpopulation.
+The last assumption is no longer safe, either --
 for example, a recent study~\citep{martin2017human}
@@
 (and hence, genome sequence) as well as an algorithm for simplifying these.
-The data structure is \emph{succinct} in the sense that its' space usage is close to optimal,
-while still allowing efficient retrieval of information (see, e.g.,
-\citet{gog2014theory}).
-The storage method and simplification algorithm are both implemented
-as improvements to the algorithmic tools and data structures in \msprime{}.
+The data structure is \emph{succinct} in the sense that its space usage is close to optimal,
+while still allowing efficient retrieval of information (see, e.g., \citet{gog2014theory}).
 We also describe how these tools can efficiently record,
 and later process, the embellished population pedigree from a forwards-time simulation.
-While providing substantial savings in computational time and space, our methods provide much
-more information than simply simulating the genomes -- the tree sequence
+While providing substantial savings in computational time and space, our methods provide 
+in principle much more information than simply simulating the genomes -- the tree sequence
 encodes all marginal genealogies of individuals living at the end of the simulation.
+These marginal genealogies enable fast data storage and processing,
+but also provide additional information that can be used to better understand
+the notoriously complex dynamics of population genetics. \revpoint{1}{4}
 Although we were motivated by a need for more efficient genomic simulations,
 these tools may prove more widely useful.
+This work originated as improvements to the algorithmic tools and data structures 
+in the coalescent simulator \msprime{}; \linelabel{msprime_rel}
+the tools are currently available bundled with the Python package \msprime{},
+but will soon be separately available as a Python package named \tskit{}.
 
@@
 \begin{figure}
-    \includegraphics[]{sims/rawspeed}
+    \includegraphics[]{sims/rawspeed_logy}
     \caption{
@@
     The left figure shows run times for standard simulations including neutral mutations.
-    The right column of panels shows run times of simulations that recorded the pedigree
+    The right column shows run times of simulations that recorded the pedigree
     and added neutral mutations afterwards.
@@
 For simplicity, we sample exactly one crossover per generation.
+Note that the \emph{table recording} portion of the algorithm does not depend on the Wright--Fisher
+nature of the population simulation;  \linelabel{ll:not_wf}
+next we will describe how to record tables from \emph{any} simulation.
 
@@
 \paragraph{Essential steps in tree sequence recording}
-To record a tree sequence, including mutations,
-a simulator must record:
+Since a tree sequence can record the history of genetic inheritance in any situation
+(requiring only unambiguous inheritance and no time travel),
+any individual-based population genetics simulator can maintain a tree sequence
+with only a little bookkeeping. \revpoint{1}{1}
+We have furthermore provided several tools to minimize this bookkeeping,
+only requiring a little one-way output at the birth of each new individual.
+Concretely, to record a tree sequence, including mutations,
+a simulator must record for each new genome
+(so, twice for each new diploid individual): \revpoint{1}{6}
 \begin{itemize}
-    \item the birth time of each individual in the Node Table,
-    \item the genetic segments each individual inherits from its parents in the Edge Table, and
-    \item the individual each new mutation occurs in, and the derived state in the Mutation Table,
-        adding also to the Site Table if the site is previously unmutated.
+    \item the birth time of the genome in the Node Table,
+    \item the segments the genome inherits from its parental genomes in the Edge Table,
+    \item the locations of any new mutations in the Site Table,
+    \item and the derived state of these new mutations in the Mutation Table
+        (as well as the identiy of this genome the mutations appeared in).
 \end{itemize}
-Recall that here we mean ``\emph{haploid} individual'', i.e., gamete.
-Before simplification, time must be translated to ``time ago'' if it is not already.
+Each of these can be simply appended to the ends of the respective tables.
+Besides this, simplification should be run every once in a while (e.g., every 100 generations).
+Before simplification, time in the Node Table must be translated to ``time ago'' if it is not already.
 (This was avoided in Algorithm~\algref{W} since there $t$ denoted
 ``time until the end of the simulation''.)
-Furthermore, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
+There are also several ``cleaning'' steps, for which we provide functions in the Tables API:
+\emph{sorting} according to several criteria for algorithmic efficiency;
+and removing any duplicate sites from the Site Table.
+After simplification, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
 being relabeled in \msprime{}'s tables as $0, 1, \ldots, N-1$,
@@
 with the node IDs recorded in the tables.
-% Finally, the first step in simplification is to \emph{sort} the tables
-% according to several criteria, for algorithmic efficiency,
-% as described in the documentation.
+
+In other words, \linelabel{ll:other_words}
+to record a tree sequence, a simulation needs only to know
+(a) which genomes recombined to produce each new genome, and how;
+(b) the locations and results of any new mutations on each genome;
+and (c) the identities of every currently alive individual at each time simplification occurs.
+
+\paragraph{Storing metadata}
+Applications may also want to store more information not fitting into an existing column of the tables,
+such as the selection coefficient of a mutation, or the sex of an individual.
+This (and, indeed, arbitrary information) can be stored in the \texttt{metadata} columns
+present in Node, Site, and Mutation tables. \revpoint{2}{8}
 
@@
 
-Our approach to simplification is based on Hudson's algorithm for simulating
-the coalescent with recombination~\citep{hudson1983properties},
-paralleling the implementation in \msprime{} \citep{kelleher2016efficient};
-an implementation in pseudocode is provided in Appendix~\ref{ss:simplify_algorithm},
+We implement simplification by starting at the end of the simulation,
+and moving back up through history,
+recording in the new tree sequence only that information necessary to construct the tree sequence
+of the specified individuals.
+This process of tracing ancestry back through time in a pedigree
+was the motivation for Hudson's coalescent simulation algorithm~\citep{hudson1983properties},
+so it is unsurprising that simplification uses many of the same tools
+as the implementation of Hudson's algorithm in \msprime{} \citep{kelleher2016efficient}.
+The main difference is that events in a coalescent simulation are random, \revpoint{1}{2}
+while here they are predetermined by history.
+An implementation in pseudocode is provided in Appendix~\ref{ss:simplify_algorithm},
 and a python implementation as supplementary information.
+
 Conceptually, this works by
@@
 (b) moving back through history,
-copying the colors of each chromosome to the portions of its' parental chromosomes
+copying the colors of each chromosome to the portions of its parental chromosomes
 from which it was inherited;
@@
 of the Wright--Fisher model.
-For an introduction to coalescent theory, see \citet{wakeley2005coalescent}.
+The arguments below are mostly self-contained,
+but for an introduction to coalescent theory, including the basic facts used below,
+see \citet{wakeley2005coalescent}.
+\linelabel{ll:coal_signpost}
 
@@
 
-\paragraph{A final note:}
-in preparing this manuscript,
-we debated a number of possible terms for the embellished pedigree,
-i.e., the ``pedigree with ancestral recombination information''.
-Etymological consensus \citep{liberman2014little} has
-``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
-(whose branching pattern resembles the bifurcation of a single parent-offspring relationship).
-An analogous term for the embellished pedigree might then be \emph{nedigree},
-from ``nid de grue'',
-as the nest of a crane is a large jumble of (forking) branches.
-We thought it unwise to use this term throughout the manuscript,
-but perhaps it will prove useful elsewhere.
+% \paragraph{A final note:}
+% in preparing this manuscript,
+% we debated a number of possible terms for the embellished pedigree,
+% i.e., the ``pedigree with ancestral recombination information''.
+% Etymological consensus \citep{liberman2014little} has
+% ``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
+% (whose branching pattern resembles the bifurcation of a single parent-offspring relationship).
+% An analogous term for the embellished pedigree might then be \emph{nedigree},
+% from ``nid de grue'',
+% as the nest of a crane is a large jumble of (forking) branches.
+% We thought it unwise to use this term throughout the manuscript,
+% but perhaps it will prove useful elsewhere.
 
@@
 \appendix
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Timing and memory use with no selection}
+\label{ss:timing_nosel}
+\renewcommand{\thefigure}{D\arabic{figure}}
+\setcounter{figure}{0}
+
+We also performed a more limited set of simulations without natural selection.
+For these we used \cpp{} as in the main text, but also provide results from our
+proof-of-concept implementation that uses \simupop{} (see
+Appendix~\ref{ss:simupop}).
+
+The total run times are shown in
+Figure~\ref{sfig:rawspeed_nosel} and for \cpp{} show the same qualitative behavior as simulations with selection
+(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to pedigree tracking is again substantial in these
+simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to pedigree tracking here with \fwdpy{}
+than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
+close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
+more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
+and~\ref{sfig:speedup_nosel}.
+For \simupop{}, the results show a qualitatively similar pattern but (1) the
+simulations for $N = 10000$ or greater did no finish in the 72 hour limit or
+exceed memory limits (more than 500 GB), and (2) the magnitude of the speedup is much less.
+The total RAM use for
+simulations without selection is shown in Figure~\ref{sfig:ramnosel}.
+% Pedigree recording again used more RAM with \fwdpp{} and less
+% RAM with \simupop{}. The reason for reduced RAM use with \simupop{} is unclear;
+% with \fwdpy{}/\fwdpp{}, however the reason for the extra RAM use is as in
+% Appendix~\ref{ss:memuse}.
+
+\begin{figure}[!h]
+    \includegraphics[]{sims/rawspeed_nosel}
+    \caption{\label{sfig:rawspeed_nosel}Total run time for a single simulation replicate as a function of region
+        length, measured as the scaled recombination parameter $\rho = 4Nr$.  The simulations here involve no natural
+    selection.}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[]{sims/speedup_nosel}
+    \caption{\label{sfig:speedup_nosel}Relative performance improvement due to pedigree tracking for simulations without
+    selection.  Data points are taken from Figure~\ref{sfig:rawspeed_nosel} and show the ratio of run times tracking
+neutral variation over the run times tracking the pedigree.}
+\end{figure}
+\begin{figure}
+    \includegraphics[]{sims/memuse_nosel}
+    \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
+    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Memory usage with selection}
+\label{ss:memuse}
+\renewcommand{\thefigure}{B\arabic{figure}}
+\setcounter{figure}{0}
+
+The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel}.  For our parameters, which involved simplifying every
+$10^3$ generations, simulation while recording pedigrees uses more RAM than a standard simulation.
+This is because with pedigree recording, a non-recombinant
+individual still requires one new edge and two new nodes to be allocated.  Without pedigree recording, no new memory is allocated for such an offspring by \fwdpp{}.
+Thus, pedigree recording accumulates a large amount of data in RAM until
+simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extra RAM needed
+can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
+require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four sets of unsimplified data are
+stored in RAM.  In practice, the RAM consumption can be minimized by simplifying more often.
+
+%% \krt{I've avoided getting into how this much RAM means parallelization is required to make more progress.}
+
+\begin{figure}
+    \includegraphics[]{sims/memuse}
+    \caption{\label{sfig:ramsel}Peak RAM use in gigabytes (GB) for simulations with selection.  The plots are from the
+        same data as in Figure \ref{fig:runtimes_selection}. Note that the y-axis scales differ considerably across
+    panels.}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%
+\section{Effect of simplification interval on run time and memory use}
+\label{ss:gcinterval}
+\renewcommand{\thefigure}{C\arabic{figure}}
+
+We performed a limited set of simulations to explore the effect of the simplification interval on run times and memory
+use. Figure~\ref{sfig:gctime} shows that run times increase for simplification intervals less than 100 generations.
+For intervals longer than 100 generations, run times are very similar, with no appreciable difference between $10^3$ and $10^4$
+generations.  As expected, the simplification interval has a near-linear effect on peak RAM use
+(Figure~\ref{sfig:gcmem}).  Taken together Figures~\ref{sfig:gctime} and~\ref{sfig:gcmem} suggest that one can tune the
+simplification interval to available RAM with little effect on run times provided that simplification doesn't occur too
+often.
+
+\begin{figure}[!h]
+    \includegraphics{sims/gc_interval/GCtime}
+    \caption{\label{sfig:gctime}The effect of simplification interval on run times.  A single replicate was done for the
+        parameters shown in the figure.  These simulations were done with the same selection parameters as in
+    Figure~\ref{fig:runtimes_selection}.}
+\end{figure}
+
+\begin{figure}[!h]
+    \includegraphics{sims/gc_interval/GCmem}
+    \caption{\label{sfig:gcmem}Peak RAM use for the simulations shown in Figure~\ref{sfig:gctime}.}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%
+\section{The \simupop{} implementation}
+\label{ss:simupop}
+
+To use the methods with \simupop{}, we created a python module
+(available at \url{https://github.com/ashander/ftprime})
+that manages communication between the forward simulator and \msprime{}.
+We implemented simulations in a Python script, usable as a command-line utility,
+taking input parameters for population size $N$ and scaled recombination rate $\rho$.
+As for the \fwdpp{} simulations above, we set scaled mutation rate $\theta=4N\mu=\rho$
+(where $\mu$ is the per-site mutation rate multiplied by the number of sites),
+and used a simplify interval of 1000 generations and simulated for $T=10N$ generations in each
+simulation.
+% introducing deleterious mutations at a rate of $\rho/100$ per generation.
+% Because \simupop{} requires explicit locations for loci, we set the number of
+% loci $n=\rho / (4Nr)$ and for these simulations without selection marked $0$ of
+% these as selected loci.
+
+We coupled \simupop{} to \msprime{} using a Python class \texttt{RecombCollector} that 1)
+tracks time, and 2) collects and parses recombination information provided by
+the \texttt{Recombinator} class of \simupop{}, adding them to a \tnodetable{} and
+\tedgetable{} upon which simplification is performed every 1000 generations (using
+\texttt{msprime.sort\_tables()} and \texttt{msprime.simplify\_tables}).
+The tables are initialized using the \texttt{TreeSequence} returned from
+\texttt{msprime.simulate($2N$, Ne=$N$, recombination\_rate=$r/2$, length=$n$)}.
+The \texttt{simuPOP.Population()} is initialized and tagged with individual IDs,
+a mapping of these IDs to the \tnodetable{} IDs is passed to
+an instance of \texttt{RecombCollector} and used internally to update the tables.
+The \texttt{simuPOP.Population()} is then evolved.
+Before mating:
+% selected mutations are created (using
+% \texttt{simuPOP.SNPMutator}) and multiplicative selection occurs (using
+% \texttt{simuPOP.PyMlSelector}) only on the selected loci;
+the internal time of the \texttt{RecombCollector} instance is updated.
+During mating: male and female parents are drawn randomly, recombination data is
+output by \simupop{} to the \texttt{RecombCollector} instance.
+After mating: every 1000 generations the tables are simplified.
+
+\texttt{RecombCollector} uses the Python class \texttt{ARGrecorder} internally
+to add data to the tables and perform simplification.
+This class contains the \tnodetable{} and \tedgetable{} and also the mapping from
+(haploid) ids to ids in the tables.
+It also provides methods (\texttt{add\_individual} and \texttt{add\_record})
+to append data to the tables.
+The only functions of
+\texttt{RecombCollector} not delegated to this internal class are computing
+haploid IDs from \simupop{}'s diploid IDs and parsing recombination data into
+data suitable for use with use with \texttt{add\_individual} and
+\texttt{add\_record}; these functions are used to stream data into the tables as
+it is received by an instance of \texttt{RecombCollector}.
+
+% \paragraph{Requirements.}
+% The simulations as implemented depend on the class \texttt{RecombCollector} provided by
+% the Python package \ftprime{} 0.0.6rc0,
+% available at \url{https://www.github.com/ashander/ftprime}.
+% Its requirements are encoded in an \texttt{environment.yml} file, compatible
+% with conda, but \msprime{} must be updated to the version listed below via pip
+% from GitHub \jda{TODO add command?}.
+%
+% In brief the requirements are:
+%
+% \begin{enumerate}
+%     \item \msprime{} 0.5.0
+%     \item \simupop{} 1.1.8.3
+%     \item python 3.5
+% \end{enumerate}
 
@@
 
-\section{Memory usage with selection}
-\label{ss:memuse}
-\renewcommand{\thefigure}{B\arabic{figure}}
-\setcounter{figure}{0}
-
-The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel}.  For our parameters, which involved simplifying every
-$10^3$ generations, simulation while recording pedigrees uses more RAM than a standard simulation.
-This is because with pedigree recording, a non-recombinant
-individual still requires one new edge and two new nodes to be allocated.  Without pedigree recording, no new memory is allocated for such an offspring by \fwdpp{}.
-Thus, pedigree recording accumulates a large amount of data in RAM until
-simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extra RAM needed
-can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
-require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four sets of unsimplified data are
-stored in RAM.  In practice, the RAM consumption can be minimized by simplifying more often.
-
-%% \krt{I've avoided getting into how this much RAM means parallelization is required to make more progress.}
-
-\begin{figure}
-    \includegraphics[]{sims/memuse}
-    \caption{\label{sfig:ramsel}Peak RAM use in gigabytes (GB) for simulations with selection.  The plots are from the
-        same data as in Figure \ref{fig:runtimes_selection}. Note that the y-axis scales differ considerably across
-    panels.}
-\end{figure}
-
+% \newpage
 \newpage
 
-\section{Effect of simplification interval on run time and memory use}
-\label{ss:gcinterval}
-\renewcommand{\thefigure}{C\arabic{figure}}
-
-We performed a limited set of simulations to explore the effect of the simplification interval on run times and memory
-use. Figure~\ref{sfig:gctime} shows that run times increase for simplification intervals less than 100 generations.
-For intervals longer than 100 generations, run times are very similar, with no appreciable difference between $10^3$ and $10^4$
-generations.  As expected, the simplification interval has a near-linear effect on peak RAM use
-(Figure~\ref{sfig:gcmem}).  Taken together Figures~\ref{sfig:gctime} and~\ref{sfig:gcmem} suggest that one can tune the
-simplification interval to available RAM with little effect on run times provided that simplification doesn't occur too
-often.
-
-\begin{figure}[!h]
-    \includegraphics{sims/gc_interval/GCtime}
-    \caption{\label{sfig:gctime}The effect of simplification interval on run times.  A single replicate was done for the
-        parameters shown in the figure.  These simulations were done with the same selection parameters as in
-    Figure~\ref{fig:runtimes_selection}.}
-\end{figure}
-
-\begin{figure}[!h]
-    \includegraphics{sims/gc_interval/GCmem}
-    \caption{\label{sfig:gcmem}Peak RAM use for the simulations shown in Figure~\ref{sfig:gctime}.}
-\end{figure}
-
-\section{Timing and memory use with no selection}
-\label{ss:timing_nosel}
-\renewcommand{\thefigure}{D\arabic{figure}}
-\setcounter{figure}{0}
-
-We also performed a more limited set of simulations without natural selection.
-For these we used \cpp{} as in the main text, but also provide results from our
-proof-of-concept implementation that uses \simupop{} (see
-Appendix~\ref{ss:simupop}).
-
-The total run times are shown in
-Figure~\ref{sfig:rawspeed_nosel} and for \cpp{} show the same qualitative behavior as simulations with selection
-(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to pedigree tracking is again substantial in these
-simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to pedigree tracking here with \fwdpy{}
-than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
-close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
-more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
-and~\ref{sfig:speedup_nosel}.
-For \simupop{}, the results show a qualitatively similar pattern but (1) the
-simulations for $N = 10000$ or greater did no finish in the 72 hour limit or
-exceed memory limits (more than 500 GB), and (2) the magnitude of the speedup is much less.
-The total RAM use for
-simulations without selection is shown in Figure~\ref{sfig:ramnosel}.
-% Pedigree recording again used more RAM with \fwdpp{} and less
-% RAM with \simupop{}. The reason for reduced RAM use with \simupop{} is unclear;
-% with \fwdpy{}/\fwdpp{}, however the reason for the extra RAM use is as in
-% Appendix~\ref{ss:memuse}.
-
-\begin{figure}[!h]
-    \includegraphics[]{sims/rawspeed_nosel}
-    \caption{\label{sfig:rawspeed_nosel}Total run time for a single simulation replicate as a function of region
-        length, measured as the scaled recombination parameter $\rho = 4Nr$.  The simulations here involve no natural
-    selection.}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[]{sims/speedup_nosel}
-    \caption{\label{sfig:speedup_nosel}Relative performance improvement due to pedigree tracking for simulations without
-    selection.  Data points are taken from Figure~\ref{sfig:rawspeed_nosel} and show the ratio of run times tracking
-neutral variation over the run times tracking the pedigree.}
-\end{figure}
-\begin{figure}
-    \includegraphics[]{sims/memuse_nosel}
-    \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
-    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
-\end{figure}
-
-\section{The \simupop{} implementation}
-\label{ss:simupop}
-
-To use the methods with \simupop{}, we created a python module
-(available at \url{https://github.com/ashander/ftprime})
-that manages communication between the forward simulator and \msprime{}.
-We implemented simulations in a Python script, usable as a command-line utility,
-taking input parameters for population size $N$ and scaled recombination rate $\rho$.
-As for the \fwdpp{} simulations above, we set scaled mutation rate $\theta=\rho$,
-used a simplify interval of 1000 generations and simulated for $T=10N$ generations in each
-simulation.
-% introducing deleterious mutations at a rate of $\rho/100$ per generation.
-% Because \simupop{} requires explicit locations for loci, we set the number of
-% loci $n=\rho / (4Nr)$ and for these simulations without selection marked $0$ of
-% these as selected loci.
-
-We coupled \simupop{} to \msprime{} using a Python class \texttt{RecombCollector} that 1)
-tracks time, and 2) collects and parses recombination information provided by
-the \texttt{Recombinator} class of \simupop{}, adding them to a \tnodetable{} and
-\tedgetable{} upon which simplification is performed every 1000 generations (using
-\texttt{msprime.sort\_tables()} and \texttt{msprime.simplify\_tables}).
-The tables are initialized using the \texttt{TreeSequence} returned from
-\texttt{msprime.simulate($2N$, Ne=$N$, recombination\_rate=$r/2$, length=$n$)}.
-The \texttt{simuPOP.Population()} is initialized and tagged with individual IDs,
-a mapping of these IDs to the \tnodetable{} IDs is passed to
-an instance of \texttt{RecombCollector} and used internally to update the tables.
-The \texttt{simuPOP.Population()} is then evolved.
-Before mating:
-% selected mutations are created (using
-% \texttt{simuPOP.SNPMutator}) and multiplicative selection occurs (using
-% \texttt{simuPOP.PyMlSelector}) only on the selected loci;
-the internal time of the \texttt{RecombCollector} instance is updated.
-During mating: male and female parents are drawn randomly, recombination data is
-output by \simupop{} to the \texttt{RecombCollector} instance.
-After mating: every 1000 generations the tables are simplified.
-
-\texttt{RecombCollector} uses the Python class \texttt{ARGrecorder} internally
-to add data to the tables and perform simplification.
-This class contains the \tnodetable{} and \tedgetable{} and also the mapping from
-(haploid) ids to ids in the tables.
-It also provides methods (\texttt{add\_individual} and \texttt{add\_record})
-to append data to the tables.
-The only functions of
-\texttt{RecombCollector} not delegated to this internal class are computing
-haploid IDs from \simupop{}'s diploid IDs and parsing recombination data into
-data suitable for use with use with \texttt{add\_individual} and
-\texttt{add\_record}; these functions are used to stream data into the tables as
-it is received by an instance of \texttt{RecombCollector}.
-
-% \paragraph{Requirements.}
-% The simulations as implemented depend on the class \texttt{RecombCollector} provided by
-% the Python package \ftprime{} 0.0.6rc0,
-% available at \url{https://www.github.com/ashander/ftprime}.
-% Its requirements are encoded in an \texttt{environment.yml} file, compatible
-% with conda, but \msprime{} must be updated to the version listed below via pip
-% from GitHub \jda{TODO add command?}.
-%
-% In brief the requirements are:
-%
-% \begin{enumerate}
-%     \item \msprime{} 0.5.0
-%     \item \simupop{} 1.1.8.3
-%     \item python 3.5
-% \end{enumerate}
-
-\newpage
-
 \section*{Python implementation of simplify (supplementary information)}
@@
 
+\ifsubmission\processdelayedfloats\fi
+
+% reviews here if toggled above
+\includereviews
+
 \end{document}

### 2018-06-03 sims/quick_plot.py

@@ -67,6 +67,48 @@ for ax in (ax_fwdpp, ax_fwdpp_arg):
 fig.tight_layout()
 plt.savefig("rawspeed.pdf")
 
+# same, on a log scale
+data = data[data['sel'] == True]
+data.sort_values(by='rho', inplace=True)
+groups = data.groupby(['engine', 'arg', 'queue', 'N'])
+fig, (ax_fwdpp, ax_fwdpp_arg) = plt.subplots(
+    1, 2, sharex=True, sharey=True)
+for name, group in groups:
+    lstyle = '-'
+    if name[0] == 'fwdpy11' and name[1] is False:
+        ax = ax_fwdpp
+    if name[0] == 'fwdpy11' and name[1] is True:
+        ax = ax_fwdpp_arg
+    m = points[name[3]]
+    if name[0] == 'simuPOP':
+        continue
+        ax = ax_simupop
+        if name[1] is True:
+            ax = ax_simupop_arg
+    mfacecolor = colors[name[3]]
+    if name[2] is True:
+        lstyle = 'dashed'
+        mfacecolor = 'none'
+    popsize = int(name[3])
+    ax.plot(group.rho, group.time / (60.**2.), marker=m,
+            ms=4, linestyle=lstyle, label=r'$N = {:.0e}$'.format(popsize),
+            color=colors[name[3]],
+            markerfacecolor=mfacecolor)
+
+ax_fwdpp.legend(loc='upper left',frameon=False)
+ax_fwdpp.set_title("With neutral mutations", fontsize='medium')
+ax_fwdpp_arg.set_title("With pedigree recording", fontsize='medium')
+ax_fwdpp.set_ylabel("Run time (hours)")
+ax_fwdpp.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
+ax_fwdpp_arg.set_xlabel('Scaled recombination rate (' + r'$\rho = 4Nr$)')
+ax_fwdpp_arg.set_xticks([1e3, 1e4, 1e5])
+for ax in (ax_fwdpp, ax_fwdpp_arg):
+    ax.set_xscale('log')
+    ax.set_yscale('log')
+fig.tight_layout()
+plt.savefig("rawspeed_logy.pdf")
+
+
 # Plot the relative speedup due to ARG tracking for the sims with selection
 
 data_arg = data[(data['arg'] == True) & (data['queue'] == False)].copy()

### 2018-06-06 review-responses.tex
update

@@
 
-We are pleased to submit a revision of our manuscript,
+We'd like to thank the reviewers for a careful reading of the paper;
+as detailed below, their helpful comments have resulted in a substantial amount of clarification
+about the overall structure and purpose of the paper.
+We hope you agree that our changes have addressed their comments and concerns.
 
@@
 We thank the reviewers for helpful feedback, and respond in detail to their points below.
-Something we did that should help address all these points
-was to make more clear the main purpose of this paper:
+In part to address all of these points,
+we have tried to clarify the main purpose of this paper, which is
 to describe algorithms and data structures for tree sequences,
 and how these are useful in simulation.
-Our main goal is not to write a simulation package.
-This is useful and important, because: 
+Our main goal is not to write a standalone simulation package.
+What we do is useful and important, because: 
 (a) we provide a fully functional and flexible, well-documented API
@@
 for recording the tree sequence.
-
-TODO: mention the tutorials?
+We'd also like to note that we are developing a set of tutorials,
+at \url{https://github.com/tskit-dev/tutorials}.
 
@@
     (like the \texttt{sort tables} functionality,
-    so that the simulator does not have to guarentee sortedness of the tables itself).
+    so that the simulator does not have to guarantee sortedness of the tables itself).
 }
@@
 \reply{
-    We are assuming that you are drawing the analogy of tree sequence simplification 
-    to the forward-backward scheme used by SPLATCHE.
-    This is a good analogy, since simplification is tracing ancestry back through time
-    as does the coalescent.
-    However, we think it's fair to say that it differs fundamentally:
+    Thanks for drawing our attention to the (good) analogy 
+    between the forward-backward scheme used by SPLATCHE
+    and the process of tree sequence simplification,
+    which we had not considered.
+    While there are indeed deep connections between this approach and tree sequence simplification, 
+    they are however fundamentally different:
     the coalescent is a stochastic process that is dual to the forwards-time process,
@@
     whereas tree sequence simplification is deterministic (just ``bookkeeping'').
-    We have added something about SPLATCHE2 to the introduction \revref.
+    We have updated the introduction to discuss this connection with SPLATCHE \revref,
+    and have discussed further the relationship of coalescence to simplification \revreffull{1}{2}.
 }
@@
 \reply{
-    We agree this makes nice motivation,
+    We agree that discussion of uses of gene trees would make nice motivation,
     but think it is somewhat outside of the scope of this paper:
     for our purposes here, it suffices to explain that they help everything go faster.
-    We've expanded a bit \revref.
+    We've expanded the discussion \revref.
 }
@@
     and closely mirrors the underlying C library.
-    We have also written documentation at \plrXXX LINK TO TUTORIALS}.
+    We have also provided a link to the  developing set of tutorials 
+    at \url{https://github.com/tskit-dev/tutorials} \llname{ll:tutorials}.
 }
@@
 \reply{
-    Thanks for the suggestions.
-    \plr{TODO: I've stuck in the "log" figure. What do you think?}
-    \plr{Kevin: how many threads?}
+    Thanks for the suggestions: we've switched the figure to a log scale,
+    and noted the number of threads (four).
+    The speedup is not that impressive for a variety of reasons, though, which
+    are hard to tease apart, so we haven't elaborated on this point.
+    (But, for the reviewer's interest: false sharing is one possible
+    issue, but perhaps $5 \times 10^4$ diploids may be too few to make
+    parallelization ``worth'' it when spawning new threads each generation.)
 }
@@
 \begin{point}{}
-"its'" on page 11.
+``its{'}'' on page 11.
 \end{point}

### 2018-06-06 forwards_paper.tex
update

@@
     (2) A population of $N$ individuals with $M$ polymorphic sites
-        can be stored in $O(N log N + M)$ space,
+        can be stored in $O(N \log N + M)$ space,
         making it feasible to store a simulation's entire final generation
@@
 the tools are currently available bundled with the Python package \msprime{},
-but will soon be separately available as a Python package named \tskit{}.
+but will soon be separately available as a package named \tskit{}, 
+which will include a Python API and an embeddable C library.
 
@@
     where the pedigree simplification steps were handled in a separate thread of execution
-    and fitness calculations were parallelized.
+    and fitness calculations were parallelized across four cores.
     Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
@@
 
-The \msprime\ Python Tables API provides a general purpose toolkit for importing
-and processing succinct tree sequences. Interoperation with Python simulators
-is then straightforward.  The implementation we benchmark here uses
-\texttt{pybind11} (\url{https://github.com/pybind/pybind11/}) to interface
-with the \fwdpp{} \cpp{} API \citep{fwdpp}. No modifications were
+The \msprime{} Python Tables API provides a general purpose toolkit for importing
+and processing succinct tree sequences, and a collection of tutorials are being developed
+at \url{https://github.com/tskit-dev/tutorials}. \linelabel{ll:tutorials}
+Interoperation with Python simulators is then straightforward.  
+The implementation we benchmark here uses \texttt{pybind11} 
+(\url{https://github.com/pybind/pybind11/}) to interface with the \fwdpp{} \cpp{} API \citep{fwdpp}. 
+No modifications were
 required to the \fwdpp{} code base; rather, we simply need to bookkeep parent/offspring labels,

### 2018-06-06 review-responses.tex
msprime -> tskit

@@
 We hope you agree that our changes have addressed their comments and concerns.
+
+There is one change of note that we should explain:
+since initial submission, we have decided to separate out the tools for dealing with tree sequences
+into a standalone package, called \tskit,
+from the coalescent simulation code (called \msprime).
+This should help avoid confusion.
+However, although we're a good ways towards this goal,
+we haven't yet completed the clean separation,
+so that currently the \tskit{} page instructs users to install \msprime{}.
+We think this slightly confusing situation is preferable to the alternative,
+which would have the terms in our paper outdated in a few months
+(when we move the tools discussed here
+from \msprime{} to \tskit{}).
 
@@
 \reply{
-    This is actually the fault of a quadratic algorithm for counting mutations
+    Good spot: this is actually the fault of a quadratic algorithm for counting mutations
     under the hood in \fwdpp{}.
-    In fact, some of us (KRT) are currently using ideas from tree sequences to fix this bottleneck.
+    In fact, some of us (KRT) have used ideas from tree sequences to fix this bottleneck,
+    and the fix will be part of the next \fwdpp{}.
 }

### 2018-06-06 forwards_paper.tex
msprime -> tskit

@@
         of deep history.
+    The software for recording and processing tree sequences is named \tskit.
 \end{abstract}
@@
 \fwdpp{} \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
-population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
+population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to around $5
 \times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
@@
 in the coalescent simulator \msprime{}; \linelabel{msprime_rel}
-the tools are currently available bundled with the Python package \msprime{},
-but will soon be separately available as a package named \tskit{}, 
-which will include a Python API and an embeddable C library.
+the software tools described here for working with tree sequences are referred to as \tskit{};
+they are currently bundled with the Python package \msprime{},
+but will soon be separately available as a Python API and an embeddable C library.
 
@@
 implemented using the forwards-time simulation library \fwdpp{} \citep{fwdpp}
-and tree sequence tools implemented in \msprime{}.
+and tree sequence tools implemented in \tskit{}.
 Then, we describe the conceptual and algorithmic foundations for the method:
-(a) a format, implemented in the \msprime{} Python API,
+(a) a format, implemented in the \tskit{} Python API,
 for recording tree sequences efficiently in several \emph{tables};
@@
 Pedigree tracking dramatically reduced runtimes, as shown in Figure~\ref{fig:runtimes_selection},
-producing a relative speedup of up to $\approx 50$ fold relative to standard simulations that track neutral mutations
+producing a relative speedup of up to around 50 fold relative to standard simulations that track neutral mutations
 (Figure~\ref{fig:relative_speedup_selection}).
@@
         The remaining panels show the specific encoding
-        of this tree sequence in the four tables (nodes, edges, sites and mutations)
-        defined by \msprime.
+        of this tree sequence in the four tables (nodes, edges, sites and mutations).
         \label{fig:example_tree_sequence}
@@
 described above) for this simulation consumed 157MiB of storage space. Using
-the \msprime\ Python API, the time required to load this file into memory was
+the \tskit{} Python API, the time required to load this file into memory was
 around 1.5 seconds, and the time required to iterate over all 1 million trees
@@
 there is only a \emph{single} tree.
-Checking this is more difficult, but is implemented in \msprime{}'s API.
-For efficiency, \msprime{} makes several other sortedness requirements on the tables,
+Checking this is more difficult, but is implemented in \tskit{}'s API.
+For efficiency, \tskit{} makes several other sortedness requirements on the tables,
 that are not necessarily satisfied by tables emitted by a forwards-time simulation.
-\msprime{}'s API includes tools to rectify this by first sorting % (using \texttt{sort\_tables})
+\tskit{}'s API includes tools to rectify this by first sorting % (using \texttt{sort\_tables})
 and then using the \texttt{simplify} algorithm described below, which works on sorted tables
-and is guaranteed to produce a valid, \msprime{}-ready tree sequence.
+and is guaranteed to produce a valid, \tskit{}-ready tree sequence.
 
 %%%%%%
-\subsection*{The \msprime\ Tables API}
+\subsection*{The \tskit{} Tables API}
 
 The facilities for working with succinct tree sequences are implemented as part
-of the \msprime\ Python API, which provides a powerful platform for processing
-tree topology and mutation data. The new portions of \msprime{} that we discuss
+of the \tskit{} Python API, which provides a powerful platform for processing
+tree topology and mutation data. The portions of \tskit{} that we discuss
 here are dedicated to tree sequence input and output using simple tables of
@@
 
-The \msprime{} Python Tables API provides a general purpose toolkit for importing
+The \tskit{} Python Tables API provides a general purpose toolkit for importing
 and processing succinct tree sequences, and a collection of tutorials are being developed
@@
 event to generate node and edge data. This information is then periodically copied
-to the \msprime\ Tables API, where it is sorted and simplified.
+to the \tskit{} Tables API, where it is sorted and simplified.
 
@@
 implementation that decouples forward simulation internals from transfer of data
-to \msprime, we also implemented a version of the simulations described in
+to \tskit, we also implemented a version of the simulations described in
 ``Simulation benchmarks'' separately in Python, described in Appendix~\ref{ss:simupop}.
@@
 After simplification, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
-being relabeled in \msprime{}'s tables as $0, 1, \ldots, N-1$,
+being relabeled in \tskit{}'s tables as $0, 1, \ldots, N-1$,
 there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
@@
 $t$ is the number of distinct trees.
-The \msprime{} API provides a method to compute $\pi$ among arbitrary subsets of the
+The \tskit{} API provides a method to compute $\pi$ among arbitrary subsets of the
 samples in a tree sequence, which took about 0.7 seconds when
@@
 or the coalescent within a known pedigree.  For such models, one
-could in principle generate tables of nodes and edges to be simplified in \msprime{}.  The resulting succinct tree
-sequence object would be in the same format as those generated by \msprime{}'s built-in simulate function, and therefore
+could in principle generate tables of nodes and edges to be simplified in \tskit{}.  The resulting succinct tree
+sequence object would be in the same format as those generated by \tskit{}'s built-in simulate function, and therefore
 compatible with existing methods for downstream analyses.
@@
 % are written in \justc{} or \cpp{} out of necessity. Therefore, the simplest and most
-% efficient approach would be to call the \msprime{} \justc{} library directly as part of
+% efficient approach would be to call the \tskit{} \justc{} library directly as part of
 % the core simulation logic, in the manner outlined in Algorithm \algref{W}. If designed
@@
 
-We implemented simulations and the connection to \msprime{} in \cpp{}, using \fwdpp{} library functions and
+We implemented simulations and the connection to \tskit{} in \cpp{}, using \fwdpp{} library functions and
 interface code using a continuum-sites model for both mutation
@@
 (available at \url{https://github.com/ashander/ftprime})
-that manages communication between the forward simulator and \msprime{}.
+that manages communication between the forward simulator and \tskit{}.
 We implemented simulations in a Python script, usable as a command-line utility,
@@
 
-We coupled \simupop{} to \msprime{} using a Python class \texttt{RecombCollector} that 1)
+We coupled \simupop{} to \tskit{} using a Python class \texttt{RecombCollector} that 1)
 tracks time, and 2) collects and parses recombination information provided by
@@
 \tedgetable{} upon which simplification is performed every 1000 generations (using
-\texttt{msprime.sort\_tables()} and \texttt{msprime.simplify\_tables}).
+\texttt{tskit.sort\_tables()} and \texttt{tskit.simplify\_tables}).
 The tables are initialized using the \texttt{TreeSequence} returned from
-\texttt{msprime.simulate($2N$, Ne=$N$, recombination\_rate=$r/2$, length=$n$)}.
+\texttt{tskit.simulate($2N$, Ne=$N$, recombination\_rate=$r/2$, length=$n$)}.
 The \texttt{simuPOP.Population()} is initialized and tagged with individual IDs,
@@
 % Its requirements are encoded in an \texttt{environment.yml} file, compatible
-% with conda, but \msprime{} must be updated to the version listed below via pip
+% with conda, but \tskit{} must be updated to the version listed below via pip
 % from GitHub \jda{TODO add command?}.
@@
 is sorted by time since the parent's birth;
-but the version implemented in \msprime{} makes different sortedness assumptions
+but the version implemented in \tskit{} makes different sortedness assumptions
 (described in the documentation).

### 2018-06-06 review-responses.tex
put back in nedigree

@@
 \reply{
-    We have removed this section.
+    We agree that the discussion of ``nedigree'' is perhaps strictly unnecessary (as we note),
+    but think that small touches like this generally improve the tone 
+    and accessibility of scientific discourse,
+    without obscuring the content.
+    So, unless the AE objects, we'd rather keep it in.
+    (And, we really do think that the field needs a better term for this concept!)
 }

### 2018-06-06 forwards_paper.tex
put back in nedigree

@@
 
-% \paragraph{A final note:}
-% in preparing this manuscript,
-% we debated a number of possible terms for the embellished pedigree,
-% i.e., the ``pedigree with ancestral recombination information''.
-% Etymological consensus \citep{liberman2014little} has
-% ``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
-% (whose branching pattern resembles the bifurcation of a single parent-offspring relationship).
-% An analogous term for the embellished pedigree might then be \emph{nedigree},
-% from ``nid de grue'',
-% as the nest of a crane is a large jumble of (forking) branches.
-% We thought it unwise to use this term throughout the manuscript,
-% but perhaps it will prove useful elsewhere.
+\paragraph{A final note:}
+in preparing this manuscript,
+we debated a number of possible terms for the embellished pedigree,
+i.e., the ``pedigree with ancestral recombination information''.
+Etymological consensus \citep{liberman2014little} has
+``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
+(whose branching pattern resembles the bifurcation of a single parent-offspring relationship).
+An analogous term for the embellished pedigree might then be \emph{nedigree},
+from ``nid de grue'',
+as the nest of a crane is a large jumble of (forking) branches.
+We thought it would be confusing to use this term throughout the manuscript,
+but perhaps it will prove useful elsewhere.
 

### 2018-06-07 forwards_paper.tex
ashander edits

@@
     on which neutral mutations can be quickly placed afterwards.
-    Recording the results of each breeding event
-    results in a linear growth in the amount of information that must be stored.
-    However, there is a great deal of redundancy in this information,
-    since the majority of individuals simulated over the course of the simulation
-    will leave no descendents in the eventual sample.
+    Recording the results of each breeding event requires storage that grows linearly with time,
+    but there is a great deal of redundancy in this information.
     We solve this storage problem by providing an algorithm to quickly `simplify'
@@
         making it feasible to store a simulation's entire final generation
-        (and its history) rather than only a subset.
+        as well as its history.
     (3) A simulation can easily be initialized with a more efficient coalescent simulation 
@@
 We have furthermore provided several tools to minimize this bookkeeping,
-only requiring a little one-way output at the birth of each new individual.
+only requiring one-way output at the birth of each new individual.
 Concretely, to record a tree sequence, including mutations,

### 2018-06-07 forwards_paper.tex
minor adjustments

@@
 The main difference is that events in a coalescent simulation are random, \revpoint{1}{2}
-while here they are predetermined by history.
+while in our simplification algorithm they are predetermined by history.
 An implementation in pseudocode is provided in Appendix~\ref{ss:simplify_algorithm},
@@
 we debated a number of possible terms for the embellished pedigree,
-i.e., the ``pedigree with ancestral recombination information''.
+i.e., the ``pedigree with ancestral recombination information'',
+the object through which each tree of a tree sequence is threaded.
 Etymological consensus \citep{liberman2014little} has

### 2018-06-07 forwards_paper.tex
jerome fixes

@@
     \item and the derived state of these new mutations in the Mutation Table
-        (as well as the identiy of this genome the mutations appeared in).
+        (as well as the identity of this genome the mutations appeared in).
 \end{itemize}
@@
 In this case, the computation is dominated by the size of the output tree sequence,
-which grows with $\log(n)$, as shown in Figure~\ref{fig:simplify_complexity}C,
+which grows with $\log(n)$, as shown in Figure~\ref{fig:simplify_complexity}C.
 
@@
 or the coalescent within a known pedigree.  For such models, one
-could in principle generate tables of nodes and edges to be simplified in \tskit{}.  The resulting succinct tree
-sequence object would be in the same format as those generated by \tskit{}'s built-in simulate function, and therefore
-compatible with existing methods for downstream analyses.
+could in principle generate tables of nodes and edges to be simplified in \tskit{}.  
+The resulting succinct tree sequence object would be in the same format as those generated by \msprime{}'s simulate function,
+and therefore compatible with existing methods for downstream analyses.
 

### 2018-06-07 sims/simplify-alg.py
tweaks for submission

@@ -85,7 +85,7 @@ def simplify(S, Ni, Ei, L):
 
     # Sort the output edges and compact them as much as possible into
     # the output table. We skip this for the algorithm listing as it's pretty mundane.
-    # TODO replace this with a calls to squash_edges() and sort_tables()
+    # Note: could be replaced with calls to squash_edges() and sort_tables()
     E = list(Eo)
     Eo.clear()
     E.sort(key=lambda e: (e.parent, e.child, e.right, e.left))

### 2018-10-12 sims/simplify-alg.py
Updated to use msprime 0.6.0 APIs.

@@ -5,6 +5,8 @@ The example works by generating an initial TreeSequence
 for a sample of 10 haplotypes using msprime.  We then
 simplify the node/edge table in that TreeSequence with
 respect to the first three samples.
+
+Requires msprime >= 0.6.0
 """
 import heapq
 import numpy as np
@@ -37,8 +39,9 @@ def simplify(S, Ni, Ei, L):
     This is an implementation of the simplify algorithm described in Appendix A
     of the paper.
     """
-    No = msprime.NodeTable()
-    Eo = msprime.EdgeTable()
+    tables = msprime.TableCollection(L)
+    No = tables.nodes
+    Eo = tables.edges
     A = [[] for _ in range(len(Ni))]
     Q = []
 
@@ -101,7 +104,7 @@ def simplify(S, Ni, Ei, L):
     j = len(E)
     Eo.add_row(E[start].left, E[j - 1].right, E[j - 1].parent, E[j - 1].child)
 
-    return msprime.load_tables(nodes=No, edges=Eo)
+    return tables.tree_sequence()
 
 
 
@@ -134,7 +137,7 @@ def verify():
 
 
 if __name__ == "__main__":
-    verify()
+    # verify()
 
     # Generate initial TreeSequence
     ts = msprime.simulate(10, recombination_rate=2, random_seed=1)

### 2018-10-12 sims/wf-figures.py

@@ -1,14 +1,17 @@
 """
 Simulation of WF model and output resulting tree sequence.
+
+Requires msprime >= 0.6.0
 """
 import random
 import msprime
+
 
 def wright_fisher(N, delta, L, T):
     """
     Direct implementation of Algorithm W.
     """
-    edges = msprime.EdgeTable()
+    tables = msprime.TableCollection(L)
     tau = []
     P = [j for j in range(N)]
     for j in range(N):
@@ -26,17 +29,16 @@ def wright_fisher(N, delta, L, T):
                 a = random.randint(0, N - 1)
                 b = random.randint(0, N - 1)
                 x = random.uniform(0, L)
-                edges.add_row(0, x, P[a], n)
-                edges.add_row(x, L, P[b], n)
+                tables.edges.add_row(0, x, P[a], n)
+                tables.edges.add_row(x, L, P[b], n)
                 n += 1
             j += 1
         P = Pp
-    nodes = msprime.NodeTable()
     P = set(P)
     for j in range(n):
-        nodes.add_row(time=tau[j], flags=int(j in P))
-    msprime.sort_tables(nodes=nodes, edges=edges)
-    return msprime.load_tables(nodes=nodes, edges=edges)
+        tables.nodes.add_row(time=tau[j], flags=int(j in P))
+    tables.sort()
+    return tables.tree_sequence()
 
 
 
@@ -55,7 +57,5 @@ if __name__ == "__main__":
     # Rescale the height so that the nodes in the new tree are at the
     # same height as those in the old tree.
     h = h * 6 / 7
-    print(t.draw(format="unicode", node_label_text=node_labels))
-    t.draw(path="wf-after.svg", node_label_text=node_labels, width=200, height=h)
-
-
+    print(t.draw(format="unicode", node_labels=node_labels))
+    t.draw(path="wf-after.svg", node_labels=node_labels, width=200, height=h)

### 2018-10-12 forwards_paper.tex
tweaks for submission
Link ftprime in main readme; correct sim setup code and sims/README txt

@@
 \newif\ifsubmission
-% \submissiontrue
-\submissionfalse
+\submissiontrue
+% \submissionfalse
 \ifsubmission
@@
     \newcommand\citep{\cite}
-    \renewcommand\includegraphics{\relax}
-    \renewcommand{\includegraphics}[2][]{\relax}
+    % \renewcommand\includegraphics{\relax}
+    % \renewcommand{\includegraphics}[2][]{\relax}
     \renewcommand{\figurename}{Fig.}
@@
 % Responses to reviews
-\input{review-response-commands}
 % set this to show line numbers and include responses to reviews or not
@@
 % \reviewresponsesfalse  % don't include them
+\input{review-response-commands}
 \newcommand{\responsefile}{review-responses}  % name of the review responses file
@@
 To use the methods with \simupop{}, we created a python module
-(available at \url{https://github.com/ashander/ftprime})
+(available at \url{https://github.com/ashander/ftprime}, version 0.0.6)
 that manages communication between the forward simulator and \tskit{}.

### 2018-10-12 sims/setup-simuPOP.sh

@@ -4,7 +4,7 @@ conda info -a
 if [[ ! -d "/opt/conda/envs/ftprime-benchmark" ]] ; then
 	conda create -q -n ftprime-benchmark python=3.5
 fi
-wget https://raw.githubusercontent.com/ashander/ftprime/0.0.6rc0/environment.yml
+wget https://raw.githubusercontent.com/ashander/ftprime/0.0.6/environment.yml
 conda env update -q -n ftprime-benchmark -f environment.yml
 source activate ftprime-benchmark
 pip install --editable=git+https://github.com/ashander/ftprime.git@master#egg=ftprime

### 2018-10-23 final_format_forwards_paper.tex
final submission version

--- 2018-06-07 forwards_paper.tex
@@
-\documentclass{article}
-\usepackage{fullpage}
-\usepackage{amsmath, amssymb}
+% Template for PLoS
+% Version 3.5 March 2018
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that 
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8x]{inputenc}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
 \usepackage[hidelinks]{hyperref}
-\usepackage[utf8]{inputenc}
-\usepackage{graphicx}
+\usepackage{nameref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
+
+%% Include all macros below
+
 \usepackage{enumitem}
 \usepackage{listings}
-% \usepackage[T1]{fontenc}
-
-\include{macros}
+
+\newcommand{\R}{\mathbb{R}}
+\newcommand{\Q}{\mathbb{Q}}
+\newcommand{\Z}{\mathbb{Z}}
+\newcommand{\N}{\mathbb{N}}
+\newcommand{\C}{\mathbb{C}}
+
+\renewcommand{\P}{\mathbb{P}}
+\newcommand{\E}{\mathbb{E}}
+\newcommand{\var}{\mathop{\mbox{Var}}}
+\newcommand{\cov}{\mathop{\mbox{cov}}}
+%\newcommand{\det}{\mathop{\mbox{det}}}
+\newcommand{\supp}{\mathop{\mbox{supp}}}
+\newcommand{\sgn}{\mathop{\mbox{sgn}}}
+\newcommand{\EE}[1]{\E\!\left[#1\right]}
+\newcommand{\PP}[1]{\P\!\left\{#1\right\}}
+\newcommand{\PPP}[2]{\P_{#1}\!\left\{#2\right\}}
+\newcommand{\EEE}[2]{\E_{#1}\!\left[#2\right]}
+\newcommand{\EEsup}[2]{\E^{#1}\!\left[#2\right]}
+
+\newcommand{\bone}{\mathbf{1}}
+
+% These macros are borrowed from TAOCPMAC.tex
+\newcommand{\slug}{\hbox{\kern1.5pt\vrule width2.5pt height6pt depth1.5pt\kern1.5pt}}
+\def\xskip{\hskip 7pt plus 3pt minus 4pt}
+\newdimen\algindent
+\newif\ifitempar \itempartrue % normally true unless briefly set false
+\def\algindentset#1{\setbox0\hbox{{\bf #1.\kern.25em}}\algindent=\wd0\relax}
+\def\algbegin #1 #2{\algindentset{#21}\alg #1 #2} % when steps all have 1 digit
+\def\aalgbegin #1 #2{\algindentset{#211}\alg #1 #2} % when 10 or more steps
+\def\alg#1(#2). {\medbreak % Usage: \algbegin Algorithm A (algname). This...
+  \noindent{\bf#1}({\it#2\/}).\xskip\ignorespaces}
+\def\kalgstep#1.{\ifitempar\smallskip\noindent\else\itempartrue
+   \hskip-\parindent\fi
+   \hbox to\algindent{\bf\hfil #1.\kern.25em}%
+   \hangindent=\algindent\hangafter=1\ignorespaces}
+
+\newcommand{\algstep}[3]{\kalgstep #1 [#2] #3 }
+\newenvironment{taocpalg}[3]{%
+\vspace{1em}%
+\algbegin Algorithm #1. ({#2}). #3 }
+{\vspace{1em}}
+
+\newcommand{\randomuniform}[0]{\mathcal{R}_U}
+\newcommand{\randomdiscrete}[0]{\mathcal{R}_D}
+\newcommand{\algref}[1]{#1}
+
+% \newcommand{\tablenotation}[1]{\mathcal{#1}}
+% \newcommand{\tableaddrow}[0]{\mbox{add}}
 \newcommand{\simupop}{\texttt{simuPOP}}
@@
 
-%% for submitting with responses to reviews
-\newif\ifsubmission
-% \submissiontrue
-\submissionfalse
-\ifsubmission
 % % stuff for submission
 \usepackage[nofiglist,nomarkers,figuresonly]{endfloat}
-    \usepackage{setspace}
-    % \doublespacing
-    \bibliographystyle{plos2015}
 \newcommand\citet{\cite}
@@
 \renewcommand{\includegraphics}[2][]{\relax}
-    \renewcommand{\figurename}{Fig.}
-    \newcommand{\Figure}{Fig }
-    \newcommand{\Figures}{Fig }
-\else
-    % for the nice version
-    \usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
-    \bibliographystyle{plainnat}
-    \newcommand{\Figure}{{Figure }}
-    \newcommand{\Figures}{{Figures }}
-\fi
-% grrr
-\newcommand{\citetx}[2][]{\ifsubmission#1 \cite{#2}\else\citet{#2}\fi}
-
-% Responses to reviews
-\input{review-response-commands}
-% set this to show line numbers and include responses to reviews or not
-\newif\ifreviewresponses
-\reviewresponsestrue  % include them
-% \reviewresponsesfalse  % don't include them
-\newcommand{\responsefile}{review-responses}  % name of the review responses file
+\newcommand{\revpoint}[2]{}
+
+%% END MACROS SECTION
 
 \begin{document}
-
-\title{Efficient pedigree recording for fast population genetics simulation}
-\author{Jerome Kelleher\footnote{Big Data Institute, University of Oxford},
-        Kevin R. Thornton\footnote{Ecology and Evolutionary Biology, University of California at Irvine},
-        Jaime Ashander\footnote{Ecology and Evolutionary Biology, University of California at Los Angeles}, and
-        Peter L. Ralph\footnote{Institute for Ecology and Evolution, University of Oregon}}
-\maketitle
-
-\begin{abstract}
+\vspace*{0.2in}
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{Efficient pedigree recording for fast population genetics simulation}
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+    Jerome Kelleher\textsuperscript{1},
+    Kevin R.\ Thornton\textsuperscript{2},
+    Jaime Ashander\textsuperscript{3},
+    Peter L.\ Ralph\textsuperscript{4,*},
+\\
+\bigskip
+\textbf{1} 
+    {Big Data Institute, University of Oxford, UK}
+\\
+\textbf{2} 
+    {Ecology and Evolutionary Biology, University of California at Irvine, USA}
+\\
+\textbf{3} 
+    {Ecology and Evolutionary Biology, University of California at Los Angeles, USA}
+\\
+\textbf{4} 
+    {Institute for Ecology and Evolution, University of Oregon, USA}
+\bigskip
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+\textbf{*} \url{plr@uoregon.edu}
+
+\end{flushleft}
+% Please keep the abstract below 300 words
+\section*{Abstract}
     In this paper we describe how to efficiently record the entire genetic history of a population
@@
     The genetic history of the population is recorded as a succinct tree sequence
-    as introduced in the software package \msprime,
+    as introduced in the software package msprime,
     on which neutral mutations can be quickly placed afterwards.
@@
         of deep history.
-    The software for recording and processing tree sequences is named \tskit.
-\end{abstract}
-
-% NOTE: I tend to use semantic linebreaks.
-% Please excuse the ragged right margins in the source.
-
-%%%%%%%%%%%%%%%%%%%%%%
+    The software for recording and processing tree sequences is named tskit.
+
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step. 
+% Author Summary not valid for PLOS ONE submissions.   
 \section*{Author Summary}
@@
 than simulations forced to explicitly carry along that information.
+
+\linenumbers
 
@@
 % When tracking neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for these simulations).
-Pedigree tracking dramatically reduced runtimes, as shown in Figure~\ref{fig:runtimes_selection},
+Pedigree tracking dramatically reduced runtimes, as shown in Fig~\ref{fig:runtimes_selection},
 producing a relative speedup of up to around 50 fold relative to standard simulations that track neutral mutations
-(Figure~\ref{fig:relative_speedup_selection}).
+(Fig~\ref{fig:relative_speedup_selection}).
 Pedigree tracking results in greater relative speedups for larger $N$
 and we observe increasing relative speedups as $4Nr$ increases for a given $N$
-(Figure~\ref{fig:relative_speedup_selection}).
+(Fig~\ref{fig:relative_speedup_selection}).
 Importantly, runtimes are approximately linear in region size $\rho$ when pedigree tracking
-(partially obscured by the log scale of the horizontal axis in Figure~\ref{fig:runtimes_selection}).
+(partially obscured by the log scale of the horizontal axis in Fig~\ref{fig:runtimes_selection}).
 %% don't know what an 'edge' is yet
@@
 In a more limited set of neutral simulations we found the same qualitative behavior,
-and a numerically larger speedup by using pedigree tracking (see Appendix~\ref{ss:timing_nosel}).
+and a numerically larger speedup by using pedigree tracking (see \nameref{ss:timing_nosel}).
 
 In our implementation, simulations with pedigree recording
-used substantially more RAM than simple forward simulations (see Appendix~\ref{ss:memuse}).
+used substantially more RAM than simple forward simulations (see \nameref{ss:memuse}).
 This is unsurprising:
@@
 the amount of required memory is mostly determined by the interval between simplification steps,
-but less frequent simplification reduces overall computation time (see Appendix~\ref{ss:gcinterval}).
+but less frequent simplification reduces overall computation time (see \nameref{ss:gcinterval}).
 In fact, our method could in some situations \emph{reduce} the amount of memory required,
@@
 \begin{figure}
-    \includegraphics[]{sims/rawspeed_logy}
+    % \includegraphics[]{sims/rawspeed_logy}
     \caption{
         \label{fig:runtimes_selection}
-    Total run time per single simulation replicate as a function of region length.
+    \textbf{Total run time per single simulation replicate as a function of region length.}
     % measured as the scaled recombination parameter $\rho = 4Nr$.
@@
 \begin{figure}
-    \includegraphics[]{sims/speedup}
+    % \includegraphics[]{sims/speedup}
     \caption{
         \label{fig:relative_speedup_selection}
-        Relative speedup of simulations due to pedigree recording.
+        \textbf{Relative speedup of simulations due to pedigree recording.}
         Each line shows the ratio of total run times of standard simulations to
         those of simulations with pedigree recording.
-        Data points are taken from Figure~\ref{fig:runtimes_selection}
+        Data points are taken from Fig~\ref{fig:runtimes_selection}
         for simulations that ran to completion in both cases.
@@
 while information about variants is recorded as \emph{sites} and \emph{mutations};
-we give an example in Figure~\ref{fig:example_tree_sequence}.
+we give an example in Fig~\ref{fig:example_tree_sequence}.
 This formulation is derived from the ``coalescence records'' encoding of tree
@@
 corresponding parent, time is measured in clock time, not in meioses.)
-The example of Figure~\ref{fig:example_tree_sequence} has five nodes:
+The example of Fig~\ref{fig:example_tree_sequence} has five nodes:
 nodes 0, 1 and 2 occur at time 0 and are the \emph{samples},
@@
     \begin{center}
-        \includegraphics[width=\textwidth]{example_tree_sequence}
+        % \includegraphics[width=\textwidth]{example_tree_sequence}
     \end{center}
     \caption{
-        An example tree sequence with three samples over a chromosome of length 10.
+        \textbf{An example tree sequence with three samples over a chromosome of length 10.}
         The leftmost panels show the tree sequence pictorially in two different ways:
@@
 of a single branch that occurs in all trees in this interval.
-The spatial extent of the edges defining the topology of Figure~\ref{fig:example_tree_sequence}
+The spatial extent of the edges defining the topology of Fig~\ref{fig:example_tree_sequence}
 are shown in the bottom left panel.
@@
 For example,
-in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
+in Fig~\ref{fig:example_tree_sequence} we have two sites, one at position
 2.5 with ancestral state `A' and the other at position 7.5 with ancestral state `G'.
@@
 unless further mutations are encountered.
-Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
+Three mutations are shown in Fig~\ref{fig:example_tree_sequence},
 illustrated by red stars.
@@
 to \tskit, we also implemented a version of the simulations described in
-``Simulation benchmarks'' separately in Python, described in Appendix~\ref{ss:simupop}.
+``Simulation benchmarks'' separately in Python, described in \nameref{ss:simupop}.
 In this proof-of-concept implementation, the simulation engine (we use \simupop{}, \citet{simupop})
@@
 Benchmarking results from this implementation are shown (alongside results from \fwdpp{})
-for simulations without selection in Appendix~\ref{ss:timing_nosel}:
+for simulations without selection in \nameref{ss:timing_nosel}:
 a relatively modest speedup of around $5 \times$ is achieved, likely due to increased overhead.
@@
     \begin{center}
-        \includegraphics{wf-before-after}
+        % \includegraphics{wf-before-after}
     \end{center}
-    \caption{An example of a marginal genealogy from a Wright-Fisher simulation
-    with $N=5$. \textbf{(A)} the original tree including all
+    \caption{
+        \textbf{An example of a marginal genealogy from a Wright-Fisher simulation with $N=5$.}
+        \textbf{(A)} the original tree including all
     intermediate nodes and dead-ends, and \textbf{(B)} the minimal tree
@@
 
-Figure~\ref{fig:wf-trees} shows
+Fig~\ref{fig:wf-trees} shows
 an example of a marginal genealogy produced by a forwards-time Wright--Fisher
@@
 individuals in the simulation leave no descendants in the present day population.
-Node 26 in Figure~\ref{fig:wf-trees}a, for example, leaves no
+Node 26 in Fig~\ref{fig:wf-trees}a, for example, leaves no
 ancestors in the current population, and so the entire path tracing back to
@@
     \begin{center}
-        \includegraphics{method_diagram}
+        % \includegraphics{method_diagram}
     \end{center}
     \caption{
-        An example of tree sequence simplification.
+        \textbf{An example of tree sequence simplification.}
         \textbf{(A)} The augmented pedigree diagram on the left
@@
 while in our simplification algorithm they are predetermined by history.
-An implementation in pseudocode is provided in Appendix~\ref{ss:simplify_algorithm},
+An implementation in pseudocode is provided in \nameref{ss:simplify_algorithm},
 and a python implementation as supplementary information.
@@
 but is reasonably close.
-Figure~\ref{fig:method_diagram} shows an example tree sequence,
+Fig~\ref{fig:method_diagram} shows an example tree sequence,
 before and after simplification,
-and Figure~\ref{fig:simplify_state} depicts the ``paint pot'' state of the algorithm
+and Fig~\ref{fig:simplify_state} depicts the ``paint pot'' state of the algorithm
 during the process of simplifying this tree sequence.
@@
 and subsequent nodes will be ordered by time since birth.
-This is seen in the red labels of Figure~\ref{fig:method_diagram}.
+This is seen in the red labels of Fig~\ref{fig:method_diagram}.
 
@@
     \begin{center}
-        \includegraphics{simplify-state-diagram}
+        % \includegraphics{simplify-state-diagram}
     \end{center}
     \caption{
-        A depiction of each state of the simplification algorithm
+        \textbf{A depiction of each state of the simplification algorithm}
         as it moves up through the embellished pedigree
-        in the example of Figure~\ref{fig:method_diagram}A.
+        in the example of Fig~\ref{fig:method_diagram}A.
         Following the ``paint pot'' description in the text,
@@
         For instance, both J and K inherit from H between 0.5 and 0.9,
-        which resulted in the first two edges of the simplified table of Figure~\ref{fig:method_diagram}C.
+        which resulted in the first two edges of the simplified table of Fig~\ref{fig:method_diagram}C.
         Later, both inherit from E between 0.2 and 0.5,
@@
 A natural question is now: how often should simplification occur?
-Limited testing (described in Appendix~\ref{ss:gcinterval})
+Limited testing (described in \nameref{ss:gcinterval})
 found that different simplification intervals affect run times by approximately
@@
 %  "that's all very good in practice, but how does it hold up in theory?"
-Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that this method
+Fig \ref{fig:runtimes_selection} and Fig \ref{fig:relative_speedup_selection} show that this method
 can dramatically improve simulation performance in practice --
@@
 The bound holds up reasonably well in practice,
-as shown in Figure~\ref{fig:simplify_complexity}.
+as shown in Fig~\ref{fig:simplify_complexity}.
 
@@
     \begin{center}
-        \includegraphics{sims/simplify-results}
+        % \includegraphics{sims/simplify-results}
     \end{center}
     \caption{
-        Time and space complexity of simplify.
+        \textbf{Time and space complexity of simplify.}
         \textbf{(A)}
@@
         Each line is one simulation, the heavy blue line gives the average,
-        and the dashed line is the upper bound of equation \eqref{eqn:edge_bound}.
+        and the dashed line is the upper bound of Eq~\eqref{eqn:edge_bound}.
         \textbf{(B)}
@@
 Empirically the algorithm is exactly linear,
-as seen in Figure~\ref{fig:simplify_complexity}B,
+as seen in Fig~\ref{fig:simplify_complexity}B,
 which shows the time required to simplify increasingly large subsets of a large tree sequence.
@@
 In this case, the computation is dominated by the size of the output tree sequence,
-which grows with $\log(n)$, as shown in Figure~\ref{fig:simplify_complexity}C.
+which grows with $\log(n)$, as shown in Fig~\ref{fig:simplify_complexity}C.
 
@@
 For instance, the large numbers of large, correlated trees produced by
-MCMC samplers used in Bayesian phylogenetics \citep[e.g.,][]{drummond2012bayesian}
+MCMC samplers used in Bayesian phylogenetics (e.g., \citet{drummond2012bayesian})
 might be compactly stored as a tree sequence,
@@
 pedigree, the distribution of coalescent times in the recent past differs from that of the unconditional coalescent
-\citep{Wakeley2012-kw}.  For populations with known pedigrees \citep[\textit{e.g.}][]{Aguillon2017-ac}, it may be of use to
+\citep{Wakeley2012-kw}.  For populations with known pedigrees (e.g., \citet{Aguillon2017-ac}), it may be of use to
 simulate transmission along such pedigrees for the purpose of inference.
@@
 Code for all simulations and figures is available at \url{https://github.com/petrelharp/ftprime_ms}.
-These made use of the GNU Scientific Library \citep[version 1.16][]{galassi2018scientific},
-pybind11 \citep[version 2.2.1][]{pybind11},
+These made use of the GNU Scientific Library (version 1.16, \citet{galassi2018scientific}),
+pybind11 (version 2.2.1, \citep{pybind11}),
 and GCC (version 4.8.5).
@@
 
+% \bibliography{references}
+
+\begin{thebibliography}{10}
+
+\bibitem{hudson1990gene}
+Hudson RR.
+\newblock Gene genealogies and the coalescent process.
+\newblock Oxford surveys in evolutionary biology. 1990;7(1):44.
+
+\bibitem{Neuhauser1997-nn}
+Neuhauser C, Krone SM.
+\newblock The genealogy of samples in models with selection.
+\newblock Genetics. 1997;145(2):519--534.
+
+\bibitem{barton2010new}
+Barton NH, Kelleher J, Etheridge AM.
+\newblock A new model for extinction and recolonization in two dimensions:
+  quantifying phylogeography.
+\newblock Evolution. 2010;64(9):2701--2715.
+
+\bibitem{kelleher2014coalescent}
+Kelleher J, Etheridge A, Barton N.
+\newblock Coalescent simulation in continuous space: Algorithms for large
+  neighbourhood size.
+\newblock Theoretical population biology. 2014;95:13--23.
+
+\bibitem{ray2010splatche2}
+Ray N, Currat M, Foll M, Excoffier L.
+\newblock {SPLATCHE2}: a spatially explicit simulation framework for complex
+  demography, genetic admixture and recombination.
+\newblock Bioinformatics. 2010;26(23):2993--2994.
+\newblock doi:{10.1093/bioinformatics/btq579}.
+
+\bibitem{martin2017human}
+Martin AR, Gignoux CR, Walters RK, Wojcik GL, Neale BM, Gravel S, et~al.
+\newblock Human demographic history impacts genetic risk prediction across
+  diverse populations.
+\newblock The American Journal of Human Genetics. 2017;100(4):635--649.
+
+\bibitem{wakeley2003gene}
+Wakeley J, Takahashi T.
+\newblock Gene genealogies when the sample size exceeds the effective size of
+  the population.
+\newblock Mol Biol Evol. 2003;20(2):208--213.
+
+\bibitem{maruvka2011recovering}
+Maruvka YE, Shnerb NM, Bar-Yam Y, Wakeley J.
+\newblock Recovering population parameters from a single gene genealogy: an
+  unbiased estimator of the growth rate.
+\newblock Mol Biol Evol. 2011;28(5):1617--1631.
+
+\bibitem{bhaskar2014distortion}
+Bhaskar A, Clark AG, Song YS.
+\newblock Distortion of genealogical properties when the sample is very large.
+\newblock Proc Natl Acad Sci USA. 2014;111(6):2385--2390.
+
+\bibitem{harris2016genetic}
+Harris K, Nielsen R.
+\newblock The Genetic Cost of {Neanderthal} Introgression.
+\newblock Genetics. 2016;203(2):881--891.
+
+\bibitem{slim}
+Messer PW.
+\newblock {SLiM}: simulating evolution with selection and linkage.
+\newblock Genetics. 2013;194(4):1037--1039.
+
+\bibitem{Sanjak2017-ko}
+Sanjak JS, Long AD, Thornton KR.
+\newblock A Model of Compound Heterozygous, {Loss-of-Function} Alleles Is
+  Broadly Consistent with Observations from {Complex-Disease} {GWAS} Datasets.
+\newblock PLoS Genet. 2017;13(1):e1006573.
+
+\bibitem{fwdpp}
+Thornton KR.
+\newblock A {C++} template library for efficient forward-time population
+  genetic simulation of large populations.
+\newblock Genetics. 2014;198(1):157--166.
+
+\bibitem{wright1931evolution}
+Wright S.
+\newblock Evolution in {Mendelian} populations.
+\newblock Genetics. 1931;16(2):97--159.
+
+\bibitem{wakeley2005coalescent}
+Wakeley J.
+\newblock Coalescent Theory, an Introduction.
+\newblock Greenwood Village, CO: Roberts and Company; 2005.
+\newblock Available from: \url{http://www.coalescentheory.com/}.
+
+\bibitem{haller2017flexible}
+Haller BC, Messer PW.
+\newblock {SLiM} 2: Flexible, Interactive Forward Genetic Simulations.
+\newblock Molecular Biology and Evolution. 2017;34(1):230--240.
+\newblock doi:{10.1093/molbev/msw211}.
+
+\bibitem{Hernandez2015-wf}
+Hernandez RD, Uricchio LH.
+\newblock {SFS\_CODE}: More Efficient and Flexible Forward Simulations; 2015.
+
+\bibitem{aberer2013rapid}
+Aberer AJ, Stamatakis A.
+\newblock Rapid forward-in-time simulation at the chromosome and genome level.
+\newblock BMC Bioinformatics. 2013;14(1):216.
+\newblock doi:{10.1186/1471-2105-14-216}.
+
+\bibitem{padhukasahasram2008exploring}
+Padhukasahasram B, Marjoram P, Wall JD, Bustamante CD, Nordborg M.
+\newblock Exploring Population Genetic Models With Recombination Using
+  Efficient Forward-Time Simulations.
+\newblock Genetics. 2008;178(4):2417--2427.
+\newblock doi:{10.1534/genetics.107.085332}.
+
+\bibitem{griffiths1991two}
+Griffiths RC.
+\newblock The two-locus ancestral graph.
+\newblock In: Selected Proceedings of the Sheffield Symposium on Applied
+  Probability. vol.~18; 1991. p. 100--117.
+
+\bibitem{griffiths1997ancestral}
+Griffiths RC, Marjoram P.
+\newblock An ancestral recombination graph.
+\newblock In: Progress in population genetics and human evolution
+  ({M}inneapolis, {MN}, 1994). vol.~87 of IMA Vol. Math. Appl. New York:
+  Springer; 1997. p. 257--270.
+\newblock Available from:
+  \url{http://www.math.canterbury.ac.nz/~r.sainudiin/recomb/ima.pdf}.
+
+\bibitem{wiuf1997number}
+Wiuf C, Hein J.
+\newblock On the number of ancestors to a {DNA} sequence.
+\newblock Genetics. 1997;147(3):1459--1468.
+
+\bibitem{wiuf1999ancestry}
+Wiuf C, Hein J.
+\newblock The ancestry of a sample of sequences subject to recombination.
+\newblock Genetics. 1999;151(3):1217--1228.
+
+\bibitem{marjoram2006coalescent}
+Marjoram P, Wall JD.
+\newblock Fast ``coalescent'' simulation.
+\newblock BMC Genet. 2006;7:16--16.
+\newblock doi:{10.1186/1471-2156-7-16}.
+
+\bibitem{wilton2015smc}
+Wilton PR, Carmi S, Hobolth A.
+\newblock The {SMC'} Is a Highly Accurate Approximation to the Ancestral
+  Recombination Graph.
+\newblock Genetics. 2015;200(1):343--355.
+\newblock doi:{10.1534/genetics.114.173898}.
+
+\bibitem{kelleher2016efficient}
+Kelleher J, Etheridge AM, McVean G.
+\newblock Efficient coalescent simulation and genealogical analysis for large
+  sample sizes.
+\newblock PLoS computational biology. 2016;12(5):e1004842.
+
+\bibitem{gog2014theory}
+Gog S, Beller T, Moffat A, Petri M.
+\newblock From theory to practice: Plug and play with succinct data structures.
+\newblock In: International Symposium on Experimental Algorithms. Springer;
+  2014. p. 326--337.
+
+\bibitem{abadi2006integrating}
+Abadi D, Madden S, Ferreira M.
+\newblock Integrating compression and execution in column-oriented database
+  systems.
+\newblock In: Proceedings of the 2006 ACM SIGMOD international conference on
+  Management of data. ACM; 2006. p. 671--682.
+
+\bibitem{walt2011numpy}
+Walt Svd, Colbert SC, Varoquaux G.
+\newblock The {NumPy} array: a structure for efficient numerical computation.
+\newblock Computing in Science \& Engineering. 2011;13(2):22--30.
+
+\bibitem{hdf5}
+{The HDF Group}. {Hierarchical Data Format, version 5}; 1997-2018.
+
+\bibitem{dask}
+{Dask Development Team}. Dask: Library for dynamic task scheduling; 2016.
+\newblock Available from: \url{http://dask.pydata.org}.
+
+\bibitem{simupop}
+Peng B, Kimmel M.
+\newblock simuPOP: a forward-time population genetics simulation environment.
+\newblock Bioinformatics. 2005;21(18):3686--3687.
+
+\bibitem{hudson1983properties}
+Hudson RR.
+\newblock Properties of a neutral allele model with intragenic recombination.
+\newblock Theor Popul Biol. 1983;23:183--201.
+
+\bibitem{watterson1975number}
+Watterson GA.
+\newblock On the number of segregating sites in genetical models without
+  recombination.
+\newblock Theor Popul Biol. 1975;7(2):256--276.
+
+\bibitem{wilkins2004separation}
+Wilkins JF.
+\newblock A Separation-of-Timescales Approach to the Coalescent in a Continuous
+  Population.
+\newblock Genetics. 2004;168(4):2227--2244.
+\newblock doi:{10.1534/genetics.103.022830}.
+
+\bibitem{barton2002neutral}
+Barton NH, Depaulis F, Etheridge AM.
+\newblock Neutral Evolution in Spatially Continuous Populations.
+\newblock Theoretical Population Biology. 2002;61(1):31--48.
+
+\bibitem{cox2002stepping}
+Cox JT, Durrett R.
+\newblock The stepping stone model: New formulas expose old myths.
+\newblock Ann Appl Probab. 2002;12(4):1348--1377.
+\newblock doi:{10.1214/aoap/1037125866}.
+
+\bibitem{drummond2012bayesian}
+Drummond AJ, Suchard MA, Xie D, Rambaut A.
+\newblock Bayesian phylogenetics with {BEAUti} and the {BEAST} 1.7.
+\newblock Mol Biol Evol. 2012;29(8):1969--1973.
+\newblock doi:{10.1093/molbev/mss075}.
+
+\bibitem{Wiuf2000-rc}
+Wiuf C, Hein J.
+\newblock The Coalescent With Gene Conversion.
+\newblock Genetics. 2000;155(1):451--462.
+
+\bibitem{Kaplan1988-in}
+Kaplan NL, Darden T, Hudson RR.
+\newblock The coalescent process in models with selection.
+\newblock Genetics. 1988;120(3):819--829.
+
+\bibitem{Kaplan1989-rt}
+Kaplan NL, Hudson RR, Langley CH.
+\newblock The ``hitchhiking effect'' revisited.
+\newblock Genetics. 1989;123(4):887--899.
+
+\bibitem{Braverman1995-gn}
+Braverman JM, Hudson RR, Kaplan NL, Langley CH, Stephan W.
+\newblock The hitchhiking effect on the site frequency spectrum of {DNA}
+  polymorphisms.
+\newblock Genetics. 1995;140(2):783--796.
+
+\bibitem{Wakeley2012-kw}
+Wakeley J, King L, Low BS, Ramachandran S.
+\newblock Gene genealogies within a fixed pedigree, and the robustness of
+  Kingman's coalescent.
+\newblock Genetics. 2012;190(4):1433--1445.
+
+\bibitem{Aguillon2017-ac}
+Aguillon SM, Fitzpatrick JW, Bowman R, Schoech SJ, Clark AG, Coop G, et~al.
+\newblock Deconstructing isolation-by-distance: The genomic consequences of
+  limited dispersal.
+\newblock PLoS Genet. 2017;13(8):e1006911.
+
+\bibitem{liberman2014little}
+Liberman A. Little triumphs of etymology: ``pedigree''; 2014.
+\newblock \url{https://blog.oup.com/2014/05/pedigree-etymology-word-origin/}.
+
+\bibitem{galassi2018scientific}
+{Galassi et al} M. GNU Scientific Library Reference Manual; 2018.
+\newblock Available from: \url{https://www.gnu.org/software/gsl/}.
+
+\bibitem{pybind11}
+Jakob W, Rhinelander J, Moldovan D. {pybind11} -- Seamless operability between
+  {C++11} and {Python}; 2016.
+
+\bibitem{Tange2011a}
+Tange O.
+\newblock GNU Parallel - The Command-Line Power Tool.
+\newblock ;login: The USENIX Magazine. 2011;36(1):42--47.
+
+\end{thebibliography}
+
+\section*{Supporting information}
+
+\paragraph*{S1 Text}
+\label{ss:timing_nosel}
+\label{ss:memuse}
+\label{ss:gcinterval}
+\label{ss:simupop}
+\label{ss:simplify_algorithm}
+{\bf Additional benchmarks and algorithm listings.}
+The supplementary text contains 
+(A) benchmarking of run time and memory usage on simulations without selection;
+(B) benchmarking of memory usage with selection;
+(C) an analysis of the effect of simplification interval on run times;
+(D) details for the \simupop{} implementation;
+and (E) more details, and a listing, of the simplification algorithm.
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 Thanks to Gil McVean, Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
-Work on this project was supported by funding from
-the Sloan Foundation and the NSF (under DBI-1262645) to PR;
-the Wellcome Trust (grant 100956/Z/13/Z) to Gil McVean;
-the NIH (R01GM115564) to KRT;
-and the USF\&WS to H.\ Bradley Shaffer.
-
-\bibliography{references}
-
-\appendix
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Timing and memory use with no selection}
-\label{ss:timing_nosel}
-\renewcommand{\thefigure}{D\arabic{figure}}
-\setcounter{figure}{0}
-
-We also performed a more limited set of simulations without natural selection.
-For these we used \cpp{} as in the main text, but also provide results from our
-proof-of-concept implementation that uses \simupop{} (see
-Appendix~\ref{ss:simupop}).
-
-The total run times are shown in
-Figure~\ref{sfig:rawspeed_nosel} and for \cpp{} show the same qualitative behavior as simulations with selection
-(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to pedigree tracking is again substantial in these
-simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to pedigree tracking here with \fwdpy{}
-than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
-close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
-more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
-and~\ref{sfig:speedup_nosel}.
-For \simupop{}, the results show a qualitatively similar pattern but (1) the
-simulations for $N = 10000$ or greater did no finish in the 72 hour limit or
-exceed memory limits (more than 500 GB), and (2) the magnitude of the speedup is much less.
-The total RAM use for
-simulations without selection is shown in Figure~\ref{sfig:ramnosel}.
-% Pedigree recording again used more RAM with \fwdpp{} and less
-% RAM with \simupop{}. The reason for reduced RAM use with \simupop{} is unclear;
-% with \fwdpy{}/\fwdpp{}, however the reason for the extra RAM use is as in
-% Appendix~\ref{ss:memuse}.
-
-\begin{figure}[!h]
-    \includegraphics[]{sims/rawspeed_nosel}
-    \caption{\label{sfig:rawspeed_nosel}Total run time for a single simulation replicate as a function of region
-        length, measured as the scaled recombination parameter $\rho = 4Nr$.  The simulations here involve no natural
-    selection.}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[]{sims/speedup_nosel}
-    \caption{\label{sfig:speedup_nosel}Relative performance improvement due to pedigree tracking for simulations without
-    selection.  Data points are taken from Figure~\ref{sfig:rawspeed_nosel} and show the ratio of run times tracking
-neutral variation over the run times tracking the pedigree.}
-\end{figure}
-\begin{figure}
-    \includegraphics[]{sims/memuse_nosel}
-    \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
-    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Memory usage with selection}
-\label{ss:memuse}
-\renewcommand{\thefigure}{B\arabic{figure}}
-\setcounter{figure}{0}
-
-The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel}.  For our parameters, which involved simplifying every
-$10^3$ generations, simulation while recording pedigrees uses more RAM than a standard simulation.
-This is because with pedigree recording, a non-recombinant
-individual still requires one new edge and two new nodes to be allocated.  Without pedigree recording, no new memory is allocated for such an offspring by \fwdpp{}.
-Thus, pedigree recording accumulates a large amount of data in RAM until
-simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extra RAM needed
-can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
-require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four sets of unsimplified data are
-stored in RAM.  In practice, the RAM consumption can be minimized by simplifying more often.
-
-%% \krt{I've avoided getting into how this much RAM means parallelization is required to make more progress.}
-
-\begin{figure}
-    \includegraphics[]{sims/memuse}
-    \caption{\label{sfig:ramsel}Peak RAM use in gigabytes (GB) for simulations with selection.  The plots are from the
-        same data as in Figure \ref{fig:runtimes_selection}. Note that the y-axis scales differ considerably across
-    panels.}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%
-\section{Effect of simplification interval on run time and memory use}
-\label{ss:gcinterval}
-\renewcommand{\thefigure}{C\arabic{figure}}
-
-We performed a limited set of simulations to explore the effect of the simplification interval on run times and memory
-use. Figure~\ref{sfig:gctime} shows that run times increase for simplification intervals less than 100 generations.
-For intervals longer than 100 generations, run times are very similar, with no appreciable difference between $10^3$ and $10^4$
-generations.  As expected, the simplification interval has a near-linear effect on peak RAM use
-(Figure~\ref{sfig:gcmem}).  Taken together Figures~\ref{sfig:gctime} and~\ref{sfig:gcmem} suggest that one can tune the
-simplification interval to available RAM with little effect on run times provided that simplification doesn't occur too
-often.
-
-\begin{figure}[!h]
-    \includegraphics{sims/gc_interval/GCtime}
-    \caption{\label{sfig:gctime}The effect of simplification interval on run times.  A single replicate was done for the
-        parameters shown in the figure.  These simulations were done with the same selection parameters as in
-    Figure~\ref{fig:runtimes_selection}.}
-\end{figure}
-
-\begin{figure}[!h]
-    \includegraphics{sims/gc_interval/GCmem}
-    \caption{\label{sfig:gcmem}Peak RAM use for the simulations shown in Figure~\ref{sfig:gctime}.}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%
-\section{The \simupop{} implementation}
-\label{ss:simupop}
-
-To use the methods with \simupop{}, we created a python module
-(available at \url{https://github.com/ashander/ftprime})
-that manages communication between the forward simulator and \tskit{}.
-We implemented simulations in a Python script, usable as a command-line utility,
-taking input parameters for population size $N$ and scaled recombination rate $\rho$.
-As for the \fwdpp{} simulations above, we set scaled mutation rate $\theta=4N\mu=\rho$
-(where $\mu$ is the per-site mutation rate multiplied by the number of sites),
-and used a simplify interval of 1000 generations and simulated for $T=10N$ generations in each
-simulation.
-% introducing deleterious mutations at a rate of $\rho/100$ per generation.
-% Because \simupop{} requires explicit locations for loci, we set the number of
-% loci $n=\rho / (4Nr)$ and for these simulations without selection marked $0$ of
-% these as selected loci.
-
-We coupled \simupop{} to \tskit{} using a Python class \texttt{RecombCollector} that 1)
-tracks time, and 2) collects and parses recombination information provided by
-the \texttt{Recombinator} class of \simupop{}, adding them to a \tnodetable{} and
-\tedgetable{} upon which simplification is performed every 1000 generations (using
-\texttt{tskit.sort\_tables()} and \texttt{tskit.simplify\_tables}).
-The tables are initialized using the \texttt{TreeSequence} returned from
-\texttt{tskit.simulate($2N$, Ne=$N$, recombination\_rate=$r/2$, length=$n$)}.
-The \texttt{simuPOP.Population()} is initialized and tagged with individual IDs,
-a mapping of these IDs to the \tnodetable{} IDs is passed to
-an instance of \texttt{RecombCollector} and used internally to update the tables.
-The \texttt{simuPOP.Population()} is then evolved.
-Before mating:
-% selected mutations are created (using
-% \texttt{simuPOP.SNPMutator}) and multiplicative selection occurs (using
-% \texttt{simuPOP.PyMlSelector}) only on the selected loci;
-the internal time of the \texttt{RecombCollector} instance is updated.
-During mating: male and female parents are drawn randomly, recombination data is
-output by \simupop{} to the \texttt{RecombCollector} instance.
-After mating: every 1000 generations the tables are simplified.
-
-\texttt{RecombCollector} uses the Python class \texttt{ARGrecorder} internally
-to add data to the tables and perform simplification.
-This class contains the \tnodetable{} and \tedgetable{} and also the mapping from
-(haploid) ids to ids in the tables.
-It also provides methods (\texttt{add\_individual} and \texttt{add\_record})
-to append data to the tables.
-The only functions of
-\texttt{RecombCollector} not delegated to this internal class are computing
-haploid IDs from \simupop{}'s diploid IDs and parsing recombination data into
-data suitable for use with use with \texttt{add\_individual} and
-\texttt{add\_record}; these functions are used to stream data into the tables as
-it is received by an instance of \texttt{RecombCollector}.
-
-% \paragraph{Requirements.}
-% The simulations as implemented depend on the class \texttt{RecombCollector} provided by
-% the Python package \ftprime{} 0.0.6rc0,
-% available at \url{https://www.github.com/ashander/ftprime}.
-% Its requirements are encoded in an \texttt{environment.yml} file, compatible
-% with conda, but \tskit{} must be updated to the version listed below via pip
-% from GitHub \jda{TODO add command?}.
-%
-% In brief the requirements are:
-%
-% \begin{enumerate}
-%     \item \msprime{} 0.5.0
-%     \item \simupop{} 1.1.8.3
-%     \item python 3.5
-% \end{enumerate}
-
-%%%%%%%%%%%%%%
-\section{Simplification algorithm}
-\label{ss:simplify_algorithm}
-
-Here we provide a concrete implementation of the simplification algorithm.
-The algorithm uses a few simple structures
-besides the node and edge tables discussed earlier. As the algorithm needs to track the mapping
-between input and output node IDs over specific genomic intervals, we use a $\ancsegment$
-type to track these individual mappings. Thus, a segment $x$ has three attributes:
-$\attrleft{x}$ and $\attrright{x}$ represent the genomic interval in the usual way,
-and $\attrnode{x}$ represents the output node ID that is assigned to this interval.
-These output node IDs correspond to colors in the ``paint pot'' analogy.
-The per-interval segments are assigned to input IDs via the $\Al$ mapping,
-which is the main state required by the algorithm. Each element $\Al[u]$
-is a list of ancestral segments, describing the output node IDs mapped to input
-node ID $u$ over a set of intervals.
-For each input node, we need to process the ancestral segments
-to find overlaps (and hence coalescences), and to update $\Al$.
-We do this using a $\priorityq{}$ type, which maintains a list of ancestral
-segments sorted by left coordinate. For a priority queue $Q$, the operation
-$\pqpush{Q}{x}$ inserts the segment $x$ into the queue, and $\pqpop{Q}$
-removes and returns the segment in $Q$ with the smallest left coordinate.
-To obtain the segment currently in the queue with the smallest left value we
-write $Q[0]$.
-Finally, note that to keep the exposition simple,
-the algorithm described here requires that the input node table
-is sorted by time since the parent's birth;
-but the version implemented in \tskit{} makes different sortedness assumptions
-(described in the documentation).
-
-\begin{taocpalg}{S}{Simplify a tree sequence}
-{Input consists of
-    a list $S$ of sample IDs,
-    a list $\Nt_I$ of nodes (ordered by birth time),
-    a list $\Et_I$ of edges,
-    and the genome length, $L$.
-    The output is a
-    nodes $\Nt_O$,
-    and list of edges $\Et_O$.
-}
-
-\algstep{S1.}{Initialisation.}{
-    Set $\Al_u \leftarrow \listnew{}$ for $0 \leq u < |\Nt_I|$. Then,
-    for each $u$ in $S$, set $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ and
-    call $\listappend{\Al[u]}{\ancsegment{0, L, v}}$. Then set $Q \leftarrow \priorityq{}$
-    and $u \leftarrow 0$.
-}
-
-\algstep{S2.}{Select parent edges.}{
-    Set $E \leftarrow \{e \in \Et_I : \attrparent{e} = u \}$.
-    Set $v \leftarrow -1$.  If $|E| = 0$, go to \algref{S9}.
-}
-
-\algstep{S3.}{Insert edge/ancestry intersections.}{
-    For each $e\in E$ and each $x \in \Al[\attrchild{e}]$,
-    if $\attrright{x} > \attrleft{e}$ and $\attrright{e} > \attrleft{x}$,
-    set $\ell \leftarrow \max(\attrleft{x}, \attrleft{e})$,
-    $r \leftarrow \min(\attrright{x}, \attrright{e})$,
-    $y \leftarrow \ancsegment{\ell, r, \attrnode{x}}$ and call
-    $\pqpush{Q}{y}$.
-}
-
-\algstep{S4.}{Find segments with minimum left coordinate.}{
-    Set $\ell \leftarrow \attrleft{Q[0]}$, $r \leftarrow L$ and $X \leftarrow \mbox{List}()$.
-    Then, while $|Q| > 0$ and $\attrleft{Q[0]} = \ell$, set $x \leftarrow \pqpop{Q}$,
-    $r \leftarrow \min(r, \attrright{x})$ and call $\listappend{X}{x}$.
-    Afterwards, if $|Q| > 0$, set $r \leftarrow \min(r, \attrleft{Q[0]})$. If
-    $|X| > 1$ go to \algref{S6}.
-}
-
-\algstep{S5.}{No overlapping segments.}{
-    Set $x \leftarrow X[0]$ and $\alpha \leftarrow x$. If $|Q| > 0$ and
-    $\attrleft{Q[0]} < \attrright{x}$, set $\alpha \leftarrow
-    \ancsegment{\attrleft{x}, \attrright{Q[0]}, \attrnode{x}}$,
-    $\attrleft{x} \leftarrow \attrleft{Q[0]}$ and call $\pqpush{Q}{x}$.
-    Go to \algref{S8}.
-}
-
-\algstep{S6.}{Overlap; new output node.}{
-    If $v = -1$, set $v \leftarrow \taddrow{\Nt_0}{\Nt_I[u]}$.
-}
-
-\algstep{S7.}{Record edges.}{
-    Set $\alpha \leftarrow \ancsegment{\ell, r, v}$. Then for each $x \in X$
-    call $\taddrow{\Et_O}{l, r, v, \attrnode{x}}$ and, if $\attrright{x} > r$,
-    set $\attrleft{x} \leftarrow r$ and call $\pqpush{Q}{x}$.
-}
-
-\algstep{S8.}{Left coordinate loop.}{
-    Call $\listappend{\Al[u]}{\alpha}$. If $|Q| > 0$ return to \algref{S4}.
-}
-
-\algstep{S9.}{Parent loop.}{
-    Set $u \leftarrow u + 1$. If $u < |\Nt_I|$, go to \algref{S2}.
-}
-
-\end{taocpalg}
-
-The algorithm begins in \algref{S1} by allocating an empty list to store ancestral
-segments for each of the $|\Nt_I|$ input nodes, and then creating the initial
-state for the samples in $S$. For each input sample with node ID $u$, we
-add a row in the output node table $\Nt_O$, and create a mapping for this
-new node $v$. After this initial state has been created, we then allocate our
-priority queue $Q$, and set the current input ID $u$ to $0$.
-Recall that the function $\taddrow{\Nt}{t}$ adds a new node to the node table $\Nt$
-with birth time $t$, and returns the ID (i.e., index) of that new node in $\Nt$.
-So, the operation $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ looks up the birth time
-for input node ID $u$ in $\Nt_I$, adds a new node with this time to $\Nt_O$,
-and stores the resulting (output) node ID in $v$.
-
-To process ancestral information moving back up through the pedigree, steps \algref{S2}--\algref{S9}
-consider each node $u$ in turn (note we are assuming that nodes are ordered by time-since-birth).
-In \algref{S2} we first find all edges in the input where the parent is equal to
-our current input ID $u$, and then set the corresponding output ID $v$ to $-1$.
-Then, in \algref{S3} we find all intersections between these input edges and the
-mapped ancestry segments of their child nodes.
-For each of these intersections we
-create a new segment $y$ and insert it into the priority queue.
-(In the paint pot analogy,
-this stores in $Q$ all segments of color that must be transferred to the ancestor
-with input ID $u$.)
-
-After finding all overlaps and filling the priority queue with the resulting ancestry
-segments, we then need to process these segments. Steps \algref{S4}--\algref{S8}
-loop over the segments in the queue, considering each in turn until the queue is empty.
-Step \algref{S4} builds a list $X$ of all segments covering the current left-most coordinate
-(along with some bookkeeping to keep track of where our next
-right coordinate $r$ is), removing them from $Q$. If there is only one element in $X$
-this means that we have no overlapping ancestral segments at this coordinate and
-we proceed to \algref{S5}.
-
-In this case, only a single segment covers the leftmost region of the segments,
-so there is no change in the mapping between input and output IDs,
-and the ancestry segment $\alpha$ is not changed (this is assigned to $\Al_u$ in \algref{S8}).
-(The color in this segment of the offspring can simply be transferred to the parent.)
-However, if other segments in $Q$ intersect with $x$
-(since $x$ is the current left-most segment, these have left coordinate less then $\attrright{x}$),
-then we can only transfer the portion of $x$ up until it overlaps with other segments.
-Therefore, we change the left coordinate
-of $x$ to this coordinate and reinsert it into $Q$ where it will be processed again
-later, effectively cutting off (and propagating) the leftmost segment of $x$.
-After doing this, we proceed to \algref{S8}, update the ancestry mapping for
-the input node $u$, and loop back to \algref{S4} if $Q$ is not empty.
-
-On the other hand, if there is more than one segment in $X$, we know that an
-overlap among the ancestral segments has occurred and we must update the output
-nodes and edges accordingly. In step \algref{S6} we first allocate a new output
-node $v$ for input node $u$, if it has not been done already (we may have many overlapping
-segments along the genome). Step \algref{S7} then continues by creating a new
-ancestry segment $\alpha$ mapping the current interval to the output node $v$.
-We then iterate over all segments in $X$, adding the appropriate output edges
-and editing and reinserting the ancestry segment $x$ into the priority queue if
-necessary, as before.
-
-% \newpage
-\newpage
-
-\section*{Python implementation of simplify (supplementary information)}
-
-\lstinputlisting[language=python]{sims/simplify-alg.py}
-
-\ifsubmission\processdelayedfloats\fi
-
-% reviews here if toggled above
-\includereviews
+% Work on this project was supported by funding from
+% the Sloan Foundation and the NSF (under DBI-1262645) to PR;
+% the Wellcome Trust (grant 100956/Z/13/Z) to Gil McVean;
+% the NIH (R01GM115564) to KRT;
+% and the USF\&WS to H.\ Bradley Shaffer.
 

### 2019-08-09 talks/poster/probgen-2018-poster.tex
added slides

--- 2018-04-12 talks/poster/evo-wibo-2018-poster.tex
@@
   \vspace{0.5em}
-     \textbf{\textcolor{DavisBlue}{Recording pedigrees for fast genome simulation and analysis}}\vspace{0.5em}}
+     \textbf{\textcolor{DavisBlue}{Tree sequences \\ for fast genome simulation and analysis}}\vspace{0.5em}}
   % Authors
@@
     {\textbf{code:}  \url{https://github.com/tskit-dev} }\\
-    {\textbf{preprint:}  \url{https://www.biorxiv.org/content/early/2018/01/26/248500} }\\
+    \vspace{0.8em}
+    {Related posters: Jerome Kelleher (tree sequences) and Phillip Messer (SLiM)}
+    % {\textbf{paper:}  \url{https://www.biorxiv.org/content/early/2018/01/26/248500} }\\
   }
@@
         \hfill
-          \includegraphics[width=12em]{UOSignature-STK-BLK} \\
+          \includegraphics[width=10em]{UOSignature-STK-BLK} \\
       \end{minipage}
@@
 
-For a set of sampled chromosomes,
-at each position along the genome there is a genealogical tree
-that says how they are related.
-
-A \textbf{tree sequence} describes this sequence of trees.
+A \textbf{tree sequence} describes
+the correlated sequence of genealogical trees
+that say how a set of chromosomes are related.
 
@@
 
-    \item The \emph{pedigree} (parental relationships) plus crossover locations
+    \item The \emph{pedigree} plus crossover locations
         would give us the tree sequence for \emph{everyone, ever}.
@@
 
-    \item This information is equivalent to the Ancestral Recombination Graph (ARG).
+    \item This is equivalent to the Ancestral Recombination Graph (ARG).
 
@@
 
-\citep{kelleher2016efficient}
+\citet{kelleher2016efficient}
 introduced the \textbf{tree sequence} data structure
 for a fast coalescent simulator, \texttt{msprime}.
+Updated in \citet{kelleher2018efficient}.
 
@@
 
+    % We updated and generalized the data structure in \citet{kelleher2018efficient}:
+
     \begin{itemize}
 
-        \item \textbf{Edges:}   Who inherits from who; only \emph{necessary} for coalescent events.
+        \item \textbf{Edges:}   Who inherits from who. \\
             \emph{Records:} interval (left, right); parent node; child node.
 
-        \item \textbf{Nodes:}   The ancestors those happen in.
-            \emph{Records:} time ago (of birth); ID (implicit).
+        \item \textbf{Nodes:}   The ancestors those happen in. \\
+            \emph{Records:} time ago (of birth); individual.
 
-        \item \textbf{Mutations:}   When state changes along the tree.
-            \emph{Records:} site it occurred at; node it occurred in; derived state.
+        \item \textbf{Mutations:}   When state changes along the tree. \\
+            \emph{Records:} site index; node index; derived state.
 
-        \item \textbf{Sites:}   Where mutations fall on the genome.
-            \emph{Records:} genomic position; ancestral (root) state; ID (implicit).
+        \item \textbf{Sites:}   Where mutations fall on the genome. \\
+            \emph{Records:} genomic position; root state.
+
+        \item \textbf{Individuals:} Optional. Containers for polyploids. \\
+            \emph{Records:} metadata; pointed to by nodes.
 
@@
     \begin{center}
-        \includegraphics[width=0.75\textwidth]{../file_size}
+        \includegraphics[width=0.75\textwidth]{storing_everyone}
     \end{center}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\headerbox{Contributions}{name=contrib,column=1,row=0}{
+\headerbox{References}{name=contrib,column=1,row=0}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    \begin{itemize}
-        \compresslist
-    \item Jerome Kelleher -- algorithms, data structures, and core development
-    \item Kevin Thornton -- implementation, profiling with \texttt{fwdpp},
-    \item Jaime Ashander -- implementation, profiling with \texttt{simuPOP}
-    \item Ben Haller -- implementation in \texttt{SLiM}
-    \item Jared Galloway -- implementation in \texttt{SLiM}
-    \end{itemize}
-
+  % \scriptsize
+  \renewcommand{\section}[2]{\vskip 0.0em}
+  \bibliographystyle{abbrvnat}
+  \setlength{\bibsep}{0.0pt}
+  \bibliography{refs}
 }
@@
 
-\includegraphics[width=7em]{../finger_right} {\Large \emph{This may be a bad idea.}}
-
 }
@@
 
-        \item add each gamete to the Node Table,
-        \item add entries to the Edge Table
-            recording which parent each gamete inherited each bit of genome from, and
-        \item add any new selected mutations to the Mutation Table 
+        \item Add new individuals and their chromosomes to the Individual and Node Tables, respectively.
+        \item Record in the Edge Table 
+            which parental chromosome each gamete inherited each bit of genome from.
+        \item Add any new selected mutations to the Mutation Table 
             and (if necessary) their locations to the Site Table.
@@
 
-\textbf{But,} we won't end up needing the entire history of everyone ever.
+\includegraphics[width=7em]{../finger_right} 
+\parbox[bottom]{20em}{
+    \vspace{-4em}
+    {\Large \emph{This is too much data.}} \\
+    \emph{\ldots but,} most of it can be \emph{simplified} away.
+}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\headerbox{Simulations, $50\times$ faster}{name=fast,column=1,below=howfwd}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    Recording, simplifying, and output of tables, is done with \texttt{C} code in \texttt{tskit}.
+
+    % The simulation was done in \texttt{fwdpp} (C++ by Kevin Thornton),
+    % connected with \texttt{pybind11} and \texttt{numpy}.
+    % \textit{Machine:} Ubuntu / 2x 2.6 GHz Intel E5-2650 CPU.
+
+    % \emph{Simulation parameters:}
+    % \begin{itemize}
+    %         \compresslist
+    %     \item  Wright-Fisher population of size $N$, for $10N$ generations
+    %     \item  neutral mutation rate $\mu$ equal to recombination rate $r$ per gamete
+    %     \item  many, weakly deleterious mutations: rate $\mu/100$ with
+    %         $s$ exponentially distributed with mean $2.5/N$.
+    % \end{itemize}
+
+    \begin{center}
+    % \includegraphics[width=\textwidth]{rawspeed}
+
+    \parbox[bottom]{0.49\textwidth}{
+        \includegraphics[width=0.45\textwidth]{speedup}
+        (above) many selected mutations;
+        (right) neutral.
+    }
+    \parbox[bottom]{0.49\textwidth}{
+        \includegraphics[width=0.45\textwidth]{slim_speed}
+    }
+    \end{center}
+
+    Simulations that recorded tree sequences
+    had neutral mutation rate set to zero during the simulation,
+    and neutral mutations added \emph{afterwards}.
+
+    \vspace{2em}
+
+    Results in \citet{haller2018treesequence,kelleher2018efficient}.
+
+    \vspace{2em}
+
+    {\Large
+    Now in SLiM and fwdpp!
+    }
+
+}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\headerbox{A general class of statistics}{name=stats,column=2,row=0}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+        Every single-site statistic is a function from genotype patterns to $\mathbb{R}$,
+        and each SNP genotype pattern is determined by 
+            the samples below the edge it occurs on.
+            \vspace{1em}
+
+    \emph{Ingredients:}
+    \begin{enumerate}
+            \compresslist
+        \item A list of $k$ subsets of samples, $I_1, \ldots, I_k$,
+            with $|I_j| = N_j$.
+        \item A weighting function $f : \mathbb{N}^k \to \mathbb{R}$.
+    \end{enumerate}
+
+    \emph{Computation} on a genome of length $L$ with with $M$ SNPs:
+    \begin{enumerate}
+            \compresslist
+        \item Let $n_i(j)$ be the number of samples in $I_j$
+            that inherit mutation $i$.
+        \item Use the weighting function to compute the weight of each mutation;
+            average these across the genome:
+            \[
+                S(w) := \dfrac{1}{L} \sum_{i=1}^M (w(n_1, \ldots, n_k)  
+                                        + w(N_1 - n_1, \ldots, N_k - n_k) .
+            \]
+    \end{enumerate}
+
+}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\headerbox{Examples}{name=examples,column=2,below=stats}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    Folded SFS:
+        $$ f_\ell(n_1) = \textbf{1}_\ell(n_1) \qquad \text{for}\; 1 \le \ell \le N_1 - 1 .$$
+
+    Divergence between $I_1$ and $I_2$:
+        $$ w(n_1, n_2) = n_1 (N_2 - n_2) / N_1 N_2 .$$
+
+    Patterson's $f_4(I_1, I_2; I_3, I_4)$:
+        \begin{align*}
+           &w(n_1, n_2, n_3, n_4) \\
+            & \qquad =
+            \frac{
+               n_1 n_3 (N_2 - n_2) (N_4 - n_4)
+                      - n_1 n_4 (N_2 - n_2) (N_3 - n_3) }{ N_1 N_2 N_3 N_4  }
+              .  \end{align*}
+
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\headerbox{Computation}{name=computation,column=2,below=examples}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    \begin{enumerate}
+            \compresslist
+        \item[0.]
+            Find $n$ for each node in the first tree.
+        \item[1.]
+            Compute $f(n)$ for each mutation on this tree,
+            and add these to the total.
+        \item[2.]
+            Update the tree, and update $n$ for each node
+            in the path from changed nodes to the root.
+        \item[3.]
+            Return to (1).
+        \item[4.]
+            When done, divide the total by the sequence length.
+    \end{enumerate}
+
+    \emph{Complexity} for $N$ samples at $L$ SNPs with $T$ trees:
+        $$ O(N + L + T \log(N)) . $$
+    This is much better than using the genotype matrix, which is $O(NL)$.
+
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\headerbox{Other notes}{name=other,column=2,below=computation}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+    \begin{itemize}
+        \item 
+            The \emph{same} statistics can be computed 
+            as weighted averages of \textbf{branch lengths} also,
+            which gives the expected value of the statistic, averaged over mutational noise.
+
+        \item 
+            This tracks who is ancestral to a set of samples;
+            \emph{a dual procedure} tracks who has inherited from a set of ancestors.
+
+        \item
+            \emph{Many statistics} can be computed on the same pass through the tree sequence
+            with little additional cost.
+
+        \item
+            The implementation computes these in \emph{windows}.
+
+    \end{itemize}
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\headerbox{Updating internal state}{name=statfig,column=2,below=other}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    \includegraphics[width=0.3\textwidth]{stats_1}
+    \includegraphics[width=0.3\textwidth]{stats_2}
+    \includegraphics[width=0.3\textwidth]{stats_3}
+
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\headerbox{}{name=trees,column=0,above=bottom,span=3}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    \includegraphics[width=0.085\textwidth]{tree0}
+    \includegraphics[width=0.085\textwidth]{tree1}
+    \includegraphics[width=0.085\textwidth]{tree2}
+    \includegraphics[width=0.085\textwidth]{tree3}
+    \includegraphics[width=0.085\textwidth]{tree4}
+    \includegraphics[width=0.085\textwidth]{tree5}
+    \includegraphics[width=0.085\textwidth]{tree6}
+    \includegraphics[width=0.085\textwidth]{tree7}
+    \includegraphics[width=0.085\textwidth]{tree8}
+    \includegraphics[width=0.085\textwidth]{tree9}
+    \includegraphics[width=0.085\textwidth]{tree10}
+
+}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\end{poster}%
+\end{document}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\headerbox{Simulations, $50\times$ faster}{name=fast,column=2,row=0}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    Recording, simplifying, and output of tables, was done with \texttt{C} code in \texttt{tskit}.
-    The simulation was done in \texttt{fwdpp} (C++ by Kevin Thornton),
-    connected with \texttt{pybind11} and \texttt{numpy}.
-    \textit{Machine:} Ubuntu / 2x 2.6 GHz Intel E5-2650 CPU.
-
-    \emph{Simulation parameters:}
-    \begin{itemize}
-            \compresslist
-        \item  Wright-Fisher population of size $N$, for $10N$ generations
-        \item  neutral mutation rate $\mu$ equal to recombination rate $r$ per gamete
-        \item  many, weakly deleterious mutations: rate $\mu/100$ with
-            $s$ exponentially distributed with mean $2.5/N$.
-    \end{itemize}
-
-    \emph{Note:}
-    simulations that recorded tree sequences ("pedigree recording")
-    had neutral mutation rate was zero,
-    but neutral mutations were added \emph{afterwards}.
-
-    \begin{center}
-    \includegraphics[width=\textwidth]{rawspeed}
-
-    \includegraphics[width=0.75\textwidth]{speedup}
-    \end{center}
-
-}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\headerbox{Fast statistics computation}{name=stats,column=2,below=fast}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    \textbf{A general class of statistics:}
-    any polynomial $f(p_1, \ldots, p_k)$ gives a statistic on $k$ populations,
-    by setting $p_i$ to the allele frequency in population $i$ and averaging $f$ across loci.
-    \emph{(Example: divergence is $p_1(1-p-2) + p-2(1-p_1)$.)}
-
-    The polynomial defines a weighting function on edges of the trees in the tree sequence,
-    and the statistic can be calculated by summing across mutations
-    weighted by the edge they occur on.
-    \emph{(Example: proportion of pairs from the two populations on opposite sides of the edge.)}
-
-    \textbf{Computation on a tree sequence:}
-    If we compute weights on the first tree, then we only need update those weights above nodes that change
-    in the second tree.
-
-}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \headerbox{Summary}{name=summary,column=2,below=stats}{
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\headerbox{}{name=references,column=2,below=summary}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-  \scriptsize
-  \renewcommand{\section}[2]{\vskip 0.0em}
-  \bibliographystyle{abbrvnat}
-  \setlength{\bibsep}{0.0pt}
-  \bibliography{../../references}
-  }
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\headerbox{}{name=trees,column=0,above=bottom,span=3}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    \includegraphics[width=0.085\textwidth]{tree0}
-    \includegraphics[width=0.085\textwidth]{tree1}
-    \includegraphics[width=0.085\textwidth]{tree2}
-    \includegraphics[width=0.085\textwidth]{tree3}
-    \includegraphics[width=0.085\textwidth]{tree4}
-    \includegraphics[width=0.085\textwidth]{tree5}
-    \includegraphics[width=0.085\textwidth]{tree6}
-    \includegraphics[width=0.085\textwidth]{tree7}
-    \includegraphics[width=0.085\textwidth]{tree8}
-    \includegraphics[width=0.085\textwidth]{tree9}
-    \includegraphics[width=0.085\textwidth]{tree10}
-
-}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\end{poster}%
-\end{document}

### 2019-08-09 talks/export-layers-svg.sh

@@ -26,4 +26,4 @@ done
 
 inkscape --without-gui --export-pdf=/dev/stdout $TMPFILE
 
-# [ -f $TMPFILE ] && rm $TMPFILE
+[ -f $TMPFILE ] && rm $TMPFILE
